### 2017-01-11 pt_characterization/data.old/paper.tex
pt char paper

\documentclass[10pt]{article}
\usepackage[margin=1in]{geometry}
 \usepackage{auto-pst-pdf}
\usepackage{graphicx}
%\usepackage{arydshln}
\usepackage{ifpdf}
\ifpdf
  \usepackage{epstopdf}
\fi
\usepackage{multirow}
\usepackage{epsfig}
\usepackage{float}
\usepackage{url}
\usepackage{color}
\usepackage{subfigure}

%\newcommand\solidrule[1][1cm]{\rule[0.5ex]{#1}{.4pt}}
%\newcommand\dashedrule{\mbox{%
%  \solidrule[2mm]\hspace{2mm}\solidrule[2mm]\hspace{2mm}\solidrule[2mm]}}

%\usepackage{hyperref}

\begin{document}
\title{A Comprehensive Study of Characterizing Program Execution Time}

\author{
Young-Kyoon Suh\\
%Department of Computer Science \\
%University of Arizona \\
%\small\tt yksuh@cs.arizona.edu
}
\maketitle

\begin{abstract}
(Tentative) Measuring execution time is a useful tool in evaluating the performance of a program. 
But it is challenging to obtain precise, accurate program execution time due to the presence of various system daemons 
and their unpredictable activities. Such activities significantly contribute to varying program execution time.
It will be very useful to predict the concrete performance of a program with different input sizes in a real situation, 
if a (probabilistic) distribution of the execution times of that program is found, considering the daemon activities.
%However, it is not easy to find such a characterization of the execution times. 
In this work, we discuss several interesting phenomenon observed to characterize program execution times.
We find that such a distribution of the execution times cannot be uniquely identified, and it will 
be more likely to be mixture of two or more models formed by different periodicity of different daemon processes. 
Finally, we discuss some remaining issues that should be resolved to successfully 
identify a distribution of program execution times. 
\end{abstract}

\section{Experiment Notes}
%% done
\begin{table}[h]
\begin{center}
\begin{tabular}{|p{3cm}||p{9cm}|p{4cm}|} \hline
Task Length & Description & Time Length\\ \cline{1-3}
\multicolumn{3}{|c|}{Regular PUT experiment. Refer to Sections~\ref{sec:emp4_summary},~\ref{sec:empv4_hist}, and~\ref{sec:empv5_hist}.} \\ \hline
PUT1$\sim$PUT64 & Runs of 1000 samples (on {\tt sodb12}). & 2013-10-14 $\sim$ 2013-10-15\\ \hline
PUT128$\sim$PUT2048 & Runs of 300 samples (on {\tt sodb12}). & 2013-12-12 $\sim$ 2013-12-21\\ \hline 
PUT4096 & A run of 300 samples (on {\tt sodb12}). & 2014-06-23 $\sim$ 2014-07-10 \\ \hline \hline
PUT8192 & Runs of 40/260 samples (on {\tt sodb12}). & 2015-04-23 $\sim$ 2015-04-27 / 2015-10-31 $\sim$ 2015-11-24\\ \hline
PUT16384 & Runs of 40/260 samples (on {\tt sodb12}). & 2015-04-23 $\sim$ 2015-04-23 / 2015-11-25 $\sim$ 2016-01-14\\ \hline 
\end{tabular}
\end{center}
\vspace{-.2in}
\caption{Notes on the regular PUT data used for the histograms\label{tab:exp_notes1}}
\end{table}
\begin{table}[H]
\begin{center}
\begin{tabular}{|p{3cm}||p{9cm}|p{4cm}|} \hline
Task Length & Description & Time Length\\ \cline{1-3}
\multicolumn{3}{|c|}{Regular PUT experiment. Refer to Section~\ref{sec:new_put}.} \\ \hline
PUT1/2 & Runs of 20k samples on {\tt sodb9}/{\tt sodb10}. & 2015-12-15 $\sim$ 2015-12-15 \\ \hline
PUT4/8 & Runs of 20k samples on {\tt sodb10}. & 2016-01-20 $\sim$ 2016-01-20  \\ \hline 
PUT16 & Runs of 2k, 4k, 8k, 16k, and 32k samples (on {\tt sodb12}). & 2016-01-25 $\sim$ 2016-02-09\\ \cline{1-3}
%PUT4096 & A dual PUT experiment with a run of 1,000 samples on {\tt sodb8}. But the dual data in the second half were missing due to a glitch in the code, which was fixed. The dual data in the first half were available for histogram. Used {\tt gettimeofday()} for measuring the elapsed time of each half of every PUT4096 (same for the following PUT2 and PUT64). & 2015-11-08 $\sim$ 2015-12-25 (roughly)\\ \hline 
\multicolumn{3}{|c|}{Dual PUT experiment. Refer to Section~\ref{sec:dual_put}.} \\ \hline
PUT4096 & A run of 500 samples on {\tt sodb8}. %Used {\tt gettimeofday()} for measuring the elapsed time of each half of every PUT4096 (same for the following PUT2 and PUT64).} 
& 2015-11-08 $\sim$ 2015-12-25 \\ \hline 
PUT2 & A run of 1k samples on {\tt sodb9}. & 2015-12-27 $\sim$ 2015-12-27 \\ \hline
%PUT64 & A run of 1k samples on {\tt sodb10} & 2015-12-27 $\sim$ 2015-12-27 \\ \hline 
PUT4$\sim$PUT32 & Runs of 1k samples on {\tt sodb9}. & 2016-01-27 $\sim$ 2016-01-31 \\ \hline
PUT64$\sim$PUT2048& Runs of 1k samples on {\tt sodb9}. & 2016-02-17 $\sim$ 2016-04-05 \\ \hline
PUT2048 & A run of 100 samples on {\tt sodb9}. & 2016-04-13 $\sim$ 2016-04-16 \\ \hline
PUT4096 & A run of 100 samples on {\tt sodb8}. & 2016-04-13 $\sim$ 2016-04-19 \\ \hline
\end{tabular}
\end{center}
\vspace{-.2in}
\caption{Notes on the new PUT experiments\label{tab:exp_notes3}}
\end{table}

\clearpage
\newpage

\section{Summary of the EMPv4 data~\label{sec:emp4_summary}}
EMPv4: Running PUT with a specific task length under a controlled environment, 
with i) daemon processes disabled, 
ii) the NTP daemon process activated,
iii) major CPU features (turbo and speedstep) disabled, and
iv) an up-to-date Linux version (RHEL 6.0) installed.

\begin{table}[h]
\centering
{
 \begin{tabular}{|l|c|c|c|c|c|c|} \hline
\multirow{2}{*}{}   & Num. of Samples & Minimum & Maximum & Average & Std. Dev. \\ 
                        & & (msec)  & (msec)  & (msec)  & (msec) \\ \hline
 PUT1  & 1,000 & 999.0 & 1,005.0 & 1,002.4 & 0.73\\ \hline

% PUT1 & 300 & 999.0 & 1,002.0 & 1,005.1 & 0.85 \\ \hline \hline

 PUT2 & 1,000 & 1,996.0 & 2,007.0 & 2,004.5 & 1.38 \\\hline

% PUT2 & 300 & 1,998.0 & 2,007.0 & 2,005.2 & 1.27 \\ \hline \hline

 PUT4 & 1,000 & 4,004.0 & 4,012.0 & 4,008.6 & 1.64\\\hline

% PUT4 & 300 & 4,006.0 & 4,012.0 & 4,009.1 & 1.52\\ \hline \hline

 PUT8 & 1,000 & 8,014.0 & 8,023.0 & 8,018.1 & 1.72 \\\hline

% PUT8 & 300 & 8,014.0 & 8,023.0 & 8,018.3 & 1.81 \\ \hline \hline

 PUT16 & 1,000 & 16,029.0 & 16,041.0 & 16,034.3 & 1.86 \\ \hline

% PUT16 & 300 & 16,029.0 & 16,041.0 & 16,034.1 & 1.86 \\ \hline \hline

 PUT32 & 1,000 & 32,064.0 & 32,084.0 & 32,068.2 & 2.05 \\ \hline

% PUT32 & 300 & 32,065.0 & 32,084.0 & 32,068.4 & 1.93 \\ \hline \hline

 PUT64 & 1,000 & 64,129.0 & 64,145.0 & 64,135.0 & 2.27 \\ \hline

% PUT64 & 300 & 64,129.0 & 64,145.0 & 64,135.2 & 2.25 \\ \hline \hline

 PUT128 & 300 & 128,244.0 & 128,260.0 & 128,251.2 & 2.32\\ \hline

 PUT256 & 300 & 256,494.0 & 256,523.0 & 256,502.3 & 3.29\\ \hline

 PUT512 & 300 & 512,995.0 & 513,152.0 & 513,005.1 & 9.41\\ \hline

 PUT1024 & 300 & 1,025,997.0 & 1,026,141.0 & 1,026,012.4 & 11.43\\ \hline

 PUT2048 & 300 & 2,051,981.0 & 2,052,156.0 & 2,052,012.0 & 11.19\\ \hline

 PUT4096 & 300 & 4,105,451.0 & 4,105,629.0 & 4,105,526.0 & 25.98\\ \hline

 PUT8192 & 40 (last Apr) & 8,207,870.0 & 8,207,967.0 & 8,207,918.0 & 21.03\\ \hline

 PUT8192 & 260 (Nov) & 8,210,940.0 & 8,211,196.0 & 8,211,049.0 & 36.60\\ \hline

 PUT16384 & 40 (last Apr) & 16,415,757.0 & 16,415,964.0 & 16,415,810.3 & 40.43\\ \hline

 PUT16384 & 260 (Nov) & 16,422,028 & 16,422,389 & 16,422,153.0 & 52.54\\ \hline
  \end{tabular}
  }

\caption{PT statistics by EMPv4 (See Table~\ref{tab:exp_notes1}.)\label{tab:empv4_stat}}
\end{table}

\begin{figure}[htp!]
	\centering
	\includegraphics[scale=0.8]{overall_pt_std.eps}
	\caption{Std. dev. of PT over increasing task length~ (See Table~\ref{tab:exp_notes1}.)\label{fig:incr_task_len}}
\end{figure}

\newpage

\section{Histograms on the EMPv4 Data~\label{sec:empv4_hist}}
The base data of the following histograms are from Table~\ref{tab:exp_notes1}.

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT1]{
		\includegraphics[scale=0.43]{1_sec_pt_hist.eps}
		\label{fig:put1_hist}
	}
	\subfigure[PT frequency on PUT2]{
		\includegraphics[scale=0.43]{2_sec_pt_hist.eps}
		\label{fig:put2_hist}
	}
	\subfigure[PT frequency on PUT4]{
		\includegraphics[scale=0.43]{4_sec_pt_hist.eps}
		\label{fig:put4_hist}
	}
	\subfigure[PT frequency on PUT8]{
		\includegraphics[scale=0.43]{8_sec_pt_hist.eps}
		\label{fig:put8_hist}
	}
	\caption{PT Histograms of PUT1 ... PUT8~\label{fig:pt_hist1}}
\end{figure}

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT16]{
		\includegraphics[scale=0.43]{16_sec_pt_hist.eps}
		\label{fig:put16_hist}
	}
	\subfigure[PT frequency on PUT32]{
		\includegraphics[scale=0.43]{32_sec_pt_hist.eps}
		\label{fig:put32_hist}
	}
	\subfigure[PT frequency on PUT64]{
		\includegraphics[scale=0.43]{64_sec_pt_hist.eps}
		\label{fig:put64_hist}
	}
	\caption{PT Histograms of PUT16 ... PUT64~\label{fig:pt_hist2}}
\end{figure}

\newpage

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT128]{
		\includegraphics[scale=0.43]{128_sec_pt_hist.eps}
		\label{fig:put128_hist}
	}
	\subfigure[PT frequency on PUT256]{
		\includegraphics[scale=0.43]{256_sec_pt_hist.eps}
		\label{fig:put256_hist}
	}
	\subfigure[PT frequency on PUT512]{
		\includegraphics[scale=0.43]{512_sec_pt_hist.eps}
		\label{fig:put512_hist}
	}
	\subfigure[PT frequency on PUT1024]{
		\includegraphics[scale=0.43]{1024_sec_pt_hist.eps}
		\label{fig:put1024_hist}
	}
	\caption{PT Histograms of PUT128 ... PUT1024~\label{fig:pt_hist3}}
\end{figure}

\newpage

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT2048]{
		\includegraphics[scale=0.43]{2048_sec_pt_hist.eps}
		\label{fig:put2048_hist}
	}
	\subfigure[PT frequency on PUT4096]{
		\includegraphics[scale=0.43]{4096_sec_pt_hist.eps}
		\label{fig:put4096_hist}
	}
	\caption{PT Histograms of PUT2048 and PUT4096~\label{fig:pt_hist4}}
\end{figure}

\newpage

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT8192 with 40 samples  (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{8192_sec_pt_hist1.eps}
		\label{fig:put8192_hist1}
	}
	\subfigure[PT frequency on PUT8192 with 260 samples (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{8192_sec_pt_hist2.eps}
		\label{fig:put8192_hist2}
	}
	\subfigure[PT frequency on PUT16384 with 40 samples (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{16384_sec_pt_hist1.eps}
		\label{fig:put16384_hist1}
	}
	\subfigure[PT frequency on PUT16384 with 260 samples (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{16384_sec_pt_hist2.eps}
		\label{fig:put16384_hist2}
	}
	\caption{PT Histograms of PUT8192 and PUT16384~\label{fig:pt_hist5}}
\end{figure}

\newpage

\section{Histograms on the EMPv5 Data~\label{sec:empv5_hist}}
The base data of the following histograms are from Table~\ref{tab:exp_notes1}. 
EMPv5(-relaxed) trims outliers from the data of each PUT by EMPv4.
To be more specific, for each run of PUT an outlier is determined 
as the one above and below the average $\pm$ *five\footnote{In the stricter version, we use *two*.}* 
standard deviations computed from the EMPv4 data.

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT1]{
		\includegraphics[scale=0.43]{1_sec_pt_hist2.eps}
		\label{fig:put1_hist2}
	}
	\subfigure[PT frequency on PUT2]{
		\includegraphics[scale=0.43]{2_sec_pt_hist2.eps}
		\label{fig:put2_hist2}
	}
	\subfigure[PT frequency on PUT4]{
		\includegraphics[scale=0.43]{4_sec_pt_hist2.eps}
		\label{fig:put4_hist2}
	}
	\subfigure[PT frequency on PUT8]{
		\includegraphics[scale=0.43]{8_sec_pt_hist2.eps}
		\label{fig:put8_hist2}
	}
	\caption{PT Histograms of PUT1 ... PUT8~\label{fig:pt_out_hist1}}
\end{figure}

\newpage

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT16]{
		\includegraphics[scale=0.43]{16_sec_pt_hist2.eps}
		\label{fig:put16_hist2}
	}
	\subfigure[PT frequency on PUT32]{
		\includegraphics[scale=0.43]{32_sec_pt_hist2.eps}
		\label{fig:put32_hist2}
	}
	\subfigure[PT frequency on PUT64]{
		\includegraphics[scale=0.43]{64_sec_pt_hist2.eps}
		\label{fig:put64_hist2}
	}
	\caption{PT Histograms of PUT16 ... PUT64~\label{fig:pt_out_hist2}}
\end{figure}

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT128]{
		\includegraphics[scale=0.43]{128_sec_pt_hist2.eps}
		\label{fig:put128_hist2}
	}
	\subfigure[PT frequency on PUT256]{
		\includegraphics[scale=0.43]{256_sec_pt_hist2.eps}
		\label{fig:put256_hist2}
	}
	\subfigure[PT frequency on PUT512]{
		\includegraphics[scale=0.43]{512_sec_pt_hist2.eps}
		\label{fig:put512_hist2}
	}
	\subfigure[PT frequency on PUT1024]{
		\includegraphics[scale=0.43]{1024_sec_pt_hist2.eps}
		\label{fig:put1024_hist2}
	}
	\caption{PT Histograms of PUT128 ... PUT1024~\label{fig:pt_out_hist3}}
\end{figure}

\newpage

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT2048]{
		\includegraphics[scale=0.43]{2048_sec_pt_hist2.eps}
		\label{fig:put2048_hist2}
	}
	\subfigure[PT frequency on PUT4096]{
		\includegraphics[scale=0.43]{4096_sec_pt_hist2.eps}
		\label{fig:put4096_hist2}
	}
	\caption{PT Histograms of PUT2048 and PUT4096~\label{fig:pt_out_hist4}}
\end{figure}

\newpage

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT8192 with 40 samples (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{8192_sec_pt_hist_out1.eps}
		\label{fig:put8192_out_hist1}
	}
	\subfigure[PT frequency on PUT8192 with 260 samples (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{8192_sec_pt_hist_out2.eps}
		\label{fig:put8192_out_hist2}
	}
	\subfigure[PT frequency on PUT16384 with 40 samples (See Table~\ref{tab:exp_notes1}.)]{
		\includegraphics[scale=0.43]{16384_sec_pt_hist_out1.eps}
		\label{fig:put16384_out_hist1}
	}
	\subfigure[PT frequency on PUT16384 with 260 samples (See Table~\ref{tab:exp_notes1}.))]{
		\includegraphics[scale=0.43]{16384_sec_pt_hist_out2.eps}
		\label{fig:put16384_out_hist2}
	}
	\caption{PT Histograms of PUT8192 and PUT16384~\label{fig:pt_out_hist5}}
\end{figure}

%%% section 5
\clearpage
\newpage

\section{Sample Size vs. Standard Deviation of PT~\label{sec:new_put}}
The base data of the following histograms are from Table~\ref{tab:exp_notes3}.

\subsection{PUT1 and PUT2~\label{sec:put1_put2}}
Table~\ref{fig:put_std} exhibits varying standard deviations 
over increasing sample size on PUT1 and PUT2.
EMPv4 is applied to the table's data.

\begin{table}[h]
\centering
{
 \begin{tabular}{|l|c|c|} \hline
\multirow{2}{*}{Num. of Samples}   & \multicolumn{2}{|c|}{Std. Dev. (msec)} \\  \cline{2-3}
						    & PUT1 & PUT2 \\  \hline\hline
1,000 & 1.07 & 1.40\\ \hline
2,000 & 1.06 & 1.39\\ \hline
3,000 & 1.07 & 1.38\\ \hline
4,000 & 1.07 & 1.37\\ \hline
5,000 & 1.07 & 1.40\\ \hline
6,000 & 1.06& 1.70\\ \hline
7,000 & 1.06& 1.65\\ \hline
8,000 & 1.07&1.62\\ \hline
9,000 & 1.07&1.60\\ \hline
10,000 & 1.07&1.58\\ \hline
11,000 & 1.08&1.57\\ \hline
12,000 & 1.08&1.56\\ \hline
13,000 & 1.08&1.54\\ \hline
14,000 & 1.08&1.53\\ \hline
15,000 & 1.08&1.52\\ \hline
16,000 & 1.08&1.51\\ \hline
17,000 & 1.08&1.50\\ \hline
18,000 & 1.08&1.50\\ \hline
19,000 & 1.08&1.50\\ \hline 
20,000 & 1.08& 1.49\\ \hline 
  \end{tabular}
  }
 \caption{Std. Dev. of PUT1 and PUT2 over increasing sample size~\label{fig:put_std}}
\end{table}

 \newpage

\begin{figure}[hp!]
	\centering
	\includegraphics[scale=0.8]{put_pt_std.eps}
	\caption{Std. dev. of PT on PUT1 and PUT2 over increasing sample size ~\label{fig:std_ss}}
\end{figure}

\begin{figure}[hp!]
	\centering
	\includegraphics[scale=0.8]{put_pt_20k_std.eps}
	\caption{Std. dev. of PT on PUT1 and PUT2 over increasing sample size ~\label{fig:detailed_std_ss}}
\end{figure}

\begin{table}
\centering
{
 \begin{tabular}{|l|l|} \hline
PUT2  & Program Time \\ \hline
{\tt incr\_work} & 2078 msecs (at the 5276th iteration)\\ \hline \hline
Daemon Processes  & Program Time \\ \hline
{\tt md0\_raid1} & 1 msec\\ \hline
{\tt proc\_monitor} & 198 msecs\\ \hline
{\tt rhn\_check} & 460 msecs\\ \hline \hline
{\bf Total} & 659 msecs\\ \hline
  \end{tabular}
  }
 \caption{The daemon processes captured at the hike of PUT2~\label{fig:daemon}}
\end{table}

\begin{figure}
	\centering
	\subfigure[PT frequency on PUT1 by EMPv4 (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put1_20k_hist.eps}
		\label{fig:put1_20k_hist1}
	}
	\subfigure[PT frequency on PUT1 by EMPv5 (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put1_20k_hist2.eps}
		\label{fig:put1_20k_hist2}
	}
	\subfigure[PT frequency on PUT2 by EMPv4  (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put2_20k_hist.eps}
		\label{fig:put2_20k_hist3}
	}
	\subfigure[PT frequency on PUT2 by EMPv5 (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put2_20k_hist2.eps}
		\label{fig:put2_20k_hist4}
	}
	\caption{PT Histograms of PUT1 and PUT2 by 20,000 trials~\label{fig:put_x_20k}}
\end{figure}

\begin{figure}
	\centering
	\subfigure[PT frequency on PUT4 by EMPv4 (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put4_20k_hist.eps}
		\label{fig:put4_20k_hist1}
	}
	\subfigure[PT frequency on PUT4 by EMPv5 (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put4_20k_hist2.eps}
		\label{fig:put4_20k_hist2}
	}
	\subfigure[PT frequency on PUT8 by EMPv4  (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put8_20k_hist.eps}
		\label{fig:put8_20k_hist3}
	}
	\subfigure[PT frequency on PUT8 by EMPv5 (See Table~\ref{tab:exp_notes3}.)]{
		\includegraphics[scale=0.43]{put8_20k_hist2.eps}
		\label{fig:put8_20k_hist4}
	}
	\caption{PT Histograms of PUT4 and PUT8 by 20,000 trials~\label{fig:put_x_20k_2}}
\end{figure}

\newpage

\subsection{PUT16~\label{sec:put16_ss}}
In this experiment we ran PUT16 up to 32,000 from 1,000 times by a factor of two.
The relaxed version of EMPv5 (called {\em EMPv5-relaxed}) uses *five* standard deviations 
whereas its strict version (called {\em EMPv5-strict})  does *two* standard deviations for a vertical gap below and above the average.
(Young: 2k samples seem most appropriate to represent the whole population of PUT16, 
in that the standard deviations by EMPv5 on the 2k sample size are almost at peak compared 
to those of the other sample sizes.)

\begin{table}[h]
\centering
{
 \begin{tabular}{|l|c|c|c|} \hline
\multirow{2}{*}{Num. of Samples}   & \multicolumn{3}{|c|}{Std. Dev. (msec)} \\  \cline{2-4}
						    & EMPv4 & EMPv5-relaxed & EMPv5-strict \\  \hline
1,000 & 1.86 & 1.86  & 1.68\\ \hline
2,000 & 2.20 & 2.12 & 1.81\\ \hline
4,000 & 2.21 & 1.89 & 1.65\\ \hline
8,000 & 2.23 & 1.97 & 1.71\\ \hline
16,000 & 2.07 & 2.00 & 1.61\\ \hline
32,000 & 1.81 & 1.75 & 1.53\\ \hline
  \end{tabular}
  }
 \caption{Standard deviations of PUT16 over increasing sample size~\label{fig:put16_std_ss_summary}}
\end{table}

%\begin{figure}[hp!]
%	\centering
%	\includegraphics[scale=0.8]{put16_std_ss.eps}
%	\caption{Std. dev. of PT on PUT16 over increasing sample size ~\label{fig:put16_std_ss_plot}}
%\end{figure}

\begin{figure}[h]
	\centering
	\subfigure[Standard deviations of PT on PUT16]{
		\includegraphics[scale=0.62]{put16_std_ss.eps}
		\label{fig:put16_std}
	}
	\subfigure[Standard deviations of PT on PUT16, in log scale]{
		\includegraphics[scale=0.62]{put16_std_ss_log.eps}
		\label{fig:put16_std_log}
	}
	\caption{Standard deviations of PT on PUT16 over increasing sample size~\label{fig:put16_std_ss_plot}}
\end{figure}

\clearpage
\newpage

\subsection{Histograms by EMPv4~\label{sec:put16_empv4}}
We apply EMPv4 to the runs of PUT16 as mentioned above.
The following histograms are the results of EMPv4.

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT16 with 1k samples]{
		\includegraphics[scale=0.43]{16_sec_pt_hist.eps}
		\label{fig:put16_1k_hist}
	}
	\subfigure[PT frequency on PUT16 with 2k samples]{
		\includegraphics[scale=0.43]{16_sec_2k_hist.eps}
		\label{fig:put16_2k_hist}
	}
	\subfigure[PT frequency on PUT16 with 4k samples]{
		\includegraphics[scale=0.43]{16_sec_4k_hist.eps}
		\label{fig:put16_4k_hist}
	}
	\subfigure[PT frequency on PUT16 with 8k samples]{
		\includegraphics[scale=0.43]{16_sec_8k_hist.eps}
		\label{fig:put16_8k_hist}
	}
	\caption{PT histogram of PUT16 by EMPv4, with the sample size increasing from 1k to 8k~\label{fig:put16_ss1_empv4}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[PT frequency on PUT16 with 16k samples]{
		\includegraphics[scale=0.43]{16_sec_16k_hist.eps}
		\label{fig:put16_16k_hist}
	}
	\subfigure[PT frequency on PUT16 with 32k samples]{
		\includegraphics[scale=0.43]{16_sec_32k_hist.eps}
		\label{fig:put16_32k_hist}
	}
	\caption{PT histogram of PUT16 by EMPv4, with the sample size increasing from 16k to 32k~\label{fig:put16_ss2_empv4}}
\end{figure}

\clearpage
\newpage

\subsection{Histograms by EMPv5~\label{sec:put16_empv5}}
We now apply EMPv5 to the same data of PUT16. 
To be more specific, we use EMPv5-strict, 
by which the following histograms are obtained.

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on PUT16 with 1k samples]{
		\includegraphics[scale=0.43]{16_sec_1k_hist2.eps}
		\label{fig:put16_1k_hist2}
	}
	\subfigure[PT frequency on PUT16 with 2k samples]{
		\includegraphics[scale=0.43]{16_sec_2k_hist2.eps}
		\label{fig:put16_2k_hist2}
	}
	\subfigure[PT frequency on PUT16 with 4k samples]{
		\includegraphics[scale=0.43]{16_sec_4k_hist2.eps}
		\label{fig:put16_4k_hist2}
	}
	\subfigure[PT frequency on PUT16 with 8k samples]{
		\includegraphics[scale=0.43]{16_sec_8k_hist2.eps}
		\label{fig:put16_8k_hist2}
	}
	\caption{PT histogram of PUT16 by EMPv5, with the sample size increasing from 1k to 8k~\label{fig:put16_ss1_empv5}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[PT frequency on PUT16 with 16k samples]{
		\includegraphics[scale=0.43]{16_sec_16k_hist2.eps}
		\label{fig:put16_16k_hist2}
	}
	\subfigure[PT frequency on PUT16 with 32k samples]{
		\includegraphics[scale=0.43]{16_sec_32k_hist2.eps}
		\label{fig:put16_32k_hist2}
	}
	\caption{PT histogram of PUT16 by EMPv5, with the sample size increasing from 16k to 32k~\label{fig:put16_ss2_empv5}}
\end{figure}

%%%% section 6
\clearpage
\newpage

\section{Dual PUT Experiment~\label{sec:dual_put}}
In this section we study the characteristics of program times measured in the dual PUT experiments, 
in which the for-loop of PUT is broken up into two equal-sized for-loops ($for_{1}$ and $for_{2}$) 
of which elapsed times are individually measured using {\tt gettimeofday()}.
The experiment is designed to see whether there exists ``internal dependency'' of measured times of the two for-loops when PUT is timed.
%For this dual experiment, we measure and compare the execution time of the first half (part I) and second half (part II) of each PUT. 
To be more specific, we compute correlation coefficients between corresponding measured times of $for_{1}$ and $for_{2}$, 
within the same run of each PUT. 
We also compute the correlation coefficients after removing some outliers determined by eye. 
In this experiment we expect that little dependency will be observed within the same run for any PUT.
Note that except dual PUT4096 (on {\tt sodb12}) all the other dual experiments were run on the same machine ({\tt sodb9}), as described in Table~\ref{tab:exp_notes3}. 
This could be one of possible reasons that the structure of dual PUT4096 looks quite different from that of the others 
although both {\tt sodb9} and {\tt sodb12} have the same machine specification.

%Note that after initial analysis on the data of dual PUT2, PUT64, and PUT4096, 
%we conducted additional experiments of dual PUT4, PUT8, PUT16, and PUT32. 
%It was because some dependency (a correlation efficient of 0.3) was observed at PUT2 and we wanted to check out 
%from which such dependency began between PUT2 and PUT64.

The base data of the following table and subsequent plots are from Table~\ref{tab:exp_notes3}.

\begin{table}[h]
\centering
{
% \begin{tabular}{|l|l|l|l|l|l|l|l|} \hline
%Dual PUT  					 & PUT2 	& PUT4 									& PUT8 	& PUT16 	& PUT32 	& PUT64 & PUT4096 \\ \hline
%Sample Size 					  & 1000  & 1000  									& 1000  	& 1000  	& 1000		& 1000 &	500\\ \hline 
%Correlation Coefficients & 0.3  	& -0.07 (-0.15 except max)	& 0.8		& 0.3 		& -0.01 	& -0.01 &	 -0.01\\ \hline 
\begin{tabular}{|c|p{2cm}|p{2cm}|c|} \hline
 & Corr. Coeff. & Corr. Coeff. with outliers removed & Sample Size  (\# of regulars)\\ \hline
PUT2 & 0.3 & 0.5 & 1,000 \\ \hline
PUT4 & -0.07 & -0.2 & 1,000\\ \hline
PUT8 & 0.8 & -0.2 & 1,000\\ \hline
PUT16 & 0.3 & -0.6 & 1,000\\ \hline
PUT32 & 0.01 & -0.4 & 1,000\\ \hline
PUT64 & 0.003 & -0.5 & 1,000\\ \hline
%PUT64(on {\tt sodb9}) & 0.003 & -0.5 & 1,000\\ \hline
%PUT64 on \tt{sodb10} & -0.01 & -0.4 & 1,000\\ \hline
PUT128 & 0.04 & -0.5 & 1,000\\ \hline
PUT256 & 0.004 & -0.1 & 1,000\\ \hline
%PUT256.{\tt{sodb12}} & 0.002 & 0.26 & 1,000 \\ \hline %(850)
PUT512 & -0.03 & -0.1 & 1,000\\ \hline
PUT1024 & 0.14 & -0.04 & 1,000\\ \hline
PUT2048 & -0.01 & -0.08 & 1,000\\ \hline \hline
PUT4096 & -0.01 & -0.2 & 500\\ \hline
  \end{tabular}
  }
 \caption{Overall statistics of dual PUT experiment~\label{fig:corr_dual_put}}
\end{table}

\clearpage
\newpage

\subsection{Scatter Plots}
In this section we plot measured times of dual PUT experiments.
We provide not only scatter plots of raw data but also those of focused clouds to further look inside. 
The focused clouds were obtained by cutting off outliers chosen by eye.

\begin{figure}[htp!]
	\centering
	\subfigure[Measured times on dual PUT2]{
		\includegraphics[scale=0.4]{dual_put2.eps}
		\label{fig:dual_put2}
	}
	\subfigure[Zoomed-in measured times on dual PUT2]{
		\includegraphics[scale=0.4]{dual_put2_trimmed.eps}
		\label{fig:dual_put2_trimmed}
	}
	\subfigure[Measured times on dual PUT4]{
		\includegraphics[scale=0.4]{dual_put4.eps}
		\label{fig:dual_put4}
	}
	\subfigure[Zoomed-in measured times on dual PUT4]{
		\includegraphics[scale=0.4]{dual_put4_trimmed.eps}
		\label{fig:dual_put4_trimmed}
	}
	\caption{Scatter plots on dual PUT2$\sim$PUT8~\label{fig:dual_put_plotting1}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[Measured times on dual PUT8]{
		\includegraphics[scale=0.35]{dual_put8.eps}
		\label{fig:dual_put8}
	}
	\subfigure[Zoomed-in measured times on dual PUT8]{
		\includegraphics[scale=0.35]{dual_put8_trimmed.eps}
		\label{fig:dual_put8_trimmed}
	}
	\subfigure[Measured times on dual PUT16]{
		\includegraphics[scale=0.35]{dual_put16.eps}
		\label{fig:dual_put16}
	}
	\subfigure[Zoomed-in measured times on dual PUT16]{
		\includegraphics[scale=0.35]{dual_put16_trimmed.eps}
		\label{fig:dual_put16_trimmed}
	}
	\subfigure[Measured times on dual PUT32]{
		\includegraphics[scale=0.35]{dual_put32.eps}
		\label{fig:dual_put32}
	}
	\subfigure[Zoomed-in measured times on dual PUT32]{
		\includegraphics[scale=0.35]{dual_put32_trimmed.eps}
		\label{fig:dual_put32_trimmed}
	}
	\caption{Scatter plots on dual PUT8$\sim$PUT32~\label{fig:dual_put_plotting3}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[Measured times on dual PUT64]{
		\includegraphics[scale=0.35]{dual_put64.eps}
		\label{fig:dual_put64}
	}
	\subfigure[Zoomed-in measured times on dual PUT64]{
		\includegraphics[scale=0.35]{dual_put64_trimmed.eps}
		\label{fig:dual_put64_trimmed}
	}
	\subfigure[Measured times on dual PUT128]{
		\includegraphics[scale=0.35]{dual_put128.eps}
		\label{fig:dual_put128}
	}
	\subfigure[Zoomed-in measured times on dual PUT128]{
		\includegraphics[scale=0.35]{dual_put128_trimmed.eps}
		\label{fig:dual_put128_trimmed}
	}
	\subfigure[Measured times on dual PUT256]{
		\includegraphics[scale=0.35]{dual_put256.eps}
		\label{fig:dual_put256}
	}
	\subfigure[Zoomed-in measured times on dual PUT256]{
		\includegraphics[scale=0.35]{dual_put256_trimmed.eps}
		\label{fig:dual_put256_trimmed}
	}
	\caption{Scatter plots on dual PUT64$\sim$PUT256~\label{fig:dual_put_plotting4}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[Measured times on dual PUT512]{
		\includegraphics[scale=0.4]{dual_put512.eps}
		\label{fig:dual_put512}
	}
	\subfigure[Zoomed-in measured times on dual PUT512]{
		\includegraphics[scale=0.4]{dual_put512_trimmed.eps}
		\label{fig:dual_put512_trimmed}
	}
	\subfigure[Measured times on dual PUT1024]{
		\includegraphics[scale=0.4]{dual_put1024.eps}
		\label{fig:dual_put1024}
	}
	\subfigure[Zoomed-in measured times on dual PUT1024]{
		\includegraphics[scale=0.4]{dual_put1024_trimmed.eps}
		\label{fig:dual_put1024_trimmed}
	}
	\caption{Scatter plots on dual PUT512 and PUT1024~\label{fig:dual_put_plotting5}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[Measured times on dual PUT2048 (run on {\tt sodb9})]{
		\includegraphics[scale=0.4]{dual_put2048.eps}
		\label{fig:dual_put2048}
	}
	\subfigure[Zoomed-in measured times on dual PUT2048 (run on {\tt sodb9})]{
		\includegraphics[scale=0.4]{dual_put2048_trimmed.eps}
		\label{fig:dual_put2048_trimmed}
	}
	%\hrulefill
\begin{center}
\line(1,0){450}
\end{center}
%\hdashline
%\dashedrule
	\subfigure[Measured times on dual PUT4096 (run on {\tt sodb8})]{
		\includegraphics[scale=0.4]{dual_put4096.eps}
		\label{fig:dual_put4096}
	}
	\subfigure[Zoomed-in measured times on dual PUT4096 (run on {\tt sodb8})]{
		\includegraphics[scale=0.4]{dual_put4096_trimmed.eps}
		\label{fig:dual_put4096_trimmed}
	}
	\caption{Scatter plots on dual PUT2048$\sim$PUT4096~\label{fig:dual_put_plotting6}}
\end{figure}

\subsubsection{Supplementary Scatter Plots}

\begin{figure}[H]
	\centering
	\subfigure[The central clump of dual PUT64]{
		\includegraphics[scale=0.4]{dual_put64_trimmed_level1.eps}
		\label{fig:fo_dual_put64_trimmed}
	}
	\subfigure[The central clump of dual PUT128]{
		\includegraphics[scale=0.4]{dual_put128_trimmed_level1.eps}
		\label{fig:fo_dual_put128_trimmed}
	}
	\subfigure[The central clump of dual PUT256]{
		\includegraphics[scale=0.4]{dual_put256_trimmed_level1.eps}
		\label{fig:fo_dual_put256_trimmed}
	}
	\subfigure[The central clump of dual PUT512]{
		\includegraphics[scale=0.4]{dual_put512_trimmed_level1.eps}
		\label{fig:fo_dual_put512_trimmed}
	}
	\caption{Focused clumps - part I~\label{fig:extra_clumps1}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[The central clump of dual PUT1024]{
		\includegraphics[scale=0.4]{dual_put1024_trimmed_level1.eps}
		\label{fig:fo_dual_put1024_trimmed}
	}
	\subfigure[The central clump of dual PUT2048]{
		\includegraphics[scale=0.4]{dual_put2048_trimmed_level1.eps}
		\label{fig:fo_dual_put2048_trimmed}
	}
\begin{center}
\line(1,0){450}
\end{center}
	\subfigure[The central clump of dual PUT4096]{
		\includegraphics[scale=0.4]{dual_put4096_trimmed_level1.eps}
		\label{fig:fo_dual_put4096_trimmed}
	}
	\caption{Focused clumps - part II~\label{fig:extra_clumps2}}
\end{figure}

\clearpage
\newpage

\subsubsection{Captured Processes on Dual PUT4096}
In this study we examine a list of processes captured in a specific region of measured times on dual PUT4096. 
The base data is from Figure~\ref{fig:dual_put4096}. 
We divide the data into four subregions by the pivot point of ``2,052,575'' msec on the $x$ and $y$ axes. 
This pivot is chosen by eye, based on Figure~\ref{fig:fo_dual_put4096_trimmed}.

Table~\ref{fig:dual4096_procs} represents a list of daemon processes captured in each of the four subregions 
while Table~\ref{fig:reg_procs_dual4096} shows a list of regions in which a specific daemon process(es) appeared.
The last column in Table~\ref{fig:dual4096_procs} indicates the average time executed by the captured daemon processes 
in each region. Table~\ref{fig:reg_procs_dual4096} also shows 
how long each daemon process was on average run whenever it appeared in a specific region.
\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|p{5cm}|p{2cm}|p{2cm}|} \hline
Subregion & Daemon Processes & Avg. Daemon (Running) Time (msecs) & \# of Samples \\ \hline
Region 1 ($x$ $<$ pivot, $y$ $<$ pivot) & {\tt bash}, {\tt id}, {\tt java}, {\tt rhn\_check}, {\tt rhsmcertd-worke}, {\tt sshd}, {\tt uname} & 3,494 & 214\\ \hline
Region 2 ($x$  $<$ pivot,  $y$ $\geq$ pivot) &  {\tt id}, {\tt java}, {\tt rhn\_check}, {\tt rhsmcertd-worke}, {\tt sshd} & 6,954 & 137\\ \hline
Region 3 ($x$  $\geq$ pivot, $y$ $<$ pivot)& {\tt java}, {\tt rhn\_check}, {\tt rhsmcertd-worke}, {\tt sshd}  & 16,350 & 128\\ \hline
Region 4 ($x$  $\geq$ pivot, $y$ $\geq$ pivot) & {\tt java}, {\tt rhn\_check}, {\tt rhsmcertd-worke}, {\tt sshd}  & 2,130 & 21\\ \hline
  \end{tabular}
  }
 \caption{Captured Daemon Processes and Their Times in Each Region~\label{fig:dual4096_procs}}
\end{table}
%{\tt proc\_monitor}
\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|} \hline
Daemon Processes & Captured Region List \\ \hline
{\tt bash} & 1 (370 msecs)\\ \hline
%{\tt grep} & 1,2,3,4 \\ \hline
{\tt id} & 1 (10 msecs), 2 (10 msecs)\\ \hline
{\tt java} & 1 (11 msecs), 2 (11 msecs), 3 (12 msecs), 4 (12 msecs)\\ \hline
{\tt rhn\_check} & 1 (6,415 msecs), 2 (25,855 msecs), 3 (21,923 msecs), 4 (7,585 msecs)\\ \hline
{\tt rhsmcertd-worke} & 1 (1,157 msecs), 2 (1,156 msecs), 3 (1,158 msecs), 4 (1,152 msecs)\\ \hline
{\tt sshd} & 1 (183 msecs), 2 (64 msecs), 3 (80 msecs), 4 (75 msecs)\\ \hline
{\tt uname} & 1 (10 msecs)\\ \hline
  \end{tabular}
  }
 \caption{Per-Daemon Appearance Region and Its Averge Running Time~\label{fig:reg_procs_dual4096}}
\end{table}

\clearpage
\newpage

\subsubsection{Analysis of Outliers on Dual PUT128}

An in-depth analysis of outliers in Figure~\ref{fig:dual_put128} was conducted. 
This analysis concerns what daemon processes with positive CPU time were captured and how much time was taken by them.
The value in the parentheses is obtained when {\tt proc\_monitor} is considered.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|p{2cm}|l|l|p{5.5cm}|p{2cm}|} \hline
Iter. \# & 1st Half (msec) & 2nd Half (msec) & Daemon Procs. & Daem. Time (msec)\\ \hline
68   & 64142 &  64701 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 18 (220)\\ \hline
180  & 64138 & 64941 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 20 (224)\\ \hline
292  & 64141 & 64706 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 19 (221)\\ \hline
351  & 64938 & 64163 & {\tt kblockd/0}, {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 19 (221)\\ \hline
435  & 64692 & 64154 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java}  (, {\tt proc\_monitor}) & 17 (219)\\ \hline
547  & 66901 & 64192 & {\tt kslowd000}, {\tt kslowd001}, {\tt java} (, {\tt proc\_monitor}) & 16 (218)\\ \hline
659  & 64699 & 64148 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 19 (221)\\ \hline
771  & 64946 & 64149 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 18 (220)\\ \hline
883  & 64700 & 64143 & {\tt kslowd000}, {\tt kslowd001}, {\tt md0\_raid1}, {\tt java} (, {\tt proc\_monitor}) & 17 (219)\\ \hline
986  & 64945 & 64144 & {\tt kslowd000}, {\tt kslowd001}, {\tt java} (, {\tt proc\_monitor}) & 16 (218)\\ \hline \hline %\cline{2-5}
Corr. eff. b/w PUT and daem. proc. times& \multicolumn{4}{|c||}{-0.67 (-0.63)} \\ \hline
%{\tt grep} & 1,2,3,4 \\ \hline
  \end{tabular}
  }
 \caption{Further Examination of Outliers on Dual PUT128 in Figure~\ref{fig:dual_put128}~\label{fig:dual_put128_stat}}
\end{table}

\clearpage
\newpage

\subsubsection{Refined Dual PUT Data}

We conduct another in-depth study on some outliers of the dual PUT512, PUT1024, PUT2048 and PUT4096 data.
The outliers are considered the ones with a daemon process(es) (e.g.,{\tt rhn\_check}) run for 
a significant amount of time (more than 500 msecs for dual PUT512 and 1,000 msecs for the other three, respectively).

\paragraph{Dual 512:} In Figure~\ref{fig:new_put512}, a total of 36 points are identified as outliers and removed, compared to that of Figure~\ref{fig:dual_put512}.

\begin{figure}[htp!]
	\centering
	\subfigure[Refined Dual PUT1512]{
		\includegraphics[scale=0.4]{dual_put512_new.eps}
		\label{fig:new_dual_put512}
	}
	\subfigure[The Focused Clump of Refined Dual PUT512]{%%%% 
		\includegraphics[scale=0.4]{dual_put512_new_trimmed.eps}
		\label{fig:fo_dual_put512_new_trimmed}
	}
	\caption{Refined Dual PUT512 with Significant {\tt rhn\_check} Removed~\label{fig:new_put512}}
\end{figure}

\paragraph{Dual 1024:} In Figure~\ref{fig:refined_dual_put1024}, just four points are removed compared to that of Figure~\ref{fig:dual_put1024}.

\begin{figure}[htp!]
	\centering
	\subfigure[Refined Dual PUT1024]{%%%%  f
		\includegraphics[scale=0.4]{dual_put1024_new.eps}
		\label{fig:refined_dual_put1024}
	}
	\subfigure[The Focused Clump of Refined Dual PUT1024]{%%%% 
		\includegraphics[scale=0.4]{dual_put1024_new_trimmed.eps}
		\label{fig:fo_refined_dual_put1024_new_trimmed}
	}
	\caption{Refined Dual PUT1024 with Significant {\tt rhn\_check} Removed~\label{fig:new_put1024}}
\end{figure}

\clearpage
\newpage

\paragraph{Dual 2048:} In Figure~\ref{fig:refined_dual_put2048}, ten points (outliers) are removed compared to that of Figure~\ref{fig:dual_put2048}.
\begin{figure}[h]
	\centering
	\subfigure[Refined Dual PUT2048]{%%%%  f
		\includegraphics[scale=0.4]{dual_put2048_new.eps}
		\label{fig:refined_dual_put2048}
	}
	\subfigure[The Focused Clump of Refined Dual PUT2048]{%%%% 
		\includegraphics[scale=0.4]{dual_put2048_new_trimmed.eps}
		\label{fig:fo_refined_dual_put2048_new_trimmed}
	}
	\caption{Refined Dual PUT2048 with {\tt rhn\_check} Removed.\label{fig:new_put2048}}
\end{figure}

%\begin{center}
%\line(1,0){480}
%\end{center}
\vspace{-.2in}

\paragraph{Dual 4096:} See Table~\ref{tab:exp_notes3} for the details of this run (on {\tt sodb8}). In Fig.~\ref{fig:refined_dual_put4096}, five points (outliers) are removed compared to that of Fig.~\ref{fig:dual_put4096}. 
{\color{blue}In particular, {\tt rhn\_check} with a significant time (771 msecs) is still included in one of the two outliers at the bottom right corner in Fig.~\ref{fig:refined_dual_put4096}.}
\begin{figure}[h]
	\centering
	\subfigure[Refined Dual PUT4096]{%%%%  five rows (57,250,331,394,456) removed
		\includegraphics[scale=0.4]{dual_put4096_new.eps}
		\label{fig:refined_dual_put4096}
	}
	\subfigure[The Focused Clump of Refined Dual PUT4096]{%%%% two more rows (123, 165) removed
		\includegraphics[scale=0.4]{dual_put4096_new_trimmed.eps}
		\label{fig:fo_refined_dual_put4096_new_trimmed}
	}
	\caption{Refined Dual PUT4096 with {\tt rhn\_check} Removed~\label{fig:new_put4096}}
\end{figure}

\clearpage
\newpage

\subsubsection{Dual PUT Data with {\tt rhn\_check} Disabled}
In this experiment we switched off the {\tt rhn\_check} daemon and then ran dual PUT2048 (on {\tt sodb9}) and PUT4096 (on {\tt sodb8}) 100 times.
For more details of the two runs, see Table~\ref{tab:exp_notes3}. 
Table~\ref{tab:rhn_effect_comp} compares the measurement quality of 
each half when {\tt rhn\_check} was enabled and disabled.

%\begin{table}[h]
%\centering
%{
% \begin{tabular}{|l|c|c|c|} \hline
%\multirow{2}{*}{Num. of Samples}   & \multicolumn{3}{|c|}{Std. Dev. (msec)} \\  \cline{2-4}
%						    & EMPv4 & EMPv5-relaxed & EMPv5-strict \\  \hline
%1,000 & 1.86 & 1.86  & 1.68\\ \hline
%2,000 & 2.20 & 2.12 & 1.81\\ \hline
%4,000 & 2.21 & 1.89 & 1.65\\ \hline
%8,000 & 2.23 & 1.97 & 1.71\\ \hline
%16,000 & 2.07 & 2.00 & 1.61\\ \hline
%32,000 & 1.81 & 1.75 & 1.53\\ \hline
%  \end{tabular}
%  }
% \caption{Standard deviations of PUT16 over increasing sample size~\label{fig:put16_std_ss_summary}}
%\end{table}

\begin{table}[h]
\centering
{
 \begin{tabular}{|l|c|c|c|c|} \hline
  														& Sample Size & Std. Dev. (msec) in 1st & Std. Dev. (msec) in 2nd & Corr. Coeff.\\ \hline
Dual PUT2048 w/ {\tt rhn\_check} &  1000 & 2659.249 & 1739.93 &  -0.008\\ \hline
Dual PUT2048 w/o {\tt rhn\_check} &  100  & 12.66 & 54.79 & 0.55 \\ \hline
Dual PUT4096 w/ {\tt rhn\_check} &  500  & 1942.12 & 1946.38 & -0.005\\ \hline
Dual PUT4096 w/o {\tt rhn\_check} &  100  & 47.26 &  73.49 & -0.22\\ \hline
  \end{tabular}
  }
 \caption{Standard deviations of dual PUT2048 and PUT4096~\label{tab:rhn_effect_comp}}
\end{table}

Figure~\ref{fig:rc_off} compares the measurements of the first and second halves of the dual PUTs.

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT2048]{
		\includegraphics[scale=0.4]{rc_off_dual1_put2048.eps}
		\label{fig:rc_off_dual1_put2048}
	}
	\subfigure[2nd half's measured times on dual PUT2048]{
		\includegraphics[scale=0.4]{rc_off_dual2_put2048.eps}
		\label{fig:rc_off_dual2_put2048}
	}
	\subfigure[1st half's measured times on dual PUT4096]{
		\includegraphics[scale=0.4]{new_dual1_put4096.eps}
		\label{fig:rc_off_dual1_put4096}
	}
	\subfigure[2nd half's measured times on dual PUT4096]{
		\includegraphics[scale=0.4]{rc_off_dual2_put4096.eps}
		\label{fig:rc_off_dual2_put4096}
	}
	\caption{Comparison of measured times on dual PUT2048 and PUT4096 with {\tt rhn\_check} off~\label{fig:rc_off}}
\end{figure}

\clearpage
\newpage

Figure~\ref{fig:new_dual_detailed} shows raw and zoomed-in scatter plots of the measurements of the dual PUTs.

\begin{figure}[H]
	\centering
	\subfigure[Measured times on dual PUT2048]{
		\includegraphics[scale=0.4]{new_dual_put2048.eps}
		\label{fig:new_dual_put2048}
	}
	\subfigure[Zoomed-in measured times on dual PUT2048]{
		\includegraphics[scale=0.4]{new_dual_put2048_trimmed.eps}
		\label{fig:new_dual_put2048_trimmed}
	}
	\subfigure[Measured times on dual PUT4096]{
		\includegraphics[scale=0.4]{new_dual_put4096.eps}
		\label{fig:new_dual_put4096}
	}
	\subfigure[Zoomed-in measured times on dual PUT4096]{
		\includegraphics[scale=0.4]{new_dual_put4096_trimmed.eps}
		\label{fig:new_dual_put4096_trimmed}
	}
	\caption{Scatter plots on dual PUT2048 and PUT4096 with {\tt rhn\_check} off~\label{fig:new_dual_detailed}}
\end{figure}

\clearpage
\newpage

\subsubsection{Influence of {\tt rhn\_check} on the Same Dual PUT}
Even though {\tt rhn\_check} is off, both halves' distributions look different. 
Dual PUT2048 reveals two bands above 1,026,300 msec in the second half (Fig.~\ref{fig:rc_off_new_dual2_put2048})
while exhibiting only three scattered points above 1,026,300 msec in the first half (Fig.~\ref{fig:first100_on_dual2_put2048}). 
In dual PUT4096, the second half (Fig.~\ref{fig:rc_off_new_dual2_put4096_v2}) also shows 
more variance than that of the first half (Fig.~\ref{fig:first100_on_dual2_put4096_v2}).

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT2048 with {\tt rhn\_check} on]{
		\includegraphics[scale=0.4]{first_100_dual1_put2048_v2.eps}
		\label{fig:first100_on_dual1_put2048_v2}
	}
	\subfigure[1st half's measured times on dual PUT2048 with {\tt rhn\_check} off]{
		\includegraphics[scale=0.4]{new_dual1_put2048_v2.eps}
		\label{fig:rc_off_new_dual1_put2048_v2}
	}
	\subfigure[2nd half's measured times on dual PUT2048 with {\tt rhn\_check} on]{
		\includegraphics[scale=0.4]{first_100_dual2_put2048.eps}
		\label{fig:first100_on_dual2_put2048}
	}
	\subfigure[2nd half's measured times on dual PUT2048 with {\tt rhn\_check} off]{
		\includegraphics[scale=0.4]{new_dual2_put2048.eps}
		\label{fig:rc_off_new_dual2_put2048}
	}
	\caption{Comparison of measured times on the same axes for dual PUT2048 in the presence/absence of {\tt rhn\_check}: some outliers excluded~\label{fig:rhn_check_effect1}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT4096 with {\tt rhn\_check} on ]{
		\includegraphics[scale=0.4]{first_100_dual1_put4096.eps}
		\label{fig:first100_on_dual1_put4096}
	}
	\subfigure[1st half's measured times on dual PUT4096 with {\tt rhn\_check} off]{
		\includegraphics[scale=0.4]{new_dual1_put4096.eps}
		\label{fig:rc_off_new_dual1_put4096}
	}
	 \subfigure[2nd half's measured times on dual PUT4096 with {\tt rhn\_check} on]{
		\includegraphics[scale=0.4]{first_100_dual2_put4096_v2.eps}
		\label{fig:first100_on_dual2_put4096_v2}
	}
	\subfigure[2nd half's measured times on dual PUT4096 with {\tt rhn\_check} off]{
		\includegraphics[scale=0.4]{new_dual2_put4096_v2.eps}
		\label{fig:rc_off_new_dual2_put4096_v2}
	}
	\caption{Comparison of measured times on the same axes for dual PUT4096 in the presence/absence of {\tt rhn\_check}: some outliers excluded~\label{fig:rhn_check_effect2}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT2048 with {\tt rhn\_check} on]{
		\includegraphics[scale=0.4]{first_100_dual1_put2048.eps}
		\label{fig:first100_on_dual1_put2048}
	}
	\subfigure[1st half's measured times on dual PUT2048 with {\tt rhn\_check} off]{
		\includegraphics[scale=0.4]{new_dual1_put2048.eps}
		\label{fig:rc_off_new_dual1_put2048}
	}
   \subfigure[2nd half's measured times on dual PUT4096 with {\tt rhn\_check} on]{
		\includegraphics[scale=0.4]{first_100_dual2_put4096.eps}
		\label{fig:first100_on_dual2_put4096}
	}
	\subfigure[2nd half's measured times on dual PUT4096 with {\tt rhn\_check} off]{
		\includegraphics[scale=0.4]{new_dual2_put4096.eps}
		\label{fig:rc_off_new_dual2_put4096}
	}
	\caption{Unrefined (full) measured times on dual PUT2048/PUT4096 in the presence/absence of {\tt rhn\_check}: Fig.~\ref{fig:first100_on_dual1_put2048} unrefined on Fig.~\ref{fig:first100_on_dual1_put2048_v2}, 
Fig.~\ref{fig:rc_off_new_dual1_put2048} on Fig.~\ref{fig:rc_off_new_dual1_put2048_v2}, 
Fig.~\ref{fig:first100_on_dual2_put4096} on Fig.~\ref{fig:first100_on_dual2_put4096_v2}, and  
Fig.~\ref{fig:rc_off_new_dual2_put4096} on Fig.~\ref{fig:rc_off_new_dual2_put4096_v2}, 	
	~\label{fig:rhn_check_effect3}}
\end{figure}

\clearpage
\newpage

\subsection{Investigation about Periodic Appearance of Daemons in dual PUT experiments}
In this section we study a phenomenon of why some daemons ({\tt rhn\_check}, {\tt kslowd000}, and {\tt kslowd001}) 
appear to kick in the second half's execution of the dual PUT with a long task length.

\subsubsection{{\tt rhn\_check}}

\paragraph{Dual PUT2048:}
We examine the trace of regular PUT1024 by \underline{EMPv6}, to check when {\tt rhn\_check} kicks in.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
1024 & 13 & {\tt rhn\_check} & 534 \\ \hline
1024 & 27 & {\tt rhn\_check} & 18372 \\ \hline
1024 & 41 & {\tt rhn\_check} & 422 \\ \hline
1024 & 55 & {\tt rhn\_check} & 522 \\ \hline
1024 & 69 & {\tt rhn\_check} & 417 \\ \hline
1024 & 83 & {\tt rhn\_check} & 522 \\ \hline
1024 & 97 & {\tt rhn\_check} & 417 \\ \hline
1024 & 111 & {\tt rhn\_check} & 518 \\ \hline
1024 & 125 & {\tt rhn\_check} & 417 \\ \hline
1024 & 135 & {\tt rhn\_check} & 522 \\ \hline
1024 & 149 & {\tt rhn\_check} & 425 \\ \hline
1024 & 163 & {\tt rhn\_check} & 528 \\ \hline
1024 & 177 & {\tt rhn\_check} & 418 \\ \hline
1024 & 191 & {\tt rhn\_check} & 523 \\ \hline
1024 & 205 & {\tt rhn\_check} & 419 \\ \hline
1024 & 219 & {\tt rhn\_check} & 522 \\ \hline
1024 & 233 & {\tt rhn\_check} & 422 \\ \hline
1024 & 247 & {\tt rhn\_check} & 526 \\ \hline
1024 & 261 & {\tt rhn\_check} & 421 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 19 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt rhn\_check} Appearance at Even Iterations of (Regular) PUT1024 with 300 measurements~\label{fig:rc_reg_put_1024_even}}
\end{table}

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
1024 & 276 & {\tt rhn\_check} & 19748 \\ \hline
1024 & 290 & {\tt rhn\_check} & 382 \\ \hline
1024 & 298 & {\tt rhn\_check} & 537 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 3 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt rhn\_check} Appearance at Odd Iterations of (Regular) PUT1024 with 300 measurements~\label{fig:rc_reg_put_1024_odd}}
\end{table}

\pagebreak
\paragraph{Dual PUT4096:}
We examine the trace of regular PUT2048 by \underline{EMPv6}, to check when {\tt rhn\_check} kicks in.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
2048 & 13 & rhn\_check & 378 \\ \hline
2048 & 27 & rhn\_check & 375 \\ \hline
2048 & 41 & rhn\_check & 374 \\ \hline
2048 & 55 & rhn\_check & 378 \\ \hline
2048 & 69 & rhn\_check & 383 \\ \hline
2048 & 89 & rhn\_check & 544 \\ \hline
2048 & 103 & rhn\_check & 542 \\ \hline
2048 & 117 & rhn\_check & 19810 \\ \hline
2048 & 131 & rhn\_check & 544 \\ \hline
2048 & 145 & rhn\_check & 540 \\ \hline
2048 & 159 & rhn\_check & 541 \\ \hline
2048 & 163 & rhn\_check & 374 \\ \hline
2048 & 177 & rhn\_check & 378 \\ \hline
2048 & 191 & rhn\_check & 373 \\ \hline
2048 & 205 & rhn\_check & 377 \\ \hline
2048 & 219 & rhn\_check & 376 \\ \hline
2048 & 233 & rhn\_check & 375 \\ \hline
2048 & 245 & rhn\_check & 378 \\ \hline
2048 & 259 & rhn\_check & 376 \\ \hline
2048 & 273 & rhn\_check & 374 \\ \hline
2048 & 287 & rhn\_check & 375 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 21 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt rhn\_check}'s Appearance at Odd Iterations of (Regular) PUT2048 with 300 measurements~\label{fig:rc_reg_put_2048_odd}}
\end{table}

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
2048 & 6 & {\tt rhn\_check} & 539 \\ \hline
2048 & 20 & {\tt rhn\_check} & 541 \\ \hline
2048 & 34 & {\tt rhn\_check} & 549 \\ \hline
2048 & 48 & {\tt rhn\_check} & 539 \\ \hline
2048 & 62 & {\tt rhn\_check} & 542 \\ \hline
2048 & 76 & {\tt rhn\_check} & 543 \\ \hline
2048 & 82 & {\tt rhn\_check} & 379 \\ \hline
2048 & 96 & {\tt rhn\_check} & 374 \\ \hline
2048 & 110 & {\tt rhn\_check} & 387 \\ \hline
2048 & 124 & {\tt rhn\_check} & 371 \\ \hline
2048 & 138 & {\tt rhn\_check} & 377 \\ \hline
2048 & 152 & {\tt rhn\_check} & 378 \\ \hline
2048 & 170 & {\tt rhn\_check} & 536 \\ \hline
2048 & 184 & {\tt rhn\_check} & 551 \\ \hline
2048 & 198 & {\tt rhn\_check} & 543 \\ \hline
2048 & 212 & {\tt rhn\_check} & 541 \\ \hline
2048 & 226 & {\tt rhn\_check} & 540 \\ \hline
2048 & 240 & {\tt rhn\_check} & 549 \\ \hline
2048 & 252 & {\tt rhn\_check} & 542 \\ \hline
2048 & 266 & {\tt rhn\_check} & 541 \\ \hline
2048 & 280 & {\tt rhn\_check} & 544 \\ \hline
2048 & 294 & {\tt rhn\_check} & 573 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 22 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt rhn\_check}'s Appearance at Even Iterations of (Regular) PUT2048 with 300 measurements~\label{fig:rc_reg_put_2048_even}}
\end{table}

\clearpage
\newpage

\subsubsection{{\tt kslowd000}}

\paragraph{Dual PUT2048:}
We examined the trace of regular PUT1024 by \underline{EMPv6} 
but had no record capturing {\tt kslowd000}.

\paragraph{Dual PUT4096:} 
We examined the trace of regular PUT2048 by \underline{EMPv6} 
but had no record capturing {\tt kslowd000}, either. 
But the daemon was found in 
every measurement of the dual PUT2048 data (Figs.~\ref{fig:rc_off_dual1_put2048} and~\ref{fig:rc_off_dual2_put2048}) by \underline{EMPv6} with {\tt rhn\_sd} ({\tt rhn\_check}) disabled.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
2048 & 1 & {\tt kslowd000} & 103 \\ \hline
2048 & 3 & {\tt kslowd000} & 103 \\ \hline
2048 & 5 & {\tt kslowd000} & 103 \\ \hline
2048 & 7 & {\tt kslowd000} & 102 \\ \hline
2048 & 9 & {\tt kslowd000} & 102 \\ \hline
2048 & 11 & {\tt kslowd000} & 104 \\ \hline
2048 & 13 & {\tt kslowd000} & 103 \\ \hline
2048 & 15 & {\tt kslowd000} & 102 \\ \hline
2048 & 17 & {\tt kslowd000} & 103 \\ \hline
2048 & 19 & {\tt kslowd000} & 103 \\ \hline
2048 & 21 & {\tt kslowd000} & 103 \\ \hline
2048 & 23 & {\tt kslowd000} & 102 \\ \hline
2048 & 25 & {\tt kslowd000} & 103 \\ \hline
2048 & 27 & {\tt kslowd000} & 103 \\ \hline
2048 & 29 & {\tt kslowd000} & 102 \\ \hline
2048 & 31 & {\tt kslowd000} & 102 \\ \hline
2048 & 33 & {\tt kslowd000} & 103 \\ \hline
2048 & 35 & {\tt kslowd000} & 103 \\ \hline
2048 & 37 & {\tt kslowd000} & 102 \\ \hline
2048 & 39 & {\tt kslowd000} & 103 \\ \hline
2048 & 41 & {\tt kslowd000} & 103 \\ \hline
2048 & 43 & {\tt kslowd000} & 103 \\ \hline
2048 & 45 & {\tt kslowd000} & 103 \\ \hline
2048 & 47 & {\tt kslowd000} & 103 \\ \hline
2048 & 49 & {\tt kslowd000} & 104 \\ \hline
2048 & 51 & {\tt kslowd000} & 102 \\ \hline
2048 & 53 & {\tt kslowd000} & 102 \\ \hline
2048 & 55 & {\tt kslowd000} & 103 \\ \hline
2048 & 57 & {\tt kslowd000} & 103 \\ \hline
2048 & 59 & {\tt kslowd000} & 102 \\ \hline
2048 & 61 & {\tt kslowd000} & 103 \\ \hline
2048 & 63 & {\tt kslowd000} & 104 \\ \hline
2048 & 65 & {\tt kslowd000} & 103 \\ \hline
2048 & 67 & {\tt kslowd000} & 102 \\ \hline
2048 & 69 & {\tt kslowd000} & 104 \\ \hline
2048 & 71 & {\tt kslowd000} & 103 \\ \hline
2048 & 73 & {\tt kslowd000} & 102 \\ \hline
2048 & 75 & {\tt kslowd000} & 102 \\ \hline
2048 & 77 & {\tt kslowd000} & 103 \\ \hline
2048 & 79 & {\tt kslowd000} & 103 \\ \hline
2048 & 81 & {\tt kslowd000} & 102 \\ \hline
2048 & 83 & {\tt kslowd000} & 104 \\ \hline
2048 & 85 & {\tt kslowd000} & 103 \\ \hline
2048 & 87 & {\tt kslowd000} & 103 \\ \hline
2048 & 89 & {\tt kslowd000} & 103 \\ \hline
2048 & 91 & {\tt kslowd000} & 103 \\ \hline
2048 & 93 & {\tt kslowd000} & 103 \\ \hline
2048 & 95 & {\tt kslowd000} & 102 \\ \hline
2048 & 97 & {\tt kslowd000} & 102 \\ \hline
2048 & 99 & {\tt kslowd000} & 103 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt kslowd000}'s Appearance at Odd Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_odd}}
\end{table}

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
2048 & 2 & {\tt kslowd000} & 103 \\ \hline
2048 & 4 & {\tt kslowd000} & 102 \\ \hline
2048 & 6 & {\tt kslowd000} & 103 \\ \hline
2048 & 8 & {\tt kslowd000} & 103 \\ \hline
2048 & 10 & {\tt kslowd000} & 103 \\ \hline
2048 & 12 & {\tt kslowd000} & 102 \\ \hline
2048 & 14 & {\tt kslowd000} & 103 \\ \hline
2048 & 16 & {\tt kslowd000} & 103 \\ \hline
2048 & 18 & {\tt kslowd000} & 102 \\ \hline
2048 & 20 & {\tt kslowd000} & 102 \\ \hline
2048 & 22 & {\tt kslowd000} & 103 \\ \hline
2048 & 24 & {\tt kslowd000} & 103 \\ \hline
2048 & 26 & {\tt kslowd000} & 102 \\ \hline
2048 & 28 & {\tt kslowd000} & 103 \\ \hline
2048 & 30 & {\tt kslowd000} & 104 \\ \hline
2048 & 32 & {\tt kslowd000} & 104 \\ \hline
2048 & 34 & {\tt kslowd000} & 102 \\ \hline
2048 & 36 & {\tt kslowd000} & 103 \\ \hline
2048 & 38 & {\tt kslowd000} & 103 \\ \hline
2048 & 40 & {\tt kslowd000} & 103 \\ \hline
2048 & 42 & {\tt kslowd000} & 102 \\ \hline
2048 & 44 & {\tt kslowd000} & 103 \\ \hline
2048 & 46 & {\tt kslowd000} & 103 \\ \hline
2048 & 48 & {\tt kslowd000} & 102 \\ \hline
2048 & 50 & {\tt kslowd000} & 103 \\ \hline
2048 & 52 & {\tt kslowd000} & 103 \\ \hline
2048 & 54 & {\tt kslowd000} & 103 \\ \hline
2048 & 56 & {\tt kslowd000} & 103 \\ \hline
2048 & 58 & {\tt kslowd000} & 103 \\ \hline
2048 & 60 & {\tt kslowd000} & 103 \\ \hline
2048 & 62 & {\tt kslowd000} & 102 \\ \hline
2048 & 64 & {\tt kslowd000} & 102 \\ \hline
2048 & 66 & {\tt kslowd000} & 103 \\ \hline
2048 & 68 & {\tt kslowd000} & 103 \\ \hline
2048 & 70 & {\tt kslowd000} & 102 \\ \hline
2048 & 72 & {\tt kslowd000} & 103 \\ \hline
2048 & 74 & {\tt kslowd000} & 104 \\ \hline
2048 & 76 & {\tt kslowd000} & 103 \\ \hline
2048 & 78 & {\tt kslowd000} & 103 \\ \hline
2048 & 80 & {\tt kslowd000} & 103 \\ \hline
2048 & 82 & {\tt kslowd000} & 103 \\ \hline
2048 & 84 & {\tt kslowd000} & 102 \\ \hline
2048 & 86 & {\tt kslowd000} & 102 \\ \hline
2048 & 88 & {\tt kslowd000} & 103 \\ \hline
2048 & 90 & {\tt kslowd000} & 103 \\ \hline
2048 & 92 & {\tt kslowd000} & 102 \\ \hline
2048 & 94 & {\tt kslowd000} & 103 \\ \hline
2048 & 96 & {\tt kslowd000} & 103 \\ \hline
2048 & 98 & {\tt kslowd000} & 103 \\ \hline
2048 & 100 & {\tt kslowd000} & 102 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt kslowd000}'s Appearance at Even Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_even}}
\end{table}

\clearpage
\newpage

\subsubsection{{\tt kslowd001}} 
We examined the trace of regular PUT2048 by \underline{EMPv6} 
but had no record capturing {\tt kslowd001}, either. 
But the daemon was found in 
every measurement of the dual PUT2048 data (Figs.~\ref{fig:rc_off_dual1_put2048} and~\ref{fig:rc_off_dual2_put2048}) by \underline{ EMPv6} with {\tt rhn\_sd} ({\tt rhn\_check}) disabled.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
2048 & 1 & {\tt kslowd001} & 104 \\ \hline
2048 & 3 & {\tt kslowd001} & 103 \\ \hline
2048 & 5 & {\tt kslowd001} & 102 \\ \hline
2048 & 7 & {\tt kslowd001} & 103 \\ \hline
2048 & 9 & {\tt kslowd001} & 103 \\ \hline
2048 & 11 & {\tt kslowd001} & 102 \\ \hline
2048 & 13 & {\tt kslowd001} & 102 \\ \hline
2048 & 15 & {\tt kslowd001} & 103 \\ \hline
2048 & 17 & {\tt kslowd001} & 103 \\ \hline
2048 & 19 & {\tt kslowd001} & 102 \\ \hline
2048 & 21 & {\tt kslowd001} & 103 \\ \hline
2048 & 23 & {\tt kslowd001} & 103 \\ \hline
2048 & 25 & {\tt kslowd001} & 103 \\ \hline
2048 & 27 & {\tt kslowd001} & 102 \\ \hline
2048 & 29 & {\tt kslowd001} & 103 \\ \hline
2048 & 31 & {\tt kslowd001} & 103 \\ \hline
2048 & 33 & {\tt kslowd001} & 102 \\ \hline
2048 & 35 & {\tt kslowd001} & 102 \\ \hline
2048 & 37 & {\tt kslowd001} & 103 \\ \hline
2048 & 39 & {\tt kslowd001} & 103 \\ \hline
2048 & 41 & {\tt kslowd001} & 103 \\ \hline
2048 & 43 & {\tt kslowd001} & 103 \\ \hline
2048 & 45 & {\tt kslowd001} & 103 \\ \hline
2048 & 47 & {\tt kslowd001} & 102 \\ \hline
2048 & 49 & {\tt kslowd001} & 102 \\ \hline
2048 & 51 & {\tt kslowd001} & 103 \\ \hline
2048 & 53 & {\tt kslowd001} & 103 \\ \hline
2048 & 55 & {\tt kslowd001} & 102 \\ \hline
2048 & 57 & {\tt kslowd001} & 103 \\ \hline
2048 & 59 & {\tt kslowd001} & 104 \\ \hline
2048 & 61 & {\tt kslowd001} & 103 \\ \hline
2048 & 63 & {\tt kslowd001} & 102 \\ \hline
2048 & 65 & {\tt kslowd001} & 103 \\ \hline
2048 & 67 & {\tt kslowd001} & 103 \\ \hline
2048 & 69 & {\tt kslowd001} & 102 \\ \hline
2048 & 71 & {\tt kslowd001} & 102 \\ \hline
2048 & 73 & {\tt kslowd001} & 103 \\ \hline
2048 & 75 & {\tt kslowd001} & 103 \\ \hline
2048 & 77 & {\tt kslowd001} & 102 \\ \hline
2048 & 79 & {\tt kslowd001} & 104 \\ \hline
2048 & 81 & {\tt kslowd001} & 103 \\ \hline
2048 & 83 & {\tt kslowd001} & 103 \\ \hline
2048 & 85 & {\tt kslowd001} & 103 \\ \hline
2048 & 87 & {\tt kslowd001} & 103 \\ \hline
2048 & 89 & {\tt kslowd001} & 103 \\ \hline
2048 & 91 & {\tt kslowd001} & 102 \\ \hline
2048 & 93 & {\tt kslowd001} & 102 \\ \hline
2048 & 95 & {\tt kslowd001} & 103 \\ \hline
2048 & 97 & {\tt kslowd001} & 103 \\ \hline
2048 & 99 & {\tt kslowd001} & 102 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt kslowd001}'s Appearance at Odd Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_odd1}}
\end{table}

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|l|l|} \hline
Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
2048 & 2 & {\tt kslowd001} & 102 \\ \hline
2048 & 4 & {\tt kslowd001} & 103 \\ \hline
2048 & 6 & {\tt kslowd001} & 103 \\ \hline
2048 & 8 & {\tt kslowd001} & 102 \\ \hline
2048 & 10 & {\tt kslowd001} & 103 \\ \hline
2048 & 12 & {\tt kslowd001} & 103 \\ \hline
2048 & 14 & {\tt kslowd001} & 102 \\ \hline
2048 & 16 & {\tt kslowd001} & 102 \\ \hline
2048 & 18 & {\tt kslowd001} & 103 \\ \hline
2048 & 20 & {\tt kslowd001} & 104 \\ \hline
2048 & 22 & {\tt kslowd001} & 102 \\ \hline
2048 & 24 & {\tt kslowd001} & 102 \\ \hline
2048 & 26 & {\tt kslowd001} & 103 \\ \hline
2048 & 28 & {\tt kslowd001} & 103 \\ \hline
2048 & 30 & {\tt kslowd001} & 104 \\ \hline
2048 & 32 & {\tt kslowd001} & 103 \\ \hline
2048 & 34 & {\tt kslowd001} & 103 \\ \hline
2048 & 36 & {\tt kslowd001} & 102 \\ \hline
2048 & 38 & {\tt kslowd001} & 102 \\ \hline
2048 & 40 & {\tt kslowd001} & 103 \\ \hline
2048 & 42 & {\tt kslowd001} & 103 \\ \hline
2048 & 44 & {\tt kslowd001} & 102 \\ \hline
2048 & 46 & {\tt kslowd001} & 102 \\ \hline
2048 & 48 & {\tt kslowd001} & 103 \\ \hline
2048 & 50 & {\tt kslowd001} & 103 \\ \hline
2048 & 52 & {\tt kslowd001} & 102 \\ \hline
2048 & 54 & {\tt kslowd001} & 103 \\ \hline
2048 & 56 & {\tt kslowd001} & 103 \\ \hline
2048 & 58 & {\tt kslowd001} & 102 \\ \hline
2048 & 60 & {\tt kslowd001} & 102 \\ \hline
2048 & 62 & {\tt kslowd001} & 103 \\ \hline
2048 & 64 & {\tt kslowd001} & 103 \\ \hline
2048 & 66 & {\tt kslowd001} & 102 \\ \hline
2048 & 68 & {\tt kslowd001} & 103 \\ \hline
2048 & 70 & {\tt kslowd001} & 103 \\ \hline
2048 & 72 & {\tt kslowd001} & 103 \\ \hline
2048 & 74 & {\tt kslowd001} & 103 \\ \hline
2048 & 76 & {\tt kslowd001} & 103 \\ \hline
2048 & 78 & {\tt kslowd001} & 103 \\ \hline
2048 & 80 & {\tt kslowd001} & 102 \\ \hline
2048 & 82 & {\tt kslowd001} & 102 \\ \hline
2048 & 84 & {\tt kslowd001} & 103 \\ \hline
2048 & 86 & {\tt kslowd001} & 103 \\ \hline
2048 & 88 & {\tt kslowd001} & 103 \\ \hline
2048 & 90 & {\tt kslowd001} & 103 \\ \hline
2048 & 92 & {\tt kslowd001} & 104 \\ \hline
2048 & 94 & {\tt kslowd001} & 103 \\ \hline
2048 & 96 & {\tt kslowd001} & 102 \\ \hline
2048 & 98 & {\tt kslowd001} & 104 \\ \hline
2048 & 100 & {\tt kslowd001} & 103 \\ \hline
\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
  \end{tabular}
  }
 \caption{{\tt kslowd001}'s Appearance at Even Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_even1}}
\end{table}

\clearpage
\newpage

\subsubsection{Captured Daemon Processes at Some Outliers}

\paragraph{Dual 4096 with {\tt rhn\_check} active:} Data based on Figure~\ref{fig:dual_put4096}.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|p{2cm}|p{8cm}|} \hline
Iter. \# & Process Name (msec)\\ \hline
56 & {\tt rhn\_check} (30031) \\ \hline
249 & {\tt rhn\_check} (30111) \\ \hline
  \end{tabular}
  }
 \caption{Captured daemon processes at outliers on dual PUT4096~\label{fig:dual_put4096_procs}}
\end{table}

\paragraph{Dual 4096 with {\tt rhn\_check} removed:} Data based on Figure~\ref{fig:new_dual_put4096}.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|p{2cm}|p{8cm}|} \hline
Iter. \# & Process Name (msec)\\ \hline
4 & {\tt rhsmcertd-worke} (230), {\tt sshd} (28), {\tt grep} (9), {\tt flush-9:127}(0.1)\\ \hline
25 & {\tt rhsmcertd-worke} (229), {\tt sshd} (27), {\tt grep} (8)\\ \hline
46 & {\tt rhsmcertd-worke} (230), {\tt sshd} (26), {\tt grep} (10), {\tt java}(1) \\ \hline
67 & {\tt rhsmcertd-worke} (231), {\tt sshd} (29), {\tt grep} (11),  {\tt java} (1) \\ \hline
  \end{tabular}
  }
 \caption{Captured daemon processes at outliers on dual PUT4096~\label{fig:no_rhn_dual_put4096_procs}}
\end{table}

\paragraph{Dual 2048 with {\tt rhn\_check} active:} Data based on Figure~\ref{fig:dual_put2048}.

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|p{2cm}|p{8cm}|} \hline
Iter. \# & Process Name (msec)\\ \hline
178 & {\tt rhn\_check} (29211), {\tt flush-9:0} (112), {\tt kslowd001} (106),  {\tt md0\_raid1} (106), 
{\tt kslowd000} (104), {\tt jbd2/md0-8} (20), {\tt rhnsd} (6),  {\tt java} (4), {\tt cifsd} (1)\\ \hline
353 & {\tt rhn\_check} (31877), {\tt flush-9:0} (146), {\tt kslowd001} (105),  {\tt kslowd000} (104), 
{\tt md0\_raid1} (99), {\tt jbd2/md0-8} (15), {\tt rhnsd} (6),  {\tt java} (3), {\tt cifsd} (1)\\ \hline
395 & {\tt rhn\_check} (32436), {\tt flush-9:0} (142), {\tt md0\_raid1} (116), 
{\tt kslowd001} (105),  {\tt kslowd000} (104), {\tt jbd2/md0-8} (11), {\tt rhnsd} (4),  {\tt java} (3), 
{\tt cifsd} (2), {\tt kblockd/0} (1) \\ \hline
  \end{tabular}
  }
 \caption{Captured daemon processes at outliers on dual PUT2048~\label{fig:dual_put2048_procs}}
\end{table}

\clearpage
\newpage

\paragraph{*Dual 2048 with {\tt rhn\_check} removed*:} Data based on Figure~\ref{fig:new_dual_put2048}.
(The average/median of {\tt PUT2048} was 2,052,306 msecs.)

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|p{2cm}|p{8cm}|} \hline
Iter. \# & Process (Time (msec))\\ \hline
%13 & {\tt PUT2048} (2,052,315) / Daemons (711): {\tt rhsmcertd-worke} (234), {\tt proc\_monitor} (200), {\tt kslowd000} (103), {\tt kslowd001} (102),  {\tt ssd} (27), {\tt md0\_raid1} (26), {\tt grep} (16), {\tt java} (2), {\tt jbd2/md0-8} (1)\\ \hline
%55 & {\tt PUT2048} (2,052,341) / Daemons (674): {\tt rhsmcertd-worke} (232), {\tt proc\_monitor} (198), {\tt kslowd000} (103), {\tt kslowd001} (102),  {\tt ssd} (26), {\tt md0\_raid1} (26), {\tt grep} (8), {\tt kblockd/0} (2),  {\tt java} (2), {\tt jbd2/md0-8} (1)\\ \hline
%99 & {\tt PUT2048} (2,052,337) / Daemons (712): {\tt rhsmcertd-worke} (233), {\tt proc\_monitor} (200), {\tt kslowd000} (103), {\tt kslowd001} (102),  {\tt ssd} (33), {\tt md0\_raid1} (27), {\tt grep} (10), {\tt java} (2), {\tt flush-9:0} (1) {\tt kblockd/0} (1) \\ \hline
13 & {\tt PUT2048} (2,052,315) / Daemons (511): {\tt rhsmcertd-worke} (234), {\tt kslowd000} (103), {\tt kslowd001} (102),  {\tt ssd} (27), {\tt md0\_raid1} (26), {\tt grep} (16), {\tt java} (2), {\tt jbd2/md0-8} (1)\\ \hline
55 & {\tt PUT2048} (2,052,341) / Daemons (476): {\tt rhsmcertd-worke} (232),  {\tt kslowd000} (103), {\tt kslowd001} (102),  {\tt ssd} (26), {\tt md0\_raid1} (26), {\tt grep} (8), {\tt kblockd/0} (2),  {\tt java} (2), {\tt jbd2/md0-8} (1)\\ \hline
99 & {\tt PUT2048} (2,052,337) / Daemons (512): {\tt rhsmcertd-worke} (233), {\tt kslowd000} (103), {\tt kslowd001} (102),  {\tt ssd} (33), {\tt md0\_raid1} (27), {\tt grep} (10), {\tt java} (2), {\tt flush-9:0} (1) {\tt kblockd/0} (1) \\ \hline
  \end{tabular}
  }
 \caption{Captured daemon processes at outliers on dual PUT2048~\label{fig:no_rhn_dual_put2048_procs}}
\end{table}

\clearpage
\newpage

\subsection{Program Time Comparison}
In this section we perform one-to-one comparison on measured times of parts I and II 
for the same iteration of each PUT.

\begin{figure}[htp!]
	\centering
	\subfigure[1st half's measured times on dual PUT2]{
		\includegraphics[scale=0.4]{dual1_put2.eps}
		\label{fig:dual1_put2}
	}
	\subfigure[2nd half's measured times on dual PUT2]{
		\includegraphics[scale=0.4]{dual2_put2.eps}
		\label{fig:dual2_put2}
	}
	\subfigure[1st half's measured times on dual PUT4]{
		\includegraphics[scale=0.4]{dual1_put4.eps}
		\label{fig:dual1_put4}
	}
	\subfigure[2nd half's measured times on dual PUT4]{
		\includegraphics[scale=0.4]{dual2_put4.eps}
		\label{fig:dual2_put4}
	}
	\caption{Comparison of measured times on dual PUT2$\sim$PUT4~\label{fig:dual_put_data1}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[1st half's measured times on dual PUT8]{
		\includegraphics[scale=0.35]{dual1_put8.eps}
		\label{fig:dual1_put8}
	}
	\subfigure[2nd half's measured times on dual PUT8]{
		\includegraphics[scale=0.35]{dual2_put8.eps}
		\label{fig:dual2_put8}
	}
	\subfigure[1st half's measured times on dual PUT16]{
		\includegraphics[scale=0.35]{dual1_put16.eps}
		\label{fig:dual1_put16}
	}
	\subfigure[2nd half's measured times on dual PUT16]{
		\includegraphics[scale=0.35]{dual2_put16.eps}
		\label{fig:dual2_put16}
	}
	\subfigure[1st half's measured times on dual PUT32]{
		\includegraphics[scale=0.35]{dual1_put32.eps}
		\label{fig:dual1_put32}
	}
	\subfigure[2nd half's measured times on dual PUT32]{
		\includegraphics[scale=0.35]{dual2_put32.eps}
		\label{fig:dual2_put32}
	}
	\caption{Comparison of measured times on dual PUT8$\sim$PUT32~\label{fig:dual_put_data2}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT64]{
		\includegraphics[scale=0.35]{dual1_put64.eps}
		\label{fig:dual1_put64}
	}
	\subfigure[2nd half's measured times on dual PUT64]{
		\includegraphics[scale=0.35]{dual2_put64.eps}
		\label{fig:dual2_put64}
	}
	\subfigure[1st half's measured times on dual PUT128]{
		\includegraphics[scale=0.35]{dual1_put128.eps}
		\label{fig:dual1_put128}
	}
	\subfigure[2nd half's measured times on dual PUT128]{
		\includegraphics[scale=0.35]{dual2_put128.eps}
		\label{fig:dual2_put128}
	}
	\subfigure[1st half's measured times on dual PUT256]{
		\includegraphics[scale=0.35]{dual1_put256.eps}
		\label{fig:dual1_put256}
	}
	\subfigure[2nd half's measured times on dual PUT256]{
		\includegraphics[scale=0.35]{dual2_put256.eps}
		\label{fig:dual2_put256}
	}
	\caption{Comparison of measured times onl dual PUT64$\sim$ PUT256~\label{fig:dual_put_data3}}
\end{figure}

\clearpage
\newpage

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT512]{
		\includegraphics[scale=0.4]{dual1_put512.eps}
		\label{fig:dual1_put512}
	}
	\subfigure[2nd half's measured times on dual PUT512]{
		\includegraphics[scale=0.4]{dual2_put512.eps}
		\label{fig:dual2_put512}
	}
	\subfigure[1st half's measured times on dual PUT1024]{
		\includegraphics[scale=0.4]{dual1_put1024.eps}
		\label{fig:dual1_put1024}
	}
	\subfigure[2nd half's measured times on dual PUT1024]{
		\includegraphics[scale=0.4]{dual2_put1024.eps}
		\label{fig:dual2_put1024}
	}
	\caption{Comparison of measured times on dual PUT512 and PUT1024~\label{fig:dual_put_data4}}
\end{figure}

\clearpage
\newpage

\begin{figure}[H]
	\centering
	\subfigure[1st half's measured times on dual PUT2048]{
		\includegraphics[scale=0.4]{dual1_put2048.eps}
		\label{fig:dual1_put2048}
	}
	\subfigure[2nd half's measured times on dual PUT2048]{
		\includegraphics[scale=0.4]{dual2_put2048.eps}
		\label{fig:dual2_put2048}
	}
	\subfigure[1st half's measured times on dual PUT4096]{
		\includegraphics[scale=0.4]{dual1_put4096.eps}
		\label{fig:dual1_put4096}
	}
	\subfigure[2nd half's measured times on dual PUT4096]{
		\includegraphics[scale=0.4]{dual2_put4096.eps}
		\label{fig:dual2_put4096}
	}
	\caption{Comparison of measured times on dual PUT2048 and PUT4096~\label{fig:dual_put_data5}}
\end{figure}

\clearpage
\newpage

\section{Successive Iterations' Dependency}
%Given a pair of successive iterations, we compute the correlation efficient of the pair. 
%E.g., given dual PUT2, the first half's and second half's measured times were (999, 1001) at iteration 974 and (1002, 1004) at iteration 975.
%The computed correlation efficient of this pair was 1. (command: {\tt cor(c(999, 1002), c(1001, 1004))})
%On the other hand, the first half's and second half's measured times were (1002, 1001) at iteration 965 and (1002, 1004) at iteration 966.
%The correlation of efficient of this pair is -1. 
%But the correlation efficient is 0 for the following pair: (1002, 1002) at iteration 102 and (1002, 1002) at iteration 103.
In this section we plot measured times of each iteration pair consisting of odd and even iterations.
Specifically, the measured times at adjacent, odd and even iterations consist of $x$ and $y$ coordinates and plotted.
The data are described in Table~\ref{tab:exp_notes1}. 
The data in Figure~\ref{fig:iter_dep3} exclude outliers chosen by eye.

\begin{figure}[htp!]
	\centering
	\subfigure[Scatter plot on PUT1]{
		\includegraphics[scale=0.44]{1_sec_ip_pt.eps}
		\label{fig:put1_iterpair}
	}
	\subfigure[Scatter plot on PUT2]{
		\includegraphics[scale=0.44]{2_sec_ip_pt.eps}
		\label{fig:put2_iterpair}
	}
	\subfigure[Scatter plot on PUT4]{
		\includegraphics[scale=0.44]{4_sec_ip_pt.eps}
		\label{fig:put4_iterpair}
	}
	\subfigure[Scatter plot on PUT8]{
		\includegraphics[scale=0.44]{8_sec_ip_pt.eps}
		\label{fig:put8_iterpair}
	}
	\caption{Iteration dependency on PUT1$\sim$PUT8~\label{fig:iter_dep1}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[Scatter plot on PUT16]{
		\includegraphics[scale=0.35]{16_sec_ip_pt.eps}
		\label{fig:put16_iterpair}
	}
	\subfigure[Scatter plot on PUT32]{
		\includegraphics[scale=0.35]{32_sec_ip_pt.eps}
		\label{fig:put32_iterpair}
	}
	\subfigure[Scatter plot on PUT64]{
		\includegraphics[scale=0.35]{64_sec_ip_pt.eps}
		\label{fig:put64_iterpair}
	}
	\subfigure[Scatter plot on PUT128]{
		\includegraphics[scale=0.35]{128_sec_ip_pt.eps}
		\label{fig:put128_iterpair}
	}
	\subfigure[Scatter plot on PUT256]{
		\includegraphics[scale=0.35]{256_sec_ip_pt.eps}
		\label{fig:put256_iterpair}
	}
	\subfigure[Scatter plot on PUT4096]{
		\includegraphics[scale=0.35]{4096_sec_ip_pt.eps}
		\label{fig:put4096_iterpair}
	}
	\caption{Iteration Dependency on PUT16$\sim$PUT256 and PUT4096~\label{fig:iter_dep2}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[Scatter plot on PUT512]{
		\includegraphics[scale=0.35]{512_sec_ip_pt.eps}
		\label{fig:put512_iterpair}
	}
	\subfigure[Scatter plot on trimmed data of PUT512]{
		\includegraphics[scale=0.35]{512_sec_ip_pt_trimmed.eps}
		\label{fig:put512_trimmed_iterpair}
	}
	\subfigure[Scatter plot on PUT1024]{
		\includegraphics[scale=0.35]{1024_sec_ip_pt.eps}
		\label{fig:put1024_iterpair}
	}
	\subfigure[Scatter plot on trimmed data of PUT1024]{
		\includegraphics[scale=0.35]{1024_sec_ip_pt_trimmed.eps}
		\label{fig:put1024_trimmed_terpair}
	}
	\subfigure[Scatter plot on PUT2048]{
		\includegraphics[scale=0.35]{2048_sec_ip_pt.eps}
		\label{fig:put2048_iterpair}
	}
	\subfigure[Scatter plot on trimmed data of PUT2048]{
		\includegraphics[scale=0.35]{2048_sec_ip_pt_trimmed.eps}
		\label{fig:put2048_trimmed_iterpair}
	}
	\caption{Iteration dependency on PUT512, PUT1024, and PUT2048~\label{fig:iter_dep3}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[Scatter plot on PUT8192 in Apr 2015]{
		\includegraphics[scale=0.44]{8192_sec_ip_pt1.eps}
		\label{fig:put8192_iterpair1}
	}
	\subfigure[Scatter plot on PUT8192 in Nov 2015]{
		\includegraphics[scale=0.44]{8192_sec_ip_pt2.eps}
		\label{fig:put8192_iterpair2}
	}
	\subfigure[Scatter plot on PUT16384 in Apr 2015]{
		\includegraphics[scale=0.44]{16384_sec_ip_pt1.eps}
		\label{fig:put16384_iterpair1}
	}
	\subfigure[Scatter plot on PUT16384 in Nov 2015]{
		\includegraphics[scale=0.44]{16384_sec_ip_pt2.eps}
		\label{fig:put16384_iterpair2}
	}
	\caption{Iteration dependency on PUT8192$\sim$PUT16384~\label{fig:iter_dep4}}
\end{figure}

%%%% section 7
\clearpage
\newpage

\section{Influence of Daemon Process on Program Time Measurement~\label{sec:daemon_put}}
In this section we investigate correlations of program times between 
PUT and a group of daemon processes.
The base data, obtained by EMPv4, are from Table~\ref{tab:exp_notes1}. 
It seems that the longer PUT, the stronger correlation of its PT with that of daemon processes.

\begin{table}[h]
\centering
{
 \begin{tabular}{|l|l|l|} \hline
PUT & Correlation Coefficient by EMPv4 & Correlation Coefficient by EMPv5-relaxed\\ \hline
PUT1 & -0.2 & -0.2\\ \hline 
PUT2 & -0.005 & -0.009\\ \hline 
PUT4 & -0.05 & -0.05\\ \hline
PUT8 & 0.1 & 0.1 \\ \hline 
PUT16 & 0.1 & 0.1\\ \hline 
PUT32 & 0.3 & 0.15\\ \hline  
PUT64 & 0.2 & 0.2 \\ \hline 
PUT128 & 0.2 & 0.2 \\ \hline 
PUT256 & 0.4 & 0.4\\ \hline 
PUT512 & 0.9 & 0.6\\ \hline 
PUT1024 & 0.9 & 0.2\\ \hline 
PUT2048 & 0.8 & 0.24\\ \hline 
PUT4096 & 0.4 & 0.4\\ \hline 
PUT8192 in Apr & 0.4 & 0.4 \\ \hline 
PUT8192 in Nov & 0.3 & 0.3 \\ \hline
PUT16384 in Apr & 0.4 & 0.8\\ \hline
PUT16384 in Nov & 0.5 & 0.5\\ \hline
  \end{tabular}
  }
 \caption{Correlation Coefficients between Program Times of Daemon and PUT by EMPv4 and EMPv5~\label{fig:corr_put_daemon}}
\end{table}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT1's PTs]{
		\includegraphics[scale=0.4]{put1_pt.eps}
		\label{fig:put1_pt}
	}
	\subfigure[PUT1's daemon PTs]{
		\includegraphics[scale=0.4]{put1_daemon_pt.eps}
		\label{fig:put1_daemon_pt}
	}
	\caption{Program times between PUT1 vs. Daemon processes~\label{fig:put_daemon_pt1}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT2's PTs]{
		\includegraphics[scale=0.35]{put2_pt.eps}
		\label{fig:put2_pt}
	}
	\subfigure[PUT2's daemon PTs]{
		\includegraphics[scale=0.35]{put2_daemon_pt.eps}
		\label{fig:put2_daemon_pt}
	}
	\subfigure[PUT4's PTs]{
		\includegraphics[scale=0.35]{put4_pt.eps}
		\label{fig:put4_pt}
	}
	\subfigure[PUT4's daemon PTs]{
		\includegraphics[scale=0.35]{put4_daemon_pt.eps}
		\label{fig:put4_daemon_pt}
	}
	\subfigure[PUT8's PTs]{
		\includegraphics[scale=0.35]{put8_pt.eps}
		\label{fig:put8_pt}
	}
	\subfigure[PUT8's daemon PTs]{
		\includegraphics[scale=0.35]{put8_daemon_pt.eps}
		\label{fig:put8_daemon_pt}
	}
	\caption{Program times between PUT2$\sim$PUT8 vs. Daemon processes~\label{fig:put_daemon_pt}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT16's PTs]{
		\includegraphics[scale=0.35]{put16_pt.eps}
		\label{fig:put16_pt}
	}
	\subfigure[PUT16's daemon PTs]{
		\includegraphics[scale=0.35]{put16_daemon_pt.eps}
		\label{fig:put16_daemon_pt}
	}
	\subfigure[PUT32's PTs]{
		\includegraphics[scale=0.35]{put32_pt.eps}
		\label{fig:put32_pt}
	}
	\subfigure[PUT32's daemon PTs]{
		\includegraphics[scale=0.35]{put32_daemon_pt.eps}
		\label{fig:put32_daemon_pt}
	}
	\subfigure[PUT64's PTs]{
		\includegraphics[scale=0.35]{put64_pt.eps}
		\label{fig:put64_pt}
	}
	\subfigure[PUT64's daemon PTs]{
		\includegraphics[scale=0.35]{put64_daemon_pt.eps}
		\label{fig:put64_daemon_pt}
	}
	\caption{Program times between PUT16$\sim$PUT64 vs. Daemon processes~\label{fig:put_daemon_pt2}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT128's PTs]{
		\includegraphics[scale=0.35]{put128_pt.eps}
		\label{fig:put128_pt}
	}
	\subfigure[PUT128's daemon PTs]{
		\includegraphics[scale=0.35]{put128_daemon_pt.eps}
		\label{fig:put128_daemon_pt}
	}
	\subfigure[PUT256's PTs]{
		\includegraphics[scale=0.35]{put256_pt.eps}
		\label{fig:put256_pt}
	}
	\subfigure[PUT256's daemon PTs]{
		\includegraphics[scale=0.35]{put256_daemon_pt.eps}
		\label{fig:put256_daemon_pt}
	}
	\subfigure[PUT512's PTs]{
		\includegraphics[scale=0.35]{put512_pt.eps}
		\label{fig:put512_pt}
	}
	\subfigure[PUT512's daemon PTs]{
		\includegraphics[scale=0.35]{put512_daemon_pt.eps}
		\label{fig:put512_daemon_pt}
	}
	\caption{Program times between PUT128$\sim$PUT512 vs. Daemon processes~\label{fig:put_daemon_pt3}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT1024's PTs]{
		\includegraphics[scale=0.35]{put1024_pt.eps}
		\label{fig:put1024_pt}
	}
	\subfigure[PUT1024's daemon PTs]{
		\includegraphics[scale=0.35]{put1024_daemon_pt.eps}
		\label{fig:put1024_daemon_pt}
	}
	\subfigure[PUT2048's PTs]{
		\includegraphics[scale=0.35]{put2048_pt.eps}
		\label{fig:put2048_pt}
	}
	\subfigure[PUT2048's daemon PTs]{
		\includegraphics[scale=0.35]{put2048_daemon_pt.eps}
		\label{fig:put2048_daemon_pt}
	}
	\subfigure[PUT4096's PTs]{
		\includegraphics[scale=0.35]{put4096_pt.eps}
		\label{fig:put4096_pt}
	}
	\subfigure[PUT4096's daemon PTs]{
		\includegraphics[scale=0.35]{put4096_daemon_pt.eps}
		\label{fig:put4096_daemon_pt}
	}
	\caption{Program times between PUT1024$\sim$PUT4096 vs. Daemon processes~\label{fig:put_daemon_pt4}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT8192's PTs in Apr 2015]{
		\includegraphics[scale=0.35]{put8192_pt1.eps}
		\label{fig:put8192_pt1}
	}
	\subfigure[PUT8192's daemon PTs in Apr 2015]{
		\includegraphics[scale=0.35]{put8192_daemon_pt1.eps}
		\label{fig:put8192_daemon_pt1}
	}
	\subfigure[PUT8192's PTs in Nov 2015]{
		\includegraphics[scale=0.35]{put8192_pt2.eps}
		\label{fig:put8192_pt2}
	}
	\subfigure[PUT8192's daemon PTs in Nov 2015]{
		\includegraphics[scale=0.35]{put8192_daemon_pt2.eps}
		\label{fig:put8192_daemon_pt2}
	}
	\subfigure[Combined PUT8192's PTs]{
		\includegraphics[scale=0.35]{put8192_pt.eps}
		\label{fig:put8192_pt}
	}
	\subfigure[Combined PUT8192's daemon PTs]{
		\includegraphics[scale=0.35]{put8192_daemon_pt.eps}
		\label{fig:put8192_daemon_pt}
	}
	\caption{Program times between PUT8192 vs. Daemon processes~\label{fig:put_daemon_pt6}}
\end{figure}

\begin{figure}[htp!]
	\centering
	\subfigure[PUT16384's PTs in Apr 2015]{
		\includegraphics[scale=0.35]{put16384_pt1.eps}
		\label{fig:put16384_pt1}
	}
	\subfigure[PUT16384's daemon PTs in Apr 2015]{
		\includegraphics[scale=0.35]{put16384_daemon_pt1.eps}
		\label{fig:put16384_daemon_pt1}
	}
	\subfigure[PUT16384's PTs in Nov 2015]{
		\includegraphics[scale=0.35]{put16384_pt2.eps}
		\label{fig:put16384_pt2}
	}
	\subfigure[PUT16384's daemon PTs in Nov 2015]{
		\includegraphics[scale=0.35]{put16384_daemon_pt2.eps}
		\label{fig:put16384_daemon_pt2}
	}
	\subfigure[Combined PUT16384's PTs]{
		\includegraphics[scale=0.35]{put16384_pt.eps}
		\label{fig:put16384_pt}
	}
	\subfigure[Combined PUT16384's daemon PTs]{
		\includegraphics[scale=0.35]{put16384_daemon_pt.eps}
		\label{fig:put16384_daemon_pt}
	}
	\caption{Program times between PUT16384 vs. Daemon processes~\label{fig:put_daemon_pt7}}
\end{figure}

%%%% section 8
\clearpage
\newpage

\section{Conclusion}

Below are the summarized observations by this study. (Order does not matter.)

%\begin{enumerate}
%\item Typically, the distribution of program times (PTs) of PUT is somewhat mixture of two models. 
%\item Outlier trimming does not well shape a normal distribution of PT.
%\item For a short task length of PUT, it seems there’s dependency between iterations in the same run of PUT.
%\item Presence and activation of daemon processes strongly contribute to creating high variance in PT measurement.
%\item The bigger task length, the stronger correlation between PTs of PUT and daemon processes.
%\item When PUT is timed affects PT distribution, due to the presence of daemon processes whose running can't be controlled.
%\item Measurement protocol is scalable with growing sample size and increasing task length.
%\end{enumerate}

\begin{itemize}

\item Typically, the distribution of program times (PTs) of PUT is somewhat mixture of two (or possibly more) models. (But only when you compare fist/second half.)
\item Outlier trimming does not well shape a normal distribution of PT. Outliers are higher non-normal. Still, we don't have a precise characterization of outliers, but doing so would be difficult because of very large, very infrequent outliers (cf. Figure~\ref{fig:std_ss}).
\item For a short task length of PUT, it seems there is dependency between iterations in the same run of PUT.
\item Presence and activation of (infrequent) daemon processes strongly contribute to creating high variance in PT
measurement, due to a few very-far outliers. This results in a very long, very thin tail with a structure we don't understand.
\item The bigger task length, the stronger correlation between PTs of PUT and daemon processes, probably because the longer the PUT runs, the more likely that there will be a daemon process. {\bf Caveat}: what if the task size is not large and the sample size is small?
\item {\bf New:} The bigger task length, the more clumps formed, probably caused by different periodicity of different daemon processes (cf. Figures~\ref{fig:dual_put_plotting4} and~\ref{fig:dual_put_plotting5}). This implies that when a given task length is big, it will give more room to capture activities of various daemon processes at different times. Accordingly, it leads to observing mixture of two (or more) models, as addressed in the second item. 
\item When PUT is timed affects PT distribution, due to the presence of daemon processes whose running
can't be controlled.
\item Measurement protocol is scalable with growing sample size and increasing task length, because the protocol can do the trimming: see above. {\bf Idea}: Run one version *many* times to estimate the distribution and what is a reasonable criterion for dropping outliers.

\end{itemize}

Below is a remaining issue(s) revealed through this study.

\begin{itemize}

\item We want a definition of ``outlier'' over a wide range of program times.

\end{itemize}

\clearpage
\newpage
\section{Appendix}
This appendix provides specific details of what daemon processes 
was captured and how much time was taken at a specific iteration revealing the most program time of a certain PUT.

\subsection{Breakdown on Program Times of Daemon Processes}
\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|} \hline
PUT256  & Program Time \\ \hline
{\tt incr\_work} & 256,514 msecs (at the 118th iteration)\\ \hline \hline
Daemon Processes  & Program Time \\ \hline
{\tt java} & 2 msecs\\ \hline
{\tt md0\_raid1} & 4 msecs\\ \hline
{\tt jbd2/md0-8} & 1 msec\\ \hline
{\tt flush-9:0} & 10 msecs\\ \hline
{\tt proc\_monitor} & 262 msecs\\ \hline
{\tt rhnsd} & 6 msecs\\ \hline
{\tt rhn\_check} & 1,944 msecs\\ \hline \hline
{\bf Total} & 2,229 msecs\\ \hline
  \end{tabular}
  }
 \caption{The daemon processes captured at the worst PT of PUT256~\label{fig:put256_daemon}}
\end{table}

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|} \hline
PUT256  & Program Time \\ \hline
{\tt incr\_work} & 513,152 msecs (at the 128th iteration)\\ \hline \hline
Daemon Processes  & Program Time \\ \hline
{\tt java} & 2 msecs\\ \hline
{\tt md0\_raid1} & 51 msecs\\ \hline
{\tt jbd2/md0-8} & 27 msecs\\ \hline
{\tt flush-9:0} & 86 msecs\\ \hline
{\tt proc\_monitor} & 270 msecs\\ \hline
{\tt rhnsd} & 6 msecs\\ \hline
{\tt rhn\_check} & 19,820 msecs\\ \hline \hline
{\bf Total} & 20,262 msecs\\ \hline
  \end{tabular}
  }
 \caption{The daemon processes captured at the worst PT of PUT512~\label{fig:put512_daemon}}
\end{table}

\begin{table}[htp!]
\centering
{
 \begin{tabular}{|l|l|} \hline
PUT4096 & Program Time \\ \hline
{\tt incr\_work} & 4,105,629 msecs (at the 284th iteration)\\ \hline \hline
Daemon Processes  & Program Time \\ \hline
{\tt events/0} & 1 msec\\ \hline
{\tt kblockd/0} & 1 msec\\ \hline
{\tt kslowd000} & 31,710 msecs\\ \hline
{\tt kslowd001} & 31,782 msecs\\ \hline
{\tt md0\_raid1} & 82 msecs\\ \hline
{\tt jbd2/md0-8} & 21 msecs\\ \hline
{\tt flush-9:0} & 79 msecs\\ \hline
{\tt proc\_monitor} & 206 msecs\\ \hline
{\tt rhnsd} & 3 msecs\\ \hline
{\tt ntpd} & 1 msec\\ \hline
{\tt java} & 2 msecs\\ \hline
{\tt rhn\_check} & 21,840 msecs\\ \hline \hline
{\bf Total} & 85,728 msecs\\ \hline
  \end{tabular}
  }
 \caption{The daemon processes captured at the worst PT of PUT4096~\label{fig:put4096_daemon}}
\end{table}

\begin{table}[H]
\centering
{
 \begin{tabular}{|l|l|} \hline
PUT8192 & Program Time \\ \hline
{\tt incr\_work} & 8,207,884 msecs (at the 244th iteration)\\ \hline \hline
Daemon Processes  & Program Time \\ \hline
{\tt kblockd/0} & 3 msecs\\ \hline
{\tt kslowd000} & 31,710 msecs\\ \hline
{\tt kslowd001} & 31,782 msecs\\ \hline
{\tt md0\_raid1} & 12 msecs\\ \hline
{\tt jbd2/md0-8} & 2 msecs\\ \hline
{\tt proc\_monitor} & 204 msecs\\ \hline
{\tt rhnsd} & 6 msecs\\ \hline
{\tt java} & 1 msec\\ \hline
%{\tt sshd} & 11 msecs\\ \hline
%{\tt sshd} & 2 msecs\\ \hline
%{\tt grep} & 5 msecs\\ \hline
%{\tt grep} & 1 msec\\ \hline
%{\tt sshd} & 13 msecs\\ \hline
%{\tt grep} & 5 msecs\\ \hline
%{\tt grep} & 1 msec\\ \hline
{\tt rhsmcertd-worke} & 114 msecs\\ \hline
{\tt rhsmcertd-worke} & 114 msecs\\ \hline
{\tt rhn\_check} & 708 msecs\\ \hline \hline
%{\bf Total} & 64,694 msecs\\ \hline
{\bf Total} & 64,656 msecs\\ \hline
  \end{tabular}
  }
 \caption{The daemon processes captured at the worst PT of PUT8192~\label{fig:put8192_daemon}}
\end{table}

\begin{table}[H]
\centering
{
 \begin{tabular}{|l|p{7cm}|} \hline
Daemon Processes  & Descriptions\\ \hline
{\tt kslowd000} ({\tt kslowd001}) & A kernel threads for performing things that take a relatively long time. ``Typically, when processing something, these items will spend a lot of time, blocking a thread on I/O, thus making that thread unavailable for doing other work." (\url{http://www.mjmwired.net/kernel/Documentation/slow-work.txt})\\ \hline
{\tt rhn\_check} & An external program for check for updates, run by {\tt rhnsd}\\ \hline 
{\tt rhnsd} & ``A background daemon process that periodically polls the Red Hat Network to see if there are any queued actions available. Typically started from the initialization ({\tt init}) scripts in {\tt /etc/init.d/rhnsd}  when it’s time to poll the Red Hat Network server for available updates and actions. The default interval is every 240 minutes. The minimum polling interval is 60 minutes. 
Any network  activity  is  done  via  the {\tt rhn\_check} utility." (\url{http://linuxcommand.org/man_pages/rhnsd8.html})
       \\ \hline 
  \end{tabular}
  }
 \caption{Descriptions of some daemon processes~\label{fig:daemon_desc}}
\end{table}

\end{document}

### 2017-01-11 pt_characterization/data.old/paper2.tex
pt char paper

--- 2017-01-11 pt_characterization/data.old/paper.tex
@@
 \usepackage{url}
-\usepackage{color}
 \usepackage{subfigure}
@@
 
-\paragraph{Dual 2048:} In Figure~\ref{fig:refined_dual_put2048}, ten points (outliers) are removed compared to that of Figure~\ref{fig:dual_put2048}.
+\paragraph{Dual 2048:} In Figure~\ref{fig:refined_dual_put2048}, ten points are removed compared to that of Figure~\ref{fig:dual_put2048}.
+
 \begin{figure}[h]
@@
 	}
-	\caption{Refined Dual PUT2048 with {\tt rhn\_check} Removed.\label{fig:new_put2048}}
-\end{figure}
-
-%\begin{center}
-%\line(1,0){480}
-%\end{center}
-\vspace{-.2in}
-
-\paragraph{Dual 4096:} See Table~\ref{tab:exp_notes3} for the details of this run (on {\tt sodb8}).
-In Fig.~\ref{fig:refined_dual_put4096}, five points (outliers) are removed compared to that of Fig.~\ref{fig:dual_put4096}.
-{\color{blue}In particular, {\tt rhn\_check} with a significant time (771 msecs) is still included in one of the two outliers at the bottom right corner in Fig.~\ref{fig:refined_dual_put4096}.}
+	\caption{Refined Dual PUT2048 with Significant {\tt rhn\_check} Removed~\label{fig:new_put2048}}
+\end{figure}
+
+\begin{center}
+\line(1,0){480}
+\end{center}
+
+\paragraph{Dual 4096:} See Table~\ref{tab:exp_notes3} for the details of this run (on {\tt sodb8}). In Figure~\ref{fig:refined_dual_put4096}, 
+five points are removed compared to that of Figure~\ref{fig:dual_put4096}.
+
 \begin{figure}[h]
@@
 	}
-	\caption{Refined Dual PUT4096 with {\tt rhn\_check} Removed~\label{fig:new_put4096}}
+	\caption{Refined Dual PUT4096 with Significant {\tt rhn\_check} Removed~\label{fig:new_put4096}}
 \end{figure}
@@
 
-\subsubsection{Influence of {\tt rhn\_check} on the Same Dual PUT}
-Even though {\tt rhn\_check} is off, both halves' distributions look different. 
-Dual PUT2048 reveals two bands above 1,026,300 msec in the second half (Fig.~\ref{fig:rc_off_new_dual2_put2048})
-while exhibiting only three scattered points above 1,026,300 msec in the first half (Fig.~\ref{fig:first100_on_dual2_put2048}). 
-In dual PUT4096, the second half (Fig.~\ref{fig:rc_off_new_dual2_put4096_v2}) also shows 
-more variance than that of the first half (Fig.~\ref{fig:first100_on_dual2_put4096_v2}).
+\section{Influence of {\tt rhn\_check} on the Same Dual PUT}
 
@@
 		\includegraphics[scale=0.4]{new_dual1_put2048_v2.eps}
-		\label{fig:rc_off_new_dual1_put2048_v2}
+		\label{fig:rc_on_new_dual1_put2048_v2}
 	}
@@
 		\includegraphics[scale=0.4]{new_dual2_put2048.eps}
-		\label{fig:rc_off_new_dual2_put2048}
-	}
-	\caption{Comparison of measured times on the same axes for dual PUT2048 in the presence/absence of {\tt rhn\_check}: some outliers excluded~\label{fig:rhn_check_effect1}}
+		\label{fig:rc_on_new_dual2_put2048}
+	}
+	\caption{Comparison of measured times on the same axes for dual PUT2048 in the presence/absence of {\tt rhn\_check}: some outliers excluded)~\label{fig:rhn_check_effect1}}
 \end{figure}
@@
 		\includegraphics[scale=0.4]{new_dual1_put4096.eps}
-		\label{fig:rc_off_new_dual1_put4096}
+		\label{fig:rc_on_new_dual1_put4096}
 	}
@@
 		\includegraphics[scale=0.4]{new_dual2_put4096_v2.eps}
-		\label{fig:rc_off_new_dual2_put4096_v2}
-	}
-	\caption{Comparison of measured times on the same axes for dual PUT4096 in the presence/absence of {\tt rhn\_check}: some outliers excluded~\label{fig:rhn_check_effect2}}
+		\label{fig:rc_on_new_dual2_put4096_v2}
+	}
+	\caption{Comparison of measured times on the same axes for dual PUT4096 in the presence/absence of {\tt rhn\_check}: some outliers excluded)~\label{fig:rhn_check_effect2}}
 \end{figure}
@@
 		\includegraphics[scale=0.4]{new_dual1_put2048.eps}
-		\label{fig:rc_off_new_dual1_put2048}
+		\label{fig:rc_on_new_dual1_put2048}
 	}
@@
 		\includegraphics[scale=0.4]{new_dual2_put4096.eps}
-		\label{fig:rc_off_new_dual2_put4096}
-	}
-	\caption{Unrefined (full) measured times on dual PUT2048/PUT4096 in the presence/absence of {\tt rhn\_check}: Fig.~\ref{fig:first100_on_dual1_put2048} unrefined on Fig.~\ref{fig:first100_on_dual1_put2048_v2},
-Fig.~\ref{fig:rc_off_new_dual1_put2048} on Fig.~\ref{fig:rc_off_new_dual1_put2048_v2}, 
-Fig.~\ref{fig:first100_on_dual2_put4096} on Fig.~\ref{fig:first100_on_dual2_put4096_v2}, and  
-Fig.~\ref{fig:rc_off_new_dual2_put4096} on Fig.~\ref{fig:rc_off_new_dual2_put4096_v2}, 	
-	~\label{fig:rhn_check_effect3}}
-\end{figure}
-
-\clearpage
-\newpage
-
-\subsection{Investigation about Periodic Appearance of Daemons in dual PUT experiments}
-In this section we study a phenomenon of why some daemons ({\tt rhn\_check}, {\tt kslowd000}, and {\tt kslowd001}) 
-appear to kick in the second half's execution of the dual PUT with a long task length.
-
-\subsubsection{{\tt rhn\_check}}
-
-\paragraph{Dual PUT2048:}
-We examine the trace of regular PUT1024 by \underline{EMPv6}, to check when {\tt rhn\_check} kicks in.
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-1024 & 13 & {\tt rhn\_check} & 534 \\ \hline
-1024 & 27 & {\tt rhn\_check} & 18372 \\ \hline
-1024 & 41 & {\tt rhn\_check} & 422 \\ \hline
-1024 & 55 & {\tt rhn\_check} & 522 \\ \hline
-1024 & 69 & {\tt rhn\_check} & 417 \\ \hline
-1024 & 83 & {\tt rhn\_check} & 522 \\ \hline
-1024 & 97 & {\tt rhn\_check} & 417 \\ \hline
-1024 & 111 & {\tt rhn\_check} & 518 \\ \hline
-1024 & 125 & {\tt rhn\_check} & 417 \\ \hline
-1024 & 135 & {\tt rhn\_check} & 522 \\ \hline
-1024 & 149 & {\tt rhn\_check} & 425 \\ \hline
-1024 & 163 & {\tt rhn\_check} & 528 \\ \hline
-1024 & 177 & {\tt rhn\_check} & 418 \\ \hline
-1024 & 191 & {\tt rhn\_check} & 523 \\ \hline
-1024 & 205 & {\tt rhn\_check} & 419 \\ \hline
-1024 & 219 & {\tt rhn\_check} & 522 \\ \hline
-1024 & 233 & {\tt rhn\_check} & 422 \\ \hline
-1024 & 247 & {\tt rhn\_check} & 526 \\ \hline
-1024 & 261 & {\tt rhn\_check} & 421 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 19 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt rhn\_check} Appearance at Even Iterations of (Regular) PUT1024 with 300 measurements~\label{fig:rc_reg_put_1024_even}}
-\end{table}
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-1024 & 276 & {\tt rhn\_check} & 19748 \\ \hline
-1024 & 290 & {\tt rhn\_check} & 382 \\ \hline
-1024 & 298 & {\tt rhn\_check} & 537 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 3 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt rhn\_check} Appearance at Odd Iterations of (Regular) PUT1024 with 300 measurements~\label{fig:rc_reg_put_1024_odd}}
-\end{table}
-
-\pagebreak
-\paragraph{Dual PUT4096:}
-We examine the trace of regular PUT2048 by \underline{EMPv6}, to check when {\tt rhn\_check} kicks in.
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-2048 & 13 & rhn\_check & 378 \\ \hline
-2048 & 27 & rhn\_check & 375 \\ \hline
-2048 & 41 & rhn\_check & 374 \\ \hline
-2048 & 55 & rhn\_check & 378 \\ \hline
-2048 & 69 & rhn\_check & 383 \\ \hline
-2048 & 89 & rhn\_check & 544 \\ \hline
-2048 & 103 & rhn\_check & 542 \\ \hline
-2048 & 117 & rhn\_check & 19810 \\ \hline
-2048 & 131 & rhn\_check & 544 \\ \hline
-2048 & 145 & rhn\_check & 540 \\ \hline
-2048 & 159 & rhn\_check & 541 \\ \hline
-2048 & 163 & rhn\_check & 374 \\ \hline
-2048 & 177 & rhn\_check & 378 \\ \hline
-2048 & 191 & rhn\_check & 373 \\ \hline
-2048 & 205 & rhn\_check & 377 \\ \hline
-2048 & 219 & rhn\_check & 376 \\ \hline
-2048 & 233 & rhn\_check & 375 \\ \hline
-2048 & 245 & rhn\_check & 378 \\ \hline
-2048 & 259 & rhn\_check & 376 \\ \hline
-2048 & 273 & rhn\_check & 374 \\ \hline
-2048 & 287 & rhn\_check & 375 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 21 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt rhn\_check}'s Appearance at Odd Iterations of (Regular) PUT2048 with 300 measurements~\label{fig:rc_reg_put_2048_odd}}
-\end{table}
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-2048 & 6 & {\tt rhn\_check} & 539 \\ \hline
-2048 & 20 & {\tt rhn\_check} & 541 \\ \hline
-2048 & 34 & {\tt rhn\_check} & 549 \\ \hline
-2048 & 48 & {\tt rhn\_check} & 539 \\ \hline
-2048 & 62 & {\tt rhn\_check} & 542 \\ \hline
-2048 & 76 & {\tt rhn\_check} & 543 \\ \hline
-2048 & 82 & {\tt rhn\_check} & 379 \\ \hline
-2048 & 96 & {\tt rhn\_check} & 374 \\ \hline
-2048 & 110 & {\tt rhn\_check} & 387 \\ \hline
-2048 & 124 & {\tt rhn\_check} & 371 \\ \hline
-2048 & 138 & {\tt rhn\_check} & 377 \\ \hline
-2048 & 152 & {\tt rhn\_check} & 378 \\ \hline
-2048 & 170 & {\tt rhn\_check} & 536 \\ \hline
-2048 & 184 & {\tt rhn\_check} & 551 \\ \hline
-2048 & 198 & {\tt rhn\_check} & 543 \\ \hline
-2048 & 212 & {\tt rhn\_check} & 541 \\ \hline
-2048 & 226 & {\tt rhn\_check} & 540 \\ \hline
-2048 & 240 & {\tt rhn\_check} & 549 \\ \hline
-2048 & 252 & {\tt rhn\_check} & 542 \\ \hline
-2048 & 266 & {\tt rhn\_check} & 541 \\ \hline
-2048 & 280 & {\tt rhn\_check} & 544 \\ \hline
-2048 & 294 & {\tt rhn\_check} & 573 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 22 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt rhn\_check}'s Appearance at Even Iterations of (Regular) PUT2048 with 300 measurements~\label{fig:rc_reg_put_2048_even}}
-\end{table}
-
-\clearpage
-\newpage
-
-\subsubsection{{\tt kslowd000}}
-
-\paragraph{Dual PUT2048:}
-We examined the trace of regular PUT1024 by \underline{EMPv6} 
-but had no record capturing {\tt kslowd000}.
-
-\paragraph{Dual PUT4096:} 
-We examined the trace of regular PUT2048 by \underline{EMPv6} 
-but had no record capturing {\tt kslowd000}, either. 
-But the daemon was found in 
-every measurement of the dual PUT2048 data (Figs.~\ref{fig:rc_off_dual1_put2048} and~\ref{fig:rc_off_dual2_put2048}) by \underline{EMPv6} with {\tt rhn\_sd} ({\tt rhn\_check}) disabled.
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-2048 & 1 & {\tt kslowd000} & 103 \\ \hline
-2048 & 3 & {\tt kslowd000} & 103 \\ \hline
-2048 & 5 & {\tt kslowd000} & 103 \\ \hline
-2048 & 7 & {\tt kslowd000} & 102 \\ \hline
-2048 & 9 & {\tt kslowd000} & 102 \\ \hline
-2048 & 11 & {\tt kslowd000} & 104 \\ \hline
-2048 & 13 & {\tt kslowd000} & 103 \\ \hline
-2048 & 15 & {\tt kslowd000} & 102 \\ \hline
-2048 & 17 & {\tt kslowd000} & 103 \\ \hline
-2048 & 19 & {\tt kslowd000} & 103 \\ \hline
-2048 & 21 & {\tt kslowd000} & 103 \\ \hline
-2048 & 23 & {\tt kslowd000} & 102 \\ \hline
-2048 & 25 & {\tt kslowd000} & 103 \\ \hline
-2048 & 27 & {\tt kslowd000} & 103 \\ \hline
-2048 & 29 & {\tt kslowd000} & 102 \\ \hline
-2048 & 31 & {\tt kslowd000} & 102 \\ \hline
-2048 & 33 & {\tt kslowd000} & 103 \\ \hline
-2048 & 35 & {\tt kslowd000} & 103 \\ \hline
-2048 & 37 & {\tt kslowd000} & 102 \\ \hline
-2048 & 39 & {\tt kslowd000} & 103 \\ \hline
-2048 & 41 & {\tt kslowd000} & 103 \\ \hline
-2048 & 43 & {\tt kslowd000} & 103 \\ \hline
-2048 & 45 & {\tt kslowd000} & 103 \\ \hline
-2048 & 47 & {\tt kslowd000} & 103 \\ \hline
-2048 & 49 & {\tt kslowd000} & 104 \\ \hline
-2048 & 51 & {\tt kslowd000} & 102 \\ \hline
-2048 & 53 & {\tt kslowd000} & 102 \\ \hline
-2048 & 55 & {\tt kslowd000} & 103 \\ \hline
-2048 & 57 & {\tt kslowd000} & 103 \\ \hline
-2048 & 59 & {\tt kslowd000} & 102 \\ \hline
-2048 & 61 & {\tt kslowd000} & 103 \\ \hline
-2048 & 63 & {\tt kslowd000} & 104 \\ \hline
-2048 & 65 & {\tt kslowd000} & 103 \\ \hline
-2048 & 67 & {\tt kslowd000} & 102 \\ \hline
-2048 & 69 & {\tt kslowd000} & 104 \\ \hline
-2048 & 71 & {\tt kslowd000} & 103 \\ \hline
-2048 & 73 & {\tt kslowd000} & 102 \\ \hline
-2048 & 75 & {\tt kslowd000} & 102 \\ \hline
-2048 & 77 & {\tt kslowd000} & 103 \\ \hline
-2048 & 79 & {\tt kslowd000} & 103 \\ \hline
-2048 & 81 & {\tt kslowd000} & 102 \\ \hline
-2048 & 83 & {\tt kslowd000} & 104 \\ \hline
-2048 & 85 & {\tt kslowd000} & 103 \\ \hline
-2048 & 87 & {\tt kslowd000} & 103 \\ \hline
-2048 & 89 & {\tt kslowd000} & 103 \\ \hline
-2048 & 91 & {\tt kslowd000} & 103 \\ \hline
-2048 & 93 & {\tt kslowd000} & 103 \\ \hline
-2048 & 95 & {\tt kslowd000} & 102 \\ \hline
-2048 & 97 & {\tt kslowd000} & 102 \\ \hline
-2048 & 99 & {\tt kslowd000} & 103 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt kslowd000}'s Appearance at Odd Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_odd}}
-\end{table}
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-2048 & 2 & {\tt kslowd000} & 103 \\ \hline
-2048 & 4 & {\tt kslowd000} & 102 \\ \hline
-2048 & 6 & {\tt kslowd000} & 103 \\ \hline
-2048 & 8 & {\tt kslowd000} & 103 \\ \hline
-2048 & 10 & {\tt kslowd000} & 103 \\ \hline
-2048 & 12 & {\tt kslowd000} & 102 \\ \hline
-2048 & 14 & {\tt kslowd000} & 103 \\ \hline
-2048 & 16 & {\tt kslowd000} & 103 \\ \hline
-2048 & 18 & {\tt kslowd000} & 102 \\ \hline
-2048 & 20 & {\tt kslowd000} & 102 \\ \hline
-2048 & 22 & {\tt kslowd000} & 103 \\ \hline
-2048 & 24 & {\tt kslowd000} & 103 \\ \hline
-2048 & 26 & {\tt kslowd000} & 102 \\ \hline
-2048 & 28 & {\tt kslowd000} & 103 \\ \hline
-2048 & 30 & {\tt kslowd000} & 104 \\ \hline
-2048 & 32 & {\tt kslowd000} & 104 \\ \hline
-2048 & 34 & {\tt kslowd000} & 102 \\ \hline
-2048 & 36 & {\tt kslowd000} & 103 \\ \hline
-2048 & 38 & {\tt kslowd000} & 103 \\ \hline
-2048 & 40 & {\tt kslowd000} & 103 \\ \hline
-2048 & 42 & {\tt kslowd000} & 102 \\ \hline
-2048 & 44 & {\tt kslowd000} & 103 \\ \hline
-2048 & 46 & {\tt kslowd000} & 103 \\ \hline
-2048 & 48 & {\tt kslowd000} & 102 \\ \hline
-2048 & 50 & {\tt kslowd000} & 103 \\ \hline
-2048 & 52 & {\tt kslowd000} & 103 \\ \hline
-2048 & 54 & {\tt kslowd000} & 103 \\ \hline
-2048 & 56 & {\tt kslowd000} & 103 \\ \hline
-2048 & 58 & {\tt kslowd000} & 103 \\ \hline
-2048 & 60 & {\tt kslowd000} & 103 \\ \hline
-2048 & 62 & {\tt kslowd000} & 102 \\ \hline
-2048 & 64 & {\tt kslowd000} & 102 \\ \hline
-2048 & 66 & {\tt kslowd000} & 103 \\ \hline
-2048 & 68 & {\tt kslowd000} & 103 \\ \hline
-2048 & 70 & {\tt kslowd000} & 102 \\ \hline
-2048 & 72 & {\tt kslowd000} & 103 \\ \hline
-2048 & 74 & {\tt kslowd000} & 104 \\ \hline
-2048 & 76 & {\tt kslowd000} & 103 \\ \hline
-2048 & 78 & {\tt kslowd000} & 103 \\ \hline
-2048 & 80 & {\tt kslowd000} & 103 \\ \hline
-2048 & 82 & {\tt kslowd000} & 103 \\ \hline
-2048 & 84 & {\tt kslowd000} & 102 \\ \hline
-2048 & 86 & {\tt kslowd000} & 102 \\ \hline
-2048 & 88 & {\tt kslowd000} & 103 \\ \hline
-2048 & 90 & {\tt kslowd000} & 103 \\ \hline
-2048 & 92 & {\tt kslowd000} & 102 \\ \hline
-2048 & 94 & {\tt kslowd000} & 103 \\ \hline
-2048 & 96 & {\tt kslowd000} & 103 \\ \hline
-2048 & 98 & {\tt kslowd000} & 103 \\ \hline
-2048 & 100 & {\tt kslowd000} & 102 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt kslowd000}'s Appearance at Even Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_even}}
-\end{table}
-
-\clearpage
-\newpage
-
-\subsubsection{{\tt kslowd001}} 
-We examined the trace of regular PUT2048 by \underline{EMPv6} 
-but had no record capturing {\tt kslowd001}, either. 
-But the daemon was found in 
-every measurement of the dual PUT2048 data (Figs.~\ref{fig:rc_off_dual1_put2048} and~\ref{fig:rc_off_dual2_put2048}) by \underline{ EMPv6} with {\tt rhn\_sd} ({\tt rhn\_check}) disabled.
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-2048 & 1 & {\tt kslowd001} & 104 \\ \hline
-2048 & 3 & {\tt kslowd001} & 103 \\ \hline
-2048 & 5 & {\tt kslowd001} & 102 \\ \hline
-2048 & 7 & {\tt kslowd001} & 103 \\ \hline
-2048 & 9 & {\tt kslowd001} & 103 \\ \hline
-2048 & 11 & {\tt kslowd001} & 102 \\ \hline
-2048 & 13 & {\tt kslowd001} & 102 \\ \hline
-2048 & 15 & {\tt kslowd001} & 103 \\ \hline
-2048 & 17 & {\tt kslowd001} & 103 \\ \hline
-2048 & 19 & {\tt kslowd001} & 102 \\ \hline
-2048 & 21 & {\tt kslowd001} & 103 \\ \hline
-2048 & 23 & {\tt kslowd001} & 103 \\ \hline
-2048 & 25 & {\tt kslowd001} & 103 \\ \hline
-2048 & 27 & {\tt kslowd001} & 102 \\ \hline
-2048 & 29 & {\tt kslowd001} & 103 \\ \hline
-2048 & 31 & {\tt kslowd001} & 103 \\ \hline
-2048 & 33 & {\tt kslowd001} & 102 \\ \hline
-2048 & 35 & {\tt kslowd001} & 102 \\ \hline
-2048 & 37 & {\tt kslowd001} & 103 \\ \hline
-2048 & 39 & {\tt kslowd001} & 103 \\ \hline
-2048 & 41 & {\tt kslowd001} & 103 \\ \hline
-2048 & 43 & {\tt kslowd001} & 103 \\ \hline
-2048 & 45 & {\tt kslowd001} & 103 \\ \hline
-2048 & 47 & {\tt kslowd001} & 102 \\ \hline
-2048 & 49 & {\tt kslowd001} & 102 \\ \hline
-2048 & 51 & {\tt kslowd001} & 103 \\ \hline
-2048 & 53 & {\tt kslowd001} & 103 \\ \hline
-2048 & 55 & {\tt kslowd001} & 102 \\ \hline
-2048 & 57 & {\tt kslowd001} & 103 \\ \hline
-2048 & 59 & {\tt kslowd001} & 104 \\ \hline
-2048 & 61 & {\tt kslowd001} & 103 \\ \hline
-2048 & 63 & {\tt kslowd001} & 102 \\ \hline
-2048 & 65 & {\tt kslowd001} & 103 \\ \hline
-2048 & 67 & {\tt kslowd001} & 103 \\ \hline
-2048 & 69 & {\tt kslowd001} & 102 \\ \hline
-2048 & 71 & {\tt kslowd001} & 102 \\ \hline
-2048 & 73 & {\tt kslowd001} & 103 \\ \hline
-2048 & 75 & {\tt kslowd001} & 103 \\ \hline
-2048 & 77 & {\tt kslowd001} & 102 \\ \hline
-2048 & 79 & {\tt kslowd001} & 104 \\ \hline
-2048 & 81 & {\tt kslowd001} & 103 \\ \hline
-2048 & 83 & {\tt kslowd001} & 103 \\ \hline
-2048 & 85 & {\tt kslowd001} & 103 \\ \hline
-2048 & 87 & {\tt kslowd001} & 103 \\ \hline
-2048 & 89 & {\tt kslowd001} & 103 \\ \hline
-2048 & 91 & {\tt kslowd001} & 102 \\ \hline
-2048 & 93 & {\tt kslowd001} & 102 \\ \hline
-2048 & 95 & {\tt kslowd001} & 103 \\ \hline
-2048 & 97 & {\tt kslowd001} & 103 \\ \hline
-2048 & 99 & {\tt kslowd001} & 102 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt kslowd001}'s Appearance at Odd Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_odd1}}
-\end{table}
-
-\begin{table}[htp!]
-\centering
-{
- \begin{tabular}{|l|l|l|l|} \hline
-Dual PUTx & Iter. \# & Proc. Name & Program Time (msec)\\ \hline
-2048 & 2 & {\tt kslowd001} & 102 \\ \hline
-2048 & 4 & {\tt kslowd001} & 103 \\ \hline
-2048 & 6 & {\tt kslowd001} & 103 \\ \hline
-2048 & 8 & {\tt kslowd001} & 102 \\ \hline
-2048 & 10 & {\tt kslowd001} & 103 \\ \hline
-2048 & 12 & {\tt kslowd001} & 103 \\ \hline
-2048 & 14 & {\tt kslowd001} & 102 \\ \hline
-2048 & 16 & {\tt kslowd001} & 102 \\ \hline
-2048 & 18 & {\tt kslowd001} & 103 \\ \hline
-2048 & 20 & {\tt kslowd001} & 104 \\ \hline
-2048 & 22 & {\tt kslowd001} & 102 \\ \hline
-2048 & 24 & {\tt kslowd001} & 102 \\ \hline
-2048 & 26 & {\tt kslowd001} & 103 \\ \hline
-2048 & 28 & {\tt kslowd001} & 103 \\ \hline
-2048 & 30 & {\tt kslowd001} & 104 \\ \hline
-2048 & 32 & {\tt kslowd001} & 103 \\ \hline
-2048 & 34 & {\tt kslowd001} & 103 \\ \hline
-2048 & 36 & {\tt kslowd001} & 102 \\ \hline
-2048 & 38 & {\tt kslowd001} & 102 \\ \hline
-2048 & 40 & {\tt kslowd001} & 103 \\ \hline
-2048 & 42 & {\tt kslowd001} & 103 \\ \hline
-2048 & 44 & {\tt kslowd001} & 102 \\ \hline
-2048 & 46 & {\tt kslowd001} & 102 \\ \hline
-2048 & 48 & {\tt kslowd001} & 103 \\ \hline
-2048 & 50 & {\tt kslowd001} & 103 \\ \hline
-2048 & 52 & {\tt kslowd001} & 102 \\ \hline
-2048 & 54 & {\tt kslowd001} & 103 \\ \hline
-2048 & 56 & {\tt kslowd001} & 103 \\ \hline
-2048 & 58 & {\tt kslowd001} & 102 \\ \hline
-2048 & 60 & {\tt kslowd001} & 102 \\ \hline
-2048 & 62 & {\tt kslowd001} & 103 \\ \hline
-2048 & 64 & {\tt kslowd001} & 103 \\ \hline
-2048 & 66 & {\tt kslowd001} & 102 \\ \hline
-2048 & 68 & {\tt kslowd001} & 103 \\ \hline
-2048 & 70 & {\tt kslowd001} & 103 \\ \hline
-2048 & 72 & {\tt kslowd001} & 103 \\ \hline
-2048 & 74 & {\tt kslowd001} & 103 \\ \hline
-2048 & 76 & {\tt kslowd001} & 103 \\ \hline
-2048 & 78 & {\tt kslowd001} & 103 \\ \hline
-2048 & 80 & {\tt kslowd001} & 102 \\ \hline
-2048 & 82 & {\tt kslowd001} & 102 \\ \hline
-2048 & 84 & {\tt kslowd001} & 103 \\ \hline
-2048 & 86 & {\tt kslowd001} & 103 \\ \hline
-2048 & 88 & {\tt kslowd001} & 103 \\ \hline
-2048 & 90 & {\tt kslowd001} & 103 \\ \hline
-2048 & 92 & {\tt kslowd001} & 104 \\ \hline
-2048 & 94 & {\tt kslowd001} & 103 \\ \hline
-2048 & 96 & {\tt kslowd001} & 102 \\ \hline
-2048 & 98 & {\tt kslowd001} & 104 \\ \hline
-2048 & 100 & {\tt kslowd001} & 103 \\ \hline
-\multicolumn{3}{|r|}{Total Appearances:} & 50 times\\ \hline
-  \end{tabular}
-  }
- \caption{{\tt kslowd001}'s Appearance at Even Iterations of dual PUT2048 with 100 measurements~\label{fig:rc_dual_put_2048_even1}}
-\end{table}
+		\label{fig:rc_on_new_dual2_put4096}
+	}
+	\caption{Complete measured times on dual PUT2048/PUT4096 in the presence/absence of {\tt rhn\_check}~\label{fig:rhn_check_effect3}}
+\end{figure}
 

### 2017-01-11 pt_characterization/paper.tex
pt char paper

\documentclass[10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{epsfig}
\usepackage{float}
\usepackage{subfigure}
\usepackage{algorithm}
\usepackage{algorithmic}
\def\ert{\mbox{{\it expRunTime}}}
\def\sertput{\mbox{{\it sendExpRunTimeToPUT}}}
\def\rpe{\mbox{{\it recordPUTExecution}}}
\def\procDiff{\mbox{{\it procDiff}}}
\def\getProcDiff{\mbox{{\it getProcDiff}}}
\def\bfps{\mbox{{\it beforeProcs}}}
\def\afps{\mbox{{\it afterProcs}}}
\def\gpi{\mbox{{\it getProcInfo}}}
\def\ect{\mbox{{\it elapsedClockTime}}}
\def\pt{\mbox{{\it ProgramTime}}}
\def\cpt{\mbox{{\it computeProgramTime}}}
\long\def\shorten#1{}
\begin{document}
\title{Characterization of Program Execution Time Distribution}

\author{
Young-Kyoon Suh\\
Department of Computer Science \\
University of Arizona \\
\small\tt \{yksuh\}@cs.arizona.edu
}
\maketitle

\section{Observed Phenomena}
In this section, we discuss interesting phenomena observed from our measurements on 
a program-under-test (PUT) with a given task length. 
PUT is a simple compute-only program that runs a nested-for loop for a given amount of time (task length) and terminates.

%For instance, PUT8 denotes PUT taking 8 seconds as its task length. 
%In this report, we use two types of program execution time: elapsed time (ET) and program time (PT). 
%ET is computed as the difference of the wall-clock times measured before and after PUT execution.
%PT is calculated as the sum of the user time and system time of PUT.

\subsection{Different Bands in Elapsed Time}
\label{sec:bands}

We introduce an interesting phenomenon regarding the elapsed time (ET) of PUT.
ET is computed as the difference of the wall-clock times measured before and after PUT execution.

\paragraph{Presumption} Little structure in ET

\paragraph{Description} Figure~\ref{fig:et_plot} shows the distribution of ET on PUT16 and PUT32. 
In this report, PUT$x$ denotes the PUT with $x$-second task length.
We observe from these data that some or many of ET records were concentrated within a certain interval, 
called {\em band} (or layer). The used data were obtained by EMPv2.

\begin{figure}[hp!]
	\centering
	\subfigure[Per-execution ET on PUT16]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/16_sec_et_all.eps}
		\label{fig:16_sec_ect}
	}
	\subfigure[Per-execution ET on PUT32]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/32_sec_et_all.eps}
		\label{fig:32_sec_ect}
	}
	\caption{Per-execution ET on PUT16 and PUT32~\label{fig:et_plot}}
\end{figure}

\paragraph{Analysis} As illustrated in Figure~\ref{fig:16_sec_ect}, 
there was the compact band formed around 16,650 msecs in the $y$-axis corresponding to most of the PUT executions. 
Another {\it weak} band around 16,900 msecs was formed as well. 
Furthermore, Figure~\ref{fig:32_sec_ect} demonstrates this similar occurrence observed in PUT32. 
Most of the $y$-axis values were clustered around 33,250 msecs. 
There were also other higher bands consisting of around 33,400 or 33,500 msecs. 
For each experiment we indeed were able to see an intensively-clustered band (formed by around the median ET measurements) 
and a few bands weakly formed above and below the main band. 
%In particular, Y-axis values belonging to a higher band than the main band seem periodic.

\paragraph{Hypothesis} ``{\it A high band seems formed 
due to daemon processes that periodically came up and ran for a moment, thus affecting 
ET at certain executions.}''

\paragraph{Test} 
When we looked at the data in PUT32, a raw ET value ({\it e.g}, 33,500 msecs) 
exceeding its median appeared at the 547th, 573th, 600th, and 626th. 
The same ET value was seen at the 731th, 758th, 916th, 942th, 968th, and 
995th. 
In other words, every 26 or 27 executions ET was higher than the median.
In PUT16, the measured high ET value was also observed every 51 executions. 
Based on these investigations, 
relatively more executed processes were included in the high ET measurements. 
Considering there were no other processes except PUT, 
we infer that some daemon processes must have contributed to forming such a high band. 
This hypothesis, therefore, is supported.

\paragraph{Status} This cause is understood.

\clearpage
\newpage

\subsection{Periodic Dips in PT}\label{sec:pd_pt}

We now turn our discussion to several phenomena regarding PT. 
PT is calculated as the sum of the user time and system time of PUT.

\paragraph{Presumption} PT is totally flat.

\paragraph{Description} 
In the measured data by EMPv2, 
we observed periodically-appearing dips. 
Figure~\ref{fig:pt_all} exhibits the distribution of PT in the experiments 
where the assigned task lengths were one, two, and four seconds.

\begin{figure}[H]
	\centering
	\subfigure[Per-execution PT on PUT1]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/1_sec_pt_all.eps}
		\label{fig:1_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/2_sec_pt_all.eps}
		\label{fig:2_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/4_sec_pt_all.eps}
		\label{fig:4_sec_pt_all}
	}
	\caption{Per-execution PT on PUT1, PUT2, and PUT4~\label{fig:pt_all}}
\end{figure}

\paragraph{Analysis}
At first, the calculated PT values were plotted around the median. 
At some point, however, the PT values dropped to the minimum, 
and then the values bounced back up to the median. 
This pattern repeated after a certain number of executions; 
namely, this dip seemed to occur every around $x$ seconds after the first appearance. 
Specifically, the dip first appeared after the 200th execution (about 200 seconds) and 
reappeared every approximately 400 seconds, as shown in Figure~\ref{fig:1_sec_pt_all}.
In PUT2, these dips first emerged around the 200th execution, 
and they reoccurred after around 800 seconds, as illustrated in Figure~\ref{fig:2_sec_pt_all}.
In Figure~\ref{fig:4_sec_pt_all}, 
the first dip was observed at around the 100th execution and repeated after approximately 800 seconds. 
Note that the $x$-axis values represents execution number, rather than actual time.
The same execution number across different tasks do not mean that the respective actual times are same. 
The 200th execution in PUT2 and PUT4 could happen around 400 seconds and 800 seconds, respectively, after the tasks start.
%Although weak, the dips seem to occur every around 400 seconds even in the eight-second task.

\paragraph{Hypothesis I} ``{\it These dips are caused by adjustments by NTP (Network Time Protocol)~\cite{Mills} used to 
synchronize its local system time to a server in the network.}''

This hypothesis is based on the assumption that PT will be faster whenever the system time is corrected 
by the NTP daemon ({\tt ntpd}).

\paragraph{Test} To confirm this, we ran some additional experiments in which {\tt ntpd} was disabled.
In these experiments, EMPv1 was used.

The experiment results are shown in Figure~\ref{fig:ntpd_off_pt}. 
Unfortunately, the dips did not go away. 
Conversely, overall dips became thinner and appeared more frequently. 
For instance, the dips shown in Figure~\ref{fig:ntp_off_1_sec_pt_all} were weaker and more frequent, 
compared with the ones in Figure~\ref{fig:1_sec_pt_all}. 
The same trend was seen in Figures~\ref{fig:ntp_off_2_sec_pt_all} and~\ref{fig:ntp_off_4_sec_pt_all} although less strong and somewhat seemingly random. 
In addition, the overall distribution of PT looked more dispersed, weakening the compactness of the bands.

\begin{figure}[H]
	\centering
	\subfigure[Per-execution PT on PUT1]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-off-turbo-on/1_sec_pt_all.eps}
		\label{fig:ntp_off_1_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-off-turbo-on/2_sec_pt_all.eps}
		\label{fig:ntp_off_2_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-off-turbo-on/4_sec_pt_all.eps}
		\label{fig:ntp_off_4_sec_pt_all}
	}
	\caption{Per-execution PT on PUT1, PUT2, and PUT4 when {\tt ntpd} was turned off~\label{fig:ntpd_off_pt}}
\end{figure}

Turning off {\tt ntpd} did not help us to figure out the cause of the dips. 
Our hypothesis, thus, is not supported.

\paragraph{Hypothesis II} ``{\it Disabling the Turbo mode set in a processor will make the dips go away.}''

Some processors support ``turbo mode'' and scale up the frequency and voltage at times. 
This hypothesis studies the impact of the frequency scaling caused by the turbo mode.

%Turbo mode dynamically increases the processor's frequency as needed by taking advantage of thermal and power headroom. 
%Speedstep is meant to reduce power usage/heat based on the load of the processor. (slow down an idle processor)
%Turbo boost increase processor speed based on load and temperature. So it will find the so called sweet spot between temperature and performance up to a certain clock speed.

\paragraph{Test} 
We found that the processor type of our experiment machine is Intel Core {\it i}7 CPU 870 @ 2.93GHz. 
This processor enabled by default the turbo mode through the feature named ``Intel$\textregistered$ Turbo Boost Technology''. 
In addition, the turbo mode can be influenced by another feature of the processor, 
called ``Enhanced Intel SpeedStep$\textregistered$ Technology'', which can reduce power usage/heat based on the load of the processor. 
To see if the Turbo mode was the cause of the dips, we disabled both features. The frequency scaling was stopped.

Figure~\ref{fig:turbo_off_pt} shows the distribution of PT on PUT1, PUT2, and PUT4. 
In these data, EMPv2 was used. 
Compared with Figure~\ref{fig:pt_all}, overall the degree of dips became much more discrete and weaker. 
In addition, the standard variations of PT on the PUT tasks reduced to 
1.65/2.78/5.75 from 2.39/4.33/7.52, respectively. 
This implies that turning off the turbo mode helped us reduce the variation of PT. 
As a result, most executions yielded flat PT measurements.
Nevertheless, the dips were still observed.

\begin{figure}[H]
	\centering
	\subfigure[Per-execution PT on PUT1]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/1_sec_pt_all.eps}
		\label{fig:tm_1_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/2_sec_pt_all.eps}
		\label{fig:tm_2_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/4_sec_pt_all.eps}
		\label{fig:tm_4_sec_pt_all}
	}
	\caption{Per-execution PT on PUT1, PUT2, PUT4, and PUT8 when the turbo mode was turned off~\label{fig:turbo_off_pt}}
\end{figure}

%As shown in Figures~\ref{fig:turbo_off_pt}, overall PT was flat although the dots look scattered due to resolution . 
%Specifically, the standard variations of PT in the 1/2/4-second tasks reduced to 1.65/2.78/5.75 from 2.39/4.33/7.52, respectively.

%Figure~\ref{fig:turbo_off_ect} shows the distribution of ET in the one-/two-/four-second tasks. 
%Interestingly, the 
%
%\begin{figure}[htp!]
%	\centering
%	\subfigure[Per-execution ET on the 1-second task]{
%		\includegraphics[scale=0.43]{figures/turbo_mode/1_sec_ect_all.eps}
%		\label{fig:tm_1_sec_ect_all}
%	}
%	\subfigure[Per-execution ET on the 2-second task]{
%		\includegraphics[scale=0.43]{figures/turbo_mode/2_sec_ect_all.eps}
%		\label{fig:tm_2_sec_ect_all}
%	}
%	\subfigure[Per-execution ET on the 4-second task]{
%		\includegraphics[scale=0.43]{figures/turbo_mode/4_sec_ect_all.eps}
%		\label{fig:tm_4_sec_ect_all}
%	}
%%	\subfigure[Per-execution PT on the 8-second task]{
%%		\includegraphics[scale=0.43]{figures/ntp-off/8_sec_pt_all.eps}
%%		\label{fig:ntp_off_8_sec_pt_all}
%%	}
%%	\caption{Per-execution PT on the 1/2/4/8-second task when {\tt ntpd} was turned off\label{fig:ntpd_off_pt}}
%	\caption{Per-execution ET on the 1/2/4-second task when Turbo mode was turned off\label{fig:turbo_off_ect}}
%\end{figure}
%
%They all look flat except a couple of executions.

\paragraph{Status} This cause is yet understood.

In subsequent sections, we show the experiment results after the Turbo mode was off.

\newpage

\subsection{Synchronized ET and PT}

\paragraph{Presumption} Totally correlated

\paragraph{Description} 
Figure~\ref{fig:sync_time} shows ET and PT measurements on PUT2 and PUT4. 
The common phenomenon is that most of the measured ET and PT data looked synchronized. 
The phenomenon was observed from the EMPv3 data.

\begin{figure}[H]
	\centering
	\subfigure[Per-execution ET on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/2_sec_et_all.eps}
		\label{fig:2_sec_ect}
	}
	\subfigure[Per-execution PT on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/2_sec_pt_all.eps}
		\label{fig:2_sec_pt_all2}
	}
	\subfigure[Per-execution ET on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/4_sec_et_all.eps}
		\label{fig:4_sec_ect}
	}
	\subfigure[Per-execution PT on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/4_sec_pt_all.eps}
		\label{fig:4_sec_pt_all2}
	}
	\caption{Per-execution ET/PT on PUT2 and PUT4~\label{fig:sync_time}}
\end{figure}

\paragraph{Analysis} In PUT2, if we overlap Figures~\ref{fig:2_sec_ect} and \ref{fig:2_sec_pt_all2}, 
we know that most executions have exactly the same ET and PT. 
This also holds true for PUT4, as shown in Figures~\ref{fig:4_sec_ect} and \ref{fig:4_sec_pt_all2}. 
The synchronization does not take place in some executions, where some other daemon processes had indeed run together.

%\paragraph{Hypothesis} ``{\it When the number of daemon processes is very low, ET and PT will be synchronized.}''. 
%
%\paragraph{Test} 
%For this test, we used EMPv2. 
%Figure~\ref{fig:daemon} shows the number of daemon processes observed when measuring the execution time of PUT2 and PUT4.
%In Figures~\ref{fig:2_sec_low_daemon} and \ref{fig:4_sec_low_daemon}, 
%we found that when fewer than three daemon processes (low) was present, 
%ET and PT measurements were exactly the same. 
%Figures~\ref{fig:2_sec_high_daemon} and \ref{fig:4_sec_high_daemon} exhibit the measurements with 
%more than three daemon processes (high). 
%For these measurements, both did not get synchronized; 
%namely, the measured ET values were greater than the measured PT values. 
%This hypothesis, therefore, is supported. 
%
%\begin{figure}[hp!]
%	\centering
%	\subfigure[Low number of daemon processes]{
%		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/2_sec_num_low_daemons.eps}
%		\label{fig:2_sec_low_daemon}
%	}
%	\subfigure[High number of daemon processes]{
%		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/2_sec_num_high_daemons.eps}
%		\label{fig:2_sec_high_daemon}
%	}
%	\subfigure[Low number of daemon processes]{
%		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/4_sec_num_low_daemons.eps}
%		\label{fig:4_sec_low_daemon}
%	}
%	\subfigure[High number of daemon processes]{
%		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/4_sec_num_high_daemons.eps}
%		\label{fig:4_sec_high_daemon}
%	}
%	\caption{Number of daemon processes per execution\label{fig:daemon}}
%\end{figure}
%
%\paragraph{Status} We understood this.

\newpage

\subsection{Outliers in PT}

\paragraph{Presumption} Large outlier in PT should not occur.

\paragraph{Description} 
This phenomenon indicates that a couple of PT measurements showed exceedingly high peaks.

\paragraph{Analysis} 
%In Figure~\ref{fig:pt_all}, we've already seen several outliers across different measurements.
%There is one outlier observed on PUT8, as shown in Figure~\ref{fig:1_sec_pt_all}, 
%and there are also several outliers seen in Figures~\ref{fig:1_sec_pt_all} and \ref{fig:4_sec_pt_all}. 
%Furthermore, we can see these abnormal PT in heavier tasks.

Figure~\ref{fig:peak_pt} exhibits the PT measurements on PUT16, PUT32, and PUT64. 
These data were obtained by EMPv3. 
There were three outliers observed at around the 90th and the 820th executions, as shown in Figure~\ref{fig:16_sec_pt_all}.
Likewise, a couple of outliers were also observed from the measurements on PUT32 and in PUT64, 
as illustrated in Figures~\ref{fig:32_sec_pt_all} and \ref{fig:64_sec_pt_all}.

\begin{figure}[H]
\centering
	\subfigure[Per-execution PT on PUT16]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/16_sec_pt_all.eps}
		\label{fig:16_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT32]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/32_sec_pt_all.eps}
		\label{fig:32_sec_pt_all}
	}
	\subfigure[Per-execution PT on PUT64]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-off/64_sec_pt_all.eps}
		\label{fig:64_sec_pt_all}
	}
	\caption{Per-execution PT on PUT8, PUT16, PUT32, and PUT64~\label{fig:peak_pt}}
\end{figure}

There are three common things that we can draw regarding this phenomenon. 
First of all, literally all the peaks are far above the main band. 
Second, every experiment reveals a few peaks. 
Lastly, their occurrence does not have any periodic cycle. In other words, 
the outliers are scattered all over the places. 
It looks irregular and appears randomly at a specific execution.

\paragraph{Hypothesis} ``{\it The outliers could be correlated with processes 
scheduled to run at some point during our experiment.}''\\

%\paragraph{Hypothesis} ``{\it Considering that there could be processes that can be scheduled 
%during the experiment, these peaks could be correlated to this.}''
%
We looked at the data revealing the peaks in every experiment. 
Interestingly, {\em CPU\_COUNT}, representing \# of context swaps, is positively 
correlated with peak. PT grows linearly with the increase of \# of context switches. 
The correlation coefficient of PT and the context switches is 0.67, which is quite high. 
(The correlation between PT and \# of run processes is also positive.) 
Hence, this hypothesis is supported.

\paragraph{Status} We have understood this.

\newpage

\subsection{Periodic Peaks in Frequency of PT}

\paragraph{Presumption} No peak except the median PT

\paragraph{Description} 
Figure~\ref{fig:pt_hist} shows the frequency of PT in the experiments 
whose respective expected running time is 4 and 8 seconds. 
This phenomenon indicates that a few peaks in frequency of PT are regularly seen. 
The phenomenon was observed from the EMPv2 data.

\begin{figure}[H]
	\centering
	\subfigure[PT frequency on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/4_sec_pt_hist.eps}
		\label{fig:4_sec_pt_hist}
	}
	\subfigure[PT frequency on PUT8]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/8_sec_pt_hist.eps}
		\label{fig:8_sec_pt_hist}
	}
	\caption{PT frequency on PUT4 and PUT8~\label{fig:pt_hist}}
\end{figure}

\paragraph{Analysis} In Figure~\ref{fig:4_sec_pt_hist}, we could see two distinct peaks 
formed around 4,138 msecs and 4,142 msecs before the highest peak around 4,155 msecs. 
This pattern was also observed in Figure~\ref{fig:8_sec_pt_hist}, 
where there were about three peaks shown at 8,275 msecs, 8,292 msecs, and 8,300 msecs 
before the highest one around 8,318 msecs. 
The peaks appeared to repeat every around 8 msecs. 
This periodicty seems formed by the same PT values on PUT4 and PUT8. 
In particular, these PT values were below the median. 
The low PT values were calculated when PT periodically went down to the minimum, 
as discussed in Section~\ref{sec:pd_pt}. 
As a consequence, periodic peaks in frequency were observed, as illustrated in Figure~\ref{fig:pt_hist}.

\paragraph{Hypothesis} ``{\it Heavier PUT will also have similar periodic peaks in frequency due to their low PT values.}''

\paragraph{Test} 
Such similar peaks in frequency periodically 
appeared in the measurements on PUT16 and PUT32, as depicted in Figure~\ref{fig:extra_pt_hist}. 
Again, EMPv2 was used for this test.

\begin{figure}[H]
	\centering
	\subfigure[PT frequency on PUT16]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/16_sec_pt_hist.eps}
		\label{fig:16_sec_pt_hist}
	}
	\subfigure[PT frequency on PUT32]{
		\includegraphics[scale=0.43]{figures/sodb8-ntp-on-turbo-on/32_sec_pt_hist.eps}
		\label{fig:32_sec_pt_hist}
	}
	\caption{PT frequency on PUT16 and PUT32~\label{fig:extra_pt_hist}}
\end{figure}

Therefore, this hypothesis is supported.

\paragraph{Status} The cause is understood.

\newpage

\subsection{Increasing Minimum and Maximum Difference in PT}
%Figure~\ref{fig:mm_pt_dist} shows the distance between the minimum and maximum PT 
%over increasing amount of work.

\paragraph{Presumption} Almost zero or very small constant

\paragraph{Description} 
This phenomenon indicates that the difference between the maximum and minimum of the measured PT values 
grows linearly over increasing task length, as shown in Figure~\ref{fig:mm_pt_dist}. 
Note that $xy$ axes are in log scale. 
This phenomenon was observed from the EMPv2 data.

% 60 =  1.778 => 0.058
% 89 = 1.9493 => 0.043
% 106 = 2.02  => 0.025
% 121 = 2.08  => 0.0146
% 195 = 2.29  => 0.0117
% 770 = 2.89  => 0.023
% 1038 = 3.01 => 0.0156

\begin{figure}[H]
\centering
\includegraphics[scale=0.8]{figures/sodb8-ntp-on-turbo-on/overall_pt_diff.eps}
\caption{Increasing difference over increasing task length~\label{fig:mm_pt_dist}}
\end{figure}

\paragraph{Analysis} As task length increases, the PT difference increases. 
Considering that the $y$-axis values are plotted in log scale, the difference grows linearly. 
It appears to be synchronized with the degree of increasing task length. 
Our expectation on PT was flat, and thus, the increasing difference was strange.

\paragraph{Hypothesis} ?

\paragraph{Test} ?

%\paragraph{Hypothesis} ``{\it Disabling the Turbo mode will show the flat PT difference.}''
%
%\paragraph{Test} Figure~\ref{fig:off_turbo_pt_dist} shows the PT difference when the Turbo mode was off.
%
%\begin{figure}[hp!]
%\centering
%\includegraphics[scale=0.9]{figures/turbo_mode/overall_pt_diff.eps}
%\caption{PT difference over increasing expected running time under the disabled Turbo mode\label{fig:off_turbo_pt_dist}}
%\end{figure}
%
%Unfortunately, overall the PT difference still grows over increasing expected running time even 
%after the mode was not enabled. It went down at the four-second task, compared to that of the two-second task. 
%However, the increase is still shown.

%\paragraph{Hypothesis} ``{\it Even if we increase the expected running time of PUT,
%the difference in PT will be constant.}''
%
%Based on our expectation that PT should be flat, we anticipated that 
%the minimum and maximum PT through 1000 executions did not differ much, regardless of how long PUT took. 
%However, we observed that the difference got bigger over increasing the degree of the task. 
%For instance, the difference started with 60 msecs in the one-second task. 
%As the expected running times increase, the difference also grows somewhat linearly. 
%The difference reaches up to around 1040 msecs in the sixty-four second work. 
%Thus, the hypothesis is not supported. 
%
\paragraph{Status} Not understood.
%
%\subsection{Miscellaneous}
%
%\begin{enumerate}
%\item Note that the X-axis in th
%\end{enumerate}
%
%
\vspace{-0.01in}

\newpage

\section{Exploring a Proper Distribution for Measured Data}

Figures~\ref{fig:extra_pt_hist1},~\ref{fig:extra_pt_hist2},~\ref{fig:extra_pt_hist3},~\ref{fig:extra_pt_hist4},~\ref{fig:extra_pt_hist5},~\ref{fig:extra_pt_hist6}, 
and~\ref{fig:extra_pt_hist7} represent the histograms of ET and PT on the 1 $\sim$ 4096 second tasks. 
Nnormal and beta distributions are fitted on the histograms as well.
For these histograms, we used EMPv5, which identified and removed outliers in its measurements.

\begin{figure}[H]
	\centering
	\subfigure[ET frequency on PUT1]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/1_sec_et_hist.eps}
		\label{fig:1_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT1]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/1_sec_pt_hist.eps}
		\label{fig:1_sec_pt_hist1}
	}
	\subfigure[ET frequency on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/2_sec_et_hist.eps}
		\label{fig:2_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT2]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/2_sec_pt_hist.eps}
		\label{fig:2_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT1 and PUT2~\label{fig:extra_pt_hist1}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[ET frequency on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/4_sec_et_hist.eps}
		\label{fig:4_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT4]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/4_sec_pt_hist.eps}
		\label{fig:4_sec_pt_hist1}
	}
	\subfigure[ET frequency on PUT8]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/8_sec_et_hist.eps}
		\label{fig:8_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT8]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/8_sec_pt_hist.eps}
		\label{fig:8_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT4 and PUT8~\label{fig:extra_pt_hist2}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[ET frequency on PUT16]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/16_sec_et_hist.eps}
		\label{fig:16_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT16]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/16_sec_pt_hist.eps}
		\label{fig:16_sec_pt_hist1}
	}
	\subfigure[ET frequency on PUT32]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/32_sec_et_hist.eps}
		\label{fig:32_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT32]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/32_sec_pt_hist.eps}
		\label{fig:32_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT16 and PUT32~\label{fig:extra_pt_hist3}}
\end{figure}

\begin{figure}
	\centering
	\subfigure[ET frequency on PUT64]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/64_sec_et_hist.eps}
		\label{fig:64_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT64]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/64_sec_pt_hist.eps}
		\label{fig:64_sec_pt_hist1}
	}
	\subfigure[ET frequency on PUT128]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/128_sec_et_hist.eps}
		\label{fig:128_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT128]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/128_sec_pt_hist.eps}
		\label{fig:128_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT64 and PUT128~\label{fig:extra_pt_hist4}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[ET frequency on PUT256]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/256_sec_et_hist.eps}
		\label{fig:256_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT256]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/256_sec_pt_hist.eps}
		\label{fig:256_sec_pt_hist1}
	}
	\subfigure[ET frequency on PUT512]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/512_sec_et_hist.eps}
		\label{fig:512_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT512]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/512_sec_pt_hist.eps}
		\label{fig:512_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT256 and PUT512~\label{fig:extra_pt_hist5}}
\end{figure}

\begin{figure}[H]
	\centering
	\subfigure[ET frequency on PUT1024]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/1024_sec_et_hist.eps}
		\label{fig:1024_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT1024]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/1024_sec_pt_hist.eps}
		\label{fig:1024_sec_pt_hist1}
	}
	\subfigure[ET frequency on PUT2048]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/2048_sec_et_hist.eps}
		\label{fig:2048_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT2048]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/2048_sec_pt_hist.eps}
		\label{fig:2048_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT1024 and PUT2048~\label{fig:extra_pt_hist6}}
\end{figure}

\begin{figure}[H]
	\centering
		\subfigure[ET frequency on PUT4096]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/4096_sec_et_hist.eps}
		\label{fig:4096_sec_et_hist1}
	}
	\subfigure[PT frequency on PUT4096]{
		\includegraphics[scale=0.43]{figures/sodb12-ntp-on-turbo-off/4096_sec_pt_hist.eps}
		\label{fig:4096_sec_pt_hist1}
	}
	\caption{Normal and beta distributions fitting onto the histograms of the measurements of PUT4096~\label{fig:extra_pt_hist7}}
\end{figure}

\bibliographystyle{abbrv}
\newcommand{\etalchar}[1]{$^{#1}$}
\begin{thebibliography}{99}
\bibitem{Mills} David L. Mills, ``Internet Time Synchronization: The Network Time Protocol,'' in {\it IEEE Trans. on Comm.}, Vol. 39, No. 10, pp.~1482--1493, October, 1991.
\end{thebibliography}
\end{document}

### 2017-01-11 pt_characterization/data.old/daemon_pt.r

@@ -1,0 +1,396 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("PUT1_pt.eps")
+ymin<- min(x$PRTIME)
+ymax <- max(x$PRTIME)
+plot(x$ITERNUM, x$PRTIME, ylim=c(ymin, ymax), 
+xlab='Iterations', ylab=expression('Process Time (ms)'))
+dev.off()
+
+postscript("PUT1_daemon_pt.eps")
+ymin<- min(x$TPRTIME-x$PRTIME)
+ymax <- max(x$TPRTIME-x$PRTIME)
+plot(x$ITERNUM, x$TPRTIME-x$PRTIME, ylim=c(ymin, ymax), 
+xlab='Iterations', ylab=expression('Process Time (ms)'))
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+postscript("PUT2_pt.eps")
+ymin<- min(x$PRTIME)
+ymax <- max(x$PRTIME)
+plot(x$ITERNUM, x$PRTIME, ylim=c(ymin, ymax+1), 
+xlab='Iterations', ylab=expression('Process Time (ms)'))
+dev.off()
+
+postscript("PUT2_daemon_pt.eps")
+ymin<- min(x$TPRTIME-x$PRTIME)
+ymax <- max(x$TPRTIME-x$PRTIME)
+plot(x$ITERNUM, x$TPRTIME-x$PRTIME, ylim=c(ymin, ymax), 
+xlab='Iterations', ylab=expression('Process Time (ms)'))
+dev.off()
+
+cor(x$TPRTIME-x$PRTIME, x$PRTIME)
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+postscript("PUT4_pt.eps")
+ymin<- min(x$PRTIME)
+ymax <- max(x$PRTIME)
+plot(x$ITERNUM, x$PRTIME, ylim=c(ymin, ymax), 
+xlab='Iterations', ylab=expression('Process Time (ms)'))
+dev.off()
+
+postscript("PUT4_daemon_pt.eps")
+ymin<- min(x$TPRTIME-x$PRTIME)
+ymax <- max(x$TPRTIME-x$PRTIME)
+plot(x$ITERNUM, x$TPRTIME-x$PRTIME, ylim=c(ymin, ymax), 
[…]

### 2017-01-11 pt_characterization/data.old/daemon_pt_empv5.r

@@ -1,0 +1,105 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+cor(x$TDIFF, x$PRTIME)
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+cor(x$TDIFF, x$PRTIME)
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+cor(x$TDIFF, x$PRTIME)
+
+
+x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+cor(x$TDIFF, x$PRTIME)
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+cor(x$TDIFF, x$PRTIME)
+
+x = read.csv(file="32_sec.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+cor(x$TDIFF, x$PRTIME)
+
[…]

### 2017-01-11 pt_characterization/data.old/dual_compile.sh

@@ -1,0 +1,4 @@
+#!/bin/sh
+gcc -o dual_incr_work -O3 dual_incr_work.c
+mv dual_incr_work ../incr_work
+

### 2017-01-11 pt_characterization/data.old/dual_daemon_anal.py

@@ -1,0 +1,34 @@
+import random
+import re
+import sys, getopt
+from sets import Set
+
+def main(argv):	
+	print 'Number of arguments:', len(sys.argv), 'arguments.'
+  	#print 'Argument List:', str(sys.argv)
+  	#if len(sys.argv) == 1:
+	#	print '[Usage] python gen_subquery.py input_file_name'
+	#	print 'Give an input query file.'
+	#	sys.exit(-1);
+   	#print '[Usage] python gen_subquery.py input1 input2 input3 input4'
+	for i in range(0,len(sys.argv)-1):
+		f = open(sys.argv[i+1], 'r')
+		in_list = [];
+		for line in f:
+			line = line.replace('\n', '');
+			in_list.append(str(line));
+		in_pred = "";
+		i = 0;
+		for item in in_list:
+			i+=1;
+			in_pred += item;
+			if i < len(in_list):
+				in_pred+=",";
+		finalStr = "IN (" + in_pred + ");";
+		print finalStr;
+if __name__ == "__main__":
+   main(sys.argv[1:])
+			
+			
+				
+			

### 2017-01-11 pt_characterization/data.old/dual_incr_work.c

@@ -1,0 +1,245 @@
+#include <stdio.h>    /* standard in and output*/
+#include <sys/socket.h> /* for socket() and socket functions*/
+#include <arpa/inet.h>  /* for sockaddr_in and inet_ntoa() */
+#include <stdlib.h>   
+#include <string.h>    
+#include <unistd.h>     /* for close() */
+#include <time.h> 
+#include <limits.h> 
+
+#define PORT_NUM 2400
+#define COEF 1460000000 // 16B
+#define MAX_INCR 420000000
+#define MAX_ITER 40
+int first_half[MAX_ITER];
+int second_half[MAX_ITER];
+char c;
+int f_iter = 0, s_iter=0;
+struct timeval tim;
+FILE* f_half = NULL;
+FILE* s_half = NULL;
+double t1=0, t2=0;
+#if 1 
+inline void foo(int exp_secs) {
+	unsigned long long i, j, x;
+	//printf("Increment work of %d seconds starts!\n", exp_secs);
+	//max = 2147483647;
+	//max = 2000000000; 
+	unsigned long long fn = (long)COEF*(long)exp_secs;
+	//printf("max iterations: %llu\n", n);
+	// first half
+	gettimeofday(&tim, NULL);
+	t1=tim.tv_sec*1000+(tim.tv_usec/1000.0);
+	for (i = 1; i <= fn/2; i++) {
+		x = 0;
+		for (j = 0; j < UINT_MAX; j++) {
+			x++;
+		}
+	}
+	gettimeofday(&tim, NULL);
+	t2=tim.tv_sec*1000+(tim.tv_usec/1000.0);
+	first_half[f_iter++] = (int)(t2-t1);
+	// second half
+	gettimeofday(&tim, NULL);
+	t1=tim.tv_sec*1000+(tim.tv_usec/1000.0);
+	for (i = 1; i <= fn/2; i++) {
+                x = 0;
+                for (j = 0; j < UINT_MAX; j++) {
+                        x++;
+                }
+        }
+	gettimeofday(&tim, NULL);
+        t2=tim.tv_sec*1000+(tim.tv_usec/1000.0);
+        second_half[s_iter++] = (int)(t2-t1);
+	//sleep(exp_secs);
+	//printf("Done!\n");
+}
+#else
+inline void foo(int exp_secs) {
+        unsigned long long i, j,k, x;
+        for (i = 1; i <= (long)exp_secs; i++) {
+                x = 0;
[…]

### 2017-01-11 pt_characterization/data.old/dual_put_hist.r

@@ -1,0 +1,1482 @@
+x = read.csv(file="new_dual2_PUT2048.dat",head=TRUE,sep="\t")
+ymax <- max(x$METIME)
+ymax#1026508
+
+
+x = read.csv(file="new_dual1_PUT4096.dat",head=TRUE,sep="\t")
+sd(x$METIME)
+setEPS()
+postscript("rc_off_dual1_put4096.eps")
+xmin<- min(x$METIME)
+xmin#2052469
+xmax<- max(x$METIME)
+xmax#2052652
+plot(x$ITERNUM, x$METIME, ylim=c(2052400, 2052900), 
+xlab='Iterations', main='Measured times in the first half of dual PUT4096 w/ rhn check off', 
+ylab=expression('Measured Time (ms)'))
+dev.off()
+
+y = read.csv(file="new_dual2_PUT4096.dat",head=TRUE,sep="\t")
+sd(y$METIME)
+setEPS()
+postscript("rc_off_dual2_put4096.eps")
+ymin<- min(y$METIME)
+ymin#2052472
+ymax<- max(y$METIME)
+ymax#2052818
+setEPS()
+plot(y$ITERNUM, y$METIME, ylim=c(2052400, 2052900), 
+xlab='Iterations', main='Measured times in the second half of dual PUT4096 w/ rhn check off', 
+ylab=expression('Measured Time (ms)'))
+dev.off()
+
+x = read.csv(file="new_dual1_PUT2048.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("rc_off_dual1_put2048.eps")
+ymin<- min(x$METIME)
+ymin#1026254
+ymax <- max(x$METIME)
+ymax#1026342
+plot(x$ITERNUM, x$METIME, ylim=c(1026250, 1026550), 
+xlab='Iterations', main='Measured times in the first half of dual PUT2048 w/ rhn check off', 
+ylab=expression('Measured Time (ms)'))
+dev.off()
+
+x = read.csv(file="new_dual1_PUT2048.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("rc_off_dual1_put2048_v2.eps")
+ymin<- min(x$METIME)
+ymin#1026254
+ymax <- max(x$METIME)
+ymax#1026342
+plot(x$ITERNUM, x$METIME, ylim=c(1026250, 1026550), 
+xlab='Iterations', main='Measured times in the first half of dual PUT2048 w/ rhn check off', 
[…]

### 2017-01-11 pt_characterization/data.old/iter_pair_pt.r

@@ -1,0 +1,334 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+postscript("1_sec_ip_pt.eps")
+odd_x <- subset(x, x$ITERNUM %% 2 == 1)
+even_x <- subset(x, x$ITERNUM %% 2 == 0)
+min_val <- min(min(odd_x$PRTIME), min(even_x$PRTIME)) - 1
+max_val <- max(max(odd_x$PRTIME), max(even_x$PRTIME)) + 1
+plot(odd_x$PRTIME, even_x$PRTIME, 
+xlim=c(min_val, max_val), ylim=c(min_val, max_val), 
+main='Iteration Dependency of PUT1', 
+xlab='Odd iteration\'s program time (ms)', ylab=expression('Even iteration\'s program time (ms)'))
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+postscript("2_sec_ip_pt.eps")
+odd_x <- subset(x, x$ITERNUM %% 2 == 1)
+even_x <- subset(x, x$ITERNUM %% 2 == 0)
+min_val <- min(min(odd_x$PRTIME), min(even_x$PRTIME)) - 4
+max_val <- max(max(odd_x$PRTIME), max(even_x$PRTIME)) + 4
+plot(odd_x$PRTIME, even_x$PRTIME, 
+xlim=c(min_val, max_val), ylim=c(min_val, max_val), 
+main='Iteration Dependency of PUT2', 
+xlab='Odd iteration\'s program time (ms)', ylab=expression('Even iteration\'s program time (ms)'))
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+postscript("4_sec_ip_pt.eps")
+odd_x <- subset(x, x$ITERNUM %% 2 == 1)
+even_x <- subset(x, x$ITERNUM %% 2 == 0)
+min_val <- min(min(odd_x$PRTIME), min(even_x$PRTIME)) - 2
+max_val <- max(max(odd_x$PRTIME), max(even_x$PRTIME)) + 2
+plot(odd_x$PRTIME, even_x$PRTIME, 
+xlim=c(min_val, max_val), ylim=c(min_val, max_val), 
+main='Iteration Dependency of PUT4', 
[…]

### 2017-01-11 pt_characterization/data.old/no_outlier_hist.r

@@ -1,0 +1,331 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist2.eps")
+x <- subset(x, x$ITERNUM <= 1000)
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+binsize=1
+xmin <-min(x$PRTIME)-1
+xmax <-max(x$PRTIME)+2
+h = hist(x$PRTIME, breaks=seq(xmin,xmax),right=F,plot=F)
+plot(h, xaxt='n', freq=TRUE,ylim=c(0,500),xlim=c(xmin,xmax),
+col="blue", main='PT frequency on PUT1', sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""),   xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin,xmax,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist2.eps")
+x <- subset(x, x$ITERNUM <= 1000)
+x_up = mean(x$PRTIME) + 5*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 5*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nrow(x)
+max(x$PRTIME)
+min(x$PRTIME)
+xmin <-min(x$PRTIME)-1
+xmax <-max(x$PRTIME)+2
+h = hist(x$PRTIME, breaks=seq(xmin,xmax),right=F,plot=F)
+max(h$counts)
+plot(h, xaxt='n', freq=TRUE,ylim=c(0,600),xlim=c(xmin,xmax),col="blue", main='PT frequency on PUT2', sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""),   xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin,xmax,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
[…]

### 2017-01-11 pt_characterization/data.old/procs_dual_put4096.py

@@ -1,0 +1,51 @@
+import random
+import re
+import sys, getopt
+from sets import Set
+
+def main(argv):	
+	#print 'Number of arguments:', len(sys.argv), 'arguments.'
+	flag = 0;
+	if flag:
+		for i in range(0,len(sys.argv)-1):
+			f = open(sys.argv[i+1], 'r')
+			in_list = [];
+			for line in f:
+				line = line.replace('\n', '');
+				in_list.append(str(line));
+				j = 0;
+				finalStr = "";
+				for item in in_list:
+					j+=1;
+					item = item.replace('_', '\_');
+					finalStr += "{\\tt " + item + "}";
+					#finalStr += item;
+					if j < len(in_list):
+						finalStr+=", ";
+			print finalStr, "\n";
+	else:
+		daemon_map = {} 			
+		for i in range(0,len(sys.argv)-1):
+			f = open(sys.argv[i+1], 'r')
+			in_list = [];
+			for line in f:
+				line = line.replace('\n', '');
+				in_list.append(str(line));
+			for item in in_list:
+				if item not in daemon_map:
+					daemon_map.setdefault(item, []);
+				daemon_map[item].append(str(i+1))
+	  	for key in sorted(daemon_map):
+			size = 0;
+			strVal ="";			
+			for val in daemon_map[key]:
+				#print "Size: ", len(daemon_map[key]);
+				strVal += val;
+				size+=1;
+				if (size < len(daemon_map[key])):
+					strVal += ",";
+			key2 = key.replace('_', '\_');
+			print "{\\tt %s} & %s \\\\ \hline" % (key2, strVal)
+				#print daemon_map;
+if __name__ == "__main__":
+   main(sys.argv[1:])

### 2017-01-11 pt_characterization/data.old/put16_std_ss.r

@@ -1,0 +1,251 @@
+x = read.csv(file="16_sec_1k.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_1k_hist.eps")
+binsize=2
+nbins = (max(x$PRTIME)-min(x$PRTIME)+1)/binsize
+max(x$PRTIME)
+min(x$PRTIME)
+h = hist(x$PRTIME, right=F,breaks=nbins,plot=F)
+max(h$counts)
+xmin <- min(x$PRTIME)-3
+xmax <- max(x$PRTIME)+3
+plot(h, xaxt='n', freq=TRUE,xlim=c(xmin,xmax),
+ylim=c(0, 500), col="green", main='PT frequency on PUT16_1K', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin,xmax,by=binsize)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+## empv5
+x = read.csv(file="16_sec_1k.dat",head=TRUE,sep="\t")
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+setEPS()
+postscript("16_sec_1k_hist2.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+max(h$counts)
+max(x$PRTIME)
+min(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+3),
+ylim=c(0, 300), col="blue", main='PT frequency on PUT16_1K (by EMPv5)', sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
+#empv4
+x = read.csv(file="16_sec_2k.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_2k_hist.eps")
+binsize=2
[…]

### 2017-01-11 pt_characterization/data.old/put_20k_run_hist.r

@@ -1,0 +1,225 @@
+x = read.csv(file="put1_20k.txt",head=TRUE,sep="\t")
+setEPS()
+postscript("put1_20k_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,12000),xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+1),col="green", main='PT frequency on PUT1_20K', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(min(x$PRTIME)-1,max(x$PRTIME)+1,by=5)
+axis(side=1, at=xaxtl, labels=xaxtl)
+#axis(side=1, at=seq(min(h$mids)-1,max(h$mids)+1,1), labels=seq(min(x$PRTIME),max(x$PRTIME)+1,by=1))
+dev.off()
+
+x = read.csv(file="put1_20k.txt",head=TRUE,sep="\t")
+setEPS()
+postscript("put1_20k_hist2.eps")
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+binsize=1
+xmin <-min(x$PRTIME)-2
+xmax <-max(x$PRTIME)+2
+h = hist(x$PRTIME, breaks=seq(xmin,xmax),right=F,plot=F)
+plot(h, xaxt='n', freq=TRUE,ylim=c(0,12000),xlim=c(xmin,xmax),col="blue", main='PT frequency on PUT1_20K', sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""),   xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin,xmax,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="put2_20k.txt",head=TRUE,sep="\t")
+#x = subset(org, org$PRTIME >= 2000)
+#y = subset(org, org$PRTIME < 2000)
[…]

### 2017-01-11 pt_characterization/data.old/put_iter_corr.r

@@ -1,0 +1,174 @@
+x = read.csv(file="dual1_PUT2.dat",head=TRUE,sep="\t")
+#x <- subset(x, x$ITERNUM <=10)
+y = read.csv(file="dual2_PUT2.dat",head=TRUE,sep="\t")
+#y <- subset(y, y$ITERNUM <=10)
+total <- c("ITER_PAIR", "CORR_VAL")
+max <- 1000
+for(i in 2:max)
+{
+	x1 <- subset(x, x$ITERNUM == i-1)
+	x2 <- subset(x, x$ITERNUM == i)
+	y1 <- subset(y, y$ITERNUM == i-1)
+	y2 <- subset(y, y$ITERNUM == i)
+	first <- rbind(x1, x2)
+	second <- rbind(y1, y2)
+	c <- cor(first$METIME, second$METIME)
+	if(is.na(c)){
+	  c <- 0
+	}
+	total <- rbind(total, c(i-1, c))
+}
+nrow(total)
+write.table(total, "2_sec_ip_cor.dat", sep="\t", row.names=FALSE, col.names=FALSE) 
+
+x = read.csv(file="2_sec_ip_cor.dat",head=TRUE,sep="\t")
+postscript("2_sec_ip_cor.eps")
+plot(x$ITER_PAIR, x$CORR_VAL, xlim=c(0, 1000), ylim=c(-2, 2), 
+xlab='Iteration Pair', main='Iteration Dependency of dual PUT2', ylab=expression('Correlation Efficient (-1 ~ 1)'))
+dev.off()
+
+
+x = read.csv(file="dual1_PUT16.dat",head=TRUE,sep="\t")
+#x <- subset(x, x$ITERNUM <=10)
+y = read.csv(file="dual2_PUT16.dat",head=TRUE,sep="\t")
+#y <- subset(y, y$ITERNUM <=10)
+total <- c("ITER_PAIR", "CORR_VAL")
+max <- 1000
+for(i in 2:max)
+{
+	x1 <- subset(x, x$ITERNUM == i-1)
+	x2 <- subset(x, x$ITERNUM == i)
+	y1 <- subset(y, y$ITERNUM == i-1)
+	y2 <- subset(y, y$ITERNUM == i)
+	first <- rbind(x1, x2)
+	second <- rbind(y1, y2)
+	c <- cor(first$METIME, second$METIME)
+	if(is.na(c)){
+	  c <- 0
+	}
+	total <- rbind(total, c(i-1, c))
+}
+nrow(total)
[…]

### 2017-01-11 pt_characterization/data.old/regular_hist.r

@@ -1,0 +1,242 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+#plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+1),col="green", main='PT frequency on PUT1', 
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+1),col="green", main='PT frequency on PUT1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(min(x$PRTIME)-1,max(x$PRTIME)+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+#axis(side=1, at=seq(min(h$mids)-1,max(h$mids)+1,1), labels=seq(min(x$PRTIME),max(x$PRTIME)+1,by=1))
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+1),col="green", main='PT frequency on PUT2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+2),col="green", main='PT frequency on PUT4', 
[…]

### 2017-01-11 pt_characterization/data.old/regular_put_iter.r

@@ -1,0 +1,62 @@
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+total <- NULL
+max <- 1000
+for(i in 2:max)
+{
+	x1 <- subset(x, x$ITERNUM == i-1)
+	x2 <- subset(x, x$ITERNUM == i)
+	c <- x2$METIME-x1$METIME
+	total <- rbind(total, c(i-1, c))
+}
+cols <- c("ITER_PAIR","DIFF")
+colnames(total) <- cols
+nrow(total)
+postscript("16_sec_ip_diff.eps")
+#write.table(total, "16_sec_iter_diff.dat", sep="\t", row.names=FALSE, col.names=FALSE) 
+plot(total,  
+main='Iteration Dependency of PUT2', 
+xlim=c(1, 999), 
+ylim=c(-550, 550), 
+xlab='Iteration Pair', 
+ylab=expression('Program time difference (ms)'))
+dev.off()
+
+
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+total <- c("ITER_PAIR", "CORR_VAL")
+max <- 10
+for(i in 2:max)
+{
+	x1 <- subset(x, x$ITERNUM == i-1)
+	x2 <- subset(x, x$ITERNUM == i)
+	c <- cor(x1$METIME, x2$METIME)
+	if(is.na(c)){
+	  c <- 0
+	}
+	total <- rbind(total, c(i-1, c))
+}
+nrow(total)
+write.table(total, "16_sec_ip_cor.dat", sep="\t", row.names=FALSE, col.names=FALSE) 
+#In cor(first$METIME, second$METIME) : the standard deviation is zero	
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+x1 <- subset(x, x$ITERNUM == 2)
+x2 <- subset(x, x$ITERNUM == 3)
+c <- cor(x1$METIME, x2$METIME)
+
+x1 <- subset(x, x$ITERNUM == 84)
+y1 <- subset(y, y$ITERNUM == 84)
+x2 <- subset(x, x$ITERNUM == 85)
+y2 <- subset(y, y$ITERNUM == 85)
+first <- rbind(x1, y1)
+second <- rbind(x2, y2)
+c <- cor(first$METIME, second$METIME)
+> first
+    ITERNUM METIME
+84       84   8019
+841      84   8019
+    ITERNUM METIME
+85       85   8015
+851      85   8019
+

### 2017-01-11 pt_characterization/figures/sodb12-ntp-on-turbo-off/put_hist.r

@@ -1,0 +1,1067 @@
+x.org = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x.sub <- subset(x.org, x.org$ITERNUM <= 300)
+x_up = mean(x.sub$METIME) + 2*sd(x.sub$METIME)
+x_dn = mean(x.sub$METIME) - 2*sd(x.sub$METIME)
+x = subset(x.sub, x.sub$METIME >= x_dn & x.sub$METIME <= x_up)
+setEPS()
+postscript("1_sec_et_hist.eps")
+h = hist(x$METIME, right=F, breaks=max(x$METIME)-min(x$METIME)+1, plot=F)
+plot(h, ylim=c(0,200), freq=TRUE,  xaxt='n', col="yellow", main='', xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids)),ceiling(max(h$mids)))
+axis(side=1, at=xaxtl, labels=xaxtl)
+xfit<-seq(min(x$METIME),max(x$METIME),length=40)
+yfit<-dnorm(xfit,mean=mean(x$METIME),sd=sd(x$METIME))
+yfit <- yfit*diff(h$mids[1:2])*length(x$METIME)
+lines(xfit, yfit, col="red", lwd=2)
+### perform normalization of x #####
+x_normal = (x$METIME-min(x$METIME))/(max(x$METIME)-min(x$METIME))
+x_normal = sort(x_normal)
+### get histogram on normalized x #####
+h_normal = hist(x_normal, breaks=length(h$breaks),right=F, plot=F)
+### get mean/variance of normalized x #####
+m = mean(x_normal)
+v = var(x_normal)
+estBetaParams <- function(mu, var) {
+  alpha <- ((1 - mu) / var - 1 / mu) * mu ^ 2
+  beta <- alpha * (1 / mu - 1)
+  return(params = list(alpha = alpha, beta = beta))
+}
+### get parameters a, b from beta distribution on normalized x #####
+r <- estBetaParams(m,v)
+a <- r[1]; b <- r[2]
+a <- as.double(a)
+b <- as.double(b)
+### get new x-axis values by beta distribution of normalized x #####
+y_beta<-dbeta(x_normal,a,b)
+### fit x_val and y_val #####
+x_val = sort(x$METIME)
[…]

### 2017-01-11 pt_characterization/figures/sodb8-ntp-off-turbo-on/all_pt.r

@@ -1,0 +1,18 @@
+
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(1020, 1100), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(2060, 2160), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(4120, 4260), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()

### 2017-01-11 pt_characterization/figures/sodb8-ntp-on-turbo-off/all_et.r

@@ -1,0 +1,26 @@
+
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(1100, 1300), pch=1, main='', xlab='# of executions', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(2200, 2500), pch=1, main='', xlab='# of executions', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(4500, 5000), pch=1, main='', xlab='# of executions', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(9000, 9300), pch=1, main='', xlab='# of executions', ylab='ET (msec)')
+dev.off()
+
+#plot(x$METIME, xlim=c(0,1000), ylim=c(min(x$METIME), max(x$METIME)), pch=1, main='', xlab='# of executions', ylab='ET (msec)')

### 2017-01-11 pt_characterization/figures/sodb8-ntp-on-turbo-off/all_pt.r

@@ -1,0 +1,42 @@
+
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(1100, 1300), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(2200, 2500), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(4500, 5000), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(9000, 9100), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(18050, 18300), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="32_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("32_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(36000, 37200), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()
+
+x = read.csv(file="64_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("64_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), ylim=c(72000, 75500), pch=1, main='', xlab='# of executions', ylab='PT (msec)')
+dev.off()

### 2017-01-11 pt_characterization/figures/sodb8-ntp-on-turbo-on/all_et.r

@@ -1,0 +1,36 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(min(x$METIME), max(x$METIME)), pch=1, main='', xlab='Iterations', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(min(x$METIME), max(x$METIME)), pch=1, main='', xlab='Iterations', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(min(x$METIME), max(x$METIME)), pch=1, main='', xlab='Iterations', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(min(x$METIME), max(x$METIME)), pch=1, main='', xlab='Iterations', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(16400,17400), pch=1, main='', xlab='Iterations', ylab='ET (msec)')
+dev.off()
+
+x = read.csv(file="32_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("32_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), ylim=c(33000, 34000), pch=1, main='', xlab='Iterations', ylab='ET (msec)')
+dev.off()
+

### 2017-01-11 pt_characterization/figures/sodb8-ntp-on-turbo-on/all_hist.r

@@ -1,0 +1,56 @@
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+h = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("4_sec_pt_hist.eps")
+counts = h$counts
+breaks = h$breaks[0:(length(h$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', col="green", log='y', type='h', lwd=5, lend=2, main='', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000,by=10)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+
+x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
+setEPS()
+h = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("8_sec_pt_hist.eps")
+counts = h$counts
+breaks = h$breaks[0:(length(h$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', col="green", log='y', type='h', lwd=5, lend=2, main='', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000,by=10)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+setEPS()
+h = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("16_sec_pt_hist.eps")
+counts = h$counts
+breaks = h$breaks[0:(length(h$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', col="green", log='y', type='h', lwd=5, lend=2, main='', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
[…]

### 2017-01-11 pt_characterization/new_data/emp_pt_hist.r

@@ -1,0 +1,236 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(min(x$PRTIME)-1,max(x$PRTIME)+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+#axis(side=1, at=seq(min(h$mids)-1,max(h$mids)+1,1), labels=seq(min(x$PRTIME),max(x$PRTIME)+1,by=1))
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+1),col="green", main='PT frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+2),col="green", main='PT frequency on INC4', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
[…]

### 2017-01-11 pt_characterization/new_data/emp_v5_hist.r

@@ -1,0 +1,341 @@
+x = read.csv(file="128_sec.dat",head=TRUE,sep="\t")
+#x <- subset(x, x$ITERNUM <= 10)
+uol = mean(x$METIME) + 2*sd(x$METIME)
+x.sub = subset(x, x$METIME <= uol)
+x.ol = subset(x, x$METIME > uol)
+nrow(x.sub)
+nrow(x.ol)
+uol = mean(x.sub$PRTIME) + 2*sd(x.sub$PRTIME)
+dol = mean(x.sub$PRTIME) - 2*sd(x.sub$PRTIME)
+x.ref = subset(x.sub, x.sub$PRTIME <= uol & x.sub$PRTIME >= dol)
+x.ol2 = subset(x.sub, x.sub$PRTIME > uol | x.sub$PRTIME < dol)
+nrow(x.ref)
+nrow(x.ol2)
+min(x.ref$PRTIME)
+max(x.ref$PRTIME)
+round(mean(x.ref$PRTIME),0)
+round(sd(x.ref$PRTIME),1)
+sd(x.ref$PRTIME)/mean(x.ref$PRTIME)
+
+### 1 sec.dat
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= 10)
+#x <- subset(x, x$ITERNUM <= 800)
+uol = mean(x$METIME) + 2*sd(x$METIME)
+x.sub = subset(x, x$METIME <= uol)
+x.ol = subset(x, x$METIME > uol)
+nrow(x.sub)
+nrow(x.ol)
+uol = mean(x.sub$PRTIME) + 2*sd(x.sub$PRTIME)
+dol = mean(x.sub$PRTIME) - 2*sd(x.sub$PRTIME)
+x.ref = subset(x.sub, x.sub$PRTIME <= uol & x.sub$PRTIME >= dol)
+x.ol2 = subset(x.sub, x.sub$PRTIME > uol | x.sub$PRTIME < dol)
+nrow(x.ref)
+nrow(x.ol2)
+min(x.ref$PRTIME)
+max(x.ref$PRTIME)
+round(mean(x.ref$PRTIME),0)
+round(sd(x.ref$PRTIME),1)
+sd(x.ref$PRTIME)/mean(x.ref$PRTIME)
+
+### 2 sec.dat
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
[…]

### 2017-01-11 pt_characterization/resources/fitting/1024_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/1024_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/1024_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,300), pch=19, main='ECT on 1024-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/1024_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', type='h', lwd=5, lend=2, main='ECT frequency on 1024-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/1024_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,300), pch=19, main='PT on 1024-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/1024_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', type='h', lwd=5, lend=2, main='PT frequency on 1024-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/128_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/128_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/128_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,300), pch=19, main='ECT on 128-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/128_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', type='h', lwd=5, lend=2, main='ECT frequency on 128-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/128_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,300), pch=19, main='PT on 128-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/128_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', type='h', lwd=5, lend=2, main='PT frequency on 128-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/16_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,100), pch=19, main='ET on 16-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("16_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', log='y', type='h', lwd=5, lend=2, main='ECT frequency on 16-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,100,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("16_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,100), pch=19, main='PT on 16-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("16_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', log='y', type='h', lwd=5, lend=2, main='PT frequency on 16-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,100, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/1_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/1_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='ET (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/1_sec_et_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', log='y', type='h', lwd=5, lend=2, main='', xlab='ET (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/1_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/1_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', log='y', type='h', lwd=5, lend=2, main='', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/2048_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/2048_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/2048_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,300), pch=19, main='ECT on 2048-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/2048_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', type='h', lwd=5, lend=2, main='ECT frequency on 2048-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/2048_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,300), pch=19, main='PT on 2048-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/2048_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', type='h', lwd=5, lend=2, main='PT frequency on 2048-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/256_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/256_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/256_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,300), pch=19, main='ECT on 256-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/256_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', type='h', lwd=5, lend=2, main='ECT frequency on 256-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/256_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,300), pch=19, main='PT on 256-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/256_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', type='h', lwd=5, lend=2, main='PT frequency on 256-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/2_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/2_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='ET (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/2_sec_et_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', log='y', type='h', lwd=5, lend=2, main='', xlab='ET (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/2_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/2_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', log='y', type='h', lwd=5, lend=2, main='', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/32_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/32_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/32_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,100), pch=19, main='ECT on 32-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/32_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', log='y', type='h', lwd=5, lend=2, main='ECT frequency on 32-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,100,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/32_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,100), pch=19, main='PT on 32-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/32_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', log='y', type='h', lwd=5, lend=2, main='PT frequency on 32-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,100, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/4096_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/4096_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/4096_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,300), pch=19, main='ECT on 4096-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/4096_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', type='h', lwd=5, lend=2, main='ECT frequency on 4096-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/4096_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,300), pch=19, main='PT on 4096-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/4096_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', type='h', lwd=5, lend=2, main='PT frequency on 4096-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/4_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/4_sec_et_all.eps")
+plot(x$METIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='ET (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/4_sec_et_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', log='y', type='h', lwd=5, lend=2, main='', xlab='ET (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/4_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/4_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', log='y', type='h', lwd=5, lend=2, main='', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,1000, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/512_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/512_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/512_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,300), pch=19, main='ECT on 512-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/512_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', type='h', lwd=5, lend=2, main='ECT frequency on 512-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/512_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,300), pch=19, main='PT on 512-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/512_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', type='h', lwd=5, lend=2, main='PT frequency on 512-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,300, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/64_sec.r

@@ -1,0 +1,34 @@
+x = read.csv(file="result_summary/64_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("result_summary/64_sec_ect_all.eps")
+plot(x$METIME, xlim=c(0,100), pch=19, main='ECT on 64-second task', xlab='# of executions', ylab='ECT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$METIME, breaks = 100, plot=F)
+postscript("result_summary/64_sec_ect_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', yaxt='n', log='y', type='h', lwd=5, lend=2, main='ECT frequency on 64-second task', xlab='ECT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(0,ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,100,by=100)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+setEPS()
+postscript("result_summary/64_sec_pt_all.eps")
+plot(x$PRTIME, xlim=c(0,100), pch=19, main='PT on 64-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+setEPS()
+hist.data = hist(x$PRTIME, breaks = 100, plot=F)
+postscript("result_summary/64_sec_pt_hist.eps")
+counts = hist.data$counts
+breaks = hist.data$breaks[0:(length(hist.data$breaks)-1)]
+plot(breaks, counts, xaxt='n', log='y', type='h', lwd=5, lend=2, main='PT frequency on 64-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+delta=max(breaks)-min(breaks)
+xaxtl <- seq(floor(min(breaks)-delta/2),ceiling(max(breaks)+delta/2),by=(ceiling((delta/2)/100)*5))
+axis(side=1, at=xaxtl, labels=xaxtl)
+axt <- seq(0,100, by=50)
+axis(side=2, at=axt, labels=axt)
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/distribution_exploration.r

@@ -1,0 +1,171 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F, breaks=max(x$PRTIME)-min(x$PRTIME)+1, plot=F)
+plot(h, ylim=c(0,600), freq=TRUE,  xaxt='n', col="green", main='PT frequency on the 1-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids)),ceiling(max(h$mids)))
+axis(side=1, at=xaxtl, labels=xaxtl)
+xfit<-seq(min(x$PRTIME),max(x$PRTIME),length=40)
+yfit<-rnorm(xfit,mean=mean(x$PRTIME),sd=sd(x$PRTIME))
+yfit <- yfit*diff(h$mids[1:2])*length(x$PRTIME)
+lines(xfit, yfit, col="blue", lwd=2)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F, breaks=max(x$PRTIME)-min(x$PRTIME)+1, plot=F)
+plot(h, ylim=c(0,600), freq=TRUE,  xaxt='n', col="green", main='PT frequency on the 2-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids))-10,ceiling(max(h$mids))+10)
+axis(side=1, at=xaxtl, labels=xaxtl)
+xfit<-seq(min(x$PRTIME),max(x$PRTIME),length=40)
+yfit<-dnorm(xfit,mean=mean(x$PRTIME),sd=sd(x$PRTIME))
+yfit <- yfit*diff(h$mids[1:2])*length(x$PRTIME)
+lines(xfit, yfit, col="blue", lwd=2)
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F, breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
[…]

### 2017-01-11 pt_characterization/resources/fitting/misc.r

@@ -1,0 +1,19 @@
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_ect.eps")
+plot(x$METIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='ET (ms)')
+dev.off()
+setEPS()
+
+x = read.csv(file="32_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("32_sec_ect.eps")
+plot(x$METIME, xlim=c(0,1000), pch=19, main='', xlab='# of executions', ylab='ET (ms)')
+dev.off()
+setEPS()
+
+setEPS()
+postscript("16_sec_pt.eps")
+plot(x$PRTIME, xlim=c(0,100), pch=19, main='PT on 16-second task', xlab='# of executions', ylab='PT (ms)')
+dev.off()
+

### 2017-01-11 pt_characterization/resources/fitting/normal_fitting.r

@@ -1,0 +1,172 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F, breaks=max(x$PRTIME)-min(x$PRTIME)+1, plot=F)
+plot(h, ylim=c(0,600), freq=TRUE,  xaxt='n', col="green", main='PT frequency on the 1-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids)),ceiling(max(h$mids)))
+axis(side=1, at=xaxtl, labels=xaxtl)
+xfit<-seq(min(x$PRTIME),max(x$PRTIME),length=40)
+yfit<-dnorm(xfit,mean=mean(x$PRTIME),sd=sd(x$PRTIME))
+yfit <- dgamma(yfit, shape=1.5)
+yfit <- yfit*diff(h$mids[1:2])*length(x$PRTIME)
+lines(xfit, yfit, col="blue", lwd=2)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F, breaks=max(x$PRTIME)-min(x$PRTIME)+1, plot=F)
+plot(h, ylim=c(0,600), freq=TRUE,  xaxt='n', col="green", main='PT frequency on the 2-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids))-10,ceiling(max(h$mids))+10)
+axis(side=1, at=xaxtl, labels=xaxtl)
+xfit<-seq(min(x$PRTIME),max(x$PRTIME),length=40)
+yfit<-dnorm(xfit,mean=mean(x$PRTIME),sd=sd(x$PRTIME))
+yfit <- yfit*diff(h$mids[1:2])*length(x$PRTIME)
+lines(xfit, yfit, col="blue", lwd=2)
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F, breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
[…]

### 2017-01-11 pt_characterization/rscripts/beta_dist.r

@@ -1,0 +1,469 @@
+###### 1 second task #######
+
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_beta.eps")
+h = hist(x$PRTIME, right=F, plot=F)
+### perform normalization of x #####
+x_normal = (x$PRTIME-min(x$PRTIME))/(max(x$PRTIME)-min(x$PRTIME))
+x_normal = sort(x_normal)
+
+### get histogram on normalized x #####
+h_normal = hist(x_normal, breaks=length(h$breaks),right=F, plot=F)
+### get mean/variance of normalized x #####
+m = mean(x_normal)
+v = var(x_normal)
+estBetaParams <- function(mu, var) {
+  alpha <- ((1 - mu) / var - 1 / mu) * mu ^ 2
+  beta <- alpha * (1 / mu - 1)
+  return(params = list(alpha = alpha, beta = beta))
+}
+### get parameters a, b from beta distribution on normalized x #####
+r <- estBetaParams(m,v)
+a <- r[1]; b <- r[2]
+a <- as.double(a)
+b <- as.double(b)
+
+### get new x-axis values by beta distribution of normalized x #####
+y_beta<-dbeta(x_normal,a,b)
+
+### fit x_val and y_val #####
+plot(h, freq=TRUE, ylim=c(0,200),xaxt='n', col="green", main='PT frequency on the 2-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids))-2,ceiling(max(h$mids))+1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+x_val = sort(x$PRTIME)
+y_val = min(h$counts) + ((y_beta)/(max(y_beta)-min(y_beta))) * (max(h$counts)-min(h$counts))
+lines(x_val,y_val, col="blue", lwd=2)
+dev.off()
+setEPS()
+
+
+###### 2 second task #######
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_beta.eps")
+h = hist(x$PRTIME, right=F, plot=F)
+### perform normalization of x #####
+x_normal = (x$PRTIME-min(x$PRTIME))/(max(x$PRTIME)-min(x$PRTIME))
+x_normal = sort(x_normal)
+
+### get histogram on normalized x #####
[…]

### 2017-01-11 pt_characterization/rscripts/beta_dist2.r

@@ -1,0 +1,75 @@
+#### 16 second task #####
+
+x = read.csv(file="16_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_pt_beta.eps")
+h = hist(x$PRTIME, right=F, plot=F)
+### perform normalization of x #####
+x_normal = (x$PRTIME-min(x$PRTIME))/(max(x$PRTIME)-min(x$PRTIME))
+### get histogram on normalized x #####
+h_normal = hist(x_normal, breaks=length(h$breaks),right=F, plot=F)
+plot(h, freq=TRUE, ylim=c(0,400),xaxt='n', col="green", main='PT frequency on the 16-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids))-2,ceiling(max(h$mids))+1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+### get mean/variance of normalized x #####
+m = mean(x_normal)
+v = var(x_normal)
+estBetaParams <- function(mu, var) {
+  alpha <- ((1 - mu) / var - 1 / mu) * mu ^ 2
+  beta <- alpha * (1 / mu - 1)
+  return(params = list(alpha = alpha, beta = beta))
+}
+### get parameters a, b from beta distribution on normalized x #####
+r <- estBetaParams(m,v)
+a <- r[1]; b <- r[2]
+a <- as.double(a)
+b <- as.double(b)
+### get new x-axis values by beta distribution of normalized x #####
+x_beta<-dbeta(x_normal,a,b)
+### get another histogram on the new x-axis values #####
+h_beta = hist(x_beta, breaks=length(h$breaks),right=F, plot=F)
+### denormalize to the raw x-axis values from the x-axis values by beta distribution #####
+x_val = min(x$PRTIME) + ((h_beta$mids)/(max(h_beta$breaks)-min(h_beta$breaks))) * (max(x$PRTIME)-min(x$PRTIME))
+### get frequency data of the histogram on the new x-axis values #####
+y_val = h_beta$counts
+### fit x_val and y_val #####
+lines(x_val,y_val, col="blue", lwd=2)
+dev.off()
+
+#### 32 second task #####
+
+x = read.csv(file="32_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("32_sec_pt_beta.eps")
+h = hist(x$PRTIME, right=F, plot=F)
[…]

### 2017-01-11 pt_characterization/rscripts/beta_dist3.r

@@ -1,0 +1,78 @@
+
+#### 16 second task #####
+
+x = read.csv(file="16_sec.txt",head=TRUE,sep="\t")
+setEPS()
+postscript("16_sec_pt_beta.eps")
+h = hist(x$PRTIME, right=F, plot=F)
+### perform normalization of x #####
+x_normal = (x$PRTIME-min(x$PRTIME))/(max(x$PRTIME)-min(x$PRTIME))
+### get histogram on normalized x #####
+h_normal = hist(x_normal, breaks=length(h$breaks),right=F, plot=F)
+plot(h, freq=TRUE, ylim=c(0,400),xaxt='n', col="green", main='PT frequency on the 16-second task', xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(floor(min(h$mids))-2,ceiling(max(h$mids))+1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+### get mean/variance of normalized x #####
+m = mean(x_normal)
+v = var(x_normal)
+estBetaParams <- function(mu, var) {
+  alpha <- ((1 - mu) / var - 1 / mu) * mu ^ 2
+  beta <- alpha * (1 / mu - 1)
+  return(params = list(alpha = alpha, beta = beta))
+}
+### get parameters a, b from beta distribution on normalized x #####
+r <- estBetaParams(m,v)
+a <- r[1]; b <- r[2]
+a <- as.double(a)
+b <- as.double(b)
+### get new x-axis values by beta distribution of normalized x #####
+x_beta<-dbeta(x_normal,a,b)
+### get another histogram on the new x-axis values #####
+h_beta = hist(x_beta, breaks=length(h$breaks),right=F, plot=F)
+### denormalize to the raw x-axis values from the x-axis values by beta distribution #####
+x_val = min(x$PRTIME) + ((h_beta$mids)/(max(h_beta$breaks)-min(h_beta$breaks))) * (max(x$PRTIME)-min(x$PRTIME))
+### get frequency data of the histogram on the new x-axis values #####
+y_val = h_beta$counts
+### fit x_val and y_val #####
+lines(x_val,y_val, col="blue", lwd=2)
+dev.off()
+
+#### 32 second task #####
+
+x = read.csv(file="32_sec.txt",head=TRUE,sep="\t")
+setEPS()
+postscript("32_sec_pt_beta.eps")
+h = hist(x$PRTIME, right=F, plot=F)
[…]

### 2017-01-11 pt_characterization/rscripts/pt_csw.r

@@ -1,0 +1,62 @@
+x = read.csv(file="result_summary/1_sec.dat",head=TRUE,sep="\t")
+postscript("result_summary/1_sec.eps")
+plot(x$IVCSW+x$VCSW, x$PRTKS, pch=19, xaxt='n', yaxt='n', main='Context Switches (CSws) vs. Program Time (PT) on the 1-second task', xlab='# of CSw', ylab='PT (ticks)')
+axt <- seq(min(x$PRTKS),max(x$PRTKS),by=1)
+axis(side=2, at=axt, labels=axt)
+xaxtl <- seq(min(x$IVCSW+x$VCSW)-10,max(x$IVCSW+x$VCSW)+10,by=10)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+setEPS()
+
+x = read.csv(file="result_summary/1_sec.dat",head=TRUE,sep="\t")
+postscript("result_summary/1_sec2.eps")
+new_x <- subset(x, x$IVCSW+x$VCSW < max(x$IVCSW+x$VCSW)) 
+plot(new_x$IVCSW+new_x$VCSW, new_x$PRTKS, pch=19, xaxt='n', yaxt='n', main='CSws vs. PT on the 1-second task (with the max CSw removed)', xlab='# of CSw', ylab='PT (ticks)')
+axt <- seq(min(new_x$PRTKS),max(new_x$PRTKS))
+axis(side=2, at=axt, labels=axt)
+xaxtl <- seq(min(new_x$IVCSW+new_x$VCSW),max(new_x$IVCSW+new_x$VCSW))
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+setEPS()
+
+x = read.csv(file="result_summary/64_sec.dat",head=TRUE,sep="\t")
+postscript("result_summary/64_sec.eps")
+plot(x$IVCSW+x$VCSW, x$PRTKS, pch=19, xaxt='n', yaxt='n', main='Context Switches (CSws) vs. Program Time (PT) on the 64-second task', xlab='# of CSw', ylab='PT (ticks)')
+axt <- seq(min(x$PRTKS),max(x$PRTKS),by=1)
+axis(side=2, at=axt, labels=axt)
+xaxtl <- seq(min(x$IVCSW+x$VCSW)-20,max(x$IVCSW+x$VCSW)+20,by=20)
+axis(side=1, at=xaxtl, labels=xaxtl)
[…]

### 2017-01-20 pt_characterization/new_data/empv5_et_hist.r
characterizing draft

@@ -1,0 +1,295 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_et_hist.eps")
+xmax <- max(x$METIME)
+xmin <- min(x$METIME)
+h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_et_hist.eps")
+xmax <- max(x$METIME)
+xmin <- min(x$METIME)
+h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_et_hist.eps")
+xmax <- max(x$METIME)
+xmin <- min(x$METIME)
+h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC4', 
[…]

### 2017-01-20 pt_characterization/new_data/empv5_pt_hist.r

@@ -1,0 +1,261 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+xmax <- max(x$PRTIME)
+xmin <- min(x$PRTIME)
+nbins <- ceiling((xmax-xmin) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+plot(h, ylim=c(0,500), xaxt='n',freq=TRUE,xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=binsize)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+3),col="green", main='PT frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist_v5.eps")
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
[…]

### 2017-01-24 pt_characterization/new_data/empv5_pt_hist.r
EMPv5 data

@@ -98,9 +98,16 @@ dev.off()
 
 x = read.csv(file="128_sec.dat",head=TRUE,sep="\t")
 setEPS()
-postscript("128_sec_pt_hist.eps")
-h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+postscript("128_sec_pt_hist_v5.eps")
+x <- subset(x, x$ITERNUM != 90 & x$ITERNUM != 202)
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
 binsize=1
+xmax <- max(x$PRTIME)
+xmin <- min(x$PRTIME)
+nbins <- ceiling((xmax-xmin) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
 plot(h, ylim=c(0,80), axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-4,max(x$PRTIME)+5),col="green", main='PT frequency on INC128', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='PT (ms)', ylab=expression('Frequency'))
@@ -108,36 +115,63 @@ dev.off()
 
 x = read.csv(file="256_sec.dat",head=TRUE,sep="\t")
 setEPS()
-postscript("256_sec_pt_hist.eps")
-h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+postscript("256_sec_pt_hist_v5.eps")
+x <- subset(x, x$ITERNUM != 6 & x$ITERNUM != 62 & x$ITERNUM != 118 & x$ITERNUM != 171 & x$ITERNUM != 202 & x$ITERNUM != 258)
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
 binsize=1
[…]

### 2017-01-25 sedona-ieice16/submission/misc/cover-letter.tex
sedona paper added

% Cover letter using letter.sty
\documentclass{letter} % Uses 10pt
%Use \documentstyle[newcent]{letter} for New Century Schoolbook postscript font
% the following commands control the margins:
\topmargin=-1in    % Make letterhead start about 1 inch from top of page 
\textheight=8in  % text height can be bigger for a longer letter
\oddsidemargin=0pt % leftmargin is 1 inch
\textwidth=6.5in   % textwidth of 6.5in leaves 1 inch for right margin

\begin{document}

\signature{Young-Kyoon Suh}           % name for signature 
\longindentation=0pt                       % needed to get closing flush left
\let\raggedleft\raggedright                % needed to get date flush left

\begin{letter}{
%Editors-in-Chief of ACM Transactions on Computer Systems (TOCS): \\
%\begin{tabular}{ll}
%\vspace{.1in}
%
%Prof. Todd C. Mowry\\
%Computer Science Department\\
%Carnegie Mellon University\\
%5000 Forbes Avenue\\
%Pittsburgh, PA 15213 \\
%USA\\ 
}

\begin{flushleft}
{\large\bf Young-Kyoon Suh}
\end{flushleft}
\medskip\hrule height 1pt
%\begin{flushright}
%\hfill Department of Computer Science, The University of Arizona, Tucson, AZ 85721 USA\\
%\hfill Tel: +1-520-370-1865
%(Please understand we cannot reveal our affiliation, as the manuscript should be double-blind.)
%\end{flushright} 
\begin{flushright}
\hfill Supercomputing R\&D Center, 
\end{flushright}
\vspace{-.15in}
\begin{flushright}
\hfill Korea Institute of Science and Technology Information (KISTI), 
\end{flushright}
\vspace{-.15in}
\begin{flushright}
\hfill Republic of Korea, 34141, 
\end{flushright}
\vspace{-.15in}
\begin{flushright}
\hfill Tel: +82-42-869-0725
\end{flushright}
\vspace{-.15in}
\begin{flushright}
\hfill Email: yksuh@kisti.re.kr
\end{flushright} 
%\vfill % forces letterhead to top of page

%\vspace{-0.5in}

\opening{Dear Editor-in-Chief:}

\noindent This letter to be submitted to IEICE Transactions on Information and Systems 
presents the following contributions. No version of this letter was submitted earlier.

\begin{itemize}

\item We provide empirical evidence that measuring program time 
can be seriously affected by extant system daemons.

\item We propose a novel timing protocol, called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons), 
that identifies infrequent, long-running daemons that impact the timing results for that program.

\item We evaluate the performance of the protocol with rigorous experiments, starting from a simple program in pure-computation mode 
to a popular industrial benchmark suite.

\item Experimental results show a strong support for the effectiveness of our SEDONA timing protocol. 
\end{itemize}

\closing{Sincerely yours,}

\end{letter}

\end{document}

### 2017-01-25 sedona-ieice16/submission/misc/paperRTS.tex
sedona paper added

\documentclass[letter]{ieice}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{balance}
\usepackage{graphicx}          % graphicx package for including ps files  
\usepackage{epsfig}
\usepackage{url}
\usepackage{subfigure}
\usepackage{algorithmic}
\usepackage{float}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{latexsym}
\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{comment}
%\usepackage{pstricks}
\usepackage{epstopdf}

\setcounter{page}{1}
%\breakauthorline{}% breaks lines after the n-th author

\field{D}
%\SpecialIssue{}
%\SpecialSection{}
%\theme{}
\title{SEDONA: A Novel Protocol for Identifying Infrequent, Long-Running Daemons on a Linux System}
%\titlenote{This work was in part supported by the EDISON program (NRF-2011-0020576).}
%\titlenote{}
\authorlist{
 \authorentry[yksuh@kisti.re.kr]{Young-Kyoon SUH}{m}{labelA}
 %\authorentry[rts@cs.arizona.edu]{Richard T. Snodgrass}{n}{labelB}}
\affiliate[labelA]{Supercomputing R\&{D} Center, KISTI, 245 Daehangno, Yuseong-gu, Daejeon, 34141, Republic of Korea}
%\affiliate[labelB]{Department of Computer Science, University of Arizona, Tucson, AZ 85721, USA}
}
\received{2016}{11}{30}
%\revised{2015}{10}{31}

\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\long\def\todo#1{\hilitey{{\bf TODO:} {\em #1}}}
\long\def\shorten#1{}
\long\def\comment#1{}

\begin{document}

\maketitle

\begin{summary}
It is challenging to accurately measure the execution-time of a given
program due to many extraneous factors. To enable better timing results
restrained, we present a novel protocol that identifies infrequent,
long-running daemons to enable effective elimination of executions
exhibiting these daemons.

\end{summary}
\begin{keywords}
Infrequent Long-running Daemon, Execution-Time Measurement
\end{keywords}

\section{Introduction}
\label{sec:intro}

Measuring program execution time is a much-used
technique for performance evaluation in computer science. 
Despite the importance of accurate and precise execution-time measurement, 
how to achieve {\em better} timing has not been well addressed. 
Surprisingly, there is considerable variability in the measured time, 
which indeed suggests the need for a better timing protocol.

Consider a simple compute-bound program, termed {\em PUT} (program-under-test), 
as shown in Fig.~\ref{alg:put}. 
This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
The $tl$ value is translated to the corresponding 
number of iterations for which that for-loop is performed.

\vspace{-.25in}
\begin{figure}[h]
\begin{center}
\begin{algorithmic}
{\bf Algorithm} PerformManyIncrements($tl$): \\
\STATE $t$ = TranslateToNumRepeats($tl$) \\
\FOR{$k$ = 1 to $t$ by 1}
	\FOR{$i$ = 1 to {UINT\_MAX}-1 by 1}
		\STATE $j$ += 1 \\
	\ENDFOR 
\ENDFOR 
\end{algorithmic}
\end{center}
\caption{Task Definition by a Program-Under-Test (PUT)\label{alg:put}}
\vspace{-.25in}
\end{figure}

\begin{figure}[t]
	\centering
	\subfigure[Initial Elapsed Time Measurement]{
		\includegraphics[width=0.32\textwidth]{128_sec_et_old_hist.eps}
		\label{fig:hist_before_et}
	}
	\subfigure[Program Time After Identified Daemon Removal]{
		\includegraphics[width=0.32\textwidth]{128_sec_pt_new_hist.eps}
		\label{fig:hist_after_pt}
	}
	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
	Program-Under-Test~\label{fig:meas_comp}}
	\vspace{-.3in}
\end{figure}

We used 128~seconds as the task length ran the PUT program (designated as PUT128) 
800~times. Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on the PUT128.
As can be seen in Fig.\ref{fig:hist_before_et}, 
most of the 800 executions measured in elapsed time, 
via  {\tt System.currentTimeMillis()} in Java, are gathered in the tallest bar 
of \hbox{128,000--129,000 msec,} with just a few executions outside of that bar. 
But these outliers seriously hurt the overall measurement quality: 
a standard deviation of 1,700 msec and a relative error of 0.03. 
\shorten{This result suggests adequately eliminating these outliers 
while retaining most of the measured executions.}

In contrast, our scheme, which we propose shortly, 
significantly reduces such variability.
The gist of the scheme is to (i)~focus on
{\em process time} (PT) rather than on elapsed time (ET) and 
(ii)~remove some executions that involve
{\em infrequent, \hbox{long-running} daemon processes}, which impact
the process time of the process. 
%(Typically, one tick is equivalent to 1/100 secs, or 1 {\tt HZ}.) 
Using PT is preferred, in that it considers 
the time taken for only a process of interest.
PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
in user and system mode, available via {\tt taskstats} C struct, 
provided by the Linux NetLink facility~\cite{Netlink}.
This use of PT and selective elimination 
results in improved 
measurement quality---in both standard deviation and relative error---
by three orders of magnitude, 
as illustrated in Fig.~\ref{fig:hist_after_pt}.

{\bf Related Work.} 
McGeoch introduced
two basic methods of measuring program time: elapsed time and CPU time~\cite{Mcgeoch12}. 
Bryant and O'Hallaron~\cite{Randal03} 
presented two timing schemes of using clock-cycle and interval counters. 
They proposed a measurement protocol, called {\em minimum-of-k}, 
that for observed elapsed ticks the minimum is chosen as the most accurate one. 
Odom et al.'s work~\cite{Odom05} focuses on timing long-running programs 
in a simulation framework via 
dynamic sampling of trace snippets during program execution. 
None of these prior works, however, takes into account the variability in 
timing and the influence of daemons that may significantly disturb the timing.

%Some prior works~\cite{Burns93,Grund11,Santos10,Wilhelm08} discuss 
%the worst case execution time (WCET) of a process. 
%The papers focused on measuring the WCET of a deadline-sensitive process 
%on real-time embedded systems while being oblivious to 
%daemons coexisting with the process. In addition, we do not consider real-time constraints.

Commercial software tools measure execution time~\cite{VTune,TimeSys,WindView}. 
Since the tools' source code is not disclosed, there is no way of figuring out 
whether they can prevent such a daemon from timing.

We previously developed a timing protocol called TTP (Tucson Timing Protocol)
for programs exhibiting I/O, in particular, query execution time in \hbox{DBMSes}~\cite{Currim}.
Our study identified a variety of Linux measures 
(e.g., user ticks, system ticks, IOWait ticks, etc.) 
relevant to timing a single query and then 
presented a structural causal model 
of explicating the variance of query time. 
Based on this model, we proposed the protocol for calculating the query time. 
Our scheme we propose in this manuscript 
is applicable to the TTP protocol, to further improve the query time calculation.

{\bf Contribution.} 
Our contributions are following.
\vspace{-0.05in}
\begin{itemize}

\item We provide empirical evidence that 
measuring program time 
can be seriously affected by extant system daemons.

\item We propose a novel timing protocol that
identifies infrequent, long-running daemons that impact the timing results for that program. 
%This scheme determines in a disciplined way cutoffs to remove executions with such daemons.

\item We evaluate the performance of the protocol with rigorous experiments, 
starting from a simple program in pure-computation mode 
to a popular industrial benchmark suite.

%\item We examine how the FIND scheme applies to process time as well as elapsed time.

\item The experimental results show a demonstrate for the effectiveness of our scheme\shorten{SEDONA: 
early eliminating dirty executions infected with such a daemon 
and eventually contributing to enhancing the overall measurement-quality}.

\end{itemize}
\vspace{-0.05in}

\noindent
The rest of this letter is organized as follows. 
Section~\ref{sec:prop_appach} elaborates on the proposed scheme. 
In the following section, we evaluate the performance of 
the scheme using real workloads.

%\vspace\fill

\section{Proposed Scheme}
\label{sec:prop_appach}

In this section we propose a novel execution-time measurement scheme, 
called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons). 
Our scheme catches and eliminates executions including daemon processes that are infrequent and 
long-running via a {\em cutoff} measure, and significantly improves measurement quality. 
The SEDONA scheme consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
%We first enumerate the steps to (a)~identify infrequent, 
%long-running daemons, using a single run of many executions of PUT128, 
%(b)~refine the list using a single run of many executions of PUT16384 (with a task length of 16K sec), 
%and then (c)~use the data from those two runs to determine the cutoff for 
%each \hbox{so-identified} daemon. 
%We can then apply these cutoffs to remove outliers from subsequent runs of any 
%arbitrary PUT and realistic workloads.
%\vspace{-0.22in}

\begin{figure}[h]
\begin{center}
\begin{algorithmic}
{\bf Algorithm} The SEDONA Timing Scheme: \\
\STATE Step 1. Set up the timing environment.
\STATE Step 2. Perform a single PUT run (specifically, PUT128) for many samples (specifically, 800).
\STATE Step 3. Consider each pair of elapsed time measurements to be a
dual-PUT measurement 
and examine a scatter-plot to see if it it displays an $L$-shape.
\STATE Step 4. Zoom into the central cluster to ensure that it is symmetric (roughly circular).
\STATE Step 5. Compute the maximum and standard deviation of the process time 
for each daemon encountered within the central cluster samples.
\STATE Step 6. Identify for each sample in the $L$-shape infrequent, long-running daemon executions. 
\STATE Step 7. Determine potentially periodic daemons based on the $L$-executions 
and for each daemon compute the minimum process time from those executions identified. 
\STATE Step 8. Perform Steps 1--6 above for a single run 
consisting of a small number of executions (specifically, 40) of PUT16384.  
\STATE Step 9. Compute the cutoffs for each identified daemon. 
\STATE Step 10. Discard an execution including a daemon of which process
time is greater than the respective cutoff time.
\end{algorithmic}
\end{center}
\caption{Summary of the SEDONA Scheme\label{alg:find}}
\vspace{-0.28in}
\end{figure}
%

%We now elaborate the SEDONA scheme with an example, to explain and justify each step.
%\vspace{-0.8in}

%\subsection{A Running Example} 
%\vspace{-0.05in}
{\bf A Running Example.} 
As motivated by our prior work~\cite{Currim}, 
we configure a timing environment by i) stopping non-critical daemons, 
ii) activating the Network Timing Protocol daemon, and 
iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1). 
%(Our machine's specification will be given in Tab~\ref{tab:machine_config}).

We then run a program-under-test (called PUT) of 
Fig.~\ref{alg:put} with a task length of 128~sec, 
termed {\em PUT128}, 800 times (Step 2). 
(We render Fig.~\ref{fig:hist_before_et} using the 800 samples from this run.)
We use 128 seconds because that is long enough to perhaps experience an infrequent daemon. 
We run it 800 times to capture infrequent daemons that perhaps run every few hours or even 
once a day. 
Note that we collect all daemon processes as well as the PUT and their measures 
through the Netlink interface from the kernel before and after each 
timing.

Fig.~\ref{fig:reg_put128} plots all the 800 elapsed times of the run of PUT128.
%We see three rows in the plot: a solid row of many samples, perhaps
%six or more samples that are just above that solid row, and two
%samples that are way above the solid row. 
The plot clearly shows three rows; that is, 
the top and middle rows represent over a dozen of outliers far from 
the rest of the samples clustered in the bottom row. 
We'll now drill down into these outliers to 
show how to reliably eliminate the indirect influence of 
some ``infrequent, long-running daemons'' on the process time (PT) 
of the PUT.

To identify such daemons, we use a novel scatter plot: 
those of {\em pairs of successive samples}.\shorten{So samples 1 and 2 
form the first pair, samples 3 and 4 form the second pair.} 
Fig.~\ref{fig:raw_put128} presents such a scatter plot 
of 400 samples of a \hbox{dual-PUT256} constructed 
from a run of the 800 PUT128 samples (Step~3). 
There are two quite obvious outliers\shorten{, corresponding to sample \# 75 and \# 634,} with ETs of 163,913 msec (rightmost) and 161,785 msec (uppermost), respectively.

\begin{figure}[t]
	\centering
	\subfigure[PUT128 with 800 Samples]{
		%\includegraphics[width=0.223\textwidth]{put128_empv4.eps}
		\includegraphics[width=0.223\textwidth]{put128_empv4_new.eps}
		\label{fig:reg_put128}
	}
	\subfigure[Dual-PUT256 Measurements]{
		%\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4.eps}
		\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4_new.eps}
		\label{fig:raw_put128}
	}
	%\vspace{-0.1in}    
	\subfigure[Zooming in on the $L$-shape]{
		\includegraphics[width=0.223\textwidth]{zooming_in_dual_et_put256_empv4.eps}
		\label{fig:dual_put256_liters}
	}
%	\subfigure[Further Zooming in on the $L$-Shape]{
%		\includegraphics[scale=0.39]{further_zooming_in_dual_et_put256_empv4.eps}
%		\label{fig:fzoom_in_dual_et_put256}
%	}
	\subfigure[The Central Cluster]{
		\includegraphics[width=0.223\textwidth]{dual_et_clump_put256_empv4.eps}
		\label{fig:dual_et_put256_cluster}
        }
    \caption{Successive Scatter plots of a PUT128 with 800 samples 
    (equivalent to a Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
    \vspace{-0.3in}     
\end{figure}

We informally term this phenomenon of a scatter plot of a dual-PUT run an
``$L$-shape,'' and attribute it to the
presence of infrequent long-running daemons.\shorten{Again, samples containing such
daemons (termed the $L$-samples) will appear
along the left y-axis or along the bottom x-axis, but will not occur in the
top right of the scatter plot of the dual-PUT.}

Figure~\ref{fig:dual_put256_liters} zooms into the lower left region,
focusing on the
tight cluster of samples.  Interestingly, this plot continues to exhibit an
$L$-shape, with perhaps a dozen or more $L$-samples in the left and bottom arms of
the ``$L$,'' and again no samples in the upper right portion of the
scatter plot.

We continue zooming until we get to Figure~\ref{fig:dual_et_put256_cluster},
which shows a central cluster (Step 4). 
We confirm the symmetry of the ET measurements in the central cluster: there
is no $L$-shape, and thus no $L$-samples, and thus no obvious infrequent
long-running daemons.\shorten{(We emphasize that these samples may have lots of
frequent daemons, as well as infrequent, {\em short-running} daemons.) There
were 384 dual-PUT256 samples in this central cluster, implying that 16 of
the PUT128 samples were $L$-samples and the 785 remaining of the original 800 PUT128
samples had no infrequent, long-running daemons, which motivates that term.}

We then perform Step~5, which computes 
the maximum process time and standard deviation 
of PT (process time, note the switch in emphasis from elapsed time 
to process time) of the daemon processes 
(e.g. {\tt cifsd}, {\tt flush-9:0}, {\tt jbd2/md0-8}, 
{\tt kblockd/0}, {\tt khugepaged}, {\tt md0\_raid1}, and {\tt ntpd})
observed in the central cluster samples in Fig.~\ref{fig:dual_et_put256_cluster}. 
\shorten{In the running example the maximum PT of those daemons was under 4 msec, 
and their standard deviation was just about 1 msec.}

In Step~6 we identify, for each daemon in the \hbox{$L$-samples}, 
those that are actual long-running daemon
executions.  We define such executions as those whose PT 
is over two standard deviations above the maximum PT for
that daemon in the central cluster samples. 
In the running example  {\tt flush-9:0}, 
{\tt jbd2/md0-8}, and {\tt md0\_raid1} 
are determined as infrequent, long-running.
We also identify ``extra'' infrequent daemons: those
found only in the $L$-samples but not in the central cluster. 
The running example reveals the following extra daemons: 
{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
{\tt rhsmcertd-worke}, and {\tt sshd}.

For each of the infrequent daemons 
we use a heuristic to determine the daemon's periodicity: 
the daemon must occur regularly in a sequence of samples.
For instance, the {\tt rhn\_check} daemon appears 
roughly every 112 samples (which corresponds to very close to every four 
hours). Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, 
{\tt md0\_raid1}, and {\tt rhn\_check}) all occur together 
(in those two outliers in Fig.~\ref{fig:raw_put128}) and 
  have a periodicity of about every 559 samples
  (5x longer, or just about 20 hours). 
%  It is important only which samples a given daemon shows up 
%  in; we don't care in this heuristic whether daemons are co-occurring.

Next, we can compute for each so-identified infrequent, long-running daemon its
minimum time in the \hbox{$L$-samples} (Step 7). 
This computation provides a rough, initial distinction of a ``long-running''
daemon, namely, the valley between the maximum PT from the central cluster 
and the minimum PT from the $L$-samples, to differentiate ``short-running'' 
from ``\hbox{long-running}'' executions of the daemon. 
For those daemons (i.e. {\tt grep}) never appearing in the central cluster,
this initial analysis concludes only that they are infrequent.

In Step 8 we repeat Steps 1--6, but instead with the much-longer running
PUT16384 (4.5 hours per sample versus 2 minutes), to see 
if any of our identified infrequent daemons are actually frequent 
at that much longer PUT execution time. 
We find some frequent daemon processes appearing in both of the 
clusters of the dual-PUT256 and dual-PUT32768: 
{\tt flush-9.0}, {\tt jbd2/md0-8}, {\tt kblock/0}, {\tt md0\_raid1}, 
and {\tt ntpd}. That said, the central cluster also contains other
  processes not seen in the dual-PUT256 central cluster: 
  {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
  {\tt rhsmcertd-worke}, and {\tt sshd}. 
  But these daemons were categorized in the \hbox{dual-PUT256} analysis as {\em
  infrequent}, several having periodicities estimated at four or twenty hours.
When the PUT had a ``short'' 
program time (in this case, two minutes), daemons with a periodicity of
hours are infrequent. But with a PUT with a ``long'' program time (in this
case, 4.5 hours), some of those daemons are now frequent, and appear in the 
central cluster.

In Step~9 we compute 
the cutoff for each of those infrequent, long-running daemons so
identified, based on the PUT128 and the 
PUT16384 as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
Here is how to compute the cutoff. For the cutoff of such a daemon with PUT128, 
we take the midpoint between the maximum of that daemon's PTs in the 
central cluster (or 0, if absent) and the minimum of those in the $L$-samples. 
For the cutoff of such a daemon with PUT16384, we do the same. 
We then compute a ``task time'' as 5\% of the inferred periodicity. 
This 5\% ensures that such infrequent daemons will impact 
only a small percentage of the shorter PUTs, while presumably being 
associated with much larger cutoffs for the very long PUTs. 
We also include daemons that (a)~were identified as infrequent and 
\hbox{long-running} from PUT128 and (b)~were not identified as so in the PUT16384 
\hbox{$L$-samples}, but may have in the \hbox{dual-PUT32768} central cluster. 
We then take the {\em maximum} of the two cutoffs for the final cutoff PT 
(the last column of Tab.~\ref{tab:final_infrequent_cutoff}).

\begin{table}[h]
\centering
{\scriptsize
\begin{tabular}{|p{1.2cm}|c|c|p{1cm}|p{1.45cm}|} \hline
Process  & Cutoff PT & Cutoff PT & Task  & {\bf Final } \\
 Name & on PUT128  & on PUT16K & Time & {\bf Cutoff PT} \\\hline
{\tt bash} & 1 msec & --- & --- & {\bf 1 msec} \\ \hline
{\tt flush-9:0} & 64 msec & --- & $< 1$ hour & {\bf 64 msec} \\
                & ---     & 48 msec & $\geq 1$ hour & {\bf 48 msec} \\ \hline
{\tt grep }     & 1 msec & 12 msec & --- & {\bf 12 msec}\\ \hline
{\tt jbd2/md0-8} & 4 msec & --- & $< 1$ hour & {\bf 4 msec} \\
                & ---     & 11 msec & $\geq 1$ hour & {\bf 11 msec} \\ \hline
{\tt md0\_raid1} & 35 msec & ---     & $< 1$ hour  & {\bf 35 msec}\\
                 & ---     & 51 msec & $\geq 1$ hour & {\bf 51 msec} \\ \hline
{\tt rhn\_check}  & 281 msec & --- & $< 12$ min & {\bf 281 msec} \\
                 &  --- & 12,828 msec & $\geq 12$ min & {\bf 12,828 msec}\\ \hline
{\tt rhnsd} & 2 msec & --- & $< 12$ min & {\bf 2 msec} \\
            & ---    & 12 msec &$\geq 12$ min & {\bf 12 msec} \\ \hline
{\tt rhsmcertd}  & 1 msec & 1 msec & --- & {\bf 1 msec} \\  \hline
{\tt rhsmcertd}  & 57 msec & --- & $< 12$ min & {\bf 57 msec} \\
{\tt -worke}           &  --- & 119 msec  & $\geq 12$ min & {\bf 119 msec}\\ \hline
{\tt sshd} & 2 msec & 23 msec & --- & {\bf 23 msec}\\ \hline
\end{tabular}
}
\caption{Collected Infrequent, Long-running Daemons and Their Final Cutoff
  Process Time \hbox{(Step~9)}\label{tab:final_infrequent_cutoff}}
  \vspace{-0.4in}
\end{table}

Based on Tab.~\ref{tab:final_infrequent_cutoff}, 
we discard any sample containing an infrequent, long-running daemon execution
over the corresponding cutoff. We thus end up dropping just fifteen of the 800
PUT128 samples and only two of the forty PUT16384 samples. 
As a result, the overall measurement quality---the standard deviation and relative error 
---for PUT128 was improved 
by three orders of magnitude (as shown in Fig.~\ref{fig:meas_comp}) 
and by about two orders of magnitude for PUT16384.
%(5661.7 vs. 229.8) 0.0003448491 vs. 8.4$\times$10$^{-6}$
%(40.4 vs. 26.7) 2.463122e-06 vs. 1.626005e-06

%\begin{table}[H]
%\centering
%{\small
%\begin{tabular}{|c|c|c|} \hline
% & \multirow{2}{*}{ET/PT} & Std. Dev. & \multirow{2}{*}{Rel. Err.} \\ 
% & & &  & (msec) &  \\ \hline
%\multirow{2}{*}{PUT128} & ET & 800 & 2.5  & 2.0$\times$10$^{-5}$\\  \cline{2-9}
%  & PT & 2.6  & 2.0$\times$10$^{-5}$\\ \hline  
%\multirow{2}{*}{PUT16384} & ET & 138.6  & 8.4$\times$10$^{-6}$\\  \cline{2-9}
% & PT & 2.7  & 1.6$\times$10$^{-6}$\\ \hline  
%\end{tabular}
%}
%\caption{Measurement Statistics of PUT128 (above) and PUT16384 (below) 
%by the SEDONA Scheme\label{tab:refined_stat}}
%\end{table}

%We additionally remove those samples 
%in which the PT measurements are greater or less 
%two standard deviations from the average of the combined samples. In this
%particular case, no additional samples were removed.

%Finally, we calculate the execution time of PUT128 as the average PT measurement in Table~\ref{tab:refined_empv5}: 
%128250 milliseconds (rounded up to the nearest millisecond, because PT is
%measured in msec). Table~\ref{tab:refined_empv5} shows the statistics of 
%the PT and ET measurements. (In the fourth column, the number of samples
%dropped by the first sanity check (15 for ET) and by the second sanity check
%(1) are indicated.) Our original goal was to
%determine using a carefully-motivated protocol the program time (PT) for
%this program under test (PUT). We see that the standard deviation and
%relative error have both been considerably reduced by this principled
%elimination of samples containing identified infrequent, long-running daemons.

%\begin{table}[H]
%\centering
%{\small
%\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|} \hline
% & \multirow{2}{*}{ET/PT} & \multirow{2}{*}{\# of Samples} & \multirow{2}{*}{\# of Ols. (=I+II)} & Min. & Max. & Avg. & Std. Dev. & \multirow{2}{*}{Rel. Err.} \\ 
% & & &  & (msec) & (msec) & (msec) & (msec) &  \\ \hline
%\multirow{2}{*}{PUT128} & ET & 800 & 16 (=15+1) & 128,245  & 128,256  & 128,250  & 2.5  & 2.0$\times$10$^{-5}$\\  \cline{2-9}
%  & PT & 800 & 20 (=15+5) & 128,245  & 128,255  & 128,250  & 2.6  & 2.0$\times$10$^{-5}$\\ \hline  
%\multirow{2}{*}{PUT16384} & ET & 40 & 6 (=2+4) & 16,416,454  & 16,417,155  & 16,416,637  & 138.6  & 8.4$\times$10$^{-6}$\\  \cline{2-9}
% & PT & 40 & 2 (=2+0) & 16,415,757  & 16,415,855  & 16,415,804  & 2.7  & 1.6$\times$10$^{-6}$\\ \hline  
%\end{tabular}
%}
%\caption{Statistics of PUT128 (above) and PUT16384 (below) by EMPv5\label{tab:refined_empv5}}
%\end{table}

\section{Evaluation}
\label{sec:eval}
\vspace{-0.07in}
We now evaluate the \hbox{performance} of the SEDONA protocol.
Our experiments were conducted on a machine
described in Table~\ref{tab:machine_config}. 
\begin{table}[h]
\vspace{-0.2in}
\begin{center}
{\scriptsize
\begin{tabular}{|l|p{7cm}|}\hline
OS & Red Hat Ent. Linux (RHEL) 6.4 with a kernel of 2.6.32 \\ \hline
CPU & Intel Core i7-870 Lynnfield 2.93GHz quad-core \hbox{processor}\shorten{ on a LGA 1156 95W motherboard}\\ \hline
RAM & 4GB of DDR3 1333 dual-channel memory\\ \hline
HDD & Western Digital Caviar Black 1TB 7200rpm SATA Drive\\ \hline
\end{tabular}
}
\end{center}
\caption{Machine Configurations\label{tab:machine_config}}
\vspace{-0.3in}
\end{table}

We evaluated the performance of the SEDONA using SPEC CPU2006 
benchmarks~\cite{specCpu2006}, providing various compute-bound real applications. 
%Thus, employing the SPEC benchmarks as real-world workloads 
%was a reasonable choice for the evaluation. 
The results are provided in Table~\ref{tab:spec_real}.
Note that in the table there are two empty bins associated with 
the {\tt 481} and {\tt 483} benchmarks. 
In the case of {\tt 481}, it was not possible to obtain its 
measurements of {\tt 481} due to some unknown error 
(perhaps caused by provided binary input data). 
As {\tt 483} concerned I/O, we excluded that benchmark. 
%Other than these two, we had successful runs of a total of 29 (out of 31) benchmarks.

Table~\ref{tab:spec_real} shows that
the SEDONA protocol (which uses PT) significantly outperformed the original 
measurement technique (using ET), termed ORG, 
on the standard deviation and relative error across the different SPEC benchmarks. 
None of the benchmarks revealed a bigger standard deviation from the SEDONA
protocol as compared to that of ORG.
%All of the benchmarks revealed a smaller standard deviation on our scheme than that of the ORG. 
Our protocol quite effectively filtered out infrequent daemon executions 
in these real-world workloads. 
Furthermore, the relative error of SEDONA was equal to or 
lower than that of ORG for slightly under an half (specifically, 12) of the benchmarks.
%For instance, we observed about a 10x margin between the two schemes for the 
%{\tt 434} workload.

SEDONA also scaled well for the SPEC workloads, 
with regard to growth of relative error as the execution time lengthened.
For the short benchmarks 
(e.g., {\tt 400}, {\tt 403}, {\tt 410}, 
{\tt 434}, {\tt 445}, and {\tt 999}: those taking under 100 sec), 
our scheme outperformed the ORG by about 3.5x, on average. 
The SEDONA continued its dominance against the conventional technique 
for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
For the long-running benchmarks (e.g., {\tt 436} and {\tt 454}, both$>$ 900 sec), 
the relative error of the \hbox{SEDONA} was slightly lower than that of the ORG.

%Lastly, we still observed a substantial standard deviation (and relative error) for 
%some benchmarks (e.g., {\tt 436} and {\tt 462}).
%The high variance seemed mainly attributed to the activities
%of the {\tt kslowd000} and {\tt kslowd001} daemon processes, as also identified in the real-world programs. 
%These two were not observed during our cutoff study. 
%The daemons cannot be switched off, as they involve kernel threads. 
%More investigation is needed to hopefully better remove 
%daemon executions interfering with timing a program.

%In summary, these results empirically demonstrate that measurement quality---accuracy,
%precision, and scalability---of EMP is also realized with real CPU-bound benchmarks.

\vspace{-0.1in}
\begin{table}[h]
\centering
{\scriptsize
\begin{tabular}{l|cc|cc} \hline
\multirow{2}{*}{Benchmarks} 
  & \multicolumn{2}{c|}{Std. Dev. (msec)} 
  & \multicolumn{2}{c}{Relative Error}\\ \cline{2-5}
  & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline		
{{\tt 400.perlbench}} & {3} 	& {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
{{\tt 401.bzip2}} & {1,185} & {1,161} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
{{\tt 403.gcc}} & {138} & {96} & {5$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
{{\tt 410.bwaves}} & {46} & {19} & {6$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
{{\tt 416.gamess}} & {965} & {876} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\%% V
{{\tt 429.mcf}} & {623} & {600} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
{{\tt 433.milc}} & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
{{\tt 434.zeusmp}} & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
{{\tt 435.gromacs}} & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
{{\tt 436.cactusADM}} & {3,914} & {3,843} & {3.37$\times$10$^{-3}$} & {3.36$\times$10$^{-3}$}\\
{{\tt 437.leslie3d}} & {1,492} & {1,475}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
{{\tt 444.namd}} & {294} & {281}  & {5$\times$10$^{-4}$} & {5$\times$10$^{-4}$}\\
{{\tt 445.gobmk}} & {91} & {28}  & {1$\times$10$^{-3}$} & {3$\times$10$^{-4}$}\\
{{\tt 447.dealII}} & {150} & {108}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
{{\tt 450.soplex}} & {99} & {91}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
{{\tt 453.povray}} & {623} & {582}  & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
{{\tt 454.calculix}} & {678} & {627}  & {3.9$\times$10$^{-4}$} & {3.7$\times$10$^{-4}$}\\
{{\tt 456.hmmer}} & {85} & {50}  & {2$\times$10$^{-4}$} & {1$\times$10$^{-4}$}\\
{{\tt 458.sjeng}} & {542} & {513} &  {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
{{\tt 459.GemsFDTD}} & {2,143} & {2,132}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
{{\tt 462.libquantum}} & {3,326} & {3,274}  & {6$\times$10$^{-3}$} & {6$\times$10$^{-3}$}\\
{{\tt 464.h264ref}} & {601} & {563}  & {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
{{\tt 465.tonto}} & {883} & {797}  & {1$\times$10$^{-3}$} & {1$\times$10$^{-3}$}\\
{{\tt 470.lbm}} & {143} & {94} & {4$\times$10$^{-4}$} & {3$\times$10$^{-4}$}\\
{{\tt 471.omnetpp}} & {2,114} & {2,072} & {6$\times$10$^{-4}$} & {6$\times$10$^{-4}$}\\
{{\tt 473.astar}} & {359} & {317} &  {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
{{\tt 481.wrf}} & \multicolumn{4}{c}{Run-time Error}\\
{{\tt 482.sphinx3}} & {2,436} & {2,392} &   {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\ %% V
{{\tt 483.xalancbmk}} & \multicolumn{4}{c}{Out-of-Scope}\\
{{\tt 998.specrand}} & {0.6} & {0.6} &  {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
{{\tt 999.specrand}} & {0.8} & {0.6} &  {6$\times$10$^{-3}$} & {5$\times$10$^{-3}$}\\ \hline
\end{tabular}
}
\caption{Performance Evaluation on the SPEC Benchmarks\label{tab:spec_real}}
\vspace{-0.33in}
\end{table}

%Figure~\ref{fig:real_performance} shows the evaluation results 
%of the proposed scheme on the running program-under-test with increasing task length in seconds.
%\begin{figure*}[t]
%	\centering
%	\subfigure[Standard Deviation]{
%		\includegraphics[width=0.45\textwidth]{all_put_std.eps}\label{fig:all_put_std}
%	} 
%	\subfigure[Relative Error]{
%		\includegraphics[width=0.45\textwidth]{all_put_re.eps}\label{fig:all_put_re}
%	}
%	\caption{Comparison of Measurement-Quality over Increasing Task Length~\label{fig:all_put}}
%\end{figure*}

%Figure~\ref{fig:real_performance} shows the evaluation results 
%of the proposed scheme on an industrial benchmark, called SPEC CPU2006~\cite{specCpu2006}.
%\begin{figure*}[t!]
%	\centering
%	\subfigure[Standard Deviations]{
%		%\resizebox{width=.8\textwidth}{!}{\includegraphics[width=140mm,height=]{spec_std.eps}\label{fig:spec_std}}
%		%\includegraphics[width=\textwidth,height=3in]{spec_std.eps}\label{fig:spec_std}
%		\includegraphics[scale=0.5]{spec_std.eps}\label{fig:spec_std}
%	} 
%	\subfigure[Relative Errors]{
%		%\vspace{-1in}
%		%\resizebox{.8\textwidth}{!}{\includegraphics[width=140mm]{spec_re.eps}\label{fig:spec_std}}
%		%\includegraphics[width=\textwidth,height=3in]{spec_re.eps}\label{fig:spec_re}
%		\includegraphics[scale=0.5]{spec_re.eps}\label{fig:spec_re}
%		%\includegraphics[scale=0.5]{spec_re.eps}\label{fig:spec_re}
%	}
%	\caption{Timing Performance over SPEC CPU2006~\label{fig:real_performance}}
%\end{figure*}

\section{Conclusion}
\label{sec:conclusion}
\vspace{-0.07in}
In this paper we \shorten{provided an empirical evidence of 
how significantly measurement quality could be hurt by 
the presence of infrequent, long-running daemon processes. 
We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
employing process time and identifying an infrequent, long-running daemon}. 
%Our evaluation using real benchmarks demonstrated the strong effectiveness of 
%the scheme in improving the measurement performance.
Our scheme is more precise and accurate than the traditional method\shorten{
as shown in the experiments}.

% In the future we plan to integrate this approach into a novel query timing 
% protocol~\cite{Currim} to further improve the protocol by 
% early identifying such a daemon and eliminating samples contaminated by that daemon.

%\section*{Acknowledgments} 
%The author give many thanks to Rick Snodgrass, Rob Meier, and John Kececioglu 
%for their exceptional supports and helpful comments to improve this paper.

{\tiny 
\bibliographystyle{ieicetr} 
\bibliography{paper}
}
\end{document}

### 2017-01-25 sedona-ieice16/submission/misc/paper_yksuh.tex
sedona paper added

--- 2017-01-25 sedona-ieice16/submission/misc/paperRTS.tex
@@
 %\theme{}
-\title{SEDONA: A Novel Protocol for Identifying Infrequent, Long-Running Daemons on a Linux System}
+\title{SEDONA: A Novel Timing Scheme for Identifying an Infrequent, Long-Running Daemons on a Linux System}
 %\titlenote{This work was in part supported by the EDISON program (NRF-2011-0020576).}
@@
 \begin{summary}
-It is challenging to accurately measure the execution-time of a given
-program due to many extraneous factors. To enable better timing results
-restrained, we present a novel protocol that identifies infrequent,
-long-running daemons to enable effective elimination of executions
-exhibiting these daemons.
+In general, it is very challenging to measure the execution-time of a given program accurately and precisely because of many extraneous factors.
+To obtain better timing results with such factors restrained, we present a novel timing scheme via identification of infrequent, long-running daemons and selective elimination of executions with so-identified daemons.
 
@@
 \begin{keywords}
-Infrequent Long-running Daemon, Execution-Time Measurement
+Infrequent, Long-running Daemon, Elimination, Execution-Time, Measurement
 \end{keywords}
@@
 
-Measuring program execution time is a much-used
-technique for performance evaluation in computer science. 
+Measuring program execution-time is one of the most popular 
+techniques for performance evaluation in computer science. 
 Despite the importance of accurate and precise execution-time measurement, 
@@
 Surprisingly, there is considerable variability in the measured time, 
-which indeed suggests the need for a better timing protocol.
+which indeed suggests soliciting a nicer timing scheme.
 
-Consider a simple compute-bound program, termed {\em PUT} (program-under-test), 
+Let's consider a simple compute-bound program, termed {\em PUT} (program-under-test), 
 as shown in Fig.~\ref{alg:put}. 
-This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
-The $tl$ value is translated to the corresponding 
-number of iterations for which that for-loop is performed.
+This program runs a nested for-loop designed to be finished in a specified task length ($tl$) (in sec). 
+In the inner loop an integer variable ($j$) gets incremented by one. 
+%That $tl$ value is translated to the corresponding 
+%number of iterations for which that for-loop is performed. 
+%The addition operation is
 
@@
 	\centering
-	\subfigure[Initial Elapsed Time Measurement]{
+	\subfigure[A Histogram of Original Measurements]{
 		\includegraphics[width=0.32\textwidth]{128_sec_et_old_hist.eps}
@@
 	}
-	\subfigure[Program Time After Identified Daemon Removal]{
+	\subfigure[A Histogram of Our Measurements]{
 		\includegraphics[width=0.32\textwidth]{128_sec_pt_new_hist.eps}
@@
 
-We used 128~seconds as the task length ran the PUT program (designated as PUT128) 
-800~times. Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on the PUT128.
+We passed 128~seconds as task length to and ran the PUT program (called PUT128) 
+800~times. Fig.~\ref{fig:meas_comp} shows two histograms of timing results on the PUT128.
 As can be seen in Fig.\ref{fig:hist_before_et}, 
@@
 In contrast, our scheme, which we propose shortly, 
-significantly reduces such variability.
-The gist of the scheme is to (i)~focus on
-{\em process time} (PT) rather than on elapsed time (ET) and 
-(ii)~remove some executions that involve
-{\em infrequent, \hbox{long-running} daemon processes}, which impact
+successfully handles such variability with useful treatments. 
+The gist of the scheme is to i) employ 
+{\em process time} (PT) rather than elapsed time (ET) and 
+ii) remove some executions including 
+{\em infrequent, \hbox{long-running} daemon processes} impacting 
 the process time of the process. 
-%(Typically, one tick is equivalent to 1/100 secs, or 1 {\tt HZ}.) 
+(Typically, one tick is equivalent to 1/100 secs, or 1 {\tt HZ}.) 
 Using PT is preferred, in that it considers 
 the time taken for only a process of interest.
-PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
+PT can be calculated as the sum of ticks 
 in user and system mode, available via {\tt taskstats} C struct, 
 provided by the Linux NetLink facility~\cite{Netlink}.
-This use of PT and selective elimination 
+The successful use of PT and selective elimination 
 results in improved 
-measurement quality---in both standard deviation and relative error---
+measurement quality---standard deviation and relative error---
 by three orders of magnitude, 
@@
 {\bf Related Work.} 
-McGeoch introduced
+There are several academic works concerning measuring program time. 
+McGeoch introduces
 two basic methods of measuring program time: elapsed time and CPU time~\cite{Mcgeoch12}. 
 Bryant and O'Hallaron~\cite{Randal03} 
-presented two timing schemes of using clock-cycle and interval counters. 
+presents two timing schemes of using clock-cycle and interval counters. 
 They proposed a measurement protocol, called {\em minimum-of-k}, 
@@
 None of these prior works, however, takes into account the variability in 
-timing and the influence of daemons that may significantly disturb the timing.
+timing and the influence of co-existing daemons that may significantly disturb the timing.
 
@@
 
-Commercial software tools measure execution time~\cite{VTune,TimeSys,WindView}. 
+There are some commercial software tools for measuring execution time~\cite{VTune,TimeSys,WindView}. 
 Since the tools' source code is not disclosed, there is no way of figuring out 
@@
 
-We previously developed a timing protocol called TTP (Tucson Timing Protocol)
+We have developed a timing protocol, called TTP (Tucson Timing Protocol), 
 for programs exhibiting I/O, in particular, query execution time in \hbox{DBMSes}~\cite{Currim}.
@@
 
-\item We provide empirical evidence that 
+\item We provide empirical evidences that 
 measuring program time 
-can be seriously affected by extant system daemons.
+can be seriously hurt by extant system daemons 
+without a proper care.
 
-\item We propose a novel timing protocol that
-identifies infrequent, long-running daemons that impact the timing results for that program. 
+\item We propose a novel timing scheme 
+to identify an infrequent, long-running daemons impacting the timing results for that program. 
 %This scheme determines in a disciplined way cutoffs to remove executions with such daemons.
 
-\item We evaluate the performance of the protocol with rigorous experiments, 
+\item We evaluate the performance of the method by rigorous experiments, 
 starting from a simple program in pure-computation mode 
@@
 
-\item The experimental results show a demonstrate for the effectiveness of our scheme\shorten{SEDONA: 
+\item The experimental results show a strong support for the effectiveness of our scheme\shorten{SEDONA: 
 early eliminating dirty executions infected with such a daemon 
@@
 
-\noindent
-The rest of this letter is organized as follows. 
+The rest of this letter is organized in the following. 
 Section~\ref{sec:prop_appach} elaborates on the proposed scheme. 
-In the following section, we evaluate the performance of 
-the scheme using real workloads.
+In turn, we evaluate the performance of 
+the scheme using real workloads. Finally, we 
+conclude this letter.
 
@@
 {\bf Algorithm} The SEDONA Timing Scheme: \\
-\STATE Step 1. Set up the timing environment.
+\STATE Step 1. Set up a timing environment.
 \STATE Step 2. Perform a single PUT run (specifically, PUT128) for many samples (specifically, 800).
-\STATE Step 3. Consider each pair of elapsed time measurements to be a
-dual-PUT measurement 
+\STATE Step 3. Consider each pair of elapsed time measurements to be a dual-PUT measurement, 
 and examine a scatter-plot to see if it it displays an $L$-shape.
@@
 \STATE Step 9. Compute the cutoffs for each identified daemon. 
-\STATE Step 10. Discard an execution including a daemon of which process
-time is greater than the respective cutoff time.
+\STATE Step 10. Discard an execution including a daemon of which process time is greater than 
+the respective cutoff time.
 \end{algorithmic}
@@
 
-%We now elaborate the SEDONA scheme with an example, to explain and justify each step.
+We now elaborate the above SEDONA scheme with an example, to explain and justify each step.
 %\vspace{-0.8in}
@@
 {\bf A Running Example.} 
-As motivated by our prior work~\cite{Currim}, 
+As illustrated in our prior work~\cite{Currim}, 
 we configure a timing environment by i) stopping non-critical daemons, 
 ii) activating the Network Timing Protocol daemon, and 
-iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1). 
+iii) switching off CPU features\cite{intel15,intelSpeed15} if any (Step 1). 
 %(Our machine's specification will be given in Tab~\ref{tab:machine_config}).
@@
 (We render Fig.~\ref{fig:hist_before_et} using the 800 samples from this run.)
-We use 128 seconds because that is long enough to perhaps experience an infrequent daemon. 
-We run it 800 times to capture infrequent daemons that perhaps run every few hours or even 
+Why PUT128? Because it is long enough to perhaps experience an infrequent daemon. 
+Why 800? To capture infrequent daemons perhaps running every few hours or even 
 once a day. 
@@
 the top and middle rows represent over a dozen of outliers far from 
-the rest of the samples clustered in the bottom row. 
+th rest of the samples clustered in the bottom row. 
 We'll now drill down into these outliers to 
@@
 To identify such daemons, we use a novel scatter plot: 
-those of {\em pairs of successive samples}.\shorten{So samples 1 and 2 
+those of pairs of successive samples.\shorten{So samples 1 and 2 
 form the first pair, samples 3 and 4 form the second pair.} 
@@
 from a run of the 800 PUT128 samples (Step~3). 
-There are two quite obvious outliers\shorten{, corresponding to sample \# 75 and \# 634,} with ETs of 163,913 msec (rightmost) and 161,785 msec (uppermost), respectively.
+There are two quite obvious outliers\shorten{, corresponding to sample \# 75 and \# 634,} with ETs of 163,913 msec (rightmost), and 161,785 msec (uppermost), respectively.
 
@@
 Next, we can compute for each so-identified infrequent, long-running daemon its
-minimum time in the \hbox{$L$-samples} (Step 7). 
-This computation provides a rough, initial distinction of a ``long-running''
-daemon, namely, the valley between the maximum PT from the central cluster 
-and the minimum PT from the $L$-samples, to differentiate ``short-running'' 
+minimum time in the $L$-samples (Step 7). 
+This computation provides a rough, initial distinction of a ``long-running'' daemon. 
+Namely, the valley between the maximum PT from the central cluster 
+and the minimum PT from the $L$-samples differentiated ``short-running'' 
 from ``\hbox{long-running}'' executions of the daemon. 
-For those daemons (i.e. {\tt grep}) never appearing in the central cluster,
+For those daemons (i.e. {\tt grep}) never appearing in the central cluster 
 this initial analysis concludes only that they are infrequent.
@@
 In Step 8 we repeat Steps 1--6, but instead with the much-longer running
-PUT16384 (4.5 hours per sample versus 2 minutes), to see 
+PUT16384 instead (4.5 hours per sample versus 2 minutes), to see 
 if any of our identified infrequent daemons are actually frequent 
@@
 we discard any sample containing an infrequent, long-running daemon execution
-over the corresponding cutoff. We thus end up dropping just fifteen of the 800
+over the corresponding cutoff. We thus end up removing the fifteen of the 800
 PUT128 samples and only two of the forty PUT16384 samples. 
@@
 by three orders of magnitude (as shown in Fig.~\ref{fig:meas_comp}) 
-and by about two orders of magnitude for PUT16384.
+and by about two orders of magnitude for PUT16384, too.
 %(5661.7 vs. 229.8) 0.0003448491 vs. 8.4$\times$10$^{-6}$
@@
 
+We now evaluate the \hbox{performance} of the SEDONA.
+
 \section{Evaluation}
@@
 \vspace{-0.07in}
-We now evaluate the \hbox{performance} of the SEDONA protocol.
-Our experiments were conducted on a machine
-described in Table~\ref{tab:machine_config}. 
+Our experiments were conducted on a machine with 
+the specification described in Table~\ref{tab:machine_config}. 
 \begin{table}[h]
@@
 
-Table~\ref{tab:spec_real} shows that
-the SEDONA protocol (which uses PT) significantly outperformed the original 
+As exhibited in Table~\ref{tab:spec_real}, overall 
+the SEDONA (using PT) significantly outperformed the original 
 measurement technique (using ET), termed ORG, 
 on the standard deviation and relative error across the different SPEC benchmarks. 
-None of the benchmarks revealed a bigger standard deviation from the SEDONA
-protocol as compared to that of ORG.
+None of the benchmarks revealed a bigger standard deviation of our scheme compared to that of the ORG.
 %All of the benchmarks revealed a smaller standard deviation on our scheme than that of the ORG. 
-Our protocol quite effectively filtered out infrequent daemon executions 
-in these real-world workloads. 
-Furthermore, the relative error of SEDONA was equal to or 
-lower than that of ORG for slightly under an half (specifically, 12) of the benchmarks.
-%For instance, we observed about a 10x margin between the two schemes for the 
-%{\tt 434} workload.
+Our timing approach filtering out infrequent daemon executions 
+was very effective for these real-world workloads. 
+Furthermore, the relative error of the SEDONA was equal to or 
+lower than that of the ORG for slightly under an half (specifically, 12) of the benchmarks. 
+For instance, we observed about a 10x margin between the two schemes for the 
+{\tt 434} workload.
 
-SEDONA also scaled well for the SPEC workloads, 
+The SEDONA scheme also scaled well for the SPEC workloads, 
 with regard to growth of relative error as the execution time lengthened.
-For the short benchmarks 
+For some lightweight benchmarks 
 (e.g., {\tt 400}, {\tt 403}, {\tt 410}, 
-{\tt 434}, {\tt 445}, and {\tt 999}: those taking under 100 sec), 
+{\tt 434}, {\tt 445}, and {\tt 999})  (taking under 100 sec), 
 our scheme outperformed the ORG by about 3.5x, on average. 
 The SEDONA continued its dominance against the conventional technique 
-for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-For the long-running benchmarks (e.g., {\tt 436} and {\tt 454}, both$>$ 900 sec), 
+for the medium-weight ones (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}) (100 $\sim$ 400 sec). 
+For the heavyweight ones (e.g., {\tt 436} and {\tt 454}) ($>$ 900 sec), 
 the relative error of the \hbox{SEDONA} was slightly lower than that of the ORG.
@@
 the presence of infrequent, long-running daemon processes. 
-We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
+We }presented a novel execution-time measurement scheme, called {\em SEDONA}\shorten{, 
 employing process time and identifying an infrequent, long-running daemon}. 
@@
 %the scheme in improving the measurement performance.
-Our scheme is more precise and accurate than the traditional method\shorten{
+Our scheme was more precise and accurate than the traditional method\shorten{
 as shown in the experiments}.

### 2017-01-25 sedona-ieice16/submission/paper.tex
sedona paper added

--- 2017-01-25 sedona-ieice16/submission/misc/paperRTS.tex
@@
 %\usepackage{pstricks}
+\usepackage{array}
 \usepackage{epstopdf}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 \field{D}
-%\SpecialIssue{}
-%\SpecialSection{}
-%\theme{}
 \title{SEDONA: A Novel Protocol for Identifying Infrequent, Long-Running Daemons on a Linux System}
 %\titlenote{This work was in part supported by the EDISON program (NRF-2011-0020576).}
-%\titlenote{}
 \authorlist{
@@
 }
-\received{2016}{11}{30}
-%\revised{2015}{10}{31}
+\received{2016}{12}{3}
 
@@
 It is challenging to accurately measure the execution-time of a given
-program due to many extraneous factors. To enable better timing results
-restrained, we present a novel protocol that identifies infrequent,
+program due to many extraneous factors. To enable better timing results\shorten{restrained},
+we present a novel protocol that identifies infrequent, 
 long-running daemons to enable effective elimination of executions 
@@
 
-Consider a simple compute-bound program, termed {\em PUT} (program-under-test), 
+Consider a simple compute-bound program, what we term {\em PUT} (program-under-test), 
 as shown in Fig.~\ref{alg:put}. 
@@
 
-\vspace{-.25in}
+\vspace{-.2in}
 \begin{figure}[h]
@@
 {\bf Algorithm} PerformManyIncrements($tl$): \\
-\STATE $t$ = TranslateToNumRepeats($tl$) \\
+\STATE $t$ = $tl$ * {CONSTANT} \\
 \FOR{$k$ = 1 to $t$ by 1}
@@
 \end{center}
-\caption{Task Definition by a Program-Under-Test (PUT)\label{alg:put}}
-\vspace{-.25in}
+\caption{Computation by a Program-Under-Test (PUT)\label{alg:put}}
+\vspace{-.2in}
 \end{figure}
 
-\begin{figure}[t]
-	\centering
-	\subfigure[Initial Elapsed Time Measurement]{
-		\includegraphics[width=0.32\textwidth]{128_sec_et_old_hist.eps}
-		\label{fig:hist_before_et}
-	}
-	\subfigure[Program Time After Identified Daemon Removal]{
-		\includegraphics[width=0.32\textwidth]{128_sec_pt_new_hist.eps}
-		\label{fig:hist_after_pt}
-	}
-	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
-	Program-Under-Test~\label{fig:meas_comp}}
-	\vspace{-.3in}
-\end{figure}
-
-We used 128~seconds as the task length ran the PUT program (designated as PUT128) 
-800~times. Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on the PUT128.
-As can be seen in Fig.\ref{fig:hist_before_et}, 
-most of the 800 executions measured in elapsed time, 
+We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
+800~times. Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on PUT128.
+Most of the 800 executions measured in elapsed time, 
 via  {\tt System.currentTimeMillis()} in Java, are gathered in the tallest bar 
-of \hbox{128,000--129,000 msec,} with just a few executions outside of that bar. 
-But these outliers seriously hurt the overall measurement quality: 
+of \hbox{128,000--129,000 msec,} with just a few executions outside of that bar, as can be seen in Fig.~\ref{fig:hist_before_et}. 
+But these outliers significantly impact the overall measurement quality: 
 a standard deviation of 1,700 msec and a relative error of 0.03. 
@@
 the process time of the process. 
-%(Typically, one tick is equivalent to 1/100 secs, or 1 {\tt HZ}.) 
 Using PT is preferred, in that it considers 
@@
 This use of PT and selective elimination 
-results in improved 
-measurement quality---in both standard deviation and relative error---
-by three orders of magnitude, 
-as illustrated in Fig.~\ref{fig:hist_after_pt}.
+was able to improve 
+measurement quality---in both standard deviation and relative error---up to by 
+three orders of magnitude for PUT128, as illustrated in Fig.~\ref{fig:hist_after_pt}.
 
+%\vspace{0.05in}
 {\bf Related Work.} 
@@
 that for observed elapsed ticks the minimum is chosen as the most accurate one. 
-Odom et al.'s work~\cite{Odom05} focuses on timing long-running programs 
-in a simulation framework via 
-dynamic sampling of trace snippets during program execution. 
+Odom et al.'s work~\cite{Odom05} focused on timing long-running programs 
+in a simulation framework via dynamic sampling of trace snippets during program execution. 
 None of these prior works, however, takes into account the variability in 
@@
 
-%Some prior works~\cite{Burns93,Grund11,Santos10,Wilhelm08} discuss 
-%the worst case execution time (WCET) of a process. 
-%The papers focused on measuring the WCET of a deadline-sensitive process 
-%on real-time embedded systems while being oblivious to 
-%daemons coexisting with the process. In addition, we do not consider real-time constraints.
+\begin{figure}[t]
+	\centering
+	\subfigure[Initial Elapsed Time Measurement]{
+		\includegraphics[width=0.31\textwidth]{128_sec_et_old_hist.eps}
+		\label{fig:hist_before_et}
+	}
+	\subfigure[Program Time After Identified Daemon Removal]{
+		\includegraphics[width=0.31\textwidth]{128_sec_pt_new_hist.eps}
+		\label{fig:hist_after_pt}
+	}
+	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
+	Program-Under-Test~\label{fig:meas_comp}}
+	\vspace{-.18in}
+\end{figure}
 
@@
 
+%\vspace{0.05in}
 {\bf Contribution.} 
 Our contributions are following.
-\vspace{-0.05in}
+\vspace{-0.12in}
 \begin{itemize}
 
-\item We provide empirical evidence that 
+\item We show empirical evidence that 
 measuring program time 
-can be seriously affected by extant system daemons.
+can be seriously affected by daemons.
 
@@
 identifies infrequent, long-running daemons that impact the timing results for that program.
-%This scheme determines in a disciplined way cutoffs to remove executions with such daemons.
 
@@
 
-%\item We examine how the FIND scheme applies to process time as well as elapsed time.
-
-\item The experimental results show a demonstrate for the effectiveness of our scheme\shorten{SEDONA: 
+\item The experimental results show a support for the effectiveness of our scheme\shorten{SEDONA: 
 early eliminating dirty executions infected with such a daemon 
@@
 long-running via a {\em cutoff} measure, and significantly improves measurement quality. 
-The SEDONA scheme consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
+The SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
 %We first enumerate the steps to (a)~identify infrequent, 
@@
 \begin{algorithmic}
-{\bf Algorithm} The SEDONA Timing Scheme: \\
+{\bf Algorithm} The SEDONA Timing Protocol: \\
 \STATE Step 1. Set up the timing environment.
 \STATE Step 2. Perform a single PUT run (specifically, PUT128) for many samples (specifically, 800).
-\STATE Step 3. Consider each pair of elapsed time measurements to be a
-dual-PUT measurement 
+\STATE Step 3. Consider each pair of elapsed time measurements to be a dual-PUT measurement 
 and examine a scatter-plot to see if it it displays an $L$-shape.
@@
 \STATE Step 9. Compute the cutoffs for each identified daemon. 
-\STATE Step 10. Discard an execution including a daemon of which process
-time is greater than the respective cutoff time.
+\STATE Step 10. Discard an execution including a daemon of which process time is greater than the respective cutoff time.
 \end{algorithmic}
 \end{center}
-\caption{Summary of the SEDONA Scheme\label{alg:find}}
-\vspace{-0.28in}
+\caption{Summary of the SEDONA Timing Protocol\label{alg:find}}
+\vspace{-0.25in}
 \end{figure}
@@
 		%\includegraphics[width=0.223\textwidth]{put128_empv4.eps}
-		\includegraphics[width=0.223\textwidth]{put128_empv4_new.eps}
+		\includegraphics[width=0.224\textwidth]{put128_empv4_new.eps}
 		\label{fig:reg_put128}
@@
 		%\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4.eps}
-		\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4_new.eps}
+		\includegraphics[width=0.224\textwidth]{dual_et_put256_empv4_new.eps}
 		\label{fig:raw_put128}
 	}
-	%\vspace{-0.1in}    
+	\vspace{0.1in}    
 	\subfigure[Zooming in on the $L$-shape]{
-		\includegraphics[width=0.223\textwidth]{zooming_in_dual_et_put256_empv4.eps}
+		\includegraphics[width=0.224\textwidth]{zooming_in_dual_et_put256_empv4.eps}
 		\label{fig:dual_put256_liters}
@@
 	\subfigure[The Central Cluster]{
-		\includegraphics[width=0.223\textwidth]{dual_et_clump_put256_empv4.eps}
+		\includegraphics[width=0.224\textwidth]{dual_et_clump_put256_empv4.eps}
 		\label{fig:dual_et_put256_cluster}
@@
     (equivalent to a Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
-    \vspace{-0.3in}     
+    \vspace{-0.16in}     
 \end{figure}
@@
 In Step 8 we repeat Steps 1--6, but instead with the much-longer running
-PUT16384 (4.5 hours per sample versus 2 minutes), to see 
-if any of our identified infrequent daemons are actually frequent 
+PUT16384 (4.5 hours per \hbox{sample} versus 2 minutes), to see 
+if any of our identified \hbox{infrequent} daemons are actually frequent 
 at that much longer PUT execution time. 
@@
   But these daemons were categorized in the \hbox{dual-PUT256} analysis as {\em
-  infrequent}, several having periodicities estimated at four or twenty hours.
+  infrequent}, several having \hbox{periodicities} estimated at four or twenty hours.
 When the PUT had a ``short'' 
 program time (in this case, two minutes), daemons with a periodicity of
-hours are infrequent. But with a PUT with a ``long'' program time (in this
+hours are \hbox{infrequent}. But with a PUT with a ``long'' program time (in this
 case, 4.5 hours), some of those daemons are now frequent, and appear in the 
@@
 Here is how to compute the cutoff. For the cutoff of such a daemon with PUT128, 
-we take the midpoint between the maximum of that daemon's PTs in the 
-central cluster (or 0, if absent) and the minimum of those in the $L$-samples. 
+we take the \hbox{midpoint} between the maximum of that daemon's PTs in the 
+central cluster (or 0, if absent) and the \hbox{minimum} of those in the $L$-samples. 
 For the cutoff of such a daemon with PUT16384, we do the same. 
@@
 \begin{tabular}{|p{1.2cm}|c|c|p{1cm}|p{1.45cm}|} \hline
-Process  & Cutoff PT & Cutoff PT & Task  & {\bf Final } \\
- Name & on PUT128  & on PUT16K & Time & {\bf Cutoff PT} \\\hline
+{\tiny Process}  & {\tiny Cutoff PT} & {\tiny Cutoff PT} & {\tiny Task}  & {\tiny{\bf Final }} \\
+{\tiny Name} & {\tiny on PUT128}  & {\tiny on PUT16K} & {\tiny Time} & {\tiny {\bf Cutoff PT}} \\\hline
+%{\tiny Process}  & {\tiny Cutoff PT} & {\tiny Cutoff PT} & {\tiny Task}  & {\tiny{\bf Final }} \\
+% Name & on PUT128  & on PUT16K & Time & {\bf Cutoff PT} \\\hline
 {\tt bash} & 1 msec & --- & --- & {\bf 1 msec} \\ \hline
@@
 Based on Tab.~\ref{tab:final_infrequent_cutoff}, 
-we discard any sample containing an infrequent, long-running daemon execution
+we discard any \hbox{sample} \hbox{containing} an infrequent, long-running daemon execution
 over the corresponding cutoff. We thus end up dropping just fifteen of the 800
 PUT128 samples and only two of the forty PUT16384 samples. 
-As a result, the overall measurement quality---the standard deviation and relative error 
+As a result, the \hbox{overall} \hbox{measurement} quality---the standard deviation and \hbox{relative} error 
 ---for PUT128 was improved 
 by three orders of magnitude (as shown in Fig.~\ref{fig:meas_comp}) 
-and by about two orders of magnitude for PUT16384.
+and by about two \hbox{orders} of magnitude for PUT16384.
 %(5661.7 vs. 229.8) 0.0003448491 vs. 8.4$\times$10$^{-6}$
@@
 
-%\begin{table}[H]
-%\centering
-%{\small
-%\begin{tabular}{|c|c|c|} \hline
-% & \multirow{2}{*}{ET/PT} & Std. Dev. & \multirow{2}{*}{Rel. Err.} \\ 
-% & & &  & (msec) &  \\ \hline
-%\multirow{2}{*}{PUT128} & ET & 800 & 2.5  & 2.0$\times$10$^{-5}$\\  \cline{2-9}
-%  & PT & 2.6  & 2.0$\times$10$^{-5}$\\ \hline  
-%\multirow{2}{*}{PUT16384} & ET & 138.6  & 8.4$\times$10$^{-6}$\\  \cline{2-9}
-% & PT & 2.7  & 1.6$\times$10$^{-6}$\\ \hline  
-%\end{tabular}
-%}
-%\caption{Measurement Statistics of PUT128 (above) and PUT16384 (below) 
-%by the SEDONA Scheme\label{tab:refined_stat}}
-%\end{table}
-
-%We additionally remove those samples 
-%in which the PT measurements are greater or less 
-%two standard deviations from the average of the combined samples. In this
-%particular case, no additional samples were removed.
-
-%Finally, we calculate the execution time of PUT128 as the average PT measurement in Table~\ref{tab:refined_empv5}: 
-%128250 milliseconds (rounded up to the nearest millisecond, because PT is
-%measured in msec). Table~\ref{tab:refined_empv5} shows the statistics of 
-%the PT and ET measurements. (In the fourth column, the number of samples
-%dropped by the first sanity check (15 for ET) and by the second sanity check
-%(1) are indicated.) Our original goal was to
-%determine using a carefully-motivated protocol the program time (PT) for
-%this program under test (PUT). We see that the standard deviation and
-%relative error have both been considerably reduced by this principled
-%elimination of samples containing identified infrequent, long-running daemons.
-
-%\begin{table}[H]
-%\centering
-%{\small
-%\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|} \hline
-% & \multirow{2}{*}{ET/PT} & \multirow{2}{*}{\# of Samples} & \multirow{2}{*}{\# of Ols. (=I+II)} & Min. & Max. & Avg. & Std. Dev. & \multirow{2}{*}{Rel. Err.} \\ 
-% & & &  & (msec) & (msec) & (msec) & (msec) &  \\ \hline
-%\multirow{2}{*}{PUT128} & ET & 800 & 16 (=15+1) & 128,245  & 128,256  & 128,250  & 2.5  & 2.0$\times$10$^{-5}$\\  \cline{2-9}
-%  & PT & 800 & 20 (=15+5) & 128,245  & 128,255  & 128,250  & 2.6  & 2.0$\times$10$^{-5}$\\ \hline  
-%\multirow{2}{*}{PUT16384} & ET & 40 & 6 (=2+4) & 16,416,454  & 16,417,155  & 16,416,637  & 138.6  & 8.4$\times$10$^{-6}$\\  \cline{2-9}
-% & PT & 40 & 2 (=2+0) & 16,415,757  & 16,415,855  & 16,415,804  & 2.7  & 1.6$\times$10$^{-6}$\\ \hline  
-%\end{tabular}
-%}
-%\caption{Statistics of PUT128 (above) and PUT16384 (below) by EMPv5\label{tab:refined_empv5}}
-%\end{table}
+\vspace\fill
 
@@
 \vspace{-0.07in}
-We now evaluate the \hbox{performance} of the SEDONA protocol.
-Our experiments were conducted on a machine
+We now evaluate the \hbox{performance} of the SEDONA \hbox{protocol}.
+Our experiments were conducted on a \hbox{machine}
 described in Table~\ref{tab:machine_config}. 
@@
 
-We evaluated the performance of the SEDONA using SPEC CPU2006 
-benchmarks~\cite{specCpu2006}, providing various compute-bound real applications. 
+We evaluated the performance of SEDONA \hbox{using} SPEC CPU2006 
+benchmarks~\cite{specCpu2006}, providing \hbox{various} compute-bound real applications. 
 %Thus, employing the SPEC benchmarks as real-world workloads 
 %was a reasonable choice for the evaluation. 
-The results are provided in Table~\ref{tab:spec_real}.
-Note that in the table there are two empty bins associated with 
-the {\tt 481} and {\tt 483} benchmarks. 
-In the case of {\tt 481}, it was not possible to obtain its 
-measurements of {\tt 481} due to some unknown error 
-(perhaps caused by provided binary input data). 
-As {\tt 483} concerned I/O, we excluded that benchmark. 
-%Other than these two, we had successful runs of a total of 29 (out of 31) benchmarks.
+The results are \hbox{provided} in Table~\ref{tab:spec_real}. 
+Note that in the table the results for {\tt 481} and {\tt 483} \hbox{benchmarks} 
+are omitted because of some runtime error and incurred I/O, respectively.
 
@@
 on the standard deviation and relative error across the different SPEC benchmarks. 
-None of the benchmarks revealed a bigger standard deviation from the SEDONA
-protocol as compared to that of ORG.
+None of the benchmarks revealed a bigger standard deviation from the SEDONA protocol as compared to that of ORG. 
 %All of the benchmarks revealed a smaller standard deviation on our scheme than that of the ORG. 
@@
 in these real-world workloads. 
-Furthermore, the relative error of SEDONA was equal to or 
-lower than that of ORG for slightly under an half (specifically, 12) of the benchmarks.
-%For instance, we observed about a 10x margin between the two schemes for the 
-%{\tt 434} workload.
+Furthermore, the relative error of SEDONA was 
+lower than that of ORG for all the benchmarks. 
+For instance, about a 10x \hbox{margin} between the two was yielded for 
+{\tt 434.zeusmp}.
 
@@
 (e.g., {\tt 400}, {\tt 403}, {\tt 410}, 
-{\tt 434}, {\tt 445}, and {\tt 999}: those taking under 100 sec), 
+{\tt 434}, {\tt 445}, and {\tt 999}: those taking \hbox{under} 100 sec), 
 our scheme outperformed the ORG by about 3.5x, on average. 
-The SEDONA continued its dominance against the conventional technique 
+The SEDONA protocol continued its dominance against the conventional technique 
 for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-For the long-running benchmarks (e.g., {\tt 436} and {\tt 454}, both$>$ 900 sec), 
-the relative error of the \hbox{SEDONA} was slightly lower than that of the ORG.
+For the long-running benchmarks (e.g., {\tt 436} and {\tt 454}, both$>$ 1,000 sec), 
+the relative error of \hbox{SEDONA} was slightly lower than that of ORG.
+
+%Lastly, the average standard deviation and relative error of the SEDONA 
+%among all the benchmarks were smaller that those of the ORG, 
+%as indicated by the last two rows. 
+%Lastly, the SEDONA outperformed the ORG by a slight margin \hbox{regarding} 
+%the plain (un-weighted) and weighted averages of 
+%the standard deviation and relative error among all the benchmarks, 
+%as indicated by the last two rows.
+%\shorten{In particular, the SEDONA outperformed the ORG regarding 
+%the weighted average by normalization by a slight margin 
+%in spite of seemingly being equal because of the rounding up to one significant digit.}
 
@@
 %The high variance seemed mainly attributed to the activities
-%of the {\tt kslowd000} and {\tt kslowd001} daemon processes, as also identified in the real-world programs. 
+%of the {\tt kslowd000} and {\tt kslowd001} daemon processes, 
+%as also identified in the real-world programs. 
 %These two were not observed during our cutoff study. 
@@
 
-%In summary, these results empirically demonstrate that measurement quality---accuracy,
-%precision, and scalability---of EMP is also realized with real CPU-bound benchmarks.
+To summarize, our proposed SEDONA protocol can achieve {\em better} accuracy,
+precision, and scalability in measuring the execution time of real \hbox{compute-bound} benchmarks 
+than the conventional timing method.
 
@@
 \centering
-{\scriptsize
-\begin{tabular}{l|cc|cc} \hline
-\multirow{2}{*}{Benchmarks} 
-  & \multicolumn{2}{c|}{Std. Dev. (msec)} 
-  & \multicolumn{2}{c}{Relative Error}\\ \cline{2-5}
-  & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline		
-{{\tt 400.perlbench}} & {3} 	& {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-{{\tt 401.bzip2}} & {1,185} & {1,161} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
-{{\tt 403.gcc}} & {138} & {96} & {5$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
-{{\tt 410.bwaves}} & {46} & {19} & {6$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
-{{\tt 416.gamess}} & {965} & {876} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\%% V
-{{\tt 429.mcf}} & {623} & {600} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-{{\tt 433.milc}} & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-{{\tt 434.zeusmp}} & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
-{{\tt 435.gromacs}} & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
-{{\tt 436.cactusADM}} & {3,914} & {3,843} & {3.37$\times$10$^{-3}$} & {3.36$\times$10$^{-3}$}\\
-{{\tt 437.leslie3d}} & {1,492} & {1,475}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-{{\tt 444.namd}} & {294} & {281}  & {5$\times$10$^{-4}$} & {5$\times$10$^{-4}$}\\
-{{\tt 445.gobmk}} & {91} & {28}  & {1$\times$10$^{-3}$} & {3$\times$10$^{-4}$}\\
-{{\tt 447.dealII}} & {150} & {108}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
-{{\tt 450.soplex}} & {99} & {91}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
-{{\tt 453.povray}} & {623} & {582}  & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
-{{\tt 454.calculix}} & {678} & {627}  & {3.9$\times$10$^{-4}$} & {3.7$\times$10$^{-4}$}\\
-{{\tt 456.hmmer}} & {85} & {50}  & {2$\times$10$^{-4}$} & {1$\times$10$^{-4}$}\\
-{{\tt 458.sjeng}} & {542} & {513} &  {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
-{{\tt 459.GemsFDTD}} & {2,143} & {2,132}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-{{\tt 462.libquantum}} & {3,326} & {3,274}  & {6$\times$10$^{-3}$} & {6$\times$10$^{-3}$}\\
-{{\tt 464.h264ref}} & {601} & {563}  & {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
-{{\tt 465.tonto}} & {883} & {797}  & {1$\times$10$^{-3}$} & {1$\times$10$^{-3}$}\\
-{{\tt 470.lbm}} & {143} & {94} & {4$\times$10$^{-4}$} & {3$\times$10$^{-4}$}\\
-{{\tt 471.omnetpp}} & {2,114} & {2,072} & {6$\times$10$^{-4}$} & {6$\times$10$^{-4}$}\\
-{{\tt 473.astar}} & {359} & {317} &  {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
-{{\tt 481.wrf}} & \multicolumn{4}{c}{Run-time Error}\\
-{{\tt 482.sphinx3}} & {2,436} & {2,392} &   {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\ %% V
-{{\tt 483.xalancbmk}} & \multicolumn{4}{c}{Out-of-Scope}\\
-{{\tt 998.specrand}} & {0.6} & {0.6} &  {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
-{{\tt 999.specrand}} & {0.8} & {0.6} &  {6$\times$10$^{-3}$} & {5$\times$10$^{-3}$}\\ \hline
+{
+\scriptsize
+%\tiny
+%\begin{tabular}{|r|r|r|r|r|r|r|} \hline
+% 		  & \multicolumn{2}{r|}{Execution Time (ms)} & \multicolumn{2}{r}{Standard} & \multicolumn{2}{|r|}{\multirow{2}{*}{Relative Error}}\\ \cline{2-3}
+%SPEC  & {ORG}  & {SEDONA} 	& \multicolumn{2}{r}{Deviation (ms)} 	 					& \multicolumn{2}{|r|}{}\\ \cline{4-7}	
+%		 & {(in ET)} &  {(in PT)} & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline
+
+%\begin{tabular}{|p{0.6cm}|p{1cm}|p{1cm}|p{0.5cm}|p{1cm}|p{0.8cm}|p{1cm}|} \hline
+%		  & \multicolumn{2}{p{2.3cm}|}{Execution Time (ms)} & \multicolumn{2}{p{1.6cm}}{Standard}  & \multicolumn{2}{|l|}{\multirow{2}{*}{Relative Error}}\\ \cline{2-3}
+%SPEC  & {ORG}  & {SEDONA} & \multicolumn{2}{p{1.6cm}}{Deviation(ms)} & \multicolumn{2}{|c|}{}\\ \cline{4-7}	
+%		 & {(in ET)} &  {(in PT)} & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline
+
+\begin{tabular}{|p{0.18cm}|R{1.14cm}|R{1.14cm}|R{0.64cm}|R{1cm}|R{0.97cm}|R{0.97cm}|} \hline
+ 		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative Error}}}\\ \cline{2-3}
+ & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SEDONA}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{}\\ \cline{4-7}	
+ & \multicolumn{1}{c|}{{\tiny{(in ET)}}} & \multicolumn{1}{c|}{{\tiny{(in PT)}}} & {\tiny{ORG}} & {\tiny{SEDONA}} & \multicolumn{1}{c|}{{\tiny{ORG}}}& \multicolumn{1}{c|}{{\tiny{SEDONA}}}\\ \hline
+{{\tt 400}} & 454 & 445 & {3} & {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+{{\tt 401}} & 536,639 & 528,517 & {1,185} & {1,161} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
+{{\tt 403}} & 26,109	& 25,695 & {138} & {96} & {5$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
+{{\tt 410}} & 7,938 & 7,801 & {46} & {19} & {6$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
+{{\tt 416}} & 1,015,342 & 999,846 & {965} & {876} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\%% V
+{{\tt 429}} & 235,811 & 232,209 & {623} & {600} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+{{\tt 433}} & 480,586 & 473,256  & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
+{{\tt 434}} & 16,495  & 16,242  & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
+{{\tt 435}} & 990,575 & 975,445  & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
+{{\tt 436}} & 1,160,742 & 1,143,078   & {3,914} & {3,843} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+{{\tt 437}} & 581,635 & 572,775 & {1,492} & {1,475}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+{{\tt 444}} & 591,201 & 582,229 & {294} & {281}  & {5$\times$10$^{-4}$} & {5$\times$10$^{-4}$}\\
+{{\tt 445}} & 84,435  & 83,164  & {91} & {28}  & {1$\times$10$^{-3}$} & {3$\times$10$^{-4}$}\\
+{{\tt 447}} & 521,846 & 513,493  & {150} & {108}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
+{{\tt 450}} & 341,030 & 335,848  & {99} & {91}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
+{{\tt 453}} & 258,797 & 254,496  & {623} & {582}  & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
+{{\tt 454}} & 1,721,804 & 1,695,613  & {678} & {627}  & {4$\times$10$^{-4}$} & {4$\times$10$^{-4}$}\\
+{{\tt 456}} & 410,533 & 404,328  & {85} & {50}  & {2$\times$10$^{-4}$} & {1$\times$10$^{-4}$}\\
+{{\tt 458}} & 589,541 & 580,591  & {542} & {513} &  {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
+{{\tt 459}} & 798,917 & 786,726  & {2,143} & {2,132}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+{{\tt 462}} & 595,188 & 586,120  & {3,326} & {3,274}  & {6$\times$10$^{-3}$} & {6$\times$10$^{-3}$}\\
+{{\tt 464}} & 649,838 & 639,939  & {601} & {563}  & {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
+{{\tt 465}} & 895,754 & 882,106  & {883} & {797}  & {1$\times$10$^{-3}$} & {1$\times$10$^{-3}$}\\
+{{\tt 470}}	& 349,830 & 344,510 & {143} & {94} & {4$\times$10$^{-4}$} & {3$\times$10$^{-4}$}\\
+{{\tt 471}} & 367,589 & 361,959  & {2,114} & {2,072} & {6$\times$10$^{-4}$} & {6$\times$10$^{-4}$}\\
+{{\tt 473}} & 362,587 & 357,090  & {359} & {317} &  {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
+%{{\tt 481}} & \multicolumn{4}{c}{Runtime Error}\\
+{{\tt 482}} & 654,208	 & 644,223  & {2,436} & {2,392} &   {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\ %% V
+%{{\tt 483}} & \multicolumn{4}{c}{Out-of-Scope}\\
+{{\tt 998}} & 128	 & 127  & {0.6} & {0.6} &  {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
+{{\tt 999}} & 128 &  127 & {0.8} & {0.6} &  {6$\times$10$^{-3}$} & {5$\times$10$^{-3}$}\\ \hline 
+\multicolumn{3}{|r|}{Averages} & 852 & 815 &  {3$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ \hline
 \end{tabular}
@@
 \end{table}
+\vspace{0.5in}
 
-%Figure~\ref{fig:real_performance} shows the evaluation results 
-%of the proposed scheme on the running program-under-test with increasing task length in seconds.
-%\begin{figure*}[t]
-%	\centering
-%	\subfigure[Standard Deviation]{
-%		\includegraphics[width=0.45\textwidth]{all_put_std.eps}\label{fig:all_put_std}
-%	} 
-%	\subfigure[Relative Error]{
-%		\includegraphics[width=0.45\textwidth]{all_put_re.eps}\label{fig:all_put_re}
-%	}
-%	\caption{Comparison of Measurement-Quality over Increasing Task Length~\label{fig:all_put}}
-%\end{figure*}
+%\vspace{-0.1in}
+%\begin{table}[h]
+%\centering
+%{
+%\scriptsize
+%%\tiny
+%\begin{tabular}{|p{1.6cm}|p{1.25cm}|p{0.8cm}|p{1cm}|p{0.8cm}|p{1cm}|} \hline
+%  \multirow{2}{*}{Benchmarks} 
+%  & Weight By
+%  & \multicolumn{2}{|p{1.8cm}|}{Std. Dev. (msec)} 
+%  & \multicolumn{2}{|p{1.7cm}|}{Relative Error}\\ \cline{3-6}
+% & Exec. Time & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline		
+%{{\tt 400.perlbench}} & {3$\times$10$^{-5}$} & {3} & {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+%{{\tt 401.bzip2}} & {4$\times$10$^{-2}$} & {1,185} & {1,161} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
+%{{\tt 403.gcc}} & 2$\times$10$^{-3}$ & {138} & {96} & {5$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
+%{{\tt 410.bwaves}} & 6$\times$10$^{-4}$ & {46} & {19} & {6$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
+%{{\tt 416.gamess}} & 7$\times$10$^{-2}$ & {965} & {876} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\%% V
+%{{\tt 429.mcf}} & 2$\times$10$^{-2}$ & {623} & {600} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+%{{\tt 433.milc}} & 3$\times$10$^{-2}$ & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
+%{{\tt 434.zeusmp}} & 1$\times$10$^{-3}$ & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
+%{{\tt 435.gromacs}} & 7$\times$10$^{-2}$ & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
+%{{\tt 436.cactusADM}} & 8$\times$10$^{-2}$  & {3,914} & {3,843} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+%{{\tt 437.leslie3d}} & 4$\times$10$^{-2}$ & {1,492} & {1,475}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+%{{\tt 444.namd}} & 4$\times$10$^{-2}$ & {294} & {281}  & {5$\times$10$^{-4}$} & {5$\times$10$^{-4}$}\\
+%{{\tt 445.gobmk}} & 5$\times$10$^{-3}$ & {91} & {28}  & {1$\times$10$^{-3}$} & {3$\times$10$^{-4}$}\\
+%{{\tt 447.dealII}} & 3$\times$10$^{-2}$ & {150} & {108}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
+%{{\tt 450.soplex}} &2$\times$10$^{-2}$ & {99} & {91}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
+%{{\tt 453.povray}} & 2$\times$10$^{-2}$ & {623} & {582}  & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
+%%{{\tt 454.calculix}} & {678} & {627}  & {3.9$\times$10$^{-4}$} & {3.7$\times$10$^{-4}$}\\
+%{{\tt 454.calculix}} & 1$\times$10$^{-1}$ & {678} & {627}  & {4$\times$10$^{-4}$} & {4$\times$10$^{-4}$}\\
+%{{\tt 456.hmmer}} & 3$\times$10$^{-2}$ & {85} & {50}  & {2$\times$10$^{-4}$} & {1$\times$10$^{-4}$}\\
+%{{\tt 458.sjeng}} & 4$\times$10$^{-2}$ & {542} & {513} &  {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
+%{{\tt 459.GemsFDTD}} & 6$\times$10$^{-2}$ & {2,143} & {2,132}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
+%{{\tt 462.libquantum}} & 4$\times$10$^{-2}$ & {3,326} & {3,274}  & {6$\times$10$^{-3}$} & {6$\times$10$^{-3}$}\\
+%{{\tt 464.h264ref}} &  5$\times$10$^{-2}$ & {601} & {563}  & {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
+%{{\tt 465.tonto}} & 6$\times$10$^{-2}$ & {883} & {797}  & {1$\times$10$^{-3}$} & {1$\times$10$^{-3}$}\\
+%{{\tt 470.lbm}} & 2$\times$10$^{-2}$ & {143} & {94} & {4$\times$10$^{-4}$} & {3$\times$10$^{-4}$}\\
+%{{\tt 471.omnetpp}} & 3$\times$10$^{-2}$ & {2,114} & {2,072} & {6$\times$10$^{-4}$} & {6$\times$10$^{-4}$}\\
+%{{\tt 473.astar}} & 3$\times$10$^{-2}$ & {359} & {317} &  {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
+%%{{\tt 481.wrf}} & \multicolumn{4}{c}{Run-time Error}\\
+%{{\tt 482.sphinx3}} & 5$\times$10$^{-2}$ & {2,436} & {2,392} &   {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\ %% V
+%%{{\tt 483.xalancbmk}} & \multicolumn{4}{c}{Out-of-Scope}\\
+%{{\tt 998.specrand}} & 9$\times$10$^{-6}$ & {0.6} & {0.6} &  {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
+%{{\tt 999.specrand}} & 9$\times$10$^{-6}$  & {0.8} & {0.6} &  {6$\times$10$^{-3}$} & {5$\times$10$^{-3}$}\\ \hline 
+%Total.   & \multicolumn{1}{c|}{1} & \multicolumn{4}{c|}{---}\\ \hline
+%Avg.   & --- & 852 & 815 &  {3$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ \hline
+%Wght. Avg. & --- & 43  & 42  &  {6$\times$10$^{-5}$} & {6$\times$10$^{-5}$}\\ \hline
+%\end{tabular}
+%}
+%\caption{Performance Evaluation on the SPEC Benchmarks\label{tab:spec_real}}
+%\vspace{-0.33in}
+%%\vspace{-0.33in}
+%\end{table}
+%\vspace{0.5in}
 
-%Figure~\ref{fig:real_performance} shows the evaluation results 
-%of the proposed scheme on an industrial benchmark, called SPEC CPU2006~\cite{specCpu2006}.
-%\begin{figure*}[t!]
-%	\centering
-%	\subfigure[Standard Deviations]{
-%		%\resizebox{width=.8\textwidth}{!}{\includegraphics[width=140mm,height=]{spec_std.eps}\label{fig:spec_std}}
-%		%\includegraphics[width=\textwidth,height=3in]{spec_std.eps}\label{fig:spec_std}
-%		\includegraphics[scale=0.5]{spec_std.eps}\label{fig:spec_std}
-%	} 
-%	\subfigure[Relative Errors]{
-%		%\vspace{-1in}
-%		%\resizebox{.8\textwidth}{!}{\includegraphics[width=140mm]{spec_re.eps}\label{fig:spec_std}}
-%		%\includegraphics[width=\textwidth,height=3in]{spec_re.eps}\label{fig:spec_re}
-%		\includegraphics[scale=0.5]{spec_re.eps}\label{fig:spec_re}
-%		%\includegraphics[scale=0.5]{spec_re.eps}\label{fig:spec_re}
-%	}
-%	\caption{Timing Performance over SPEC CPU2006~\label{fig:real_performance}}
-%\end{figure*}
-
-\section{Conclusion}
-\label{sec:conclusion}
-\vspace{-0.07in}
-In this paper we \shorten{provided an empirical evidence of 
-how significantly measurement quality could be hurt by 
-the presence of infrequent, long-running daemon processes. 
-We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
-employing process time and identifying an infrequent, long-running daemon}. 
-%Our evaluation using real benchmarks demonstrated the strong effectiveness of 
-%the scheme in improving the measurement performance.
-Our scheme is more precise and accurate than the traditional method\shorten{
-as shown in the experiments}.
-
+%\section{Conclusion and Future Work}
+%\label{sec:conclusion}
+%%\vspace{-0.07in}
+%In this paper we \shorten{provided an empirical evidence of 
+%how significantly measurement quality could be hurt by 
+%the presence of infrequent, long-running daemon processes. 
+%We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
+%employing process time and identifying an infrequent, long-running daemon}. 
+%%Our evaluation using real benchmarks demonstrated the strong effectiveness of 
+%%the scheme in improving the measurement performance.
+%Our scheme is more precise and accurate than the traditional method\shorten{
+%as shown in the experiments}.
 % In the future we plan to integrate this approach into a novel query timing 
@@
 
+\section{Conclusion and Future Work}
+\label{sec:conclusion}
+\vspace{-0.07in}
+We \shorten{provided an empirical evidence of 
+how significantly measurement quality could be hurt by 
+the presence of infrequent, long-running daemon processes. 
+We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
+employing process time and identifying an infrequent, long-running daemon}, 
+which is more precise and \hbox{accurate} than the traditional method\shorten{
+as shown in the experiments}.
+Our plan is to integrate SEDONA into the query timing protocol~\cite{Currim}.
+%Our evaluation using real benchmarks demonstrated the strong effectiveness of 
+%the scheme in improving the measurement performance.
+%Our scheme is more precise and accurate than the traditional method\shorten{
+%as shown in the experiments}.
+% In the future we plan to integrate this approach into a novel query timing 
+% protocol~\cite{Currim} to further improve the protocol by 
+% early identifying such a daemon and eliminating samples contaminated by that daemon.
+
+%\vspace\fill
+
 %\section*{Acknowledgments} 
 %The author give many thanks to Rick Snodgrass, Rob Meier, and John Kececioglu 
-%for their exceptional supports and helpful comments to improve this paper.
+%for their sincere supports.
 
+\vspace{-0.03in}
 {\tiny 
@@
 \bibliography{paper}
+%\vspace{-0.2in}
 }
+
+\vspace\fill
 \end{document}

### 2017-01-25 sedona-ieice16/submission/data/fig_script.r

@@ -1,0 +1,334 @@
+x = read.csv(file="all_spec_before.dat",head=TRUE,sep="\t")
+y = read.csv(file="all_spec_after.dat",head=TRUE,sep="\t")
+specCpu.std <- cbind(ORG_ET_std=(x$STD_ET),FIND_ET_std=(y$STD_PT))
+specCpu.std <- round(specCpu.std,1)
+setEPS()
+postscript("spec_std.eps")
+b<-barplot(t(specCpu.std[1:31,]), log="y", yaxt="n", main="", beside=TRUE, 
+names.arg=c("400","401","403","410","416","429","433","434","435","436","437",
+	    "444","445","447","450","453","454","456","458","459","462","464",
+	    "465","470","471","473","481","482","483","998","999"), 
+	las=2,xlab="SPEC CPU2006",ylab="Standard deviation in log scale (msec)", ylim=c(0.1,8192), col=c("tan2","blue"))
+#legend(x=85.1,y=8000, c("ORG: ET", "FIND: PT"),cex=.8, col=c("tan2","blue"),pch=c(22,0,0))
+legend(x=9,y=8000, c("ORG-ET", "FIND-PT"),cex=.8, col=c("tan2","blue"),pch=c(22,0,0))
+ticks <- seq(-1, 16, by=1)
+axis(2, at=c(-1,1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024, 2048, 4096, 8192,16384))
+dev.off()
+
+x = read.csv(file="all_spec_before.dat",head=TRUE,sep="\t")
+y = read.csv(file="all_spec_after.dat",head=TRUE,sep="\t")
+specCpu.std <- cbind(ORG_ET_re=(x$RE_ET),FIND_ET_re=(y$RE_PT))
+#specCpu.std <- round(specCpu.std,1)
+setEPS()
+postscript("spec_re.eps")
+b <- barplot(t(specCpu.std[1:31,]), log="y", yaxt="n", main="", beside=TRUE, 
+names.arg=c("400","401","403","410","416","429","433","434","435","436","437","444", "445","447","450","453","454","456","458","459","462","464","465","470","471","473","481","482","483","998","999"), 
[…]

### 2017-01-25 sedona-ieice16/submission/data/spec_eval.r

@@ -1,0 +1,674 @@
+x = read.csv(file="spec_eval.dat",head=TRUE,sep="\t")
+y <- subset(x, x$RE_et > x$RE_pt)
+nrow(y)
+y <- subset(x, x$RE_et == x$RE_pt)
+nrow(y)
+y <- subset(x, x$RE_et < x$RE_pt)
+nrow(y)
+specCpu.std <- cbind(ET_std=(x$RE_et),PT_std=(x$RE_pt))
+setEPS()
+postscript("spec_re.eps")
+b <- barplot(t(specCpu.std[1:31,]), log="y", yaxt="n", main="", beside=TRUE, 
+names.arg=c("400","401","403","410","416","429","433","434","435","436","437","444", "445","447","450","453","454","456","458","459","462","464","465","470","471","473","481","482","483","998","999"), 
+las=2,xlab="SPEC CPU2006",ylab="Relative error in log scale (msec)", ylim=c(0.000001,0.1), col=c("tan2","blue"))
+legend(x=1,y=0.05, c("ET", "PT"),cex=.8, col=c("tan2","blue"),pch=c(22,0,0))
+ticks <- seq(-6, -1, by=1)
+labels <- sapply(ticks, function(i) as.expression(bquote(10^ .(i))))
+axis(2, at=c(0.000001,0.00001,0.0001,0.001,0.01,0.1), labels=labels)
+dev.off()
+
+x = read.csv(file="spec_eval.dat",head=TRUE,sep="\t")
+specCpu.std <- cbind(ET_std=(x$SD_et),PT_std=(x$SD_pt))
+setEPS()
+postscript("spec_std.eps")
+b <- barplot(t(specCpu.std[1:31,]), log="y", yaxt="n", main="", beside=TRUE, 
+names.arg=c("400","401","403","410","416","429","433","434","435","436","437","444", "445","447","450","453","454","456","458","459","462","464","465","470","471","473","481","482","483","998","999"), 
+las=2,xlab="SPEC CPU2006",ylab="Standard deviation in log scale (msec)", ylim=c(0.1,100000), col=c("tan2","blue"))
[…]

### 2017-01-25 pt_characterization/new_data/empv5_et_hist.r
empv5 et data

@@ -1,12 +1,16 @@
 x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
 setEPS()
-postscript("1_sec_et_hist.eps")
-xmax <- max(x$METIME)
-xmin <- min(x$METIME)
-h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+postscript("1_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-binsize=1
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
 plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='ET (ms)', ylab=expression('Frequency'))
@@ -16,13 +20,17 @@ dev.off()
 
 x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
 setEPS()
-postscript("2_sec_et_hist.eps")
-xmax <- max(x$METIME)
-xmin <- min(x$METIME)
-h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+postscript("2_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-binsize=1
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
[…]

### 2017-01-25 pt_characterization/new_data/empv5_pt_hist.r

@@ -2,11 +2,14 @@ x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
 setEPS()
 postscript("1_sec_pt_hist_v5.eps")
 binsize=1
-xmax <- max(x$PRTIME)
-xmin <- min(x$PRTIME)
-nbins <- ceiling((xmax-xmin) / binsize)
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
 h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
-plot(h, ylim=c(0,500), xaxt='n',freq=TRUE,xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, ylim=c(0,1000), xaxt='n',freq=TRUE,xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), xlab='PT (ms)', ylab=expression('Frequency'))
 xaxtl <- seq(xmin-1,xmax+1,by=binsize)
 axis(side=1, at=xaxtl, labels=xaxtl)
@@ -152,11 +155,6 @@ xaxtl <- seq(xmin,xmax,by=5)
 axis(side=1, at=xaxtl, labels=xaxtl)
 dev.off()
 
-1024	     27
-	1024	     88
-	1024	    252
-	1024	    276
-
 x = read.csv(file="1024_sec.dat",head=TRUE,sep="\t")
 setEPS()
 postscript("1024_sec_pt_hist_v5.eps")
@@ -181,7 +179,7 @@ dev.off()
 
 x = read.csv(file="2048_sec.dat",head=TRUE,sep="\t")
 setEPS()
-postscript("2048_sec_pt_hist_V5.eps")
+postscript("2048_sec_pt_hist_v5.eps")
 x <- subset(x, x$ITERNUM <= 300)
 x <- subset(x, x$ITERNUM != 117)
 x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)

### 2017-02-01 sedona-ieice16/revision/reviews/review_response.tex
SEDONA revision

\documentclass[10pt,letterpaper]{article}

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{algorithmic}
\usepackage{alltt}
\usepackage{array}
\usepackage[cmex10]{amsmath}
\usepackage{amssymb}
%\usepackage[caption=false]{caption}
\usepackage[dvipsnames]{color}
%\usepackage{colortbl}
%\usepackage{enumitem}
%\usepackage{eqparbox}
%\usepackage{fixltx2e}
%\usepackage{float}
%\usepackage{floatflt}
%\usepackage{mdwmath}
%\usepackage{mdwtab}
%\usepackage{multirow}
\usepackage{stfloats}
%\usepackage[font=footnotesize]{subfig}
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
%\usepackage{subfigure} 
\usepackage{times}
\usepackage{url}
\usepackage{verbatim} 
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{color}

\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\long\def\todo#1{\hilite{{\bf TODO:} {\em #1}}}

\setlength{\oddsidemargin}{0in}
\setlength{\textwidth}{6.5in}

% new commands: from file -- to permit re-use
%\input{newcommands.tex}

%Hyphenation problems
\hyphenation{name-space}
\hyphenation{schema-location}
\hyphenation{xml-lint}
\hyphenation{time-stamp}
\hyphenation{time-stamps}
\hyphenation{time-stamped}

\urlstyle{rm}
\graphicspath{{./figures/}}

\definecolor{gray}{rgb}{0.5,0.5,0.5}
%\newcommand{\rev}[1]{\textcolor{blue}{#1}}
%\newcommand{\quo}[1]{\textcolor{gray}{#1}}
%\newcommand{\rev}[1]{\vspace{3mm} \noindent {\em {\color{blue}{#1}}}}

\newenvironment{myindentpar}[1]%
{\begin{list}{}
         {\vspace{10pt}
					\setlength{\leftmargin}{#1}}
          \item[]
}
{\end{list}}
\newcommand{\rev}[1]{\begin{myindentpar}{.25in} {\em {\color{blue}{#1}}}\end{myindentpar}}

\newenvironment{myindentparQ}[1]%
{\begin{list}{}
         {\setlength{\leftmargin}{#1}}
          \item[]
}
{\end{list}}
\newcommand{\quo}[1]{\begin{myindentparQ}{.25in} {{\color{gray}{#1}}}\end{myindentparQ}}

\begin{document}

\title{Author's Reply}
\author{}
\maketitle

\section*{Overview}\label{sec:overview}
I appreciate these quite thorough reviews. I appreciate the feedback and suggestions.

\newcounter{RP}

\clearpage
\section*{Reviewer}\label{sec:rev1}

% 1
\rev{
$<<$ Reviewer's comments to the author(s) $>>$

The SEDONA is based on the probability that any daemons run during the
measurement. And cutoff time is based on the actual measurement time.
The proposed approach is interesting, but I cannot understand why the
author employs such statistics and heuristic way, and how efficient the
proposed approach will work. Please refer to my comments, and consider
to revise the manuscript and re-submit .
}

I understand your concern. I respond to individually below. 
Please also note that your (excellent) comments are incorporated into the revision.

\rev{
Comments: 
(1) In the last of the manuscript, the author describes ``Our plan is to
integrate SEDONA into the query timing protocol [8]." If the research
goal is the combination of the SEDONA and the query timing protocol,
the proposed approach may be suitable. However, the goal is not
clearly described in the manuscript. The readers will recognize the
SEDONA is a generic protocol. If the research goal is the combination
of this work and the prior work, the author should mention so.
}

I understand that the goal of this paper is not well articulated. 
Yes, SEDONA is a generic protocol, which can be applied to any arbitrary Linux system 
when it comes to reducing variability of the execution time of a program of interest. 
I now state the actual goal, marked in blue, in the introduction section.

Note also that the mention of my plan in the future work 
is just one aspect of the extension of the SEDONA protocol. 
I claim that the combination is not the ultimate research goal of this paper. 
To avoid confusion, I removed that sentence.

\rev{
(2) If the proposed approach is for a generic purpose, there is a strong
doubt about usability. The improvement of the SEDONA is little. And the
extent of the each benchmark’s improvement described in Table 3
varies widely. Thus, the experimental results does not confirm whether
the SEDONA is widely practical or is useful. The description to clarify
the SEDONA’s characteristics is needed.
}

\rev{ 
(3) The improvement of the SEDONA is little. And the extent of the each
benchmark’s improvement described in Table 3 varies widely. Thus, the
experimental results does not confirm whether the SEDONA is widely
practical or is useful.
}

This review is the same as (2). Please refer to our response of (2) for this comment.

\rev{ 
(4) Since the explanation of ``dual-PUT'' is not enough, I can not well
understand the algorithm shown in Fig. 3. Similarly, what is “1st
Half’s ET (ms)” attached at Fig. 4(b), 4(c), 4(d) ? I understood only that
the label is related with “dual-PUT”. These words are important in the
manuscript. So, please keep in mind to write understandably.
}

\end{document}

### 2017-02-02 sedona-ieice16/revision/paper.tex
SEDONA revision

--- 2017-01-25 sedona-ieice16/submission/paper.tex
@@
 }
-\received{2016}{12}{3}
+\received{2017}{2}{28}
 
@@
 \begin{summary}
-It is challenging to accurately measure the execution-time of a given
-program due to many extraneous factors. To enable better timing results\shorten{restrained},
-we present a novel protocol that identifies infrequent, 
-long-running daemons to enable effective elimination of executions 
-exhibiting these daemons.
+%It is challenging to accurately measure the execution-time of a given
+%program due to many extraneous factors. To enable better timing results\shorten{restrained},
+%we present a novel protocol that identifies infrequent, 
+%long-running daemons to enable effective elimination of executions 
+%exhibiting these daemons.
+{\color{blue}Measuring program execution time 
+a \hbox{much-used} technique for performance evaluation in computer science. 
+Without a proper care, however, timed results may vary a lot, 
+thus making it hard to trust their validity. 
+In this paper we propose a novel timing protocol to significantly 
+reduce such variability.}
 
@@
 how to achieve {\em better} timing has not been well addressed. 
-Surprisingly, there is considerable variability in the measured time, 
-which indeed suggests the need for a better timing protocol.
+{\color{blue} Surprisingly, there is considerable variability in the measured time.
+%which indeed suggests the need for a better timing protocol.
+The goal of this paper is to propose
+a better timing protocol that can significantly reduce such variability 
+and thus enable better timing results without a doubt of their validity.}
 
@@
 This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
-The $tl$ value is translated to the corresponding 
-number of iterations for which that for-loop is performed.
+%The $tl$ value is translated to the corresponding 
+%number of {\color{blue}repetitions} for which that for-loop is performed. 
+{\color{blue}The $tl$ value is used to compute 
+the number of repetitions ($t$) for which that for-loop is performed 
+to reach the specified task length.}
 
@@
 
+%\begin{figure}[t]
+%	\centering
+%	\subfigure[Initial Elapsed Time Measurement]{
+%		%\includegraphics[width=0.31\textwidth]{128_sec_et_old_hist.eps}
+%		\includegraphics[width=0.31\textwidth]{128_sec_et_old.eps}
+%		\label{fig:hist_before_et}
+%	}
+%	\subfigure[Program Time After Our Protocol]{
+%		%\includegraphics[width=0.31\textwidth]{128_sec_pt_new_hist.eps}
+%		\includegraphics[width=0.31\textwidth]{128_sec_pt_new.eps}
+%		\label{fig:hist_after_pt}
+%	}
+%	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
+%	Program-Under-Test~\label{fig:meas_comp}}
+%	\vspace{-.18in}
+%\end{figure}
+
 \begin{figure}[t]
@@
 	\subfigure[Initial Elapsed Time Measurement]{
-		\includegraphics[width=0.31\textwidth]{128_sec_et_old_hist.eps}
+		%\includegraphics[width=0.31\textwidth]{128_sec_et_old_hist.eps}
+		\includegraphics[width=0.224\textwidth]{128_sec_et_old.eps}
 		\label{fig:hist_before_et}
 	}
-	\subfigure[Program Time After Identified Daemon Removal]{
-		\includegraphics[width=0.31\textwidth]{128_sec_pt_new_hist.eps}
+	\subfigure[Program Time After Our Protocol]{
+		%\includegraphics[width=0.31\textwidth]{128_sec_pt_new_hist.eps}
+		\includegraphics[width=0.224\textwidth]{128_sec_pt_new.eps}
 		\label{fig:hist_after_pt}

### 2017-02-02 sedona-ieice16/revision/data/fig_script.r

@@ -42,12 +42,17 @@ mean(x$METIME)
 y <- subset(x, x$METIME < 160000) 
 max(y$METIME)
 #x <- subset(x, x$ITERNUM != 90 & x$ITERNUM != 202) 
-sd(x$METIME)/mean(x$METIME)
+re <- sd(x$METIME)/mean(x$METIME)
 setEPS()
 postscript("128_sec_et_old.eps")
 plot(x$METIME, xlim=c(0,max(x$ITERNUM)), ylim=c(old_xmin, old_xmax), 
 	pch=1, main='', 
-	xlab='Iteration', ylab='Elapsed Time (msec)')
+	xlab='Repetition', ylab='Elapsed Time (msec)')
+options(scipen = 999)
+legend(380, 171500, c(as.expression(paste("# of total samples = ",nrow(x))), 
+		    as.expression(paste("standard deviation = ",round(sd(x$METIME),1),"msec")), 
+		    as.expression(paste("relative error = ",round(re,5)))
+))
 dev.off()
 
 x = read.csv(file="128_sec.dat",head=TRUE,sep="\t")
@@ -111,17 +116,30 @@ x$ITERNUM != 634 & x$ITERNUM != 663 & x$ITERNUM != 746 & x$ITERNUM != 775
 #     12661 grep 		  451	       1
 #     12652 grep 		  451	       1
 #     12658 bash 		  451	       1
-#)
+#
+)
 min(x$PRTIME)
-old_xmin <- 128240
+old_xmin <- 120000
 max(x$PRTIME)
-old_xmax <- 128260
+old_xmax <- 170000
 mean(x$PRTIME)
 min(x$PRTIME)
 max(x$PRTIME)
 mean(x$PRTIME)
 sd(x$PRTIME)## 2.6
-sd(x$PRTIME)/mean(x$PRTIME)### 2.061408e-05
+re <- sd(x$PRTIME)/mean(x$PRTIME)### 2.061408e-05
+postscript("128_sec_pt_new.eps")
+plot(x$PRTIME, xlim=c(0,max(x$ITERNUM)), ylim=c(old_xmin, old_xmax), 
+	pch=1, main='', 
+	xlab='Repetition', ylab='Process Time (msec)'
+)
+options(scipen = 999)
+legend(380, 171500, c(as.expression(paste("# of retained samples = ",nrow(x))), 
+		    as.expression(paste("standard deviation = ",round(sd(x$PRTIME),1),"msec")), 
[…]

### 2017-02-02 sedona-ieice16/revision/paper.tex
SEDONA updates

@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on PUT128.
-Most of the 800 executions measured in elapsed time, 
-via  {\tt System.currentTimeMillis()} in Java, are gathered in the tallest bar 
-of \hbox{128,000--129,000 msec,} with just a few executions outside of that bar, as can be seen in Fig.~\ref{fig:hist_before_et}. 
-But these outliers significantly impact the overall measurement quality: 
-a standard deviation of 1,700 msec and a relative error of 0.03. 
-\shorten{This result suggests adequately eliminating these outliers 
+800~times. 
+%Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on PUT128.
+{\color{blue}Fig.~\ref{fig:meas_comp} shows two plots of the timing results on PUT128.
+%Most of the 800 repetitions measured in elapsed time, 
+%via  {\tt System.currentTimeMillis()} in Java, are gathered in the tallest bar 
+%of \hbox{128,000--129,000 msec,} with just a few executions outside of that bar, 
+%as can be seen in Fig.~\ref{fig:hist_before_et}. 
+Most of the 800 repetitions measured in elapsed time, 
+via  {\tt System.currentTimeMillis()} in Java, form the central band 
+of \hbox{128,000--129,000 ms,} as can be seen in Fig.~\ref{fig:hist_before_et}. 
+There also exist over a dozen of repetitions above that band, called outliers.
+But these outliers {\em critically} impact the overall measurement quality: 
+a standard deviation of 1,730 ms and a relative error of 0.01.
+This result suggests adequately eliminating these outliers 
 while retaining most of the measured executions.}
 
-In contrast, our scheme, which we propose shortly, 
-significantly reduces such variability.
-The gist of the scheme is to (i)~focus on
-{\em process time} (PT) rather than on elapsed time (ET) and 
-(ii)~remove some executions that involve
-{\em infrequent, \hbox{long-running} daemon processes}, which impact
-the process time of the process. 
+{\color{blue}
+Our protocol, which we propose shortly, on the contrary, 
+effectively identifies and removes those outliers. 
+In effect, the protocol considerably reduces such variability.
+The protocol identified and removed fifteen repetitions 
+that had {\em infrequent, \hbox{long-running} daemon processes}, 
+impacting the execution time of PUT128. 
+In addition, the protocol utilized {\em process time} (PT) (of PUT128) 
+rather than elapsed time (ET). 
 Using PT is preferred, in that it considers 
@@
 provided by the Linux NetLink facility~\cite{Netlink}.
-This use of PT and selective elimination 
-was able to improve 
-measurement quality---in both standard deviation and relative error---up to by 
-three orders of magnitude for PUT128, as illustrated in Fig.~\ref{fig:hist_after_pt}.
+As a result, none of the retained timing results was outside of 
+the central band in Fig.~\ref{fig:hist_after_pt}, and thus 
+the measurement quality---in both standard deviation and relative error---was 
+improved up to by three orders of magnitude for PUT128, 
+as shown in Fig.~\ref{fig:hist_after_pt}.
+}
+
+%The key idea to the protocol is to (i)~utilize
+%{\em process time} (PT) rather than on elapsed time (ET) and 
+%(ii)~throw out some executions that involve
+%{\em infrequent, \hbox{long-running} daemon processes}, which impact
+%the process time of the process. 
+%Using PT is preferred, in that it considers 
+%the time taken for only a process of interest.
+%PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
+%in user and system mode, available via {\tt taskstats} C struct, 
+%provided by the Linux NetLink facility~\cite{Netlink}.
+%This use of PT and selective elimination 
+%was able to improve 
+%measurement quality---in both standard deviation and relative error---up to by 
+%three orders of magnitude for PUT128, as illustrated in Fig.~\ref{fig:hist_after_pt}. 
+%}
 

### 2017-02-03 sedona-ieice16/revision/paper.tex
SEDONA updates

@@
 
+{\color{blue}
+We compared the performance of ORG and \hbox{SEDONA} on 
+real programs such as insertion sort, termed SORT and 
+matrix multiplication (in column major), termed MM. 
+SEDONA overwhelmed ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
+As workload size increased, 
+the performance gap between SEDONA and ORG got increasingly widened, reaching up to by 6x.
+
+\begin{figure}[htp!]
+	\centering
+	\subfigure[SORT - Standard Deviation]{
+		\includegraphics[scale=0.31]{sort_std.eps}
+		\label{fig:sort_std}
+	}
+	\subfigure[SORT - Relative Error]{
+		\includegraphics[scale=0.31]{sort_re.eps}
+		\label{fig:sort_re}
+	}
+	\subfigure[MM - Standard Deviation]{
+		\includegraphics[scale=0.31]{matc_std.eps}
+		\label{fig:matc_std}
+	}
+	\subfigure[MM - Relative Error]{
+		\includegraphics[scale=0.31]{matc_re.eps}
+		\label{fig:matc_re}
+	}
+	\vspace{-.1in}
+	\caption{Performance Comparison on Real-world Programs~\label{fig:synprog_test}}
+\vspace{-.2in}
+\end{figure}
+}
+
 Table~\ref{tab:spec_real} shows that

### 2017-02-03 sedona-ieice16/revision/paper.tex
SEDONA paper updates

@@
 how to achieve {\em better} timing has not been well addressed. 
-{\color{blue} Surprisingly, there is considerable variability in the measured time.
+Surprisingly, there is considerable variability in the measured time.
 %which indeed suggests the need for a better timing protocol.
-The goal of this paper is to propose
+{\color{red} The goal of this paper is to propose
 a better timing protocol that can significantly reduce such variability 
@@
 {\color{blue}
-Our protocol, which we propose shortly, on the contrary, 
-effectively identifies and removes those outliers. 
-In effect, the protocol considerably reduces such variability.
-The protocol identified and removed fifteen repetitions 
-that had {\em infrequent, \hbox{long-running} daemon processes}, 
-impacting the execution time of PUT128. 
-In addition, the protocol utilized {\em process time} (PT) (of PUT128) 
-rather than elapsed time (ET). 
-Using PT is preferred, in that it considers 
-the time taken for only a process of interest.
-PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
-in user and system mode, available via {\tt taskstats} C struct, 
+Our protocol, which we characterize shortly, on the contrary, 
+exactly identified and removed those outliers. 
+In effect, the protocol considerably reduced such variability, 
+as illustrated in Fig.~\ref{fig:hist_after_pt}. 
+Specifically, regarding standard deviation and relative error 
+the protocol presented a significant performance gain up to by 
+three orders of magnitude for the same PUT128 results.
+}
+
+{\color{red}
+{\bf Characteristics.} There are several distinctions of our protocol. 
+First, we utilize {\em process time} (PT) rather than on elapsed time (ET). 
+The use of PT is preferred, as it takes into account 
+the time taken for only a ``process'' (or program) of interest.
+PT is defined as the sum of ticks (where one tick is equal to 10 msec)
+in user and system mode. 
+These tick measures are obtainable via {\tt taskstats} C struct, 
 provided by the Linux NetLink facility~\cite{Netlink}. 
-As a result, none of the retained timing results was outside of 
-the central band in Fig.~\ref{fig:hist_after_pt}, and thus 
-the measurement quality---in both standard deviation and relative error---was 
-improved up to by three orders of magnitude for PUT128, 
-as shown in Fig.~\ref{fig:hist_after_pt}.
+Second, the protocol throws out some repetitions that involve
+{\em infrequent, \hbox{long-running} daemon processes}. 
+We observe that such daemon processes substantially impact 
+the execution time of the process.  
+By eliminating those repetitions, the protocol greatly improves
+the overall measurement quality, which would otherwise be biased by them.
 }
+%The protocol thus can considerably reduce such variability.
+%The protocol identified and removed fifteen repetitions 
+%that had {\em infrequent, \hbox{long-running} daemon processes}, 
+%impacting the execution time of PUT128. 
+%As a result, none of the retained timing results was outside of 
+%the central band in Fig.~\ref{fig:hist_after_pt}, and thus 
+%the measurement quality---in both standard deviation and relative error---was 
+%improved up to by three orders of magnitude for PUT128, 
+%as shown in Fig.~\ref{fig:hist_after_pt}.
+
+%In addition, the protocol utilized {\em process time} (PT) (of PUT128) 
+%rather than elapsed time (ET). 
+%Using PT is preferred, in that it considers 
+%the time taken for only a process of interest.
+%PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
+%in user and system mode, available via {\tt taskstats} C struct, 
+%provided by the Linux NetLink facility~\cite{Netlink}.
 

### 2017-02-06 sedona-ieice16/revision/reviews/review_response.tex
SEDONA updates

@@
 
-I understand your concern. I respond to individually below. 
-Please also note that your (excellent) comments are incorporated into the revision.
+Your concerns are well understood. I respond to individually below. 
+%Your comments are incorporated into the revision.
 
@@
 
-I understand that the goal of this paper is not well articulated. 
-Yes, SEDONA is a generic protocol, which can be applied to any arbitrary Linux system 
-when it comes to reducing variability of the execution time of a program of interest. 
-I now state the actual goal, marked in blue, in the introduction section.
+The described plan (in the previous paper) is one possibility in the future, 
+but that is not stated as this paper's research goal. 
+Our goal of this paper is to propose a {\em generic, better} timing protocol 
+against the conventional approach of utilizing elapsed time. 
+The goal is successfully achieved by significantly reducing variability of 
+the execution time of a program of interest in our experiments. 
+I now clearly state this goal, marked in red, at the last sentence in Section~1.
 
-Note also that the mention of my plan in the future work 
-is just one aspect of the extension of the SEDONA protocol. 
-I claim that the combination is not the ultimate research goal of this paper. 
-To avoid confusion, I removed that sentence.
+%Note also that the mention of my plan in the future work 
+%is just one aspect of the extension of the SEDONA protocol. 
+%I claim that the combination is not the ultimate research goal of this paper. 
+%To avoid confusion, I removed that sentence.
 
@@
 
+Your concern is understood, but it is hard to agree with the comment that the improvement is little, for the following reasons.
+First, for a certain workload such as {\tt 434}, 
+our protocol reduced the variability was reduced up to by about 10x (from 75 ms to 7 ms), 
+which is *not* small. (Please refer to the last sentence 
+in the first paragraph in the second column on page~4.)
+There were also some other workloads such as {\tt 410} and {\tt 445} 
+in which the improvement reached about 3x, respectively. 
+Furthermore, every benchmark's standard deviation by our protocol 
+was less than that of the conventional method. 
+The relative error of the benchmarks was on average improved 
+by 1.5x as well.
+
+Second, our protocol scaled well over increasing execution time. 
+(Please refer to the second paragraph in the right column on page~4.)
+The relative error of our protocol was still smaller than that of the ORG method 
+for a long-running benchmark like {\tt 436} and {\tt 454} 
+as well as for a medium-length or a short-length one. 
+Such scalability was still observed in our new experiment, 
+which I describe shortly, as shown in Fig.~5.
+
+Note also that the ET results by the ORG method were based 
+on the settings taking care of several identified timing factors, such as
+(i)~deactivating non-critical daemons, 
+(ii)~switching on the network timing protocol daemon, and 
+(iii)~kernel upgrade. 
+(The description of these settings is marked in red in the paragraph below Fig.~1.) 
+In the well-controlled measurement environment, our protocol further 
+reduced the variability of execution time against the ORG method
+in such various benchmarks, which actually demonstrated its general purpose and wide usability.
+
+Lastly, the results from a new experiment provide 
+another strongly evidence that our protocol is widely practical and useful. 
+In the new experiment we evaluated the performance of our protocol 
+on some real-world programs like insertion sort and matrix multiplication (in column major). 
+The experimental results showed that the performance gap between the original and our methods 
+was increasingly widened, e.g., up to by 6x, as workload level increased.
+
+Therefore, I strongly claim that these various experiments confirm 
+the wide acceptance and usability of SEDONA as an alternative to the ET-based measurement technique.
+
 \rev{ 
@@
 
-This review is the same as (2). Please refer to our response of (2) for this comment.
+This review is basically the same as (2). Please refer to our response of (2) for this comment. 
+One note. You may feel that the extent of the improvement varies widely, 
+because the SPEC CPU2006 includes a set of very different compute-bound benchmarks. 
+But even if so, we found that all the benchmarks in the suite benefited from our timing protocol, and in addition, 
+for some workloads, our protocol significantly reduced their variability up to by 10x.
 
@@
 (4) Since the explanation of ``dual-PUT'' is not enough, I can not well
-understand the algorithm shown in Fig. 3. Similarly, what is “1st
+understand the algorithm shown in Fig. 3. 
+}
+
+More descriptions of the term ``dual-PUT'' is provided in Section~2. 
+\rev{
+Similarly, what is “1st
 Half’s ET (ms)” attached at Fig. 4(b), 4(c), 4(d) ? I understood only that

### 2017-02-06 sedona-ieice16/revision/paper.tex
SEDONA updates

@@
 \documentclass[letter]{ieice}
+\usepackage{balance}
+\usepackage{flushend}
 \usepackage[ruled,linesnumbered]{algorithm2e}
@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. 
+800~times. {\color{red}Note that PUT128 was run on a machine, with 
+a specification illustrated in Table~\ref{tab:machine_config}, 
+in a controlled environment 
+with (i)~many non-critical daemon processes deactivated, (ii)~the 
+network timing protocol daemon enabled, and (iii)~a kernel up-to-date.}
+
 %Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on PUT128.
@@
 of \hbox{128,000--129,000 ms,} as can be seen in Fig.~\ref{fig:hist_before_et}. 
-There also exist over a dozen of repetitions above that band, called outliers.
-But these outliers {\em critically} impact the overall measurement quality: 
+Although the environment restrained some identified timing factors, 
+there exist over a dozen of repetitions above that band, called outliers. 
+Indeed, these outliers {\em critically} impact the overall measurement quality: 
 a standard deviation of 1,730 ms and a relative error of 0.01.
@@
 Our contributions are following.
-\vspace{-0.12in}
+%\vspace{-0.12in}
 \begin{itemize}
@@
 \item We evaluate the performance of the protocol with rigorous experiments, 
-starting from a simple program in pure-computation mode 
-to a popular industrial benchmark suite.
+starting from a simple program in pure-computation mode {\color{blue}through 
+some memory-intensive real-world programs} to a popular industrial benchmark suite.
 
@@
 
-\begin{figure}[t]
+\begin{figure}[ht]
 	\centering
@@
     (equivalent to a Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
-    \vspace{-0.16in}     
+    \vspace{-0.2in}     
 \end{figure}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \centering
@@
   Process Time \hbox{(Step~9)}\label{tab:final_infrequent_cutoff}}
-  \vspace{-0.4in}
+  \vspace{-0.2in}
 \end{table}
@@
 
-\vspace\fill
+%\vspace\fill
 
@@
 the performance gap between SEDONA and ORG got increasingly widened, reaching up to by 6x.
-
-\begin{figure}[htp!]
-	\centering
-	\subfigure[SORT - Standard Deviation]{
-		\includegraphics[scale=0.31]{sort_std.eps}
-		\label{fig:sort_std}
-	}
-	\subfigure[SORT - Relative Error]{
-		\includegraphics[scale=0.31]{sort_re.eps}
-		\label{fig:sort_re}
-	}
-	\subfigure[MM - Standard Deviation]{
-		\includegraphics[scale=0.31]{matc_std.eps}
-		\label{fig:matc_std}
-	}
-	\subfigure[MM - Relative Error]{
-		\includegraphics[scale=0.31]{matc_re.eps}
-		\label{fig:matc_re}
-	}
-	\vspace{-.1in}
-	\caption{Performance Comparison on Real-world Programs~\label{fig:synprog_test}}
-\vspace{-.2in}
-\end{figure}
 }
@@
 {\tt 434.zeusmp}.
+
+\begin{figure}[t]
+	\centering
+	\subfigure[{\color{blue}SORT - Standard Deviation}]{
+		\includegraphics[scale=0.31]{sort_std.eps}
+		\label{fig:sort_std}
+	}
+	\subfigure[{\color{blue}SORT - Relative Error}]{
+		\includegraphics[scale=0.31]{sort_re.eps}
+		\label{fig:sort_re}
+	}
+	\subfigure[{\color{blue}MM - Standard Deviation}]{
+		\includegraphics[scale=0.31]{matc_std.eps}
+		\label{fig:matc_std}
+	}
+	\subfigure[{\color{blue}MM - Relative Error}]{
+		\includegraphics[scale=0.31]{matc_re.eps}
+		\label{fig:matc_re}
+	}
+	\vspace{-.1in}
+	\caption{{\color{blue}Performance Comparison on Real-world Programs~\label{fig:synprog_test}}}
+\vspace{-.2in}
+\end{figure}
 
@@
 \vspace{-0.1in}
-\begin{table}[h]
+\begin{table}[t]
 \centering
@@
 \caption{Performance Evaluation on the SPEC Benchmarks\label{tab:spec_real}}
-\vspace{-0.33in}
+\vspace{-0.25in}
 \end{table}
-\vspace{0.5in}
+%\vspace{0.5in}
 
@@
 
+\balance
+
 \vspace{-0.03in}

### 2017-02-06 sedona-ieice16/revision/reviews/review_response.tex
SEDONA updates almost done

@@
 
-Your concerns are well understood. I respond to individually below. 
+Your concerns are well understood. I found such statistics and heuristic approaches 
+very useful to catch what system daemons are infrequent, long-running. 
+Also, the approaches make our protocol generic, in that 
+such statistics and heuristics can be computed and 
+applied on any Linux system. The experimental results show that 
+our protocol is more efficient than the traditional method relying on elapsed time.
+
+I now respond to individually below.
 %Your comments are incorporated into the revision.
@@
 
-The described plan (in the previous paper) is one possibility in the future, 
-but that is not stated as this paper's research goal. 
-Our goal of this paper is to propose a {\em generic, better} timing protocol 
-against the conventional approach of utilizing elapsed time. 
-The goal is successfully achieved by significantly reducing variability of 
-the execution time of a program of interest in our experiments. 
+I understand your confusion. The previously described plan was suggested as one of the possibilities in the future work. But that is not the research goal of this paper. 
+As you already recognized, the goal of this paper is to propose a {\em generic, better} timing protocol against the conventional approach of utilizing elapsed time, 
+to enable better timing results. This goal is successfully achieved 
+by significantly reducing variability in timing as demonstrated in our various experiments.
+%The goal is successfully achieved by significantly reducing variability of 
+%the execution time of a program of interest in our experiments. 
 I now clearly state this goal, marked in red, at the last sentence in Section~1.
+
+To avoid confusion, I removed the last mention. Also, the first paragraph on page~2, 
+which used to provide an explanation of a possible combination of the query timing protocol 
+with SEDONA, was removed.
 
@@
 
-Note also that the ET results by the ORG method were based 
-on the settings taking care of several identified timing factors, such as
-(i)~deactivating non-critical daemons, 
-(ii)~switching on the network timing protocol daemon, and 
-(iii)~kernel upgrade. 
-(The description of these settings is marked in red in the paragraph below Fig.~1.) 
-In the well-controlled measurement environment, our protocol further 
-reduced the variability of execution time against the ORG method
-in such various benchmarks, which actually demonstrated its general purpose and wide usability.
+Note also that the ET results by the ORG method were also based 
+on the settings taking care of several identified timing factors as
+described in the third paragraph in Section~2.
+Under such a well-configured timing environment, our protocol further 
+reduced the variability of execution time compared to the ORG method
+in such various benchmarks, 
+which actually demonstrated its general purpose and wide usability.
 
@@
 because the SPEC CPU2006 includes a set of very different compute-bound benchmarks. 
-But even if so, we found that all the benchmarks in the suite benefited from our timing protocol, and in addition, 
-for some workloads, our protocol significantly reduced their variability up to by 10x.
+But even if so, we found that all the benchmarks in the suite benefited from our timing protocol, 
+and in addition, for some workloads, our protocol significantly reduced their variability up to by 10x.
 
@@
 (4) Since the explanation of ``dual-PUT'' is not enough, I can not well
-understand the algorithm shown in Fig. 3. 
-}
-
-More descriptions of the term ``dual-PUT'' is provided in Section~2. 
-\rev{
-Similarly, what is “1st
+understand the algorithm shown in Fig. 3. Similarly, what is “1st
 Half’s ET (ms)” attached at Fig. 4(b), 4(c), 4(d) ? I understood only that
@@
 
+Your point is well taken. 
+The term of ``dual-PUT'' represents 
+a pair of two consecutive samples of a run of a (regular) PUT. 
+(The length of that regular PUT is virtually doubled as dual-PUT.) More descriptions of the term ``dual-PUT'' are provided with an easy example 
+in the second paragraph in the right column on page~2. 
+In addition, the 1st half indicates the first element of the pair 
+whereas the 2nd half represents the second element of the same pair. 
+The label of ``1st Half's ET (ms)'' is renamed as ``Odd Samples' ET(ms),'' 
+for better clarification. Please refer to the following (third) paragraph.
+
 \end{document}

### 2017-02-06 sedona-ieice16/revision/paper.tex
SEDONA updates almost done

@@
 }
-\received{2017}{2}{28}
+\received{2017}{2}{10}
 
@@
 \begin{summary}
-%It is challenging to accurately measure the execution-time of a given
-%program due to many extraneous factors. To enable better timing results\shorten{restrained},
-%we present a novel protocol that identifies infrequent, 
-%long-running daemons to enable effective elimination of executions 
-%exhibiting these daemons.
 {\color{blue}Measuring program execution time 
@@
 This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
-%The $tl$ value is translated to the corresponding 
-%number of {\color{blue}repetitions} for which that for-loop is performed. 
 {\color{blue}The $tl$ value is used to compute 
-the number of repetitions ($t$) for which that for-loop is performed 
+the number of iterations ($t$) for which that for-loop is performed 
 to reach the specified task length.}
@@
 with (i)~many non-critical daemon processes deactivated, (ii)~the 
-network timing protocol daemon enabled, and (iii)~a kernel up-to-date.}
+network timing protocol daemon enabled, and (iii)~an up-to-date kernel.}
 
-%Fig.~\ref{fig:meas_comp} shows two histograms of the timing results on PUT128.
 {\color{blue}Fig.~\ref{fig:meas_comp} shows two plots of the timing results on PUT128.
-%Most of the 800 repetitions measured in elapsed time, 
-%via  {\tt System.currentTimeMillis()} in Java, are gathered in the tallest bar 
-%of \hbox{128,000--129,000 msec,} with just a few executions outside of that bar, 
-%as can be seen in Fig.~\ref{fig:hist_before_et}. 
-Most of the 800 repetitions measured in elapsed time, 
+Most of the 800 executions measured in elapsed time, 
 via  {\tt System.currentTimeMillis()} in Java, form the central band 
 of \hbox{128,000--129,000 ms,} as can be seen in Fig.~\ref{fig:hist_before_et}. 
-Although the environment restrained some identified timing factors, 
-there exist over a dozen of repetitions above that band, called outliers. 
-Indeed, these outliers {\em critically} impact the overall measurement quality: 
-a standard deviation of 1,730 ms and a relative error of 0.01.
-This result suggests adequately eliminating these outliers 
-while retaining most of the measured executions.}
+Despite the well-controlled timing environment, there exist over a dozen of executions above that band, or {\em outliers}.
+Indeed, these outliers {\em critically} impact the overall measurement quality: a standard deviation of 1,730 ms and a relative error of 0.01.
+This result suggests adequately eliminating these outliers while retaining most of the executions.}
 
@@
 provided by the Linux NetLink facility~\cite{Netlink}. 
-Second, the protocol throws out some repetitions that involve
+Second, the protocol throws out some executions that involve
 {\em infrequent, \hbox{long-running} daemon processes}. 
@@
 the execution time of the process.  
-By eliminating those repetitions, the protocol greatly improves
+By eliminating those executions, the protocol greatly improves
 the overall measurement quality, which would otherwise be biased by them.
 }
-%The protocol thus can considerably reduce such variability.
-%The protocol identified and removed fifteen repetitions 
-%that had {\em infrequent, \hbox{long-running} daemon processes}, 
-%impacting the execution time of PUT128. 
-%As a result, none of the retained timing results was outside of 
-%the central band in Fig.~\ref{fig:hist_after_pt}, and thus 
-%the measurement quality---in both standard deviation and relative error---was 
-%improved up to by three orders of magnitude for PUT128, 
-%as shown in Fig.~\ref{fig:hist_after_pt}.
 
-%In addition, the protocol utilized {\em process time} (PT) (of PUT128) 
-%rather than elapsed time (ET). 
-%Using PT is preferred, in that it considers 
-%the time taken for only a process of interest.
-%PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
-%in user and system mode, available via {\tt taskstats} C struct, 
-%provided by the Linux NetLink facility~\cite{Netlink}.
-
-%The key idea to the protocol is to (i)~utilize
-%{\em process time} (PT) rather than on elapsed time (ET) and 
-%(ii)~throw out some executions that involve
-%{\em infrequent, \hbox{long-running} daemon processes}, which impact
-%the process time of the process. 
-%Using PT is preferred, in that it considers 
-%the time taken for only a process of interest.
-%PT can be calculated as the sum of ticks (where one tick is equal to 10 msec)
-%in user and system mode, available via {\tt taskstats} C struct, 
-%provided by the Linux NetLink facility~\cite{Netlink}.
-%This use of PT and selective elimination 
-%was able to improve 
-%measurement quality---in both standard deviation and relative error---up to by 
-%three orders of magnitude for PUT128, as illustrated in Fig.~\ref{fig:hist_after_pt}. 
-%}
-
-%\vspace{0.05in}
 {\bf Related Work.} 
@@
 
-%\begin{figure}[t]
-%	\centering
-%	\subfigure[Initial Elapsed Time Measurement]{
-%		%\includegraphics[width=0.31\textwidth]{128_sec_et_old_hist.eps}
-%		\includegraphics[width=0.31\textwidth]{128_sec_et_old.eps}
-%		\label{fig:hist_before_et}
-%	}
-%	\subfigure[Program Time After Our Protocol]{
-%		%\includegraphics[width=0.31\textwidth]{128_sec_pt_new_hist.eps}
-%		\includegraphics[width=0.31\textwidth]{128_sec_pt_new.eps}
-%		\label{fig:hist_after_pt}
-%	}
-%	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
-%	Program-Under-Test~\label{fig:meas_comp}}
-%	\vspace{-.18in}
-%\end{figure}
-
 \begin{figure}[t]
 	\centering
-	\subfigure[Initial Elapsed Time Measurement]{
-		%\includegraphics[width=0.31\textwidth]{128_sec_et_old_hist.eps}
+	\subfigure[{\color{blue}Initial Elapsed Time Measurement}]{
 		\includegraphics[width=0.224\textwidth]{128_sec_et_old.eps}
@@
 	}
-	\subfigure[Program Time After Our Protocol]{
-		%\includegraphics[width=0.31\textwidth]{128_sec_pt_new_hist.eps}
+	\subfigure[{\color{blue}Program Time After Our Protocol}]{
 		\includegraphics[width=0.224\textwidth]{128_sec_pt_new.eps}
@@
 	}
-	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
+	\caption{{\color{blue}An Example of Measurement Results on a \linebreak \hbox{Compute-bound}} 
 	Program-Under-Test~\label{fig:meas_comp}}
@@
 
-We previously developed a timing protocol called TTP (Tucson Timing Protocol)
-for programs exhibiting I/O, in particular, query execution time in \hbox{DBMSes}~\cite{Currim}.
-Our study identified a variety of Linux measures 
-(e.g., user ticks, system ticks, IOWait ticks, etc.) 
-relevant to timing a single query and then 
-presented a structural causal model 
-of explicating the variance of query time. 
-Based on this model, we proposed the protocol for calculating the query time. 
-Our scheme we propose in this manuscript 
-is applicable to the TTP protocol, to further improve the query time calculation.
-
-%\vspace{0.05in}
+\vspace{0.05in}
 {\bf Contribution.} 
 Our contributions are following.
-%\vspace{-0.12in}
+\vspace{-0.05in}
 \begin{itemize}
@@
 \item We show empirical evidence that 
-measuring program time 
+measuring execution time 
 can be seriously affected by daemons.
@@
 
-\item We evaluate the performance of the protocol with rigorous experiments, 
-starting from a simple program in pure-computation mode {\color{blue}through 
-some memory-intensive real-world programs} to a popular industrial benchmark suite.
+\item {\color{blue}We evaluate the performance of the protocol with rigorous experiments 
+using real-world programs.}
 
@@
 The rest of this letter is organized as follows. 
-Section~\ref{sec:prop_appach} elaborates on the proposed scheme. 
-In the following section, we evaluate the performance of 
-the scheme using real workloads.
+{\color{blue}We next elaborate on the proposed timing protocol. 
+In turn, we evaluate the performance of the scheme using real workloads. 
+Finally, we conclude this letter.}
 
@@
 
-In this section we propose a novel execution-time measurement scheme, 
+In this section we propose a novel execution time measurement scheme, 
 called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons).
-Our scheme catches and eliminates executions including daemon processes that are infrequent and 
-long-running via a {\em cutoff} measure, and significantly improves measurement quality. 
-The SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
-%We first enumerate the steps to (a)~identify infrequent, 
-%long-running daemons, using a single run of many executions of PUT128, 
-%(b)~refine the list using a single run of many executions of PUT16384 (with a task length of 16K sec), 
-%and then (c)~use the data from those two runs to determine the cutoff for 
-%each \hbox{so-identified} daemon. 
-%We can then apply these cutoffs to remove outliers from subsequent runs of any 
-%arbitrary PUT and realistic workloads.
-%\vspace{-0.22in}
+{\color{blue} Our scheme (1) utilizes PT (process time) to avoid absorbing timing noises from other co-running processes and (2) catches and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
+%These two features contribute to significantly enhancing measurement quality.
+}
+
+Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
+{\color{blue}Note that this protocol can be applied to any arbitrary Linux system.}
 
@@
 \end{figure}
-%
 
-%We now elaborate the SEDONA scheme with an example, to explain and justify each step.
-%\vspace{-0.8in}
-
-%\subsection{A Running Example} 
-%\vspace{-0.05in}
 {\bf A Running Example.} 
@@
 iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1).
-%(Our machine's specification will be given in Tab~\ref{tab:machine_config}).
 
@@
 
-Fig.~\ref{fig:reg_put128} plots all the 800 elapsed times of the run of PUT128.
+Fig.~\ref{fig:reg_put128} plots all the 800 elapsed times (ETs) of the run of PUT128.
 %We see three rows in the plot: a solid row of many samples, perhaps
@@
 show how to reliably eliminate the indirect influence of 
-some ``infrequent, long-running daemons'' on the process time (PT) 
-of the PUT.
+some ``infrequent, long-running daemons'' on PT (\hbox{process} time) of the PUT.
 
-To identify such daemons, we use a novel scatter plot: 
-those of {\em pairs of successive samples}.\shorten{So samples 1 and 2 
-form the first pair, samples 3 and 4 form the second pair.} 
+To identify such daemons, we use a novel \hbox{scatter} plot: 
+those of {\em pairs of successive samples}{\color{blue}. 
+So \hbox{samples}~\#{1} and~\#{2} form the first pair, 
+samples~\#{3} and~\#{4} form the second pair.
+Such a pair is termed ``\hbox{{\it dual-PUT}}.'' 
+A sample of \hbox{dual-PUT256}, for instance, 
+consists of two consecutive (odd and even) 
+samples of PUT128: 
+sample~\#{1} of dual-PUT256 is equivalent 
+to samples~\#{1} and \#{2} of PUT128.}
+
 Fig.~\ref{fig:raw_put128} presents such a scatter plot 
-of 400 samples of a \hbox{dual-PUT256} constructed 
+of 400 samples of a run of \hbox{dual-PUT256} constructed 
 from a run of the 800 PUT128 samples (Step~3). 
-There are two quite obvious outliers\shorten{, corresponding to sample \# 75 and \# 634,} with ETs of 163,913 msec (rightmost) and 161,785 msec (uppermost), respectively.
+{\color{blue} The $xy$-plane corresponds to the dual-PUT samples; 
+for example, \hbox{sample}~\#{2} of a run of dual-PUT256 
+is plotted as a point with samples \#{3} and~\#{4} of a run of PUT128 
+on the respective $x$ and $y$ axes. In that sense 
+the $x$ ($y$) axis is named as ``Odd (Even) Sample's ET.''}
+There are two quite obvious outliers with ETs 
+of 163,913 ms (rightmost) and 161,785 ms (uppermost), respectively.
 
@@
 	}
-	\subfigure[Dual-PUT256 Measurements]{
+	\subfigure[Dual-PUT256 Samples]{
 		%\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4.eps}
-		\includegraphics[width=0.224\textwidth]{dual_et_put256_empv4_new.eps}
+		\includegraphics[width=0.224\textwidth]{dual_et_put256_empv4.eps}
 		\label{fig:raw_put128}
 	}
-	\vspace{0.1in}    
+	%\vspace{-0.1in}    
 	\subfigure[Zooming in on the $L$-shape]{
@@
 	}
-%	\subfigure[Further Zooming in on the $L$-Shape]{
-%		\includegraphics[scale=0.39]{further_zooming_in_dual_et_put256_empv4.eps}
-%		\label{fig:fzoom_in_dual_et_put256}
-%	}
 	\subfigure[The Central Cluster]{
@@
     (equivalent to a Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
-    \vspace{-0.2in}     
+    \vspace{-0.3in}     
 \end{figure}
@@
 $L$-shape, with perhaps a dozen or more $L$-samples in the left and bottom arms of
-the ``$L$,'' and again no samples in the upper right portion of the
+the ``$L$'', and again no samples in the upper right portion of the
 scatter plot.
@@
 
-\section{Conclusion and Future Work}
+\section{Conclusion}
 \label{sec:conclusion}
@@
 the presence of infrequent, long-running daemon processes. 
-We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
+We }proposed a novel timing protocol called {\em SEDONA}\shorten{, 
 employing process time and identifying an infrequent, long-running daemon}, 
-which is more precise and \hbox{accurate} than the traditional method\shorten{
+which is more precise and \hbox{accurate} than the traditional method relying on elapsed time\shorten{
 as shown in the experiments}.
-Our plan is to integrate SEDONA into the query timing protocol~\cite{Currim}.
+%Our plan is to integrate SEDONA into the query timing protocol~\cite{Currim}.
 %Our evaluation using real benchmarks demonstrated the strong effectiveness of 

### 2017-02-06 sedona-ieice16/revision/data/fig_script.r

@@ -1,36 +1,3 @@
-x = read.csv(file="all_spec_before.dat",head=TRUE,sep="\t")
-y = read.csv(file="all_spec_after.dat",head=TRUE,sep="\t")
-specCpu.std <- cbind(ORG_ET_std=(x$STD_ET),FIND_ET_std=(y$STD_PT))
-specCpu.std <- round(specCpu.std,1)
-setEPS()
-postscript("spec_std.eps")
-b<-barplot(t(specCpu.std[1:31,]), log="y", yaxt="n", main="", beside=TRUE, 
-names.arg=c("400","401","403","410","416","429","433","434","435","436","437",
-	    "444","445","447","450","453","454","456","458","459","462","464",
-	    "465","470","471","473","481","482","483","998","999"), 
-	las=2,xlab="SPEC CPU2006",ylab="Standard deviation in log scale (msec)", ylim=c(0.1,8192), col=c("tan2","blue"))
-#legend(x=85.1,y=8000, c("ORG: ET", "FIND: PT"),cex=.8, col=c("tan2","blue"),pch=c(22,0,0))
-legend(x=9,y=8000, c("ORG-ET", "FIND-PT"),cex=.8, col=c("tan2","blue"),pch=c(22,0,0))
-ticks <- seq(-1, 16, by=1)
-axis(2, at=c(-1,1, 2, 4, 8, 16, 32, 64, 128, 256, 512, 1024, 2048, 4096, 8192,16384))
-dev.off()
-
-x = read.csv(file="all_spec_before.dat",head=TRUE,sep="\t")
-y = read.csv(file="all_spec_after.dat",head=TRUE,sep="\t")
-specCpu.std <- cbind(ORG_ET_re=(x$RE_ET),FIND_ET_re=(y$RE_PT))
-#specCpu.std <- round(specCpu.std,1)
-setEPS()
-postscript("spec_re.eps")
-b <- barplot(t(specCpu.std[1:31,]), log="y", yaxt="n", main="", beside=TRUE, 
-names.arg=c("400","401","403","410","416","429","433","434","435","436","437","444", "445","447","450","453","454","456","458","459","462","464","465","470","471","473","481","482","483","998","999"), 
[…]

### 2017-02-06 sedona-ieice16/revision/reviews/review_response.tex
New updates

@@
 a pair of two consecutive samples of a run of a (regular) PUT. 
-(The length of that regular PUT is virtually doubled as dual-PUT.) More descriptions of the term ``dual-PUT'' are provided with an easy example 
+(The length of that regular PUT is virtually doubled as dual-PUT.) More descriptions of the term ``\hbox{dual-PUT}'' are provided with an easy example 
 in the second paragraph in the right column on page~2. 

### 2017-02-06 sedona-ieice16/revision/paper.tex
New updates

@@
   infrequent}, several having \hbox{periodicities} estimated at four or twenty hours.
-When the PUT had a ``short'' 
+When {\color{blue}PUT128} had a ``short'' 
 program time (in this case, two minutes), daemons with a periodicity of
-hours are \hbox{infrequent}. But with a PUT with a ``long'' program time (in this
-case, 4.5 hours), some of those daemons are now frequent, and appear in the 
-central cluster.
+hours are \hbox{infrequent}. But with {\color{blue}PUT16384} 
+with a ``long'' program time (in this case, 4.5 hours), some of those daemons are now frequent, and appear in the central cluster.
 

### 2017-02-07 sedona-ieice16/revision/reviews/review_response.tex
Revision done

--- 2017-02-06 sedona-ieice16/revision/reviews/review_response.tex
@@
 
-Your concerns are well understood. I found such statistics and heuristic approaches 
-very useful to catch what system daemons are infrequent, long-running. 
-Also, the approaches make our protocol generic, in that 
-such statistics and heuristics can be computed and 
-applied on any Linux system. The experimental results show that 
+Your concerns are well understood. 
+Indeed, our statistics and heuristics greatly help us 
+identify what daemons are infrequent, long-running on our system 
+(as described in the fourth paragraph in the left column on page~3). 
+Applying such statistics and heuristics can be done to any Linux system, 
+which make our protocol generic. Furthermore, the experimental results show that 
 our protocol is more efficient than the traditional method relying on elapsed time.
 
+Note that my responses to your comments 
+are incorporated into the revision and marked in {\color{red}red}. 
+There are some new and modified contents in the revision, 
+which are marked in {\color{blue}blue} for better distinction.
+
 I now respond to individually below.
-%Your comments are incorporated into the revision.
 
@@
 a pair of two consecutive samples of a run of a (regular) PUT. 
-(The length of that regular PUT is virtually doubled as dual-PUT.) More descriptions of the term ``dual-PUT'' are provided with an easy example 
-in the second paragraph in the right column on page~2. 
+(The task length of that regular PUT is virtually doubled, extended to the corresponding 
+dual-PUT.) More descriptions of the term ``\hbox{dual-PUT}'' are provided with a simple example in the second paragraph in the right column on page~2. 
 In addition, the 1st half indicates the first element of the pair 
-whereas the 2nd half represents the second element of the same pair. 
+whereas the 2nd half the second element of the same pair. 
 The label of ``1st Half's ET (ms)'' is renamed as ``Odd Samples' ET(ms),'' 
-for better clarification. Please refer to the following (third) paragraph.
+for better clarification. Please refer to the subsequent (third) paragraph.
 

### 2017-02-07 sedona-ieice16/revision/cover_letter.tex
Revision done

--- 2017-01-25 sedona-ieice16/submission/misc/cover-letter.tex
@@
 \begin{flushright}
-\hfill Supercomputing R\&D Center, 
+%\hfill National Institute of Supercomputing and Networking, 
 \end{flushright}
@@
 \begin{flushright}
-\hfill Republic of Korea, 34141, 
+\hfill Republic of Korea, 34141
 \end{flushright}
@@
 \begin{flushright}
-\hfill Email: yksuh@kisti.re.kr
+\hfill Email: youngkyoon.suh@gmail.com
 \end{flushright} 
@@
 
-%\vspace{-0.5in}
+\vspace{0.5in}
 
-\opening{Dear Editor-in-Chief:}
+\opening{Dear Prof. Rajkumar Buyya:}
 
-\noindent This letter to be submitted to IEICE Transactions on Information and Systems 
-presents the following contributions. No version of this letter was submitted earlier.
+\noindent
 
-\begin{itemize}
-
-\item We provide empirical evidence that measuring program time 
-can be seriously affected by extant system daemons.
-
-\item We propose a novel timing protocol, called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons), 
-that identifies infrequent, long-running daemons that impact the timing results for that program.
-
-\item We evaluate the performance of the protocol with rigorous experiments, starting from a simple program in pure-computation mode 
-to a popular industrial benchmark suite.
-
-\item Experimental results show a strong support for the effectiveness of our SEDONA timing protocol. 
-\end{itemize}
+Attached is a minor revision that responds to the first reviewer's comments. 
+We marked in red some text in response to his/her comments. 
+The revision also includes some texts marked in blue that are added or updated. 
+Note also that we replaced the term of {\em PUT} with another term, called {\em INC}, 
+in order to prevent readers from getting confused with a general concept of the program under test. 
+Many thanks to you for taking sincere care of our manuscript during a series of reviews.
 

### 2017-02-07 sedona-ieice16/revision/paper.tex
Revision done

--- 2017-02-06 sedona-ieice16/revision/paper.tex
@@
 Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
-{\color{blue}Note that this protocol can be applied to any arbitrary Linux system.}
+{\color{red}Note that this protocol can be applied to any arbitrary Linux system.}
 
@@
 To identify such daemons, we use a novel \hbox{scatter} plot: 
-those of {\em pairs of successive samples}{\color{blue}. 
+those of {\em pairs of successive samples}{\color{red}. 
 So \hbox{samples}~\#{1} and~\#{2} form the first pair, 
@@
 from a run of the 800 PUT128 samples (Step~3). 
-{\color{blue} The $xy$-plane corresponds to the dual-PUT samples; 
+{\color{red} The $xy$-plane corresponds to the dual-PUT samples; 
 for example, \hbox{sample}~\#{2} of a run of dual-PUT256 
@@
         }
-    \caption{Successive Scatter plots of a PUT128 with 800 samples 
-    (equivalent to a Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
+    \caption{Successive Scatter Plots of PUT128 with 800 samples 
+    (equivalent to Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
     \vspace{-0.3in}     
@@
 This computation provides a rough, initial distinction of a ``long-running''
-daemon, namely, the valley between the maximum PT from the central cluster 
+%daemon, namely, the valley between the maximum PT from the central cluster 
+daemon, or the valley between the maximum PT from the central cluster 
 and the minimum PT from the $L$-samples, to differentiate ``short-running'' 
@@
 We find some frequent daemon processes appearing in both of the 
-clusters of the dual-PUT256 and dual-PUT32768: 
+clusters of dual-PUT256 and dual-PUT32768{\color{blue}, each consisting of 
+pairs of two successive samples of PUT128 and PUT16384}: 
 {\tt flush-9.0}, {\tt jbd2/md0-8}, {\tt kblock/0}, {\tt md0\_raid1}, 
 and {\tt ntpd}. That said, the central cluster also contains other
-  processes not seen in the dual-PUT256 central cluster: 
+  processes not seen in the \hbox{dual-PUT256} central cluster: 
   {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
@@
   infrequent}, several having \hbox{periodicities} estimated at four or twenty hours.
-When the PUT had a ``short'' 
-program time (in this case, two minutes), daemons with a periodicity of
-hours are \hbox{infrequent}. But with a PUT with a ``long'' program time (in this
-case, 4.5 hours), some of those daemons are now frequent, and appear in the 
-central cluster.
+When {\color{blue}PUT128} had a ``short'' 
+program time (or, two minutes), daemons with a periodicity of
+hours are \hbox{infrequent}. But with {\color{blue}PUT16384} 
+with a ``long'' program time (or, 4.5 hours), some of those daemons are now frequent, and appear in the central cluster.
 
@@
 the cutoff for each of those infrequent, long-running daemons so
-identified, based on the PUT128 and the 
+identified, based on the runs of PUT128 and the 
 PUT16384 as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
@@
   Process Time \hbox{(Step~9)}\label{tab:final_infrequent_cutoff}}
-  \vspace{-0.2in}
+  \vspace{-0.4in}
 \end{table}
@@
 \vspace{-0.07in}
-We now evaluate the \hbox{performance} of the SEDONA \hbox{protocol}.
+We now evaluate the \hbox{performance} of SEDONA 
+compared to that of the conventional timing scheme, termed ORG, based on 
+elapsed time.
 Our experiments were conducted on a \hbox{machine}
@@
 
-We evaluated the performance of SEDONA \hbox{using} SPEC CPU2006 
-benchmarks~\cite{specCpu2006}, providing \hbox{various} compute-bound real applications. 
-%Thus, employing the SPEC benchmarks as real-world workloads 
-%was a reasonable choice for the evaluation. 
-The results are \hbox{provided} in Table~\ref{tab:spec_real}. 
+{\color{blue}
+First, we compared the performance of the two approaches 
+using real-world programs such as insertion sort, termed SORT and 
+matrix multiplication (in column major), termed MM. 
+In this comparison SEDONA overwhelmed ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
+As workload size increased, the performance gap between the two 
+got increasingly widened, reaching up to by 6x.
+}
+
+In turn, our evaluation proceeded with SPEC CPU2006 
+benchmarks~\cite{specCpu2006}, providing various \linebreak 
+\hbox{compute-bound} real applications. The results are \hbox{provided} in Table~\ref{tab:spec_real}. 
 Note that in the table the results for {\tt 481} and {\tt 483} \hbox{benchmarks} 
@@
 {\color{blue}
-We compared the performance of ORG and \hbox{SEDONA} on 
-real programs such as insertion sort, termed SORT and 
-matrix multiplication (in column major), termed MM. 
-SEDONA overwhelmed ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
-As workload size increased, 
-the performance gap between SEDONA and ORG got increasingly widened, reaching up to by 6x.
+Table~\ref{tab:spec_real} shows that 
+SEDONA outperformed ORG,
+on the standard deviation and relative error across the very different SPEC benchmarks. 
+To be more specific, all the benchmarks revealed a smaller standard deviation 
+from SEDONA as compared to that of ORG. 
+Our timing protocol quite effectively filtered out infrequent daemon executions 
+in the industrial workloads. The relative error of SEDONA was also 
+lower than that of ORG for almost every benchmark.
+{\color{red}About a 10x \hbox{margin} between the two was yielded for {\tt 434}, for example.}
 }
-
-Table~\ref{tab:spec_real} shows that
-the SEDONA protocol (which uses PT) significantly outperformed the original 
-measurement technique (using ET), termed ORG, 
-on the standard deviation and relative error across the different SPEC benchmarks. 
-None of the benchmarks revealed a bigger standard deviation from the SEDONA protocol as compared to that of ORG. 
-%All of the benchmarks revealed a smaller standard deviation on our scheme than that of the ORG. 
-Our protocol quite effectively filtered out infrequent daemon executions 
-in these real-world workloads. 
-Furthermore, the relative error of SEDONA was 
-lower than that of ORG for all the benchmarks. 
-For instance, about a 10x \hbox{margin} between the two was yielded for 
-{\tt 434.zeusmp}.
 
@@
 
+%\vspace{-.15in}
 SEDONA also scaled well for the SPEC workloads, 
@@
 {\tt 434}, {\tt 445}, and {\tt 999}: those taking \hbox{under} 100 sec), 
-our scheme outperformed the ORG by about 3.5x, on average. 
+our scheme outperformed the ORG method by about 3.5x, on average. 
 The SEDONA protocol continued its dominance against the conventional technique 
 for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-For the long-running benchmarks (e.g., {\tt 436} and {\tt 454}, both$>$ 1,000 sec), 
-the relative error of \hbox{SEDONA} was slightly lower than that of ORG.
-
-%Lastly, the average standard deviation and relative error of the SEDONA 
-%among all the benchmarks were smaller that those of the ORG, 
-%as indicated by the last two rows. 
-%Lastly, the SEDONA outperformed the ORG by a slight margin \hbox{regarding} 
-%the plain (un-weighted) and weighted averages of 
-%the standard deviation and relative error among all the benchmarks, 
-%as indicated by the last two rows.
-%\shorten{In particular, the SEDONA outperformed the ORG regarding 
-%the weighted average by normalization by a slight margin 
-%in spite of seemingly being equal because of the rounding up to one significant digit.}
-
-%Lastly, we still observed a substantial standard deviation (and relative error) for 
-%some benchmarks (e.g., {\tt 436} and {\tt 462}).
-%The high variance seemed mainly attributed to the activities
-%of the {\tt kslowd000} and {\tt kslowd001} daemon processes, 
-%as also identified in the real-world programs. 
-%These two were not observed during our cutoff study. 
-%The daemons cannot be switched off, as they involve kernel threads. 
-%More investigation is needed to hopefully better remove 
-%daemon executions interfering with timing a program.
+{\color{blue}Even for the long-running benchmarks (e.g., {\tt 436}, {\tt 416}, and {\tt 454}, both$>$ 1,000 sec), 
+the relative error of \hbox{SEDONA} was still lower than that of ORG.}
 
 To summarize, our proposed SEDONA protocol can achieve {\em better} accuracy,
-precision, and scalability in measuring the execution time of real \hbox{compute-bound} benchmarks 
-than the conventional timing method.
+precision, and scalability in measuring the execution time of real \hbox{compute-bound} benchmarks than the conventional timing method. 
+{\color{blue}The experimental results demonstrate 
+the general applicability of SEDONA to timing any CPU-bound program.}
 
@@
 \scriptsize
-%\tiny
-%\begin{tabular}{|r|r|r|r|r|r|r|} \hline
-% 		  & \multicolumn{2}{r|}{Execution Time (ms)} & \multicolumn{2}{r}{Standard} & \multicolumn{2}{|r|}{\multirow{2}{*}{Relative Error}}\\ \cline{2-3}
-%SPEC  & {ORG}  & {SEDONA} 	& \multicolumn{2}{r}{Deviation (ms)} 	 					& \multicolumn{2}{|r|}{}\\ \cline{4-7}	
-%		 & {(in ET)} &  {(in PT)} & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline
-
-%\begin{tabular}{|p{0.6cm}|p{1cm}|p{1cm}|p{0.5cm}|p{1cm}|p{0.8cm}|p{1cm}|} \hline
-%		  & \multicolumn{2}{p{2.3cm}|}{Execution Time (ms)} & \multicolumn{2}{p{1.6cm}}{Standard}  & \multicolumn{2}{|l|}{\multirow{2}{*}{Relative Error}}\\ \cline{2-3}
-%SPEC  & {ORG}  & {SEDONA} & \multicolumn{2}{p{1.6cm}}{Deviation(ms)} & \multicolumn{2}{|c|}{}\\ \cline{4-7}	
-%		 & {(in ET)} &  {(in PT)} & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline
-
 \begin{tabular}{|p{0.18cm}|R{1.14cm}|R{1.14cm}|R{0.64cm}|R{1cm}|R{0.97cm}|R{0.97cm}|} \hline
@@
 {{\tt 433}} & 480,586 & 473,256  & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-{{\tt 434}} & 16,495  & 16,242  & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
+{{\tt 434}} & 16,495  & 16,242  & {\color{red}{75}} & {\color{red}{7}} & {\color{red}{5$\times$10$^{-3}$}} & {\color{red}{5$\times$10$^{-4}$}}\\
 {{\tt 435}} & 990,575 & 975,445  & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
@@
 \end{table}
-%\vspace{0.5in}
 
-%\vspace{-0.1in}
-%\begin{table}[h]
-%\centering
-%{
-%\scriptsize
-%%\tiny
-%\begin{tabular}{|p{1.6cm}|p{1.25cm}|p{0.8cm}|p{1cm}|p{0.8cm}|p{1cm}|} \hline
-%  \multirow{2}{*}{Benchmarks} 
-%  & Weight By
-%  & \multicolumn{2}{|p{1.8cm}|}{Std. Dev. (msec)} 
-%  & \multicolumn{2}{|p{1.7cm}|}{Relative Error}\\ \cline{3-6}
-% & Exec. Time & {ORG} & {SEDONA} & {ORG} & {SEDONA}\\ \hline		
-%{{\tt 400.perlbench}} & {3$\times$10$^{-5}$} & {3} & {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-%{{\tt 401.bzip2}} & {4$\times$10$^{-2}$} & {1,185} & {1,161} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
-%{{\tt 403.gcc}} & 2$\times$10$^{-3}$ & {138} & {96} & {5$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
-%{{\tt 410.bwaves}} & 6$\times$10$^{-4}$ & {46} & {19} & {6$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
-%{{\tt 416.gamess}} & 7$\times$10$^{-2}$ & {965} & {876} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\%% V
-%{{\tt 429.mcf}} & 2$\times$10$^{-2}$ & {623} & {600} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-%{{\tt 433.milc}} & 3$\times$10$^{-2}$ & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-%{{\tt 434.zeusmp}} & 1$\times$10$^{-3}$ & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
-%{{\tt 435.gromacs}} & 7$\times$10$^{-2}$ & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
-%{{\tt 436.cactusADM}} & 8$\times$10$^{-2}$  & {3,914} & {3,843} & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-%{{\tt 437.leslie3d}} & 4$\times$10$^{-2}$ & {1,492} & {1,475}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-%{{\tt 444.namd}} & 4$\times$10$^{-2}$ & {294} & {281}  & {5$\times$10$^{-4}$} & {5$\times$10$^{-4}$}\\
-%{{\tt 445.gobmk}} & 5$\times$10$^{-3}$ & {91} & {28}  & {1$\times$10$^{-3}$} & {3$\times$10$^{-4}$}\\
-%{{\tt 447.dealII}} & 3$\times$10$^{-2}$ & {150} & {108}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
-%{{\tt 450.soplex}} &2$\times$10$^{-2}$ & {99} & {91}  & {3$\times$10$^{-4}$} & {2$\times$10$^{-4}$}\\
-%{{\tt 453.povray}} & 2$\times$10$^{-2}$ & {623} & {582}  & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\
-%%{{\tt 454.calculix}} & {678} & {627}  & {3.9$\times$10$^{-4}$} & {3.7$\times$10$^{-4}$}\\
-%{{\tt 454.calculix}} & 1$\times$10$^{-1}$ & {678} & {627}  & {4$\times$10$^{-4}$} & {4$\times$10$^{-4}$}\\
-%{{\tt 456.hmmer}} & 3$\times$10$^{-2}$ & {85} & {50}  & {2$\times$10$^{-4}$} & {1$\times$10$^{-4}$}\\
-%{{\tt 458.sjeng}} & 4$\times$10$^{-2}$ & {542} & {513} &  {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
-%{{\tt 459.GemsFDTD}} & 6$\times$10$^{-2}$ & {2,143} & {2,132}  & {3$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
-%{{\tt 462.libquantum}} & 4$\times$10$^{-2}$ & {3,326} & {3,274}  & {6$\times$10$^{-3}$} & {6$\times$10$^{-3}$}\\
-%{{\tt 464.h264ref}} &  5$\times$10$^{-2}$ & {601} & {563}  & {9$\times$10$^{-4}$} & {9$\times$10$^{-4}$}\\
-%{{\tt 465.tonto}} & 6$\times$10$^{-2}$ & {883} & {797}  & {1$\times$10$^{-3}$} & {1$\times$10$^{-3}$}\\
-%{{\tt 470.lbm}} & 2$\times$10$^{-2}$ & {143} & {94} & {4$\times$10$^{-4}$} & {3$\times$10$^{-4}$}\\
-%{{\tt 471.omnetpp}} & 3$\times$10$^{-2}$ & {2,114} & {2,072} & {6$\times$10$^{-4}$} & {6$\times$10$^{-4}$}\\
-%{{\tt 473.astar}} & 3$\times$10$^{-2}$ & {359} & {317} &  {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
-%%{{\tt 481.wrf}} & \multicolumn{4}{c}{Run-time Error}\\
-%{{\tt 482.sphinx3}} & 5$\times$10$^{-2}$ & {2,436} & {2,392} &   {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\ %% V
-%%{{\tt 483.xalancbmk}} & \multicolumn{4}{c}{Out-of-Scope}\\
-%{{\tt 998.specrand}} & 9$\times$10$^{-6}$ & {0.6} & {0.6} &  {4$\times$10$^{-3}$} & {4$\times$10$^{-3}$}\\
-%{{\tt 999.specrand}} & 9$\times$10$^{-6}$  & {0.8} & {0.6} &  {6$\times$10$^{-3}$} & {5$\times$10$^{-3}$}\\ \hline 
-%Total.   & \multicolumn{1}{c|}{1} & \multicolumn{4}{c|}{---}\\ \hline
-%Avg.   & --- & 852 & 815 &  {3$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ \hline
-%Wght. Avg. & --- & 43  & 42  &  {6$\times$10$^{-5}$} & {6$\times$10$^{-5}$}\\ \hline
-%\end{tabular}
-%}
-%\caption{Performance Evaluation on the SPEC Benchmarks\label{tab:spec_real}}
-%\vspace{-0.33in}
-%%\vspace{-0.33in}
-%\end{table}
-%\vspace{0.5in}
-
-%\section{Conclusion and Future Work}
-%\label{sec:conclusion}
-%%\vspace{-0.07in}
-%In this paper we \shorten{provided an empirical evidence of 
-%how significantly measurement quality could be hurt by 
-%the presence of infrequent, long-running daemon processes. 
-%We }presented a novel execution-time measurement scheme called {\em SEDONA}\shorten{, 
-%employing process time and identifying an infrequent, long-running daemon}. 
-%%Our evaluation using real benchmarks demonstrated the strong effectiveness of 
-%%the scheme in improving the measurement performance.
-%Our scheme is more precise and accurate than the traditional method\shorten{
-%as shown in the experiments}.
-% In the future we plan to integrate this approach into a novel query timing 
-% protocol~\cite{Currim} to further improve the protocol by 
-% early identifying such a daemon and eliminating samples contaminated by that daemon.
-
+{\color{red}
 \section{Conclusion}
@@
 employing process time and identifying an infrequent, long-running daemon}, 
-which is more precise and \hbox{accurate} than the traditional method relying on elapsed time\shorten{
+more precise and \hbox{accurate} than the traditional method, 
+and performed its rigorous empirical evaluation\shorten{
 as shown in the experiments}.
-%Our plan is to integrate SEDONA into the query timing protocol~\cite{Currim}.
-%Our evaluation using real benchmarks demonstrated the strong effectiveness of 
-%the scheme in improving the measurement performance.
-%Our scheme is more precise and accurate than the traditional method\shorten{
-%as shown in the experiments}.
-% In the future we plan to integrate this approach into a novel query timing 
-% protocol~\cite{Currim} to further improve the protocol by 
-% early identifying such a daemon and eliminating samples contaminated by that daemon.
-
-%\vspace\fill
-
-%\section*{Acknowledgments} 
-%The author give many thanks to Rick Snodgrass, Rob Meier, and John Kececioglu 
-%for their sincere supports.
+}
 

### 2017-02-07 sedona-ieice16/revision/cover_letter.tex
SEDONA final updates

@@
 
-\opening{Dear Prof. Rajkumar Buyya:}
+\opening{Dear Editor-in-Chief:}
 
@@
 
-Attached is a minor revision that responds to the first reviewer's comments. 
-We marked in red some text in response to his/her comments. 
-The revision also includes some texts marked in blue that are added or updated. 
-Note also that we replaced the term of {\em PUT} with another term, called {\em INC}, 
-in order to prevent readers from getting confused with a general concept of the program under test. 
-Many thanks to you for taking sincere care of our manuscript during a series of reviews.
+I am re-submitting a revised manuscript that responds to a previous reviewer's comments. 
+I mark in red some text in response to the reviewer's comments. 
+This revision also includes some text marked in blue that are added or updated, 
+compared to its previous version.
+
+Here are highlights of differences between the revision and the previous version.
+
+\begin{itemize}
+
+\item Research goal: the goal of this manuscript is clearly stated upfront in Section~1.
+
+\item Characteristics: the distinctions of our protocol are discussed 
+in the paragraph of ``Characteristics" in Section~1.
+
+\item Timing environment: the description of our timing environment 
+is moved upfront to Section~1 (from Section~2), 
+to emphasize the fact that even though a timing environment was well configured, 
+there still existed significant variability in the traditional measurement method.
+
+\item Terminology: the definition of ``dual-PUT" is provided with its example in Section~2.
+
+\item Experiments: the results from a new experiment are provided in Fig.~5 on page~4.
+
+\end{itemize}
+
+Many thanks to you for handling our manuscript.
 

### 2017-02-07 sedona-ieice16/revision/paper.tex
SEDONA final updates

@@
 a specification illustrated in Table~\ref{tab:machine_config}, 
-in a controlled environment 
-with (i)~many non-critical daemon processes deactivated, (ii)~the 
-network timing protocol daemon enabled, and (iii)~an up-to-date kernel.}
+in the same timing environment as used in our prior work~\cite{Currim}, 
+with i)~non-critical daemons deactivated, 
+ii)~the Network Timing Protocol daemon switched on,  
+iii)~particular CPU features\cite{intel15,intelSpeed15} turned off, and 
+and iv)~an up-to-date kernel installed.}
 
@@
 in a simulation framework via dynamic sampling of trace snippets during program execution. 
-None of these prior works, however, takes into account the variability in 
-timing and the influence of daemons that may significantly disturb the timing.
+But none of these prior works considers the variability in 
+timing and the impact of daemons that may disturb the timing a lot.
 
@@
 {\bf A Running Example.} 
-As motivated by our prior work~\cite{Currim}, 
-we configure a timing environment by i) stopping non-critical daemons, 
-ii) activating the Network Timing Protocol daemon, and 
-iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1).
+%As motivated by our prior work~\cite{Currim}, 
+%we configure a timing environment by i) stopping non-critical daemons, 
+%ii) activating the Network Timing Protocol daemon, and 
+%iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1). 
+{\color{blue}
+First, we set up the same measurement environment as described earlier in Section~\ref{sec:intro}, 
+to eliminate known timing factors in advance (Step~1).}
 

### 2017-02-07 sedona-ieice16/revision/reviews/review_response.tex
Final updates before Rick

@@
 \section*{Overview}\label{sec:overview}
-I appreciate these quite thorough reviews. I appreciate the feedback and suggestions.
+I appreciate these quite thorough reviews and suggestions.
 
@@
 
-Your concerns are well understood. 
-Indeed, our statistics and heuristics greatly help us 
-identify what daemons are infrequent, long-running on our system 
+I understand your concerns.
+Indeed, the used statistics and heuristics greatly helped me 
+identify what daemons were infrequent, long-running on our system 
 (as described in the fourth paragraph in the left column on page~3). 
 Applying such statistics and heuristics can be done to any Linux system, 
-which make our protocol generic. Furthermore, the experimental results show that 
-our protocol is more efficient than the traditional method relying on elapsed time.
-
-Note that my responses to your comments 
-are incorporated into the revision and marked in {\color{red}red}. 
-There are some new and modified contents in the revision, 
-which are marked in {\color{blue}blue} for better distinction.
+which indeed makes our protocol generic. Furthermore, the experimental results show that 
+the proposed protocol is more efficient than the traditional method relying on elapsed time.
 
 I now respond to individually below.
+Note that my responses to your comments are incorporated into the revision and marked in {\color{red}red}.
+There are also some new or modified contents in the revision, which are marked in {\color{blue}blue}.
 
@@
 
-I understand your confusion. The previously described plan was suggested as one of the possibilities in the future work. But that is not the research goal of this paper. 
-As you already recognized, the goal of this paper is to propose a {\em generic, better} timing protocol against the conventional approach of utilizing elapsed time, 
+I understand your confusion. The previously described plan was suggested as one of the possibilities in the future work. But that was not stated as the ultimate goal of this paper. 
+As you already recognized, my goal is to propose a {\em generic, better} timing protocol against the conventional approach of utilizing elapsed time, 
 to enable better timing results. This goal is successfully achieved 
-by significantly reducing variability in timing as demonstrated in our various experiments.
-%The goal is successfully achieved by significantly reducing variability of 
-%the execution time of a program of interest in our experiments. 
-I now clearly state this goal, marked in red, at the last sentence in Section~1.
+by significantly reducing variability in timing as demonstrated in our various experiments. I now clearly state this goal, marked in red, in the last sentence in Section~1.
 
@@
 with SEDONA, was removed.
-
-%Note also that the mention of my plan in the future work 
-%is just one aspect of the extension of the SEDONA protocol. 
-%I claim that the combination is not the ultimate research goal of this paper. 
-%To avoid confusion, I removed that sentence.
 
@@
 
-Your concern is understood, but it is hard to agree with the comment that the improvement is little, for the following reasons.
+I understand this concern. But I do not agree with the comment that the improvement is little, for the following reasons.
 First, for a certain workload such as {\tt 434}, 
-our protocol reduced the variability was reduced up to by about 10x (from 75 ms to 7 ms), 
-which is *not* small. (Please refer to the last sentence 
+our protocol reduced the variability up to by about 10x 
+(from 75 ms to 7 ms), 
+which cannot be said to be small. (Please refer to the last sentence 
 in the first paragraph in the second column on page~4.)
@@
 (Please refer to the second paragraph in the right column on page~4.)
-The relative error of our protocol was still smaller than that of the ORG method 
-for a long-running benchmark like {\tt 436} and {\tt 454} 
+The relative error of our protocol was lower than that of the traditional method 
+for a long-running benchmark such as {\tt 416}, {\tt 436}, and {\tt 454} 
 as well as for a medium-length or a short-length one. 
-Such scalability was still observed in our new experiment, 
-which I describe shortly, as shown in Fig.~5.
+A new experiment, 
+which I describe shortly, also revealed 
+such scalability of our protocol, as shown in Fig.~5.
 
-Note also that the ET results by the ORG method were also based 
-on the settings taking care of several identified timing factors as
-described in the third paragraph in Section~2.
-Under such a well-configured timing environment, our protocol further 
-reduced the variability of execution time compared to the ORG method
-in such various benchmarks, 
-which actually demonstrated its general purpose and wide usability.
+(Note that the ET results by the conventional measurement were already 
+improved enough, as they were measured based 
+on the settings handling several identified timing factors mentioned in the third paragraph in Section~1.
+Under such a well-configured timing environment, our protocol could further reduce the variability of execution time in such various benchmarks.
+This actually demonstrates its general purpose and wide usability.)
 
-Lastly, the results from a new experiment provide 
-another strongly evidence that our protocol is widely practical and useful. 
+Lastly, the new experimental results provide 
+another strong evidence that our protocol is widely practical and useful. 
 In the new experiment we evaluated the performance of our protocol 
 on some real-world programs like insertion sort and matrix multiplication (in column major). 
-The experimental results showed that the performance gap between the original and our methods 
+The results showed that the performance gap between the original and our timing schemes  
 was increasingly widened, e.g., up to by 6x, as workload level increased.
@@
 Therefore, I strongly claim that these various experiments confirm 
-the wide acceptance and usability of SEDONA as an alternative to the ET-based measurement technique.
+the wide acceptance and usability of SEDONA as an alternative to the ET-based conventional measurement technique.
 
@@
 
-This review is basically the same as (2). Please refer to our response of (2) for this comment. 
-One note. You may feel that the extent of the improvement varies widely, 
-because the SPEC CPU2006 includes a set of very different compute-bound benchmarks. 
-But even if so, we found that all the benchmarks in the suite benefited from our timing protocol, 
-and in addition, for some workloads, our protocol significantly reduced their variability up to by 10x.
+This review is basically the same as (2).
+Please refer to our response of (2) for this comment.
+One note, though.
+You may feel that the extent of the improvement varies widely.
+That is because the SPEC CPU2006 includes a set of very different compute-bound benchmarks.
+We also found that all the different benchmarks in the suite benefited from our timing protocol, and for a certain workload, our protocol significantly reduced their variability up to by 10x as mentioned earlier.
 
@@
 Your point is well taken. 
-The term of ``dual-PUT'' represents 
+More descriptions of the term ``\hbox{dual-PUT}'' are provided with a simple example in the second paragraph in the right column on page~2.  
+In brief, the term of ``dual-PUT'' represents 
 a pair of two consecutive samples of a run of a (regular) PUT. 
-(The task length of that regular PUT is virtually doubled, extended to the corresponding 
-dual-PUT.) More descriptions of the term ``\hbox{dual-PUT}'' are provided with a simple example in the second paragraph in the right column on page~2. 
+(The task length of that regular PUT is virtually doubled, extended to the corresponding dual-PUT.)
+
 In addition, the 1st half indicates the first element of the pair 

### 2017-02-07 sedona-ieice16/revision/paper.tex
Final updates before Rick

@@
 for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-{\color{blue}Even for the long-running benchmarks (e.g., {\tt 436}, {\tt 416}, and {\tt 454}, both$>$ 1,000 sec), 
+{\color{blue}Even for the long-running benchmarks (e.g., {\tt 416}, {\tt 436}, and {\tt 454}, both$>$ 1,000 sec), 
 the relative error of \hbox{SEDONA} was still lower than that of ORG.}

### 2017-02-08 sedona-ieice16/revision/reviews/review_response.tex
Applying Rick's comments

@@
 \section*{Overview}\label{sec:overview}
-I appreciate these quite thorough reviews and suggestions.
+I appreciate this detailed review and included suggestions.
 
@@
 
-I understand your concerns.
-Indeed, the used statistics and heuristics greatly helped me 
+Thank you for your comments. Indeed, the used statistics and heuristics greatly helped me 
 identify what daemons were infrequent, long-running on our system 
@@
 
-I now respond to individually below.
-Note that my responses to your comments are incorporated into the revision and marked in {\color{red}red}.
-There are also some new or modified contents in the revision, which are marked in {\color{blue}blue}.
+I now respond to individually below. Note that my responses to your comments are 
+incorporated into the revision and marked in {\color{blue}blue}.
 
@@
 
-I understand your confusion. The previously described plan was suggested as one of the possibilities in the future work. But that was not stated as the ultimate goal of this paper. 
-As you already recognized, my goal is to propose a {\em generic, better} timing protocol against the conventional approach of utilizing elapsed time, 
+The previously-described plan was suggested as one of the possibilities 
+in the future work. But that was not stated as the ultimate goal of this paper. 
+As you already recognized, my goal is to propose a {\em generic, better} timing protocol against the 
+conventional approach of utilizing elapsed time, 
 to enable better timing results. This goal is successfully achieved 
-by significantly reducing variability in timing as demonstrated in our various experiments. I now clearly state this goal, marked in red, in the last sentence in Section~1.
+by significantly reducing variability in timing as demonstrated in our various experiments. 
+I now clearly state this goal in the last sentence in Section~1.
 
-To avoid confusion, I removed the last mention. Also, the first paragraph on page~2, 
+To avoid confusion, I removed the last mention. Also, the first paragraph of the previous manuscript on page~2, 
 which used to provide an explanation of a possible combination of the query timing protocol 
@@
 
-I understand this concern. But I do not agree with the comment that the improvement is little, for the following reasons.
+Such a doubt may be raised. But I do not agree with the comment that the improvement is little, 
+for the following reasons.
 First, for a certain workload such as {\tt 434}, 
@@
 Furthermore, every benchmark's standard deviation by our protocol 
-was less than that of the conventional method. 
+was less than that of the extant method. 
 The relative error of the benchmarks was on average improved 
@@
 
-(Note that the ET results by the conventional measurement were already 
+(Note that the ET results by the traditional measurement were already 
 improved enough, as they were measured based 
-on the settings handling several identified timing factors mentioned in the third paragraph in Section~1.
-Under such a well-configured timing environment, our protocol could further reduce the variability of execution time in such various benchmarks.
+on the settings handling several identified timing factors mentioned in the third paragraph 
+in Section~1. Under such a well-configured timing environment, our protocol could further 
+reduce the variability of execution time in such various benchmarks. 
 This actually demonstrates its general purpose and wide usability.)
@@
 
-Therefore, I strongly claim that these various experiments confirm 
-the wide acceptance and usability of SEDONA as an alternative to the ET-based conventional measurement technique.
+Therefore, I claim that these various experiments confirm 
+the wide acceptance and usability of SEDONA as an alternative to the 
+ET-based extant measurement technique.
 
@@
 
-This review is basically the same as (2).
+This comment is basically the same as (2).
 Please refer to our response of (2) for this comment.
@@
 
-Your point is well taken. 
-More descriptions of the term ``\hbox{dual-PUT}'' are provided with a simple example in the second paragraph in the right column on page~2.  
+I understand your confusion. More descriptions of the term ``\hbox{dual-PUT}'' are now 
+provided with a simple example in the second paragraph in the right column on page~2.  
 In brief, the term of ``dual-PUT'' represents 

### 2017-02-08 sedona-ieice16/revision/cover_letter.tex
Applying Rick's comments

@@
 \begin{flushright}
-\hfill Email: youngkyoon.suh@gmail.com
+\hfill Email: yksuh@kisti.re.kr
 \end{flushright} 
@@
 
-\opening{Dear Editor-in-Chief:}
+\opening{Dear Dr. Koichi Mouri (Associated Editor):}
 
@@
 
-I am re-submitting a revised manuscript that responds to a previous reviewer's comments. 
-I mark in red some text in response to the reviewer's comments. 
-This revision also includes some text marked in blue that are added or updated, 
-compared to its previous version.
+I am submitting a revised manuscript that responds to a previous reviewer's comments. 
+In this revised manuscript I mark in blue text that was changed or added in response to the reviewer's comments, 
+compared to its previous (rejected) version (paper ID number: 2016EDL8232).
 
@@
 
-Many thanks to you for handling our manuscript.
+Many thanks to you for handling my manuscript.
 

### 2017-02-08 sedona-ieice16/revision/paper.tex
Applying Rick's comments

@@
 \begin{summary}
-{\color{blue}Measuring program execution time 
+{\color{blue}Measuring program execution time is 
 a \hbox{much-used} technique for performance evaluation in computer science. 
-Without a proper care, however, timed results may vary a lot, 
+Without proper care, however, timed results may vary a lot, 
 thus making it hard to trust their validity. 
@@
 %which indeed suggests the need for a better timing protocol.
-{\color{red} The goal of this paper is to propose
+{\color{blue} The goal of this paper is to propose
 a better timing protocol that can significantly reduce such variability 
@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. {\color{red}Note that PUT128 was run on a machine, with 
+800~times. {\color{blue}Note that PUT128 was run on a machine, with 
 a specification illustrated in Table~\ref{tab:machine_config}, 
@@
 
-{\color{red}
+{\color{blue}
 {\bf Characteristics.} There are several distinctions of our protocol. 
@@
 In turn, we evaluate the performance of the scheme using real workloads. 
-Finally, we conclude this letter.}
+\shorten{Finally, we conclude this letter.}
+}
 
@@
 called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons). 
-{\color{blue} Our scheme (1) utilizes PT (process time) to avoid absorbing timing noises from other co-running processes and (2) catches and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
+{\color{blue} Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
 %These two features contribute to significantly enhancing measurement quality.
@@
 Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
-{\color{red}Note that this protocol can be applied to any arbitrary Linux system.}
+{\color{blue}Note that this protocol can be applied to any arbitrary Linux system.}
 
@@
 To identify such daemons, we use a novel \hbox{scatter} plot: 
-those of {\em pairs of successive samples}{\color{red}. 
-So \hbox{samples}~\#{1} and~\#{2} form the first pair, 
-samples~\#{3} and~\#{4} form the second pair.
+those of {\em pairs of successive samples}{\color{blue}. 
+So samples~\#{1} and~\#{2} form the first pair and samples~\#{3} and~\#{4} 
+form the second pair.
 Such a pair is termed ``\hbox{{\it dual-PUT}}.'' 
@@
 from a run of the 800 PUT128 samples (Step~3). 
-{\color{red} The $xy$-plane corresponds to the dual-PUT samples; 
-for example, \hbox{sample}~\#{2} of a run of dual-PUT256 
+{\color{blue} The $xy$-plane corresponds to the dual-PUT samples. 
+For example, \hbox{sample}~\#{2} of a run of dual-PUT256 
 is plotted as a point with samples \#{3} and~\#{4} of a run of PUT128 
@@
     (equivalent to Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
-    \vspace{-0.3in}     
+    \vspace{-0.2in}     
 \end{figure}
@@
 
+{\color{blue}
 For each of the infrequent daemons 
-we use a heuristic to determine the daemon's periodicity: 
+we use a heuristic to determine the daemon's periodicity:}
 the daemon must occur regularly in a sequence of samples.
@@
 When {\color{blue}PUT128} had a ``short'' 
-program time (or, two minutes), daemons with a periodicity of
+program time {\color{blue}(or, two minutes)}, daemons with a periodicity of
 hours are \hbox{infrequent}. But with {\color{blue}PUT16384} 
-with a ``long'' program time (or, 4.5 hours), some of those daemons are now frequent, and appear in the central cluster.
+with a ``long'' program time {\color{blue}(or, 4.5 hours)}, some of those daemons are now frequent, and appear in the central cluster.
 
@@
 the cutoff for each of those infrequent, long-running daemons so
-identified, based on the runs of PUT128 and the 
-PUT16384 as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
+identified, based on {\color{blue}the runs of PUT128 and 
+PUT16384} as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
 Here is how to compute the cutoff. For the cutoff of such a daemon with PUT128, 
@@
 matrix multiplication (in column major), termed MM. 
-In this comparison SEDONA overwhelmed ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
+In this comparison SEDONA bettered ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
 As workload size increased, the performance gap between the two 
-got increasingly widened, reaching up to by 6x.
+were increasingly widened, reaching up to by 6x.
 }
@@
 lower than that of ORG for almost every benchmark.
-{\color{red}About a 10x \hbox{margin} between the two was yielded for {\tt 434}, for example.}
+{\color{blue}About a 10x \hbox{margin} between the two resulted from {\tt 434}, for example.}
 }
@@
 for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-{\color{blue}Even for the long-running benchmarks (e.g., {\tt 416}, {\tt 436}, and {\tt 454}, both$>$ 1,000 sec), 
+{\color{blue}Even for the long-running benchmarks (e.g., {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
 the relative error of \hbox{SEDONA} was still lower than that of ORG.}
@@
 {{\tt 433}} & 480,586 & 473,256  & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-{{\tt 434}} & 16,495  & 16,242  & {\color{red}{75}} & {\color{red}{7}} & {\color{red}{5$\times$10$^{-3}$}} & {\color{red}{5$\times$10$^{-4}$}}\\
+{{\tt 434}} & 16,495  & 16,242  & {\color{blue}{75}} & {\color{blue}{7}} & {\color{blue}{5$\times$10$^{-3}$}} & {\color{blue}{5$\times$10$^{-4}$}}\\
 {{\tt 435}} & 990,575 & 975,445  & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
@@
 
-{\color{red}
+{\color{blue}
 \section{Conclusion}
@@
 \vspace{-0.07in}
-We \shorten{provided an empirical evidence of 
-how significantly measurement quality could be hurt by 
-the presence of infrequent, long-running daemon processes. 
-We }proposed a novel timing protocol called {\em SEDONA}\shorten{, 
-employing process time and identifying an infrequent, long-running daemon}, 
-more precise and \hbox{accurate} than the traditional method, 
-and performed its rigorous empirical evaluation\shorten{
-as shown in the experiments}.
+%We \shorten{provided an empirical evidence of 
+%how significantly measurement quality could be hurt by 
+%the presence of infrequent, long-running daemon processes. 
+%We }proposed a novel timing protocol called {\em SEDONA}\shorten{, 
+%employing process time and identifying an infrequent, long-running daemon}, 
+%more precise and \hbox{accurate} than the traditional method, 
+%and performed its rigorous empirical evaluation\shorten{
+%as shown in the experiments}.
+We proposed a novel timing protocol called {\em SEDONA}, and 
+performed an empirical evaluation to show that 
+it is more precise and accurate than the extant method.
 }

### 2017-02-08 sedona-ieice16/revision/data/fig_script.r

@@ -16,7 +16,8 @@ plot(x$METIME, xlim=c(0,max(x$ITERNUM)), ylim=c(old_xmin, old_xmax),
 	pch=1, main='', 
 	xlab='Iteration', ylab='Elapsed Time (msec)')
 options(scipen = 999)
-legend(380, 171500, c(as.expression(paste("# of total samples = ",nrow(x))), 
+op <- par(cex = 1.5)
+legend(160, 155500, c(as.expression(paste("# of total samples = ",nrow(x))), 
 		    as.expression(paste("standard deviation = ",round(sd(x$METIME),1),"msec")), 
 		    as.expression(paste("relative error = ",round(re,5)))
 ))
@@ -101,7 +102,8 @@ plot(x$PRTIME, xlim=c(0,max(x$ITERNUM)), ylim=c(old_xmin, old_xmax),
 	xlab='Iteration', ylab='Process Time (msec)'
 )
 options(scipen = 999)
-legend(380, 171500, c(as.expression(paste("# of retained samples = ",nrow(x))), 
+op <- par(cex = 1.5)
+legend(200, 155500, c(as.expression(paste("# of retained samples = ",nrow(x))), 
 		    as.expression(paste("standard deviation = ",round(sd(x$PRTIME),1),"msec")), 
 		    as.expression(paste("relative error = ",round(re,5)))
 ))

### 2017-02-08 sedona-ieice16/revision/reviews/review_response.tex
review response modified

@@
 Second, our protocol scaled well over increasing execution time. 
-(Please refer to the second paragraph in the right column on page~4.)
+(Please refer to the bottom paragraph in the left column on page~4.)
 The relative error of our protocol was lower than that of the traditional method 

### 2017-02-08 pt_characterization/new_data/empv5_pt_hist.r
New draft

@@ -223,7 +223,8 @@ dev.off()
 x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
 setEPS()
 postscript("8192_sec_pt_hist1_v5.eps")
-x <- subset(x, x$ITERNUM <= 40) 
+#x <- subset(x, x$ITERNUM <= 40) 
+x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
 x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
 x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
 x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
@@ -234,26 +235,7 @@ nbins <- ceiling((xmax-xmin) / binsize)
 h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
 xmin <-xmin-17
 xmax <-xmax+15
-plot(h, ylim=c(0,10), xaxt='n',freq=TRUE,xlim=c(xmin,xmax),col="green", main='PT frequency on INC8192', 
-sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), xlab='PT (ms)', ylab=expression('Frequency'))
-xaxtl <- seq(xmin,xmax,by=binsize)
-axis(side=1, at=xaxtl, labels=xaxtl)
-dev.off()
-
-x = read.csv(file="8192_sec2.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("8192_sec_pt_hist2.eps")
-x <- subset(x, x$ITERNUM >40 & x$ITERNUM <= 300)
-binsize=20
-nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
-h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
-max(h$counts)
-min(x$PRTIME)
-max(x$PRTIME)
-max(x$PRTIME)-min(x$PRTIME)
-xmin <-min(x$PRTIME)-20
-xmax <-max(x$PRTIME)+24
-plot(h, ylim=c(0,60), xaxt='n',freq=TRUE,xlim=c(xmin,xmax),col="green", main='PT frequency on INC8192 in Nov 2015', 
[…]

### 2017-02-09 sedona-ieice16/revision/reviews/review_response.tex
SEDONA updates fitting 4 pages

@@
 identify what daemons were infrequent, long-running on our system 
-(as described in the fourth paragraph in the left column on page~3). 
+(as described in the third paragraph in the left column on page~3). 
 Applying such statistics and heuristics can be done to any Linux system, 
@@
 I now respond to individually below. Note that my responses to your comments are 
-incorporated into the revision and marked in {\color{blue}blue}.
+incorporated into the revision.
 
@@
 which cannot be said to be small. (Please refer to the last sentence 
-in the first paragraph in the second column on page~4.)
+in the fourth paragraph in Section~3 on page~4.)
 There were also some other workloads such as {\tt 410} and {\tt 445} 
@@
 Second, our protocol scaled well over increasing execution time. 
-(Please refer to the bottom paragraph in the left column on page~4.)
+(Please refer to the fifth paragraph in Section~3 on page~4.)
 The relative error of our protocol was lower than that of the traditional method 

### 2017-02-09 sedona-ieice16/revision/resources/cover_letter.tex
SEDONA updates fitting 4 pages

--- 2017-02-08 sedona-ieice16/revision/cover_letter.tex
@@
 I am submitting a revised manuscript that responds to a previous reviewer's comments. 
-In this revised manuscript I mark in blue text that was changed or added in response to the reviewer's comments, 
-compared to its previous (rejected) version (paper ID number: 2016EDL8232).
-
-Here are highlights of differences between the revision and the previous version.
+Here are highlights of differences between this revision and its previous version (paper ID number: 2016EDL8232).
 

### 2017-02-09 sedona-ieice16/revision/resources/paper_4p.tex
SEDONA updates fitting 4 pages

--- 2017-02-08 sedona-ieice16/revision/paper.tex
@@
 }
-\received{2017}{2}{10}
+\received{2017}{2}{9}
 
@@
 {\color{blue}
-{\bf Characteristics.} There are several distinctions of our protocol. 
+{\bf Characteristics.} There are several important specifics of our protocol. 
 First, we utilize {\em process time} (PT) rather than on elapsed time (ET). 
@@
 \noindent
-The rest of this letter is organized as follows. 
-{\color{blue}We next elaborate on the proposed timing protocol. 
-In turn, we evaluate the performance of the scheme using real workloads. 
+{\color{blue}The rest is organized as follows. 
+We next elaborate on the proposed timing protocol. In turn we evaluate the performance of the scheme using real workloads. 
 \shorten{Finally, we conclude this letter.}
@@
 First, we set up the same measurement environment as described earlier in Section~\ref{sec:intro}, 
-to eliminate known timing factors in advance (Step~1).}
+to eliminate known timing factors (Step~1).}
 
@@
     (equivalent to Dual-PUT256 with 400 samples) in Steps~2---4~\label{fig:put128_plot}} 
-    \vspace{-0.2in}     
+    \vspace{-0.3in}     
 \end{figure}
@@
 the maximum process time and standard deviation 
-of PT (process time, note the switch in emphasis from elapsed time 
-to process time) of the daemon processes 
-(e.g. {\tt cifsd}, {\tt flush-9:0}, {\tt jbd2/md0-8}, 
-{\tt kblockd/0}, {\tt khugepaged}, {\tt md0\_raid1}, and {\tt ntpd})
+of PT \shorten{(process time, note the switch in emphasis from elapsed time 
+to process time)} of the daemon processes
+%(e.g. {\tt cifsd}, {\tt flush-9:0}, {\tt jbd2/md0-8}, 
+%{\tt kblockd/0}, {\tt khugepaged}, {\tt md0\_raid1}, and {\tt ntpd})
+such as {\tt cifsd}, etc. 
 observed in the central cluster samples in Fig.~\ref{fig:dual_et_put256_cluster}. 
@@
 are determined as infrequent, long-running.
-We also identify ``extra'' infrequent daemons: those
-found only in the $L$-samples but not in the central cluster. 
-The running example reveals the following extra daemons: 
-{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
-{\tt rhsmcertd-worke}, and {\tt sshd}.
+%We also identify ``extra'' infrequent daemons: those
+%found only in the $L$-samples but not in the central cluster. 
+%The running example reveals the following extra daemons: 
+%{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
+%{\tt rhsmcertd-worke}, and {\tt sshd}.
+{\color{blue}We also identify ``extra'' infrequent daemons, 
+found only in the $L$-samples but not in the central cluster.}
 
@@
 roughly every 112 samples (which corresponds to very close to every four 
-hours). Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, 
-{\tt md0\_raid1}, and {\tt rhn\_check}) all occur together 
-(in those two outliers in Fig.~\ref{fig:raw_put128}) and 
-  have a periodicity of about every 559 samples
-  (5x longer, or just about 20 hours). 
+hours). 
+%Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, 
+%{\tt md0\_raid1}, and {\tt rhn\_check}) all occur together 
+%(in those two outliers in Fig.~\ref{fig:raw_put128}) and 
+%  have a periodicity of about every 559 samples
+%  (5x longer, or just about 20 hours). 
+{\color{blue}Four others all occur together and have a periodicity of about every 559 samples 
+(5x longer, or just about 20 hours)}. 
 %  It is important only which samples a given daemon shows up 
@@
 clusters of dual-PUT256 and dual-PUT32768{\color{blue}, each consisting of 
-pairs of two successive samples of PUT128 and PUT16384}: 
+pairs of two successive samples of PUT128 and PUT16384}.\shorten{: 
 {\tt flush-9.0}, {\tt jbd2/md0-8}, {\tt kblock/0}, {\tt md0\_raid1}, 
-and {\tt ntpd}. That said, the central cluster also contains other
-  processes not seen in the \hbox{dual-PUT256} central cluster: 
-  {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
-  {\tt rhsmcertd-worke}, and {\tt sshd}. 
+and {\tt ntpd}.} 
+%That said, the central cluster also contains other
+%  processes not seen in the \hbox{dual-PUT256} central cluster: 
+%  {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
+%  {\tt rhsmcertd-worke}, and {\tt sshd}. 
+That said, the central cluster also contains other
+  processes not seen in the \hbox{dual-PUT256} central cluster{\color{blue}, 
+  such as {\tt grep}, {\tt rhn\_check}, etc.}
   But these daemons were categorized in the \hbox{dual-PUT256} analysis as {\em
@@
 
+\vspace{-0.2in}
 \begin{table}[ht]
 \centering
-{\scriptsize
+{\tiny
 \begin{tabular}{|p{1.2cm}|c|c|p{1cm}|p{1.45cm}|} \hline
@@
   Process Time \hbox{(Step~9)}\label{tab:final_infrequent_cutoff}}
-  \vspace{-0.4in}
+  \vspace{-0.3in}
 \end{table}
@@
 \begin{center}
-{\scriptsize
+{\tiny
 \begin{tabular}{|l|p{7cm}|}\hline
@@
 \caption{Machine Configurations\label{tab:machine_config}}
-\vspace{-0.3in}
+\vspace{-0.35in}
 \end{table}
@@
 }
+
+\vspace{-.2in}
+\begin{figure}[h]
+	\centering
+	\subfigure[{\color{blue}SORT - Standard Deviation}]{
+		\includegraphics[scale=0.31]{sort_std.eps}
+		\label{fig:sort_std}
+	}
+%	\subfigure[{\color{blue}SORT - Relative Error}]{
+%		\includegraphics[scale=0.31]{sort_re.eps}
+%		\label{fig:sort_re}
+%	}
+	\subfigure[{\color{blue}MM - Standard Deviation}]{
+		\includegraphics[scale=0.31]{matc_std.eps}
+		\label{fig:matc_std}
+	}
+%	\subfigure[{\color{blue}MM - Relative Error}]{
+%		\includegraphics[scale=0.31]{matc_re.eps}
+%		\label{fig:matc_re}
+%	}
+	\vspace{-.1in}
+	\caption{{\color{blue}Performance Comparison on Real-world Programs~\label{fig:synprog_test}}}
+\vspace{-.2in}
+\end{figure}
 
@@
 }
-
-\begin{figure}[t]
-	\centering
-	\subfigure[{\color{blue}SORT - Standard Deviation}]{
-		\includegraphics[scale=0.31]{sort_std.eps}
-		\label{fig:sort_std}
-	}
-	\subfigure[{\color{blue}SORT - Relative Error}]{
-		\includegraphics[scale=0.31]{sort_re.eps}
-		\label{fig:sort_re}
-	}
-	\subfigure[{\color{blue}MM - Standard Deviation}]{
-		\includegraphics[scale=0.31]{matc_std.eps}
-		\label{fig:matc_std}
-	}
-	\subfigure[{\color{blue}MM - Relative Error}]{
-		\includegraphics[scale=0.31]{matc_re.eps}
-		\label{fig:matc_re}
-	}
-	\vspace{-.1in}
-	\caption{{\color{blue}Performance Comparison on Real-world Programs~\label{fig:synprog_test}}}
-\vspace{-.2in}
-\end{figure}
 
@@
 {
-\scriptsize
-\begin{tabular}{|p{0.18cm}|R{1.14cm}|R{1.14cm}|R{0.64cm}|R{1cm}|R{0.97cm}|R{0.97cm}|} \hline
- 		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative Error}}}\\ \cline{2-3}
- & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SEDONA}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{}\\ \cline{4-7}	
- & \multicolumn{1}{c|}{{\tiny{(in ET)}}} & \multicolumn{1}{c|}{{\tiny{(in PT)}}} & {\tiny{ORG}} & {\tiny{SEDONA}} & \multicolumn{1}{c|}{{\tiny{ORG}}}& \multicolumn{1}{c|}{{\tiny{SEDONA}}}\\ \hline
+\tiny
+\begin{tabular}{|p{0.1cm}|R{1.1cm}|R{1.1cm}|R{0.6cm}|R{0.6cm}|R{0.97cm}|R{0.97cm}|} \hline
+ 		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative}}}\\ \cline{2-3}
+ & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SED.}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{{\tiny Error}}\\ \cline{4-7}	
+ & \multicolumn{1}{c|}{{\tiny{(in ET)}}} & \multicolumn{1}{c|}{{\tiny{(in PT)}}} & {\tiny{ORG}} & {\tiny{SED.}} & \multicolumn{1}{c|}{{\tiny{ORG}}}& \multicolumn{1}{c|}{{\tiny{SED.}}}\\ \hline
 {{\tt 400}} & 454 & 445 & {3} & {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\

### 2017-02-09 sedona-ieice16/revision/resources/paper_blue.tex
SEDONA updates fitting 4 pages

--- 2017-02-09 sedona-ieice16/revision/resources/paper_4p.tex
@@
 Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
-{\color{blue}Note that this protocol can be applied to any arbitrary Linux system.}
+{\color{blue}Note that this protocol (algorithm) is applicable to any arbitrary Linux system.}
 
@@
 the maximum process time and standard deviation 
-of PT \shorten{(process time, note the switch in emphasis from elapsed time 
+of PT\shorten{(process time, note the switch in emphasis from elapsed time 
 to process time)} of the daemon processes
@@
 %{\tt kblockd/0}, {\tt khugepaged}, {\tt md0\_raid1}, and {\tt ntpd}) 
-such as {\tt cifsd}, etc. 
+(i.e. {\tt flush-9:0}) 
 observed in the central cluster samples in Fig.~\ref{fig:dual_et_put256_cluster}. 
@@
 %{\tt rhsmcertd-worke}, and {\tt sshd}.
-{\color{blue}We also identify ``extra'' infrequent daemons, 
-found only in the $L$-samples but not in the central cluster.}
+{\color{blue}We also identify ``extra'' infrequent daemons: 
+{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
+{\tt rhsmcertd-worke}, and {\tt sshd}, those 
+found only in the $L$-samples but not in the central cluster. 
+}
 
@@
 For each of the infrequent daemons 
-we use a heuristic to determine the daemon's periodicity:}
+we use a heuristic to determine the daemon's periodicity:
 the daemon must occur regularly in a sequence of samples.
-For instance, the {\tt rhn\_check} daemon appears 
-roughly every 112 samples (which corresponds to very close to every four 
-hours). 
+For instance, {\tt rhn\_check} appears 
+roughly every 112 samples 
+%(which corresponds to very close to every four hours). 
+(or almost every four hours). 
 %Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, 
@@
 %  (5x longer, or just about 20 hours). 
-{\color{blue}Four others all occur together and have a periodicity of about every 559 samples 
-(5x longer, or just about 20 hours)}. 
+Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, {\tt md0\_raid1}, and {\tt rhn\_check}) 
+all occur together and have a periodicity of about every 559 samples 
+(5x longer, or just about 20 hours). 
 %  It is important only which samples a given daemon shows up 
 %  in; we don't care in this heuristic whether daemons are co-occurring.
+}
 
@@
 We find some frequent daemon processes appearing in both of the 
-clusters of dual-PUT256 and dual-PUT32768{\color{blue}, each consisting of 
+clusters of \hbox{dual-PUT256} and \hbox{dual-PUT32768}{\color{blue}, each consisting of 
 pairs of two successive samples of PUT128 and PUT16384}.\shorten{: 
@@
 and {\tt ntpd}.} 
-%That said, the central cluster also contains other
-%  processes not seen in the \hbox{dual-PUT256} central cluster: 
-%  {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
-%  {\tt rhsmcertd-worke}, and {\tt sshd}. 
 That said, the central cluster also contains other
-  processes not seen in the \hbox{dual-PUT256} central cluster{\color{blue}, 
-  such as {\tt grep}, {\tt rhn\_check}, etc.}
+  processes not seen in the \hbox{dual-PUT256} central cluster: 
+  {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
+  {\tt rhsmcertd-worke}, and {\tt sshd}. 
   But these daemons were categorized in the \hbox{dual-PUT256} analysis as {\em
@@
 For the short benchmarks 
-(e.g., {\tt 400}, {\tt 403}, {\tt 410}, 
+(e.g. {\tt 400}, {\tt 403}, {\tt 410}, 
 {\tt 434}, {\tt 445}, and {\tt 999}: those taking \hbox{under} 100 sec), 
@@
 The SEDONA protocol continued its dominance against the conventional technique 
-for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-{\color{blue}Even for the long-running benchmarks (e.g., {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
+for the medium-length benchmarks (e.g. {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
+{\color{blue}Even for the long-running benchmarks (e.g. {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
 the relative error of \hbox{SEDONA} was still lower than that of ORG.}
@@
 \begin{tabular}{|p{0.1cm}|R{1.1cm}|R{1.1cm}|R{0.6cm}|R{0.6cm}|R{0.97cm}|R{0.97cm}|} \hline
- 		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative}}}\\ \cline{2-3}
- & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SED.}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{{\tiny Error}}\\ \cline{4-7}	
+ 		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative Error}}}\\ \cline{2-3}
+ & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SED.}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{{\tiny}}\\ \cline{4-7}	
  & \multicolumn{1}{c|}{{\tiny{(in ET)}}} & \multicolumn{1}{c|}{{\tiny{(in PT)}}} & {\tiny{ORG}} & {\tiny{SED.}} & \multicolumn{1}{c|}{{\tiny{ORG}}}& \multicolumn{1}{c|}{{\tiny{SED.}}}\\ \hline

### 2017-02-09 sedona-ieice16/revision/paper.tex
Final Submission

@@
 \begin{summary}
-{\color{blue}Measuring program execution time is 
+Measuring program execution time is 
 a \hbox{much-used} technique for performance evaluation in computer science. 
@@
 In this paper we propose a novel timing protocol to significantly 
-reduce such variability.}
+reduce such variability.
 
@@
 %which indeed suggests the need for a better timing protocol.
-{\color{blue} The goal of this paper is to propose
+The goal of this paper is to propose
 a better timing protocol that can significantly reduce such variability 
-and thus enable better timing results without a doubt of their validity.}
+and thus enable better timing results without a doubt of their validity.
 
@@
 This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
-{\color{blue}The $tl$ value is used to compute 
+The $tl$ value is used to compute 
 the number of iterations ($t$) for which that for-loop is performed 
-to reach the specified task length.}
+to reach the specified task length.
 
@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. {\color{blue}Note that PUT128 was run on a machine, with 
+800~times. Note that PUT128 was run on a machine, with 
 a specification illustrated in Table~\ref{tab:machine_config}, 
 in the same timing environment as used in our prior work~\cite{Currim}, 
-with i)~non-critical daemons deactivated, 
-ii)~the Network Timing Protocol daemon switched on,  
-iii)~particular CPU features\cite{intel15,intelSpeed15} turned off, and 
-and iv)~an up-to-date kernel installed.}
+with (i)~non-critical daemons deactivated, 
+(ii)~the Network Timing Protocol daemon switched on,  
+(iii)~particular CPU features~\cite{intel15,intelSpeed15} turned off, 
+and (iv)~an up-to-date kernel installed.
 
-{\color{blue}Fig.~\ref{fig:meas_comp} shows two plots of the timing results on PUT128.
+Fig.~\ref{fig:meas_comp} shows two plots of the timing results on PUT128.
 Most of the 800 executions measured in elapsed time, 
@@
 Indeed, these outliers {\em critically} impact the overall measurement quality: a standard deviation of 1,730 ms and a relative error of 0.01.
-This result suggests adequately eliminating these outliers while retaining most of the executions.}
+This result suggests adequately eliminating these outliers while retaining most of the executions.
 
-{\color{blue}
 Our protocol, which we characterize shortly, on the contrary, 
@@
 three orders of magnitude for the same PUT128 results.
-}
 
-{\color{blue}
-{\bf Characteristics.} There are several distinctions of our protocol. 
+{\bf Characteristics.} There are several important specifics of our protocol. 
 First, we utilize {\em process time} (PT) rather than on elapsed time (ET). 
@@
 the overall measurement quality, which would otherwise be biased by them.
-}
 
@@
 	\centering
-	\subfigure[{\color{blue}Initial Elapsed Time Measurement}]{
+	\subfigure[Initial Elapsed Time Measurement]{
 		\includegraphics[width=0.224\textwidth]{128_sec_et_old.eps}
@@
 	}
-	\subfigure[{\color{blue}Program Time After Our Protocol}]{
+	\subfigure[Program Time After Our Protocol]{
 		\includegraphics[width=0.224\textwidth]{128_sec_pt_new.eps}
@@
 	}
-	\caption{{\color{blue}An Example of Measurement Results on a \linebreak \hbox{Compute-bound}} 
+	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
 	Program-Under-Test~\label{fig:meas_comp}}
@@
 
-\item {\color{blue}We evaluate the performance of the protocol with rigorous experiments 
-using real-world programs.}
+\item We evaluate the performance of the protocol with rigorous experiments 
+using real-world programs.
 
@@
 \noindent
-The rest of this letter is organized as follows. 
-{\color{blue}We next elaborate on the proposed timing protocol. 
-In turn, we evaluate the performance of the scheme using real workloads. 
+The rest is organized as follows. 
+We next elaborate on the proposed timing protocol. In turn we evaluate the performance of the scheme using real workloads. 
 \shorten{Finally, we conclude this letter.}
-}
 
@@
 called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons). 
-{\color{blue} Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
+Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
 %These two features contribute to significantly enhancing measurement quality.
-}
 
 Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}. 
-{\color{blue}Note that this protocol can be applied to any arbitrary Linux system.}
+Note that this protocol (algorithm) is applicable to any arbitrary Linux system.
 
@@
 {\bf A Running Example.} 
-%As motivated by our prior work~\cite{Currim}, 
-%we configure a timing environment by i) stopping non-critical daemons, 
-%ii) activating the Network Timing Protocol daemon, and 
-%iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1). 
-{\color{blue}
-First, we set up the same measurement environment as described earlier in Section~\ref{sec:intro}, 
-to eliminate known timing factors in advance (Step~1).}
+First, we set up the same measurement environment as described earlier in \hbox{Section}~\ref{sec:intro}, 
+to eliminate known timing factors (Step~1).
 
@@
 
-Fig.~\ref{fig:reg_put128} plots all the 800 elapsed times (ETs) of the run of PUT128.
-%We see three rows in the plot: a solid row of many samples, perhaps
-%six or more samples that are just above that solid row, and two
-%samples that are way above the solid row. 
+Fig.~\ref{fig:reg_put128} (= Fig.~\ref{fig:hist_before_et}) 
+plots all the 800 elapsed times (ETs) of PUT128.
 The plot clearly shows three rows; that is, 
@@
 the rest of the samples clustered in the bottom row. 
-We'll now drill down into these outliers to 
+We will now drill down into these outliers to 
 show how to reliably eliminate the indirect influence of 
-some ``infrequent, long-running daemons'' on PT (\hbox{process} time) of the PUT.
+some ``infrequent, \hbox{long-running} daemons'' on PT (\hbox{process} time) of the PUT.
 
 To identify such daemons, we use a novel \hbox{scatter} plot: 
-those of {\em pairs of successive samples}{\color{blue}. 
+those of {\em pairs of successive samples}. 
 So samples~\#{1} and~\#{2} form the first pair and samples~\#{3} and~\#{4} 
@@
 sample~\#{1} of dual-PUT256 is equivalent 
-to samples~\#{1} and \#{2} of PUT128.}
+to samples~\#{1} and~\#{2} of PUT128.
 
@@
 from a run of the 800 PUT128 samples (Step~3). 
-{\color{blue} The $xy$-plane corresponds to the dual-PUT samples. 
+The $xy$-plane corresponds to the dual-PUT samples. 
 For example, \hbox{sample}~\#{2} of a run of dual-PUT256 
-is plotted as a point with samples \#{3} and~\#{4} of a run of PUT128 
+is plotted as a point with samples~\#{3} and~\#{4} of a run of PUT128 
 on the respective $x$ and $y$ axes. In that sense 
-the $x$ ($y$) axis is named as ``Odd (Even) Sample's ET.''}
+the $x$ ($y$) axis is named as ``Odd (Even) Sample's ET.''
 There are two quite obvious outliers with ETs 
@@
 	\subfigure[PUT128 with 800 Samples]{
-		%\includegraphics[width=0.223\textwidth]{put128_empv4.eps}
 		\includegraphics[width=0.224\textwidth]{put128_empv4_new.eps}
@@
 	\subfigure[Dual-PUT256 Samples]{
-		%\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4.eps}
 		\includegraphics[width=0.224\textwidth]{dual_et_put256_empv4.eps}
@@
 the maximum process time and standard deviation 
-of PT (process time, note the switch in emphasis from elapsed time 
-to process time) of the daemon processes 
-(e.g. {\tt cifsd}, {\tt flush-9:0}, {\tt jbd2/md0-8}, 
-{\tt kblockd/0}, {\tt khugepaged}, {\tt md0\_raid1}, and {\tt ntpd})
+of PT\shorten{(process time, note the switch in emphasis from elapsed time 
+to process time)} of the daemon processes
+(i.e. {\tt flush-9:0}) 
 observed in the central cluster samples in Fig.~\ref{fig:dual_et_put256_cluster}.
-\shorten{In the running example the maximum PT of those daemons was under 4 msec, 
-and their standard deviation was just about 1 msec.}
 
@@
 {\tt jbd2/md0-8}, and {\tt md0\_raid1} 
-are determined as infrequent, long-running.
-We also identify ``extra'' infrequent daemons: those
+are determined as infrequent, \hbox{long-running}.
+We also identify ``extra'' infrequent daemons: 
+{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
+{\tt rhsmcertd-worke}, and {\tt sshd}, those 
 found only in the $L$-samples but not in the central cluster.
-The running example reveals the following extra daemons: 
-{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
-{\tt rhsmcertd-worke}, and {\tt sshd}.
 
-{\color{blue}
 For each of the infrequent daemons 
-we use a heuristic to determine the daemon's periodicity:}
+we use a heuristic to determine the daemon's periodicity:
 the daemon must occur regularly in a sequence of samples.
-For instance, the {\tt rhn\_check} daemon appears 
-roughly every 112 samples (which corresponds to very close to every four 
-hours). Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, 
-{\tt md0\_raid1}, and {\tt rhn\_check}) all occur together 
-(in those two outliers in Fig.~\ref{fig:raw_put128}) and 
-  have a periodicity of about every 559 samples
+For instance, {\tt rhn\_check} appears 
+roughly every 112 samples (or almost every four hours). 
+Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, {\tt md0\_raid1}, and {\tt rhn\_check}) 
+all occur together and have a periodicity of about every 559 samples 
 (5x longer, or just about 20 hours).
-%  It is important only which samples a given daemon shows up 
-%  in; we don't care in this heuristic whether daemons are co-occurring.
 
@@
 This computation provides a rough, initial distinction of a ``long-running''
-%daemon, namely, the valley between the maximum PT from the central cluster 
 daemon, or the valley between the maximum PT from the central cluster 
@@
 For those daemons (i.e. {\tt grep}) never appearing in the central cluster,
-this initial analysis concludes only that they are infrequent.
+this initial \hbox{analysis} concludes only that they are infrequent.
 
@@
 We find some frequent daemon processes appearing in both of the 
-clusters of dual-PUT256 and dual-PUT32768{\color{blue}, each consisting of 
-pairs of two successive samples of PUT128 and PUT16384}: 
-{\tt flush-9.0}, {\tt jbd2/md0-8}, {\tt kblock/0}, {\tt md0\_raid1}, 
-and {\tt ntpd}. That said, the central cluster also contains other
+clusters of \hbox{dual-PUT256} and \hbox{dual-PUT32768}, each consisting of 
+pairs of two successive samples of PUT128 and PUT16384. 
+That said, the central cluster also contains other
   processes not seen in the \hbox{dual-PUT256} central cluster: 
@@
   infrequent}, several having \hbox{periodicities} estimated at four or twenty hours.
-When {\color{blue}PUT128} had a ``short'' 
-program time {\color{blue}(or, two minutes)}, daemons with a periodicity of
-hours are \hbox{infrequent}. But with {\color{blue}PUT16384} 
-with a ``long'' program time {\color{blue}(or, 4.5 hours)}, some of those daemons are now frequent, and appear in the central cluster.
+When PUT128 had a ``short'' 
+program time (or, two minutes), daemons with a periodicity of
+hours are \hbox{infrequent}. But with PUT16384 
+with a ``long'' program time (or, 4.5 hours), 
+some of those daemons are now frequent, and appear in the central cluster.
 
@@
 the cutoff for each of those infrequent, long-running daemons so
-identified, based on {\color{blue}the runs of PUT128 and 
-PUT16384} as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
+identified, based on the runs of PUT128 and 
+PUT16384 as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
 Here is how to compute the cutoff. For the cutoff of such a daemon with PUT128, 
@@
 {\tiny Name} & {\tiny on PUT128}  & {\tiny on PUT16K} & {\tiny Time} & {\tiny {\bf Cutoff PT}} \\\hline
-%{\tiny Process}  & {\tiny Cutoff PT} & {\tiny Cutoff PT} & {\tiny Task}  & {\tiny{\bf Final }} \\
-% Name & on PUT128  & on PUT16K & Time & {\bf Cutoff PT} \\\hline
 {\tt bash} & 1 msec & --- & --- & {\bf 1 msec} \\ \hline
@@
 and by about two \hbox{orders} of magnitude for PUT16384.
-%(5661.7 vs. 229.8) 0.0003448491 vs. 8.4$\times$10$^{-6}$
-%(40.4 vs. 26.7) 2.463122e-06 vs. 1.626005e-06
-
-%\vspace\fill
 
@@
 We now evaluate the \hbox{performance} of SEDONA 
-compared to that of the conventional timing scheme, termed ORG, based on 
+compared to that of the original timing scheme, termed ORG, based on 
 elapsed time.
@@
 \begin{center}
-{\scriptsize
+{\tiny
 \begin{tabular}{|l|p{7cm}|}\hline
 OS & Red Hat Ent. Linux (RHEL) 6.4 with a kernel of 2.6.32 \\ \hline
-CPU & Intel Core i7-870 Lynnfield 2.93GHz quad-core \hbox{processor}\shorten{ on a LGA 1156 95W motherboard}\\ \hline
+CPU & Intel Core i7-870 Lynnfield 2.93GHz quad-core \hbox{processor}\\ \hline
 RAM & 4GB of DDR3 1333 dual-channel memory\\ \hline
@@
 \caption{Machine Configurations\label{tab:machine_config}}
-\vspace{-0.3in}
+\vspace{-0.35in}
 \end{table}
 
-{\color{blue}
 First, we compared the performance of the two approaches 
@@
 matrix multiplication (in column major), termed MM. 
-In this comparison SEDONA bettered ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
+In this comparison SEDONA outperformed ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
 As workload size increased, the performance gap between the two 
 were increasingly widened, reaching up to by 6x.
+
+\vspace{-.2in}
+\begin{figure}[h]
+	\centering
+	\subfigure[SORT - Standard Deviation]{
+		\includegraphics[scale=0.31]{sort_std.eps}
+		\label{fig:sort_std}
 	}
+	\subfigure[MM - Standard Deviation]{
+		\includegraphics[scale=0.31]{matc_std.eps}
+		\label{fig:matc_std}
+	}
+	\vspace{-.1in}
+	\caption{Performance Comparison on Real-world Programs~\label{fig:synprog_test}}
+\vspace{-.25in}
+\end{figure}
 
@@
 
-{\color{blue}
 Table~\ref{tab:spec_real} shows that 
-SEDONA outperformed ORG,
+SEDONA bettered ORG,
 on the standard deviation and relative error across the very different SPEC benchmarks. 
@@
 lower than that of ORG for almost every benchmark.
-{\color{blue}About a 10x \hbox{margin} between the two resulted from {\tt 434}, for example.}
-}
+For example, about a 10x \hbox{margin} between the two resulted from {\tt 434}.
 
-\begin{figure}[t]
-	\centering
-	\subfigure[{\color{blue}SORT - Standard Deviation}]{
-		\includegraphics[scale=0.31]{sort_std.eps}
-		\label{fig:sort_std}
-	}
-	\subfigure[{\color{blue}SORT - Relative Error}]{
-		\includegraphics[scale=0.31]{sort_re.eps}
-		\label{fig:sort_re}
-	}
-	\subfigure[{\color{blue}MM - Standard Deviation}]{
-		\includegraphics[scale=0.31]{matc_std.eps}
-		\label{fig:matc_std}
-	}
-	\subfigure[{\color{blue}MM - Relative Error}]{
-		\includegraphics[scale=0.31]{matc_re.eps}
-		\label{fig:matc_re}
-	}
-	\vspace{-.1in}
-	\caption{{\color{blue}Performance Comparison on Real-world Programs~\label{fig:synprog_test}}}
-\vspace{-.2in}
-\end{figure}
-
-%\vspace{-.15in}
 SEDONA also scaled well for the SPEC workloads, 
@@
 The SEDONA protocol continued its dominance against the conventional technique 
-for the medium-length benchmarks (e.g., {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-{\color{blue}Even for the long-running benchmarks (e.g., {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
-the relative error of \hbox{SEDONA} was still lower than that of ORG.}
+for the medium-length benchmarks (e.g. {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
+Even for the long-running benchmarks (e.g. {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
+the relative error of \hbox{SEDONA} was still lower than that of ORG.
 
 To summarize, our proposed SEDONA protocol can achieve {\em better} accuracy,
-precision, and scalability in measuring the execution time of real \hbox{compute-bound} benchmarks than the conventional timing method. 
-{\color{blue}The experimental results demonstrate 
-the general applicability of SEDONA to timing any CPU-bound program.}
+precision, and scalability in measuring the execution time of real \hbox{compute-bound} programs 
+than the ET-based existing method. 
+The experimental results demonstrate 
+the general applicability of SEDONA to timing any CPU-bound program.
 
-\vspace{-0.1in}
+%\vspace{-0.in}
 \begin{table}[t]
@@
 {
-\scriptsize
-\begin{tabular}{|p{0.18cm}|R{1.14cm}|R{1.14cm}|R{0.64cm}|R{1cm}|R{0.97cm}|R{0.97cm}|} \hline
+\tiny
+\begin{tabular}{|p{0.1cm}|R{1.1cm}|R{1.1cm}|R{0.6cm}|R{0.6cm}|R{0.92cm}|R{0.92cm}|} \hline
  		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative Error}}}\\ \cline{2-3}
- & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SEDONA}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{}\\ \cline{4-7}	
- & \multicolumn{1}{c|}{{\tiny{(in ET)}}} & \multicolumn{1}{c|}{{\tiny{(in PT)}}} & {\tiny{ORG}} & {\tiny{SEDONA}} & \multicolumn{1}{c|}{{\tiny{ORG}}}& \multicolumn{1}{c|}{{\tiny{SEDONA}}}\\ \hline
+ & \multicolumn{1}{c|}{{\tiny{ORG}}}  & \multicolumn{1}{c|}{{\tiny{SED.}}} & \multicolumn{2}{c|}{{\tiny Deviation (ms)}}  & \multicolumn{2}{c|}{{\tiny}}\\ \cline{4-7}	
+ & \multicolumn{1}{c|}{{\tiny{(in ET)}}} & \multicolumn{1}{c|}{{\tiny{(in PT)}}} & {\tiny{ORG}} & {\tiny{SED.}} & \multicolumn{1}{c|}{{\tiny{ORG}}}& \multicolumn{1}{c|}{{\tiny{SED.}}}\\ \hline
 {{\tt 400}} & 454 & 445 & {3} & {2} & {6$\times$10$^{-3}$} & {3$\times$10$^{-3}$}\\
@@
 {{\tt 433}} & 480,586 & 473,256  & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-{{\tt 434}} & 16,495  & 16,242  & {\color{blue}{75}} & {\color{blue}{7}} & {\color{blue}{5$\times$10$^{-3}$}} & {\color{blue}{5$\times$10$^{-4}$}}\\
+{{\tt 434}} & 16,495  & 16,242  & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
 {{\tt 435}} & 990,575 & 975,445  & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
@@
 
-{\color{blue}
 \section{Conclusion}
@@
 \vspace{-0.07in}
-%We \shorten{provided an empirical evidence of 
-%how significantly measurement quality could be hurt by 
-%the presence of infrequent, long-running daemon processes. 
-%We }proposed a novel timing protocol called {\em SEDONA}\shorten{, 
-%employing process time and identifying an infrequent, long-running daemon}, 
-%more precise and \hbox{accurate} than the traditional method, 
-%and performed its rigorous empirical evaluation\shorten{
-%as shown in the experiments}.
 We proposed a novel timing protocol called {\em SEDONA}, and 
@@
 it is more precise and accurate than the extant method.
-}
 

### 2017-02-10 sedona-ieice16/revision/reviews/review_response.tex
Final Submission
Modified Revision of SEDONA

@@
 I now respond to individually below. Note that my responses to your comments are 
-incorporated into the revision.
+incorporated into the revision and marked in {\color{blue}blue}.
 
@@
 whereas the 2nd half the second element of the same pair. 
-The label of ``1st Half's ET (ms)'' is renamed as ``Odd Samples' ET(ms),'' 
+The label of ``1st (2nd) Half's ET (ms)'' is renamed as ``Odd (Even) Samples' ET(ms),'' 
 for better clarification. Please refer to the subsequent (third) paragraph.

### 2017-02-10 sedona-ieice16/revision/paper.tex
Modified Revision of SEDONA

@@
 }
-\received{2017}{2}{9}
+\received{2017}{2}{10}
 
@@
 \begin{summary}
+{\color{blue}
 Measuring program execution time is 
@@
 reduce such variability.
+}
 
@@
 Surprisingly, there is considerable variability in the measured time.
-%which indeed suggests the need for a better timing protocol.
+{\color{blue}
 The goal of this paper is to propose
@@
 and thus enable better timing results without a doubt of their validity.
+}
 
@@
 This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
+{\color{blue}
 The $tl$ value is used to compute 
 the number of iterations ($t$) for which that for-loop is performed 
-to reach the specified task length.
+to reach the specified task length.}
 
@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. Note that PUT128 was run on a machine, with 
+800~times. 
+{\color{blue}
+Note that PUT128 was run on a machine, with 
 a specification illustrated in Table~\ref{tab:machine_config}, 
@@
 the overall measurement quality, which would otherwise be biased by them.
+}
 
@@
 	\centering
-	\subfigure[Initial Elapsed Time Measurement]{
+	\subfigure[{\color{blue}Initial Elapsed Time Measurement}]{
 		\includegraphics[width=0.224\textwidth]{128_sec_et_old.eps}
@@
 	}
-	\subfigure[Program Time After Our Protocol]{
+	\subfigure[{\color{blue}Program Time After Our Protocol}]{
 		\includegraphics[width=0.224\textwidth]{128_sec_pt_new.eps}
@@
 	}
-	\caption{An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
-	Program-Under-Test~\label{fig:meas_comp}}
+	\caption{{\color{blue}An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
+	Program-Under-Test}~\label{fig:meas_comp}}
 	\vspace{-.18in}
@@
 
+{\color{blue}
 \item We evaluate the performance of the protocol with rigorous experiments 
 using real-world programs.
+}
 
@@
 \noindent
+{\color{blue}
 The rest is organized as follows. 
 We next elaborate on the proposed timing protocol. In turn we evaluate the performance of the scheme using real workloads. 
-\shorten{Finally, we conclude this letter.}
-
-%\vspace\fill
+}
 
@@
 called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons). 
+{\color{blue}
 Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
-%These two features contribute to significantly enhancing measurement quality.
+}
 
 Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}.
-Note that this protocol (algorithm) is applicable to any arbitrary Linux system.
+{\color{blue} 
+Note that this protocol (algorithm) is applicable to any arbitrary Linux system.}
 
@@
 {\bf A Running Example.} 
+{\color{blue}
 First, we set up the same measurement environment as described earlier in \hbox{Section}~\ref{sec:intro}, 
 to eliminate known timing factors (Step~1).
+}
 
@@
 
-Fig.~\ref{fig:reg_put128} (= Fig.~\ref{fig:hist_before_et}) 
-plots all the 800 elapsed times (ETs) of PUT128.
+%Fig.~\ref{fig:reg_put128} (= Fig.~\ref{fig:hist_before_et}) 
+%plots all the 800 elapsed times (ETs) of PUT128.
+%The plot clearly shows three rows; that is, 
+%the top and middle rows represent over a dozen of outliers far from 
+%the rest of the samples clustered in the bottom row. 
+%We will now drill down into these outliers to 
+%show how to reliably eliminate the indirect influence of 
+%some ``infrequent, \hbox{long-running} daemons'' on PT (\hbox{process} time) of the PUT.
+Fig.~\ref{fig:reg_put128} plots all the 800 elapsed times (ETs) of the run of PUT128.
 The plot clearly shows three rows; that is, 
@@
 
+{\color{blue}
 To identify such daemons, we use a novel \hbox{scatter} plot: 
@@
 to samples~\#{1} and~\#{2} of PUT128.
+}
 
@@
 from a run of the 800 PUT128 samples (Step~3). 
+{\color{blue}
 The $xy$-plane corresponds to the dual-PUT samples. 
@@
 the $x$ ($y$) axis is named as ``Odd (Even) Sample's ET.''
+}
 There are two quite obvious outliers with ETs 
@@
 		\label{fig:raw_put128}
-	}
-	%\vspace{-0.1in}    
+	}\vspace{-0.15in}
 	\subfigure[Zooming in on the $L$-shape]{
@@
 ``$L$-shape,'' and attribute it to the
-presence of infrequent long-running daemons.\shorten{Again, samples containing such
-daemons (termed the $L$-samples) will appear
-along the left y-axis or along the bottom x-axis, but will not occur in the
-top right of the scatter plot of the dual-PUT.}
+presence of infrequent long-running daemons.
 
@@
 is no $L$-shape, and thus no $L$-samples, and thus no obvious infrequent
-long-running daemons.\shorten{(We emphasize that these samples may have lots of
-frequent daemons, as well as infrequent, {\em short-running} daemons.) There
-were 384 dual-PUT256 samples in this central cluster, implying that 16 of
-the PUT128 samples were $L$-samples and the 785 remaining of the original 800 PUT128
-samples had no infrequent, long-running daemons, which motivates that term.}
+long-running daemons.
 
@@
 the maximum process time and standard deviation 
-of PT\shorten{(process time, note the switch in emphasis from elapsed time 
-to process time)} of the daemon processes
+of PT (process time, note the switch in emphasis from ET to PT) of the daemon processes
 (i.e. {\tt flush-9:0}) 
@@
 are determined as infrequent, \hbox{long-running}.
+{\color{blue}
 We also identify ``extra'' infrequent daemons: 
@@
 (5x longer, or just about 20 hours). 
+}
 
@@
 We find some frequent daemon processes appearing in both of the 
-clusters of \hbox{dual-PUT256} and \hbox{dual-PUT32768}, each consisting of 
-pairs of two successive samples of PUT128 and PUT16384. 
+clusters of \hbox{dual-PUT256} and \hbox{dual-PUT32768}{\color{blue}, each consisting of 
+pairs of two successive samples of PUT128 and PUT16384.}
 That said, the central cluster also contains other
@@
   infrequent}, several having \hbox{periodicities} estimated at four or twenty hours.
+{\color{blue}
 When PUT128 had a ``short'' 
@@
 hours are \hbox{infrequent}. But with PUT16384 
-with a ``long'' program time (or, 4.5 hours), 
+with a ``long'' program time (or, 4.5 hours),} 
 some of those daemons are now frequent, and appear in the central cluster.
@@
 the cutoff for each of those infrequent, long-running daemons so
-identified, based on the runs of PUT128 and 
-PUT16384 as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
+identified, based on {\color{blue}the runs of PUT128 and 
+PUT16384} as collected in Tab.~\ref{tab:final_infrequent_cutoff}. 
 Here is how to compute the cutoff. For the cutoff of such a daemon with PUT128, 
@@
 
+{\color{blue}
 First, we compared the performance of the two approaches 
@@
 were increasingly widened, reaching up to by 6x.
+}
 
@@
 	\centering
-	\subfigure[SORT - Standard Deviation]{
+	\subfigure[{\color{blue}SORT - Standard Deviation}]{
 		\includegraphics[scale=0.31]{sort_std.eps}
@@
 	}
-	\subfigure[MM - Standard Deviation]{
+	\subfigure[{\color{blue}MM - Standard Deviation}]{
 		\includegraphics[scale=0.31]{matc_std.eps}
@@
 
+{\color{blue}
 Table~\ref{tab:spec_real} shows that 
@@
 For example, about a 10x \hbox{margin} between the two resulted from {\tt 434}.
+}
 
@@
 for the medium-length benchmarks (e.g. {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
+{\color{blue}
 Even for the long-running benchmarks (e.g. {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
 the relative error of \hbox{SEDONA} was still lower than that of ORG.
+}
 
@@
 than the ET-based existing method. 
+{\color{blue}
 The experimental results demonstrate 
-the general applicability of SEDONA to timing any CPU-bound program.
+the general applicability of SEDONA to timing any CPU-bound program.}
 
@@
 {{\tt 433}} & 480,586 & 473,256  & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-{{\tt 434}} & 16,495  & 16,242  & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
+{{\tt 434}} & 16,495  & 16,242  & {\color{blue}{75}} & {\color{blue}{7}} & {{\color{blue}5$\times$10$^{-3}$}} & {{\color{blue}5$\times$10$^{-4}$}}\\
 {{\tt 435}} & 990,575 & 975,445  & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
@@
 
+{\color{blue}
 \section{Conclusion}
@@
 performed an empirical evaluation to show that 
-it is more precise and accurate than the extant method.
+it is more precise and accurate than the extant method.}
 

### 2017-02-10 sedona-ieice16/revision/resources/paper_black.tex
Modified Revision of SEDONA

--- 2017-02-09 sedona-ieice16/revision/paper.tex
@@
 Surprisingly, there is considerable variability in the measured time.
-%which indeed suggests the need for a better timing protocol.
 The goal of this paper is to propose
@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. Note that PUT128 was run on a machine, with 
+800~times. 
+Note that PUT128 was run on a machine, with 
 a specification illustrated in Table~\ref{tab:machine_config}, 
@@
 We next elaborate on the proposed timing protocol. In turn we evaluate the performance of the scheme using real workloads.
-\shorten{Finally, we conclude this letter.}
-
-%\vspace\fill
 
@@
 Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
-%These two features contribute to significantly enhancing measurement quality.
 

### 2017-02-10 sedona-ieice16/revision/resources/paper_blue.tex
Modified Revision of SEDONA

@@
 \begin{summary}
-{\color{blue}Measuring program execution time is 
+{\color{blue}
+Measuring program execution time is 
 a \hbox{much-used} technique for performance evaluation in computer science. 
@@
 In this paper we propose a novel timing protocol to significantly 
-reduce such variability.}
+reduce such variability.
+}
 
@@
 Surprisingly, there is considerable variability in the measured time.
-%which indeed suggests the need for a better timing protocol.
-{\color{blue} The goal of this paper is to propose
+{\color{blue}
+The goal of this paper is to propose
 a better timing protocol that can significantly reduce such variability 
-and thus enable better timing results without a doubt of their validity.}
+and thus enable better timing results without a doubt of their validity.
+}
 
@@
 This program runs a nested for-loop with a specified task length ($tl$) (in seconds). 
-{\color{blue}The $tl$ value is used to compute 
+{\color{blue}
+The $tl$ value is used to compute 
 the number of iterations ($t$) for which that for-loop is performed 
@@
 We used 128~seconds as the task length and ran the PUT program (termed PUT128) 
-800~times. {\color{blue}Note that PUT128 was run on a machine, with 
+800~times. 
+{\color{blue}
+Note that PUT128 was run on a machine, with 
 a specification illustrated in Table~\ref{tab:machine_config}, 
 in the same timing environment as used in our prior work~\cite{Currim}, 
-with i)~non-critical daemons deactivated, 
-ii)~the Network Timing Protocol daemon switched on,  
-iii)~particular CPU features\cite{intel15,intelSpeed15} turned off, and 
-and iv)~an up-to-date kernel installed.}
+with (i)~non-critical daemons deactivated, 
+(ii)~the Network Timing Protocol daemon switched on,  
+(iii)~particular CPU features~\cite{intel15,intelSpeed15} turned off, 
+and (iv)~an up-to-date kernel installed.
 
-{\color{blue}Fig.~\ref{fig:meas_comp} shows two plots of the timing results on PUT128.
+Fig.~\ref{fig:meas_comp} shows two plots of the timing results on PUT128.
 Most of the 800 executions measured in elapsed time, 
@@
 three orders of magnitude for the same PUT128 results.
-}
 
-{\color{blue}
 {\bf Characteristics.} There are several important specifics of our protocol. 
@@
 	}
-	\caption{{\color{blue}An Example of Measurement Results on a \linebreak \hbox{Compute-bound}} 
-	Program-Under-Test~\label{fig:meas_comp}}
+	\caption{{\color{blue}An Example of Measurement Results on a \linebreak \hbox{Compute-bound} 
+	Program-Under-Test}~\label{fig:meas_comp}}
 	\vspace{-.18in}
@@
 
-\item {\color{blue}We evaluate the performance of the protocol with rigorous experiments 
-using real-world programs.}
+{\color{blue}
+\item We evaluate the performance of the protocol with rigorous experiments 
+using real-world programs.
+}
 
@@
 \noindent
-{\color{blue}The rest is organized as follows. 
+{\color{blue}
+The rest is organized as follows. 
 We next elaborate on the proposed timing protocol. In turn we evaluate the performance of the scheme using real workloads. 
-\shorten{Finally, we conclude this letter.}
 }
-
-%\vspace\fill
 
@@
 called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons). 
-{\color{blue} Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
-%These two features contribute to significantly enhancing measurement quality.
+{\color{blue}
+Our scheme (1) utilizes PT (process time) to avoid absorbing timing noise from other co-running processes and (2)~identifies and eliminates executions including daemon processes that are infrequent and \hbox{long-running} via a {\em cutoff} measure.
 }
@@
 Our SEDONA protocol consists of a total of ten steps, as described in Fig.~\ref{alg:find}.
-{\color{blue}Note that this protocol (algorithm) is applicable to any arbitrary Linux system.}
+{\color{blue} 
+Note that this protocol (algorithm) is applicable to any arbitrary Linux system.}
 
@@
 {\bf A Running Example.} 
-%As motivated by our prior work~\cite{Currim}, 
-%we configure a timing environment by i) stopping non-critical daemons, 
-%ii) activating the Network Timing Protocol daemon, and 
-%iii) switching off particular CPU features\cite{intel15,intelSpeed15} if any (Step 1). 
 {\color{blue}
-First, we set up the same measurement environment as described earlier in Section~\ref{sec:intro}, 
-to eliminate known timing factors (Step~1).}
+First, we set up the same measurement environment as described earlier in \hbox{Section}~\ref{sec:intro}, 
+to eliminate known timing factors (Step~1).
+}
 
@@
 
+%Fig.~\ref{fig:reg_put128} (= Fig.~\ref{fig:hist_before_et}) 
+%plots all the 800 elapsed times (ETs) of PUT128.
+%The plot clearly shows three rows; that is, 
+%the top and middle rows represent over a dozen of outliers far from 
+%the rest of the samples clustered in the bottom row. 
+%We will now drill down into these outliers to 
+%show how to reliably eliminate the indirect influence of 
+%some ``infrequent, \hbox{long-running} daemons'' on PT (\hbox{process} time) of the PUT.
 Fig.~\ref{fig:reg_put128} plots all the 800 elapsed times (ETs) of the run of PUT128.
-%We see three rows in the plot: a solid row of many samples, perhaps
-%six or more samples that are just above that solid row, and two
-%samples that are way above the solid row. 
 The plot clearly shows three rows; that is, 
@@
 the rest of the samples clustered in the bottom row. 
-We'll now drill down into these outliers to 
+We will now drill down into these outliers to 
 show how to reliably eliminate the indirect influence of 
-some ``infrequent, long-running daemons'' on PT (\hbox{process} time) of the PUT.
+some ``infrequent, \hbox{long-running} daemons'' on PT (\hbox{process} time) of the PUT.
 
+{\color{blue}
 To identify such daemons, we use a novel \hbox{scatter} plot: 
-those of {\em pairs of successive samples}{\color{blue}. 
+those of {\em pairs of successive samples}. 
 So samples~\#{1} and~\#{2} form the first pair and samples~\#{3} and~\#{4} 
@@
 sample~\#{1} of dual-PUT256 is equivalent 
-to samples~\#{1} and \#{2} of PUT128.}
+to samples~\#{1} and~\#{2} of PUT128.
+}
 
@@
 from a run of the 800 PUT128 samples (Step~3). 
-{\color{blue} The $xy$-plane corresponds to the dual-PUT samples. 
+{\color{blue}
+The $xy$-plane corresponds to the dual-PUT samples. 
 For example, \hbox{sample}~\#{2} of a run of dual-PUT256 
-is plotted as a point with samples \#{3} and~\#{4} of a run of PUT128 
+is plotted as a point with samples~\#{3} and~\#{4} of a run of PUT128 
 on the respective $x$ and $y$ axes. In that sense 
-the $x$ ($y$) axis is named as ``Odd (Even) Sample's ET.''}
+the $x$ ($y$) axis is named as ``Odd (Even) Sample's ET.''
+}
 There are two quite obvious outliers with ETs 
@@
 	\subfigure[Dual-PUT256 Samples]{
-		%\includegraphics[width=0.223\textwidth]{dual_et_put256_empv4.eps}
 		\includegraphics[width=0.224\textwidth]{dual_et_put256_empv4.eps}
 		\label{fig:raw_put128}
-	}
-	%\vspace{-0.1in}    
+	}\vspace{-0.15in}
 	\subfigure[Zooming in on the $L$-shape]{
@@
 the maximum process time and standard deviation 
-of PT\shorten{(process time, note the switch in emphasis from elapsed time 
-to process time)} of the daemon processes
-%(e.g. {\tt cifsd}, {\tt flush-9:0}, {\tt jbd2/md0-8}, 
-%{\tt kblockd/0}, {\tt khugepaged}, {\tt md0\_raid1}, and {\tt ntpd}) 
+of PT (process time, note the switch in emphasis from ET to PT) of the daemon processes
 (i.e. {\tt flush-9:0}) 
 observed in the central cluster samples in Fig.~\ref{fig:dual_et_put256_cluster}.
-\shorten{In the running example the maximum PT of those daemons was under 4 msec, 
-and their standard deviation was just about 1 msec.}
 
@@
 {\tt jbd2/md0-8}, and {\tt md0\_raid1} 
-are determined as infrequent, long-running.
-%We also identify ``extra'' infrequent daemons: those
-%found only in the $L$-samples but not in the central cluster. 
-%The running example reveals the following extra daemons: 
-%{\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
-%{\tt rhsmcertd-worke}, and {\tt sshd}.
-{\color{blue}We also identify ``extra'' infrequent daemons: 
+are determined as infrequent, \hbox{long-running}.
+{\color{blue}
+We also identify ``extra'' infrequent daemons: 
 {\tt bash}, {\tt grep}, {\tt rhn\_check}, {\tt rhnsd}, {\tt rhsmcertd}, 
@@
 found only in the $L$-samples but not in the central cluster.
-}
 
-{\color{blue}
 For each of the infrequent daemons 
@@
 For instance, {\tt rhn\_check} appears 
-roughly every 112 samples 
-%(which corresponds to very close to every four hours). 
-(or almost every four hours). 
-%Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, 
-%{\tt md0\_raid1}, and {\tt rhn\_check}) all occur together 
-%(in those two outliers in Fig.~\ref{fig:raw_put128}) and 
-%  have a periodicity of about every 559 samples
-%  (5x longer, or just about 20 hours). 
+roughly every 112 samples (or almost every four hours). 
 Four others ({\tt flush-9:0}, {\tt jbd2/md0-8}, {\tt md0\_raid1}, and {\tt rhn\_check}) 
@@
 (5x longer, or just about 20 hours). 
-%  It is important only which samples a given daemon shows up 
-%  in; we don't care in this heuristic whether daemons are co-occurring.
 }
@@
 clusters of \hbox{dual-PUT256} and \hbox{dual-PUT32768}{\color{blue}, each consisting of 
-pairs of two successive samples of PUT128 and PUT16384}.\shorten{: 
-{\tt flush-9.0}, {\tt jbd2/md0-8}, {\tt kblock/0}, {\tt md0\_raid1}, 
-and {\tt ntpd}.} 
+pairs of two successive samples of PUT128 and PUT16384.}
 That said, the central cluster also contains other
@@
   infrequent}, several having \hbox{periodicities} estimated at four or twenty hours.
-When {\color{blue}PUT128} had a ``short'' 
-program time {\color{blue}(or, two minutes)}, daemons with a periodicity of
-hours are \hbox{infrequent}. But with {\color{blue}PUT16384} 
-with a ``long'' program time {\color{blue}(or, 4.5 hours)}, some of those daemons are now frequent, and appear in the central cluster.
+{\color{blue}
+When PUT128 had a ``short'' 
+program time (or, two minutes), daemons with a periodicity of
+hours are \hbox{infrequent}. But with PUT16384 
+with a ``long'' program time (or, 4.5 hours),} 
+some of those daemons are now frequent, and appear in the central cluster.
 
@@
 OS & Red Hat Ent. Linux (RHEL) 6.4 with a kernel of 2.6.32 \\ \hline
-CPU & Intel Core i7-870 Lynnfield 2.93GHz quad-core \hbox{processor}\shorten{ on a LGA 1156 95W motherboard}\\ \hline
+CPU & Intel Core i7-870 Lynnfield 2.93GHz quad-core \hbox{processor}\\ \hline
 RAM & 4GB of DDR3 1333 dual-channel memory\\ \hline
@@
 matrix multiplication (in column major), termed MM. 
-In this comparison SEDONA bettered ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
+In this comparison SEDONA outperformed ORG, as can be seen in Fig.~\ref{fig:synprog_test}. 
 As workload size increased, the performance gap between the two 
@@
 	}
-%	\subfigure[{\color{blue}SORT - Relative Error}]{
-%		\includegraphics[scale=0.31]{sort_re.eps}
-%		\label{fig:sort_re}
-%	}
 	\subfigure[{\color{blue}MM - Standard Deviation}]{
@@
 	}
-%	\subfigure[{\color{blue}MM - Relative Error}]{
-%		\includegraphics[scale=0.31]{matc_re.eps}
-%		\label{fig:matc_re}
-%	}
 	\vspace{-.1in}
-	\caption{{\color{blue}Performance Comparison on Real-world Programs~\label{fig:synprog_test}}}
-\vspace{-.2in}
+	\caption{Performance Comparison on Real-world Programs~\label{fig:synprog_test}}
+\vspace{-.25in}
 \end{figure}
@@
 Table~\ref{tab:spec_real} shows that 
-SEDONA outperformed ORG,
+SEDONA bettered ORG,
 on the standard deviation and relative error across the very different SPEC benchmarks. 
@@
 lower than that of ORG for almost every benchmark.
-{\color{blue}About a 10x \hbox{margin} between the two resulted from {\tt 434}, for example.}
+For example, about a 10x \hbox{margin} between the two resulted from {\tt 434}.
 }
 
-%\vspace{-.15in}
 SEDONA also scaled well for the SPEC workloads, 
@@
 for the medium-length benchmarks (e.g. {\tt 447}, {\tt 456}, {\tt 470}, and {\tt 473}).
-{\color{blue}Even for the long-running benchmarks (e.g. {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
-the relative error of \hbox{SEDONA} was still lower than that of ORG.}
+{\color{blue}
+Even for the long-running benchmarks (e.g. {\tt 416}, {\tt 436}, and {\tt 454}, both $>$1,000 sec), 
+the relative error of \hbox{SEDONA} was still lower than that of ORG.
+}
 
 To summarize, our proposed SEDONA protocol can achieve {\em better} accuracy,
-precision, and scalability in measuring the execution time of real \hbox{compute-bound} benchmarks than the conventional timing method. 
-{\color{blue}The experimental results demonstrate 
+precision, and scalability in measuring the execution time of real \hbox{compute-bound} programs 
+than the ET-based existing method. 
+{\color{blue}
+The experimental results demonstrate 
 the general applicability of SEDONA to timing any CPU-bound program.}
 
-\vspace{-0.1in}
+%\vspace{-0.in}
 \begin{table}[t]
@@
 \tiny
-\begin{tabular}{|p{0.1cm}|R{1.1cm}|R{1.1cm}|R{0.6cm}|R{0.6cm}|R{0.97cm}|R{0.97cm}|} \hline
+\begin{tabular}{|p{0.1cm}|R{1.1cm}|R{1.1cm}|R{0.6cm}|R{0.6cm}|R{0.92cm}|R{0.92cm}|} \hline
  		  & \multicolumn{2}{c|}{{\tiny Execution Time (ms)}} & \multicolumn{2}{c|}{{\tiny Standard}}   & \multicolumn{2}{c|}{\multirow{2}{*}{{\tiny Relative Error}}}\\ \cline{2-3}
@@
 {{\tt 433}} & 480,586 & 473,256  & {743} & {725} & {2$\times$10$^{-3}$} & {2$\times$10$^{-3}$}\\ %% V
-{{\tt 434}} & 16,495  & 16,242  & {\color{blue}{75}} & {\color{blue}{7}} & {\color{blue}{5$\times$10$^{-3}$}} & {\color{blue}{5$\times$10$^{-4}$}}\\
+%{{\tt 434}} & 16,495  & 16,242  & {75} & {7} & {5$\times$10$^{-3}$} & {5$\times$10$^{-4}$}\\
+{{\color{blue}{\tt 434}}} & 16,495  & 16,242  & {\color{blue}{75}} & {\color{blue}{7}} & {{\color{blue}5$\times$10$^{-3}$}} & {{\color{blue}5$\times$10$^{-4}$}}\\
 {{\tt 435}} & 990,575 & 975,445  & {947} & {900} & {1$\times$10$^{-3}$} & {9$\times$10$^{-4}$}\\
@@
 \vspace{-0.07in}
-%We \shorten{provided an empirical evidence of 
-%how significantly measurement quality could be hurt by 
-%the presence of infrequent, long-running daemon processes. 
-%We }proposed a novel timing protocol called {\em SEDONA}\shorten{, 
-%employing process time and identifying an infrequent, long-running daemon}, 
-%more precise and \hbox{accurate} than the traditional method, 
-%and performed its rigorous empirical evaluation\shorten{
-%as shown in the experiments}.
 We proposed a novel timing protocol called {\em SEDONA}, and 
 performed an empirical evaluation to show that 
-it is more precise and accurate than the extant method.
-}
+it is more precise and accurate than the extant method.}
 

### 2017-02-13 pt_characterization/machine_cons_check/paper.tex
machine consistency check

\documentclass[10pt]{article}
\usepackage[margin=1in]{geometry}
 \usepackage{auto-pst-pdf}
\usepackage{graphicx}
%\usepackage{arydshln}
\usepackage{ifpdf}
\ifpdf
  \usepackage{epstopdf}
\fi
\usepackage{multirow}
\usepackage{epsfig}
\usepackage{float}
\usepackage{url}
\usepackage{color}
\usepackage{subfigure}

%\newcommand\solidrule[1][1cm]{\rule[0.5ex]{#1}{.4pt}}
%\newcommand\dashedrule{\mbox{%
%  \solidrule[2mm]\hspace{2mm}\solidrule[2mm]\hspace{2mm}\solidrule[2mm]}}

%\usepackage{hyperref}

\begin{document}
\title{Machine Homogeneity Check on EMPv5}

\author{
Young-Kyoon Suh\\
}
\maketitle

\section{Experiment Notes}
%% done
\begin{table}[h]
\begin{center}
\begin{tabular}{|p{2cm}|p{3cm}|p{6cm}|p{4cm}|} \hline
Machine & Task Length & Description & Time Length\\ \hline
{\tt sodb8} &  INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
{\tt sodb9} &  INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
{\tt sodb10} & INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
{\tt sodb12} & INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
\end{tabular}
\end{center}
\vspace{-.2in}
\caption{Detailed description of INC data used for histograms\label{tab:exp_notes}}
\end{table}

The following sections presents experimental results on a machine basis. 
We used four nodes in a cluster. 
Each section exhibits histograms of elapsed time (ET) and process time (PT) 
over increasing INC's task lengths on each individual machine.

%\begin{table}[h]
%\begin{center}
%\begin{tabular}{|l|l|l|l}|} \hline
%Machine & {\tt sodb8} & {\tt sodb9} & {\tt sodb10} & {\tt sodb12} \\ \hline
%1  & \\ \hline
%2  & \\ \hline
%4  & \\ \hline
%8  & \\ \hline
%16 & \\ \hline
%32 & \\ \hline
%64 & \\ \hline
%128 & \\ \hline
%\end{tabular}
%\end{center}
%\caption{Machine-wise Run Statistics: Standard Deviation\label{tab:std_machine}}
%\end{table}

\input{sodb8/res.tex}
\input{sodb9/res.tex}
\input{sodb10/res.tex}
\input{sodb12/res.tex}

\end{document}

### 2017-02-13 pt_characterization/machine_cons_check/sodb10/res.tex
machine consistency check

\newpage

\section{{\tt sodb10}~\label{sec:sodb10_hist}} 
This section exhibits histograms on the EMPv5 data obtained on {\tt sodb10}. 
The detailed description of the base data are from Table~\ref{tab:exp_notes}.

\subsection{ET}

\begin{figure}[hp!]
	\centering
	\subfigure[ET frequency on INC1]{
		\includegraphics[scale=0.43]{sodb10/1_sec_et_hist_v5.eps}
		\label{fig:inc1_et_hist_v5}
	}
	\subfigure[ET frequency on INC2]{
		\includegraphics[scale=0.43]{sodb10/2_sec_et_hist_v5.eps}
		\label{fig:inc2_et_hist_v5}
	}
	\subfigure[ET frequency on INC4]{
		\includegraphics[scale=0.43]{sodb10/4_sec_et_hist_v5.eps}
		\label{fig:inc4_et_hist_v5}
	}
	\subfigure[ET frequency on INC8]{
		\includegraphics[scale=0.43]{sodb10/8_sec_et_hist_v5.eps}
		\label{fig:inc8_et_hist_v5}
	}
	\caption{ET Histograms of INC1 ... INC8~\label{fig:s10_et_hist1}}
\end{figure}

\begin{figure}[hp!]
	\centering
	\subfigure[ET frequency on INC16]{
		\includegraphics[scale=0.43]{sodb10/16_sec_et_hist_v5.eps}
		\label{fig:inc16_et_hist_v5}
	}
	\subfigure[ET frequency on INC32]{
		\includegraphics[scale=0.43]{sodb10/32_sec_et_hist_v5.eps}
		\label{fig:inc32_et_hist_v5}
	}
	\subfigure[ET frequency on INC64]{
		\includegraphics[scale=0.43]{sodb10/64_sec_et_hist_v5.eps}
		\label{fig:inc64_et_hist_v5}
	}
	\subfigure[ET frequency on INC128]{
		\includegraphics[scale=0.43]{sodb10/128_sec_et_hist_v5.eps}
		\label{fig:inc128_et_hist_v5}
	}
	\caption{ET Histograms of INC16 ... INC128~\label{fig:s10_et_hist2}}
\end{figure}

\newpage

\subsection{PT}

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on INC1]{
		\includegraphics[scale=0.43]{sodb10/1_sec_pt_hist_v5.eps}
		\label{fig:inc1_hist_v5}
	}
	\subfigure[PT frequency on INC2]{
		\includegraphics[scale=0.43]{sodb10/2_sec_pt_hist_v5.eps}
		\label{fig:inc2_hist_v5}
	}
	\subfigure[PT frequency on INC4]{
		\includegraphics[scale=0.43]{sodb10/4_sec_pt_hist_v5.eps}
		\label{fig:inc4_hist_v5}
	}
	\subfigure[PT frequency on INC8]{
		\includegraphics[scale=0.43]{sodb10/8_sec_pt_hist_v5.eps}
		\label{fig:inc8_hist_v5}
	}
	\caption{PT Histograms of INC1 ... INC8~\label{fig:s10_pt_hist1}}
\end{figure}

\begin{figure}[hp!]
	\centering
	\subfigure[PT frequency on INC16]{
		\includegraphics[scale=0.43]{sodb10/16_sec_pt_hist_v5.eps}
		\label{fig:inc16_hist_v5}
	}
	\subfigure[PT frequency on INC32]{
		\includegraphics[scale=0.43]{sodb10/32_sec_pt_hist_v5.eps}
		\label{fig:inc32_hist_v5}
	}
	\subfigure[PT frequency on INC64]{
		\includegraphics[scale=0.43]{sodb10/64_sec_pt_hist_v5.eps}
		\label{fig:inc64_hist_v5}
	}
	\subfigure[PT frequency on INC128]{
		\includegraphics[scale=0.43]{sodb10/128_sec_pt_hist_v5.eps}
		\label{fig:inc128_hist_v5}
	}
	\caption{PT Histograms of INC16 ... INC64~\label{fig:s10_pt_hist2}}
\end{figure}

### 2017-02-13 pt_characterization/machine_cons_check/sodb12/res.tex
machine consistency check

--- 2017-02-13 pt_characterization/machine_cons_check/sodb10/res.tex
@@
 
-\section{{\tt sodb10}~\label{sec:sodb10_hist}} 
-This section exhibits histograms on the EMPv5 data obtained on {\tt sodb10}. 
+\section{{\tt sodb12}~\label{sec:sodb12_hist}} 
+This section exhibits histograms on the EMPv5 data obtained on {\tt sodb12}. 
 The detailed description of the base data are from Table~\ref{tab:exp_notes}.
@@
 	\subfigure[ET frequency on INC1]{
-		\includegraphics[scale=0.43]{sodb10/1_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/1_sec_et_hist_v5.eps}
 		\label{fig:inc1_et_hist_v5}
@@
 	\subfigure[ET frequency on INC2]{
-		\includegraphics[scale=0.43]{sodb10/2_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/2_sec_et_hist_v5.eps}
 		\label{fig:inc2_et_hist_v5}
@@
 	\subfigure[ET frequency on INC4]{
-		\includegraphics[scale=0.43]{sodb10/4_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/4_sec_et_hist_v5.eps}
 		\label{fig:inc4_et_hist_v5}
@@
 	\subfigure[ET frequency on INC8]{
-		\includegraphics[scale=0.43]{sodb10/8_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/8_sec_et_hist_v5.eps}
 		\label{fig:inc8_et_hist_v5}
 	}
-	\caption{ET Histograms of INC1 ... INC8~\label{fig:s10_et_hist1}}
+	\caption{ET Histograms of INC1 ... INC8~\label{fig:s12_et_hist1}}
 \end{figure}
@@
 	\subfigure[ET frequency on INC16]{
-		\includegraphics[scale=0.43]{sodb10/16_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/16_sec_et_hist_v5.eps}
 		\label{fig:inc16_et_hist_v5}
@@
 	\subfigure[ET frequency on INC32]{
-		\includegraphics[scale=0.43]{sodb10/32_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/32_sec_et_hist_v5.eps}
 		\label{fig:inc32_et_hist_v5}
@@
 	\subfigure[ET frequency on INC64]{
-		\includegraphics[scale=0.43]{sodb10/64_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/64_sec_et_hist_v5.eps}
 		\label{fig:inc64_et_hist_v5}
@@
 	\subfigure[ET frequency on INC128]{
-		\includegraphics[scale=0.43]{sodb10/128_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/128_sec_et_hist_v5.eps}
 		\label{fig:inc128_et_hist_v5}
 	}
-	\caption{ET Histograms of INC16 ... INC128~\label{fig:s10_et_hist2}}
+	\caption{ET Histograms of INC16 ... INC128~\label{fig:s12_et_hist2}}
 \end{figure}
@@
 	\subfigure[PT frequency on INC1]{
-		\includegraphics[scale=0.43]{sodb10/1_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/1_sec_pt_hist_v5.eps}
 		\label{fig:inc1_hist_v5}
@@
 	\subfigure[PT frequency on INC2]{
-		\includegraphics[scale=0.43]{sodb10/2_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/2_sec_pt_hist_v5.eps}
 		\label{fig:inc2_hist_v5}
@@
 	\subfigure[PT frequency on INC4]{
-		\includegraphics[scale=0.43]{sodb10/4_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/4_sec_pt_hist_v5.eps}
 		\label{fig:inc4_hist_v5}
@@
 	\subfigure[PT frequency on INC8]{
-		\includegraphics[scale=0.43]{sodb10/8_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/8_sec_pt_hist_v5.eps}
 		\label{fig:inc8_hist_v5}
 	}
-	\caption{PT Histograms of INC1 ... INC8~\label{fig:s10_pt_hist1}}
+	\caption{PT Histograms of INC1 ... INC8~\label{fig:s12_pt_hist1}}
 \end{figure}
@@
 	\subfigure[PT frequency on INC16]{
-		\includegraphics[scale=0.43]{sodb10/16_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/16_sec_pt_hist_v5.eps}
 		\label{fig:inc16_hist_v5}
@@
 	\subfigure[PT frequency on INC32]{
-		\includegraphics[scale=0.43]{sodb10/32_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/32_sec_pt_hist_v5.eps}
 		\label{fig:inc32_hist_v5}
@@
 	\subfigure[PT frequency on INC64]{
-		\includegraphics[scale=0.43]{sodb10/64_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/64_sec_pt_hist_v5.eps}
 		\label{fig:inc64_hist_v5}
@@
 	\subfigure[PT frequency on INC128]{
-		\includegraphics[scale=0.43]{sodb10/128_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb12/128_sec_pt_hist_v5.eps}
 		\label{fig:inc128_hist_v5}
 	}
-	\caption{PT Histograms of INC16 ... INC64~\label{fig:s10_pt_hist2}}
+	\caption{PT Histograms of INC16 ... INC64~\label{fig:s12_pt_hist2}}
 \end{figure}

### 2017-02-13 pt_characterization/machine_cons_check/sodb8/res.tex
machine consistency check

--- 2017-02-13 pt_characterization/machine_cons_check/sodb12/res.tex
@@
 
-\section{{\tt sodb12}~\label{sec:sodb12_hist}} 
-This section exhibits histograms on the EMPv5 data obtained on {\tt sodb12}. 
+\section{{\tt sodb8}~\label{sec:sodb8_hist}} 
+This section exhibits histograms on the EMPv5 data obtained on {\tt sodb8}. 
 The detailed description of the base data are from Table~\ref{tab:exp_notes}.
@@
 	\subfigure[ET frequency on INC1]{
-		\includegraphics[scale=0.43]{sodb12/1_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/1_sec_et_hist_v5.eps}
 		\label{fig:inc1_et_hist_v5}
@@
 	\subfigure[ET frequency on INC2]{
-		\includegraphics[scale=0.43]{sodb12/2_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/2_sec_et_hist_v5.eps}
 		\label{fig:inc2_et_hist_v5}
@@
 	\subfigure[ET frequency on INC4]{
-		\includegraphics[scale=0.43]{sodb12/4_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/4_sec_et_hist_v5.eps}
 		\label{fig:inc4_et_hist_v5}
@@
 	\subfigure[ET frequency on INC8]{
-		\includegraphics[scale=0.43]{sodb12/8_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/8_sec_et_hist_v5.eps}
 		\label{fig:inc8_et_hist_v5}
 	}
-	\caption{ET Histograms of INC1 ... INC8~\label{fig:s12_et_hist1}}
+	\caption{ET Histograms of INC1 ... INC8~\label{fig:s8_et_hist1}}
 \end{figure}
@@
 	\subfigure[ET frequency on INC16]{
-		\includegraphics[scale=0.43]{sodb12/16_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/16_sec_et_hist_v5.eps}
 		\label{fig:inc16_et_hist_v5}
@@
 	\subfigure[ET frequency on INC32]{
-		\includegraphics[scale=0.43]{sodb12/32_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/32_sec_et_hist_v5.eps}
 		\label{fig:inc32_et_hist_v5}
@@
 	\subfigure[ET frequency on INC64]{
-		\includegraphics[scale=0.43]{sodb12/64_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/64_sec_et_hist_v5.eps}
 		\label{fig:inc64_et_hist_v5}
@@
 	\subfigure[ET frequency on INC128]{
-		\includegraphics[scale=0.43]{sodb12/128_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/128_sec_et_hist_v5.eps}
 		\label{fig:inc128_et_hist_v5}
 	}
-	\caption{ET Histograms of INC16 ... INC128~\label{fig:s12_et_hist2}}
+	\caption{ET Histograms of INC16 ... INC128~\label{fig:s8_et_hist2}}
 \end{figure}
@@
 	\subfigure[PT frequency on INC1]{
-		\includegraphics[scale=0.43]{sodb12/1_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/1_sec_pt_hist_v5.eps}
 		\label{fig:inc1_hist_v5}
@@
 	\subfigure[PT frequency on INC2]{
-		\includegraphics[scale=0.43]{sodb12/2_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/2_sec_pt_hist_v5.eps}
 		\label{fig:inc2_hist_v5}
@@
 	\subfigure[PT frequency on INC4]{
-		\includegraphics[scale=0.43]{sodb12/4_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/4_sec_pt_hist_v5.eps}
 		\label{fig:inc4_hist_v5}
@@
 	\subfigure[PT frequency on INC8]{
-		\includegraphics[scale=0.43]{sodb12/8_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/8_sec_pt_hist_v5.eps}
 		\label{fig:inc8_hist_v5}
 	}
-	\caption{PT Histograms of INC1 ... INC8~\label{fig:s12_pt_hist1}}
+	\caption{PT Histograms of INC1 ... INC8~\label{fig:s8_pt_hist1}}
 \end{figure}
@@
 	\subfigure[PT frequency on INC16]{
-		\includegraphics[scale=0.43]{sodb12/16_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/16_sec_pt_hist_v5.eps}
 		\label{fig:inc16_hist_v5}
@@
 	\subfigure[PT frequency on INC32]{
-		\includegraphics[scale=0.43]{sodb12/32_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/32_sec_pt_hist_v5.eps}
 		\label{fig:inc32_hist_v5}
@@
 	\subfigure[PT frequency on INC64]{
-		\includegraphics[scale=0.43]{sodb12/64_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/64_sec_pt_hist_v5.eps}
 		\label{fig:inc64_hist_v5}
@@
 	\subfigure[PT frequency on INC128]{
-		\includegraphics[scale=0.43]{sodb12/128_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb8/128_sec_pt_hist_v5.eps}
 		\label{fig:inc128_hist_v5}
 	}
-	\caption{PT Histograms of INC16 ... INC64~\label{fig:s12_pt_hist2}}
+	\caption{PT Histograms of INC16 ... INC64~\label{fig:s8_pt_hist2}}
 \end{figure}

### 2017-02-13 pt_characterization/machine_cons_check/sodb9/res.tex
machine consistency check

--- 2017-02-13 pt_characterization/machine_cons_check/sodb8/res.tex
@@
 
-\section{{\tt sodb8}~\label{sec:sodb8_hist}} 
-This section exhibits histograms on the EMPv5 data obtained on {\tt sodb8}. 
+\section{{\tt sodb9}~\label{sec:sodb9_hist}} 
+This section exhibits histograms on the EMPv5 data obtained on {\tt sodb9}. 
 The detailed description of the base data are from Table~\ref{tab:exp_notes}.
@@
 	\subfigure[ET frequency on INC1]{
-		\includegraphics[scale=0.43]{sodb8/1_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/1_sec_et_hist_v5.eps}
 		\label{fig:inc1_et_hist_v5}
@@
 	\subfigure[ET frequency on INC2]{
-		\includegraphics[scale=0.43]{sodb8/2_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/2_sec_et_hist_v5.eps}
 		\label{fig:inc2_et_hist_v5}
@@
 	\subfigure[ET frequency on INC4]{
-		\includegraphics[scale=0.43]{sodb8/4_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/4_sec_et_hist_v5.eps}
 		\label{fig:inc4_et_hist_v5}
@@
 	\subfigure[ET frequency on INC8]{
-		\includegraphics[scale=0.43]{sodb8/8_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/8_sec_et_hist_v5.eps}
 		\label{fig:inc8_et_hist_v5}
 	}
-	\caption{ET Histograms of INC1 ... INC8~\label{fig:s8_et_hist1}}
+	\caption{ET Histograms of INC1 ... INC8~\label{fig:s9_et_hist1}}
 \end{figure}
@@
 	\subfigure[ET frequency on INC16]{
-		\includegraphics[scale=0.43]{sodb8/16_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/16_sec_et_hist_v5.eps}
 		\label{fig:inc16_et_hist_v5}
@@
 	\subfigure[ET frequency on INC32]{
-		\includegraphics[scale=0.43]{sodb8/32_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/32_sec_et_hist_v5.eps}
 		\label{fig:inc32_et_hist_v5}
@@
 	\subfigure[ET frequency on INC64]{
-		\includegraphics[scale=0.43]{sodb8/64_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/64_sec_et_hist_v5.eps}
 		\label{fig:inc64_et_hist_v5}
@@
 	\subfigure[ET frequency on INC128]{
-		\includegraphics[scale=0.43]{sodb8/128_sec_et_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/128_sec_et_hist_v5.eps}
 		\label{fig:inc128_et_hist_v5}
 	}
-	\caption{ET Histograms of INC16 ... INC128~\label{fig:s8_et_hist2}}
+	\caption{ET Histograms of INC16 ... INC128~\label{fig:s9_et_hist2}}
 \end{figure}
@@
 	\subfigure[PT frequency on INC1]{
-		\includegraphics[scale=0.43]{sodb8/1_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/1_sec_pt_hist_v5.eps}
 		\label{fig:inc1_hist_v5}
@@
 	\subfigure[PT frequency on INC2]{
-		\includegraphics[scale=0.43]{sodb8/2_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/2_sec_pt_hist_v5.eps}
 		\label{fig:inc2_hist_v5}
@@
 	\subfigure[PT frequency on INC4]{
-		\includegraphics[scale=0.43]{sodb8/4_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/4_sec_pt_hist_v5.eps}
 		\label{fig:inc4_hist_v5}
@@
 	\subfigure[PT frequency on INC8]{
-		\includegraphics[scale=0.43]{sodb8/8_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/8_sec_pt_hist_v5.eps}
 		\label{fig:inc8_hist_v5}
 	}
-	\caption{PT Histograms of INC1 ... INC8~\label{fig:s8_pt_hist1}}
+	\caption{PT Histograms of INC1 ... INC8~\label{fig:s9_pt_hist1}}
 \end{figure}
@@
 	\subfigure[PT frequency on INC16]{
-		\includegraphics[scale=0.43]{sodb8/16_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/16_sec_pt_hist_v5.eps}
 		\label{fig:inc16_hist_v5}
@@
 	\subfigure[PT frequency on INC32]{
-		\includegraphics[scale=0.43]{sodb8/32_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/32_sec_pt_hist_v5.eps}
 		\label{fig:inc32_hist_v5}
@@
 	\subfigure[PT frequency on INC64]{
-		\includegraphics[scale=0.43]{sodb8/64_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/64_sec_pt_hist_v5.eps}
 		\label{fig:inc64_hist_v5}
@@
 	\subfigure[PT frequency on INC128]{
-		\includegraphics[scale=0.43]{sodb8/128_sec_pt_hist_v5.eps}
+		\includegraphics[scale=0.43]{sodb9/128_sec_pt_hist_v5.eps}
 		\label{fig:inc128_hist_v5}
 	}
-	\caption{PT Histograms of INC16 ... INC64~\label{fig:s8_pt_hist2}}
+	\caption{PT Histograms of INC16 ... INC64~\label{fig:s9_pt_hist2}}
 \end{figure}

### 2017-02-13 pt_characterization/machine_cons_check/sodb10/et_hist_sodb10.r

@@ -1,0 +1,171 @@
+### sodb10
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb10/pt_hist_sodb10.r

@@ -1,0 +1,155 @@
+### sodb10
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- 250
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb12/et_hist_sodb12.r

@@ -1,0 +1,171 @@
+### sodb12
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb12/pt_hist_sodb12.r

@@ -1,0 +1,155 @@
+### sodb12
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- 150
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb8/et_hist_sodb8.r

@@ -1,0 +1,170 @@
+### sodb8
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb8/pt_hist_sodb8.r

@@ -1,0 +1,154 @@
+### sodb8
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC2', 
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb9/et_hist_sodb9.r

@@ -1,0 +1,171 @@
+### sodb9
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
[…]

### 2017-02-13 pt_characterization/machine_cons_check/sodb9/pt_hist_sodb9.r

@@ -1,0 +1,155 @@
+### sodb9
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- 250
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
[…]

### 2017-02-13 pt_characterization/machine_cons_check/paper.tex
Add some plots for machine homogeneity check

@@
 \begin{document}
-\title{Machine Homogeneity Check on EMPv5}
+\title{Empirical Check of Machine Homogeneity on Timing Protocol}
 
@@
 
+\section{Description}
+In this work we do a check to see if the result on the same experiment is different 
+on different nodes, each with the same machine specification.
+
 \section{Experiment Notes}
+We provide a short description of our experimental runs. 
+In these experiments we used four nodes ({\tt sodb8}, {\tt sodb9}, {\tt sodb10}, and {\tt sodb12} ) 
+in the same cluster.
+
 %% done
@@
 
-The following sections presents experimental results on a machine basis. 
-We used four nodes in a cluster. 
-Each section exhibits histograms of elapsed time (ET) and process time (PT) 
-over increasing INC's task lengths on each individual machine.
+\paragraph{Measurement Quality Across the SoDB Nodes:} 
+Fig.~\ref{fig:machine_comp} shows 
+the standard deviation and relative error of 
+elapsed time (ET) and process time (PT) on the different SoDB nodes 
+as the task length of INC increases.
 
-%\begin{table}[h]
-%\begin{center}
-%\begin{tabular}{|l|l|l|l}|} \hline
-%Machine & {\tt sodb8} & {\tt sodb9} & {\tt sodb10} & {\tt sodb12} \\ \hline
-%1  & \\ \hline
-%2  & \\ \hline
-%4  & \\ \hline
-%8  & \\ \hline
-%16 & \\ \hline
-%32 & \\ \hline
-%64 & \\ \hline
-%128 & \\ \hline
-%\end{tabular}
-%\end{center}
-%\caption{Machine-wise Run Statistics: Standard Deviation\label{tab:std_machine}}
-%\end{table}
+\begin{figure}[t]
+	\centering
+	\subfigure[Standard Deviation - ET]{
+		\includegraphics[scale=0.6]{overall/machine_et_std.eps}
+        \label{fig:pt_std}
+    }
+    \subfigure[Relative Error - ET]{
+        \includegraphics[scale=0.6]{overall/machine_et_re.eps}
+        \label{fig:pt_re}
+    }
+	\subfigure[Standard Deviation - PT]{
+		\includegraphics[scale=0.6]{overall/machine_pt_std.eps}
+        \label{fig:pt_std}
+    }
+    \subfigure[Relative Error - PT]{
+        \includegraphics[scale=0.6]{overall/machine_pt_re.eps}
+        \label{fig:pt_re}
+    }
+    \caption{Measurement Quality Comparison among Different SoDB Machines}
+    \label{fig:machine_comp}
+\end{figure}
+
+\newpage
+
+Each of the following sections exhibits histograms of ET and PT 
+over increasing INC's task lengths on each individual node.
 
@@
 
+\newpage
+
+\section{Conclusion}
+My recommendation is to avoid {\tt sodb8} and to use the rest of 
+the SoDB nodes for the rest of our experiments. Further investigation is 
+needed to examine why the experiment results on {\tt sodb8} are different 
+than those of the other nodes.
+
 \end{document}

### 2017-02-16 azdblab-spe17/paper.tex
azdblab she paper

\documentclass[times]{speauth}
\usepackage{moreverb}
\usepackage{url}
\usepackage{textcomp}
\usepackage{amssymb,amsmath}
\usepackage{balance}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{subfigure}
\usepackage{float}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{latexsym}
\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{comment}
\usepackage{graphicx}           
\usepackage{epsfig}
\usepackage{algorithmic}
\usepackage{listings}
\usepackage{multirow}
\usepackage{etoolbox}
%\def\ert{\mbox{{\it expRunTime}}}
%\def\rpe{\mbox{{\it recordImageDiff}}}
%\def\procDiff{\mbox{{\it procDiff}}}
%\def\uintMax{\mbox{{\tt UINT\_MAX}}}
%\def\numRep{\mbox{{\tt NUM\_REPETITIONS}}}
%\def\DiffA{\mbox{{\it Diff$_{A}$}}}
%\def\DiffB{\mbox{{\it Diff$_{B}$}}}
%\def\getProcDiff{\mbox{{\it getProcDiff}}}
%\def\bfps{\mbox{{\it beforeImage}}}
%\def\afps{\mbox{{\it afterImage}}}
%\def\gpi{\mbox{{\it getProcInfo}}}
%\def\stm{\mbox{{\it startTime}}}
%\def\gtm{\mbox{{\it getTime}}}
%\def\prog{\mbox{{\it userProgram}}}
%\def\eput{\mbox{{\it executeProgram}}}
%\def\etm{\mbox{{\it endTime}}}
%\def\ect{\mbox{{\it elapsedClockTime}}}
%\def\et{\mbox{{\it elapsedTime}}}
%\def\pt{\mbox{{\it ProgramTime}}}
%\def\cpt{\mbox{{\it computeProgramTime}}}
%\def\inc{\mbox{INC}}
\def\azdblab{\mbox{{\sc AZDBLab}}}
\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\long\def\todo#1{\hilitey{{\bf TODO: [}} {\em #1}\hilitey{]}}
\long\def\shorten#1{}
\makeatletter
\def\@xfootnote[#1]{
  \protected@xdef\@thefnmark{#1}
  \@footnotemark\@footnotetext}
\makeatother
%c2j: Conference to Journal: first parameter is conference, second is journal
% For Conferences: c2j#1#2{#1}
% For Journals: c2j#1#2{2}
\long\def\c2j#1#2{#1}
\def\volumeyear{2017}
\begin{document}

\runningheads{Y-K.~Suh, R.~T.~Snodgrass}{$\azdblab$: A Laboratory Information System for Large-Scale Empirical DBMS Studies}

\title{$\azdblab$: A Laboratory Information System \\ for Large-Scale Empirical DBMS Studies}

\author{Young-Kyoon Suh\affil{1}\corrauth, Richard T.~Snodgrass\affil{1}
}

\address{
\affilnum{1}University of Arizona, Tucson, AZ 85721
}

\corraddr{KISTI, 245 Daehak-ro, Yuseong-gu, Daejeon, 34141, Rep. of KOREA (current address).
E-mail: \hbox{yksuh}@kisti.re.kr}

\begin{abstract}
In database field, scientific approach has been much less prominent while very stong
mathematical and engineering work has been done.
Understanding database query optimizers is of critical importance in a sense that it can provide
great insights into their improvements by looming which parts should be reexamined.
However, there have been few systems for supporting this scientific approach in which one can
simultaneously run and monitor a variety of experiments on DBMSes and analyze the
results to test his/her hypotheses.

In this manuscript we present a novel DBMS-oriented research infrastructure, 
called {\em Arizona Database Laboratory} ($\azdblab$), to assist database researchers 
in conducting a {\em large-scale} empirical study across multiple \hbox{DBMSes} in a {\em scientific} manner. 
For them to test their hypotheses on the behavior of query \hbox{optimizers}, 
$\azdblab$ can run and monitor a large-scale \hbox{experiment} with thousands (or millions) of queries on 
\hbox{different} \hbox{DBMSes}. 
Furthermore, $\azdblab$ can help users \hbox{automatically} analyze these queries. 
We elaborate in detail how $\azdblab$ seamlessly supports scientific research on a variety of relational \hbox{DBMSes}.
\end{abstract}

\keywords{}

\maketitle

\section{Introduction\label{sec:introduction}}

\section{Background}\label{sec:background}

\subsection{Accuracy and Precision in Timing}\label{sec:prec_accuracy}

\section{Conclusion}
In this paper, we presented a novel laboratory information system 
for empirical studies of different relational DBMSes.

\section{Future Work}\label{sec:future_work}

\section{Acknowledgments}
We give special thanks to Rui Zhang, for his initial development on $\azdblab$. 
We also thank Ricardo Carlos, Preetha Chatterjee, Pallavi
Chilappagari, Jennifer Dempsey, David Gallup, Kevan Holdaway, Andrey
Kvochko, Siou Lin, Adam Robertson, Lopamudra Sarangi, Linh Tran, Cheng Yi, and Man Zhang for
their contributions to the $\azdblab$. and Phil Kaslo, Tom Lowry, and John
Luiten for constructing and maintaining our experimental instrument\c2j{}{,
  a laboratory of six machines and associated software}. 
This research was supported in part by NSF grants \hbox{IIS-0639106} and \hbox{IIS-1016205}.

\bibliographystyle{wileyj}
%\bibliography{paper}

\end{document}

### 2017-02-16 azdblab-spe17/paper2.tex
azdblab she paper

%\documentclass{sig-alternate}
\documentclass{vldb}
\usepackage{url}
%\usepackage{algorithm}
%\usepackage{algorithmic}
%\usepackage{minted}
\usepackage{subfigure}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}

\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\newcommand{\INDSTATE}[1][1]{\STATE\hspace{#1\algorithmicindent}}
\long\def\todo#1{\hilitey{{\bf TODO:} {\em #1}}}
\long\def\shorten#1{}
\def\azdb{\hbox{\sc AZDBLab}}
%\def\azdb{\hbox{\sc DBLab}}
\def\QatC{Q{@}C}
\long\def\comment#1{}
%c2j: Conference to Journal: first parameter is conference, second is journal
% For Conferences: c2j#1#2{#1}
% For Journals: c2j#1#2{2}
\long\def\c2j#1#2{#1}

%\conferenceinfo{SIGMOD'14,} {June 22--27, 2014, Salt Lake City, Utah, USA.}
%\CopyrightYear{2014}
%\crdata{978-1-4503-2037-5/13/06}
%\clubpenalty=10000
%\widowpenalty = 10000

\begin{document}
\pagenumbering{gobble}

%don't want date printed
\date{}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
%\title{A Laboratory Information System for Empirical Studies on a Variety of DBMSes}
\title{AZDBLab: A Laboratory Information System for Large-Scale Empirical DBMS Studies}
%\numberofauthors{1}
%\author{
% DRAFT
%}
%
\numberofauthors{3}
\author{
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
%% e-mail address with \email.
% 1st author
\alignauthor
Young-Kyoon Suh\\
%       \affaddr{Department of Computer Science}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\ 
       \affaddr{\small\tt yksuh@cs.arizona.edu}
%2nd author
\alignauthor
Richard T.~Snodgrass\\
%       \affaddr{Department of Computer Science}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\
       \affaddr{\small\tt rts@cs.arizona.edu}
%3rd author
%\alignauthor
%Rui Zhang\\
%       \affaddr{Dataware Ventures}\\
%%       \affaddr{Torrence, CA}\\
%       \email{\small\tt ruizhang@datawareventures.com}
\alignauthor
Rui Zhang\\
%       \affaddr{Department of Computer Science}\\
       \affaddr{Dataware Ventures}\\
%       \affaddr{Tucson, AZ}\\
       \affaddr{\small\tt ruizhang@datawareventures.com}
\shorten{
\and 
%%4th author
\alignauthor
Matthew Wong Johnson\\
%      \affaddr{Department of Computer Science}\\
       \affaddr{Univ. of California, San Diego}\\
%       \affaddr{San Diego, CA}\\
       \email{\small\tt mwj@email.arizona.edu}
%%%5th author
\alignauthor
Andrey Kvochko\\
%       \affaddr{Department of Computer Science}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\
       \affaddr{\small\tt akvochko@cs.arizona.edu}
%6th author
\alignauthor
Benjamin Dicken\\
%       \affaddr{Department of Computer Science}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\
       \affaddr{\small\tt bddicken@cs.arizona.edu}
}
}
%\date{%\vspace*{-1cm}}

\maketitle
%\vspace*{-3in}

\begin{abstract}%\todo{Young: Rewrite this after finalizing body.} 
%Understanding a variety of behaviors present in DBMSes 
%can give an importance implication database query optimizers to improve. 
%There has been no system to study these aspects across different DBMSes. 
%This scientific approach leads to fundamental improvements 
%of database query optimizers. 
%To address this ,
%There has been no system of allowing database researchers to 
%simultaneously run experiments on different DBMSes and seamlessly analyze the results, 
%to better understand a variety of behaviours of database query optimizers.

%In database field, scientific approach has been much less prominent while very strong 
%mathematical and engineering work has been done. 
%Understanding database query optimizers better, for instance, is of critical importance, 
%as it can provide great insights into their improvements by looming 
%which parts should be re-examined. 
%However, there has been few systems for supporting such a scientific approach.

In the database field, while very strong 
mathematical and engineering work has been done, the 
scientific approach has been much less prominent. 
\shorten{Understanding database query optimizers better, for instance, is of critical importance, 
as it can provide insights into their improvements by \hbox{identifying} which parts should be re-examined.}
The deep understanding of query optimizers obtained through the scientific approach can lead to better engineered designs.
Unlike other domains, there have been few {\em DBMS-dedicated} laboratories, 
focusing on such scientific investigation.

%In this demonstration, we present a novel laboratory information system, 
%called {\em Arizona Database Laboratory} ($\azdb$), to assist database researchers 
%to empirically study DBMSes in a scientific manner. 
%For one to test their hypotheses on the behavior of query optimizers, 
%our system can allow him or her to simultaneously run and monitor a variety of experiments on multiple DBMSes. 
%Furthermore, the system can later automatically analyze the results. 
%The audience will interact with $\azdb$ through the stand-alone 
%application or the mobile app to get to conduct an experiment. 
%Subsequently, the audience will visually 
%see the analysis results on the completed experiment in the form of a paper.

In this demonstration, we present a novel DBMS-oriented research infrastructure, 
called {\em Arizona Database Laboratory} \linebreak ($\azdb$), to assist database researchers 
in conducting a {\em large-scale} empirical study across multiple \hbox{DBMSes}\shorten{ in a {\em scientific} manner}. 
For them to test their hypotheses on the behavior of query \linebreak \hbox{optimizers}, 
$\azdb$ can run and monitor a large-scale \hbox{experiment} with thousands (or millions) of queries on \linebreak \hbox{different} \hbox{DBMSes}. 
Furthermore, $\azdb$ can help users \linebreak \hbox{automatically} analyze these queries. 
In the demo, the \linebreak \hbox{audience} will interact with $\azdb$ through the stand-alone
\hbox{application} and the mobile app to conduct such a large-scale \hbox{experiment} for a study. 
The \hbox{audience}  will then run a Tucson \hbox{Timing} \hbox{Protocol} analysis on the finished \linebreak \hbox{experiment} and then see the analysis (data sanity check and timing) \hbox{results}.

%
%Understanding database query optimizers is of critical importance in a sense that it can provide 
%great insights into what parts of the optimizers should be examined for their improvements. 
%However, this scientific approach has been much less prominent in database field while very stong 
%mathematical and engineering work has been done. 
%For instance, there has been no system of allowing database researchers to 
%simultaneously run experiments on different DBMSes and seamlessly analyze the results. 
%To support this scientific, empirical study on DBMSes, in this paper we introduce a novel laboratory information system, 
%called {\em Arizona Database Laboratory} ($\azdb$). 
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\section{Introduction}\label{sec:intro}
%Understanding database query optimizers is of critical importance in a sense that it can provide 
%great insights into what parts of the optimizers should be examined for their improvements. 
%Also, it can ultimately derive predictive theories about how optimizers as a general class behave, which will be very 
%useful in the enhanced design of query optimizers. 
%Despite these strengths, this scientific approach has been much less prominent in database field while very stong 
%mathematical and engineering work has been done. 
%For instance, there has been no system of allowing database researchers to 
%simultaneously run experiments on different DBMSes and seamlessly analyze the results. 
%To support this scientific, empirical study on DBMSes, we introduce a novel laboratory information system, 
%called {\em Arizona Database Laboratory} ($\azdb$). 
%In database field, scientific approach has been much less prominent while very strong 
%engineering and mathematical work has been done. 
In the database field, while very strong mathematical and engineering work has been done, the 
scientific approach has been much less prominent. 
Much work has focused on \linebreak \hbox{proposing} new algorithms for optimizing DBMS performance 
and on building system components for new needs, but the \hbox{community} has not devoted 
much attention on \hbox{scientifically} understanding \hbox{DBMS} as an {\em experiment subject}. 
\shorten{\hbox{Indeed}, understanding database query optimizers better is of critical importance, 
as it can provide insights into their improvements by identifying which parts should be re-examined. 
Also, it can ultimately derive predictive theories about the database optimizer behavior, 
thereby leading to their enhanced designs.} 
The deep understanding of query optimizers obtained through the \linebreak \hbox{scientific} approach 
can lead to better engineered designs.

There, however, have been few {\em DBMS-dedicated} \linebreak \hbox{laboratories}
for supporting such \hbox{scientific} \hbox{investigation}, while prior work mainly has focused on 
%networks ({\it e.g.,} 
%\hbox{PlanetLab}~\cite{Peterson2006planetlab}\shorten{\footnote{http://www.planet-lab.org/}}), 
%\hbox{sensors} ({\it e.g.,} 
%\hbox{MoteLab}~\cite{WA05MoteLab}\shorten{\footnote{http://motelab.eecs.harvard.edu/}}), 
%\hbox{smartphones} ({\it e.g.,} 
%\hbox{SmartLab}~\cite{Larkou13SmartLab}\shorten{\footnote{http://smartlab.cs.ucy.ac.cy/}}, 
%\hbox{PhoneLab}~\cite{Nandugudi13PhoneLab}\shorten{\footnote{http://www.phone-lab.org/}}, 
%\hbox{CrowdLab}~\cite{cuervo2011crowdlab}), 
%{\it etc}. 
\hbox{networks} and smartphones as we will discuss in Section~\ref{sec:related_work}.

In this demonstration, we present a novel DBMS-oriented \hbox{research} infrastructure, 
called {\em Arizona Database Laboratory} ($\azdb$), to assist database researchers 
to conduct a {\em large-scale} empirical study across {\em multiple} \hbox{DBMSes}\shorten{ in a {\em scientific} manner}.

$\azdb$ presents to a scientist an electronic lab \linebreak \hbox{notebook}. 
Within this environment, the user can \hbox{design} massive \hbox{experiments} with \hbox{thousands} (or millions) of queries. 
%(Running TPC benchmarks within $\azdb$ is not yet considered, but it will be possible in the future.)
The user can then begin and monitor many runs of the \hbox{experiments} \shorten{tens (or hundreds or more)} 
on different \hbox{DBMSes} over a long period of time.

In addition, $\azdb$ provides an integrated and robust database experiment environment. 
In $\azdb$, a user can record both independent variables (those controlled by the \hbox{experimenter}) 
and dependent variables (those \hbox{resulting} from the \hbox{experiment}) and perform analyses on the \linebreak \hbox{experiment} data. 
For instance, the user can easily \hbox{combine} the \hbox{completed} runs in a study and then analyze in an \linebreak \hbox{automated} way for the study the query execution (QE) \linebreak \hbox{results}. 
\shorten{Note that every query at a given cardinality gets executed 10 times, and each QE \hbox{result} contains not only QE-specific \hbox{information} (query plan and measured time) but also the \hbox{compressed} detailed \hbox{information} of \hbox{approximately} two \hbox{hundred} processes per QE in those runs. (The raw QE \hbox{result} size is about 568KB, but when compressed, it is \hbox{reduced} to \hbox{approximately} 32KB.)}
In addition, \hbox{$\azdb$} has been sufficiently \hbox{robust} to collect data over 8,277 hours (\hbox{almost} a year) 
running about 2.4 million query executions.

%$\azdb$ consists of about 60K source lines of code and currently supports experiments on seven \hbox{DBMSes}, 
%four commercial and three open-source. 
%To the scientist, $\azdb$ presents an electronic laboratory notebook. 
%Within this environment, one can design and specify experiments, 
%schedule these experiments, monitor them when they are running, 
%record both independent variables (those controlled by the experimenter) and 
%dependent variables (those \hbox{resulting} from the experiment) and perform analyses on the data. 
%\hbox{$\azdb$} runs on a hardware lab of dedicated machines, 
%one each for each subject DBMS and one to host the DBMS that stores the lab notebooks, called {\em labshelves}. 
%%Having dedicated hardware allows us to worry less about other processes running on the machines 
%%that could dirty the results (these machines are only for experiments), and allows us to run extensive 
%%experiments taking days or weeks. 
%Having dedicated hardware provides a clear environment and allows us to run extensive experiments taking days or weeks.

The design and implementation of $\azdb$ has been very challenging. 
$\azdb$ has been developed for seven years by many people, each contributing to different pieces of code. 
\shorten{As of March 20, 2014,} $\azdb$ has reached 60K source lines of code. 
\hbox{$\azdb$} runs on a hardware lab of dedicated machines, 
one each for each subject DBMS and one to host the DBMS that stores the lab notebooks on a {\em labshelf}. 
%Having dedicated hardware allows us to worry less about other processes running on the machines 
%that could dirty the results (these machines are only for experiments), and allows us to run extensive 
%experiments taking days or weeks. 
%Having dedicated \hbox{hardware} provides a clear environment and allows us to run extensive experiments taking days or weeks. 
Having \hbox{dedicated} \hbox{hardware} provides a clear environment for \hbox{running} large-scale, extensive experiments taking days, weeks, or months.
A total of seven relational \hbox{DBMSes} (four \hbox{commercial} and three open-source) supporting 
SQL and JDBC are \hbox{currently} integrated into $\azdb$
We will plug more \hbox{DBMSes} into $\azdb$ for more solid scientific \hbox{studies}. 
\shorten{We should say that considering that the DBMSes have different \hbox{implementations} and \hbox{distinct} \hbox{characteristics}, 
it was never easy for us to plug in these \hbox{DBMSes} to and fully work with $\azdb$.}

The key contributions of this demonstration are as follows.

\begin{itemize}
\item  We present a novel research infrastructure, $\azdb$, dedicated for a large-scale scientific DBMS study. 
\linebreak $\azdb$ provides a rigorous \hbox{abstraction} for \linebreak \hbox{empirically} \hbox{studying} across
multiple \hbox{DBMSes} \shorten{in} from a \linebreak \hbox{variety} of \hbox{perspectives} (termed {\em database ergalics}~\cite{ergalics})\shorten{, like the other laboratories actively operating in \hbox{different} domains}.

%\vspace\fill

\item $\azdb$ supports seamless {\em data provenance} \linebreak \hbox{collection} within an \hbox{empirical} DBMS study. 
A \linebreak database \hbox{researcher} can \hbox{design} and run a substantial \hbox{experiment} with many queries, 
see the query \hbox{execution} \hbox{results}, perform data \hbox{sanity} check and \hbox{analysis} and 
make tables, figures, and graphs for the study in an \hbox{automated}, \hbox{integrated} \hbox{fashion}. 
Note that the data provenance of the study is collected into a \hbox{labshelf}, 
\hbox{managed} by a \hbox{central} DBMS server. %Therefore, a user can easily keep track of how data is produced through $\azdb$. 
%$\azdb$ knows which version of a protocol is used for the analysis.

%\item $\azdb$ supports running experiments across multiple \hbox{DBMSes} (termed {\em database ergalics}). 
%The same or different experiments can be simultaneously run on different \hbox{DBMSes} within $\azdb$. 
\item For conducting large-scale experiments, $\azdb$ \linebreak \hbox{provides} several {\em decentralized \hbox{monitoring}} \linebreak schemes: 
a stand-alone Java application (named \linebreak {\em \hbox{Observer}})\shorten{(running on a separate machine not \hbox{participating} in experiments)}, 
an {\em Ajax}~\cite{ajax} web app, and a mobile app.
\shorten{Also, she can be notified via {\em watcher} when an error or an exception occurs during experiment. 
This frees her from watching the experiments at all times.}

\item $\azdb$ provides a reusable GUI architecture. \linebreak \hbox{Observer}'s GUI consists of the {\em tree nodes} in the left-hand side and the {\em tabbed viewers} in the right-hand side. The tree nodes represent \hbox{experiment} data, while the corresponding tabbed viewers provide the detailed \hbox{information} on the selected data. 
We have adapted the \hbox{GUI} to several diverse projects.

\item $\azdb$ provides rich {\em extensibility}. A variety of \hbox{{\em plugins}} provides $\azdb$ 
with various analytics and \hbox{different} views of the collected query execution data.

%\item We introduce a novel {\em laboratory information system}, called {\em Arizona Database Laboratory} ($\azdb$), for supporting empirical studies on a variety of DBMSes.

%\item We present various {\em plugin}s enriching the extensibility of $\azdb$.

%\item We propose a {\em robust} methodology of running an experiment in $\azdb$.

%\item We present the automated result analysis process incorporated into $\azdb$.

\end{itemize}

\section{Motivation}\label{sec:motivation}

There are questions concerning fundamental limits of \linebreak \hbox{DBMS} 
architectures that simply cannot be answered by \hbox{investigating} a single \hbox{algorithm} or even a single DBMS. 
Rather, \hbox{addressing} such questions requires the development of {\em \hbox{predictive} models across multiple \hbox{DBMSes}}.

The objective behind this {\em scientific} approach is to \linebreak \hbox{understand} \hbox{DBMSes} 
as a {\em general} class of computational \hbox{artifacts}, 
to come up with insights and ultimately with \linebreak \hbox{predictive} models 
about how such systems, again, as a \linebreak \hbox{general} class, behave by studying multiple \hbox{DBMSes} at a time. 
These models are articulated and thoroughly tested, in order to \hbox{understand} more deeply the behavior of query \hbox{optimization} and \hbox{evaluation} across multiple \hbox{DBMSes}. 
They can be \hbox{eventually} used to improve \hbox{DBMSes} through \linebreak \hbox{engineering} efforts 
that benefit from the fundamental \linebreak \hbox{understanding} by this \hbox{perspective}.

%$\azdb$ has been designed and developed over seven years to help us 
%reach this overarching goal: to {\em \hbox{predict} important characteristics} of \hbox{DBMSes} to determine {\em fundamental limits}.}  
$\azdb$ has been around over seven years to help us achieve this overarching goal: to {\em predict important 
\linebreak \hbox{characteristics}} of \hbox{DBMSes} to determine \hbox{{\em fundamental}} \linebreak \hbox{{\em limits}}. 
$\azdb$ allows us to perform substantial \linebreak \hbox{experiments} (with thousands or more of queries) that \linebreak \hbox{quantitatively} study these fundamental questions \hbox{concerning} many of the \hbox{components} of a \hbox{DBMS}. 
These cover (i) \linebreak \hbox{{\em cardinality}} \hbox{{\em estimation}} (identifying what affects the accuracy of \hbox{cardinality} \hbox{estimates}), 
(ii) {\em operator impact} (\hbox{characterizing} how specific types of \hbox{operators}, {\it e.g.}, join, projection, \hbox{sorting}, 
\hbox{affect} the \hbox{accuracy} of cardinality \hbox{estimates}, \hbox{execution} time estimates, and \hbox{optimal} plan selection), and 
(iii) {\em \hbox{execution} plan search space} (determining its detailed inner \hbox{structure}).

As an \hbox{experiment} tool, $\azdb$ coordinates the \linebreak \hbox{running}, data \hbox{collection} and analysis of such large-scale \linebreak \hbox{experiments}. 
It has been broadened to support \hbox{experiment} \hbox{replication}, \hbox{provenance} maintenance, more variety of 
%\linebreak \hbox{experiment} \hbox{scenarios} (to be discussed in Section~\ref{sec:observer}), and more \hbox{DBMSes}. 
\linebreak \hbox{experiment} \hbox{scenarios}, and more \hbox{DBMSes}.

\section{Related Work}\label{sec:related_work}

Although many labs have been established and widely used, their application domains are different from \hbox{DBMSes}. \hbox{PlanetLab}~\cite{Peterson2006planetlab} 
is known as one of the most \hbox{popular} \linebreak \hbox{platforms} for network research. It allows people to 
deploy and assess network service~\cite{Peterson2006planetlab}. 
\hbox{MoteLab}~\cite{WA05MoteLab} was a \hbox{public} 
\hbox{testbed} for wireless sensor network researchers. 
Its \hbox{service} was stopped in January 2014.

There are also available \hbox{mobile} computing research labs: \hbox{CrowdLab}~\cite{cuervo2011crowdlab}, \hbox{SmartLab}~\cite{Larkou13SmartLab}, \shorten{\hbox{PhoneLab}~\cite{Nandugudi13PhoneLab},} etc. %%% should etc be in italic?
None of these labs are, however, \hbox{{\em DBMS-centric}}.\shorten{, and there has been few laboratories dedicated for empirically studying across DBMSes.}

One of the closest tools related to $\azdb$ is Picasso~\cite{Reddy05}.
The Picasso tool can also help study query execution plan choices of (three) 
different \hbox{DBMSes}' cost-based query \linebreak \hbox{optimizers}, by generating plan and cost diagrams for \hbox{TPC-H} queries. 
Both Picasso and $\azdb$ systems use the \linebreak ``{\tt EXPLAIN}'' \hbox{command}, to see the execution plan of a query submitted to \hbox{DBMSes}. 
%However, unlike $\azdb$, that system has no functionalities of actually executing queries and collecting the query execution provenance for further analysis. 
Unlike $\azdb$, the Picasso tool neither actually runs queries nor collects the query \hbox{execution} provenance for further \hbox{analysis}.

Using $\azdb$, our empirical \hbox{DBMS} study 
has yielded two structural causal models; one that \hbox{identifies} across four \hbox{DBMSes} some of the causes of varying query time \hbox{measures}~\cite{Currim} and one that identifies several of the causal factors of \linebreak \hbox{suboptimality}, when the DBMSes choose a wrong plan. Also, we are currently working on developing another structural causal model to explain DBMS thrashing~\cite{Thomasian93}.
\shorten{We plan to add several more \hbox{commercial} \hbox{DBMSes} and open source \hbox{DBMSes} in the future.}

%In the next section, we will present the overall \hbox{architectural} components of $\azdb$. 
%We end with an outline of the demo.

%The scope of this investigation will be significantly broadened to consider many of the components of
%a modern DBMS, including (i) cardinality estimation (identifying what affects the accuracy of 
%cardinality estimates), (ii) operator impact (characterizing how specific types of operators-e.g., joins, projection, 
%sorting—affect the accuracy of cardinality estimates, execution time estimates, and optimal plan selection), 
%(iii) execution plan search space (determining its detailed inner structure), and (iv) transaction and 
%concurrency control (specifying how throughput and disk utilization depend on multiprogramming level, how the 
%presence of hot spots affect this interaction, and when thrashing occurs).

\section{AZDBLab System Overview}\label{sec:architecture} 
Figure~\ref{fig:azdblab_arch} presents the architecture of $\azdb$. 
In this section, we describe each component of $\azdb$.

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.34]{./figures/azdblab_arch}
%\vspace{-0.3in}
\caption{$\azdb$ architecture~\label{fig:azdblab_arch}}
%\vspace{-0.2in}
\end{figure}

\subsection{LabShelves}
A labshelf comprehensively stores all the {\em data provenance} \hbox{related} to experiments. 
\shorten{It is created along with a {\em versioned-schema} complying with the {\em 7-W model}~\cite{Ram}.} 
It is a fully append-only database~\cite{Snodgrass99}.
The schema of a labshelf \hbox{captures} {\em who}, {\em what}, {\em when}, {\em which}, {\em where}, {\em why}, and {\em how}, \hbox{complying} with the {\em 7-W model}~\cite{Ram}. 
%That said, there is provenance that the system does not yet capture. 
The labshelf schema has been evolving for collecting and \hbox{analyzing} a \hbox{variety} of data relevant to QE.
The labshelf data are currently managed by a dedicated DBMS server \hbox{running} on a separate machine.  
\shorten{separated from machines hosting \hbox{experiment} \hbox{subjects}.}

%\vspace\fill

\begin{figure*}[htp!]
\centering
\includegraphics[scale=0.3]{./figures/observer_gui_new}
\caption{$\azdb$ Observer\label{fig:observer_gui}}
\vspace{-0.1in}
\end{figure*}

\subsection{Decentralized Monitoring Schemes}
In this section, we present a variety of novel, decentralized monitoring schemes being in use in $\azdb$.
%These monitoring tools help a user manage and monitor experiments on a remote site.

\subsubsection{Observer}\label{sec:observer}
%Observer provides a database researcher with GUI for loading, scheduling, and running an experiment, and later analyzing 
%the completed experiment results. It is a stand-alone Java application.
Observer is an integrated \hbox{GUI} for \hbox{conducting} experiments involving a large number of queries and \hbox{analyzing} the \hbox{results}. 
The GUI is implemented with Java Swing API. When \linebreak \hbox{launching} \hbox{Observer}, a user can see the main \hbox{GUI} in \hbox{Figure}~\ref{fig:observer_gui}.

%When launching Observer, he is asked to choose a labshelf. 
%Observer then connects to the labshelf server to load and expose to him the data stored in the chosen labshelf.
%The data include a labshelf user(s), notebook(s), experiment(s), pending, running, paused, and aborted experiment run(s), 
%query execution results in the experiment run(s), and executor(s) currently in use.
%An experiment run is an instance of an experiment assigned to a chosen DBMS started at a specific time.

%\begin{figure}[htp!]
%\centering
%\includegraphics[scale=0.18]{./figures/observer_gui_new}
%\caption{$\azdb$ observer\label{fig:observer_gui}}
%\end{figure}

%For instance, under `{\tt Users}' node, there are two labshelf users, 
%named `{\tt Sabah}' and `{\tt yksuh}', created in the labshelf. 
%User `yksuh' has created `{\tt SIGMOD2014}' notebook, under which 
%there are two experiment scenarios, named `{\tt helloworldquery}' and `{\tt onepass}'.
%A scenario defines what an experiment concerns. 
%For example, in the {\tt onepass} scenario, while varying the cardinalities, 
%if we detect different query plans, we execute the same query at adjacent cardinalities (apart from 10K).}s:`{\tt exhaustive}' and `{\tt onepass}'.

The \hbox{Observer} \hbox{GUI} has a {\em tree-structured} form. 
A left-hand tree node is associated with data \hbox{corresponding} to \hbox{experiment} provenance. 
For instance, `{\tt Users}' node \hbox{contains} users in a labshelf selected by the user. 
In the figure, \hbox{under} the `{\tt Users}' node, we can see a \hbox{labshelf} user (node), named `{\tt yksuh}'\shorten{, 
\hbox{under} which there exists a \hbox{notebook} node, named `{\tt VLDB2014}', which \hbox{represents} a lab \hbox{notebook} on the \hbox{labshelf}}. 
In the right-hand side, more detailed information on a \hbox{chosen} left-hand node is \hbox{provided} in tabbed views. 
For the selected plan node (`{\tt Plan \# 2}') to be discussed shortly, 
the \hbox{corresponding}, expanded plan tree is exhibited with \linebreak \hbox{various} {\em operator-wise} cost estimates, as shown in Figure~\ref{fig:observer_gui}.

%In the figure, for instance, under `{\tt Users}' node, there is one labshelf user, named `{\tt yksuh}', created in the labshelf. 
%User `{\tt yksuh}' has created `{\tt SIGMOD2014}' notebook, under which there are two scenarios, named `{\tt helloworldquery}' and `{\tt onepass}'.
%A scenario defines what an experiment is concerned about and how the experiment should be conducted. 
%For instance, the \linebreak `{\tt onepass}' scenario studies the suboptimality of a set of {\em select}-{\em project}-{\em join} queries 
%on one to four tables, each consisting of four integer columns. 
%Three of the tables are called {\em fixed} tables and have 1M rows each, while 
%the other table is called {\em variable}.
%%In the scenario, while decrementing the cardinality of the variable table from 2M rows to 10K the cardinality of the variable table by 10K, 
%In the scenario, while decrementing the cardinality of the variable table from 2M rows to 10K rows by decrements of 10K, 
%if the plan of a query being studied is changed between adjacent cardinalities of the table, 
%we execute the query at both cardinalities. Later, we check if the execution time at a lower cardinality 
%is greater than that of a higher cardinality, and if so, the query is reported as {\em suboptimal}.

This GUI allows the user to proceed with running a \linebreak {\em scenario-based} \hbox{experiment}. 
A scenario specifies {\em what an \hbox{experiment} is \hbox{concerned} about} and {\em what specific steps should be executed}. 
To run the scenario on $\azdb$, the user first needs to write the corresponding code in Java. 
For example, a \hbox{scenario} source code, called `{\tt onepass}', can be written to study the query \hbox{{\em suboptimality}} \hbox{phenomenon} such that when the \hbox{execution} plans of a query change between two adjacent \hbox{cardinalities} (called a {\em change point}), the actual elapsed time of that query at a lower \hbox{cardinality} is greater than that of a higher \hbox{cardinality}. 
\shorten{\linebreak `{\tt onepass}' is named after our strategy of decrementing the cardinality only in one direction from the maximum to the mininum.)} 
Once the \hbox{scenario} code is ready, it can be brought as a {\em plugin} into $\azdb$.

%In the running scenario, he can define four tables, each consisting of four integer columns. 
%Three of the tables are called {\em fixed} tables and have 1M rows each, while the other table is called {\em variable} and initially has 2M rows. 
%While decrementing the cardinality of the variable table by 10K, 
%if the plan of a query being studied is changed between adjacent cardinalities of the table, 
%the sceario executes the query at both cardinalities. Later, he can check if the execution time at a lower cardinality 
%is greater than that of a higher cardinality. If so, the query is reported as {\em suboptimal}.

%In Figure~\ref{fig:observer_gui}, under the `{\tt onepass}' scenario on the \hbox{left-hand} side, there are two loaded 
%experiments, named `{\tt op-1M-10q-1}' and `{\tt op-1M-1q-1}', each of which utilizes the same scenario but has different parameters. 
%In the names, we used the convention where `{\tt op}' stands for the `{\tt onepass}' scenario, 
%`{\tt 10q}' or `{\tt 1q}' refers to the number of queries in the experiments, respectively. 
%`{\tt 1M}' denotes the base cardinality of a table used in the experiment. 
%Lastly, `{\tt 1}' indicates the query set number.
%%Under `{\tt Completed Runs}', two {\tt onepass} runs are finished on DBMS {\tt X}\footnote{The license of the commercial system forbids us to disclose it.}. 
%Under `{\tt Completed Runs}', two {\tt onepass} runs have finished on DBMS {\tt X}. 
%While studying query \#0 ({\tt SELECT t0.id3, t1.id2 FROM ft\_HT3 t3, ft\_HT1 t1, ft\_HT3 t2, ft\_HT1 t0  WHERE (t3.id3=t1.id2 AND t1.id2=t2.id1 AND t2.id1 \linebreak =t0.id4}), 
%the scenario reported a total of six change points, and $\azdb$ recorded 
%each query plan and its associated execution time (although not shown in the figure). 
%In the right hand side, we show one of its query plan trees, denoted as `{\tt Plan \# 2}', 
%with each operator's estimates collected from DBMS {\tt X}.

Besides, the user needs to write and load into $\azdb$ a simple XML experiment specification, which 
contains \linebreak \hbox{scenario} name, schema definition, \hbox{table} population \linebreak \hbox{direction}, query details, etc. 
Figure~\ref{fig:observer_gui} shows several \linebreak \hbox{experiments} such as `{\tt op-1M-1q-1}' under the `{\tt onepass}' node.

\shorten{the `{\tt onepass}' scenario node under `{\tt VLDB2014}'}
%`{\tt op}' stands for the `{\tt onepass}' scenario. 
%`{\tt 1q}' refer to the number of queries to be run for the experiments. 
%`{\tt 1M}' denotes the base cardinality of tables in the experiments, and 
%`{\tt 1}' indicates a certain query set number.

The user can then schedule any of these experiments on a specific \hbox{DBMS}. 
This scheduled experiment instance is termed as `{\em experiment run}' and its status is in `{\em pending}'. 
\hbox{Observer} updates its GUI to show the pending run.

If an \hbox{executor}, to be discussed in Section~\ref{sec:executor}, has any assigned \hbox{pending} run,
the executor starts to execute that pending run, whose status then is updated to `{\em running}'. 
As the run proceeds, the user can monitor the run's \hbox{status} through this GUI\shorten{as well as through 
the other schemes to be described in next sections}. 
Note that the user can also {\em pause} and {\em \hbox{resume}} the \hbox{running} run if necessary. 
Once the run is \linebreak \hbox{completed}, the user can check the experiment results of the run. 
In \hbox{Figure}~\ref{fig:observer_gui}, two completed onepass runs of \hbox{DBMS} {\tt X} and 
\hbox{PostgreSQL} \hbox{DBMS} are illustrated. 
\shorten{Regarding the completed run on \hbox{DBMS} {\tt X}, we can see that an \hbox{executor} stored a total of six change points while running query \#0\shorten{({\tt SELECT t0.id3, t1.id2 FROM ft\_HT3 t3, ft\_HT1 t1, ft\_HT3 t2, ft\_HT1 t0 WHERE (t3.id3=t1.id2 AND t1.id2=t2.id1 AND t2.id1=t0.id4})} over varying \hbox{cardinalities} of {\tt ft\_HT1} table on \hbox{DBMS} {\tt X}. 
Among the observed plans of query \#0, the `{\tt Plan \# 2}' tree is shown in \hbox{Figure}~\ref{fig:observer_gui}.}
\shorten{For each cardinality of change points, the executor recorded the execution plan of query \#0, ran the query, and measured and recorded into $\azdb$ the elapsed time with collected process information (that will be used for further analysis), which is not shown in the figure.}
For another study, we have implemented a new scenario and an analysis plugin to observe a totally separate 
phenomenon, thrashing~\cite{Thomasian93}.

%
%In addition, under `{\tt Running Runs}' nodes on the left-hand side of the figure, 
%we can monitor the six runs concurrently executing on \hbox{DBMSes} {\tt X}, {\tt Y}, {\tt Z}, and {\tt W} 
%on machines {\tt sodb6}, {\tt sodb8}, {\tt sodb9}, and {\tt sodb12}, as well as PostgreSQL and MySQL on machines {\tt sodb10} and {\tt sodb11}. 
%($\azdb$ disallows more than two runs to be running at a time on any one physical machine.) 
%Also, under `{\tt Paused Runs}' node, there is one run paused on MySQL on machines {\tt sodb12}. 
%Lastly, the GUI shows aborted runs, loaded various plugins, running executors, and paused executors, if any.

%Assume that the login user created a labshelf user (`Young-Kyoon'). 
%The labshelf user has a notebook (`SIGMOD$\_$2014'), under which he has run the two experiment {\em scenario}
%\footnote{A scenario directs how an experiment on a DBMS should be conducted. 
%The experiment contains a set of SPJ queries concerning from one to four tables, each 
%consisting of four integer columns. 
%One of the tables is called {\em variable} table,
%which is initially populated with two times more rows than the specified cardinality (e.g., 30K) 
%and incrementally decreased at a certain granularity (20K). 
%The other ones called {\em fixed} one, which has its fixed cardinality (e.g., 30K). 
%In the {\tt exhaustive} scenario each query is run at every cardinality while varying the cardinality of 
%the variable table (i.e. 60K) at the granularity. 
%In the {\tt onepass} scenario, while varying the cardinalities, 
%if we detect different query plans, we execute the same query at adjacent cardinalities (apart from 10K).}s:`{\tt exhaustive}' and `{\tt onepass}'. 
%Due to lack of space, Figure~\ref{fig:observer_gui} shows only the completed {\tt exhaustive} experiment (named `eh-30K-2q-1') 
%run of 2 queries on Microsoft SQL Server. 
%Specifically, in the right-hand side the query 1's plan tree is shown when the variable table's cardinality was 40K; 
%in the query plan tree, we can see each query operators' estimates as well. 
%In addition, under `Running Runs' nodes, we can monitor the five experiment runs currently running on each different DBMS, 
%and under `Paused/Aborted Runs' nodes, we can also see one run paused on IBM DB2, and two aborted runs on MySQL and oracle, respectively. 
%
%`Loaded Plugins' shows different type of plugins used in $\azdb$. 
%`Experiment Subject Plugins' represents different DBMSes, treated as {\em experiment subject}, 
%`Scenario Plugins' node includes different experiment scenarios, and lastly, 
%`Evaluation Plugins' node contains a variety of evaluation plugins for further analysis of completed runs. 
%
%Finally, `Executors' node shows different executors running on the same machine where a DBMS gets installed. 
%We will shortly discuss executors in more details.

%\subsubsection{Web Application - Ajax}
\subsubsection{Web Apps}
$\azdb$ also provides a web app using Ajax~\cite{ajax}. 
A user can access the web app running on an $\azdb$ web server 
and make a request through a web browser. 
The web app then invokes an \hbox{AjaxManager} (a Java class), 
which interacts with the \hbox{labshelf} DBMS server to pull the requested data in a labshelf.
The server then responds to the user with the data.
%Finally, the page will be dynamically updated to display the received data as done in the Observer GUI. 
%In this way, a user can launch new experiments and subsequently monitor how they are going through the web application. 
%Due to lack of space, we do not show its snapshot, but the audience will be able to 
%watch experiment runs through Ajax application. 
The web app \hbox{provides} the same functionalities and GUI as \hbox{Observer}, 
without requiring direct access to the labshelf server, thereby achieving greater security.
%achieving greater security\shorten{, 
%without requiring direct access to the labshelf DBMS server, thereby achieving greater security}.

\subsubsection{Mobile Apps}
%\todo{Ask Ben.} 
To more flexibly monitor the executing runs we also make mobile apps available for the user.
\shorten{is also available for a user.}
We have built the mobile apps on both \hbox{Android} and \hbox{iOS}.

The mobile apps provide simple functionalities, compared to \hbox{Observer}. 
%The \hbox{functionalities} of the apps are simpler than those of \hbox{Observer}. 
The user cannot conduct query execution data \hbox{analysis} through the mobile apps, 
but the user can set up a convenient \hbox{monitoring} environment on the mobile apps.

%The user can select a \hbox{labshelf} and view \hbox{currently} \hbox{pending}, running, and paused runs, 
%as well as \hbox{executors} \hbox{being} used for the labshelf, via the apps.

The mobile apps also make a request to the same \linebreak 
$\azdb$ web server, 
which invokes methods from \linebreak \hbox{AjaxManager}. 
As mentioned earlier, \hbox{AjaxManager} then connects to the labshelf server and 
retrieves the data \linebreak \hbox{corresponding} to the user's request regarding 
user \hbox{logins}, and \hbox{experiment} and \hbox{executor} \hbox{statuses}.

% via an \hbox{API} written in \hbox{JSP} and \hbox{Java}

%The mobile apps through the web server can connect to the same labshelf DBMS server as used in \hbox{Observer}, 
%to retrieve \hbox{information} about user \hbox{logins}, \hbox{experiment} and \hbox{executor} \hbox{statuses}.

%To respond to the user's request, 
%the corresponding \hbox{AjaxManager}'s method is invoked to \hbox{retrieve} \hbox{information} about user 
%\hbox{logins}, \hbox{experiment} and \hbox{executor} \hbox{statuses}.
%However, the apps allow the user to select a labshelf and view \hbox{currently} pending, running, and paused runs 
%as well as \hbox{executors} \hbox{being} used for this labshelf. 
%which in turn calls methods from \hbox{AjaxManager} to \hbox{retrieve} \hbox{information} about user 
%\hbox{logins}, \hbox{experiment} and \hbox{executor} \hbox{statuses}. 
%Communication happens with the server via \texttt{GET} and \texttt{PUT} \texttt{http} requests.

\begin{figure}[htp!]
	\centering
	\subfigure[Main]{
		\includegraphics[scale=0.17]{./figures/main}
		\label{fig:main}
	}
	\subfigure[Exp. Runs]{
		\includegraphics[scale=0.17]{./figures/experiment_run}
		\label{fig:exp_runs}
	}
	\subfigure[A running run]{
		\includegraphics[scale=0.17]{./figures/run_description}
		\label{fig:exp_details}
	}
	\vspace{-0.1in}
	\caption{$\azdb$ mobile app\label{fig:mobile_app}}
	%\vspace{-0.1in}
\end{figure}

Figure~\ref{fig:mobile_app} illustrates the $\azdb$ mobile (iOS) app. \linebreak
When a user starts the mobile app, the user is presented with a login screen, into which the user's credentials must be provided. 
%The credentials are validated through the JSP API. 
%User logins are stored on the server in XML files (one per user login) 
%that contains the user name, password hash, and superuser status.
Once the credentials are validated, the user can see the main screen consisting of four menus, as shown in \hbox{Figure}~\ref{fig:main}. 
`{\tt Experiment Runs}' lists all the runs in a \hbox{chosen} labshelf, as illustrated in \hbox{Figure}~\ref{fig:exp_runs}. 
%(The runs are the ones as shown in Figure~\ref{fig:observer_gui}.)
An item in blue \hbox{indicates} a currently running run, one in white a \hbox{pending} run, and one in gray a paused run. 
%To get a full list of information about an experiment, he can choose one from the list, 
%and a full detail of the experiment will be brought up as a new page, as illustrated in Figure~\ref{fig:exp_details}. 
A running run on \hbox{PostreSQL} is exhibited in Figure~\ref{fig:exp_details}.
%This browsing and viewing process is essentially the same for viewing executors. 
The `{\tt Executors}' menu provides currently running \hbox{executors}, the `{\tt Observer}' menu shows 
the same view as the web app's one, and the `{\tt About}' menu provides a description of $\azdb$.

%In order to allow this application to run on multiple mobile OSs with minimal effort, 
%we built it using the PhoneGap development library. Using PhoneGap, 
%we were able to write the majority of the software using platform-agnostic web standards 
%such as HTML and CSS for the UI and Javascript and JQuery for the networking and other general logic. 
%We have tested the application on both iOS and Android, 
%and we plan on testing in on the Windows Phone OS in the future.

%\subsubsection{Watcher}
%%\todo{Ask Jen.}
%Watcher is a utility program used for watching an executor's behaviors during an experiment. 
%It frees a user from keeping an eye at all times on how the experiment goes. 
%Watcher immediately sends a user via email the notification messages when we encounter the following situations 
%where 1) an executor gets stopped, 2) a DBMS unexpectedly shuts down, 3) a running experiment gets paused by an exception, and 
%4) the labshelf server is not responsive.

%Watcher sends a user the notification via his email, so that the user can go to the terminal of the machine and 
%fix the problem. Currently, it reports

\subsection{Executor}\label{sec:executor}
%An executor is a stand-alone \hbox{Java} \hbox{application} 
%and \hbox{conducts} an experiment on a \hbox{DBMS}, which is co-located with the \hbox{executor}, as illustrated in Figure~\ref{fig:azdblab_arch}. 
%It is a stand-alone Java application as observer, and it works with a chosen experiment subject plugin.
%The executor utilizes a \hbox{DBMS} experiment \hbox{subject} plugin. 
An executor \hbox{conducts} an experiment on a co-located \linebreak \hbox{DBMS}, as illustrated in Figure~\ref{fig:azdblab_arch}. 
The executor is a stand-alone Java application, utilizing a \hbox{DBMS} experiment \hbox{subject} plugin. 
A user can launch an executor and schedule a \linebreak \hbox{pending} run to the executor. 
The executor can begin the \hbox{experiment} by \hbox{loading} the run's scenario and experiment subject \hbox{plugins}. 
The \hbox{executor} then creates and populates tables, executes queries, records QE results into $\azdb$, and finishes the run, 
as explained in \hbox{Section~\ref{sec:observer}}. 
If any \hbox{exception} occurs during the \hbox{experiment}, the executor can pause the run. Later, the user can unpause the run. 
\hbox{Multiple} executors can be in action, perhaps using distinct instances of the same \hbox{DBMS}, 
each on a different \hbox{machine}. 
To avoid \linebreak \hbox{undesirable} latency by network traffic, we ensure that the executor must run on the same machine hosting a \hbox{DBMS}. 
%measures query time, and records into $\azdb$ the measured time with collected process information.

%When the researcher launches an executor, the executor interacts with the labshelf server every five seconds, 
%to check if there is any pending experiment run assigned to the executor as mentioned in Section~\ref{fig:observer_gui}. 
%If so, the executor starts the experiment; it loads necessary plugins like onepass scenario or experiment subject, 
%creates and populates tables along with the experiment specification, executes a query on the tables, 
%measures query time, and records into $\azdb$ the measured time with collected process information. 
%During the experiment, if the executor encounters any exception occurs, the executor will pause the ongoing experiment run.

%\vspace\fill

\section{Demonstration}\label{sec:demo_overview} 
Our demo consists of two parts: 1) running experiments with hundreds of queries on different \hbox{DBMSes} and 2) then analyzing QE 
results from the completed runs. 
%We describe what steps the audience will do.

%\subsection{Running an experiment}
%The goal of this demonstration is to show how an experiment can be run and managed in $\azdb$. 
The goal of the first part of this demo is to show how a database \hbox{researcher} can schedule and run a large-scale experiment in $\azdb$. 
The steps are as follows.

%\pagebreak

{\bf Step 1}: An auditor launches \hbox{Observer} and selects a \linebreak \hbox{labshelf}. 
The user then creates a \hbox{labshelf} user and the user's lab notebook. 
In turn, the user loads into the user's \hbox{notebook} an XML experiment specification guiding table population and referring to queries.

{\bf Step 2}: The user selects a \hbox{DBMS}, schedules a run of the \hbox{experiment} on the \hbox{DBMS} and then launches an executor for that DBMS. 
After that, the user can see on console that the experiment gets started on the DBMS. 
The user may add other \hbox{DBMSes} in order to run the same experiment.

{\bf Step 3}: The user monitors the run's status via \hbox{Observer} (or the \hbox{iOS} app). At the end, the user will see the run done.

%\subsection{Analyzing experiment results}
In the second part of this demo, the audience will see how the user can conduct a study on the completed runs. %Here are the demo steps:

{\bf Step 4}: The user creates a {\em paper} under her notebook and the study under that paper in the \hbox{Observer} \hbox{GUI}.

{\bf Step 5}: For the study, the user chooses the completed runs via dialog box and 
executes \hbox{Tucson} \hbox{Timing} \hbox{Protocol} (\hbox{TTP}) on the runs. 
The automated protocol performs a \hbox{series} of {\em sanity checks} on QEs of the runs, shows \hbox{validation} results, and calculates query time on the passed QEs.

%{\bf Step 3}: The user finally produces a paper based on the analyzed results. 
{\bf Step 6}: The user finally produces a \hbox{PDF} document \linebreak \hbox{containing} the protocol analysis results. 
The user can also \hbox{create} graphs showing the estimated costs of a plan operator (e.g., hash join) over increasing cardinalities. 
The analysis results and graphs can be used for producing the paper.

\shorten{In this demo, the audience will see the comprehensive provenance of the data used by, and generated for a study.} 
\shorten{Thanks to this feature, a researcher does not have to worry about the data source of the graphs or tables included in the study.}
%For instance, a graph can be included in a paper, but one may not know where its data source come from. 
%$\azdb$ keeps every provenance of where data comes
%
%The data source of the graphs can be found in $\azdb$, so that a user does not have to worry about where 
%his data come from.

%In general, the analysis on the experiment runs can include graphs. $\azdb$ helps link a specific study 
%to 
%Since $\azdb$ stores 
%This demo will show how seamlessly $\azdb$ manages provenance of data used in an analysis for a paper. 
%the provenance of where data come from, 
%
% concerns multiple experiment runs. From these runs 
%The graphs are generated based on the data obtained through experiment runs. 
%
%used for a paper.
%
%The graphs will visualize the estimated cost of a chosen plan operator 
%(like hash join) over increasing cardinality. 
%
%In general, 
\shorten{
\section{Summary}\label{sec:summary}
In this demo, we present $\azdb$, a {\em DBMS-oriented} lab for large-scale empirical studies across multiple \hbox{DBMSes}. 
\shorten{$\azdb$ gives database researchers full \hbox{support} for managing a large-scale experiment with hundreds of queries and automatically analyzing and visualizing the query execution results.}
\shorten{$\azdb$ allows database researchers to run experiments simultaneously on the \hbox{DBMSes} and 
%automatically collects all data provenance regarding the experiments for further analysis. 
automatically collect all provenance regarding the experiments for further analysis. 
Our demo presents the audience with an actual experiment scenario on \hbox{DBMSes} running on our dedicated machines 
%and shows the audience how the experiment can be monitored, and then the results can be analyzed in $\azdb$. 
and shows the audience how the experiment can be monitored and the results can be analyzed in $\azdb$.}
}

%\vspace{-0.3ex}
\vspace{-1.2ex}

\section{Acknowledgement}
We thank to T.~Buchanan, S.~Currim, B.~Dicken, \linebreak M.~\hbox{Johnson}, P.~Kaslo, A.~Kvochko, T.~Lowry, 
H.~Zuniga, and others for their contributions to $\azdb$.

\vspace{-0.9ex}
%\vspace{\vfill{*}}

\bibliographystyle{abbrv}
\newcommand{\etalchar}[1]{$^{#1}$}
%{\small
\begin{thebibliography}{99}
\vspace{0.1em}
\bibitem
{ajax}
W3C, ``XMLHttpRequest'', January 2014.

\bibitem
{ergalics}
R. T. Snodgrass, “Database Ergalics”, \url{http://cs.}\linebreak\url{arizona.edu/~rts/ergalics}, viewed Mar 28, 2014.

\bibitem
{Currim}
S.~Currim, R. T. Snodgrass, Y-K. Suh, R.~Zhang, M.~Johnson, and C.~Yi, ``DBMS Metrology: Measuring Query Time,'' in {\em SIGMOD}, 
pp.~261--272, 2013.

%\bibitem
%{Currim2}
%S.~Currim, et. al., ``A Causal Model of DBMS Suboptimality,'' in {\em Proceedings of the ACM SIGMOD conference},
%In submission, 2014.

\bibitem
{WA05MoteLab}
G.~Werner-Allen, P.~Swieskowski, and M.~Welsh, ``MoteLab: A Wireless Sensor Network Testbed,'' in {\em IPSN}, pp.~483--488, 2005.
%
\shorten{\bibitem
{Nandugudi13PhoneLab}
A.~Nandugudi, A.~Maiti, T.~Ki, F.~Bulut, M.~Demirbas, T.~Kosar, C.~Qiao, S.~Y.~Ko, and G.~Challen, ``PhoneLab: A Large Programmable Smartphone Testbed,'' in {\em SENSEMINE}, pp.~1--6, 2013.}
%
\bibitem
{Larkou13SmartLab}
G.~Larkou, C.~Costa, P.~G.~Andreou, A.~Konstantinidis, and D.~Zeinalipour-Yazti, ``Managing Smartphone Testbeds with SmartLab,'' in {\em LISA}, pp.~115--132, 2013.
%
\bibitem
{Peterson2006planetlab}
L.~Peterson, A.~Bavier, M.~E.~Fiuczynski, and S.~Muir, ``Experiences Building PlanetLab,'' in {\em OSDI}, pp.~351--366, 2006.
%
\bibitem
{cuervo2011crowdlab}
E.~Cuervo, P.~Gilbert, B.~Wu, and O.~P.~Cox, ``\hbox{Crowdlab:} An Architecture for Volunteer Mobile Testbeds,'' in {\em COMSNETS}, pp.~1--10, 2011.
%
\bibitem
{Ram}
S.~Ram and J.~Liu, ``Understanding the Semantics of Data Provenance to Support Active Conceptual Modeling,'' in {\em ACML}, pp.~1--12, 2006.
%
\bibitem 
{Snodgrass99}
R. T. Snodgrass, {\bf Developing Time-Oriented Database Applications in SQL}, \shorten{Morgan 
Kaufmann Publishers, Inc., San Francisco, CA,} July 1999\shorten{, 504+xxiv pages}.
%
\bibitem
{Reddy05}
R.~Naveen and J.~R.~Haritsa, ``Analyzing Plan Diagrams of Database Query Optimizers,'' in {\em VLDB}, 
pp.~1228--1239, 2005.

\bibitem
{Thomasian93}
A.~Thomasian, ``Two-Phase Locking Performance and Its Thrashing Behavior,'' in {\em ACM TODS}, 
pp.~579--625, 1993.

\end{thebibliography}
%}
\end{document}

### 2017-02-16 azdblab-spe17/template.tex
azdblab she paper

\documentclass[conference]{IEEEtran}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{epsfig}
\usepackage{minted}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}

\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\newcommand{\INDSTATE}[1][1]{\STATE\hspace{#1\algorithmicindent}}
\long\def\todo#1{\hilitey{{\bf TODO:} {\em #1}}}
\long\def\shorten#1{}
\def\azdb{\hbox{\sc AZDBLab}}
%\def\azdb{\hbox{\sc DBLab}}
\def\QatC{Q{@}C}
\long\def\comment#1{}
%c2j: Conference to Journal: first parameter is conference, second is journal
% For Conferences: c2j#1#2{#1}
% For Journals: c2j#1#2{2}
\long\def\c2j#1#2{#1}

\begin{document}
%don't want date printed
\date{}

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{{\sc AZDBLab} : A Laboratory Information System for Empirical Studies on DBMSes}
%
\author{
DRAFT
%\IEEEauthorblockN{Young-Kyoon Suh, Richard T. Snodgrass}
%\IEEEauthorblockA{{\it Dept. of Computer Science}\\
%{\it University of Arizona}\\
%{\it Tucson, Arizona, USA, 85721}\\
%{\tt \{yksuh, rts\}@cs.arizona.edu}}
}

\maketitle

\begin{abstract} 
%Understanding a variety of behaviors present in DBMSes 
%can give an importance implication database query optimizers to improve. 
%There has been no system to study these aspects across different DBMSes. 
%This scientific approach leads to fundamental improvements 
%of database query optimizers. 
%To address this ,
%There has been no system of allowing database researchers to 
%simultaneously run experiments on different DBMSes and seamlessly analyze the results, 
%to better understand a variety of behaviours of database query optimizers. 
In database field, scientific approach has been much less prominent while very stong 
mathematical and engineering work has been done. 
Understanding database query optimizers is of critical importance in a sense that it can provide 
great insights into their improvements by looming which parts should be reexamined. 
However, there has been no system for supporting this scientific approach in which one can 
simultaneously run and monitor a variety of experiments on DBMSes and analyze the 
results to test his/her hypotheses. 
In this paper, we introduce a novel laboratory information system, 
called {\em Arizona Database Laboratory} ({\sc AZDBLab}), of allowing database researchers 
to empirically study DBMSes. 
%
%Understanding database query optimizers is of critical importance in a sense that it can provide 
%great insights into what parts of the optimizers should be examined for their improvements. 
%However, this scientific approach has been much less prominent in database field while very stong 
%mathematical and engineering work has been done. 
%For instance, there has been no system of allowing database researchers to 
%simultaneously run experiments on different DBMSes and seamlessly analyze the results. 
%To support this scientific, empirical study on DBMSes, in this paper we introduce a novel laboratory information system, 
%called {\em Arizona Database Laboratory} ({\sc AZDBLab}). 
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\section{Introduction}\label{sec:intro}
%Understanding database query optimizers is of critical importance in a sense that it can provide 
%great insights into what parts of the optimizers should be examined for their improvements. 
%Also, it can ultimately derive predictive theories about how optimizers as a general class behave, which will be very 
%useful in the enhanced design of query optimizers. 
%Despite these strengths, this scientific approach has been much less prominent in database field while very stong 
%mathematical and engineering work has been done. 
%For instance, there has been no system of allowing database researchers to 
%simultaneously run experiments on different DBMSes and seamlessly analyze the results. 
%To support this scientific, empirical study on DBMSes, we introduce a novel laboratory information system, 
%called {\em Arizona Database Laboratory} ({\sc AZDBLab}). 
In database field, scientific approach has been much less prominent while very stong 
mathematical and engineering work has been done. 
Understanding database query optimizers is of critical importance in a sense that it can provide 
great insights into their improvements by looming which parts should be reexamined. 
Also, it can ultimately derive predictive theories about the database optimizer behavior, thereby leading to their 
enhanced design. 
However, there has been no system for supporting this scientific approach in which one can 
simultaneously run and monitor a variety of experiments on DBMSes and analyze the 
results to test his/her hypotheses. 
In this paper, we introduce a novel laboratory information system, 
called {\em Arizona Database Laboratory} ({\sc AZDBLab}), of 
allowing database researchers to empirically study DBMSes.

{\sc AZDBLab} consists of 45K source lines of code, and currently supports experiments on five DBMSes, 
three commercial and two open-source. 
 {\sc AZDBLab} is effectively an electronic laboratory notebook. 
Within this environment, one can design and specify experiments, 
schedule these experiments, monitor them when they are running, 
record both independent variables (those controlled by the experimenter) and 
dependent variables (those resulting from the experiment), and do analyses on the data. 
{\sc AZDBLab} runs on a hardware lab of dedicated machines, 
one each for each subject DBMS and one to host the DBMS that stores the lab notebooks, called {\em labshelves}. 
Having dedicated hardware allows us to worry less about other processes running on the machines 
that could dirty the results (these machines are only for experiments), and allows us to run extensive 
experiments taking days or weeks.

The key contributions of this paper are as follows.
\begin{itemize}
\item We introduce a novel {\em laboratory information system}, called {\em Arizona Database Laboratory} ({\sc AZDBLab}), 
for supporting empirical studies on a variety of DBMSes.

\item We present various {\em plugin}s enriching the extensibility of {\sc AZDBLab}.

\item We propose a {\em robust} methodology of running an experiment in {\sc AZDBLab}.

\item We present the automated result analysis process incorporated into {\sc AZDBLab}.

\end{itemize}

\section{Motivation}\label{sec:motivation} 
There are questions concerning fundamental limits of database management system (DBMS) 
architectures that simply cannot be answered by investigating a single algorithm or even a single DBMS. 
Rather, addressing such questions requires the development of {\em predictive models across multiple DBMSes}.

The objective is to understand database management systems as a {\em general} class of computational artifacts, 
to come up with insights and ultimately with predictive models about how such systems, again, as a 
general class, behave, by studying multiple DBMSes at a time. 
These models are articulated and thoroughly tested, in order to 
understand more deeply the behavior of query optimization and evaluation across multiple DBMSes. 
They can be eventually used to improve DBMSes through engineering efforts 
that benefit from the fundamental understanding that this perspective can provide.

$\azdb$ has been designed and developed over seven years, to help us 
reach this overarching goal - to {\em predict important characteristics} of DBMSes to determine {\em fundamental limits}.  
$\azdb$ allows us to perform experiments for quantitatively studying these fundamental questions concerning 
many of the components\footnote{These cover (i) {\em cardinality estimation} (identifying what affects the accuracy of cardinality estimates), 
(ii) {\em operator impact} (characterizing how specific types of operators - e.g., {\em joins}, {\em projection}, 
{\em sorting} affect the accuracy of cardinality estimates, execution time estimates, and optimal plan selection),
(iii) execution plan search space (determining its detailed inner structure)} of a DBMS. 
As an experiment tool, $\azdb$ coordinates the running, data collection and analysis of such experiments. 
It will also be broadened to support experiment replication, provenance maintenance, a more variety of experiment 
scenarios (to be discussed), and more experiment subjects (equivalent to DBMSes). 
Thanks to this research infrastructure by $\azdb$, our empirical study on DBMSes 
has already yielded structural models that identifies across two commercial and two open source DBMSes 
some of the causes of varying query time measures\cite{Currim} and suboptimality (when the DBMS chooses a wrong plan).

In the next section, we will present the overall architectural components of $\azdb$.

%The scope of this investigation will be significantly broadened to consider many of the components of
%a modern DBMS, including (i) cardinality estimation (identifying what affects the accuracy of 
%cardinality estimates), (ii) operator impact (characterizing how specific types of operators-e.g., joins, projection, 
%sorting—affect the accuracy of cardinality estimates, execution time estimates, and optimal plan selection), 
%(iii) execution plan search space (determining its detailed inner structure), and (iv) transaction and 
%concurrency control (specifying how throughput and disk utilization depend on multiprogramming level, how the 
%presence of hot spots affect this interaction, and when thrashing occurs).

\section{Architecture}\label{sec:architecture}
As shown in Figure~\ref{fig:azdblab_arch}, $\azdb$ consists of {\em labshelves}, 
{\em executor},  {\em observer} and {\em plugins}. 
We describe each of the components of $\azdb$ in the next sections.

\begin{figure}[t]
\centering
\includegraphics[scale=0.4]{./figures/azdblab_arch}
\caption{{\sc AZDBLab} archiecture~\label{fig:azdblab_arch}}
\end{figure}

\subsection{LabShelves}
A labshelf abstracts experimental data. It retains a versioned-schema 
and complies with the {\em 7-W model}~\cite{Ram}. 
Namely, the schema for a labshelf captures {\em who}, {\em what}, {\em when}, {\em which}, {\em where}, 
{\em why}, and {\em how}. It is a fully append-only database~\cite{Snodgrass99}. 
That said, there is provenance 
that the system does not yet capture. The schema also evolves as we need to capture more data for our study. 
The labshelf data are stored and managed by a dedicated server, 
running separated from other $\azdb$ components.

\begin{figure*}[htp!]
\centering
\includegraphics[scale=0.33]{./figures/observer_gui}
\caption{{\sc AZDBLab} observer\label{fig:observer_gui}}
\end{figure*}

\subsection{Executor}
Executor is the $\azdb$ component that executes experiment queries 
and stores the query execution results into a labshelf. 
To avoid undesirable latency by network traffic, we ensure that the executor must be running 
on the same machine running a DBMS as an experiment subject.

\subsection{Observer}
Observer is the $\azdb$ component that provides a login user 
with GUI necessary to run an experiment and analyze the experiment results. 
When launching observer, the login user is asked to choose a labshelf. 
Observer then loads data from the chosen labshelf and expose them to 
the login user if any. The data, which can be also created by the login user, 
includes a labshelf user(s), a notebook(s), an experiment(s), 
an experiment run(s), and query execution results in the experiment run(s).

Figure~\ref{fig:observer_gui} illustrates the GUI screenshot of observer. 
The following nodes in the LHS of the GU are organized in a tree-structured way 
(although the list is flattened due to too deep nesting).

\begin{itemize}
	\item User Node

	\item Notebook Node

	\item Scenario Node

	\item Experiment Node

	\item Data Definition Node

	\item Query Generation Definition Node

	\item Completed Run Node

	\item Query Node

	\item Plan Node

	\item Papers Node

	\item Analysis Node

	\item Analytic Specification Node

	\item Aspect Specification Node

	\item Defined Queries Node

	\item Aspect Definitions Node

	\item Analytic Definitions Node

	\item Pending Run Node

	\item Running Run Node

	\item Paused Run Node

	\item Aborted Run Node

	\item Plugin Node

	\item Executor Node

\end{itemize}

\subsubsection{Decentralized Architecture}
Discuss the decentralized architecture of {\sc AZDBLab}. 
(Ben may describe this section.)

\begin{itemize}

\item Web Application

\item Phone Application
\end{itemize}

%% Describe the role of observer
%Observer provides a login user with GUI to run an experiment. 
%When launching observer, the login user is asked to select a labshelf. 
%Then, observer loads data stored in the chosen labshelf. 
%The data includes a labshelf user(s), a notebook(s), an experiment(s), an 
%experiment run(s), and query execution results in the experiment run(s). 
%The login user can also create a

\subsection{Plugins}
The most outstanding feature of {\sc AZDBLab} is its 
unlimited extensibility by a variety of plugins. 
In this section, we discuss how a plugin is loaded into {\sc AZDBLab} 
and present different types of plugins and their roles.

\subsubsection{Plugin Management}

Each plugin is developed as a {\em .jar} file. 
It is put into {\em plugins} subdirectory. 
When observer is launched, {\em master plugin manager} 
examines each plugin class in the plugin sub-directory 
where {\sc AZDBLab} is located.  Also, 
the manager classifies the identified along with 
its type to be discussed in the next section. 
%By preserving independence of the experiment
%% Describe a variety of plugins

\subsubsection{Plugin Type}

\begin{itemize}

\item Experiment Subject : {\sc AZDBLab} treats a DBMS as {\em experiment subject}.

\item Scenario : A {\em scenario} defines an experiment scenario. 
It has two kinds of subclasses as follows.

\begin{itemize}
\item {\em ScenarioBasedOnCardinality} defines a scenario taking each table cardinality as a unit of task 
\item {\em ScenarioBasedOnQuery} defines a scenario taking each query as a unit of task 
\end{itemize}

\item Evaluation : $\azdb$ provides {\em evaluation plugins} for analyzing query execution results. 
Currently, there are two types of evaluation plugins in $\azdb$.

\begin{itemize}
\item Query Stat (Evaluation) plugin parses the process information string (provided by Process Analyzer) and 
populates the per-query execution table and per-process (within the same query execution) information tables. 
The per-query execution table stores numbers of phantom/started/stopped processes, number of faults, and 
overall user/low priority/system mode, idle task, IO wait, and IRQ, Soft IRQ, and steal and stolen ticks.

%(\item Sanity Check (Evaluation) plugin performs sanity checks on completed query executions 
%and populates the violation results. The sanity checks include number of queries not executed, 
%excess
%)

\item Process Information (Evaluation) plugin extracts 
the name and frequency of processes observed at a completed run or at a notebook. 
This process information is referenced by Process Analyzer, which will be discussed soon, 
so that the most frequently executed processes will be checked first for minimizing 
increasing their ticks during process examination.

\end{itemize}

\end{itemize}

\subsubsection{Versioning}
Refer to {\sc AZDBLab} guides.

\subsection{Utilities}
In this section, we introduce a few utilities in $\azdb$.

\subsubsection{{\sc AZDBLab} Labshelf Creator}
Literally, {\sc AZDBLab} Labshelf Creator is responsible for creating all the tables under a specific labshelf version 
of {\sc AZDBLab}. It asks and encrypts labshelf name, password and JDBC connection string 
and then creates an XML file having the input data and lastly saves the XML file as a {\em plugin} in the {\em plugins} directory.

\subsubsection{Process Analyzer}
Process analyzer provides the information of all the processes running around query execution. 
Specifically, it compares running-process snapshots before and after query execution and 
records the difference into a {\em string} along with a query execution result. 
\todo{Refer to metrology paper.}

\subsubsection{Logger}
$\azdb$ logger extends {\tt Logger} of {\tt log4j}.  
As other loggers, its primary purpose is to record into a file important running events of observer or executor 
and to check what was happening when they got hung or killed.

\subsubsection{Watcher}
$\azdb$ watcher is designed to be immediately notified of when something bad happens while executor runs an experiment. 
\todo{Jen can write this subsection.}

\section{Running Experiments}\label{sec:running_exp} 
In this section, we describe how an experiment is written and run in $\azdb$. 
Figure~\ref{fig:exp_proc} depicts the overall experiment procedure in $\azdb$.  
An $\azdb$ user first needs to write his/her own (experiment) {\em scenario}. 
Along with the scenario, he should define how data should be populated 
(data population spec), how table cardinalities should vary (table cardinality variation spec), 
and how queries are generated (query generation spec). 
By collecting these specs, an experiment spec can be written. 
In turn, the experiment can be loaded and run through observer. Once the experiment is 
complete, he can analyze the experiment results.

\begin{figure}[t]
%\begin{tabbing}
%\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=
%\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\kill
\begin{center}
\begin{algorithmic}
{\bf Algorithm} $runExperiment$($labUser$, $labNoteBook$, $expName$, $dbms$) : \\
	\STATE $experiment$ $\leftarrow$ $getExperiment$($labUser$, $labNoteBook$, $expName$)\\
    \STATE $expRun$ $\leftarrow$ $getExperimentRun$($experiment$)\\
    \STATE $expSubject$ $\leftarrow$ $expRun$.$getExperimentSubject$($dbms$) \\
    \STATE $expSubject$.$dropAllInstalledTables$()\\
    \STATE $scenario$ $\leftarrow$ $expRun$.$getScenario$()\\ 
    \STATE $scenario$.$executeExperiment$()\\
\end{algorithmic}
\caption{Run an experiment\label{alg:run_exp}}
\end{center}
%\end{tabbing}
\end{figure}

\subsection{Scenario} 
A scenario specifies experiment details of what one wishes to observe from DBMSes. 
It consists of a series of steps, each of which determines its specific task.

\begin{figure}[t]
%\begin{tabbing}
%\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=
%\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\=\quad\quad\kill
\begin{center}
\begin{algorithmic}
{\bf Algorithm} $executeExperiment$() : \\
\STATE $wait$ $\leftarrow$ 1 \\
\STATE $fixTables$ $\leftarrow$ $expRun$.$getFixTables$() \\
\STATE $varTables$ $\leftarrow$ $expRun$.$getVariableTables$() \\
\STATE $populateFixTables$($fixTables$) \\
\STATE $queries$ $\leftarrow$ $expRun$.$getQueries$() \\
\STATE $cards$  $\leftarrow$ $expRun$.$getCardinalities$()\\
\IF{Scenario is based on query}
	\STATE $tasks$ $\leftarrow$ $queries$
\ELSE
	\STATE $tasks$ $\leftarrow$ $cards$
\ENDIF
\FOR{$t$ $\leftarrow$ 1 {\bf to} {$\vert$}$tasks${$\vert$}}
	\IF{$checkToBePaused$()}
		\STATE $pauseExperiment$()
	\ENDIF
	\WHILE{$wait$ $\leq$ 10}
		\STATE\hspace{-0.3cm} $try$ \{
	    \IF{This scenario is based on query}
			\STATE $analyzeQuery$($queries$[$t$], $cards$, $varTables$)
		\ELSE
			\STATE $analyzeCard$($queries$, $cards$[$t$], $varTables$)
		\ENDIF
		\STATE$break$ \\
		\STATE\hspace{-0.3cm}\}$catch$($Exception$ $ex$)\{
			\STATE $sleep$($wait$*1000)
			\STATE  $wait$ $\leftarrow$ $wait${$\times$}2
		  	\IF{$\log_{2}$($wait$) == 10}
		  		\STATE  $pauseExperiment$()
		  	\ENDIF
		\STATE\hspace{-0.3cm} \} 
	\ENDWHILE 	
\ENDFOR
\STATE $dropExperimentTables$($fixTables$, $varTables$)\\
\end{algorithmic}
\caption{Execute an experiment scenario\label{alg:scenario}}
\end{center}
%\end{tabbing}
\end{figure}

There is no limit of number of scenarios allowed in {\sc AZDBLab}, but 
we present the following three scenarios used in our analysis.

\subsubsection{UniquePlan} 
In {\tt UniquePlan} Scenario, we test if the same query plan 
is detected at the same cardinality. To achieve this, 
we first populate the variable-table with the maximum cardinality like 2M rows, 
and then delete the rows with a granularity of 10K until the table goes down to 
the minimum cardinality like 10K rows. 
Afer that, we get a query plan associated with the 10K cardinality. 
We repeat this process, and eventually compare two plans. 
If they are not consistent with each other, then the scenario reports an error.

\begin{figure}[t]
\begin{center}
\begin{algorithmic}
{\bf Algorithm} analyzeQuery($q$, $cards$, $varTables$): 
\STATE $initialize$($varTables$) \\
\FOR{$card$ $\leftarrow$ 2M to 10K by 10K in $cards$}
   \STATE $deleteRows$($card$) \\
\ENDFOR
\STATE $plan$1 = $getQueryPlan$($q$, $card$) \\ 
\FOR{$card$ $\leftarrow$ 2M to 10K by 10K in $cards$}
   \STATE $deleteRows$($card$) \\
\ENDFOR
\STATE $plan$2 = $getQueryPlan$($q$, $card$) \\ 
\IF{$plan$1 $\neq$ $plan$2}
	\STATE Report plan difference at $card$. \\
\ENDIF
\end{algorithmic}
\caption{ {\tt UniquePlan} scenario\label{alg:uniquplan}}
\end{center}
\end{figure}

\subsubsection{Exhaustive}
In this scenario, we explore the monotonicity of query execution time 
as the cardinality of a variable table increases. 
Specifically, we populate a variable table with the maximum cardinality like 2M rows, 
and then delete the rows with a granularity of 10K until the table goes down to 
the minimum cardinality like 10K rows. Whenever the cardinality changes, 
a query plan associated with the cardinality is obtained, and the query gets executed 10 times. 
Each query execution result is recorded into a labshelf. 
As we examine a query plan at every cardinality within the variation, 
we call this scenario {\em exhaustive}.

\begin{figure}[t]
\begin{center}
\begin{algorithmic}
{\bf Algorithm} analyzeCard($queries$, $card$, $varTables$): 
\STATE $initialize$($varTables$) \\
\FOR{$q$ $\in$ $queries$}
   \STATE $plan$ = $getQueryPlan$($q$, $card$) \\ 
   \FOR{$i$ $\leftarrow$ 1 {\bf to} 10}
         \STATE $timeSingleQueryExecution$($i$, $q$,  $plan$, $card$) \\
   \ENDFOR
\ENDFOR
\end{algorithmic}
\caption{ {\tt Exhaustive} scenario\label{alg:exhaustive}}
\end{center}
\end{figure}

\subsubsection{OnePass} 
In {\tt OnePass} Scenario, a query is studied by keeping track
of the {\em states} on the variable-table.
In other words, we locate different query plans at adjacent cardinalities
defined by the {\em even} or {\em odd} states,
and query executions are made at both of the states simultaneously,
while decreasing the variable-table cardinality in one pass.

Indeed, the adjacent cardinalities are not really called even and odd,
but they are next to each other.
They, thus, are simply separated by whether a given cardinality over
granularity is even or odd.
For instance, the even state will have 2M, 1.98M, 1.96M, $\cdots$, 1M cardinalities
in order, while the odd state will take 1.99M, 1.97M, $\cdots$, 1.01M cardinalities
(decremented by granularity (10K) from the even state's cardinality).

Each state initializes its table(s) same as the variable-table(s).
In addition, it has the {\em pre-compiled} query plan at its own cardinality
so that the given query can be executed by the {\em prepared} plan at the cardinality
without any compiling overhead.

While stepping down from the max to the min cardinality,
if different query plans are observed at the even and odd states,
then the prepared query execution at both states is made at the same time
unless the query has been executed at the cardinality before.
In this regard, for repeatability we time the same query up to 10 times.
The timing results at each state are inserted into the labshelf for
analysis.
For more details about timing single query execution, refer to the DBMS metrology paper~\cite{Currim}.

\begin{figure}[t]
\begin{center}
\begin{algorithmic}
{\bf Algorithm} analyzeQuery($q$, $cards$, $varTables$): 
\STATE $evenState$.$initialize$($varTables$) \\
\STATE $oddState$.$initialize$($varTables$) \\
\FOR{$card$ $\leftarrow$ 2M to 10K by 10K in $cards$}
        \IF{$\frac{card}{10K}$ is $even$}
                \STATE $evenState$.$card$ $\leftarrow$ $card$ \\
        \STATE $evenState$.$plan$ $\leftarrow$ \\
        		\hspace{14.0mm}$evenState$.$getPreparedQueryPlan$($q$) \\
    \ELSE
            \STATE $oddState$.$card$ $\leftarrow$ $card$ \\
            \STATE $oddState$.$plan$ $\leftarrow$\\
            \hspace{14.0mm} $oddState$.$getPreparedQueryPlan$($q$) \\
    \ENDIF
    \IF{($evenState$.$plan$ $\neq$ $oddState$.$plan$)}
                \IF{$\neg${$evenState$}.$isExecuted$()}
                        \FOR{$i$ $\leftarrow$ 1 {\bf to} 10}
                                \STATE $timeSingleQueryExecution$($i$, $q$,  \\
                                                \hspace{14.0mm}$evenState$.$plan$, $evenState$.$card$) \\
                        \ENDFOR
                \ENDIF
                \IF{$\neg${$oddState$}.$isExecuted$()}
                        \FOR{$i$ $\leftarrow$ 1 {\bf to} 10}
                                \STATE $timeSingleQueryExecution$($i$, $q$,  \\
                                                \hspace{14.0mm}$oddState$.$plan$, $oddState$.$card$) \\
                        \ENDFOR
                \ENDIF
        \ENDIF
\ENDFOR
\end{algorithmic}
\caption{ {\tt OnePass} scenario\label{alg:onepass}}
\end{center}
\end{figure}

%\begin{comment}
%\subsection{Summary}
%We first applied these scenarios to DBMS A through E with a few runs of 100 queries.
%As a result, DBMSes A and B showed the repeatability for all queries.
%In particular, they were not affected by which of the above scenarios
%was run or how many times the scenario was run.
%On the contrary, the other DBMSes C, D and E failed to retain repeatability.
%They were sensitive to queries, runs and scenarios.
%
%As a last check, we also took advantage of CLI (Command-Line Interface) on
%the unrepeatable DBMSes in order to see directly the 1M plan without passing
%through the scenarios.
%DBMS C showed the same plan at a cardinality for a query at all times.
%However, the other two DBMSes D and E produced different plans for the same
%query at times.
%
%The empirical observation about repeatability can be summarized as follows.
%Some heuristic algorithms may be exploited for the optimizers of DBMS
%D and E, which may affect the enumeration and decision of query plan.
%(To figure out this, we made several attempts to look for {\tt heuristic} keyword
%from the source code of DBMS D and E optimizers.
%However, it was not clear of where it was used.)
%DBMS C has relatively plenty of operators compared with other DBMSes, so
%its compatible operators may show up at different times, considering that plans
%built by such operators would have the same execution cost.
%In the meantime, DBMS A and B were repeatable.
%
%Based on the observation from all the DBMSes, we eventually ensure
%repeatability (and monotonicity) by using {\tt exhaustive} scenario.
%In {\tt exhaustive}, multiple query executions are made at each cardinality,
%varying the max cardinality to the min one; namely, for a query we get a plan
%and execute the query multiple times, saying 10, upon decreasing the cardinality.
%Instead of the incremental 10K-row-deletion,
%changing the cardinality is done by copying from the max table as many rows as needed
%so that all pages are in full with the rows.
%This setup allows us to check if any of the multiple executions produces
%different plans at the same cardinality (and see if the query execution
%times are monotonically decreasing as cardinality decreases).
%Each DBMS ran {\tt exhaustive} with a run of 10 queries, and it had
%no case such that different plans were detected at the same cardinality
%while a query got executed repeatedly.
%\end{comment}

%In this scenario, each query gets executed 10 times at adjacent cardinalities where different query plans are detected 
%while varying cardinalities. 
%Figure~\ref{fig:exp_proc} depicts the overall experiment procedure in $\azdb$.  
%Before starting the experiment, a $\azdb$ user 
%needs to write his/her own {\em scenario as discussed before. 
%Along with the scenario, he also needs to define how data should be 
%populated usin {\em data population spec}, how cardinalities are varied usin {\em cardinality variation spec}, 
%and whether queries should be generated or ones pre-defined using {\em query generation spec}. 
%Based on these specs in concert, he writes an experiment spec. 
%When observer is launched, he can loads the experiment spec and start to 
%run the experiment on any DBMS provided by $\azdb$. Once the experiment is completed, 
%he can do analysis on the experiment results.

%Describe the start-to-end experiment procedure using observer and executor. 
%We need some screen shots of showing pending, running, resuming, pausing runs 
%spawn from an experiment created under a notebook belonging to a user. 
%In this section, we discuss how an experiment is written and run in $\azdb$.

%\subsection{Experiment}
%Figure~\ref{fig:exp_proc} depicts the experiment procedure in $\azdb$.

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.4]{./figures/exp_proc}
\caption{Experiment and analysis procedure~\label{fig:exp_proc}}
\end{figure}

\subsection{Experiment Specification}
Table~\ref{tab:exp_spec} exemplifies an experiment specification.

\begin{table}[htp!]
\begin{center}
{
\scriptsize
\begin{tabular}{|p{8.5cm}|} \hline
\begin{minted}{xml}
<experiment name="op-1M-100q-1" scenario="onepass">
  <description>
    The baseline test of onepass scenario, with 100 
    predefined queries. Four experiment tables, 
    each with 1M rows. One variable table is used.
  </description>
  <dataDefinitionReference href="TestDataDef1M.xml"/>
  <tableConfiguration>
    <variableTableSet searchMethod="linear" 
    searchGranularity="10000">
      <table name="HT1" seed="1999">
        <cardinality hypotheticalMinimum="1"
         hypotheticalMaximum="2000000"/>
      </table>
    </variableTableSet>
    <fixedTableSet>
      <table name="HT2" seed="2999">
        <cardinality hypothetical="actual"/>
      </table>
      <table name="HT3" seed="3999">
        <cardinality hypothetical="actual"/>
      </table>
      <table name="HT4" seed="4999">
        <cardinality hypothetical="actual"/>
      </table>
    </fixedTableSet>
  </tableConfiguration>
   <queryDefinitionReference numberQueries="100" 
   type="predefinedQueries" href="100_q_1.xml"/>
</experiment>
\end{minted}
\\ \hline \end{tabular}
}
\caption{Example of an Experiment Specification~\label{tab:exp_spec}}
\end{center}
\end{table}

\paragraph{Data Definition}
Table~\ref{tab:data_spec} exemplifies an data population specification.

\begin{table}[htp!]
\begin{center}
{
\scriptsize
\begin{tabular}{|p{8.5cm}|} \hline
\begin{minted}{xml}
<dataDefinition name="ft">
 <documentation>
 A database that has tables with cardinalities 
 that are a power of 10.  From 1 to 1,000,000.  
 There are only numeric column data in these tables.
 </documentation>
 <table name="HT1" cardinality="1000000">
  <column name="id1" dataType="number" dataLength="20" 
  dataGenerationType="sequential" distributionMinimum="0" 
  distributionMaximum="1000000" inPrimaryKey="true"/>
  ...
  <column name="id4" dataType="number" dataLength="20" 
  dataGenerationType="sequential" distributionMinimum="0" 
  distributionMaximum="1000000" inPrimaryKey="false"/>
 </table>
 ...
 <table name="HT4" cardinality="1000000">
 ...
 </table>
</dataDefinition>
\end{minted}
\\ \hline \end{tabular}
}
\caption{Example of a Data Population Specification~\label{tab:data_spec}}
\end{center}
\end{table}

\paragraph{Query Definition}
Table~\ref{tab:query_spec} exhibits an query generation specification.

\begin{table}[htp!]
\begin{center}
{
\scriptsize
\begin{tabular}{|p{8.5cm}|} \hline
\begin{minted}{xml}
<queryDefinition numberQueries="100" 
javaClass="DefaultGrammarSpecificGenerator"
schemaFileName="/grammar.xsd">
    <grammar>
      <select maxColumns="4"/>
      <from maxNumCorrelationNames="4"/>
      <where cartesianPossible="false" maxPredicates="3">
        <binaryOperators>
          <operator symbol="="/>
        </binaryOperators>
        <binaryLogicalOperators>
          <operator symbol="AND"/>
        </binaryLogicalOperators>
        <unaryLogicalOperators>
          <operator symbol="NOT" usePercentage="0"/>
        </unaryLogicalOperators>
      </where>
      <aggregate function_name="SUM" percentage="50"/>
    </grammar>
  </queryDefinition>

\end{minted}
\\ \hline \end{tabular}
}
\caption{Example of an Query Generation Specification~\label{tab:query_spec}}
\end{center}
\end{table}

Table~\ref{tab:predefined_query_spec} shows a set of predefined queries.

\begin{table}[htp!]
\begin{center}
{
\scriptsize
\begin{tabular}{|p{8.7cm}|} \hline
\begin{minted}{xml}
<predefinedQueries name = "100_q_1" >
<query sql="SELECT t3.id3, t3.id4, t0.id4, t3.id1 
FROM ft_HT4 t1, ft_HT1 t0, ft_HT3 t3, ft_HT2 t2  
WHERE (t1.id1=t0.id1 AND t0.id1=t3.id2 AND t3.id2=t2.id4)"/>
...
<query sql="SELECT t0.id2, SUM(t2.id4)  
FROM ft_HT1 t0, ft_HT2 t3, ft_HT1 t2, ft_HT1 t1  
WHERE (t0.id3=t3.id3 AND t3.id3=t2.id4 AND t2.id4=t1.id1) 
GROUP BY t0.id2"/>
</predefinedQueries>
\end{minted}
\\ \hline \end{tabular}
}
\caption{Example of a Pre-defined Query Specification~\label{tab:predefined_query_spec}}
\end{center}
\end{table}

\subsection{Experiment Run}
Experiment 
Figure~\ref{fig:exp_run_state} depicts the state change of an experiment run.

\begin{figure}[htp!]
\centering
\includegraphics[scale=0.4]{./figures/exp_run_state}
\caption{Experiment run state diagram\label{fig:exp_run_state}}
\end{figure}

\section{Result Analysis}\label{sec:result_analysis}
Describe how Tucson protocol~\cite{Currim} is incorporated into {\sc AZDBLab}, 
and experiment results are passed to {\bf R} to do further analysis. 
The following items should be discussed in this section.

\begin{itemize}

\item Per-execution and per-process information extraction

\item (Discontinuity visualization)

\item Tucson protocol application (sanity check)
%is applied to refine our experiment results. The application procedure of the protocol are automatically

\item Correlation analysis using R

\end{itemize}

\section{Related Work}\label{sec:related}
Provide a list of systems related to {\sc AZDBLab}.

\section{Summary and Future Work}\label{sec:summary}
Conclude this paper and talk about what needs to be done in future.

\shorten{\section{Acknowledgments}
This research was supported in part by NSF grants\linebreak \hbox{IIS-0639106},
IIS-0415101, and EIA-0080123. We thank Ricardo Carlos, Preetha Chatterjee, Pallavi
Chilappagari, Jennifer Dempsey, David Gallup, Kevan Holdaway, Andrey
Kvochko, Siou Lin, Adam Robertson,
Lopamudra Sarangi, Linh Tran, Cheng Yi, and Man Zhang for
their contributions to the \hbox{\azdb} and Phil Kaslo, Tom Lowry, and John
Luiten for constructing and maintaining our experimental instrument\c2j{}{,
  a laboratory of six machines and associated software}.
}

%\vfill\pagebreak
%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.

% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\bibliographystyle{./IEEEtran,./IEEEabrv}
\newcommand{\etalchar}[1]{$^{#1}$}
\begin{thebibliography}{99}

\vspace{0.1em}
\bibitem
{Currim}
S.~Currim, et. al., ``DBMS Metrology: Measuring Query Time'', in {\em Proceedings of the ACM SIGMOD conference},
pp.~261--272, 2013.
\bibitem
{Ram}
S. Ram and J. Liu, ``Understanding the Semantics of Data Provenance to Support Active 
Conceptual Modeling'', in {Proceedings of the International Workshop on Active Conceptual
Modeling of Learning}, P. P. Chen and Y. Wong (eds), LNCS 4512, Springer Verlag, 2007.
\bibitem
{Snodgrass99}
R. T. Snodgrass, ``Developing Time-Oriented Database Applications in SQL'', Morgan
Kaufmann Publishers, Inc., San Francisco, CA, July 1999, 504+xxiv pages
\end{thebibliography}
\end{document}

### 2017-02-17 subopt-tods16/paper.tex
SubOpt paper updates
- Updated Tables IV, V
- Updated Fig. 10

\documentclass[prodmode,acmtods]{acmsmall}
\pdfminorversion=6

%Metadata Information
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2016}
\acmMonth{2}

\usepackage{url}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphics}
% \usepackage[total={6.5in,9in}, top=1.25in,left=0.9in]{geometry}
%\usepackage[text={6.5in,9in},centerpage,includefoot]{geometry}
% \usepackage{setspace}
\usepackage{color}
\usepackage{url}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{amssymb}

\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\long\def\todo#1{\hilitey{{\bf TODO:} {\em #1}}}
\long\def\shorten#1{}

%For last edit: revise and resubmit for SIGMOD'14
\long\def\shortenF#1#2{}%First is letter indicating what has been removed

%For double blind, keep the second argument. To turn off doubleblind, keep
%the first
\long\def\doubleblind#1#2{#2}
\def\azdb{\doubleblind{\hbox{\sc AZDBLab}}{\hbox{\sc DBLab}}}
\def\QatC{Q{@}C}
\long\def\comment#1{}
%c2j: Conference to Journal: first parameter is conference, second is journal
% For Conferences: c2j#1#2{#1}
% For Journals: c2j#1#2{2}
\long\def\c2j#1#2{#2}

\begin{document}

\title{%\vspace*{-0.75in}
A Causal Model of \hbox{DBMS} Suboptimality\\
}
\markboth{\ }{A Causal Model of \hbox{DBMS} Suboptimality}

% possible, but not really needed or used for PVLDB:
%\subtitle{[Extended Abstract]
%\titlenote{A full version of this paper is available as\textit{Author's Guide to Preparing ACM GIG Proceedings Using \LaTeX$2_\epsilon$\ and BibTeX} at \texttt{www.acm.org/eaddress.htm}}}

% ****************** AUTHORS **************************************

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

%bug is vldb.cls gives an additional authors section for greater than 3...
\doubleblind{\numberofauthors{3}}{}
 %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.

\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
\doubleblind{% 1st. author
Richard T.~Snodgrass\\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\
       \email{\small\tt rts@cs.arizona.edu}
\alignauthor
% 2nd. author
\alignauthor
Sabah Currim\\
       \affaddr{Alumni Office}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\
%       %\vspace*{-0.05in}
       \email{\small\tt scurrim@email.arizona.edu}
% 3rd. author
\alignauthor
Young-Kyoon Suh\\
       \affaddr{Dept. of Computer Science}\\
       \affaddr{University of Arizona}\\
%       \affaddr{Tucson, AZ}\\
       \email{\small\tt yksuh@cs.arizona.edu}
}{}}

\begin{abstract}
The query optimization phase within a \hbox{DBMS} ostensibly finds the fastest query
execution plan from a potentially large set of enumerated plans, all of
which correctly compute the specified query. Infrequently the optimizer
selects the wrong plan, for a variety of reasons. {\em Suboptimality} is
indicated by the existence of a query plan that performs more efficiently
than the DBMS' chosen plan, for the same query. From the engineering
perspective, it is of critical importance to understand the prevalence
of suboptimality and its causal factors.
We study these aspects across \hbox{DBMSes} to identify the
underlying causes. We propose a novel structural causal model to explicate the
relationship between various factors in query optimization and
suboptimality. Our model associates suboptimality with the factors of
complexity of the query and optimizer and concomitant unanticipated
interactions among the components of the optimizer. This model induces a number of specific
hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We observe that suboptimality prevalence
correlates positively with the number of operators available in the \hbox{DBMS}
(one proxy for optimizer complexity) and with the number of plans generated
by the optimizer (another proxy for optimizer complexity) and with
the number of correlation names (a measure of query complexity), providing
empirical support for this model as well as implications for fundamental
improvements of these optimizers. The results of this study also uncovered evidence of a potential fundamental
limitation of cost-based optimization.\todo{be more specific!}
This paper thus provides
a new methodology to study mature query optimizers, that of empirical
generalization, proposes a novel causal model for query suboptimality, and
tests several hypotheses deriving from this causal model.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]
\terms{}
\keywords{}
\acmformat{\doubleblind{Sabah Currim, Richard T.~Snodgrass, Young-Kyoon Suh, and
 Rui Zhang, 2015, \hbox{DBMS} Metrology: Measuring Query Time.}{A Causal Model of
    \hbox{DBMS} Suboptimality}}

\doubleblind{\begin{bottomstuff}
This work was supported in part by NSF grants \hbox{IIS-0639106},
IIS-0415101, and EIA-0080123. 
%
%Author's addresses: G. Zhou, Computer Science Department,
\end{bottomstuff}}{}
\maketitle

\section{Introduction}
DBMSes underlie all information systems and hence optimizing their performance is of
critical importance. The DBMS's query optimizer plays an important role. But
what if the optimizer {\em doesn't}: what if it selects the wrong plan?

This paper provides a thorough investigation into \hbox{DBMS} {\em suboptimality},
when the \hbox{DBMS} chooses a slower plan over a faster plan for a given query. We systematically examine the factors influencing the number of suboptimal
queries. There could be multiple causes of the suboptimality. One possible
cause could be some peculiarity within the tens of thousands of lines of
code of that query optimizer. Another possible cause could be the query's
complexity. A third possible reason could be some fundamental limitation
{\em within the general architecture of cost-based optimization} that will
always render a good number of queries suboptimal. Prior research in other domains shows that
increasing the complexity negatively influences
performance~\cite{campbell88,moody98}.

To better understand the impact of different factors on suboptimality of
query performance and the interaction between operators, especially in a
dynamic environment, an experimental approach is needed.
Our research introduces a novel approach
to better understand the factors influencing query performance in
\hbox{DBMSes}. Based on existing research and general knowledge of \hbox{DBMSes}, we
propose an innovative predictive model to understand the presence and influence
of suboptimality in query evaluation. We use an experimental methodology with
the \hbox{DBMS} as a subject to test our hypotheses with respect to factors
influencing suboptimality, applying to 
our experiment data collected over about a \hbox{cumulative} 30~months of query
executions (over 21,000 hours to run almost a million query
executions).\todo{Young: Table I says something different: over 16,000 hours
  ro run almost two million query executions which would correspond to two
  years continuous execution}
Our research falls within creative development of
new evaluation methods and metrics for artifacts, which were identified as
important design-science contributions~\cite{hevner04}.

\pagebreak
The key contributions of this paper are as follows.
\begin{itemize} 
\item We use an innovative {\em methodology} that treats \hbox{DBMSes} as experimental subjects.

\item We find that for a surprisingly large portion of queries, the plan
  chosen  by the query optimizer is not the best plan, for some cardinality,
  of the underlying tables.

\item We propose and test a {\em predictive model} for \hbox{DBMSes} to better understand the factors causing suboptimality.

\item \todo{Rick:Mention generational result}

\item The predictive model and these experimental results suggest several specific
  engineering directions.
\end{itemize}

This paper takes a scientifically rigorous approach to an
area previously dominated by the engineering perspective, that of database
query optimization.  Our goal is to \hbox{understand} cost-based query optimizers
as a {\em general} class of computational artifacts and to come up with
insights and ultimately with predictive theories about how such optimizers,
again, as a general class, behave.  These theories can be used to further
improve \hbox{DBMSes} through engineering \hbox{efforts} that benefit from the fundamental
understanding that the scientific perspective can provide.

One might ask, shouldn't the task of optimizing queries be left to DBAs?  In
databases, and especially in data warehouses, the number of users writing
and running queries has been growing exponentially. This growth is aided, in
part, by the drag and drop query tools provided by the different
systems. For example, subject areas allow business users to write queries
without knowing anything about the underlying database structure. This,
coupled with constantly growing data presents new challenges for tuning. For
example, \doubleblind{the data warehouse deployed at the University of
  Arizona}{a large data warehouse we are familiar with} runs about
30,000 queries on a daily basis. 
Also, tuning a subject area or tables for one group of queries can
negatively impact the performance of other queries. Query optimization experts often take hours to tune and test existing canned
queries; the amount expended on this one system for manual query optimization
approaches \$100K/year. Therefore, we argue that it is important to
understand how existing query optimizers can be further improved, and
indeed, fundamental limitations inherent in these optimizers.

\begin{figure*}[t]
\centering
%originally 30pc
%\epsfig{figure=figures/plan769.pdf,width=36pc}
\includegraphics[width=30pc]{figures/subopt_flutter.pdf}
%\includegraphics[scale=0.42]{figures/subopt_flutter.pdf}
\caption{An Example of Suboptimality and Fluttering\label{fig:query769}}
\end{figure*}

We focus here on the effectiveness of query
optimization. The query optimization phase within a \hbox{DBMS} ostensibly finds the fastest
query execution plan from a potentially large set of enumerated plans, all of
which correctly compute the specified query. Occasionally the \hbox{``optimizer''}
selects a suboptimal plan, for a variety of reasons. We define {\em
  suboptimality} as the existence of a query plan that performs more
efficiently than the plan chosen by the query optimizer. From the engineering
perspective, it is of critical importance to understand the phenomenon of
suboptimality.

We study these aspects \c2j{}{across \hbox{DBMSes}} to identify the
underlying causes. We have developed an initial predictive causal model that
identifies four 
factors that may play a role in suboptimality. The ultimate goal is to {\em
  understand} a component within a \hbox{DBMS}, its cost-based optimizer, through the
articulation and empirical testing of a general scientific theory.

In Section~\ref{sec:related} we briefly summarize the vast amount of related
work in query optimization to establish the technical basis for our study.
The following section introduces the methodology we will follow, that of
empirical generalization. We then present a predictive, causal model of
suboptimality and empirically test eight specific hypotheses derived from that
model\c2j{. These are the first scientific results that we are aware of that
  test a general causal model of \hbox{DBMSes}.}{ on three \hbox{DBMSes}.
  These are the first scientific results that we are aware of that apply
  {\em across} \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
  specific algorithm.} This model has implications for research in
engineering more efficient \hbox{DBMSes}.

\section{Motivation}\label{sec:motivation}

Consider a simple select-project-join (SPJ) query, with a few attributes in
the SELECT clause, a few tables referenced in the FROM clause,
and a few equality predicates in the WHERE clause. This query might be an
excerpt from a more complex query, with the tables being intermediate
results.

%%%% query info: runid: 8917, dbms: pgsql, querynum:13, exp: op-1M-100q-6
%\vspace*{-2ex}
\noindent
\hspace{3ex}{\small\begin{verbatim}
        SELECT t0.id1, t0.id2, t2.id4, t1.id1 
        FROM ft_HT3 t2, ft_HT2 t1, ft_HT1 t0 
        WHERE (t2.id4=t1.id1 AND t2.id1=t0.id1)
\end{verbatim}
}

%\vspace*{-2ex}
\noindent
The optimizer generates
different plans for this query as the cardinality of the {\tt
  ft\_HT1} table varies, an experiment that we will elaborate later in depth.

The upper graph in Figure~\ref{fig:query769} represents the plans chosen by
a common DBMS as the cardinality of {\tt FT\_HT1} decreases from 2M tuples
to 10K tuples in units of 10K tuples. The \hbox{x-axis} depicts the estimated
cardinality and the y-axis  a plan chosen for an interval of
cardinalities. So Plan~P0 was chosen for 2M tuples, switching to Plan~P1 at
1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through the
sequence P0, P1, P0, P1, P2, P0, P2, and finally P3 at 40,000 tuples.

The lower graph in Figure~\ref{fig:query769} indicates the query times
executed at adjacent cardinalities when the plan changed, termed the
``query-at-cardinality'' (Q@C) time. For some transitions, the Q@C time at
the larger cardinality was also larger, as expected. But for other
transitions, emphasized in red ovals, the Q@C time at the larger cardinality
was smaller, such as the transition from plan P1 at 1,650,000 to P0 at 1,640,000
tuples. Such pairs identify suboptimal plans. For the pair at 720,000
tuples, P0 required 2.35sec whereas P1 at a larger cardinality required only
2.41sec. This query exhibits seven plan change points, two of which are suboptimal.

This query also illustrates an interesting phenomenon, in which the
optimizer returns to an {\em earlier} plan. Sometimes the
optimizer starts oscillating between two plans, sometimes even switching
back and forth when the cardinality estimate changes by a small
percentage. The example query showed returning to P0 twice and to P1 and to
P2 each once.
We call this phenomenon, in which
the query optimizer returns to a previous plan,
``query optimizer flutter,'' or simply ``flutter.''

We have found through our experiments that flutter and suboptimality are all
around us: {\em every} \hbox{DBMS} that we have examined, including several
open source \hbox{DBMSes} and several proprietary
\hbox{DBMSes}, covering much of the installed base worldwide, exhibit these
phenomena, even when optimizing very simple queries. In the Confirmatory
Experiment described in detail in Section~\ref{sec:experiments}, we started
with 6,966 query instances (a query run on a specific \hbox{DBMS}) after our extensive query measurement
protocol~\doubleblind{\todo{ref SIGMOD'13 paper!}}{}. While about 20\%,
1,491, of these query instances contained only one
plan, a few of the other query instances switched plans at almost every change in cardinality (we varied the
cardinality in increments of 10K tuples, a total of 200 cardinalities): see
Figure~\ref{fig:planchanges}. Slightly over half, 4,093, exhibited
suboptimality somewhere along those 200 cardinalities; a few had
many changes to a plan that was in fact suboptimal, as indicated in
Figure~\ref{fig:suboptplanchanges}, across all four \hbox{DBMSes} considered.

\begin{figure}[h!]\centering
\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

\begin{figure}[t]\centering
\includegraphics[width=0.60\textwidth]{figures/zi_ncpq.pdf}
\caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
  number of changes to a {\em sub-optimal}
  plan\label{fig:suboptplanchanges}}
\vspace*{-2ex}
\end{figure}

One oft-stated observation is that the role of query optimization is not to
get the {\em best} plan, but rather to get a plan that is acceptably
good. Figure~\ref{fig:suboptcumulative} shows
the cumulative distribution of the relative amount of suboptimality (where
an $x$-value of 100 denotes that the query ran 100\% slower than the optimal
query, that is, twice as long). The good news is that 2,738 query instances,
or 67\% of the suboptimal queries, exhibited only a small degree of
suboptimality: less than 30\%. The challenge is that over fifth of all
queries (1,355) exhibited a significant amount of suboptimality ($>= 30\%$); the
relative slowdown can be quite large for some queries.

We started with 7,640 query instances (a particular query running on a
particular DBMS, cf.~Experiment 7 of Table~\ref{tab:run_stat}
in Section~\ref{sec:experiments}. After our protocol, we were left with
  6,966 query instances, of which 5,475 had at least one change point, so those are
  the ones we consider further. Of those, 1,382 had {\em no} suboptimality, so
4,093 (75\%) had some suboptimality.

1,951 (36\% of the query instances with a change point) have the higher cardinality
running at least 20\% faster than the lower cardinality. That means that
2,142 query instances (slightly over half of the suboptimal queries) were barely
suboptimal ($<$20\% slower) and about a third (1,355) were considerably
suboptimal (where the higher cardinality ran at least 30\% faster than the
lower cardinality).

\begin{figure}\centering
\includegraphics[width=0.60\textwidth]{figures/rel_diff.pdf}
\caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
  percentage relative suboptimality\label{fig:suboptcumulative}}
\end{figure}

We emphasize three important points. First, we used a sophisticated query
measurement methodology that reduces the measurement variance, so that the
query plans we identify as suboptimal definitely are so. Second, these
results are over four \hbox{DBMSes}, and thus, such phenomena are not dependent on
a particular implementation of cost-based optimization. Rather, they seem to
be common to {\em any} cost-based optimizer, independent of the specific
cardinality \hbox{estimation} or plan costing or plan enumeration algorithm or
implementation. Third, suboptimal plans are {\em not} the result of poor
coding or of inadequate algorithms. We view query optimization in modern
\hbox{DBMSes} as an engineering marvel, especially given the complexity of the SQL
language and the requirements and expectations of \hbox{DBMS} users, who often
demand that important queries simply not get slower with a new release of
the \hbox{DBMS}. Rather, the prevalence of suboptimality observed here is a
reflection of the complexity of the task of query optimization.

%\pagebreak
This is where the shift to a new methodology comes in. We want to understand
cost-based optimization deeply, in this case to determine if a fundamental
limitation exists, which means that we need to study multiple instances of
that optimization architecture, to achieve generalizable results. To do so,
we will articulate a predictive causal model for how
and in what circumstances suboptimality arises and will provide compelling
evidence via \hbox{hypothesis} testing that this model characterizes the
\hbox{behavior} of query optimizers in general. This is what is meant by ``empirical
generalization'' and why it is needed to answer such questions. In
  the long history of research in database query optimization, or even of
  databases in general, our model and its hypothesis tests are the first
  predictive results that we are aware of that apply {\em across}
  \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
  specific algorithm. This is due to the adoption of a new research
  methodology, that of empirical generalization. Such a causal model can
\hbox{provide} guidance to the community about where fundamental research is needed
and to the \hbox{DBMS} engineers about where to focus their efforts.

\section{Related Work}\label{sec:related}

SQL has emerged as the {\em de facto} and {\em de jure} standard relational
database query and update language. It emerged as an ANSI and ISO standard
in 1986--7, with a major extension and refinement as SQL-92~\cite{Melton93} 
and an even larger extension as SQL:1999~\cite{Melton03} and refinement as
SQL:2008~\cite{SQL2008} and SQL:2011.

There has been extensive work in query optimization over the last 40
years~\cite{Ioannidis96,Jarke84}. Query optimization and evaluation
proceeds in several general steps~\cite{Ramakrishnan03}. First, the SQL query is 
translated into alternative query evaluation plans based on the 
relational algebra via {\em query enumeration}. The cost
of each plan is then estimated and the plan with the lowest estimated cost
is chosen. These steps comprise query optimization, specifically {\em
cost-based query optimization}~\cite{Selinger}. The selected query plan is
then evaluated by the query execution engine which implements a set of
physical operators, often several for each logical operator such as
join~\cite{Graefe93}.

An influential survey~\cite{Chaudhuri98} does a superb job of
capturing the major themes that have pursued in the hundreds of articles
published on this general topic. Chaudhuri reviews the many techniques and
approaches that have been developed to represent the query plans, to
enumerate equivalent query plans, to handle some of the more complex lexical
constructs of SQL, to statistically summarize the base data, and to compute the
cost of evaluation plans. He also mentions some of the theoretical work
(which is much less prevalent) to understand the computational complexity of
these algorithms. Most of this research may be classified as adopting an
engineering perspective: how can we architect a query optimizer ``where 
(1)~the search space includes plans that have {\em low cost} (2)~the costing
technique is {\em accurate} (3)~the enumeration algorithm is {\em
  efficient}. Each of these three tasks is nontrivial and that is why
building a good optimizer is an enormous undertaking.'' \cite[page~35]{Chaudhuri98}

To determine the best query access plan, the cost model estimates the
execution time of each plan. There is a vibrant literature on this
subject~\hbox{\cite{Ioannidis03,Mannino88}}, including proposals for
histograms, sampling, and parametric methods. Again, most of these papers
are engineering studies, providing new techniques that improve on the
state-of-the-art through increased accuracy or performance. There have also
been a few mathematical results, such as ``the task of estimating distinct
values is {\em provably} error prone, i.e., for any estimation scheme, there
exists a database where the error is significant''~\cite{Chaudhuri98}.

An optimizer for a language like SQL must contend with a huge search space of
complex queries. Its first objective must be {\em correctness}: that the
resulting query evaluation plan produce the correct result for the
query. A secondary but clearly very important objective is {\em efficiency}; after
all, that is the raison d'etre for this phase. As is well known, the name
for this phase is an exaggeration, as existing optimizers do not produce
provably optimal plans. That said, the
query optimizers of prominent \hbox{DBMSes} generally do a superb job of producing
the best query evaluation plan for most queries. This performance is the
result of a fruitful collaboration between the research community and
developers.

Early investigation of plan suboptimality resulted in approaches such as
dynamic query-reoptimization~\cite{Avnur,kabra98}, which exploit more
accurate runtime statistics that appear while a query is being executed, to
steer in-flight plan reoptimization. The very presence of such a radical
change to the normal optimize-execute sequence indicates that plan
suboptimality was of interest to some researchers.

However, even with great effort over decades, optimizers as a general class
are still poorly understood. As has been observed, ``query optimization has
acquired the dubious reputation of being something of a black
art''~\cite{Babcock05}. DeWitt has gone farther, stating that ``query
optimizers [do] a terrible job of producing reliable, good plans [for
  complex queries] without a lot of hand
tuning''~\cite[page~59]{winslett02}. And as we will see, suboptimality is
possible even when considering only simple queries.

While this paper does not provide direct solutions to address
suboptimality, we envision that by adopting our proposed predictive model
for suboptimality, engineering practice, such as dynamic reoptimization just
mentioned may benefit. We elaborate on this subject in
Sections~\ref{sec:root} and \ref{sec:engineering}, where we discuss the engineering implications
of our causal model.

\section{A Model of Suboptimality}\label{sec:model}

The purpose of query optimization is to generate optimal plans.  So
why would suboptimality occur in the first place? Query optimizers are
highly complex, comprised of tens or hundreds of thousands of lines of code. There
are several reasons for this complexity. First, an optimizer must
contend with the richness of the SQL language, whose definition
requires about 2000 pages~\cite{SQL2008}, with a multiple of
linguistic features. Second, an optimizer must contend with the
\hbox{richness} of the physical operators available to it. \hbox{DBMSes} have a
range of algorithms available to evaluate each of many algebraic
operators. Third, the optimizer must contend with an exponential
number of query evaluation plans. Kabra and DeWitt~\cite{kabra98}
identify several other sources of complexity: inaccurate statistics on
the underlying tables and insufficient information about the runtime
system: ``amount of available resources (especially memory), the load
on the system, and the values of host language
variables.''~(page~106). They also mention user-defined data types,
methods, and operators allowed by the newer object-relational
systems~\cite{Melton03}. Thus, the task of optimization is very
complex, with the result that the optimizers themselves consist of a
collection of ``components,'' that is, the rules or heuristics that it
uses during optimization, with each of these components being itself
complex.

We wish to understand the causal factors of suboptimality, through a
predictive model that explicitly states the interactions between these
causal factors. We test this model through experiments over tens\shorten{14,700} of thousand
of queries and hundreds of thousands of query executions\shorten{2.4M}, showing that there
is strong support for this model. We then extract engineering implications
from the model, suggestions for the most productive places to look to reduce
suboptimality and thus to improve existing query optimizers.

\begin{figure*}[tb]
\centering
%originally 30pc
%\epsfig{figure=figures/model.pdf,width=30pc}
\includegraphics[width=30pc]{figures/subopt_model.pdf}
\caption{Predictive Model of Suboptimality\label{fig:model}}
\vspace*{3ex}
\end{figure*}

Here we examine the hypothesized influence that each independent
variable will have on the one dependent variable, query suboptimality (with
some of the the influences mediated by one of the constructs). In
the next section we will operationalize these variables, explaining how
each is controlled or measured.

\subsection{Constructs in the Model}

The model concerns five general
constructs that we \hbox{hypothesize} will play a role in suboptimality: optimizer
complexity, schema complexity, query complexity, data complexity, and plan space complexity.
Two constructs include several specific variables that contribute to that
construct as a whole. Our model distills many of the widely-held assumptions
about query optimization. Our contribution is the specific \hbox{structure} of
the model and the specific operationalization of the
\hbox{factors} included in the model.

Our theory, which we will investigate in some detail and which is the
basis for our predictive model, is that suboptimality is due in part
to the complexity of the optimizer and concomitant unanticipated
interactions between the different components used within the 
optimizer with various sources of complexity in producing a good (hopefully
optimal) plan.  We argue that with the proliferation of concerns and
variability that an optimizer must contend with, it is extremely
difficult to ensure that for an arbitrary query that the optimizer
will {\em always} pick the right plan. There are many sources of complexity
that may challenge one or more components of the optimizer; our theory
implicates specific sources as contributing to observed
suboptimal plans.

The model depicted in Figure~\ref{fig:model} suggests specific factors that may
impact the prevalence of suboptimality. This model has one dependent
variable, {\em suboptimality}, on the far right. We observe this dependent
variable in our experiments by determining whether each particular query,
run on a particular \hbox{DBMS} and using a particular schema,
is suboptimal at any cardinality of the input table. In Section~\ref{sec:operationalization}, we
delve into the details of how we operationalize this and the other variables
we now examine.

\subsection{Variables in the Model}
The model has several independent
variables which influence suboptimality.

For the construct of {\em optimizer
complexity} we have one independent variable, ``\hbox{Number} of operators
available (in the \hbox{DBMS}).'' We can manipulate this variable by our choice of
\hbox{DBMS}: each \hbox{DBMS} has a set of operators available to its query
evaluator and available to the query optimizer to use with in query plans.

We have one variable for the construct of {\em schema complexity}:
``Presence of secondary indexes.'' If true, that means that every
non-primary-key attribute, for the table for which we vary the cardinality
(termed the {\em variable table}) as well as for the other three tables
(termed the {\em fixed
tables}), has a secondary index associated with it. The rationale is that the
presence of secondary indexes expands the number of possible plans: each
predicate in the query can often be mapped to one or more operators
utilizing that index.

For the construct of {\em query complexity} we have identified four
variables: ``Number of correlation names'', ``Presence of aggregate'',
``Presence of primary key attribute'', and ``Presence of subquery". For each independent
variable, we have interventional control in our experiments, in that we can
manipulate the values of these variables through the construction of the
actual query to be optimized. The first is the number of correlation names
defined in the FROM clauses. The second is whether a single aggregate operator
(that is, either {\tt Max} or {\tt Avg}) appears in the SELECT clause.
The third
is whether a primary key attribute appears in at least one predicate in the
WHERE clause. The fourth is whether a single subquery appears in the
WHERE clause; that subquery evaluates to a single value that is
equality-compared with an attribute, in the where clause. The rationale is
that each factor may expand the number of plans or otherwise render the search for
the optimal plan more complex.

We have one variable for the construct of {\em data complexity}: that of
``Percentage of skewed data''.  Skew is generally defined as how values are
distributed with a column of a table. We refine this to ``how many duplicate
values are present,'' with zero skew meaning that there are no duplicate
values present and 100\% skew implies that there is but one value in the
entire column. The presence of skew complicates query time estimation, which
in turn complicates the search for the optimal plan.

In the middle of the figure is the construct of {\em plan space
complexity}. Given a particular query, its measurable
complexity will impact the total number of candidate plans considered
by the optimizer. However, this latter factor is not directly
observable, again, especially for proprietary \hbox{systems}. However, we {\em
can} measure the number of plans actually generated by the optimizer
when presented with different cardinalities of the underlying tables.
We term this set of plans the ``effective plan
space'' and term the number of such plans, the cardinality of the
effective plan space, or ``{\em CEPS}''. Similarly, we cannot directly observe
the cost model utilized by the optimizer for the operators it considers, but
can classify the cost model of each operator as continuous (smoothly varying
with cardinality of its input(s) and without jumps) or discontinuous
(sometimes producing observable jumps in the predicted
cost), and can thus count ``the number of operators with
discontinuity.'' (Such operators have also been termed
``nonlinear''~\cite{Hulgen03}, though we focus on a more specific aspect of
discontinuity.) Finally, the variable ``Number of repeats'' is the number
of times a plan is reused across the cardinality of the variable table. For
example, if the sequence of plans for a query as the cardinality increases
is $A ~ B ~ B ~ B ~ C ~ C ~ A ~ B ~ B~ B ~ C ~ B ~ C ~ C ~ C ~ B$, then the
``Number of repeats'' will be six: removing sequential duplicates results in $A ~ B ~ C
~ A ~ B ~ C ~ B ~ C ~ B$, with the last six distinct plans being duplicates
of the first three. Interestingly, CEPS plus the number of repeats gives you the
number of plans in the sequence, again, after removing sequential
duplicates, nine plans in this particular case.

Plan space complexity is an {\em intervening} variable, in that it is dependent
on some of the variables on its left but is observable and thus  exerts influence on the dependent
variable on its right. We can observe these variables within experiments and
indirectly influence their value but
cannot directly intervene to specify their value. For example, we can influence the CEPS through
manipulating the values for the independent variables of optimizer and query
complexity, but cannot directly set a value for CEPS within an
experiment.

\subsection{Interactions in the Model}\label{sec:interactions}
Given these six constructs and eleven specific variables depicted in
Figure~\ref{fig:model}, let's now examine the causal relationships between
these variables, depicted as directed arrow between variables (a line
originating or ending at a construct is interpreted as originating or ending at all
variables in that construct). Such relationships are specific {\em interactions} between
the the constructs (in particular, between their associated variables), as hypothesized by
this predictive causal model.

One causal factor of this model is the optimizer complexity. We hypothesize
that optimizer complexity has influence over suboptimality indirectly, via
plan space complexity. We hypothesize that an optimizer with a larger number
of available operators will generate more plans and hence increase plan
space complexity.

\vspace{0.6em}\noindent
{\bf Hypothesis 1}: Number of operators available will be positively correlated with, (a)~Number of
repeats, (b)~CEPS, and (c)~Number of operators with discontinuity.

\vspace{0.6em}We now turn to query complexity, a construct associated with four
independent variables. As with the optimizer complexity construct, we
include in our model an
indirect interaction through plan space complexity.

A higher value of each of these specific variables implies a more complex
query.  We expect a strong relationship between Number of correlation names
to CEPS (because it is well-known that the number of potential join
combinations is exponential to number of correlation names), to number of
repeats (number of repeats could partially track CEPS), and number of
operators with discontinuity (for the same reason).

We also expect a positive relationship between presence of aggregate (as
that will definitely add at least one operator to the query plan), presence
of primary key attribute, and
presence of subquery.

\vspace{0.6em}\noindent
{\bf Hypothesis 2}: Number of correlation names will be
strongly correlated with (a)~Number of repeats, (b)~CEPS, and (c)~Number of
operators with discontinuity. Presence of aggregate will be positively
correlated those those three variables (correlations d--f) and similarly
with Presence of primary key attribute
(correlation g--i) and Presence of subquery (correlations j--l).

\vspace{0.6em}
We don't feel that skewed data (in the ``data complexity''
construct) will impact plan space
complexity, but rather that it will negatively impact accuracy of plan cost
estimation, and thus increase suboptimality.

\vspace{0.6em}\noindent
{\bf Hypothesis 3}: Percentage of skewed data will be negatively correlated with Suboptimality.

\vspace{0.6em}
Let's now turn to the plan space complexity construct. We hypothesize a positive
correlation between the two variables associated with this construct. As the
number of plans considered by the optimizer increases, so should CEPS, which
could increase the number of operators with discontinuity that is observed.

The model also has two interactions within plan space
complexity, between 
CEPS and number of operators with discontinuity variables and between number of repeats and CEPS.

We hypothesize that greater plan space complexity should make it more
difficult to optimize the query.  It is important to note that the
occurrences of sub-optimality is not necessarily dependent on the presence
of a discontinuous plan operator. However, we predict that when
discontinuous plan operators appear in candidate plans, they may introduce
complexity to query optimization, especially at the cardinality where
discontinuity is possible. This is because the performance of such operators
is sensitive to the input size in a rather complex way. Any inaccuracy in
the estimation of plan statistics may lead the optimizer to select a suboptimal
plan.

\vspace{0.6em}\noindent
{\bf Hypothesis 4}: Plan space complexity (CEPS) and Number of operators
with discontinuity will be positively correlated.

\vspace{0.6em}
Similarly, a plan chosen by the optimizer once will be reconsidered at a
later cardinality and perhaps chosen, following
the past experience of selecting and using that plan among many other plans.
Even if many different plans are already used, 
for the same reason the optimizer may revisit a pool of 
the previously used plans and choose one of them than to explore other new plans, 
given the complexity of plan cost estimation. 
Then the larger CEPS, the more times the optimizer repeats using the same plans. 
We thus hypothesize that the number of repeats has a positive correlation with CEPS.

\vspace{0.6em}\noindent
{\bf Hypothesis 5}: The Number of repeats will positively correlate with CEPS.

\vspace{0.6em}
The schema complexity construct consists of the variable ``Presence of
secondary indexes.'' Such indexes provide opportunities for the optimizer to
consider more candidate plans that use these indexes, due to the additional
query evaluation operators now applicable. Those additional plans enable the optimizer to
possibly do a better job, while also adding complexity to the optimization
process. We hypothesize that this factor has a more complex role in the
model, serving as a {\em moderator} of two interactions previously
introduced. We hypothesize that the overall effect of the secondary indexes
is to increase the strength of the interaction between query
complexity to plan space complexity.

\vspace{0.6em}\noindent
{\bf Hypothesis 6}: Presence of secondary indexes will strengthen
the correlations between the Query Complexity construct, specifically, the
variables Number of
correlation names, Presence of aggregate, Presence of primary key
attribute, and Presence of subquery, and
the Plan Space Complexity construct, specifically, variables Number of repeats, CEPS, and Number
of operators with discontinuity (correlations a--l listed above in
Hypothesis 2).

\vspace{0.6em}
Finally, we hypothesize that queries with a high plan space complexity
present challenges to the query optimizer, and thus increase the chance that
the query optimizer picks a suboptimal plan. This is a fairly direct
connection for CEPS and Number of operators with discontinuity. For the
Number of repeats variable, a query exhibiting a high number of repeats has
more opportunities for suboptimality at some cardinality, just because the
number of repeats is an indication that the query optimizer is struggling.

\vspace{0.6em}\noindent
{\bf Hypothesis 7}: Suboptimality will be positively correlated with Plan
space complexity, that is, (a)~Number of repeats, (b)~CEPS, and (c)~Number
of operators with discontinuity.

\vspace{0.6em}
We have just {\em described} how
suboptimality might arise, through a theory and its elaborated causal model,
which implies seven specific hypotheses.
We now need to move to {\em prediction}.
How might we test such a model?

The first step to test this model is to {\em operationalize} each
variable. In the next section we describe explicitly how each variable is
defined. For the independent variables, we must be able to intervene, that is,
set their values before the experimental test commences. For the latent and
dependent variables, we need to be able to measure their values during each experiment.

\section{Variable Operationalization}\label{sec:operationalization}

In this section we specify how we operationalized each of the seven
variables in the model.  Recall that each independent variable is a property
of a \hbox{DBMS} (number of operators available), of the schema (presence
  of secondary indexes), of a query (number of correlation names,
presence of an aggregate, presence of primary key attribute, and
presence of subquery), of the data (percentage of skewed data), of
the plan space (number of repeats, CEPS, and number of discontinuous
operators). There is 
one dependent variable of our model (suboptimality).

It is important to note that all manipulation must be done {\em
  outside} the \hbox{DBMS}. For proprietary systems we do not have access to the
internal code. We do not know ({\em cannot} know) all the plans that were
considered, nor the details of how the plans are selected. But such access
is not needed; indeed, to be able to study a phenomenon across many \hbox{DBMSes},
such access is not practical. But by designing the experiment to examine
the plans that each \hbox{DBMS} actually produces, and thus to examine phenomena
that can be externally visible, we can obtain valuable insights into general
classes of computational artifacts. Our ultimate goal is to make statements
that hold across cost-based optimizers in general, thereby moving up the
$y$-axis of Figure~\ref{fig:empirical}.

\begin{figure*}[t]
\centering
\includegraphics[width=0.60\textwidth]{figures/understanding.pdf}%0.40
\caption{Empirical Generalization\label{fig:empirical}}
\end{figure*}

\subsection{Optimizer Complexity}
By ``number of operators available in the \hbox{DBMS}'' we mean the \hbox{number} of operators
available for selection, projection, join and aggregate functions. Our
experiments intervene on this variable by selecting a particular \hbox{DBMS} on
which to evaluate each query. Across the available \hbox{DBMSes}, as the number of
operators available increases, the complexity of the optimizer increases
because it has to choose between more operators.

The {\tt EXPLAIN PLAN} facility specifies the operators(s) \hbox{employed} in that
plan. For each \hbox{DBMS}, we collect all the unique operators from all the plans
and count the number of these distinct operators, each used by for least one
query at at least one cardinality by that \hbox{DBMS}. The number of operators used
by a \hbox{DBMS} ranged from 8 to 53 across all the queries and data sets
used in the Exhaustive, Exhaustive with Keys, and Exploratory Experiments,
to be discussed in Section~\ref{sec:experiments}.

\subsection{Schema Complexity}
The presence of secondary indexes is easy to operationalize. We generate two
databases, one without any secondary indexes and one with a key specified
for each table on the non-key (that is, other than the first) attribute.

\subsection{Query Complexity}\label{sec:querycomplexity}
Query complexity is also relatively easy to operationalize. We randomly generate queries,
such as the example presented in Section~\ref{sec:motivation}. Each query
is a select-project-join-aggregate query, with a few attributes in the
SELECT clause, a few tables referenced in the FROM clause, a few
equality predicates in the WHERE clause, and zero or one aggregate
functions in the SELECT clause. As such, some of the complexities
mentioned by Kabra and \hbox{DeWitt~\cite{kabra98},} such as
user-defined data types, methods, and operators, are not
considered. Concerning presence of primary key attribute, if this independent
variable was set to 1, we ensured that there was at least one primary key
attribute in one of the comparisons in the WHERE condition.

The queries were generated
by a simple algorithm\shorten{, the pseudo-code for which is shown in
Figure~\ref{alg:querygen}}. 
\shorten{The query generator has four main components, which are {\tt
  generateSelectClause()}, {\tt generateFromClause()}, {\tt
  generateWhereClause()}, {\tt generateGroupByClause()}. , and {\tt buildAggregateStatement()}.}
The SELECT clause will contain from one to four
attributes, hence, an average of 2.5 attributes. The number of
correlation names
in the FROM clause varied from one to four, with duplication of tables
allowed (duplicate table names within a FROM clause implies a self-join).
In the queries that were generated, from one to four unique tables were
mentioned in the FROM clause. Somewhat fewer tables were mentioned than the
number of correlation names, as the presence of self-joins reduced the
number of unique tables referenced by the queries.

For the query in Section~\ref{sec:motivation}, the number of correlation
names is 3, the presence of an aggregate
is false (0), the presence of primary key attribute is false (0),
and the presence of subqueries is false (0).

%\pagebreak
\shorten{In {\tt generateWhereClause()}, in which all the joins will appear,
  we}We ensure that Cartesian products are
eliminated\c2j{.}{\shorten{({\tt cartesianPossible="false"})}}  We do this by
connecting all the correlation names that appear in the FROM statement via
equi-joins on random attributes. The comparisons are all
equality operators. To ensure that the queries are as simple as possible, we
do not include any additional predicates in the WHERE clause.  \c2j{}{This
  is realized by setting the attributes {\tt maxIsAbsolute} to {\tt true}
  and {\tt complexUsePercentage} to {\tt 100}.  Basically, ``complex''
  predicates eliminate the Cartesian product, and by setting complex
  predicates as ``absolute,'' no additional predicates are included except
  for those which are necessary for eliminating Cartesian product.} Also
for simplicity, we include neither disjunctions nor negations.

\shorten{In the \todo{Correct name: generateGroupByClause}, where the aggregate
functions and
grouping occurs, the presence or absence of a group by clause is
recorded. We expect the
complexity of the query to increase when the first aggregate function
is introduced. (We don't consider in this model whether
additional aggregate functions in the same SELECT clause add
complexity}

The presence of subquery was 0 or 1, with 0 indicating no
subquery. For each query needing a subquery, we picked a separate generated query and
rendered it as a subquery to replace an attribute in the where clause of the
original generated query. As an example when this variable was one, we
started with query qs1-2, shown here.

\begin{verbatim}
SELECT t0.id2, SUM(t1.id3)
FROM ft_HT2 t0, ft_HT2 t1
WHERE (t0.id2=t1.id1)
GROUP BY t0.id2
\end{verbatim}

We then generated another simple select-project (SP) 
query at random concerning the variable table 
and simply replaced one side (in this case, the right side) with the generated entire 
query as a subquery, to produce this final query.

\begin{verbatim}
SELECT t0.id2, SUM(t1.id3)
FROM ft_HT2 t0, ft_HT2 t1
WHERE (t0.id2 IN (SELECT t2.id3 FROM ft_HT1 t2)) 
GROUP BY t0.id2
\end{verbatim}
We thus have a maximum of only one level of nesting.

\subsection{Data Complexity}\label{sec:datacomplexity}
This independent construct has one variable: percentage of skewed data.
%\todo{Rick:write up operationalization}

Skew has a very specific definition in statistics, 
involving elongating the left or right tail of a distribution, 
thereby moving the mean left or right of the median 
(in a symmetric distribution the mean = median = mode).

But we start with a distribution without a tail: 
the uniform distribution: the values from 1 to 2 million (2M). 
We consider this to be a skew of 0 (no skew). 
At the other end of the spectrum is one in which all the values are identical, 
or a skew of 1.0.

We define the skew as the reciprocal of the number of distinct values. 
For 2M distinct values, the skew would be 1/2M, which is practically 0. 
For 1 distinct value, the skew would be 1.0. For 2 distinct values, 
the skew would be 0.5. For 10 distinct values, the skew would be 0.1.

We can generate the table of 2M rows by generating values sequentially from
1 to the number of distinct values.  This creates a "span of values." We
repeat this for the second span if necessary, and on and on, until we have
2M values in all.

When varying the cardinality, we remove 10K values from the variable table
and then copy those tuples to a new table to ensure that every page is as
full as possible (that is, 100\% load factor).  This gives us a table of
1.99K tuples. (We then get a query plan for this table.)  We repeat this
removal process until the final cardinality reaches 10K.

The way we effect the 10K removal is as follows. 
The key idea is to remove individual spans until we've deleted 10K
values. Since we don't touch the remaining spans, the number of distinct
values does not change, and so the skew remains constant.

Let's first illustrate with a skew factor of 1.0. 
This translates to exactly 1 value for the entire table, or 2M spans.
To shorten, we remove 10K spans, equal to 10K values. 
This removal keeps the skew at 1.0, due to the unique values in the
remaining spans.
At the end, 10K spans, each with a single value, remain.

What about a skew factor of 0.1? 
This skew factor translates to 10 distinct values. 
So we generate 200K spans, each with 10 values. 
To shorten, we delete 1,000 spans, which is equivalent to removing 10K values. 
As before, this does not change the skew.
At the end, the table will have 1,000 spans with 10K values, retaining a
skew factor of 0.1.

Let's see an example at the other end of the spectrum, of a skew factor of
1/2M.  This skew factor, which is almost close to 0, gets translated to 2M
distinct values.  So we generate a single span of 2M values.  To shorten the
table, it makes no sense to remove this single span in its entirety. But we
can remove 10K values from this single span.  Note that this changes the
skew to 1/1.99M, then eventually to 1/10K, which is still very close to zero
(the skew has changed from .0000005 to .00001).

We use five skewness values: 0 (approximately), 0.001, 0.1, 0.5, and 1.0,
for this independent variable.

\subsection{Plan Space Complexity}\label{sec:discontinuity}

This explanatory construct includes three variables.

As discussed in Section~\ref{sec:model}, the ``cardinality of the effective plan
space'' (CEPS) is the number of plans selected as optimal for that query
being evaluated on one of the 200 cardinalities for the variable table. It
is ``effective'' because it was chosen, as opposed to all of the
plans that were considered but not chosen. (Recall that for proprietary
\hbox{DBMSes}, we do not have access to such plans.) Note that we count only
distinct plans. As we saw in Figure~\ref{fig:query769}, fluttering queries
return to a previous plan. This particular query has a CEPS of 4.

The number of repeats is just the number of times a plan associated with a
smaller cardinality is repeated, after removing sequential
duplicates.

We also wanted to evaluate the contribution of cost model non-linearity to
suboptimality. We found that it is possible to assemble, from outside the
\hbox{DBMS}, an approximation of the cost formula for that operator, and thus
directly observe whether it is non-linear. Specifically, the result of SQL's
{\tt EXPLAIN PLAN} (a result that is particular to each \hbox{DBMS}) includes the
{\em estimated cost} of each stage of the plan. That cost (estimate) is a
dependent variable, one that can be observed.

To operationalize the ``number of operators with discontinuity'' we
classify each \hbox{DBMS} operator as either continuous or discontinuous, to be
defined shortly.  We then examine the queries
to identify the plan change points, which provide the {\em effective plan
  space}, a set of plans chosen for one or
more cardinalities of the variable table. For each distinct query plan in the
effective plan space, we
count the number of discontinuous plan operators that appear in that plan. Some of
these plans may contribute no such operators, some may contribute one such
operator, and some may contribute several such operators. We then sum the
counts of the distinct plans in the effective plan space, yielding an
integer as the operationalization of ``number of discontinuous operators''
for each query.  This count per query varied from 0 to 85 for the
queries we tested in the confirmatory analysis.

To classify each operator as continuous or discontinuous, we use the data
from the Exhaustive Experiment. We
used a small subset of queries to be used later in testing the model, to
attempt to observe the same operators as encountered in that
study. We
collected all query plans at all possible cardinalities (200 in all) and
looked for ``jumps,'' that is, when the cost model for an operator in the
plan exhibited discontinuity. Jumps are determined from the estimated cost
extracted from each plan operator. We provide an example of a jump below. If
a jump is observed, we classify that particular plan operator as
discontinuous.

A {\em jump} is a pair of close cardinalities (in our case, separated by 10K
tuples) in which the query optimizer's cost model is {\em discontinuous},
that is, does not smoothly increase from the lower cardinality to the upper
cardinality. To identify jumps in the first step, we examine the slopes
between each pair of consecutive cardinalities (again, separated by 10K
tuples). We expect that the slope (that is, the first derivative) is
well-behaved and thus does not change much for consecutive cardinalities. 
A jump thus indicates a large deviation in the second derivative of the 
cost model across the two cardinalities.

The Exhaustive experiment gathers the cost model for each operator in each
plan for each cardinality (in our case, for cardinalities ranging from 10K
to 2M in steps of 10K, or 200 points).  Figure~\ref{fig:discontinuity}
presents for a single query\shorten{({\em query4}, exhaustive-10q-1,
  oracle)}, the {\em cost} of each of the hash-join operators utilized in
each plan at a particular cardinality. Each distinct point type depicts an
individual hash-join operator. Note that when two identical query plans
appear at two different cardinalities, we consider all the plan operators
found at the same position within these two plans to be identical. This
figure shows only a small portion of the 200 cardinalities, and is typical
(we generally observed jumps at low cardinalities for these particular queries and
relation sizes).  As shown by this figure, the hash-join operator
represented by `+' has two discontinuous jumps, both appearing below
cardinality 150K. The other hash-join operator represented by $\times$ has one
discontinuous jump.  Between the jumps, the behavior is linear. It is our
guess that
such a jump is caused by the transition between one pass of a disk-based hashing
technique to two passes (and indeed for one of the operators, perhaps the one
lower in the operator tree, the transition to three passes, hence, the two
jumps). That said, all
that matters for our predictive model is that operators that experience
such discontinuity in their cost models present opportunities for
suboptimality.

%\begin{figure}[t]\centering
\begin{figure}[t]
\centering
%\includegraphics[width=0.70\textwidth]{figures/discontinuity_400K.pdf}%0.40
\includegraphics[width=0.60\textwidth]{figures/discontinuity_500K.pdf}%0.40
\caption{An Example of Discontinuous Plan \hbox{Operators (Hash-Join)}}
\label{fig:discontinuity}
\end{figure}

\shorten{It is interesting that jumps occurred only at smaller cardinalities. This
might depend on the strategy of how the \hbox{DBMS} utilizes the main memory
buffer.  Assume that the \hbox{DBMS} attempts to use as little memory as possible.
One possibility is that when the cardinality is low, memory allocation is
done in small chunks. Hence as the cardinality grows, more passes are
required. However, when tables get larger, bigger
chunks of memory would be allocated, requiring the operator to take longer
to fill up the available memory. We again emphasize that our focus is on predicting
suboptimality. Perhaps memory allocation has a role in operator
discontinuity, which our model predicts to have a role in query
suboptimality.
}
To identify such jumps, we examine the second derivative (the change in the
first derivative, that is, the change in the slope). A jump will be indicated by a larger than normal second
derivative. By identifying a sudden change in the second derivative, we can
effectively spot the cardinality at which an operator in a plan becomes
discontinuous.

Formally, we compute the slope ({\em S}) of the {\em estimated} cost
(of the cost model, formalized as $C_{\hbox{\small\em card}}$,
where {\em card} is the input
cardinality) between each pair of consecutive cardinalities as
$S_{\hbox{\small\em card}} = (C_{\hbox{\small\em card} + 10K} -
C_{\hbox{\small\em card}}) / 10K$~.
This computes a series of slope values.
We then compute the standard deviation of all the slope
values.  For example, examining Figure~\ref{fig:discontinuity}, the slopes
are small except at three places, one for the green operator and two for the
red operator. By identifying the slope values that are greater than one
standard deviation over the average value, the discontinuous operators can
be identified: A ``discontinuous operator'' is one for which a jump is
observed in that operator in one of the plans for at least one of the input
queries.

\subsection{Suboptimality}\label{sec:suboptimality}
We now consider the one dependent variable.
How might suboptimality be observed? We have developed a system, \azdb, that
allows us to perform experiments to study this phenomenon of suboptimality.
\hbox{\azdb} submits queries to the \hbox{DBMS}, while varying
the cardinality of one of the tables,
requesting in each case the evaluation plan chosen by the \hbox{DBMS}.
This is done using the
{\tt EXPLAIN} SQL facility available in modern \hbox{DBMSes}. (The Picasso system
also used this facility to visualize the plan space chosen by a \hbox{DBMS}
optimizer~\cite{harish07,Haritsa10}.) We can then compare
the performance (execution time) of various plans for the query, to identify those
situations when a suboptimal plan was chosen, when in fact there was a
different plan that was semantically equivalent to the chosen plan (that is,
yielded the identical result) but which ran faster.

We modify the cardinality to produce multiple execution
plans for a given query. For one of the \hbox{DBMSes}, we can modify the stored
table statistics directly. For the other \hbox{DBMSes}, we had to do so indirectly,
by varying the size of the table and running the optimizer on tables of
different size. As we vary the cardinality, we
collects all the plans that the optimizer felt were appropriate for that
query.

Our definition of suboptimality assumes that
actual execution time for any query plan is {\em monotonic non-decreasing}, that is,
unchanged or increasing as the cardinality increases.  The intuitive justification is that at the higher
cardinality, the plan {\em has} to do more work, in terms of CPU time and/or
I/O time, to process the greater number of tuples. (For an operator having a discontinuous cost model, as the
anticipated input cardinality grows, there will be {\em jumps}, in which the
predicted cost is temporarily much greater. Section~\ref{sec:discontinuity}
explains further how we specifically
operationalize this property. Also note that we don't consider SQL
operators such as {\tt EXCEPT} that are not monotonic.)
We formalize this property as follows.

\vspace{1em}\noindent
{\bf Definition:} {\em Strict Monotonicity}: given a query $Q$ and an actual
cardinality $a$,

\quad\quad\quad\quad\quad$\forall p \in \hbox{\em plans} (p) \;\forall c > a \;( \hbox{\em time} (p,
c) \geq \hbox{\em time} (p,a) )$

\vspace{1em}\noindent
where {\em plans}($p$) is the set of plans generated by the optimizer and
{\em time}($p$, $c$) is the execution time of plan $p$ on the data set with
the varying table at cardinality $c$.\hfill$\blacksquare$

\vspace{1em}
\noindent
Note that the comparison is with the same plan $p$, occurring at higher
cardinalities.

To test this assumption, we ran an experiment that we term
``Monotonicity.'' This experiment considered 60 queries, chosen from the
pool of queries
generated for testing suboptimality, and timed them for cardinalities from 10K
to 2M tuples, in steps of 10K tuples (hence, we used 200 cardinalities), for
each \hbox{DBMS} (for one \hbox{DBMS} that was very slow, we started with 30K tuples, as
discussed in Section~\ref{sec:experiments}). We varied the cardinality of the variable table by starting with
the maximum size, running the queries, then deleting 10K tuples and repeating.
We performed an ``{\tt ANALYZE TABLE}'' function to force
the \hbox{DBMS} to update the table's statistics to be accurate before actually
evaluating the query. \todo{Young: we did this for all experiments, not just
  monotonicity, right?}
We then examined every case where the same plan for a query was run at different cardinalities; that is,
we compared two query times measured by {\tt currentTimeMillis()} in {\sc
  Java} per a pair of different cardinalities at which the same plan is
observed during the study.   We expected that as the cardinality decreased,
runtime would monotonically decrease. However, due to the variance in query
time measurement observed even when the cardinality was identical, we encountered
spurior violations.

Assuming a normal distribution for our time measurements, 95\% of 
the distribution falls between $-2*\sigma$ and $+2*\sigma$. 
Therefore, to statistically
infer with a 95\% confidence interval that a violation occurred, 
we relaxed our definition of monotonicity to the following.

\vspace{1em}\noindent
{\bf Definition:} {\em Non-Strict Monotonicity}: given a query $Q$, an actual
cardinality $C$, and the standard deviation of the query executions for
cardinality C as $\sigma$, $Q$ is non-strict monotonic if $\forall p \in
\hbox{\em plans} (p) \;\forall c^{\prime} > C $,

\quad\quad\quad\quad\quad$\hbox{\em time} (p, c^{\prime}) \times (1+\sigma_{c^{\prime}}) \geq
\hbox{\em time} (p,C) \times (1-\sigma_{C})~.~~\hfill\blacksquare$

\vspace{1em}
As we will see in Section~\ref{sec:experiments}, we observed only 3,347
violations (0.73\%) of non-strict monotonicity, for the largest experiment
(Experiment 7: Confirmatory) across all
the \hbox{DBMSes} we studied, justifying our conclusion that the \hbox{DBMSes} under
study are indeed monotonic.

We can now turn to suboptimality. Recall that the monotonicity test examines
two adjacent {\QatC}s for which the {\em same plan} is observed. To detect
suboptimality, we look for adjacent {\QatC}s with {\em different plans}, termed
a ``change point,'' mentioned earlier. We look for such change points where the computed
query time at the {\em upper} cardinality is {\em smaller} than the computed
query time at the {\em lower} cardinality. Say the lower cardinality used
Plan {\em A} and the upper cardinality exhibited Plan {\em B}. Had the \hbox{DBMS}
query optimizer selected Plan {\em B} for the lower cardinality, the query time would have been smaller than that for
Plan {\em A}, which follows directly from the
monotonicity assumption. The conclusion is that for the lower cardinality, the
optimizer picked the less efficient plan, and thus, this query exhibits
\hbox{suboptimality}. Note that since this approach cannot consider plans
that were never chosen, it very likely misses some suboptimal plans (for
which there was a better plan not seen), and thus produces a
conservative estimate of suboptimality.

Our definition of suboptimality compares the computed runtimes 
and standard deviations at the cardinality just before the change
point (designated as $n-1$) and at the change point (that is, $n$).

The query is said to be suboptimal if
%\hspace*{3em}
$\hbox{\em time}_{n-1} - 0.5 \cdot \hbox{\em stddev}_{n-1} \geq \hbox{\em
time}_n + 0.5 \cdot \hbox{\em stddev}_n$~.\\
At each Q@C pair, Suboptimality is coded as four levels (0--3), based on the distance in 
standard deviations, upto 3 standard deviations. 
\shorten{\begin{itemize}
\item Not suboptimal, i.e., 0, when
$$\hbox{\em time}_{n-1} - 0.5 \cdot \hbox{\em
stddev}_{n-1} < \hbox{\em time}_n + 0.5 \cdot \hbox{\em stddev}_n $$
\item Low suboptimality, i.e., 1, when
$$\hbox{\em time}_{n-1} - 0.5 \cdot \hbox{\em
stddev}_{n-1} \geq \hbox{\em time}_n + 0.5 \cdot \hbox{\em stddev}_n$$
and
$$\hbox{\em time}_{n-1} 1 1.0 \cdot \hbox{\em stddev}_{n-1} < \hbox{\em time}_n +
1.0 \cdot \hbox{\em stddev}_n$$
\item Moderate suboptimality, i.e., 2, when
$$\hbox{\em time}_{n-1} - 1.0
\cdot \hbox{\em stddev}_{n-1} \geq \hbox{\em time}_n + 1.0 \cdot \hbox{\em
stddev}_n$$
and
$$\hbox{\em time}_{n-1} - 1.5 \cdot \hbox{\em stddev}_{n-1} < \hbox{\em time}_n +
1.5 \cdot \hbox{\em stddev}_n$$
\item High suboptimality, i.e., 3, when
$$\hbox{\em time}_{n-1} - 1.5
\cdot \hbox{\em stddev}_{n-1} \geq \hbox{\em time}_n + 1.5 \cdot\hbox{\em stddev}_n$$
\end{itemize}

\noindent}
We then sum this value over all pairs of cardinalities with different plans
(the change points) for the query.

\todo{Sabah: this paragraph talks about a suboptimality of up to 133, but the
  above papra says there are just 4 levels. That is inconsistent. The reason is that the raw value ranges from 0 to 133. 
After the log transformation, the values range from 0 and 2.13. Also, do
  we still use a log transformation?}
Note that while \azdb\ examines the plan at every cardinality, it only has
to actually execute the query at pairs of cardinality that are change
points.  Since many fewer {\QatC}s were involved, we could try many more
queries than the \hbox{Exhaustive} Experiment. In the Exploratory Experiment to be
described in Section~\ref{sec:experiments}, the value of suboptimality ranged
from 0 (no suboptimality) to 133, with the majority between 0 and 9. A full 56\% of
the queries were suboptimal.  \hbox{Because} the occurrence of large values of this
measure was so rare, we did a log transformation: $\log_{10} (1 + \mbox{\em
  subopt})$ in the confirmatory analysis.

\section{Testing the Causal Model}

\c2j{}{\subsection{Experimental Setup}\label{sec:setup}}
\shorten{As mentioned, we used \azdb\ to run the experiments. This infrastructure,
now comprising around 44K lines of code, handles all the details of
generating the sample data, generating the actual queries, collecting the
various plans generated by the optimizer, running the queries, and storing
the results in a ``labshelf,'' a set of tables managed by
a separate \hbox{DBMS}. This infrastructure also performs the many steps
required to ensure repeatability.

We have assembled a hardware lab of dedicated
machines, one \c2j{for each DBMS}{each for the three \hbox{DBMSes}} and one to run the \hbox{DBMS} used
to store the lab shelves. Having dedicated hardware allows us to worry
less about other processes running on the machines that could dirty the
results (these machines are {\em only} for experiments), and allows us to
run extensive experiments involving days or weeks of computation. We have
used this software and hardware lab to perform a number of experiments
examining the phenomenon of suboptimality.
}
\shorten{{\azdb} utilizes experiments defined in XML. It supports {\em experimental
  scenarios}, which are short (perhaps 100 lines of code) Java subclasses
that provide details of how to run kinds of experiments.}

The measurements were collected using Tucson Timing
Protocol\linebreak Version~1 (TTPv1)~\cite{TTPv1} and Version~2 (TTPv2)~\cite{TTPv2} on a suite of five machines, each an
\hbox{Intel} Core i7-870 Lynnfield 2.93GHz quad-core processor on a LGA 1156 95W
motherboard with 4GB of DDR3 1333 dual-channel memory and Western Digital Caviar
Black 1TB 7200rpm SATA hard drive, running Red Hat \hbox{Enterprise} Linux
Server release 5.8 (Tikanga) for TTPv1 and release 6.4 (Santiago) for TTPv2, with a kernel of 2.6.32-358.18.1.  The protocol
provided calculated query evaluation time, including computation and I/O time,
in msec. Both protocols were utilized exactly as specified.

No runs violated the experiment-wide sanity checks. For the largest
experiment, the Confirmatory experiment 7 discussed below, approximately
10.5\% of the query exectuions and 5.1\%
of the queryies-at-cardinality (Q@Cs) were dropped due to query execution
and Q@C sanity checks. As a result, excessive variation in calculated query
time was observed in only 0.003\% of the Q@Cs. A total of 0.35\% of Q@C adjacent
pairs violated relaxed monotonicity and 0.68\% of the Q@Cs violated strict
monotonicity, which is acceptable.

\doubleblind{In fact, over the course of the research described here, we
  found that currently available techniques did not offer sufficient query time
  precision, and so we developed TTPv1~\cite{TTPv1} to provide that needed
  precision. As our research in suboptimality prooceeded, we realized that
  to get a actually recognized suboptimality for two adjacent cardinalities
  separated by only 10,000 rows (just a few pages), or even in the case of
  MySQL, 300 rows, we needed even finer resolution. This provided the
  impetus for the massive effort to develop TTPv2~\cite{TTPv2}, which did
  indeed provide the needed precision.}{We
  thank the developers of this protocol and of the software that enabled the
  running of many, many queries for providing us this software.}

In these experiments, we  installed each disk directly
on the machine that also runs the DBMS, ensured a cold cache (disk drive,
disk controller, O/S, and DBMS buffer), and discarded any sequence of query
executions that appear to be the result of query result caching. We also
ensured within-run plan repeatability.

In the following, we describe in detail the data used by our experiments and
the experimental scenarios we defined. More details on the data sets used
can be found in Appendix~\ref{sec:app}.

\subsection{Data sets}\label{sec:datasets}
We generate our experiment data set randomly in the experiments. However, we
use seeds to control the random data generator so that it can produce
repeatable data as required. (Randomly generated data will probably be
easier for the optimizer to deal with than skewed data.)

Our experiment data set consists of relational tables. There are two types of
tables. The first is a ``fixed table'' that, once created and
populated, will never be modified in the future.  In contrast, the second
type is a ``variable table.'' We alter the cardinality\c2j{ of
  this table}{, physically, of such tables} as the experiments are being performed. 
%\todo{Young: please check above}
\c2j{We used a}{We organized three configurations for the data sets. The first configuration
is a small} data set with four tables, each with four integer-typed
attributes. Each table is populated with one million rows, and thus the size
of each table is roughly 16Mbyte. We also produced a version of the data set with primary
keys (of the first attribute) for all tables. \todo{Young: is the following
  true'' I don't think so...: Second, to
perform queries with more relations, we create eight tables. In this case,
each table has the same configuration as in the first data set. Third, to
test queries on large tables, we configure each table with 64 integer-typed
attributes. In this data set, we create four tables, but the size of each one is
512MByte.}

\subsection{The Experiments}\label{sec:experiments}

We are interested in predicting the behavior of \hbox{DBMSes} through our
model. We selected four relational
\hbox{DBMSes}, some open source and some proprietary, that are representative of
the relational \hbox{DBMS} market. Each was used in its stock configuration.

In each experiment, we varied the
cardinality from 2M (maximum) to 10K (minimum), in increments of 10K. For
the one \hbox{DBMS} that was slower than the others and was timing out for
the majority of the queries when run between 10K and 2M, we reduced the
size of the tables and varied the cardinality from 60K (maximum) to 300 
(minimum), in increments of 300.

Utilizing JDBC to manipulate independent variables from outside the
  \hbox{DBMS} allows us to empirically generalize by moving up the $y$ axis of
  Figure~\ref{fig:empirical}, from one system to several systems and then to
  a general theory. We don't reveal the identity of the \hbox{DBMS} we studied, for two reasons. First, commercial \hbox{DBMSes}
include in their user agreements requirements not to release 
performance data. This is detrimental to science, but we have no choice but
to live with that restriction. However, in some sense the specific \hbox{DBMS}
doesn't matter, as we are studying phenomena about cost-based optimizers
{\em in general}, and so are interested in making statements that apply to
all the experimental subjects in our study.

\begin{table}[t]
\tbl{Experiments 1--7: Detailed Run Statistics\label{tab:run_stat}}
{%
\resizebox{140mm}{!}
{
\begin{tabular}{c|c|c|c|c|c|c|c}
& {\em Experiment}& {\em Protocol} & {\em Cumulative} & {\em Number of}&{\em
    Number} &{\em Number}& {\em Number of}\\
& & & {\em Hours} & {\em Query Instances}&{\em of Q@Cs} &{\em of QEs}&{\em Retained QEs}\\
\hline
1 & Monotonicity 		& --- & 38 & 60 & 12,000 & 12,000 & 12,000\\
2 & Initial Exhaustive 		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
3 & Exhaustive 			& TTPv2 & 1,544 & 160 & 32,000 & 320,000 & 319,980\\
4 & Exhaustive with Keys 	& --- & 28 & 200 & 40,000 & --- & ---\\
5 & Initial Exploratory 	& TTPv1 & 560 & 780 & 8,842 & 88,420 & 68,891\\%40512+26367+2012
6 & Exploratory 		& TTPv2 & 1,663 & 1,200 & 12,560 & 125,600 & 114,377\\
7 & Confirmatory 		& TTPv2 & 11,375 & 7,640 & 99,558 & 995,580 & 890,631\\
\multicolumn{3}{c|}{Total}	& 16,880 & 10,200 & 236,960 & 1,861,600 & 1,650,666\\
\end{tabular}
}
}
\end{table}

Table~\ref{tab:run_stat} exhibits the statistics of running a number of 
queries in our experiments. All but the last column list the number of query
instances, etc., gathered by the protocol for each experiment. The last
column gives the number of query executions retained by the protocol. One of
the primary goals of TTPv2 was to reduce the number of QEs discarded due to
phantom processes that were indirectly detected. Experiments 3, 6, and 7
benefited from that protocol.

We performed six separate experiments, each looking at a different
aspect. It is important to emphasize that while the {\em queries} all came
from the same query pool and while the data sets were also shared by the
experiments, the {\em query executions} for the five experiments are
disjoint. As a side comment, we mention that for all four \hbox{DBMSes}, the query
plans generated by a \hbox{DBMS} for a particular \QatC\ of a particular query
varied between the experiments, but not between the query executions (QEs) of that query
instance at that cardinality, by virtue of the way we designed the
measurement protocol.

The first experiment, termed ``Monotonicity,'' was described in
Section~\ref{sec:suboptimality}. This experiment ran quickly, as it
only involved 12,000 QEs. That experiment helped us realize that we needed
to be much more sophisticated in our approach to timing queries.

\doubleblind{After we developed our six-step TTPv1 protocol,}{When TTPv1
    became available,} we performed our
second experiment, termed ``Initial Exhaustive,'' which more accurately tested the
monotonicity assumption, as also described in Section~\ref{sec:suboptimality}.
This experiment involved 160 query instances, 32,000 \QatC s, and thus
320,000 QEs, requiring 1,672 cumulative hours.  The (very small) percentage
of strict monotonicity violations observed was consistent with the remaining
variance of the query time measurement, concluding that none of
the \hbox{DBMSes} violated monotonicity. This also provides a validation of
our definition of suboptimality, which requires monotonicity.

We used the plans generated from Initial Exhaustive to classify operators as
continuous or discontinuous, as discussed in detail in
Section~\ref{sec:discontinuity}. Again we note that this experiment used a subset of
the set of {\em queries} (but {\em not} query executions) from the query pool,
to ensure that we see the same operators as encountered in that
study.

We later reran this experiment using TTPv2 (we term this simply
``Exhaustive''). Note that the number of retained QEs went up a lot (that is
the main benefit of TTPv2), while the number of hourse required actually
went down (another benefit).

We also ran an experiment (number 3 in the table) on the Exhaustive query set but using the data
set with primary keys, termed ``Exhaustive with Keys.'' However, in this
case we did not actually execute 
the queries, but rather just examined the plans that were returned from the
\hbox{DBMS} to identify additional discontinuous operators. \comment{Thus this third
experiment was very fast.}

The fourth experiment was for initial exploratory analysis of a prior
version of the causal model. (This fourth experiment used
the first version of the Tucson Timing Protocol (TTPv1)~\cite{TTPv1}.)
The earlier model had fewer independent
variables yet also had more complex interactions. Specifically, the model
did not include the independent query complexity variables of presence of
secondary indexes, presence of subquery, nor the schema complexity
independent variable of presence of secondary indexes, nor the plan space
complexity independent variable of number of repeats. The prior model also
had presence of primary keys in the schema complexity construct rather than
the query complexity construct. In reformulating that independent variable,
we were able to remove a complex mediating moderator between the plan
space complexity and suboptimality constructs.

%% The following shorten corresponds to TTPv1.
The fifth experiment was an ``Initial Exploratory'' analysis of the causal model, run on
a representative sample of queries and data sets: (a)~600 queries of one
\hbox{DBMS}, (b)~120 queries of another \hbox{DBMS}, (c)~10 queries
from each other two \hbox{DBMSes}, all without primary keys defined, and (d)~10 queries
from each of the four \hbox{DBMSes} on the primary key data set, for exploration
across all combinations, a total of 780 query instances. These query
We ran the \hbox{DBMSes} on the {\em
  change points}: consecutive cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
plans. The protocol retained about 78\% of the QEs.

%% The following corresponds to TTPv2.
We later reran the exploratory analysis using TTPv2, this time on a larger sample of queries and data sets:
(a)~200 queries of four \hbox{DBMSes} without primary keys defined and
(b)~100 queries from each of the four \hbox{DBMSes} on the primary key dataset,
for exploration across all combinations, a total of 1,200 query instances (= 300 $\times$ 4 (DBMSes)).
We termed this sixth experiment ``Exploratory.''  We ran the DBMSes on the {\em
  change points}: consecutive cardinalities (10K apart; 300 for one DBMS) with different
plans. This experiment required 1,663 hours. The experimental
methodology retaining 12,100 {\QatC}s (3.7\% were dropped), covering 1,123 query
instances (6.4\% were dropped). Only 443 strict monotonicity violations (0.78\%) and 301 relaxed
ones (0.54\%) were observed. This exploratory analysis allowed us to refine
the operationalizations.

The seventh and final experiment was used for confirmatory analysis of the model and thus was
called ``Confirmatory.'' This was the most time-consuming of
the experiments. Here we
used (a)~800 queries on the data set without primary keys defined and (b)~510 of
those queries that had joins on the primary key attributes, on the data set
with primary keys, 
(c)~100 queries on the data set without skew and primary keys,  
(d)~100 queries with a subquery on the data set without primary keys,  
(e)~100 queries on the data set with primary keys and secondary indexes defined, 
(f)~100 queries with a subquery on the data set with primary keys defined,  
(e)~100 queries with a subquery on the data set with primary keys and secondary indexes defined, and 
(g)~100 queries on the data set with primary keys and secondary indexes defined, 
for a total of 1,910 queries and a total of 7,640 query instances (over the
four DBMSes).
Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed, roughly 13 per query.
\todo{Young, where did this number come from?:}The protocol accurately timed 53,547
{\QatC} pairs as change points.\

In this Confirmatory Experiment, we observed 3,347 strict
monotonicity violations (0.74\%), and 1,966 relaxed monotonic violations
(0.43\%) (out of a total of 452,684 pairs), which
provides further confidence that monotonicity also applies to
operators found in queries over data in various contexts (as described above) 
and that our operationalization of suboptimality is a valid one.

In the remainder of this section, we focus using the measured independent,
latent, and dependent variables in the Confirmatory experiment
to test predictions that arise out of our causal model in Figure~\ref{fig:model}.

\subsection{Descriptive Statistics}
Several initial conclusions can be drawn from this confirmatory experiment, which was the
result of several years of programming effort and about 30 months of
experimental runs. First, {\em every} \hbox{DBMS} exhibits suboptimality. Thus, this
phenomenon is likely to be a fundamental aspect of either the algorithm
(cost-based optimization) or the creator of the algorithm (human information
processing). Our model includes both effects. Perhaps surprisingly, 
%a little over half (1579 queries out of the 3056 query instances
more than half (3,933 queries out of the 6,983\todo{Rick: I have 6967 in my set since 16 were dropped 
because of missing one side of change point}  
query instances that emerged from the measurement protocol) exhibited suboptimality
somewhere in the range of cardinality of the varying table. 
Most instances of those queries (3,370) had the meximum value of suboptimality at a Q@C pair (i.e., level 3). 
\shorten{a big number for a query could be many pairs, each at a small
  distance, or a few pairs, each at level 3. How many queries had at least
  one pair with subopt>=3? Answer: 670 in 5.20, 36 in 5.3, 531 in 6.0. So,
  total of 1237; query used: select count(*) FROM (select distinct experimentid, experimentname, dbms, runid,
querynum from CF_Analysis_SuboptDefn_Ver1 where subopt>=3);
query used 7_10: select count(*) FROM (select distinct experimentid, experimentname, dbms, runid,
querynum from Cnfm_Analysis_SuboptDefn where subopt_sd>=3);}
CEPS ranged from 1 to 24,
plans across the cardinality range. The number of discontinuous operators observed in plans for a query
averaged 9.4, with 85 being the maximum. Number of repeats ranged from 0 to 108, with the mean being 4.99.

\subsection{Correlational Analysis}
We tested Hypotheses 1--7 using the strength and significance of correlations
of variables involved. These hypotheses 
predicts eleven positive relationships. Table~\ref{tab:hyptest} lists the hypotheses followed by the
correlation observed when testing each hypothesis.  (``NS'' denotes not
significant at the 0.05 level, the accepted standard for
significance. ``---'' denotes no prediction arising from the model.) As can
be seen, most of the predictions (nine) arising from the causal model are
supported and significant.  The two
exceptions are Hypotheses 1 and 2. Hypothesis 1 was not supported because
the correlation between number of operators available and suboptimality was negative,
while we predicted positive correlation.  \hbox{Hypothesis}~2 was not significant.

\subsection{Regression Analysis}
\todo{Sabah: revisit this entire section, as we changed our model}
We ran a regression over the independent variables of the
model that predict suboptimality over the data from the Confirmatory experiment. 
Our model explained \% of the variance of the suboptimality dependent
variable.

We also did regressions on the causal variables for CEPS, number of repeats and for number of
operators with discontinuity. Our model explained \% of the
variance for CEPS, \% of the variance for number of repeats, and \% of the variance for
discontinuity. \shorten{need path analysis}

Hypothesis 5 predicts that the presence of secondary indexes will be a moderator,
affecting the interactions between query complexity with plan space
complexity. It thus provides
predictions that the strength of such interactions will decrease in the
presence of secondary keys. When
secondary indexes were not specified, our model explains 
\shorten{\% of the variance
of suboptimality,} \% of the variance of CEPS, and \% of the variance
of discontinuity, with all of the independent variables. When
secondary indexes were defined on the underlying tables, our model explains
\% of the variance of suboptimality, \% of the variance of CEPS, and
\% of the variance of discontinuity. These findings are all consistent with our
hypothesis predicting a positive moderation effect of primary keys.

\todo{Sabah: revisit. Not sure about the purpose of this paragraph}
We also examined the individual contributions to each of these
three variables (e.g., suboptimality) from their causal variables (in the
case of suboptimality, from the four variables of correlation names, presence
of an aggregate, CEPS, and discontinuity). 
Table~\ref{tab:hyptest8} shows that the strength of all interactions goes
down as predicted, except for ones involving aggregates. Recall
that the predictive model indicates weaker effects generally for aggregates,
as they are evaluated late in the query, and the primary key attributes are
not necessarily included in the grouping attributes.

\todo{Sabah: redo hypotheses in these tables, for example, H1 doesn't go to
  subopt, then add the confirmatory numbers after we discuss them}
\begin{table}[t]
\tbl{Testing Hypotheses 1--7: Correlations on the Confiratory Study\label{tab:hyptest}}{%
%\begin{center}
\resizebox{85mm}{!}{
\begin{tabular}{c|c|c|c|c}
{\em Variable}& {\em Suboptimality}&{\em Repeats}&{\em CEPS} & {\em Discontinuity}\\
\hline
Operators in \hbox{DBMS} & ---&  {\bf H1a:}$-$0.16 & {\bf H1b:} & {\bf H1c:} \\
Correlation names & {\bf }  & {\bf H2a:} 0.54 & {\bf H2b:}  &{\bf H2c:} 0.49\\
Skew & {\bf H3:}  & --- & ---  & ---\\
Presence of aggregate & & {\bf H2d:}  & {\bf H2e:}  & {\bf H2f:} \\
Presence of primary key & & {\bf H2g:}  & {\bf H2h:}  & {\bf H2i:} \\
Presence of subquery & & {\bf H2j:}  & {\bf H2k:}  & {\bf H2l:} \\
Repeats& {\bf H7a:}  & --- & {\bf H6:}  & ---\\
CEPS &  {\bf H7b:} 0. & --- &  --- &{\bf H4:}  \\
Discontinuity & {\bf H7c:} 0.31 & --- & --- & ---\\
\end{tabular}
}
%\end{center}
}
\end{table}

\begin{table}[t]
\tbl{Testing Hypothesis 5: Interaction Strength on the Confirmatory Study\label{tab:hyptest8}}{%
%\begin{center}
\resizebox{85mm}{!}{
\begin{tabular}{c|c|c|c|c|c|c}
{\em Variable}& \multicolumn{2}{|c|}{{\em Repeats}}&\multicolumn{2}{|c|}{{\em CEPS}} & \multicolumn{2}{|c}{{\em Discontinuity}}\\
%\hline
{\em }&{Not SI\em }&{SI\em }&{Not SI\em }&{SI\em }&{Not SI\em }&{SI\em }\\
%&{\em }&{\em }&{\em }&{\em }\\
\hline
Correlation names &  & 0. &0. &0.  & 0. & 0. \\
Presence of aggregate  & 0. & 0. & 0. & 0. & 0. & 0. \\
Presence of primary key  &  & 0. & --- & --- & --- & --- \\
Presence of subquery  &  & 0. & --- & ---  & --- & --- 
\end{tabular}
}
%\end{center}
}
\end{table}

\shorten{
In our experiments, we noticed that some of the \hbox{DBMSes} might at
different times pick a different plan for the same query at the same
cardinality. We addressed this in our experiment by running the optimizer
(with a JDBC {\tt PreparedStatement}) and then executing the query ten times
in fast succession. Each time we did an {\tt EXPLAIN PLAN}  to ensure that
the originally-chosen plan was indeed used (it was).

That said, at the next cardinality, the optimizer went into action again,
and so could have chosen a {\em different} plan for that new cardinality.
Our supposition is that such a {\em heuristic} optimizer introduces
some variance in the cardinality dependent variable.

To test this, we ran regressions on just the \hbox{DBMS} that was demonstrably {\em
  not} heuristic, for the common case of primary keys. Interestingly, only
the presence of an aggregate and CEPS were significant at the 0.05 level, yet
our model explains fully 53\% of the suboptimality variance (and 56\% of the
CEPS variance and \todo{Sabah:??} of the discontinuity variance). This might imply
that heuristic optimization itself introduces perhaps 10--20% of the variance.

The size of the regression coefficients also provides useful
information. From the regressions run on all of the data, for the
statistically significant causal factors, the highest coefficients, all in
predicting the suboptimality dependent variable, were CEPS at 0.42, number
of tables involved at 0.23, number of discontinuous variables at 0.16,
number of operators available in \hbox{DBMS} at 0.06. Note though that number of correlation
variables is highly correlated with CEPS; the regression will apportion the variance of the dependent
variable among these factors.
}

\subsection{Summary of Model Testing}
In our experimental design, we started with a structural causal model that
encapsulates our theory for how cost-based query optimizers might select a
suboptimal plan for a query at a cardinality. This model implies the eight
specific hypotheses listed in Section~\ref{sec:model}. We then performed
four experiments to refine our operationalizations (such as discontinuity,
via the Exhaustive Experiment and number of operators available with the additional
Exhaustive with Keys Experiment), to test fundamental assumptions (such as
monotonicity, via the Monotonicity Experiment), and to test and make minor
refinements to our model (via the Exploratory Experiment). While exploring
the data, one must be cognizant of the possibility of Type 1 errors: false
positives that lead one to believe a relationship exists when it doesn't.

To control for Type 1 errors, we then performed the Confirmatory 
Experiment on a completely different data set consisting of many more 
query instances, 7,640 in all, running on four \hbox{DBMSes} that each
utilize cost-based query optimization, to test
our refined model. Statistical inference is only possible in confirmatory
analysis, where the model and hypotheses are selected a priori.

Correlational analysis provides significant support for the model, except
for Hypotheses 1 and 2, with the implication that the influence of the
number of operators available remains largely unexplored. Via regression
analysis, the model explains 34\% of the variance of suboptimality (overall:
with primary keys a little lower and without primary keys a little higher
due to that moderating effect), 31\% of CEPS, and 35\% of discontinuity. In
all but one case, the causal influence of independent and latent variables is significant
($p<0.05$). The direction of the regression coefficients, again in all but one
case as predicted, also provides strong support for the model.

Now that the causal model has been found to be supported by the confirmatory
analysis, we turn to possible implications of this model.

\subsection{Identifying Root Causes of Suboptimality}\label{sec:root}

Our goal in this paper has been to understand cost-based query optimizers
as a {\em general} class of computational artifacts and to articulate and test
a predictive model characterizing how such optimizers,
again, as a general class, behave.  This model can be used to further
improve \hbox{DBMSes} through engineering \hbox{efforts} that benefit from the fundamental
understanding that the scientific perspective can provide.

Our model includes three causal factors of suboptimality: \hbox{DBMS} optimizer
complexity, query complexity, and plan space complexity. Of these factors,
the regression \hbox{coefficient} that was highest was for CEPS
(cf.~Table~\ref{tab:hyptest8}: 0.51--0.54, normalized) The next \hbox{highest}
regression factor was number of tables involved, which is highly correlated
with CEPS (cf.~Table~\ref{tab:hyptest}: 0.54). These two observations imply that the
number of plans being considered is a major determinate of
suboptimality, indicating that choosing among many plans is hard, despite
many decades of research and development.

The next most influential factor is the number of discontinuous operators
(cf.~Table~\ref{tab:hyptest8}: 0.40--0.41, normalized),
\hbox{implicating} the cost model.

\shorten{Sabah, how does the following sound?}The one following that is the number of operators provided by the \hbox{DBMS}.
This factor is totally correlated with the \hbox{DBMS}, and so
is confounded with other specificities of the coding of each \hbox{DBMS}. That said,
it is still the case that as the number of operators goes up, across
all the \hbox{DBMSes} studied, the amount of suboptimality increases. This
\hbox{factor} also strongly impacts CEPS, and so indirectly impacts suboptimality
through that path.

These factors implicate two broad root causes of suboptimality across \hbox{DBMSes}:
(i)~the cost model and (ii)~the plan search process.

\section{Diminishing Returns?}
In order to increase query performance, \hbox{DBMSes} are extended over time with
new \hbox{relational} query operators. Also over time, \hbox{DBMSes} are extended with new
storage and indexing structures,
which themselves elicit new query operators. Each subsequent generation of
the \hbox{DBMS} thus supports an ever-expanding collection of operators,
thus the current version is simply the most recent within a series of \hbox{DBMS}
{\em generations}.

\subsection{A Gedanken Experiment}\label{sec:gedanken}
Consider a Gedanken experiment that analyzes the plan for each query at
each cardinality, that is, for each Q@C, for each \hbox{DBMS} generation. In early
generations, there will be fewer plan changes as the cardinality increases
for a given query, just \hbox{because} there are fewer operators available. For
later generations, some of the Q@Cs will be associated with different plans,
enabled by the new operators added by subsequent generations. As our causal
model in Figure~\ref{fig:model} predicts, the effective plan space for a
query may grow in subsequent generations, due to the increased number of
operators available.

In many (hopefully most) cases, the new plan will be more efficient than
staying with the previous plan selected in a previous generation. After all,
that is the very reason the new operator(s) were added to that generation.
However, in the presence of suboptimality,
sometimes the new plan at that Q@C is {\em not} preferable,
as that \hbox{DBMS} generation's query optimizer selected the wrong
plan.  Indeed, the query optimizer also evolves and improves with each new
generation, in part to minimize the chance of selecting the wrong plan.

Our predictive model suggests that suboptimality is causally
impacted by the independent variable of CEPS (cardinality of effective plan
space\todo{Rick:Connect with number of operators available and with Section~\ref{sec:root}}. A
possible implication is that as the generations of the \hbox{DBMS} add more and more
operators, suboptimality will also increase. \todo{Rick:No: causality is with
  CEPS, not number of operators!}

\todo{Rick:Make point that our causal analysis is across \hbox{DBMSes} with different
  numbers of operators. Here we're talking about {\em within} a single \hbox{DBMS},
  across multiple generations, each with an additional operator, and thus a
  higher CEPS. In fact, it would be useful to see if generation number was correlated with CEPS.}

The underlying question then becomes, does an additional operator made
available in a subsequent generation of the \hbox{DBMS} actually help or hurt?

\subsection{Simulating \hbox{DBMS} Generations}
While we do not have access to prior generations of our \hbox{DBMSes} (which is why
the previous discussion was in the form of a Gedanken experiments), we can
approximate this with an experiment using data already collected on the
current version available for each \hbox{DBMS}, to {\em simulate} the prior
generations having a fewer number of operators, and thus a smaller set of
realizable query plans. (The effect will certainly be smaller in our
simulation, as we are nonetheless using the most recent {\em query
  optimizer}n in all of the simulated generations.)

Our data, drawn from the 8,840 query instances and their 112,118 Q@Cs in both the
exploratory and confirmatory experiments of the previous study, 
consists of pairs of adjacent plans
(let's refer to them as the {\em lower} plan and the {\em upper} plan) for
the same query at the {\em lower} and {\em upper} cardinalities) that are
consecutive (that is, separated by the minimum cardinality, either 10,000 or
300 rows), each with an actual execution time. All of Q@Cs having
cardinalities between the upper plan of one pair and the lower plan
of the next pair are associated with that same plan, guaranteed by the
process in which we chose those Q@Cs for actually timing.

For this generational experiment, we associate with each Q@C a generation
(a positive integer) that
is the earliest generation containing all the operators in the
plan. We also only consider Q@C pairs (with adjacent Q@Cs) where the lower
plan has a generation distinct from the upper plan.  Say the upper
generation is earlier than that of the lower plan, as illustrated in
Figure~\ref{fig:negback}. In this figure, as we scan from left to right in
increasing cardinality, we first encounter Plan A from generation 2 at the
first cardinality of 10K rows, then Plan B sometime later, then even later,
a pair of Q@Cs with Plan B at 900K rows and Plan A at the
adjacent 910K rows, then much later Plan A again at 1320K rows (this is the
measured Q@C that has Plan A that is closest to the measurement at 910K rows).

\begin{figure*}[t]
%\centering
%originally 30pc
\includegraphics[width=30pc]{figures/negback.pdf}
\caption{An Example of a Negative Backward Delta\label{fig:negback}}
\end{figure*}

In this case, the thinking goes that, had we been
in the \hbox{DBMS} generation 2, the optimizer would have selected Plan A
throughout, because Plan B simply wasn't possible (as it involves an
operator not present in generation 2). Then when generation 3 was created by
adding an operator, the optimized chose Plan B for 900K. The reason Plan B
was chosen is that it is faster than Plan A at that cardinality, shown in
the figure by extrapolating the time down from 910K down to 900K (we will
revisit this extrapolation shortly).

In this particular case, Plan B is more appropriate (faster) than Plan A,
but that is not the only possibility. Sometimes the later generation with
more operators available chooses the {\em wrong} plan, one that is slower
than the plan chosen by the earlier generation, due to the suboptimality
we've observed. (We'll examine an example shortly.)

The question then becomes, does a plan change by a subsequent generation
(enabled by one or more additional operators) represent a win (runs faster)
or a loss (runs slower, because of suboptimality)? More broadly, do the Q@Cs
in the aggregate enabled by each succeeding generation continue to overcome
the increasing burden of suboptimality?

\subsection{Characterizing \hbox{DBMS} Generations}
To which generation do we assign each \hbox{DBMS} operator?

Note that we don't actually know the specific order in which the operators
were added to each \hbox{DBMS}. But even if we did, that order was somewhat
arbitrary, with a host of considerations going into those decisions over the
years. Given that we are using the same optimizer for each such defined
generation, we'll adopt a more systematic ordering of the operators. We
start by gathering all {\em single-operator} plans, all two-operator plans,
and so forth, and order the generation by the prevalence of their
\hbox{appearance} in these query plans.

Specifically, for each \hbox{DBMS} we designate the first
generation to contain the single operator that maximizes the number of plans
at change points, that is, maximizing the number of Q@Cs, containing just that
operator. So for example, all the plans generated by MySQL that have exactly
one operator involve just the Full Table Scan operator.  So no choice was
needed: we designate the
first generation as just having that one operator. Any plan with just that
operator can be constructed by MySQL generation 1, as well as by any
subsequent generation (as each generation includes that initial operator).

We then examine the plans containing exactly two operators (in the case of
MySQL, one of which is
full table scan, as a two-operator plan not having that operator would
necessarily be constructed by generation 3 or higher). Using MySQL again, there are two sets of
such plans, those with the full table scan and full table scan with join
operators (888 Q@Cs) and
those with the full table scan and ref operators (112 Q@Cs). We pick full
table scan with join as the operator added by Generation 2, given its
prevalence of Q@Cs (a total of 1705 in the full set). Each
subsequent generation adds that operator that maximizes the number of plans
that operator will eventual enable. So Generation 3 adds the eq\_ref operator,
as that operator enables 633 plans eventually. Generation 4 adds the ref
operator and Generation 5 adds the index operator. There are six plans that
involve all five operators.

The generations of MySQL thus can be characterized from the Q@Cs we encountered: five distinct combinations of two
operators, covering 1086 Q@Cs, six combinations of three operators (1126 Q@Cs),
two combinations of four operators (104 Q@Cs), and (naturally) exactly one
combination of all five operators (6 Q@Cs). Thus for MySQL our data implies
a simulation of five distinct generations, each adding an operator.

\begin{align*}
g_1 &= \{ \hbox{\em full table scan} \}\\
g_2 &= \{ \hbox{\em full table scan}, \hbox{\em scan with join} \}\\
g_3 &= \{ \hbox{\em full table scan}, \hbox{\em scan with join}, \hbox{\em eq\_ref} \}\\
g_4 &= \{ \hbox{\em full table scan}, \hbox{\em scan with join}, \hbox{\em
  eq\_ref}, \hbox{\em ref} \}\\
g_5 &= \{ \hbox{\em full table scan}, \hbox{\em scan with join}, \hbox{\em
  eq\_ref}, \hbox{\em ref}, \hbox{\em index} \}
\end{align*}
For the four \hbox{DBMSes} in our study, the number of generations ranged from five
to thirteen.

\subsection{Using Change Points}
We now consider how to use data already collected, that is the {\em change
  points} discussed in Section~\ref{sec:experiments}: consectutive
cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
plans. Specifically, how can change points be used to evaluate the
effectiveness of different generations of a \hbox{DBMS}?

Define $\hbox{\em ops}(p)$ for a given plan $p$ to be the set of operators
present in that plan, with some operators perhaps repeated in that plan. 
For example, plan $p_1$ may be composed of a {\em select} of an {\em agg} of a {\em select} over a
relation. Hence,  $\hbox{\em ops}({p_1}) = \{\hbox{\em select}, \hbox{\em
    agg}\}$. A generation $g$ is {\em applicable} to a plan $p$ if
  $\hbox{\em ops}(p) \subseteq g$,  and so it is the case that $g_4$  is applicable to $p_1$. By
  definition, if a generation is applicable to a given plan, it is
  applicable to all subsequent generations. The smallest such generation is
  termed the {\em minimally applicable generation}, or {\em mingen}, so
  $\hbox{\em mingen}(p_1) = g_3$.

\shorten{We couldn't make this one work...
\todo{Young:replace in the above para with an actual SQL example query that was associated
  with two MySQL generations in a Q@C pair. use the generation numbers from
  the next para}

\todo{Rick: I found one example but the cardinality information 
was different than what's stated in the current prose...}

%%runid = 1097, querynum = 46, max_card = 60000, upper_card = 29100, lower_ard = 28800, upper_gen = 4, lower_gen = 5
%%max_cqt: 230, upper_cqt: 110, lower_cqt: 120, lower_extrapolated_cqt: 110
%% experimentname: op-pk-30K-100q-idx, pk: true, sec_idx: true
%% plan 0: ALL:Full Table Scan,ref
%% plan 1: eq_ref,index,ref
%% plan 0 at upper_card, plan 1 at lower_card
\noindent
\hspace{3ex}{\small\begin{verbatim}
        SELECT t3.id2, t3.id4, t0.id3, t2.id1 
        FROM ft_HT3 t2, ft_HT1 t1, ft_HT2 t3, ft_HT1 t0  
        WHERE  (t2.id2=t1.id2 AND t1.id2=t3.id1 AND t3.id1=t0.id4)
\end{verbatim}}
}

First consider pairs for which $\hbox{\em
mingen}(\hbox{\em lower}) > \hbox{\em mingen}(\hbox{\em
  upper})$. An example is shown in Figure~\ref{fig:posback}. In this case,
the set of operators in Plan A for the Q@C at cardinality 900K requires at
least generation 3, whereas the set of operators for Plan B for the Q@C at
cardinality 910K requires but generation 2. (Note that the generations here
are consecutive, that that is not required. Sometimes the generations of
pairs of Q@Cs are wildly different.)

\begin{figure*}[t]
%\centering
%originally 30pc
\includegraphics[width=30pc]{figures/posback.pdf}
\caption{An Example of a Positive Backward Delta\label{fig:posback}}
\end{figure*}

To examine the wisdom of picking Plan A for 900K (whose time was measured as
the {\em higher} point shown (the one above the dashed line), we find the closest Q@C
also having Plan A. There are two illustrated here, at the minimum and
maximum cardinalities, of 10K and 2000K, respectively, with the closer one
at 10K. (If there is no such change
point, we can't do the extrapolation).

We use the slope of the dashed line between the measured query time of Plan
B at cardinalities 10K and 910K to get an estimated query time at 900k.
This realizes an estimate of how fast the query using Plan B would have run
on the database having the variable table with a cardinality of 900K. In
this particular case, Plan B looks like it would have run {\em faster} (in
fact, was was faster even at 910K in this particular example).

As illustrated in Figure~\ref{fig:query769}, plans can be discontinuous,
because their asymptotic complexity is $O(n \lg n)$ (though this often turns into
a sequence of linear relationships, as illustrated by this
figure). This behavior will, for plan B present at
the two granularities of 10K and 910K, slightly overestimate the runtime of
that plan at 900K, and thus {\em underestimate} the penalty of going with
Plan A rather than Plan B. That said,
given that we are extrapolating from the actual run time at a very
close cardinality, our extrapolated estimate should be close.

We term this a {\em backward extrapolation}, because we are extrapolating
backward from a cardinality of 910K to 900K.

With this estimate, we can compute the {\em relative delta}, defined as the
measured time of Plan A (the chosen one) at the lower cardinality minus the
extrapolated time of Plan B at that same cardinality, divided by the
measured time of Plan B. The relative delta is thus scaled by original
measured time at that granularity, and thus has a value between -1 and 1.

In this case, the relative delta is positive (the measured time is greater
than the extrapolated time, which indicates that the optimizer chose the
{\em wrong plan}: here, Plan B should have been chosen.  A negative
relative delta implies that the additional operator(s) available to the
minimally applicable generation of the upper plan were indeed beneficial,
that the measured time was lower. (In such cases, we again divide by the
larger value.)  A
positive relative delta implies the presence of suboptimality: the wrong
plan was chosen by the \hbox{DBMS} generation, perhaps due to the greater number of
operators available. (Note that this is a slightly more expansive definition
of suboptimality than that used earlier and illustrated in
Figure~\ref{fig:query769}.)

There are thus seven orthogonal possibilities for each change point (that
is, a pair of adjacent Q@Cs, a total of 66,769 pairs/change points).

\begin{enumerate}
\item The pair of Q@Cs share the same generation, that is, 
$\hbox{\em mingen}(\hbox{\em lower}) \neq
\hbox{\em mingen}(\hbox{\em upper})$ (62,903 Q@C pairs, 94.2\%), which we don't consider further.

\item There is no other cardinality having the same plan as that with the
later generation, in which case we also ignore this Q@C (768 Q@C pairs,
1.2\%).

\item The extrapolation yielded a computed query
time that was negative, which we also drop (10 pairs, 0.01\%).

\item The extrapolation was from above and indicates
no suboptimality, termed a {\em negative backward relative delta}, as
  exemplified in Figure~\ref{fig:negback}, examined earlier (492 pairs,
  0.7\%).

\item The extrapolation indicates no suboptimality, termed a {\em positive backward
    relative delta}, exemplified in Figure~\ref{fig:posback} (583 pairs,
  0.9\%).

\item The extrapolation was from below (consider Figure~\ref{fig:negback} but with
  Plan A from generation 5; we would then need to extrapolate from the
  closest Plan B, which is {\em at a smaller cardinality}, in a {\em forward
    direction}, to compute a {\em negative forward relative delta} (1218
  pairs, 1.8\%).

\item A extrapolation was a {\em positive forward relative delta} (795 pairs, 1.2\%).
\end{enumerate}
%1 of rel del: 2, -1 of rel del: 21 (neg backward:3, pos backward:18)
From these seven possibilities, we thus retain (4) and (6), which indicate a
suboptimal plan at the higher generation (1710 pairs), and (5) and (7), which indicate a
non-suboptimal plan at the higher generation 1378 pairs).

This analysis is for a single pair of adjacent Q@Cs for a single query
running on a specific \hbox{DBMS},
providing a relative delta for the minimally applicable generation. This is a percentage difference, and so is not \hbox{affected}
by the absolute magnitude of the run time nor of the cardinality in
question. Indeed, because it is a percentage difference, the relative delta
is not affected by the query nor even which \hbox{DBMS} is involved. We associate
this relative delta with the later generation, for it is that generation
which had the choice between the two plans for that query.

\subsection{Trends Across \hbox{DBMS} Generations}
Consider what might the {\em average relative delta} behave across the
generations, that is, how
the average of the relative deltas associated with each generation will
change with successive generations.
The average relative delta is a characterization of the aggregate
impact of that generation, providing a quantitative estimate of the benefit
of adding that operator. (We use average so that each point is not impacted
by the number of pairs over which that point is computed.)

What does our causal model in Figure~\ref{fig:model} say about this? That
causal model asserts that as the number of operators increases, the latent
measures in plan space complexity increase, which therefore impacts
suboptimality, also in a positive direction.

Our Gedanken experiment in
Section~\ref{sec:gedanken} takes this predicted behavior and predicts taht
as the generations of the \hbox{DBMS} contain successively greater numbers of
operators, suboptimality will also increase.

Drilling down into this interaction, the next question is:   By
how much is the increase in efficiency enabled by a new operator (for those
Q@Cs utilizing a plan containing that operator) greater than the decrease in
efficiency resulting from suboptimality (for a subset of those Q@Cs)
resulting from adding that operator?  Given that each new operator will
improve a shrinking subset of Q@Cs, while the suboptimality occurs in a
portion over an expanding subset of Q@Cs, it seems that there must be a
point where the decrease due to suboptimality obviates the increase enabled
by the new operator. How close are modern \hbox{DBMSes} to that limit?

This analysis implies that each additional operator will result in a net {\em
  decrease} in average relative delta, with the possible result that the advantage of that
additional operator exceeded the net impact of suboptimality. We thus
predict that the first derivative of a relationship between average relative delta over
increasing generation will be negative, a natural result of the efforts of
\hbox{DBMS} developers to increase performance over successive generations of their
\hbox{DBMS}.

That said, each new operator is applicable to a successively smaller portion
of the queries, and perhaps over a successively smaller portion of the
cardinality space.  (We note in passing that the space of queries over which
our particular experiment ranged is a rather small, though oft-used, portion
of the exceeding complex SQL definition: over 3000 pages~\cite{Melton11}.)
As already noted, our causal model predicts that the prevalence of suboptimality
would increase as the number of operators available increased. We thus predict that
the {\em second derivative} of the average relative delta for increasing
generation will be {\em positive}, that is, that the curve will start to
level off. If that indeed occurs, it raises the
possibility of the curve either asymptotically approaching a horizontal line
(the first derivative approaching 0), or worse: the first derivative
changing to positive, with the average time over the queries we are studying
actually {\em increasing} with \hbox{DBMS} generation.

We partition the relevant four sets of change points into two groups: (i)~those
with a {\em positive} relative delta (either forward or backward), denoting a
{\em suboptimal decision} by the query optimizer at the later generation
(correcspeonding to a higher generation number)
and (ii)~those with a {\em negative} relative delta (either forward or
backward) relative delta, indicating a {\em non-suboptimal decision} by the
query optimizer at the later generation. (We can't state unequivicably that
the plan is optimal, because there may be yet another plan involving the
operators within that generation that is even faster.)

Let's first examine those change points for which the query optimizer made a
good decision, the non-suboptimal change points; see
Table~\ref{tab:non-subopt}.

\begin{table}[t]
\tbl{Non-Suboptimal Change Points\label{tab:non-subopt}}{%
%\begin{center}
\begin{tabular}{c|c|c|c}
{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}\\
& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}\\
\hline
1 & --- & --- & 0.00\\
2 & --- & --- & 0.00\\
3 & --- & --- & 0.00\\
4 & 24 & 0.28 & 0.28\\
5 & 282 &0.118 & 0.130\\
6 & 309 &0.274 & 0.203\\
7 & 336 & 0.317 & 0.243\\
8 & 43 & 0.20 & 0.241\\
9 & 4 & 0.1 & 0.241\\
10 & 17 &0.29 & 0.241\\
11 & 357 & 0.178 & 0.225\\
12 & 6 & 0.2 & 0.225\\
13 & 0 & --- & 0.225\\
{\em Cumulative}  &1378 & ---& 0.225\\
\end{tabular}
%\end{center}
}
\end{table}

The first thing to note is that this data aggregates the results over the
four \hbox{DBMSes}, which had generations ranging from five to 13. Thus this is the
first study we are aware of that compares the generations trends of \hbox{DBMSes}
over quite disparate code bases, thus getting at fundamental trends.

Overall, only a small number of change points, 1378, or about 2\% of the
total, satisfied the
requirements listed above, including having a different generation number for
the lower and upper plans. In fact, there were no such change points for the
first three generations nor the last generation.

While the numbers jump around a good bit, the thing to focus on is the {\em
  trend} as the generations evolve.  The basic trend is a slightly
decreasing relative delta, with the cumulative starting off at a high of
-0.28 (recall that high negative is good) and slowly decreasing to -0.23,
resulting from larger negative numbers trending to lower negative
numbers. The least squares slope is -0.024, which is consistent with the
transition from a generation being net benefitial to actually being slightly
suboptimal in generalat around generation 10.

This general behavior matches our prediction arising from the structural causal model that it
gets harder to squeeze out performance gains as operators are added to the
\hbox{DBMS} over successive generations.

We now examine those change points for which the query optimizer made a poor
decision, in that we can surmize that there was a better plan (the one right
next to it in the adjacent Q@C pair). Table~\ref{tab:subopt} provides the
same information across the \hbox{DBMS} generations for the suboptimal change
points: those for which the
  relative deltas are positive, indicating a poor decision. While there are
  still only a small number of change points, there are a greater number of
  {\em suboptimal} change points, with more relatively showing up at more
  recent generations. But more strikingly, the cumulative relative delta has
  the opposite behavior to the non-suboptimal change points: it {\em
    increases} over the generation, starting at 0.08 and ending at 0.25, a
  value {\em higher} than that of the non-suboptimal change points.

\begin{table}
[t]
\tbl{Suboptimal Change Points\label{tab:subopt}}{%
%\begin{center}
\begin{tabular}{c|c|c|c}
{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}\\
& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}\\
\hline
1 & --- & --- & 0.00\\
2 & --- & --- & 0.00\\
3 & --- & --- & 0.00\\
4 & 2 & -0.2 & -0.2\\
5 & 211 & -0.076 & -0.077\\
6 & 86 & -0.16 & -0.102\\
7 & 263 & -0.303 & -0.196\\
8 & 273 & -0.303 & -0.231\\
9 & 3 & -0.04 & -0.230\\
10 & 168 & -0.319 & -0.245\\
11 & 639 & -0.237 & -0.242\\
12 & 53 & -0.35 & -0.245\\
13 & 12 & -0.40 & -0.246\\
{\em Cumulative} & 1710 & --- & -0.246\\
\end{tabular}
%\end{center}
}
\end{table}

This is also as predicted: the optimizer is struggling with both more
options (plans over the available operators) to select from and a diminished
opportunity to make a significant improvement.

Looking at this struggle from a different perspective, consider the {\em
  maxium number of change points per query}, or {\em CPQ}, from the perspective of
\hbox{DBMS} generations. For each query, count the number of Q@Cs at each
generation (so for instance query Q1 could have 5 Q@Cs at generation 1, 7 Q@Cs at
generation 2, 17 at generation 5, etc.). We then then take the maximum CPQ
over the queries (so perhaps 17 is the maximum for generation 5 over all
queries). Our hypothesis is  that as the generation increases, the maximum
query flutter will increase, which then influences the maximum CPQ. The results are shown in Table~\ref{tab:maxcpq}.

\begin{table}
[t]
\tbl{Max Change Points Per Query, Per Generation\label{tab:maxcpq}}{%
%\begin{center}
\begin{tabular}{c|cr}
{\em Generation}&{\em Maximum Change Points}\\
& {\em Per Query}\\
\hline
1 & 1\\
2 & 66\\
3 & 115\\
4 & 136\\
5 & 132\\
6 & 141\\
7 & 42\\
8 & 172\\
9 & 2\\
10 & 12\\
11 & 55\\
12 & 29\\
13 & 1\\
{\em Cumulative} & 172\\
\end{tabular}
%\end{center}
}
\end{table}

What we notices is that the maximum CPQ {\em does} increase with generation,
up through generation 8, after which it falls off dramatically. In
retrospect, this fall-off makes sense. There can be only 200 Q@Cs for a
query, because we look for query plan changes only every 10,000 tuples up to
a maximum of 2M tuples. As the generation increases, there is less ``room''
for more plans (as some fraction of the previously generated plans will
still be quite good or even optimal), and so less opportunity for flutter
which will show up as a large maximum CPQ. And indeed, at generation 8,
there is a query that has an astonishingly high number of change points,
172, all of that generation.

There's another limitation we're hitting, one that is much more fundamental.

\begin{figure*}[t]
%\centering
%originally 30pc
\includegraphics[width=30pc]{figures/gen_benefit.pdf}
\caption{Cumulative Average Relative Delta over Generations\label{fig:benefit}}
\end{figure*}

The center graph shows the ``net benefit,'' that is, the cumulative average
relative delta
of both the suboptimal decision and non-suboptimal decision change
points. (Note that each point is not simply the sum of the non-suboptimal
and suboptimal cumulative
deltas, as the number of change points differ for each generation, but rather the cumulative
relative delta over all of the change points for that and prior generations.) There is a dramatic
rise at generation 5, with the rest of the generations presenting a
relatively smooth growth, moving from negative relative delta (not
suboptimal) to positive relative delta (suboptimal), crossing around
generation 10. This highly aggregated result, extracting the 3000-odd
change-point pairs from almost 100-thousand Q@Cs having the specified
properties of interest, implies that these four particular \hbox{DBMSes}, taken
together, might be close to hitting the wall, where adding an operator
actually slows down the average query.

More specifically, we can't say whether any of these DBMSes have actually
transitioned to where, for this class of queries, the errors of the
suboptimal change points overwhelm the benefits of the non-suboptimal change
points. Presumably, the DBMS vendors have done extensive tests to ensure
that the operator added to each generation did in fact effect a speedup on
the representative workloads that they use in evaluating their optimizer
enhancements. However, we {\em can} say that (a)~the trends observed
strongly point to a decreasing benefit and an increasing cost. as predicted
by the simple arguments made above, and (b)~if current DBMSes haven't yet
reached the point of diminishing returns, that possibility exists.

We should emphasize all the provisos mentioned earlier. We are looking at
quite simple queries, over a quite limited range of data, with only a small
percentage (3\%) of change points identified in our experimental
protocol. On the other hand, simpler queries are thought to be easier to
generate good plans than more complex queries (as Hypothesis 2 in
Section~\ref{sec:interactions} states), relational data is generally much less uniform
in its structure (we used only integers), values (the values in our
tables are quite evenly distributed), schema (the schemas of our tables
are identical), and range of sizes (the smallest table is 1/200-th of the
largest table), which should minimize the suboptimality.

\section{Engineering Implications}\label{sec:engineering}
We studied a particular
phenomenon, that of the \hbox{optimizer} selecting a wrong plan, indicated by the existence of a query plan that performs more efficiently
than the DBMS's chosen plan, for the same query. From the engineering
\hbox{perspective}, it is of critical importance to understand the prevalence
of suboptimality and its causal factors. The genesis of our model was a sense that suboptimality
is caused in part by the inherent complexity of these system and the
concomitant unanticipated interactions between various rules in the
optimizer.

Through a series of experiments managed by \azdb, we uncovered
several surprising results that provide systematic clues as to where current optimizers come up
short and how they can be further improved.

\begin{itemize}
\item For many queries, a majority of the ones we
considered, the optimizer picked the wrong plan for at least one
cardinality, even when the cardinality estimates were completely accurate
and even for our rather simple queries.

\item A quarter of the queries exhibited significant suboptimality ($>= 20\%$ of
        the runtime) at some cardinality.
\end{itemize}
\noindent
These two results indicate that there is still research needed on
this topic. Fortunately, the causal model helps point out
specifically where that research should be focused.

\begin{itemize}

\item Many queries exhibited {\em query fluttering}, in which the query
  optimizer returned to a previous plan at a higher cardinality.
\item Some queries exhibited significant {\em query thrashing}, with a plan change at
         almost every cardinality. While this phenomenon was first visualized
         by Haritsa et al.~\cite{harish07,Haritsa10} on some complex
         queries, we have shown that it is present even in a surprising
         percentage of simple queries.

\item Furthermore, some queries exhibited many changes
{\em to a suboptimal plan} as the \hbox{cardinality} was varied.
\end{itemize}
\noindent
These particular queries, as well as those of the right-hand side of
Figure~\ref{fig:suboptcumulative} exhibiting a large degree of suboptimality, can be a starting point for identifying the root
cause(s) of query thrashing. The phenomenon can be investigated initially on a
per-DBMS basis. Our methodology could then be
used to test proposed causal mechanisms of query thrashing across \hbox{DBMSes}, to
ascertain the generality of any proposed solutions.

\begin{itemize}
\item The causal model and our experimental results suggest that more
research is needed to improve the cost model of {\em discontinuous
operators}.

\item We also show that it may well be useful to explicitly take {\em \hbox{cardinality} estimate
  uncertainty} into account.

\item This research indicates that aggregates are {\em not} a problem, so
  that aspect of query optimization is in good shape.

\todo{Rick: evaluate following shorten:}\shorten{REVISE to characterize
  more precisely the relative
advantage or disadvantage of heuristic search\shorten{, and \hbox{(c)~research} is needed to
identify the inflection point where adding an operator decreases overall
\hbox{DBMS} performance}raise several unsettling concerns. These particularly simple
queries used only a few plans within any \hbox{DBMS}. Other more complex queries
will certainly use more plans. These results imply that as a \hbox{DBMS} becomes
more sophisticated, by adding operators, it also become more unstable, as
the prevalence of suboptimality grows. Perhaps there is a fixed level of
complexity beyond which a cost-based optimizer will be inexorably unstable.}
\end{itemize}

We see that costing of \hbox{discontinuous} operators is a
root cause of query suboptimality, and is thus particularly challenging to a
query optimizer. If the cost model is even a little bit off, the optimizer
might be on the wrong side of the ``jump,'' thereby selecting the wrong
plan.  That the presence of discontinuous operators has such a high
regression coefficient provides a quite specific guideline: more research is
needed to improve the accuracy of the cost model for such operators, such as
careful calibration that tunes the cost model with more accurate resource
knowledge, including the global memory capacity available, as well as to
improve the algorithms that allocate those buffer pages to specific
operators.

Concerning the plan search
process, another identified root cause of suboptimality,\shorten{one preliminary finding was that a heuristic
search process was implicated in 10--20\% of optimizer variance.} in cases
where the \hbox{DBMS} is not as sure about the cardinalities of the underlying
relations or the speed of the disk (e.g.,~if such relations migrated
frequently~\cite{reiss03}), 
perhaps the optimizer should\shorten{not try as hard to produce out the ``perfect'' plan. 
It might help if optimizers instead}
explicitly take uncertainty into account. Indeed, others have started to
\hbox{argue} that uncertainties in the query planning process should be
acknowledged and exploited~\cite{Babcock05}.

\shorten{The number of operators a \hbox{DBMS} supports is implicated in both root causes,
as well as explicitly in the causal model. Consider the following {\em
  Gedanken experiment}. We add an \hbox{operator} to a \hbox{DBMS}. Doing so will certainly
reduce the running time of some queries, those that benefit from the new
operator, and will not affect the running time of those query plans that do
not include the new operator. That sounds like a sure-fire path to improving
the performance the \hbox{DBMS}.

However, due to the factors discussed above, sometimes the query
optimizer will choose a non-optimal plan. Our analysis shows that as the number
of operators goes up, a suboptimal plan is chosen more frequently, so
the percentage of suboptimal plans will increase. Also, each new operator will
impact a smaller set of queries, but the set of suboptimal queries will grow.
At some point, the beneficial impact of each new operator will be
offset by the increased prevalence of suboptimality, and the \hbox{DBMS} will get
{\em slower, on average}. Of course, that inflection point is different for
each \hbox{DBMS}, but the fact that suboptimality is widespread across all the
\hbox{DBMSes} for the simple queries in this paper raises the possibility that we
are already there.
}

We mentioned dynamic query optimization in Section~\ref{sec:related}
earlier.  Dynamic query-reoptimization normally requires a significant
amount of information to be recorded during query execution, which can incur
non-negligible overhead on the overall query
performance~\cite{Avnur,kabra98}.  We envision that by utilizing the
proposed predictive model for suboptimality, it may be possible to enhance
reoptimization techniques such that given a particular query, a particular
data distribution, and a specific plan operator, just the important
statistics that affect the operator's performance can be identified and
should be recorded, thereby reducing the overhead of bookkeeping irrelevant
\hbox{information}.

Hence, the methodology introduced in this paper suggests fairly specifically where additional
engineering is needed (the cost model of discontinuous
operators and accommodating cardinality estimate uncertainty) and is not
needed (costing of aggregation).

\todo{Rick:discuss implications of generational study}

We emphasize that this section of this paper, considering engineering
implications of the underlying causal model, contrasts with the rest of the
paper, whose focus is on the science and on understanding these complex
systems at a fundamental level.  Good engineering should be built on solid
scientific results.

\section{Summary}\label{sec:summary}

This paper studies an important
component of a \hbox{DBMS}, the query optimizer. This component is an amazingly
sophisticated piece of code, but is still not well understood after decades
of research and development.

This paper makes the following contributions in
an \hbox{attempt} to gain new understanding of this component.
\begin{itemize}
\item Shows that even for simple queries, over a simple schema and
  relatively small range of sizes, the prevalence of {\em query
  suboptimality}, {\em query flutter}, and {\em query thrashing}, three problems that have not been systematically investigated across \hbox{DBMSes}, is
  high, and thus there is still research needed on this mature topic of
  query optimization.

\item Introduces a new {\em methodological
perspective} that treats \hbox{DBMSes} as experimental subjects within {\em empirical generalization}.

\item Proposes {\em operationalizations} of several relevant measures that apply
  even to proprietary \hbox{DBMSes}, as well as an overarching {\em predictive
    model} that attempts a causal explanation of suboptimality, encoding
   some of what is known about query optimization.

\doubleblind{\item Utilizes a novel {\em research infrastructure}, \azdb, to express, run, and
  analyze experiments.}{}

\item Tests {\em seven hypotheses} deductively derived from the predictive causal model. A
correlational analysis and a regression analysis
provided {\em strong support} for our model, across \hbox{DBMSes}, thus
qualitatively confirming what was informally known.

\item Provides a {\em space of hypotheses} that in concert through further
  investigation can refine and support (or not) this model or suggest a
  better model.

\item Uncovers compelling {\em evidence} (a)~that
suboptimality correlates with two operationalizations of query complexity,
(b)~that suboptimality correlates with two operationalizations of plan space
complexity, (c)~that query complexity is a contributor to plan space
\hbox{complexity,} and (d)~that schema complexity, as operationalized by the
presence of primary key attributes, moderates these three
interactions.\todo{Check again}

\todo{Keep the following shorten?}\shorten{\item A separate analysis suggested that heuristic plan selection
  appears to also be a source of suboptimality.}

\item For the kinds of queries we looked at, the factors that we
identified in our model: optimizer complexity, query complexity, and plan
space complexity, in concert predicted a significant portion of the variance
of suboptimality.  It is doubtful that any other factor, as yet unknown,
will itself predict as much variance as the factors we studied in this
paper. That said, it is certain that there remain several unknown causal factors;
identifying those factors will undoubtedly also have important engineering
implications.

\item Articulates for the first time a possible {\em upper bound} on the
  number of operators a \hbox{DBMS} may be able to support, given that the
  empirical evidence suggests that additional operators speed up a smaller
  and smaller portion of the query/cardinality space while incurring an
  increasing chance of suboptimality over the remaining space, which is
  growing.

\item Applies a novel experiment over the pairs of adjacent Q@Cs to show
  \todo{Rick:what?}

\item Provides a {\em path toward scientific progress} in the
  understanding of a key enabling technology. 
  It is important to emphasize that our model doesn't apply to just one implementation of the
algorithm or to one \hbox{DBMS}.  Rather, it is quite broad, applying to any \hbox{DBMS}
with a cost-based optimizer.

\item Thereby identifies {\em specific directions for engineering interventions}.

\end{itemize}
This paper thus suggests a framework of casual model elaboration and
directed engineering efforts that could reduce the prevalence of query suboptimality to an acceptably
low level.

\section{Future Work}\label{sec:future}
There are several directions we wish to take this work. With the model refinements
we propose here, additional directed pointers to engineering efforts should
emerge.

\todo{Rick: keep following shorten?}\shorten{restate: Again, the purpose of our work is to present
  a methodological perspective that can be further improved to study more
  complex interactions present in the input to query optimizers.}

\shorten{Operationally, doing experiments over four \hbox{DBMSes} was challenging. We
had to install and configure each of these \hbox{DBMSes} and get our
experimental apparatus to work uniformly across these systems. But in terms
of testing the predictions, four is a small number, which meant that we had
insufficient data to resolve Hypothesis 2 and had a low correlation for
Hypothesis 1. We plan to augment \azdb\ to enable other \hbox{DBMSes} to be
studied; our theory makes specific predictions on what we may observe with
these other systems. Specifically, the interaction between number of \hbox{DBMS}
operators and both plan complexity and suboptimality is difficult to tease
apart, as that independent variable is entirely correlated to \hbox{DBMS}. Adding
\hbox{DBMSes} to our empirical study should help understand this interaction
better. In addition, we plan to apply path analysis, to separate direct and
indirect effects.}

We want to look into query flutter and thrashing in greater detail, as those phenomena
provide concrete indicators of problematic optimizer behavior. One
possible methodological approach is to utilize SQL optimizer hints, such as
``{\tt \*+ SEMIJOIN}'' in MySQL and ``{\tt enable\_hashjoin(false)}'' in PostgreSQL,
to encourage the optimizer to produce more plans at a given cardinality,
that can then be timed to make more explict the entire plan space (recall that
CEPS, the cardinality of the effective plan space, is no greater and
probably much smaller than the cardinality of the plan space).

The investigations in this paper were based on very simple SPJ (select-project-join)
queries.  We wish to also consider {\em schema complexity}: 
foreign keys and indexes and {\em query complexity}: complex predicates, subqueries,
and user-defined data types, methods, and operators. \c2j{}{We also want
  to manipulate \hbox{DBMSes} internally (at least for those that are open-source),
  turning on and off the rules and observing suboptimality.} Our causal
model is extensible, in that we can add other factors, as long as their proper
operationalization can be established, and additional causal links.
So for
sub-queries, we can add nesting level, type of sub-query (scalar,
correlated, etc.), and number of ``effective joins''  (as some queries can
be rewritten into joins with the outer level).
It would also be interesting to study how a suboptimal subquery can affect
the suboptimality of the containing query.  For instance, is it true that if
many subqueries are themselves suboptimal, does that causally impact whether
the overall query is suboptimal? Finally, we may be able to determine
whether certain query rewrite techniques are employed within each \hbox{DBMS},
introducing another ``\hbox{DBMS} complexity'' factor into our model.

Kabra and
\hbox{DeWitt} have identified another source of complexity: inaccurate
statistics on the underlying tables and insufficient information about the
runtime system: ``amount of available resources (especially memory), the
load on the system, and the values of host language
variables.''~\cite[p.~106]{kabra98}.  Might there be other unanticipated
interactions, that are unknown simply because they haven't been
looked for?

\shorten{During our long-running experiments, the operating systems on which
the experimental subject \hbox{DBMSes} are installed were upgraded. This event caused
a significant increase in the number of phantom processes observed
during query executions, forcing us to eliminate many query executions for
statistical analysis, in that our model does not yet factor in a proper
measurement for the effects of phantom processes. But our proposed model
can be enhanced so that various additional parameters, such as operating system
configurations, can be examined under regression analyses, once these
parameters are properly operationalized.}

By employing an extensible causal model, many complex factors can be studied
via a systematic, statistically sound, scientific manner to better
understand the causal factors and their interactions.

In addition to the refinements of the causal model just discussed, our
research has provided specific directions for implementation interventions:
refining the cost model of discontinuous operators, improving buffer
allocation algorithms, and accommodating cardinality estimate uncertainty.

\todo{Rick: revise: talk about generational study, how we're hitting a wall,
  and how it might be necessary to fundamentally revisit query optimization,
  to come up with an entirely new approach that is less impacted by number
  of operators and by CEPS and thus avoids the flutter phenomenon.}

The methodology introduced in this paper and the causal model that results
has suggested both the scope of the problem of query suboptimality and a
number of specific engineering efforts that can now be carried out. Our
ultimate goal is a refined causal model that can fully explain how query
suboptimality arises in cost-based optimizers, thereby enabling engineering
solutions that reduce and eventually
effectively eliminate query suboptimality.

\doubleblind{\todo{Rick:Consider also mentioning somewhere Young's ergalics study
  on another kind of thrashing.}}{}

\doubleblind{\section{Acknowledgments}
This research was supported in part by NSF grants \hbox{IIS-0639106},
IIS-0415101, and EIA-0080123. We thank Rui Zhang for his help in initiating
this research, Ricardo Carlos, Preetha Chatterjee, Pallavi
Chilappagari, Jennifer Dempsey, David Gallup, Kevan Holdaway, Matthew Wong
Johnson, Andrey
Kvochko, Siou Lin, Adam Robertson,
Lopamudra Sarangi, Linh Tran, Cheng Yi, and Man Zhang for
their contributions to the \hbox{\azdb} and Phil Kaslo, Tom Lowry, and John
Luiten for constructing and maintaining our experimental instrument\c2j{}{,
  a laboratory of six machines and associated software}.
}{}

%\section{Electronic Appendix}
%The online electronic appendix to this article can be accessed in the ACM
%Digital\linebreak Library. That appendix lists...
Appendix~\ref{sec:app} provides further details on the experiments.

%\vfill\pagebreak
%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.

\bibliographystyle{abbrv}
\newcommand{\etalchar}[1]{$^{#1}$}
\begin{thebibliography}{99}

\vspace{0.1em}
\bibitem
%{Avnur:2000:ECA:342009.335420}
{Avnur}
R.~Avnur and J.~M.~Hellerstein, ``Eddies: Continuously Adaptive Query
Processing'', in {\em Proceedings of the ACM SIGMOD Conference},
pp.~261--272, 2000.

\bibitem
%[Babcock and Chaudhuri 2005]
{Babcock05}
B.~Babcock and S.~Chaudhuri, ``Towards a Robust Query Optimizer: A
Principled and Practical Approach,'' in {\em Proceedings of the ACM SIGMOD
  Conference}, pp.~119--130, Baltimore, Maryland, 2005.

\bibitem
% [Campbell, 1988 #829]
{campbell88}
D.~J.~Campbell, ``Task Complexity: A Review and Analysis,'' {\em Academy of Management}, 13(1), pp.~40--52, 1988.

\bibitem
%[Chaudhuri 1998]
{Chaudhuri98}
S.~Chaudhuri, ``An Overview of Query Optimization in Relational Systems,'' in {\em Proceedings of the ACM PODS Conference}, pp.~34--43, Seattle, WA, 1998.

\shorten{\bibitem
%[Cohen 1995]
{cohenbook}
P.~Cohen, {\bf Empirical Methods for Artificial Intelligence}, MIT Press, 1995.
}

\bibitem{TTPv1}
S.~ Currim, R.~T.~Snodgrass, Y.~-K.~Suh, and R.~Zhang,
``DBMS Metrology: Measuring Query Time,'' in {\em Proceedings of the ACM
  SIGMOD Conference}, pp.~421--432, June 2013.

\bibitem{TTPv2}
S.~ Currim, R.~T.~Snodgrass, Y.~-K.~Suh, and R.~Zhang,
``DBMS Metrology: Measuring Query Time,'' ACM {\em Transactions
on Database Systems}, 42+8 pages, October, 2016.

\shorten{\bibitem
%[Davies 1973]
{davies}
J.~T.~Davies, {\bf The Scientific Approach}, Academic Press, New York, 1973.

\bibitem
%[]
{gamma95}
E.~Gamma, R.~Helms, R.~Johnson, and J.~Vlissides, {\bf Design Patterns:
  Elements of Reusable Object-Oriented Software}, Addison-Wesley Publishing
Company, 1995.
}
\bibitem
%[Graefe 1993]
{Graefe93}
G.~Graefe, ``Query Evaluation Techniques for Large Databases,'' {\em ACM
  Computing Surveys} 25(2), pp.~73--170, June 1993.

\bibitem
%[Harish, Darera, and Haritsa 2007]
{harish07}
D.~Harish, P.~Darera, and J.~R.~Haritsa, ``On the Production of Anorexic
Plan Diagrams,'' in {\em Proceedings of the VLDB Conference},
pp.~1081--1092, 2007.

\bibitem{Haritsa10}
J.~R.~Haritsa, ``The Picasso Database Query Optimizer Visualizer,'' {\em
  PVLDB} 3(2):1517--1520, 2010.

\bibitem
% [Hevner et al 2004]
{hevner04}
A.~R.~Hevner, S.~T.~March, J.~Park, and S.~Ram, ``Design Science in
Information Systems Research,'' {\em MIS Quarterly} 28(1):75--105, 2004.

\bibitem{Hulgen03}
A.~Hulgen and S. Sudarshan, ``AniPQO: almost non-intrusive query optimization
for nonlinear cost functions,'' in {\em Proceedings of the VLDB Conference},
pp.~766-777, 2003,

\bibitem{Ioannidis96}
Y.~Ioannidis, ``Query Optimization,'' {\em ACM Computing Surveys}
23(1):121--123, June 1996.

\bibitem
%[Ioannidis 2003]
{Ioannidis03}
Y.~Ioannidis, ``The History of Histograms (abridged)'', in {\em Proceedings
  of the VLDB Conference}, pp.~19--30, September 2003.

\bibitem{SQL2008}
ISO, ``ISO SQL:2008 International Standard,'' 2008.

\bibitem{Jarke84}
M.~Jarke and J.~Koch, ``Query Optimization in Database Systems,'' {\em ACM
  Computing Surveys} 16(2):111--152, June 1984.
\bibitem
%[]
{kabra98}
N.~Kabra and D.~J.~DeWitt, ``Efficient mid-query re-optimization of
sub-optimal query execution plans,'' in {\em Proceedings of the ACM SIGMOD Conference}, pp.~106--117, 1998.

\bibitem
%[Mannino et al.~1988]
{Mannino88}
M.~V.~Mannino, P.~Chu, and T.~Sagar, ``Statistical Profile Estimation in
Database Systems,'' {\em ACM Computing Surveys}, 20(3), pp.~192--221, 1988.

\bibitem
%[Melton 2003]
{Melton03}
J.~Melton, {\bf Advanced SQL:1999}, Morgan Kaufmann, 2003.

\bibitem
%[Melton and Simon 1993]
{Melton93}
J.~Melton and A.~R.~Simon, {\bf Understanding the New SQL: A Complete
  Guide}, Morgan Kaufmann, 1993.

\bibitem{Melton11}
J.~Melton (editor), ISO/IEC 9075, Database Language SQL:2011 Part 2:
SQL/Foundation, December, 2011.
\bibitem
%[Moody 1998]
{moody98} D.~L.~Moody, ``Metrics for Evaluating the Quality of Entity
Relationship Models,'' {\em Proceedings of International Conference on
  Conceptual Modeling}, pp.~211--225, Springer, Singapore, 1998.

\bibitem
%[Ramakrishnan \& Gehrke 2003]
{Ramakrishnan03}
R.~Ramakrishnan and J.~Gehrke, {\bf Database Management Systems}, Third
Edition, 2003.

\shorten{\bibitem
%[Reddy and Haritsa 2005]
{reddy05}
N.~Reddy and J.~R.~Haritsa, ``Analyzing Plan Diagrams of Database Query
Optimizers,'' in {\em Proceedings of the VLDB Conference}, Trondheim,
Norway, pp.~1228--1240, September, 2005.
}
\bibitem
%[Reiss and Kanungo 2003]
{reiss03}
F.~R.~Reiss and T.~Kanungo, ``A Characterization of the Sensitivity of Query
Optimization to Storage Access Cost Parameters,'' in {\em Proceedings of the
  ACM SIGMOD Conference}, pp.~385--396, 2003.

\bibitem
%[Selinger 1979]
{Selinger}
P.~G.~Selinger, M.~M.~Astrahan, D.~D.~Chamberlin, R.~A.~Lori, and
T.~G.~Price, ``Access Path Selection in a Relational Database System,''
in {\em Proceedings of the ACM SIGMOD Conference}, pp.~23--34, 1979.

\bibitem
%[]
{winslett02}
M.~Winslett, ``David DeWitt Speaks Out,'' {\em ACM SIGMOD
  Record} 31(2), pp.~50--62, June 2002.

%\bibitem
%%{Kabra:1998:EMR:276304.276315}
%{Kabra}
%N.~Kabra, D.~J.~DeWitt, ``Efficient Mid-Query Re-Optimization of Sub-Optimal Query Execution Plans'', in {\em Proceedings of the ACM SIGMOD Conference},
%pp.~106--117, 1998.

%\bibitem
%%{Avnur:2000:ECA:342009.335420}
%{Avnur}
%R.~Avnur, J.~M.~Hellerstein, ``Eddies: Continuously Adaptive Query Processing'',
%in {\em Proceedings of the ACM SIGMOD Conference}, pp.~261--272, 2000.

\end{thebibliography}

\newpage
\appendix
\section{Details on the Experiments}\label{sec:app}
Table~\ref{tab:run_stat} lists the run statistics of the seven experiments
  used in this paper. In this appendix we provide more detailed information
  on the experiments.

Table~\ref{tab:run_stat2} gives some of those details. We'll walk through
the columns in succession.\todo{Young: please fill in last column}

\begin{table}[t]
\tbl{Experiments 1--7: Detailed Run Statistics\label{tab:run_stat2}}
{%
\resizebox{140mm}{!}
{
\begin{tabular}{c|c|c|c|c|c|c}
& {\em Experiment}& {\em Data Sets} & {\em Lab Shelves} & {\em What was} &{\em Was query}&{\em Number of}\\
&                 & {\em Used}      &                   & {\em Examined?}&{\em Timed?}   &{\em Retained (Raw) Q@Cs}\\
\hline
1 & Monotonicity 	& A & 6.0     &all cardinalities&yes& 12000 (12000)\\
2 & Initial Exhaustive 	& A & 5.19 + 6.0&all cardinalities&yes& 27948 (32000)\\
3 & Exhaustive 		& A & 7.1     &all cardinalities&no & 29515 (32000)\\
4 & Exhaustive with Keys & B& 6.0     &change points&no& --- \\
5 & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8171 (8842)\\
6 & Exploratory 	& A + B&7.1   &change points&yes& 12100 (12560)\\
7 & Confirmatory 	&A + B + C +D&7.1&change points&yes& 94502 (99558)\\
\end{tabular}
}
}
\end{table}

The third column states the data set used in each experiment, that is, the
specific tables being queried. There are four data sets, named A, B, C, and
D.

We first discuss the features shared between the four data sets. As introduced in
Section~\ref{sec:motivation}, the queries referenced tables {\tt ft\_HT1},
{\tt ft\_HT2}, {\tt ft\_HT3}, and {\tt ft\_HT4}. All four tables contain four
columns, each of type integer. The specific values of the
rows for all four columns depend on the percentage of skewed
data. Section~\ref{sec:datacomplexity} provides the algorithm for generated the
values for different values of skew; this algorithm is used in each column,
so all four columns in any row will have identical values.\todo{Young: is
  this true?}

There was one version of the
last three tables, for use with MySQL, with cardinality 60K, and one version
for the rest of the DBMSes, with cardinality 2M.

We generate 200 versions of {\tt ft\_HT1}, termed the {\em variable table}. For MySQL, these version contain
300, 600, 900, 1200, $\ldots$, 59,700, and 60,000 rows; for the rest of the
DBMSes, these versions contain 10,000, 20,000, 30,000, $\ldots$, 1,970,000,
and 2M rows, as introduced in Section~\ref{sec:motivation}.

We now differentiate the data sets, elaboring on the discussion in Section~\ref{sec:datasets}--\ref{sec:experiments}. Data Set C is perhaps the simplest to
describe: it specifies no primary key, has no duplicate rows, and has no skew
(of course, for any of the four tables). Data Set A differs from Data Set C
only in that there is skew.  As summarized at the end of
Section~\ref{sec:datacomplexity}, we use five skewness values: 0
(approximately), 0.001, 0.1, 0.5, and 1.0.

Data Set B is similar to Data Set A, adding the specification of the first
column as the primary key. And Data Set D is similar to Data Set B, adding
the specification that the other three columns should each be associated
with a secondary index, only only that one column.
We see the confirmatory examined a much larger variation of data sets than
the exploratory studies.

The next column of Table~\ref{tab:run_stat2} concerns the {\em Lab
  Shelf}. \hbox{\azdb} utilizes the metaphor of a bookshelf of lab
notebooks. Here, each shelf is associated with a version of \hbox{\azdb} itself. For the experiments in this paper, we used at vearious times over
the last three years labh shelves (that is, program versions) 5.19, 5.2,
6.0, and 7.1. Versions 5.19 and 5.2 were very similar; both implemented
TTPv1.  Version 6.0 also implemented TTPv1, but collected more query
measures that were not relevant for this paper. Version 7.1 implemented
TTPv2.

The \hbox{\azdb\ system also includes support} for {\em experiment scenarios}, which
are Java code that actually performs the experiment, such as varying the
cardinality and running different queries on the data. The only difference
in the scenario code across these experiments was in accommodating the
details of the data set (that is, creating secondary indexes and data skew)
and adding robustness features, such as using exponential backoff when the
network connection to the DBMS was temporarily lost.

The bottom line is that while the lab shelf and experiment scenario varied
somewhat, the only important aspect was the {\em Protocol} column of
Table~\ref{tab:run_stat}.

We now turn to the fifth column, ``what was examined?'' Here there are just two
possibilities, all 200 cardinalities or just the cardinalities at which the
query plan changed. The sixth column, ``wheat was timed?'', indicates that experiments three and four, Exhaustive and
Exhaustive with Keys, described in Sections~\ref{sec:discontinuity} and \ref{sec:experiments}, just
collected query plans, and thus took much less time to run.

The following sets of queries were used in the seven experiments.

\todo{Young: how does the following look? Any changes needed?}
\begin{description}
\item[QSa] 100 queries over the four tables, generated as described in Section~\ref{sec:querycomplexity}
\item[QSb] 100 queries, generated the same way
\item[QSc] 100 queries
\item[QSd] 100 queries
\item[QSe] 100 queries
\item[QSf] 100 queries
\item[QSg] 100 queries
\item[QSh] 100 queries
\item[QSi] 100 queries
\item[QSj] 100 queries
\item[QSl] A primary key query set consisting of 230 queries from {\em
  QSa}--{\em QSf}
\item[QSm] A primary key query set consisting of 160 queries: 40 queries
  drawn from each of {\em QSg}--{\em QSj}
%\item[QSlm] A primary key query set consisting of 390 queries from QSl and QSm
\item[QSn] A primary key query set 4 with 110 new queries
\item[QSo] A no-skew query set consisting of 100 queries without aggregates for no skew
\item[QSp] A subquery query set consisting of 100 queries, each with a subquery
\item[QSq] A subquery query set consisting of 100 queries, each with a subquery
\item[QSr] A primary key query set consisting of 100 queries drawn from {\em
  QSl} and {\em QSm}
\item[QSab'] 100 queries drawn from {\em QSa} and {\em QSb} for which the where
     selection involved a primary key column
\todo{Young what does this mean: ``This query set was revised to QSI and
  QSm.''? Als, it doesn't seem like this set is used anywhere...}
\end{description}

\todo{Rick: discuss last column}

Experiment 1 (Monotonicity) used the first 50 queries from QSa
for one DBMS plus the first six queries from QSa and two queries each from
{\em QSd} and {\em QSe} for MySQL, for a total of 60 query instances.

Experiment 2 (Initial Exhaustive) used the first 50 queries from {\em QSa}
for the three other DBMSes plus the ten queries for MySQL from Experiment 1, for
a total of 160 query instances. Experiment 3 (Exhaustive) used the same 160 queries.

Experiment 4 (Exhaustive with Keys) used the first 50 queries from {\em QSa}
for all four DBMSes, for a total of 200  query instances.

Experiment 5 (Initial Exploratory) used the (100) queries from {\em
  QSa}--{\em QSf} (for
one DBMS), the first 20 queries from {\em QSa}--{\em QSf} (for another DBMS), the first
10 (primay key) queries from {\em QSa} (for all four DBMSes), and the first 10
(non-primary key) queries from {\em QSa} (for the other two DBMSes), for a total of 780 query instances.

Experiment 6 (Exploratory) used {\em QSa} and {\em QSb}, plus the first 100 (primary
key) queries from {\em QSl} and {\em QSm}, for all four DBMSes, for a total of 1200 query
instances.

Experiment 7 (Confirmatory) used {\em QSc}--{\em QSr}, along with {\em QSp} for primary key and
subquery, {\em QSp} for subquery, {\em QSp} and {\em QSq} for primary key and secondary index and
subquery), all across all four DBMSes, for a total of 7,640 query instances.

\shorten{
\newpage
\todo{This is old material}

\input{OldTests.tex}

\input{Repeatability.tex}

\input{Measuring.tex}

\input{ExperimentSetup.tex}

\input{Discontinuous.tex}

\input{Results.tex}
}
\end{document}

% LocalWords:  PVLDB

### 2017-02-17 subopt-tods16/paper.tex
Sabah's updates

@@
 Presence of subquery & & {\bf H2j:}  & {\bf H2k:}  & {\bf H2l:} \\
-Repeats& {\bf H7a:}  & --- & {\bf H6:}  & ---\\
+Repeats& {\bf H7a:}  & --- & {\bf H5:}  & ---\\
 CEPS &  {\bf H7b:} 0. & --- &  --- &{\bf H4:}  \\
@@
 \begin{table}[t]
-\tbl{Testing Hypothesis 5: Interaction Strength on the Confirmatory Study\label{tab:hyptest8}}{%
+\tbl{Testing Hypothesis 6: Interaction Strength on the Confirmatory Study\label{tab:hyptest8}}{%
 %\begin{center}

### 2017-02-23 pt_characterization/recent_data/paper.tex
characterizing draft

--- 2017-02-13 pt_characterization/machine_cons_check/paper.tex
@@
 \begin{document}
-\title{Machine Homogeneity Check on EMPv5}
+\title{Histogram on Recent Runs by EMPv5}
 
@@
 
+\section{Description}
+This document presents histograms based on the most recent runs 
+of a program under test, called {\em INC}, over its increasing task lengths.
+
 \section{Experiment Notes}
+Table~\ref{tab:exp_notes} provides a short description of our experimental runs, 
+on which the following histograms are based.
+
 %% done
@@
 Machine & Task Length & Description & Time Length\\ \hline
-{\tt sodb8} &  INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
-{\tt sodb9} &  INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
-{\tt sodb10} & INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
-{\tt sodb12} & INC1$\sim$INC128 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10\\ \hline
+{\tt sodb9} &  INC1$\sim$INC1024 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10 / 2017-02-12 $\sim$ 2017-02-19\\ \hline
+{\tt sodb10} & INC2048 & Runs of 300 samples & 2017-02-12 $\sim$ 2017-02-20\\ \hline
 \end{tabular}
@@
 
-The following sections presents experimental results on a machine basis. 
-We used four nodes in a cluster. 
-Each section exhibits histograms of elapsed time (ET) and process time (PT) 
-over increasing INC's task lengths on each individual machine.
+Now we show histograms of elapsed time (ET) and process time (PT) 
+over increasing INC's task lengths.
 
-%\begin{table}[h]
-%\begin{center}
-%\begin{tabular}{|l|l|l|l}|} \hline
-%Machine & {\tt sodb8} & {\tt sodb9} & {\tt sodb10} & {\tt sodb12} \\ \hline
-%1  & \\ \hline
-%2  & \\ \hline
-%4  & \\ \hline
-%8  & \\ \hline
-%16 & \\ \hline
-%32 & \\ \hline
-%64 & \\ \hline
-%128 & \\ \hline
-%\end{tabular}
-%\end{center}
-%\caption{Machine-wise Run Statistics: Standard Deviation\label{tab:std_machine}}
-%\end{table}
+\newpage
 
-\input{sodb8/res.tex}
 \input{sodb9/res.tex}
-\input{sodb10/res.tex}
-\input{sodb12/res.tex}
 

### 2017-02-23 pt_characterization/recent_data/sodb9/res.tex
characterizing draft

--- 2017-02-13 pt_characterization/machine_cons_check/sodb9/res.tex
@@
-\newpage
-
-\section{{\tt sodb9}~\label{sec:sodb9_hist}} 
-This section exhibits histograms on the EMPv5 data obtained on {\tt sodb9}. 
+\section{Histograms~\label{sec:sodb9_hist}} 
+This section exhibits histograms on the EMPv5 data obtained when 
+the task length of INC increases from 1 second to 2048 seconds. 
 The detailed description of the base data are from Table~\ref{tab:exp_notes}.
@@
 
+\begin{figure}[hp!]
+	\centering
+	\subfigure[ET frequency on INC256]{
+		\includegraphics[scale=0.43]{sodb9/256_sec_et_hist_v5.eps}
+		\label{fig:inc256_hist_v5}
+	}
+	\subfigure[ET frequency on INC512]{
+		\includegraphics[scale=0.43]{sodb9/512_sec_et_hist_v5.eps}
+		\label{fig:inc512_hist_v5}
+	}
+	\subfigure[ET frequency on INC1024]{
+		\includegraphics[scale=0.43]{sodb9/1024_sec_et_hist_v5.eps}
+		\label{fig:inc1024_hist_v5}
+	}
+	\subfigure[ET frequency on INC2048]{
+		\includegraphics[scale=0.43]{sodb10/2048_sec_et_hist_v5.eps}
+		\label{fig:inc2048_hist_v5}
+	}
+	\caption{ET Histograms of INC256 ... INC2048~\label{fig:s9_et_hist3}}
+\end{figure}
+
 \newpage
@@
 	}
-	\caption{PT Histograms of INC16 ... INC64~\label{fig:s9_pt_hist2}}
+	\caption{PT Histograms of INC16 ... INC128~\label{fig:s9_pt_hist2}}
 \end{figure}
+
+\begin{figure}[hp!]
+	\centering
+	\subfigure[PT frequency on INC256]{
+		\includegraphics[scale=0.43]{sodb9/256_sec_pt_hist_v5.eps}
+		\label{fig:inc256_hist_v5}
+	}
+	\subfigure[PT frequency on INC512]{
+		\includegraphics[scale=0.43]{sodb9/512_sec_pt_hist_v5.eps}
+		\label{fig:inc512_hist_v5}
+	}
+	\subfigure[PT frequency on INC1024]{
+		\includegraphics[scale=0.43]{sodb9/1024_sec_pt_hist_v5.eps}
+		\label{fig:inc1024_hist_v5}
+	}
+	\subfigure[PT frequency on INC2048]{
+		\includegraphics[scale=0.43]{sodb10/2048_sec_pt_hist_v5.eps}
+		\label{fig:inc2048_hist_v5}
+	}
+	\caption{PT Histograms of INC256 ... INC2048~\label{fig:s9_pt_hist3}}
+\end{figure}

### 2017-02-23 subopt-tods16/paper_young.tex
characterizing draft

--- 2017-02-17 subopt-tods16/paper.tex
@@
 execution plan from a potentially large set of enumerated plans, all of
-which correctly compute the specified query. Infrequently the optimizer
+which correctly compute the specified query. Occasionally the optimizer
 selects the wrong plan, for a variety of reasons. {\em Suboptimality} is
 indicated by the existence of a query plan that performs more efficiently
-than the DBMS' chosen plan, for the same query. From the engineering
+than the DBMS's chosen plan, for the same query. From an engineering
 perspective, it is of critical importance to understand the prevalence
 of suboptimality and its causal factors.
-We study these aspects across \hbox{DBMSes} to identify the
-underlying causes. We propose a novel structural causal model to explicate the
+We propose a novel structural causal model to explicate the
 relationship between various factors in query optimization and
@@
 interactions among the components of the optimizer. This model induces a number of specific
-hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We observe that suboptimality prevalence
+hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We
+observe that \todo{Rick: check these}suboptimality prevalence
 correlates positively with the number of operators available in the \hbox{DBMS}
@@
 improvements of these optimizers. The results of this study also uncovered evidence of a potential fundamental
-limitation of cost-based optimization.\todo{be more specific!}
+limitation of cost-based optimization.\todo{Rick:be more specific! Couple
+  with contributions and conclusions}
 This paper thus provides
@@
 \keywords{}
-\acmformat{\doubleblind{Sabah Currim, Richard T.~Snodgrass, Young-Kyoon Suh, and
- Rui Zhang, 2015, \hbox{DBMS} Metrology: Measuring Query Time.}{A Causal Model of
+\acmformat{\doubleblind{Richard T.~Snodgrass, Sabah Currim, and Young-Kyoon
+    Suh, 2017, A Causal Model of \hbox{DBMS} Suboptimality}{A Causal Model of
     \hbox{DBMS} Suboptimality}}
@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 DBMSes underlie all information systems and hence optimizing their performance is of
@@
 
-This paper provides a thorough investigation into \hbox{DBMS} {\em suboptimality},
-when the \hbox{DBMS} chooses a slower plan over a faster plan for a given query. We systematically examine the factors influencing the number of suboptimal
-queries. There could be multiple causes of the suboptimality. One possible
-cause could be some peculiarity within the tens of thousands of lines of
-code of that query optimizer. Another possible cause could be the query's
-complexity. A third possible reason could be some fundamental limitation
-{\em within the general architecture of cost-based optimization} that will
-always render a good number of queries suboptimal. Prior research in other domains shows that
-increasing the complexity negatively influences
-performance~\cite{campbell88,moody98}.
+This paper provides a thorough investigation into \hbox{DBMS} {\em
+  suboptimality}: when the \hbox{DBMS} chooses a slower plan over a faster
+plan for a given query. We systematically examine the factors influencing
+the number of suboptimal queries. There could be multiple causes of the
+suboptimality. One possible cause could be some peculiarity within the tens
+of thousands of lines of code of that query optimizer. Another possible
+cause could be the query's complexity. Prior research in other domains shows
+that increasing complexity negatively influences
+performance~\cite{campbell88,moody98}. A third possible reason could be some
+fundamental limitation {\em within the general architecture of cost-based
+  optimization} that will always render a number of queries suboptimal.
 
@@
 \hbox{DBMSes}. Based on existing research and general knowledge of \hbox{DBMSes}, we
-propose an innovative predictive model to understand the presence and influence
-of suboptimality in query evaluation. We use an experimental methodology with
-the \hbox{DBMS} as a subject to test our hypotheses with respect to factors
-influencing suboptimality, applying to 
-our experiment data collected over about a \hbox{cumulative} 30~months of query
-executions (over 21,000 hours to run almost a million query
-executions).\todo{Young: Table I says something different: over 16,000 hours
-  ro run almost two million query executions which would correspond to two
-  years continuous execution}
+propose an innovative predictive model to better understand the presence and influence
+of suboptimality in query evaluation. We use an experimental methodology on
+a collection of \hbox{DBMSes} as subjects to test our hypotheses with respect to factors
+influencing suboptimality, utilizing
+experiment data collected over a \hbox{cumulative} 16,000 hours (over two
+years) of query executions.
 Our research falls within creative development of
@@
 \item We find that for a surprisingly large portion of queries, the plan
-  chosen  by the query optimizer is not the best plan, for some cardinality,
+  chosen  by the query optimizer is not the best plan, for some cardinality
   of the underlying tables.
 
-\item We propose and test a {\em predictive model} for \hbox{DBMSes} to better understand the factors causing suboptimality.
-
-\item \todo{Rick:Mention generational result}
+\item We propose a {\em predictive model} for \hbox{DBMSes} to better
+  understand the factors causing suboptimality.
+
+\item We test the seven quite specific hypotheses that arise from that
+  model across a wide range of DBMSes, queries, and data.
+
+\item We show that the hypothesized interactions are \todo{Rick:check after Sabah} supported,
+  lending credibility to our particular model.
+
+\item Through an innovative
+  analysis, we track the net cumulative benefit of a succession of operators
+  added to the DBMS, and show that current relational DBMSes may have
+  already reached the stage where it is difficult to provide performance
+  improvement via new operators, and in fact performance may degrade.
 
@@
   engineering directions.
+
+\item A subsequent analysis implies that a new approach, fundamentally
+  different from that utilized over the paper forty years, may be required
+  to get past a fundamental limitation uncovered in this research.
 \end{itemize}
@@
 understand how existing query optimizers can be further improved, and
-indeed, fundamental limitations inherent in these optimizers.
+indeed, whether  fundamental limitation inherent in these optimizers exist
+and whether such limitations are indeed already being encountered.
 
@@
 We study these aspects \c2j{}{across \hbox{DBMSes}} to identify the
-underlying causes. We have developed an initial predictive causal model that
+underlying causes. We have developed a predictive causal model that
 identifies four 
@@
 The following section introduces the methodology we will follow, that of
-empirical generalization. We then present a predictive, causal model of
-suboptimality and empirically test eight specific hypotheses derived from that
-model\c2j{. These are the first scientific results that we are aware of that
-  test a general causal model of \hbox{DBMSes}.}{ on three \hbox{DBMSes}.
-  These are the first scientific results that we are aware of that apply
-  {\em across} \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
-  specific algorithm.} This model has implications for research in
-engineering more efficient \hbox{DBMSes}.
+{\em empirical generalization}~\cite{cohenbook}. We then present in
+Sections~\ref{sec:model} and~\ref{sec:operationalization} a predictive,
+causal model of suboptimality and state seven specific hypotheses derived
+from that model. We then test these hypotheses across a number of queries,
+data, and cardinalities, whose results provide strong support for the
+validity of the proposed model.  These are the first scientific results that
+we are aware of that apply {\em across} \hbox{DBMSes}, rather than on a
+single, specific \hbox{DBMS} or on a specific algorithm. In
+Section~\ref{sec:diminishing} presents a follow-on analysis that
+demonstrates a fundamental limit to the number of operators that a
+cost-based optimizer can support, and indicates that perhaps extant DBMSes
+have hit that limit. This model has implications listed in
+Section~\ref{sec:engineering} for research in engineering more efficient
+\hbox{DBMSes}.
 
@@
 cardinalities. So Plan~P0 was chosen for 2M tuples, switching to Plan~P1 at
-1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through the
+1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through
+the plan
 sequence P0, P1, P0, P1, P2, P0, P2, and finally P3 at 40,000 tuples.
@@
 Experiment described in detail in Section~\ref{sec:experiments}, we started
-with 6,966 query instances (a query run on a specific \hbox{DBMS}) after our extensive query measurement
-protocol~\doubleblind{\todo{ref SIGMOD'13 paper!}}{}. While about 20\%,
-1,491, of these query instances contained only one
+with 6,967 query instances (a query run on a specific \hbox{DBMS}) after \doubleblind{our}{an} extensive query measurement
+protocol~\cite{TTPv1,TTPv2} applied its extensive sanity checks. While about
+20\%, or 1,491, of these query instances contained only one
 plan, a few of the other query instances switched plans at almost every change in cardinality (we varied the
 cardinality in increments of 10K tuples, a total of 200 cardinalities): see
-Figure~\ref{fig:planchanges}. Slightly over half, 4,093, exhibited
+Figure~\ref{fig:planchanges}. Slightly over half, or 3,933 query instances, exhibited
 suboptimality somewhere along those 200 cardinalities; a few had
@@
 \begin{figure}[h!]\centering
-\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
 \caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
@@
 \caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
-  number of changes to a {\em sub-optimal}
-  plan\label{fig:suboptplanchanges}}
+  number of changes to a {\em suboptimal} plan\label{fig:suboptplanchanges}}
 \vspace*{-2ex}
@@
 get the {\em best} plan, but rather to get a plan that is acceptably
-good. Figure~\ref{fig:suboptcumulative} shows
+good. (Thus, the very term ``query optimizer'' is aspirational rather than
+accurate.) Figure~\ref{fig:suboptcumulative} shows
 the cumulative distribution of the relative amount of suboptimality (where
@@
 suboptimality: less than 30\%. The challenge is that over fifth of all
-queries (1,355) exhibited a significant amount of suboptimality ($>= 30\%$); the
-relative slowdown can be quite large for some queries.
-
-We started with 7,640 query instances (a particular query running on a
+queries (1,355) exhibited a significant amount of suboptimality ($\geq
+30\%$). The relative slowdown can thus be quite large for some queries.
+
+We started with 7,640 query instances, each of which is a particular query running on a
 particular DBMS, cf.~Experiment 7 of Table~\ref{tab:run_stat}
 in Section~\ref{sec:experiments}. After our protocol, we were left with
-  6,966 query instances, of which 5,475 had at least one change point, so those are
+  6,967 query instances, of which 5,475 had at least one change point, so those are
   the ones we consider further. Of those, 1,382 had {\em no} suboptimality, so
@@
 
-We emphasize three important points. First, we used a sophisticated query
+We emphasize four important points.
+\begin{itemize}
+\item First, we used a sophisticated query
 measurement methodology that reduces the measurement variance, so that the
-query plans we identify as suboptimal definitely are so. Second, these
+query plans we identify as suboptimal definitely are so.
+\item Second, these
 results are over four \hbox{DBMSes}, and thus, such phenomena are not dependent on
@@
 cardinality \hbox{estimation} or plan costing or plan enumeration algorithm or
-implementation. Third, suboptimal plans are {\em not} the result of poor
+implementation.
+\item Third, suboptimal plans are {\em not} the result of poor
 coding or of inadequate algorithms. We view query optimization in modern
@@
 reflection of the complexity of the task of query optimization.
-
-%\pagebreak
-This is where the shift to a new methodology comes in. We want to understand
-cost-based optimization deeply, in this case to determine if a fundamental
-limitation exists, which means that we need to study multiple instances of
-that optimization architecture, to achieve generalizable results. To do so,
-we will articulate a predictive causal model for how
-and in what circumstances suboptimality arises and will provide compelling
-evidence via \hbox{hypothesis} testing that this model characterizes the
+\item Fourth, we wanted to understand whether a fundamental
+limitation exists in the prevalent approach to query optimization utilized
+by DBMSes generally, and certainly by the four DBMSes that we studies.
+\end{itemize}
+
+This is why we needed a new methodology. We want to understand
+cost-based optimization deeply. This means that we need to go beyond the
+examination of a single query optimizer, as is done in almost every paper
+over the forty-year history of query optimization research, to study multiple instances of
+that optimization architecture, in an effort to achieve generalizable
+results.
+
+Thus, in this paper,
+we articulate a predictive causal model for how
+and in what circumstances suboptimality arises and provide compelling
+evidence via \hbox{hypothesis} testing that this model accurately characterizes the
 \hbox{behavior} of query optimizers in general. This is what is meant by ``empirical
-generalization'' and why it is needed to answer such questions. In
+generalization''~\cite{cohenbook} and why it is needed to answer such questions. In
   the long history of research in database query optimization, or even of
@@
   \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
-  specific algorithm. This is due to the adoption of a new research
-  methodology, that of empirical generalization. Such a causal model can
+  specific algorithm. Such a DBMS-agnostic, though paradigmatic, causal model can
 \hbox{provide} guidance to the community about where fundamental research is needed
@@
 
+\todo{Rick: For Science article, add history: IMS->System R,
+  procedural->declarative, Selinger's query opt algorithm enabling rise of
+  relational model}
+\todo{Rick: For TODS, shorten history, probably remove SQL cites, just one
+  sentence about central role of Selinger}
 SQL has emerged as the {\em de facto} and {\em de jure} standard relational
@@
 and an even larger extension as SQL:1999~\cite{Melton03} and refinement as
-SQL:2008~\cite{SQL2008} and SQL:2011.
+SQL:2008~\cite{SQL2008} and SQL:2011~\cite{Melton11}.
 
 There has been extensive work in query optimization over the last 40
-years~\cite{Ioannidis96,Jarke84}. Query optimization and evaluation
+years~\cite{Ioannidis96,Jarke84}, during which a particular quite effective
+paradigm had taken hold, in both open-source and proprietary DBMSes. In this
+over-arching paradigm, query optimization and evaluation
 proceeds in several general steps~\cite{Ramakrishnan03}. First, the SQL query is 
@@
 resulting query evaluation plan produce the correct result for the
-query. A secondary but clearly very important objective is {\em efficiency}; after
-all, that is the raison d'etre for this phase. As is well known, the name
+query. This objective must be ensured both by the initial SQL-to-relational
+algebra translator and by the subsequent query enumerator. A secondary but clearly very important objective is {\em efficiency}; after
+all, that is the raison d'etre for this phase. As is well known and has been
+alluded to already, the name
 for this phase is an exaggeration, as existing optimizers do not produce
@@
   complex queries] without a lot of hand
-tuning''~\cite[page~59]{winslett02}. And as we will see, suboptimality is
-possible even when considering only simple queries.
+tuning''~\cite[page~59]{winslett02}. And as we will see, suboptimality may
+occur in sophisticated query optimizers even when considering only simple queries.
 
 While this paper does not provide direct solutions to address
-suboptimality, we envision that by adopting our proposed predictive model
+suboptimality, we envision that by following up on the implications of our proposed predictive model
 for suboptimality, engineering practice, such as dynamic reoptimization just
-mentioned may benefit. We elaborate on this subject in
+mentioned, may benefit. We elaborate on this subject in
 Sections~\ref{sec:root} and \ref{sec:engineering}, where we discuss the engineering implications
@@
 variables: ``Number of correlation names'', ``Presence of aggregate'',
-``Presence of primary key attribute'', and ``Presence of subquery". For each independent
+``Presence of primary key attribute'', and ``Presence of subquery''. For each independent
 variable, we have interventional control in our experiments, in that we can
@@
 In the middle of the figure is the construct of {\em plan space
-complexity}. Given a particular query, its measurable
+complexity}. Given a particular query, its
 complexity will impact the total number of candidate plans considered
-by the optimizer. However, this latter factor is not directly
+by the optimizer. This variable is not directly
 observable, again, especially for proprietary \hbox{systems}. However, we {\em
@@
 ``nonlinear''~\cite{Hulgen03}, though we focus on a more specific aspect of
-discontinuity.) Finally, the variable ``Number of repeats'' is the number
+discontinuity.) Finally, the variable ``Number of repeats'' {\em is}
+directly measurable as the number
 of times a plan is reused across the cardinality of the variable table. For
@@
 number of plans in the sequence, again, after removing sequential
-duplicates, nine plans in this particular case.
-
-Plan space complexity is an {\em intervening} variable, in that it is dependent
-on some of the variables on its left but is observable and thus  exerts influence on the dependent
+duplicates, nine plans in this particular case. Thus, we associate with the latent construct
+of plan space complexity three measurable (that is, indirect) variables. Our
+  model stipulates that the construct, and thus the three associated with
+  this construct, intervene between the constructs of optimizer complexity,
+  schema complexity and query complexity and the construct of suboptimality.
+
+Plan space complexity is an {\em intervening} construct, in that it is dependent
+on some of the constructs on its left but is observable and thus exerts influence on the dependent
 variable on its right. We can observe these variables within experiments and
@@
 complexity, between 
-CEPS and number of operators with discontinuity variables and between number of repeats and CEPS.
+CEPS and number of operators with discontinuity variables and between number of repeats.
 
@@
 difficult to optimize the query.  It is important to note that the
-occurrences of sub-optimality is not necessarily dependent on the presence
+occurrences of suboptimality is not necessarily dependent on the presence
 of a discontinuous plan operator. However, we predict that when
@@
 \vspace{0.6em}\noindent
-{\bf Hypothesis 4}: Plan space complexity (CEPS) and Number of operators
-with discontinuity will be positively correlated.
+{\bf Hypothesis 4}: Plan space complexity (CEPS) and (a)~Number of operators
+with discontinuity and (b) Number of repeats will be positively correlated.
 
@@
 presence of an aggregate, presence of primary key attribute, and
-presence of subquery), of the data (percentage of skewed data), of
+presence of subquery), of the data (percentage of skewed data), and of
 the plan space (number of repeats, CEPS, and number of discontinuous
@@
 This independent construct has one variable: percentage of skewed data.
-%\todo{Rick:write up operationalization}
 
@@
 
-We define the skew as the reciprocal of the number of distinct values. 
-For 2M distinct values, the skew would be 1/2M, which is practically 0. 
-For 1 distinct value, the skew would be 1.0. For 2 distinct values, 
-the skew would be 0.5. For 10 distinct values, the skew would be 0.1.
+We define the skew quite simply: ``the reciprocal of the number of distinct
+values,'' so $0 < \hbox{\em skew} \leq 1$.
+For 2M distinct values, the skew would be $\frac{1}{2M}= 0.000005$, which is practically 0. 
+For exactly one distinct value, the skew would be 1.0. For two distinct values, 
+the skew would be 0.5. For ten distinct values, the skew would be 0.1.
 
 We can generate the table of 2M rows by generating values sequentially from
-1 to the number of distinct values.  This creates a "span of values." We
+1 to the number of distinct values.  This creates a ``span of values.'' We
 repeat this for the second span if necessary, and on and on, until we have
@@
 
-We use five skewness values: 0 (approximately), 0.001, 0.1, 0.5, and 1.0,
+We use five skewness values: 0.000005 (which we term only slightly
+misleadingly as ``no skew''), 0.001, 0.1, 0.5, and 1.0,
 for this independent variable.
@@
 cardinality. To identify jumps in the first step, we examine the slopes
-between each pair of consecutive cardinalities (again, separated by 10K
+between each pair of adjacent cardinalities (again, separated by 10K
 tuples). We expect that the slope (that is, the first derivative) is
-well-behaved and thus does not change much for consecutive cardinalities. 
+well-behaved and thus does not change much for adjacent cardinalities. 
 A jump thus indicates a large deviation in the second derivative of the 
@@
 where {\em card} is the input
-cardinality) between each pair of consecutive cardinalities as
+cardinality) between each pair of adjacent cardinalities as
 $S_{\hbox{\small\em card}} = (C_{\hbox{\small\em card} + 10K} -
@@
 the \hbox{DBMS} to update the table's statistics to be accurate before actually
-evaluating the query. \todo{Young: we did this for all experiments, not just
-  monotonicity, right?}
-We then examined every case where the same plan for a query was run at different cardinalities; that is,
-we compared two query times measured by {\tt currentTimeMillis()} in {\sc
-  Java} per a pair of different cardinalities at which the same plan is
-observed during the study.   We expected that as the cardinality decreased,
-runtime would monotonically decrease. However, due to the variance in query
+evaluating the query (we did this for all experiments).
+We expected that as the cardinality decreased,
+the runtime would also monotonically decrease. However, due to the variance in query
 time measurement observed even when the cardinality was identical, we encountered
-spurior violations.
+spurious violations.
 
@@
 At each Q@C pair, Suboptimality is coded as four levels (0--3), based on the distance in 
-standard deviations, upto 3 standard deviations. 
+standard deviations, up to three standard deviations (we chose this cutoff
+because 99.7\% of all Q@C pairs are less than or equal to three standard
+deviations). 
 \shorten{\begin{itemize}
@@
 \noindent}
-We then sum this value over all pairs of cardinalities with different plans
-(the change points) for the query.
-
-\todo{Sabah: this paragraph talks about a suboptimality of up to 133, but the
-  above papra says there are just 4 levels. That is inconsistent. The reason is that the raw value ranges from 0 to 133. 
-After the log transformation, the values range from 0 and 2.13. Also, do
-  we still use a log transformation?}
+We then sum this value over all Q@C pairs with different plans
+(the change points) to arrive at a single integer for the query instance.
+
 Note that while \azdb\ examines the plan at every cardinality, it only has
@@
 points.  Since many fewer {\QatC}s were involved, we could try many more
-queries than the \hbox{Exhaustive} Experiment. In the Exploratory Experiment to be
+queries than the \hbox{Exhaustive} \hbox{Experiment}. In the Exploratory Experiment to be
 described in Section~\ref{sec:experiments}, the value of suboptimality ranged
@@
 
-No runs violated the experiment-wide sanity checks. For the largest
+No run violated the experiment-wide sanity checks. For the largest
 experiment, the Confirmatory experiment 7 discussed below, approximately
-10.5\% of the query exectuions and 5.1\%
-of the queryies-at-cardinality (Q@Cs) were dropped due to query execution
+10.5\% of the query executions and 5.1\%
+of the queries-at-cardinality (Q@Cs) were dropped due to query execution
 and Q@C sanity checks. As a result, excessive variation in calculated query
@@
   precision, and so we developed TTPv1~\cite{TTPv1} to provide that needed
-  precision. As our research in suboptimality prooceeded, we realized that
+  precision. As our research in suboptimality proceeded, we realized that
   to get a actually recognized suboptimality for two adjacent cardinalities
@@
 In the following, we describe in detail the data used by our experiments and
-the experimental scenarios we defined. More details on the data sets used
+the experimental scenarios we defined. More details on the data sets and
+experiment scenarios used
 can be found in Appendix~\ref{sec:app}.
@@
 \subsection{Data sets}\label{sec:datasets}
-We generate our experiment data set randomly in the experiments. However, we
+We generate our experiment data set randomly in each of the experiments. However, we
 use seeds to control the random data generator so that it can produce
-repeatable data as required. (Randomly generated data will probably be
-easier for the optimizer to deal with than skewed data.)
+repeatable data as required.
 
@@
   this table}{, physically, of such tables} as the experiments are being performed. 
-%\todo{Young: please check above}
 \c2j{We used a}{We organized three configurations for the data sets. The first configuration
@@
 of each table is roughly 16Mbyte. We also produced a version of the data set with primary
-keys (of the first attribute) for all tables. \todo{Young: is the following
-  true'' I don't think so...: Second, to
-perform queries with more relations, we create eight tables. In this case,
-each table has the same configuration as in the first data set. Third, to
-test queries on large tables, we configure each table with 64 integer-typed
-attributes. In this data set, we create four tables, but the size of each one is
-512MByte.}
+keys (of the first attribute) for all tables.
 
@@
 7 & Confirmatory 		& TTPv2 & 11,375 & 7,640 & 99,558 & 995,580 & 890,631\\
-\multicolumn{3}{c|}{Total}	& 16,880 & 10,200 & 236,960 & 1,861,600 & 1,650,666\\
+\multicolumn{3}{c|}{\em Total}	& 16,880 & 10,200 & 236,960 & 1,861,600 & 1,650,666\\
 \end{tabular}
@@
 ``Exhaustive''). Note that the number of retained QEs went up a lot (that is
-the main benefit of TTPv2), while the number of hourse required actually
+the main benefit of TTPv2), while the number of hours required actually
 went down (another benefit).
@@
 from each of the four \hbox{DBMSes} on the primary key data set, for exploration
-across all combinations, a total of 780 query instances. These query
+across all combinations, a total of 780 query instances. These query sets
+are enumerated in more detail in Appendix~\ref{sec:app}.
 We ran the \hbox{DBMSes} on the {\em
-  change points}: consecutive cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
+  change points}: adjacent cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
 plans. The protocol retained about 78\% of the QEs.
@@
 (a)~200 queries of four \hbox{DBMSes} without primary keys defined and
-(b)~100 queries from each of the four \hbox{DBMSes} on the primary key dataset,
+(b)~100 queries from each of the four \hbox{DBMSes} on the primary key data set,
 for exploration across all combinations, a total of 1,200 query instances (= 300 $\times$ 4 (DBMSes)).
 We termed this sixth experiment ``Exploratory.''  We ran the DBMSes on the {\em
-  change points}: consecutive cardinalities (10K apart; 300 for one DBMS) with different
+  change points}: adjacent cardinalities (10K apart; 300 for one DBMS) with different
 plans. This experiment required 1,663 hours. The experimental
@@
 (e)~100 queries on the data set with primary keys and secondary indexes defined, 
-(f)~100 queries with a subquery on the data set with primary keys defined,  
-(e)~100 queries with a subquery on the data set with primary keys and secondary indexes defined, and 
-(g)~100 queries on the data set with primary keys and secondary indexes defined, 
+(f)~100 queries with a subquery on the data set with primary keys defined, and 
+(e)~200 queries with a subquery on the data set with primary keys and secondary indexes defined,
 for a total of 1,910 queries and a total of 7,640 query instances (over the
-four DBMSes).
-Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed, roughly 13 per query.
-\todo{Young, where did this number come from?:}The protocol accurately timed 53,547
-{\QatC} pairs as change points.\
-
-In this Confirmatory Experiment, we observed 3,347 strict
-monotonicity violations (0.74\%), and 1,966 relaxed monotonic violations
-(0.43\%) (out of a total of 452,684 pairs), which
-provides further confidence that monotonicity also applies to
-operators found in queries over data in various contexts (as described above) 
-and that our operationalization of suboptimality is a valid one.
-
-In the remainder of this section, we focus using the measured independent,
-latent, and dependent variables in the Confirmatory experiment
-to test predictions that arise out of our causal model in Figure~\ref{fig:model}.
+four \hbox{DBMSes}).
+Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed,
+roughly 13 per query. The protocol accurately timed
+both sides of adjacent {\QatC}s (53,547 change points in all).
+
+In this Confirmatory Experiment, we observed 3,347 strict monotonicity
+violations (0.74\%), and 1,966 relaxed monotonicity violations (0.43\%) (out
+of a total of 452,684 pairs
+of two Q@Cs having identical plans and query instance), which provides
+further confidence that monotonicity also applies to operators found in
+queries over data in various contexts (as described above) and that our
+operationalization of suboptimality is a valid one.
+
+In the remainder of this section, we focus using the measured independent
+and dependent variables in the Confirmatory experiment
+to test the predictions that arise out of our causal model in Figure~\ref{fig:model}.
 
@@
 result of several years of programming effort and about 30 months of
-experimental runs. First, {\em every} \hbox{DBMS} exhibits suboptimality. Thus, this
-phenomenon is likely to be a fundamental aspect of either the algorithm
-(cost-based optimization) or the creator of the algorithm (human information
-processing). Our model includes both effects. Perhaps surprisingly, 
-%a little over half (1579 queries out of the 3056 query instances
-more than half (3,933 queries out of the 6,983\todo{Rick: I have 6967 in my set since 16 were dropped 
-because of missing one side of change point}  
+experimental runs, summarized in Table~\ref{tab:run_stat}.
+
+Let illustrate the descriptive statistics with the Confirmatory
+experiment. In that experiment, we started with 7,640 query instances. These
+resulted in 99,558 Q@Cs, with 10 each query executions (QEs). There were 890,631
+retained QEs, which resulted in 6,983 retained query
+instances (equivalently, 657 query instances were dropped by the
+protocol). Additionally, 16 query instances were dropped {\em after} the
+protocol because they were missing data for the only change point in the
+query. Therefore, we were left with 6,967 queries.
+
+First, perhaps surprisingly, 
+more than half (3,933 queries out of the 6,967
 query instances that emerged from the measurement protocol) exhibited suboptimality
 somewhere in the range of cardinality of the varying table. 
-Most instances of those queries (3,370) had the meximum value of suboptimality at a Q@C pair (i.e., level 3). 
+{\em Every} \hbox{DBMS} exhibits suboptimality.
+
+Secondly, most (3,370) of those suboptimal queries (86\%) had the maximum value of
+suboptimality  (i.e., level 3) at some Q@C pair.
 \shorten{a big number for a query could be many pairs, each at a small
@@
 querynum from Cnfm_Analysis_SuboptDefn where subopt_sd>=3);}
-CEPS ranged from 1 to 24,
-plans across the cardinality range. The number of discontinuous operators observed in plans for a query
-averaged 9.4, with 85 being the maximum. Number of repeats ranged from 0 to 108, with the mean being 4.99.
-
-\subsection{Correlational Analysis}
+
+This phenomenon (query suboptimality) is likely to be a fundamental aspect
+of either the algorithm (cost-based optimization) or the creator of the
+algorithm (human information processing). Our model includes both effects.
+
+Third, concerning the causal factors of suboptimality in our model,
+\begin{itemize}
+\item the cardinality of the effective plan space (CEPS) ranged from 1 to 24
+plans across the cardinality range,
+\item the number of discontinuous operators observed in plans for a query
+averaged 9.4, with 85 being the maximum, and
+\item the number of repeats ranged from 0 to 108, with the mean being 4.99.
+\end{itemize}
+
+\subsection{Correlational Analysis}\label{sec:corr}
 We tested Hypotheses 1--7 using the strength and significance of correlations
@@
 significant at the 0.05 level, the accepted standard for
-significance. ``---'' denotes no prediction arising from the model.) As can
+significance. ``---'' denotes no prediction arising from the model.)
+\todo{Sabah: update after filling out Table II:}As can
 be seen, most of the predictions (nine) arising from the causal model are
@@
 
-\subsection{Regression Analysis}
-\todo{Sabah: revisit this entire section, as we changed our model}
+\subsection{Regression Analysis}\label{sec:regression}
+\todo{Sabah: revisit this entire section, with confirmatory analysis numbers}
 We ran a regression over the independent variables of the
@@
 
-Hypothesis 5 predicts that the presence of secondary indexes will be a moderator,
-affecting the interactions between query complexity with plan space
+Hypothesis 6 predicts that the presence of secondary indexes will be a moderator,
+strengthen the interactions between query complexity with plan space
 complexity. It thus provides
-predictions that the strength of such interactions will decrease in the
+predictions that the strength of such interactions will increase in the
 presence of secondary keys. When
@@
 
-\todo{Sabah: revisit. Not sure about the purpose of this paragraph}
-We also examined the individual contributions to each of these
-three variables (e.g., suboptimality) from their causal variables (in the
-case of suboptimality, from the four variables of correlation names, presence
-of an aggregate, CEPS, and discontinuity). 
 Table~\ref{tab:hyptest8} shows that the strength of all interactions goes
-down as predicted, except for ones involving aggregates. Recall
+up as predicted, except for ones involving aggregates. Recall
 that the predictive model indicates weaker effects generally for aggregates,
@@
 
-\todo{Sabah: redo hypotheses in these tables, for example, H1 doesn't go to
-  subopt, then add the confirmatory numbers after we discuss them}
 \begin{table}[t]
-\tbl{Testing Hypotheses 1--7: Correlations on the Confiratory Study\label{tab:hyptest}}{%
+\tbl{Testing Hypotheses 1--7: Correlations on the Confirmatory Study\label{tab:hyptest}}{%
 %\begin{center}
@@
 Repeats& {\bf H7a:}  & --- & {\bf H5:}  & ---\\
-CEPS &  {\bf H7b:} 0. & --- &  --- &{\bf H4:}  \\
+CEPS &  {\bf H7b:} 0. & {\bf H4b:} &  --- &{\bf H4a:}  \\
 Discontinuity & {\bf H7c:} 0.31 & --- & --- & ---\\
@@
 encapsulates our theory for how cost-based query optimizers might select a
-suboptimal plan for a query at a cardinality. This model implies the eight
+suboptimal plan for a query at a cardinality. This model implies the seven
 specific hypotheses listed in Section~\ref{sec:model}. We then performed
-four experiments to refine our operationalizations (such as discontinuity,
-via the Exhaustive Experiment and number of operators available with the additional
-Exhaustive with Keys Experiment), to test fundamental assumptions (such as
-monotonicity, via the Monotonicity Experiment), and to test and make minor
-refinements to our model (via the Exploratory Experiment). While exploring
-the data, one must be cognizant of the possibility of Type 1 errors: false
-positives that lead one to believe a relationship exists when it doesn't.
-
-To control for Type 1 errors, we then performed the Confirmatory 
+six experiments,
+\begin{itemize}
+\item to refine our operationalizations: specifically discontinuity,
+via the Initial \hbox{Exhaustive} and Exhaustive \hbox{Experiments}, and number of operators available, via the
+Exhaustive with Keys \hbox{Experiment},
+\item to test fundamental assumptions, specifically
+monotonicity, via the Monotonicity \hbox{Experiment}, and
+\item to test and make minor
+refinements to our model: the Initial Exploratory and \hbox{Exploratory}
+Experiments.
+\end{itemize}
+During this exploration, \doubleblind{once we completed the TTPv2 protocol,}{when the TTPv2 protocol became available,} we reran
+\hbox{experiments} to avail ourselves of the increased precision due to a much
+higher percentage of retained QEs and thus Q@Cs and query instances.
+
+Throughout these six experiments, we were cognizant of the
+possibility of Type 1 errors: false positives that lead one to believe a
+relationship exists when it doesn't.
+To control for such errors, we then performed the final Confirmatory 
 Experiment on a completely different data set consisting of many more 
@@
 
+\todo{Sabah: update with final results:}
 Correlational analysis provides significant support for the model, except
@@
 Now that the causal model has been found to be supported by the confirmatory
-analysis, we turn to possible implications of this model.
+\hbox{analysis}, we turn to possible implications of this model.
 
@@
 as a {\em general} class of computational artifacts and to articulate and test
-a predictive model characterizing how such optimizers,
+a predictive model \hbox{characterizing} how such optimizers,
 again, as a general class, behave.  This model can be used to further
@@
 
+\todo{Sabah: update with final numbers:}
 Our model includes three causal factors of suboptimality: \hbox{DBMS} optimizer
@@
 the regression \hbox{coefficient} that was highest was for CEPS
-(cf.~Table~\ref{tab:hyptest8}: 0.51--0.54, normalized) The next \hbox{highest}
+(cf.~Table~\ref{tab:hyptest}: 0.51--0.54, normalized) The next \hbox{highest}
 regression factor was number of tables involved, which is highly correlated
@@
 
-\section{Diminishing Returns?}
-In order to increase query performance, \hbox{DBMSes} are extended over time with
-new \hbox{relational} query operators. Also over time, \hbox{DBMSes} are extended with new
-storage and indexing structures,
-which themselves elicit new query operators. Each subsequent generation of
-the \hbox{DBMS} thus supports an ever-expanding collection of operators,
-thus the current version is simply the most recent within a series of \hbox{DBMS}
-{\em generations}.
+\section{Diminishing Returns?}\label{sec:diminishing}
+In order to increase query performance, \hbox{DBMSes} are extended over time
+with new \hbox{relational} query operators. Also over time, \hbox{DBMSes}
+are extended with new storage and indexing structures, which themselves
+elicit new query operators. For example, nested loop join was probably the
+first implementation of this needed algebraic operator. B-tree index join
+was only possible when such indexes were added (quite early in the history
+of modern DBMSes). Adding hashing to a DBMS enables a hash join operator;
+adding hash indexes enables (perhaps several variants of) hash-index join. Each
+subsequent generation of the \hbox{DBMS} thus supports an ever-expanding
+collection of operators, with the current incarnation the most recent
+within a series of \hbox{DBMS} {\em generations}.
 
 \subsection{A Gedanken Experiment}\label{sec:gedanken}
-Consider a Gedanken experiment that analyzes the plan for each query at
-each cardinality, that is, for each Q@C, for each \hbox{DBMS} generation. In early
-generations, there will be fewer plan changes as the cardinality increases
-for a given query, just \hbox{because} there are fewer operators available. For
-later generations, some of the Q@Cs will be associated with different plans,
-enabled by the new operators added by subsequent generations. As our causal
-model in Figure~\ref{fig:model} predicts, the effective plan space for a
-query may grow in subsequent generations, due to the increased number of
-operators available.
-
-In many (hopefully most) cases, the new plan will be more efficient than
-staying with the previous plan selected in a previous generation. After all,
-that is the very reason the new operator(s) were added to that generation.
-However, in the presence of suboptimality,
-sometimes the new plan at that Q@C is {\em not} preferable,
-as that \hbox{DBMS} generation's query optimizer selected the wrong
-plan.  Indeed, the query optimizer also evolves and improves with each new
+Consider a Gedanken experiment that examines the plan for each query at
+each cardinality, that is, each Q@C, {\em for each \hbox{DBMS} generation}. In early
+generations, there will be few plan changes for a given query as the cardinality varies
+simply \hbox{because} there are few operators available. For
+subsequent generations, some of the Q@Cs will be associated with different plans,
+enabled by the new operators that were added.
+
+In many (hopefully most) cases, a new plan selected by a generation will be
+more efficient than the plan selected by the immediate previous
+generation. After all, that is the very reason the new operator(s) were
+added to that subsequent generation.  However, in the presence of
+suboptimality, sometimes the new plan at that Q@C is {\em not} preferable,
+as that \hbox{DBMS} generation's query optimizer selected the wrong plan.
+Indeed, the query optimizer also evolves and improves with each new
 generation, in part to minimize the chance of selecting the wrong plan.
 
-Our predictive model suggests that suboptimality is causally
-impacted by the independent variable of CEPS (cardinality of effective plan
-space\todo{Rick:Connect with number of operators available and with Section~\ref{sec:root}}. A
-possible implication is that as the generations of the \hbox{DBMS} add more and more
-operators, suboptimality will also increase. \todo{Rick:No: causality is with
-  CEPS, not number of operators!}
-
-\todo{Rick:Make point that our causal analysis is across \hbox{DBMSes} with different
-  numbers of operators. Here we're talking about {\em within} a single \hbox{DBMS},
-  across multiple generations, each with an additional operator, and thus a
-  higher CEPS. In fact, it would be useful to see if generation number was correlated with CEPS.}
+Our predictive model in Figure~\ref{fig:model} suggests that suboptimality is causally impacted by the
+independent variable of CEPS (cardinality of effective plan space), and CEPS
+is naturally connected to the number of operators available\todo{Rick:
+  connect with Section~\ref{sec:root} after Sabah updates that section; also
+  discuss the exponential increase in CEPS possible}. A possible implication
+is that as the evolution of generations of the \hbox{DBMS} add more and more
+operators, suboptimality will also increase.
+
+It is important to emphasize that our causal analysis to this point has been
+{\em across multiple \hbox{DBMSes}}, each with a different set of
+operators and even number of operators. Let's reprise from Section~\ref{sec:intro} on
+page~\pageref{sec:intro}: Our goal, up to this point, has been ``to \hbox{understand} cost-based query
+optimizers
+as a {\em general} class of computational artifacts and to come up with
+insights and ultimately with predictive theories about how such optimizers,
+again, as a general class, behave.''
+
+The present discussion has a quite different focus. Here we're talking about
+the query optimizer {\em within} a single \hbox{DBMS}, across multiple
+generations, each generation adding one or more operators.
+
+Let's then refine our Gedanken experiment. Let's assume each subsequent
+generation adds a single operator. Starting with a fixed set of queries,
+for each DBMS, we run each generation on each Q@C and then sum up the query
+times to compute a {\em per-generation (total) time} for that DBMS. We can
+also sum over all four DBMSes, to see how this  {\em overall per-generation
+  time} (a single number for each generation number) varies with generation.
 
 The underlying question then becomes, does an additional operator made
-available in a subsequent generation of the \hbox{DBMS} actually help or hurt?
+available in a subsequent generation of the \hbox{DBMS} actually help or
+hurt? More fully, is the predicted increase in suboptimality originating
+from that added operator, a causal effect discovered and validated by the process
+of {\em science}, specifically the application of \hbox{empirical} generalization illustrated
+in Figure~\ref{fig:empirical}, compensated for by the increased performance
+afforded by that operator, a benefit realized by the application of {\em
+  engineering}, the articulation, elaboration, and implementation of new
+storage structures and query evaluation algorithms?
+
+Fortunately, the engineering has already been done: current DBMSes have at
+their disposal multiple query operators, which have been perfected over the
+decades. Adopting the engineering perspective would predict that the overall per-generation
+time would {\em monotonically decrease as the generations add query evaluation
+operators}.
+
+Interestingly, adopting the scientific perspective reaches a different conclusion. Our
+causal model in Figure~\ref{fig:model} predicts that as the number of
+operators increases, the latent measures of plan space complexity increase,
+which therefore also {\em increase} suboptimality. As the generations of the
+\hbox{DBMS} contain successively greater numbers of operators, suboptimality
+will also increase with generation number.  Given that each new operator
+will improve a {\em shrinking}
+subset of Q@Cs for any given query, while the suboptimality occurs in a portion over an
+{\em expanding} subset of Q@Cs, our causal model implies that the curve of
+per-generation time will fall with each successive generation, but then
+start to level off as suboptimality becomes more prevalent (specifically,
+the prediction is that the second derivative is either. Eventually, the
+per-generation time will either asymptotically approach a horizontal line
+(with the first derivative approaching 0 from below), or worse: the first
+derivative changing to positive, with the per-generation time over the
+queries we are studying actually {\em increasing} with \hbox{DBMS}
+generation. In either case, the causal model predicts a point where {\em the
+\hbox{increase} in execution time due to suboptimality obviates the decrease
+enabled by the new operator}.
+
+Put in simpler terms, engineering considerations alone imply that DBMSes can
+continue to improve, whereas also adopting the scientific perspective predicts
+that eventually each DBMS will hit a wall, beyond which improvement is not possible.
+
+Does such a limit exist, and if so, how close are modern \hbox{DBMSes} to that limit?
 
@@
 While we do not have access to prior generations of our \hbox{DBMSes} (which is why
-the previous discussion was in the form of a Gedanken experiments), we can
+the previous discussion was in the form of a Gedanken experiment), we can
 approximate this with an experiment using data already collected on the
 current version available for each \hbox{DBMS}, to {\em simulate} the prior
-generations having a fewer number of operators, and thus a smaller set of
-realizable query plans. (The effect will certainly be smaller in our
-simulation, as we are nonetheless using the most recent {\em query
-  optimizer}n in all of the simulated generations.)
-
-Our data, drawn from the 8,840 query instances and their 112,118 Q@Cs in both the
-exploratory and confirmatory experiments of the previous study, 
-consists of pairs of adjacent plans
-(let's refer to them as the {\em lower} plan and the {\em upper} plan) for
-the same query at the {\em lower} and {\em upper} cardinalities) that are
-consecutive (that is, separated by the minimum cardinality, either 10,000 or
-300 rows), each with an actual execution time. All of Q@Cs having
-cardinalities between the upper plan of one pair and the lower plan
-of the next pair are associated with that same plan, guaranteed by the
-process in which we chose those Q@Cs for actually timing.
+generations, each successively having a fewer number of operators, and thus a smaller set of
+realizable query plans. (The effect will probably be smaller in our
+simulation, as we are nonetheless using the most recent query
+optimizer in all of the simulated generations; only the available operators
+will vary.)
+
+In the next section, we'll explain how we characterize the generations of
+each DBMS. For now assume that each DBMS is associated with a series of
+generations, with each generation having one more operator. So generation 1
+of the DBMS has just one operator, generation 2 has two operators, etc.
+
+Our data consists of the 8,840 query instances and their 112,118 Q@Cs in
+both the exploratory and confirmatory experiments (Experiments 6 and 7) of
+the previous study. Specifically, we focus on consists of pairs of adjacent
+Q@Cs (let's refer to them as the {\em lower} Q@C and the {\em upper} Q@C,
+and thus we also have a {\em lower plan} and an {\em upper plan}) for the
+same query at the {\em lower} and {\em upper} cardinalities, that are
+adjacent (that is, separated by the minimum cardinality, either 10,000 or
+300 rows), each with an actual execution time. All Q@Cs for that query and
+for that DBMS having cardinalities between the upper Q@C of one pair and the
+lower Q@C of the next pair are associated with that same plan. This will be
+the upper plan, guaranteed by the process in which we chose those Q@Cs for
+actually timing (recall from Section~\ref{sec:motivation} that we start from
+the highest cardinality, that of 2M tuples, looking for changes in the
+plan).
 
@@
 (a positive integer) that
-is the earliest generation containing all the operators in the
+is the earliest generation of the DBMS that contains all the operators in the
 plan. We also only consider Q@C pairs (with adjacent Q@Cs) where the lower
-plan has a generation distinct from the upper plan.  Say the upper
+plan has a generation distinct from that of the upper plan.  Say the upper
 generation is earlier than that of the lower plan, as illustrated in
-Figure~\ref{fig:negback}. In this figure, as we scan from left to right in
-increasing cardinality, we first encounter Plan A from generation 2 at the
-first cardinality of 10K rows, then Plan B sometime later, then even later,
-a pair of Q@Cs with Plan B at 900K rows and Plan A at the
-adjacent 910K rows, then much later Plan A again at 1320K rows (this is the
-measured Q@C that has Plan A that is closest to the measurement at 910K rows).
+Figure~\ref{fig:negback}. In this figure, as we scan from right to left in
+decreasing cardinality, we first encounter Plan A from generation 2 at the
+highest cardinality of 1320 rows (this is the
+measured Q@C that has Plan A that is closest to the measurement at 910K rows), then a pair of Q@Cs with Plan A at 910K rows and Plan B at the
+adjacent 900K rows, then later Plan A again at 10K rows.
 
@@
 
-In this particular case, Plan B is more appropriate (faster) than Plan A,
+In this particular case, as just mentioned, Plan B is more appropriate (faster) than Plan A,
 but that is not the only possibility. Sometimes the later generation with
@@
 The question then becomes, does a plan change by a subsequent generation
-(enabled by one or more additional operators) represent a win (runs faster)
+(\hbox{enabled} by the additional operator(s) in that generation) represent a win (runs faster)
 or a loss (runs slower, because of suboptimality)? More broadly, do the Q@Cs
@@
 generation, we'll adopt a more systematic ordering of the operators. We
-start by gathering all {\em single-operator} plans, all two-operator plans,
+start by gathering all {\em single-operator} plans, all {\em two-operator} plans,
 and so forth, and order the generation by the prevalence of their
@@
 
-We then examine the plans containing exactly two operators (in the case of
-MySQL, one of which is
-full table scan, as a two-operator plan not having that operator would
-necessarily be constructed by generation 3 or higher). Using MySQL again, there are two sets of
-such plans, those with the full table scan and full table scan with join
-operators (888 Q@Cs) and
-those with the full table scan and ref operators (112 Q@Cs). We pick full
-table scan with join as the operator added by Generation 2, given its
-prevalence of Q@Cs (a total of 1705 in the full set). Each
+We then examine the plans containing exactly two operators.
+Using MySQL again, there are two such : one with the Full Table Scan and
+Full Table Scan with Join operators (888 Q@Cs) and
+one with the Full Table Scan and Ref operators (112 Q@Cs). Full
+Table Scan with Join is thus the operator added by Generation 2, given its
+prevalence of Q@Cs. Each
 subsequent generation adds that operator that maximizes the number of plans
-that operator will eventual enable. So Generation 3 adds the eq\_ref operator,
-as that operator enables 633 plans eventually. Generation 4 adds the ref
-operator and Generation 5 adds the index operator. There are six plans that
+that operator will eventual enable. So Generation 3 adds the Eq\_Ref operator,
+as that operator enables 633 plans eventually. Generation 4 adds the Ref
+operator and Generation 5 adds the Index operator. 
 involve all five operators.
@@
 operators, covering 1086 Q@Cs, six combinations of three operators (1126 Q@Cs),
-two combinations of four operators (104 Q@Cs), and (naturally) exactly one
-combination of all five operators (6 Q@Cs). Thus for MySQL our data implies
+two combinations of four operators (104 Q@Cs), and exactly one
+combination of all five operators (6 Q@Cs).\shorten{ Thus for MySQL our data implies
 a simulation of five distinct generations, each adding an operator.
@@
   eq\_ref}, \hbox{\em ref}, \hbox{\em index} \}
-\end{align*}
+\end{align*}}
 For the four \hbox{DBMSes} in our study, the number of generations ranged from five
@@
 
-\subsection{Using Change Points}
-We now consider how to use data already collected, that is the {\em change
-  points} discussed in Section~\ref{sec:experiments}: consectutive
-cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
-plans. Specifically, how can change points be used to evaluate the
-effectiveness of different generations of a \hbox{DBMS}?
-
-Define $\hbox{\em ops}(p)$ for a given plan $p$ to be the set of operators
-present in that plan, with some operators perhaps repeated in that plan. 
-For example, plan $p_1$ may be composed of a {\em select} of an {\em agg} of a {\em select} over a
-relation. Hence,  $\hbox{\em ops}({p_1}) = \{\hbox{\em select}, \hbox{\em
-    agg}\}$. A generation $g$ is {\em applicable} to a plan $p$ if
-  $\hbox{\em ops}(p) \subseteq g$,  and so it is the case that $g_4$  is applicable to $p_1$. By
-  definition, if a generation is applicable to a given plan, it is
-  applicable to all subsequent generations. The smallest such generation is
-  termed the {\em minimally applicable generation}, or {\em mingen}, so
-  $\hbox{\em mingen}(p_1) = g_3$.
-
-\shorten{We couldn't make this one work...
-\todo{Young:replace in the above para with an actual SQL example query that was associated
-  with two MySQL generations in a Q@C pair. use the generation numbers from
-  the next para}
-
-\todo{Rick: I found one example but the cardinality information 
-was different than what's stated in the current prose...}
-
-%%runid = 1097, querynum = 46, max_card = 60000, upper_card = 29100, lower_ard = 28800, upper_gen = 4, lower_gen = 5
-%%max_cqt: 230, upper_cqt: 110, lower_cqt: 120, lower_extrapolated_cqt: 110
-%% experimentname: op-pk-30K-100q-idx, pk: true, sec_idx: true
-%% plan 0: ALL:Full Table Scan,ref
-%% plan 1: eq_ref,index,ref
-%% plan 0 at upper_card, plan 1 at lower_card
-\noindent
-\hspace{3ex}{\small\begin{verbatim}
-        SELECT t3.id2, t3.id4, t0.id3, t2.id1 
-        FROM ft_HT3 t2, ft_HT1 t1, ft_HT2 t3, ft_HT1 t0  
-        WHERE  (t2.id2=t1.id2 AND t1.id2=t3.id1 AND t3.id1=t0.id4)
-\end{verbatim}}
-}
-
-First consider pairs for which $\hbox{\em
-mingen}(\hbox{\em lower}) > \hbox{\em mingen}(\hbox{\em
-  upper})$. An example is shown in Figure~\ref{fig:posback}. In this case,
-the set of operators in Plan A for the Q@C at cardinality 900K requires at
-least generation 3, whereas the set of operators for Plan B for the Q@C at
-cardinality 910K requires but generation 2. (Note that the generations here
-are consecutive, that that is not required. Sometimes the generations of
-pairs of Q@Cs are wildly different.)
-
-\begin{figure*}[t]
-%\centering
-%originally 30pc
-\includegraphics[width=30pc]{figures/posback.pdf}
-\caption{An Example of a Positive Backward Delta\label{fig:posback}}
-\end{figure*}
-
-To examine the wisdom of picking Plan A for 900K (whose time was measured as
-the {\em higher} point shown (the one above the dashed line), we find the closest Q@C
-also having Plan A. There are two illustrated here, at the minimum and
-maximum cardinalities, of 10K and 2000K, respectively, with the closer one
-at 10K. (If there is no such change
-point, we can't do the extrapolation).
-
-We use the slope of the dashed line between the measured query time of Plan
-B at cardinalities 10K and 910K to get an estimated query time at 900k.
-This realizes an estimate of how fast the query using Plan B would have run
-on the database having the variable table with a cardinality of 900K. In
-this particular case, Plan B looks like it would have run {\em faster} (in
-fact, was was faster even at 910K in this particular example).
-
-As illustrated in Figure~\ref{fig:query769}, plans can be discontinuous,
-because their asymptotic complexity is $O(n \lg n)$ (though this often turns into
-a sequence of linear relationships, as illustrated by this
-figure). This behavior will, for plan B present at
-the two granularities of 10K and 910K, slightly overestimate the runtime of
-that plan at 900K, and thus {\em underestimate} the penalty of going with
-Plan A rather than Plan B. That said,
-given that we are extrapolating from the actual run time at a very
-close cardinality, our extrapolated estimate should be close.
-
-We term this a {\em backward extrapolation}, because we are extrapolating
-backward from a cardinality of 910K to 900K.
-
-With this estimate, we can compute the {\em relative delta}, defined as the
-measured time of Plan A (the chosen one) at the lower cardinality minus the
-extrapolated time of Plan B at that same cardinality, divided by the
-measured time of Plan B. The relative delta is thus scaled by original
-measured time at that granularity, and thus has a value between -1 and 1.
-
-In this case, the relative delta is positive (the measured time is greater
-than the extrapolated time, which indicates that the optimizer chose the
-{\em wrong plan}: here, Plan B should have been chosen.  A negative
-relative delta implies that the additional operator(s) available to the
-minimally applicable generation of the upper plan were indeed beneficial,
-that the measured time was lower. (In such cases, we again divide by the
-larger value.)  A
-positive relative delta implies the presence of suboptimality: the wrong
-plan was chosen by the \hbox{DBMS} generation, perhaps due to the greater number of
-operators available. (Note that this is a slightly more expansive definition
-of suboptimality than that used earlier and illustrated in
-Figure~\ref{fig:query769}.)
-
-There are thus seven orthogonal possibilities for each change point (that
-is, a pair of adjacent Q@Cs, a total of 66,769 pairs/change points).
-
-\begin{enumerate}
-\item The pair of Q@Cs share the same generation, that is, 
-$\hbox{\em mingen}(\hbox{\em lower}) \neq
-\hbox{\em mingen}(\hbox{\em upper})$ (62,903 Q@C pairs, 94.2\%), which we don't consider further.
-
-\item There is no other cardinality having the same plan as that with the
-later generation, in which case we also ignore this Q@C (768 Q@C pairs,
-1.2\%).
-
-\item The extrapolation yielded a computed query
-time that was negative, which we also drop (10 pairs, 0.01\%).
-
-\item The extrapolation was from above and indicates
-no suboptimality, termed a {\em negative backward relative delta}, as
-  exemplified in Figure~\ref{fig:negback}, examined earlier (492 pairs,
-  0.7\%).
-
-\item The extrapolation indicates no suboptimality, termed a {\em positive backward
-    relative delta}, exemplified in Figure~\ref{fig:posback} (583 pairs,
-  0.9\%).
-
-\item The extrapolation was from below (consider Figure~\ref{fig:negback} but with
-  Plan A from generation 5; we would then need to extrapolate from the
-  closest Plan B, which is {\em at a smaller cardinality}, in a {\em forward
-    direction}, to compute a {\em negative forward relative delta} (1218
-  pairs, 1.8\%).
-
-\item A extrapolation was a {\em positive forward relative delta} (795 pairs, 1.2\%).
-\end{enumerate}
-%1 of rel del: 2, -1 of rel del: 21 (neg backward:3, pos backward:18)
-From these seven possibilities, we thus retain (4) and (6), which indicate a
-suboptimal plan at the higher generation (1710 pairs), and (5) and (7), which indicate a
-non-suboptimal plan at the higher generation 1378 pairs).
-
-This analysis is for a single pair of adjacent Q@Cs for a single query
-running on a specific \hbox{DBMS},
-providing a relative delta for the minimally applicable generation. This is a percentage difference, and so is not \hbox{affected}
-by the absolute magnitude of the run time nor of the cardinality in
-question. Indeed, because it is a percentage difference, the relative delta
-is not affected by the query nor even which \hbox{DBMS} is involved. We associate
-this relative delta with the later generation, for it is that generation
-which had the choice between the two plans for that query.
-
-\subsection{Trends Across \hbox{DBMS} Generations}
-Consider what might the {\em average relative delta} behave across the
-generations, that is, how
-the average of the relative deltas associated with each generation will
-change with successive generations.
-The average relative delta is a characterization of the aggregate
-impact of that generation, providing a quantitative estimate of the benefit
-of adding that operator. (We use average so that each point is not impacted
-by the number of pairs over which that point is computed.)
-
-What does our causal model in Figure~\ref{fig:model} say about this? That
-causal model asserts that as the number of operators increases, the latent
-measures in plan space complexity increase, which therefore impacts
-suboptimality, also in a positive direction.
-
-Our Gedanken experiment in
-Section~\ref{sec:gedanken} takes this predicted behavior and predicts taht
-as the generations of the \hbox{DBMS} contain successively greater numbers of
-operators, suboptimality will also increase.
-
-Drilling down into this interaction, the next question is:   By
-how much is the increase in efficiency enabled by a new operator (for those
-Q@Cs utilizing a plan containing that operator) greater than the decrease in
-efficiency resulting from suboptimality (for a subset of those Q@Cs)
-resulting from adding that operator?  Given that each new operator will
-improve a shrinking subset of Q@Cs, while the suboptimality occurs in a
-portion over an expanding subset of Q@Cs, it seems that there must be a
-point where the decrease due to suboptimality obviates the increase enabled
-by the new operator. How close are modern \hbox{DBMSes} to that limit?
-
-This analysis implies that each additional operator will result in a net {\em
-  decrease} in average relative delta, with the possible result that the advantage of that
-additional operator exceeded the net impact of suboptimality. We thus
-predict that the first derivative of a relationship between average relative delta over
-increasing generation will be negative, a natural result of the efforts of
-\hbox{DBMS} developers to increase performance over successive generations of their
-\hbox{DBMS}.
-
-That said, each new operator is applicable to a successively smaller portion
-of the queries, and perhaps over a successively smaller portion of the
-cardinality space.  (We note in passing that the space of queries over which
-our particular experiment ranged is a rather small, though oft-used, portion
-of the exceeding complex SQL definition: over 3000 pages~\cite{Melton11}.)
-As already noted, our causal model predicts that the prevalence of suboptimality
-would increase as the number of operators available increased. We thus predict that
-the {\em second derivative} of the average relative delta for increasing
-generation will be {\em positive}, that is, that the curve will start to
-level off. If that indeed occurs, it raises the
-possibility of the curve either asymptotically approaching a horizontal line
-(the first derivative approaching 0), or worse: the first derivative
-changing to positive, with the average time over the queries we are studying
-actually {\em increasing} with \hbox{DBMS} generation.
-
-We partition the relevant four sets of change points into two groups: (i)~those
-with a {\em positive} relative delta (either forward or backward), denoting a
-{\em suboptimal decision} by the query optimizer at the later generation
-(correcspeonding to a higher generation number)
-and (ii)~those with a {\em negative} relative delta (either forward or
-backward) relative delta, indicating a {\em non-suboptimal decision} by the
-query optimizer at the later generation. (We can't state unequivicably that
-the plan is optimal, because there may be yet another plan involving the
-operators within that generation that is even faster.)
-
-Let's first examine those change points for which the query optimizer made a
-good decision, the non-suboptimal change points; see
-Table~\ref{tab:non-subopt}.
-
-\begin{table}[t]
-\tbl{Non-Suboptimal Change Points\label{tab:non-subopt}}{%
-%\begin{center}
-\begin{tabular}{c|c|c|c}
-{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}\\
-& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}\\
-\hline
-1 & --- & --- & 0.00\\
-2 & --- & --- & 0.00\\
-3 & --- & --- & 0.00\\
-4 & 24 & 0.28 & 0.28\\
-5 & 282 &0.118 & 0.130\\
-6 & 309 &0.274 & 0.203\\
-7 & 336 & 0.317 & 0.243\\
-8 & 43 & 0.20 & 0.241\\
-9 & 4 & 0.1 & 0.241\\
-10 & 17 &0.29 & 0.241\\
-11 & 357 & 0.178 & 0.225\\
-12 & 6 & 0.2 & 0.225\\
-13 & 0 & --- & 0.225\\
-{\em Cumulative}  &1378 & ---& 0.225\\
-\end{tabular}
-%\end{center}
-}
-\end{table}
-
-The first thing to note is that this data aggregates the results over the
-four \hbox{DBMSes}, which had generations ranging from five to 13. Thus this is the
-first study we are aware of that compares the generations trends of \hbox{DBMSes}
-over quite disparate code bases, thus getting at fundamental trends.
-
-Overall, only a small number of change points, 1378, or about 2\% of the
-total, satisfied the
-requirements listed above, including having a different generation number for
-the lower and upper plans. In fact, there were no such change points for the
-first three generations nor the last generation.
-
-While the numbers jump around a good bit, the thing to focus on is the {\em
-  trend} as the generations evolve.  The basic trend is a slightly
-decreasing relative delta, with the cumulative starting off at a high of
--0.28 (recall that high negative is good) and slowly decreasing to -0.23,
-resulting from larger negative numbers trending to lower negative
-numbers. The least squares slope is -0.024, which is consistent with the
-transition from a generation being net benefitial to actually being slightly
-suboptimal in generalat around generation 10.
-
-This general behavior matches our prediction arising from the structural causal model that it
-gets harder to squeeze out performance gains as operators are added to the
-\hbox{DBMS} over successive generations.
-
-We now examine those change points for which the query optimizer made a poor
-decision, in that we can surmize that there was a better plan (the one right
-next to it in the adjacent Q@C pair). Table~\ref{tab:subopt} provides the
-same information across the \hbox{DBMS} generations for the suboptimal change
-points: those for which the
-  relative deltas are positive, indicating a poor decision. While there are
-  still only a small number of change points, there are a greater number of
-  {\em suboptimal} change points, with more relatively showing up at more
-  recent generations. But more strikingly, the cumulative relative delta has
-  the opposite behavior to the non-suboptimal change points: it {\em
-    increases} over the generation, starting at 0.08 and ending at 0.25, a
-  value {\em higher} than that of the non-suboptimal change points.
-
-\begin{table}
-[t]
-\tbl{Suboptimal Change Points\label{tab:subopt}}{%
-%\begin{center}
-\begin{tabular}{c|c|c|c}
-{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}\\
-& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}\\
-\hline
-1 & --- & --- & 0.00\\
-2 & --- & --- & 0.00\\
-3 & --- & --- & 0.00\\
-4 & 2 & -0.2 & -0.2\\
-5 & 211 & -0.076 & -0.077\\
-6 & 86 & -0.16 & -0.102\\
-7 & 263 & -0.303 & -0.196\\
-8 & 273 & -0.303 & -0.231\\
-9 & 3 & -0.04 & -0.230\\
-10 & 168 & -0.319 & -0.245\\
-11 & 639 & -0.237 & -0.242\\
-12 & 53 & -0.35 & -0.245\\
-13 & 12 & -0.40 & -0.246\\
-{\em Cumulative} & 1710 & --- & -0.246\\
-\end{tabular}
-%\end{center}
-}
-\end{table}
-
-This is also as predicted: the optimizer is struggling with both more
-options (plans over the available operators) to select from and a diminished
-opportunity to make a significant improvement.
-
-Looking at this struggle from a different perspective, consider the {\em
-  maxium number of change points per query}, or {\em CPQ}, from the perspective of
-\hbox{DBMS} generations. For each query, count the number of Q@Cs at each
-generation (so for instance query Q1 could have 5 Q@Cs at generation 1, 7 Q@Cs at
-generation 2, 17 at generation 5, etc.). We then then take the maximum CPQ
+\subsection{Number of Change Points Per Generation}\label{sec:CPQ}
+As an illustrative example of how we can look at DBMS query optimizer
+performance through the lens of change points, let's consider  the {\em
+  maximum number of change points per query}, or {\em maximum CPQ}, from the
+perspective of
+\hbox{DBMS} generations.
+
+For each query, we count the number of Q@Cs at each
+generation, so for instance a query might have five Q@Cs at generation 1, seven Q@Cs at
+generation 2, and 17 Q@Cs at generation 5. We then then take the maximum CPQ
 over the queries (so perhaps 17 is the maximum for generation 5 over all
@@
 
-\begin{table}
-[t]
+\begin{table}[h]
 \tbl{Max Change Points Per Query, Per Generation\label{tab:maxcpq}}{%
@@
 
-What we notices is that the maximum CPQ {\em does} increase with generation,
+What we notice is that the maximum CPQ {\em does} increase with generation,
 up through generation 8, after which it falls off dramatically. In
@@
 there is a query that has an astonishingly high number of change points,
-172, all of that generation.
-
-There's another limitation we're hitting, one that is much more fundamental.
+172, all within that generation.
+
+This analysis indicates that there is something concerning that seems to get
+critical around generation 8.  And indeed, for this data set, we'll see in
+Section~\ref{sec:hitthewall} that the efficacy of the query optimizer
+bottoms out around that generation.
+
+\subsection{Using Change Points}\label{sec:usingCP}
+We now consider how to use data already collected, that is the {\em change
+  points} discussed in Section~\ref{sec:experiments}: adjacent
+cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
+plans. Specifically, how can change points be used to evaluate the
+effectiveness of different generations of a \hbox{DBMS}?
+
+Define $\hbox{\em ops}(p)$ for a given plan $p$ to be the set of operators
+present in that plan, with some operators perhaps repeated in that plan. 
+A generation $g$ (also a set of operators) is {\em applicable} to a plan $p$ if
+$\hbox{\em ops}(p) \subseteq g$. By
+definition, if a generation is applicable to a given plan, it is
+applicable to all subsequent generations. The smallest such generation is
+termed the {\em minimally applicable generation}, or {\em mingen}.
+\shorten{
+\shorten{We couldn't make this one work...
+\todo{Young:replace in the above para with an actual SQL example query that was associated
+  with two MySQL generations in a Q@C pair. use the generation numbers from
+  the next para}
+
+\todo{Rick: I found one example but the cardinality information 
+was different than what's stated in the current prose...}
+
+%%runid = 1097, querynum = 46, max_card = 60000, upper_card = 29100, lower_ard = 28800, upper_gen = 4, lower_gen = 5
+%%max_cqt: 230, upper_cqt: 110, lower_cqt: 120, lower_extrapolated_cqt: 110
+%% experimentname: op-pk-30K-100q-idx, pk: true, sec_idx: true
+%% plan 0: ALL:Full Table Scan,ref
+%% plan 1: eq_ref,index,ref
+%% plan 0 at upper_card, plan 1 at lower_card
+\noindent
+\hspace{3ex}{\small\begin{verbatim}
+        SELECT t3.id2, t3.id4, t0.id3, t2.id1 
+        FROM ft_HT3 t2, ft_HT1 t1, ft_HT2 t3, ft_HT1 t0  
+        WHERE  (t2.id2=t1.id2 AND t1.id2=t3.id1 AND t3.id1=t0.id4)
+\end{verbatim}}
+}
+}
+As change point consists of a pair of adjacent Q@Cs, we have two generations
+to consider. Lets start with pairs for which $\hbox{\em
+mingen}(\hbox{\em lower}) > \hbox{\em mingen}(\hbox{\em
+  upper})$. An example is shown in Figure~\ref{fig:posback}. In this case,
+the set of operators in Plan A for the Q@C at cardinality 900K requires at
+least generation 3, whereas the set of operators for Plan B for the Q@C at
+cardinality 910K requires but generation 2. (Note that the generations in
+this example
+are consecutive, but that is not required. Sometimes the generations within a
+pair are wildly different.)
+
+\begin{figure*}[t]
+%\centering
+%originally 30pc
+\includegraphics[width=30pc]{figures/posback.pdf}
+\caption{An Example of a Positive Backward Delta\label{fig:posback}}
+\end{figure*}
+
+To examine the wisdom of picking Plan A for 900K (whose time was measured as
+the {\em higher} point shown (the one above the dashed line), we find the closest Q@C
+also having Plan A. There are two illustrated here, at cardinalities 30K and
+2000K, respectively, with the closer one at 30K. (If there is no such Q@C,
+we can't do the extrapolation, and simply remove that change point from
+further consideration.)
+
+We use the slope of the dashed line between the measured query time of Plan
+B at cardinalities 30K and 910K to get an estimated query time at 900k.
+This realizes an estimate of how fast the query using Plan B would have run
+on the database having the variable table with a cardinality of 900K. In
+this particular case, Plan B looks like it would have run {\em faster} (in
+fact, was was faster even at 910K in this particular example).
+
+As illustrated in Figure~\ref{fig:discontinuity}, plans can be
+discontinuous, an example being external merge sort. For that operator, as
+the cardinality increases, the number of passes  will at some point by
+incremented, but the time is linear for a set number of passes.
+Such behavior will, for plan B present at
+the two granularities of 30K and 910K, slightly overestimate the runtime of
+that plan at 900K, and thus {\em underestimate} the penalty of going with
+Plan A rather than Plan B. That said,
+given that we are extrapolating from the actual run time at a very
+close cardinality, our extrapolated estimate should be close.
+
+We term this situation a {\em backward extrapolation}, because we are extrapolating
+backward from a cardinality of 910K to one of 900K.
+
+With this extrapolation, we can compute the {\em relative delta}, defined as the
+measured time of Plan A (the chosen one) at the lower cardinality (900K) minus the
+extrapolated time of Plan B at that same cardinality, divided by the
+measured time of Plan A. The relative delta is thus scaled by original
+measured time at that granularity, and thus has a maximum value of 1.
+
+In this case, the relative delta is positive (the measured time for the plan
+associated with the higher generation is greater than the extrapolated
+time), which indicates that the optimizer chose the {\em wrong plan}: here,
+Plan B should have been chosen.
+
+On the other hand, a {\em negative} relative delta (where the measured
+time for the plan associated with the higher generation is lower than the
+extrapolated time, such as that illustrated in Figure~\ref{fig:negback})
+implies that the additional operator(s) available to the minimally
+applicable generation of the upper plan were indeed beneficial, in that that
+plan was faster. (In such cases, we again divide by the larger
+value, so the minimum is~-1.)
+
+Summarizing, a
+positive relative delta (extrapolated in either the forward or backward direction) implies the presence of suboptimality: the wrong
+plan was chosen by the \hbox{DBMS} generation, perhaps due to the greater number of
+operators available. A negative relative delta (forward or backward) implies the {\em absence} of
+suboptimality: the chosen plan was faster at that cardinality than the one
+chosen at the adjacent cardinality. (Note that this is a slightly more expansive definition
+of suboptimality than that used earlier and illustrated in
+Figure~\ref{fig:query769}.)
+
+There are six orthogonal possibilities for each change point (that
+is, a pair of adjacent Q@Cs), a total of 66,769 pairs/change
+points from the Exploratory and Confirmatory Experiments. (There is also the
+case of a query instance containing a lone Q@C, which we don't consider further.)
+
+\begin{enumerate}
+\item The pair of Q@Cs share the same generation: $\hbox{\em mingen}(\hbox{\em lower}) =
+\hbox{\em mingen}(\hbox{\em upper})$  (62,903 Q@C pairs, 94.2\%), which we don't consider further.
+
+\item The extrapolation yielded a computed query
+time that was negative, which we also drop (10 pairs, 0.01\%).
+
+\item The extrapolation from above and indicating
+no suboptimality, termed a {\em negative backward relative
+  delta}, as
+  exemplified in Figure~\ref{fig:negback}, examined earlier (492 pairs,
+  0.7\%).
+
+\item The extrapolation, termed a {\em positive backward
+    relative delta} and exemplified in Figure~\ref{fig:posback}, indicates
+  a suboptimal plan (583 pairs, 0.9\%).
+
+\item The extrapolation was a {\em negative forward relative delta},
+  indicating no suboptimality (1218 pairs, 1.8\%).
+
+\item A extrapolation was from below (consider Figure~\ref{fig:negback} but with
+  Plan A from generation 5; we would then need to extrapolate from the
+  closest Plan B, which is {\em at a smaller cardinality}), in a {\em forward
+    direction}, to compute a {\em positive forward relative delta},
+  indicating a suboptimal plan (795 pairs, 1.2\%).
+\end{enumerate}
+%1 of rel del: 2, -1 of rel del: 21 (neg backward:3, pos backward:18)
+From these six possibilities, we thus retain (3) and (5), which indicate a
+\hbox{suboptimal} plan at the higher generation (1710 pairs), and (4) and (6), which indicate a
+non-suboptimal plan at the higher generation (1378 pairs).
+
+\subsection{Trends Across \hbox{DBMS} Generations}\label{sec:trends}
+The analysis in the previous section is for {\em a single pair of adjacent Q@Cs for a single query
+running on a specific \hbox{DBMS}},
+providing a {\em relative delta} for the minimally \hbox{applicable} generation. A
+positive relative delta indicates that the later generation chose a
+suboptimal plan; a negative relative delta indicates the later generation
+did not. The relative delta is a percentage difference, and so is not \hbox{affected}
+by the absolute magnitude of the run time nor by the cardinality in
+question. Indeed, because it is a percentage difference, the relative delta
+is not affected by the query nor even which \hbox{DBMS} is involved. We associate
+this relative delta with the later generation, for it is that generation
+which had the choice between the two plans for that query.
+
+Consider how the {\em average} relative delta might behave across successive
+generations.
+The average relative delta is a characterization of the aggregate
+impact of that generation, providing a quantitative estimate of the benefit
+of adding that operator. (We use average so that each point is not impacted
+by the number of pairs over which that point is computed.)  To emphasize,
+the average aggregate delta aggregates over change points, queries, and even
+DBMSes, providing a single average relative delta for each generation.
+
+What does our causal model in Figure~\ref{fig:model}, supported by the
+correlational and regression analyses of the Confirmatory Experiment in
+Sections~\ref{sec:corr} and \ref{sec:regression}, say about this? That
+causal model asserts that as the number of operators increases in
+subsequent, the latent measures in plan space complexity increase, which
+therefore impacts suboptimality, also in a positive direction.
+
+Our Gedanken experiment in
+Section~\ref{sec:gedanken} takes this behavior and predicts that
+as the generations contain successively greater numbers of
+operators, suboptimality will also increase.
+
+If we plot the performance of the DBMS on the {\em y}-axis, say as the
+total time for a workload consisting of a set of queries over a
+prescribed data set, for a sequence of {\em DBMS generations} listed on the
+{\em x}-axis, the average relative delta is in some way a characterization
+of the {\em slope} of this relationship. A negative average relative delta
+implies that the indicated generation is doing a good job, with less
+suboptimality, and so the total workload execution time will go down; a
+positive average relative delta implies that the suboptimal decisions are
+dominating, indicated by the total workload execution time going up for that
+generation.
+
+We expect that the average relative delta for the first few generations will
+be negative, reflecting new operators that improves some plans, a natural
+result of the \hbox{efforts} of \hbox{DBMS} developers to increase performance over
+successive generations of their \hbox{DBMS}.  With a negative slope, the
+performance plot will decrease over successive DBMS generations.
+
+That said, all is not rosy in this picture. Each new operator is applicable
+to a successively smaller portion of the queries, and perhaps over a
+successively smaller portion of the cardinality space.  As already noted,
+our causal model predicts that the prevalence of suboptimality would
+increase as the number of operators available increased.  It seems that even
+with DBMS implementers doing smart things, these two considerations predict
+that the average relative delta (itself a slope of the performance curve)
+will {\em increase}, as suboptimality (a positive relative delta) becomes
+more prevalent.
+
+This analysis suggests then that the performance curve will thus start to
+level off. That raises the
+possibility of the performance curve either asymptotically approaching a horizontal line
+(the first derivative approaching 0), or worse: the first derivative (the
+average relative delta)
+changing to positive, with the average time over the queries we are studying
+actually {\em increasing} with \hbox{DBMS} generation.
+
+So the question comes down to this:  {\em by
+how much} is the increase in efficiency enabled by a new operator (for those
+Q@Cs utilizing a plan containing that operator) greater than the decrease in
+efficiency resulting from suboptimality (for a subset of those Q@Cs)
+resulting from adding that operator? Is there a
+point where the decrease due to suboptimality obviates the increase enabled
+by the new operator: a DBMS generation where the average relative delta
+becomes positive? How close might modern \hbox{DBMSes} be to that limit?
+
+\subsection{Have Modern DBMSes Hit The Wall?}\label{sec:hitthewall}
+We partition the relevant four sets of change points discussed in the
+Section~\ref{sec:usingCP} into two groups: (i)~those
+with a {\em positive} relative delta (either forward or backward), denoting a
+{\em suboptimal decision} by the query optimizer at the later generation
+(corresponding to a higher generation number)
+and (ii)~those with a {\em negative} relative delta (either forward or
+backward) relative delta, indicating a {\em non-suboptimal decision} by the
+query optimizer at the later generation. (We can't state unequivocally that
+the plan is optimal, because there may be yet another plan involving the
+operators within that generation that is even faster.)
+
+Let's first examine those change points for which the query optimizer made a
+good decision, the non-suboptimal change points; see
+Table~\ref{tab:non-subopt}.
+
+The first thing to note is that this data aggregates the results over the
+four \hbox{DBMSes}, which had generations ranging from five to 13. Thus this is the
+first study we are aware of that compares the generational trends of \hbox{DBMSes}
+over quite disparate code bases, thus getting at fundamental trends.
+
+Overall, only a small number of change points, 1378, or about 2\% of the
+total, satisfied the requirements listed above, including having a different
+generation number for the lower and upper plans. In fact, there were no such
+change points for the first three generations nor for the last
+generation. The second column states the number of change points added by
+that generation and the average relative delta across just those change
+points. The last column states the {\em cumulative relative delta} for that
+generation, defined as
+the sum of the average relative delta for those change points associated
+with plans having only operators in its generation, divided by the number of
+change points. This is the relevant number for a generation, as it includes
+all plans that would have been emitted by that generation, using the
+operators at that generation's disposal.
+
+While the numbers jump around quite a bit, especially for the first few
+generations. The thing to focus on is the last column, where the cumulative
+relative delta starting off at a low of
+-0.28 (recall that low negative value is good, as it indicates the
+additional operator was effective at lowering the execution time) and slowly decreasing to -0.23:
+larger negative numbers trending to lower negative
+numbers.
+
+This general behavior matches our prediction arising from the structural causal model that it
+gets harder to squeeze out performance gains as operators are added to the
+\hbox{DBMS} over successive generations.
+
+We now examine those change points for which the query optimizer made a poor
+decision, in that we can surmise that there was a better plan (the one right
+next to it in the adjacent Q@C pair). Table~\ref{tab:subopt} provides the
+same information across the \hbox{DBMS} generations for the suboptimal change
+points: those for which the
+  relative deltas are positive, indicating a poor decision, as the query
+  time increased. While there are
+  still only a small number of change points, there are a greater number of
+  {\em suboptimal} change points, with more relatively showing up at more
+  recent generations. But more strikingly, the cumulative relative delta has
+  the opposite behavior to the non-suboptimal change points: it {\em
+    increases} over the generations, starting at 0.08 and ending at 0.25, a
+  value {\em higher} than that of the non-suboptimal change points.
+
+The last column brings the non-suboptimal and suboptimal together, stating
+the {\em net cumulative relative delta}, gathering all of the change points
+with plans that would have been generated by that DBMS generation
+together. We see that the net starts out \hbox{at -0.24}, indicating that the new
+operators are {\em decreasing} the query time for the
+workload. Unfortunately, this happy situation starts to deteriorate: with
+each successive generation, the improvement is less. This is as predicted:
+the optimizer is struggling with both more options (plans over the available
+operators) to select from and a diminished opportunity to make a significant
+improvement.
+
+\begin{table}[t]
+\tbl{Non-Suboptimal Change Points\label{tab:non-subopt}}{%
+%\begin{center}
+\begin{tabular}{c|c|c|c}
+{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}\\
+& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}\\
+\hline
+1 & --- & --- & 0.00\\
+2 & --- & --- & 0.00\\
+3 & --- & --- & 0.00\\
+4 & 24 & -0.28 & -0.28\\
+5 & 282 &-0.118 & -0.130\\
+6 & 309 &-0.274 & -0.203\\
+7 & 336 & -0.317 & -0.243\\
+8 & 43 & -0.20 & -0.241\\
+9 & 4 & -0.1 & -0.241\\
+10 & 17 &-0.29 & -0.241\\
+11 & 357 & -0.178 & -0.225\\
+12 & 6 & -0.2 & -0.225\\
+13 & 0 & --- & -0.225\\
+{\em Cumulative}  &1378 & ---& -0.225\\
+\end{tabular}
+%\end{center}
+}
+\end{table}
+
+\begin{table}
+[t]
+\tbl{Suboptimal Change Points\label{tab:subopt}}{%
+%\begin{center}
+\begin{tabular}{c|c|c|c|c}
+{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}&{\em Net Cumulative}\\
+& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}& {\em Relative Delta}\\
+\hline
+1 & --- & --- & 0.00&---\\
+2 & --- & --- & 0.00&---\\
+3 & --- & --- & 0.00&---\\
+4 & 2 & 0.2 & 0.2&-0.24\\
+5 & 211 & 0.076 & 0.077&-0.045\\
+6 & 86 & 0.16 & 0.102&-0.103\\
+7 & 263 & 0.303 & 0.196&-0.0800\\
+8 & 273 & 0.303 & 0.231&-0.0256\\
+9 & 3 & 0.04 & 0.230&-0.0257\\
+10 & 168 & 0.319 & 0.245&0.008\\
+11 & 639 & 0.237 & 0.242&0.0296\\
+12 & 53 & 0.35 & 0.245&0.0346\\
+13 & 12 & 0.40 & 0.246&0.0360\\
+{\em Cumulative} & 1710 & --- & 0.246&0.0360\\
+\end{tabular}
+%\end{center}
+}
+\end{table}
 
@@
 
-The center graph shows the ``net benefit,'' that is, the cumulative average
-relative delta
-of both the suboptimal decision and non-suboptimal decision change
-points. (Note that each point is not simply the sum of the non-suboptimal
-and suboptimal cumulative
-deltas, as the number of change points differ for each generation, but rather the cumulative
-relative delta over all of the change points for that and prior generations.) There is a dramatic
-rise at generation 5, with the rest of the generations presenting a
-relatively smooth growth, moving from negative relative delta (not
-suboptimal) to positive relative delta (suboptimal), crossing around
-generation 10. This highly aggregated result, extracting the 3000-odd
-change-point pairs from almost 100-thousand Q@Cs having the specified
-properties of interest, implies that these four particular \hbox{DBMSes}, taken
-together, might be close to hitting the wall, where adding an operator
-actually slows down the average query.
-
-More specifically, we can't say whether any of these DBMSes have actually
+At Generation 10, the
+net actually becomes very slightly positive: meaning that the performance
+curve has hit a minimum and is now on its way up, towards slower
+performance. This trend increases in later generations.
+
+Figure~\ref{fig:benefit} tells this story graphically, with the net benefit
+in the center, having a least squares slope of -0.024, which is consistent with the
+transition from a generation being net beneficial to actually being slightly
+suboptimal around generation 10 and more so at later generations.
+
+This highly aggregated result, extracting 3000-odd change-point pairs having
+the specified properties of interest stated in Section~\ref{sec:trends} and
+drawn from almost \hbox{100-thousand} Q@Cs implies that these four particular
+\hbox{DBMSes}, taken together, might be close to or have already hit the wall, where
+adding an operator actually slows down the average query.
+
+And we recall the analysis in Section~\ref{sec:CPQ} in which a related
+problem cropped up around generation 8.
+
+It is important to emphasize that we can't say whether any of these DBMSes have actually
 transitioned to where, for this class of queries, the errors of the
@@
 
-We should emphasize all the provisos mentioned earlier. We are looking at
+We should also reiterate all the provisos mentioned earlier. In all of these
+experiments, we are looking at
 quite simple queries, over a quite limited range of data, with only a small
@@
 Section~\ref{sec:interactions} states), relational data is generally much less uniform
-in its structure (we used only integers), values (the values in our
+in its structure (e.g., we used only integer data), values (the values in our
 tables are quite evenly distributed), schema (the schemas of our tables
-are identical), and range of sizes (the smallest table is 1/200-th of the
-largest table), which should minimize the suboptimality.
+are identical and extremely simple), and range of sizes (the smallest table is 1/200-th of the
+largest table); all of these factors should independently, and certainly in concert, minimize the suboptimality.
 
@@
 We studied a particular
-phenomenon, that of the \hbox{optimizer} selecting a wrong plan, indicated by the existence of a query plan that performs more efficiently
+phenomenon, suboptimality, when the \hbox{optimizer} selecting a wrong
+plan. This phenomenon is indicated by the existence of a query plan that performs more efficiently
 than the DBMS's chosen plan, for the same query. From the engineering
 \hbox{perspective}, it is of critical importance to understand the prevalence
-of suboptimality and its causal factors. The genesis of our model was a sense that suboptimality
+of suboptimality and its causal factors. The genesis of our predictive model was a sense that suboptimality
 is caused in part by the inherent complexity of these system and the
@@
 
-Through a series of experiments managed by \azdb, we uncovered
+Through a series of experiments managed by our laboratory instrument
+management system, \azdb, carried out across several years, we uncovered
 several surprising results that provide systematic clues as to where current optimizers come up
@@
 cardinality, even when the cardinality estimates were completely accurate
-and even for our rather simple queries.
-
-\item A quarter of the queries exhibited significant suboptimality ($>= 20\%$ of
+and even for our quite simple queries.
+
+\item A quarter of the queries exhibited significant suboptimality ($\geq 20\%$ of
         the runtime) at some cardinality.
@@
 \item This research indicates that aggregates are {\em not} a problem, so
-  that aspect of query optimization is in good shape.
-
-\todo{Rick: evaluate following shorten:}\shorten{REVISE to characterize
+  that aspect of query optimization is in good shape.\todo{Rick: check with
+    Sabah's results and modify around here if needed.}
+
+\shorten{REVISE to characterize
   more precisely the relative
@@
 
-\shorten{The number of operators a \hbox{DBMS} supports is implicated in both root causes,
-as well as explicitly in the causal model. Consider the following {\em
-  Gedanken experiment}. We add an \hbox{operator} to a \hbox{DBMS}. Doing so will certainly
-reduce the running time of some queries, those that benefit from the new
-operator, and will not affect the running time of those query plans that do
-not include the new operator. That sounds like a sure-fire path to improving
-the performance the \hbox{DBMS}.
-
-However, due to the factors discussed above, sometimes the query
-optimizer will choose a non-optimal plan. Our analysis shows that as the number
-of operators goes up, a suboptimal plan is chosen more frequently, so
-the percentage of suboptimal plans will increase. Also, each new operator will
-impact a smaller set of queries, but the set of suboptimal queries will grow.
-At some point, the beneficial impact of each new operator will be
-offset by the increased prevalence of suboptimality, and the \hbox{DBMS} will get
-{\em slower, on average}. Of course, that inflection point is different for
-each \hbox{DBMS}, but the fact that suboptimality is widespread across all the
-\hbox{DBMSes} for the simple queries in this paper raises the possibility that we
-are already there.
-}
-
 We mentioned dynamic query optimization in Section~\ref{sec:related}
@@
 
-\todo{Rick:discuss implications of generational study}
+The generational study in Section~\ref{sec:diminishing} though implies that
+the challenge in daunting. That study validates the implications of the
+causal model in Figure~\ref{fig:model}, which correctly predict  the almost
+inexorable rise in net cumulative relative delta, which implies that the
+suboptimizer will eventually hit the wall where it is no longer improving.
 
@@
 systems at a fundamental level.  Good engineering should be built on solid
-scientific results.
+scientific results. This paper focuses on the latter.
 
@@
 This paper makes the following contributions in
-an \hbox{attempt} to gain new understanding of this component.
+an \hbox{attempt} to gain new understanding of this
+component.\todo{Rick:make consistent with list at the beginning of the paper}
 \begin{itemize}
@@
 
-\item Provides a {\em space of hypotheses} that in concert through further
-  investigation can refine and support (or not) this model or suggest a
-  better model.
-
 \item Uncovers compelling {\em evidence} (a)~that
@@
 presence of primary key attributes, moderates these three
-interactions.\todo{Check again}
-
-\todo{Keep the following shorten?}\shorten{\item A separate analysis suggested that heuristic plan selection
+interactions.\todo{Rick:Check again}
+
+\shorten{\item A separate analysis suggested that heuristic plan selection
   appears to also be a source of suboptimality.}
@@
 space complexity, in concert predicted a significant portion of the variance
-of suboptimality.  It is doubtful that any other factor, as yet unknown,
+of suboptimality\todo{Rick: state. Do we want to keep the following?}.  It is doubtful that any other factor, as yet unknown,
 will itself predict as much variance as the factors we studied in this
@@
 
-\item Applies a novel experiment over the pairs of adjacent Q@Cs to show
-  \todo{Rick:what?}
+\item Applies a novel experiment over pairs of adjacent Q@Cs to show that
+  such an upper bound is probably and may have already been reached with one
+  or more our our subject DBMSes.
+
+\item Identifies {\em specific directions for engineering interventions}.
 
@@
 
-\item Thereby identifies {\em specific directions for engineering interventions}.
-
 \end{itemize}
 This paper thus suggests a framework of casual model elaboration and
-directed engineering efforts that could reduce the prevalence of query suboptimality to an acceptably
-low level.
+directed engineering efforts.
 
 \section{Future Work}\label{sec:future}
-There are several directions we wish to take this work. With the model refinements
+There are several directions this work could be take. \todo{RICK IS HERE}With the model refinements
 we propose here, additional directed pointers to engineering efforts should
-emerge.
+emerge.\todo{Rick:DOn't say ``we'', but just say the research needs to be done}
 
@@
 to encourage the optimizer to produce more plans at a given cardinality,
-that can then be timed to make more explict the entire plan space (recall that
+that can then be timed to make more explicit the entire plan space (recall that
 CEPS, the cardinality of the effective plan space, is no greater and
@@
 N.~Kabra and D.~J.~DeWitt, ``Efficient mid-query re-optimization of
-sub-optimal query execution plans,'' in {\em Proceedings of the ACM SIGMOD Conference}, pp.~106--117, 1998.
+suboptimal query execution plans,'' in {\em Proceedings of the ACM SIGMOD Conference}, pp.~106--117, 1998.
 
@@
 
+\bibitem{Melton11}
+J.~Melton (editor), ISO/IEC 9075, Database Language SQL:2011 Part 2:
+SQL/Foundation, December, 2011.
+
 \bibitem
@@
 
-\bibitem{Melton11}
-J.~Melton (editor), ISO/IEC 9075, Database Language SQL:2011 Part 2:
-SQL/Foundation, December, 2011.
 \bibitem
@@
 Table~\ref{tab:run_stat2} gives some of those details. We'll walk through
-the columns in succession.\todo{Young: please fill in last column}
+the columns in succession.
+
+The third column states the data set used in each experiment, that is, the
+specific tables being queried. There are four data sets, named A, B, C, and
+D.
+
+We first discuss the features shared between the four data sets. As
+introduced in Section~\ref{sec:motivation}, the queries referenced tables
+{\tt ft\_HT1}, {\tt ft\_HT2}, {\tt ft\_HT3}, and {\tt ft\_HT4}. All four
+tables contain four columns, each of type integer. The specific values of
+the rows for all but the first column depend
+on the percentage of skewed data. Section~\ref{sec:datacomplexity} provides
+the algorithm for generating the values for different values of skew; this
+algorithm is used in columns two--four, which for any row will have
+identical values. The first column holds a unique integer starting from 1
+and going to 60K or 2M, for use in an optional primary key.
+
+There was one version of the
+last three tables, for use with MySQL, with cardinality 60K, and one version
+for the rest of the DBMSes, with cardinality 2M.
+
+We generate 200 versions of {\tt ft\_HT1}, termed the {\em variable table}. For MySQL, these version contain
+300, 600, 900, 1200, $\ldots$, 59,700, and 60,000 rows; for the rest of the
+DBMSes, these versions contain 10,000, 20,000, 30,000, $\ldots$, 1,970,000,
+and 2M rows, as introduced in Section~\ref{sec:motivation}.
+
+We now differentiate the data sets, elaborating on the discussion in Sections~\ref{sec:datasets}--\ref{sec:experiments}. Data Set C is perhaps the simplest to
+describe: it specifies no primary key, has no duplicate rows, and has no skew
+(of course, for any of the four tables). Data Set A differs from Data Set C
+only in that there is skew.  As summarized at the end of
+Section~\ref{sec:datacomplexity}, we use five skewness values: 0
+(approximately), 0.001, 0.1, 0.5, and 1.0.
+
+Data Set B is similar to Data Set A, adding the specification of the first
+column as the primary key. And Data Set D is similar to Data Set B, adding
+the specification that the other three columns should each be associated
+with a secondary index, only only that one column.
+We see the confirmatory examined a much larger variation of data sets than
+the exploratory studies.
+
+The next column of Table~\ref{tab:run_stat2} concerns the {\em Lab
+  Shelf}. \hbox{\azdb} utilizes the metaphor of a bookshelf of lab
+notebooks. Here, each shelf is associated with a version of \hbox{\azdb} itself. For the experiments in this paper, we used at various times over
+the last three years lab shelves (that is, program versions) 5.19, 5.2,
+6.0, and 7.1. Versions 5.19 and 5.2 were very similar; both implemented
+TTPv1.  Version 6.0 also implemented TTPv1, but collected more query
+measures that were not relevant for this paper. Version 7.1 implemented
+TTPv2.
+
+The \hbox{\azdb\ system also includes support} for {\em experiment scenarios}, which
+are Java code that actually performs the experiment, such as varying the
+cardinality and running different queries on the data. The only difference
+in the scenario code across these experiments was in accommodating the
+details of the data set (that is, creating secondary indexes and data skew)
+and adding robustness features, such as using exponential backoff when the
+network connection to the DBMS was temporarily lost.
+
+The bottom line is that while the lab shelf and experiment scenario varied
+somewhat, the only important aspect was the {\em Protocol} column of
+Table~\ref{tab:run_stat}.
+
+We now turn to the fifth column, ``what was examined?'' Here there are just two
+possibilities, all 200 cardinalities or just the cardinalities at which the
+query plan changed. The sixth column, ``what was timed?'', indicates that experiments three and four, Exhaustive and
+Exhaustive with Keys, described in Sections~\ref{sec:discontinuity} and \ref{sec:experiments}, just
+collected query plans, and thus took much less time to run.
+
+The last column states how many Q@Cs the experiments measured (each with 10
+QEs), termed {\em raw}, and how many Q@Cs were retained after the protocol
+implied by the fourth column and listed explicitly in the third column of
+Table~\ref{tab:run_stat} dropped query executions and Q@Cs via its many
+  sanity checks.
 
@@
 & {\em Experiment}& {\em Data Sets} & {\em Lab Shelves} & {\em What was} &{\em Was query}&{\em Number of}\\
-&                 & {\em Used}      &                   & {\em Examined?}&{\em Timed?}   &{\em Retained (Raw) Q@Cs}\\
+&                 & {\em Used}      &                   & {\em Examined?}&{\em Timed?}   &{\em Retained (Raw)}\\
+&&&&&& {\em Q@Cs}\\
 \hline
-1 & Monotonicity 	& A & 6.0     &all cardinalities&yes& 12000 (12000)\\
-2 & Initial Exhaustive 	& A & 5.19 + 6.0&all cardinalities&yes& 27948 (32000)\\
-3 & Exhaustive 		& A & 7.1     &all cardinalities&no & 29515 (32000)\\
-4 & Exhaustive with Keys & B& 6.0     &change points&no& --- \\
-5 & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8171 (8842)\\
-6 & Exploratory 	& A + B&7.1   &change points&yes& 12100 (12560)\\
-7 & Confirmatory 	&A + B + C +D&7.1&change points&yes& 94502 (99558)\\
+1 & Monotonicity 	& A & 6.0     &all cardinalities&yes& 12,000 (12,000)\\
+2 & Initial Exhaustive 	& A & 5.19 + 6.0&all cardinalities&yes& 27,948 (32,000)\\
+3 & Exhaustive 		& A & 7.1     &all cardinalities&no & 29,515 (32,000)\\
+4 & Exhaustive with Keys & B& 6.0     &change points   &no& --- \\
+5 & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8,171 (8,842)\\
+6 & Exploratory 	& A + B&7.1   &change points   &yes& 12,100 (12,560)\\
+7 & Confirmatory 	&A + B + C +D&7.1&change points&yes& 94,502 (99,558)\\
+\multicolumn{2}{c|}{\em Total}         &            & &               &   & 184,236 (196,960)\\
 \end{tabular}
@@
 
-The third column states the data set used in each experiment, that is, the
-specific tables being queried. There are four data sets, named A, B, C, and
-D.
-
-We first discuss the features shared between the four data sets. As introduced in
-Section~\ref{sec:motivation}, the queries referenced tables {\tt ft\_HT1},
-{\tt ft\_HT2}, {\tt ft\_HT3}, and {\tt ft\_HT4}. All four tables contain four
-columns, each of type integer. The specific values of the
-rows for all four columns depend on the percentage of skewed
-data. Section~\ref{sec:datacomplexity} provides the algorithm for generated the
-values for different values of skew; this algorithm is used in each column,
-so all four columns in any row will have identical values.\todo{Young: is
-  this true?}
-
-There was one version of the
-last three tables, for use with MySQL, with cardinality 60K, and one version
-for the rest of the DBMSes, with cardinality 2M.
-
-We generate 200 versions of {\tt ft\_HT1}, termed the {\em variable table}. For MySQL, these version contain
-300, 600, 900, 1200, $\ldots$, 59,700, and 60,000 rows; for the rest of the
-DBMSes, these versions contain 10,000, 20,000, 30,000, $\ldots$, 1,970,000,
-and 2M rows, as introduced in Section~\ref{sec:motivation}.
-
-We now differentiate the data sets, elaboring on the discussion in Section~\ref{sec:datasets}--\ref{sec:experiments}. Data Set C is perhaps the simplest to
-describe: it specifies no primary key, has no duplicate rows, and has no skew
-(of course, for any of the four tables). Data Set A differs from Data Set C
-only in that there is skew.  As summarized at the end of
-Section~\ref{sec:datacomplexity}, we use five skewness values: 0
-(approximately), 0.001, 0.1, 0.5, and 1.0.
-
-Data Set B is similar to Data Set A, adding the specification of the first
-column as the primary key. And Data Set D is similar to Data Set B, adding
-the specification that the other three columns should each be associated
-with a secondary index, only only that one column.
-We see the confirmatory examined a much larger variation of data sets than
-the exploratory studies.
-
-The next column of Table~\ref{tab:run_stat2} concerns the {\em Lab
-  Shelf}. \hbox{\azdb} utilizes the metaphor of a bookshelf of lab
-notebooks. Here, each shelf is associated with a version of \hbox{\azdb} itself. For the experiments in this paper, we used at vearious times over
-the last three years labh shelves (that is, program versions) 5.19, 5.2,
-6.0, and 7.1. Versions 5.19 and 5.2 were very similar; both implemented
-TTPv1.  Version 6.0 also implemented TTPv1, but collected more query
-measures that were not relevant for this paper. Version 7.1 implemented
-TTPv2.
-
-The \hbox{\azdb\ system also includes support} for {\em experiment scenarios}, which
-are Java code that actually performs the experiment, such as varying the
-cardinality and running different queries on the data. The only difference
-in the scenario code across these experiments was in accommodating the
-details of the data set (that is, creating secondary indexes and data skew)
-and adding robustness features, such as using exponential backoff when the
-network connection to the DBMS was temporarily lost.
-
-The bottom line is that while the lab shelf and experiment scenario varied
-somewhat, the only important aspect was the {\em Protocol} column of
-Table~\ref{tab:run_stat}.
-
-We now turn to the fifth column, ``what was examined?'' Here there are just two
-possibilities, all 200 cardinalities or just the cardinalities at which the
-query plan changed. The sixth column, ``wheat was timed?'', indicates that experiments three and four, Exhaustive and
-Exhaustive with Keys, described in Sections~\ref{sec:discontinuity} and \ref{sec:experiments}, just
-collected query plans, and thus took much less time to run.
-
+\todo{Rick: I (Young) updated the following. Please review it.}
 The following sets of queries were used in the seven experiments.
 
-\todo{Young: how does the following look? Any changes needed?}
 \begin{description}
@@
   drawn from each of {\em QSg}--{\em QSj}
-%\item[QSlm] A primary key query set consisting of 390 queries from QSl and QSm
-\item[QSn] A primary key query set 4 with 110 new queries
-\item[QSo] A no-skew query set consisting of 100 queries without aggregates for no skew
+\item[QSn] A primary key query set consisting of 110 new queries
+\item[QSo] A query set consisting of 100 queries without aggregates (for Data Set C)
 \item[QSp] A subquery query set consisting of 100 queries, each with a subquery
 \item[QSq] A subquery query set consisting of 100 queries, each with a subquery
-\item[QSr] A primary key query set consisting of 100 queries drawn from {\em
-  QSl} and {\em QSm}
-\item[QSab'] 100 queries drawn from {\em QSa} and {\em QSb} for which the where
-     selection involved a primary key column
-\todo{Young what does this mean: ``This query set was revised to QSI and
-  QSm.''? Als, it doesn't seem like this set is used anywhere...}
+\item[QSr] A query set consisting of 390 queries drawn from {\em QSl} + {\em QSm}
+\item[QSs] A primary key query set consisting of 100 queries drawn from {\em QSr}
 \end{description}
-
-\todo{Rick: discuss last column}
 
@@
 one DBMS), the first 20 queries from {\em QSa}--{\em QSf} (for another DBMS), the first
-10 (primay key) queries from {\em QSa} (for all four DBMSes), and the first 10
+10 (primary key) queries from {\em QSa} (for all four DBMSes), and the first 10
 (non-primary key) queries from {\em QSa} (for the other two DBMSes), for a total of 780 query instances.
@@
 Experiment 6 (Exploratory) used {\em QSa} and {\em QSb}, plus the first 100 (primary
-key) queries from {\em QSl} and {\em QSm}, for all four DBMSes, for a total of 1200 query
+key) queries from {\em QSr}, for all four DBMSes, for a total of 1200 query
 instances.
 
-Experiment 7 (Confirmatory) used {\em QSc}--{\em QSr}, along with {\em QSp} for primary key and
-subquery, {\em QSp} for subquery, {\em QSp} and {\em QSq} for primary key and secondary index and
-subquery), all across all four DBMSes, for a total of 7,640 query instances.
+Experiment 7 (Confirmatory) used {\em QSc}--{\em QSj}, 
+along with {\em QSr} except the (first 100) queries included in Experiment 6, 
+{\em QSn} for primary key (for two runs, or 220 queries), 
+{\em QSo} for no data skew, 
+{\em QSp} for primary key and subquery, 
+{\em QSp} for subquery, 
+{\em QSp} and {\em QSq} for primary key and secondary index and subquery, 
+{\em QSs} for primary key and secondary index, 
+all across all four DBMSes, for a total of 7,640 query instances.
 

### 2017-02-23 pt_characterization/new_data/emp_et_hist.r

@@ -1,295 +1,0 @@
-x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("1_sec_et_hist.eps")
-xmax <- max(x$METIME)
-xmin <- min(x$METIME)
-h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
-ymax <- max(h$counts)
-ymax <- ceiling(ymax/100)*100
-binsize=1
-plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
-sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
-xlab='ET (ms)', ylab=expression('Frequency'))
-xaxtl <- seq(xmin-1,xmax+1,by=1)
-axis(side=1, at=xaxtl, labels=xaxtl)
-dev.off()
-
-x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("2_sec_et_hist.eps")
-xmax <- max(x$METIME)
-xmin <- min(x$METIME)
-h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
-ymax <- max(h$counts)
-ymax <- ceiling(ymax/100)*100
-binsize=1
-plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
-sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
-xlab='ET (ms)', ylab=expression('Frequency'))
-xaxtl <- seq(xmin-1,xmax+1,by=1)
-axis(side=1, at=xaxtl, labels=xaxtl)
-dev.off()
-
-x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("4_sec_et_hist.eps")
-xmax <- max(x$METIME)
-xmin <- min(x$METIME)
-h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
-ymax <- max(h$counts)
-ymax <- ceiling(ymax/100)*100
-binsize=1
-plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC4', 
[…]

### 2017-02-23 pt_characterization/new_data/emp_pt_hist.r

@@ -1,236 +1,0 @@
-x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("1_sec_pt_hist.eps")
-h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
-binsize=1
-plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+1),col="green", main='PT frequency on INC1', 
-sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
-xlab='PT (ms)', ylab=expression('Frequency'))
-xaxtl <- seq(min(x$PRTIME)-1,max(x$PRTIME)+1,by=1)
-axis(side=1, at=xaxtl, labels=xaxtl)
-#axis(side=1, at=seq(min(h$mids)-1,max(h$mids)+1,1), labels=seq(min(x$PRTIME),max(x$PRTIME)+1,by=1))
-dev.off()
-
-x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("2_sec_pt_hist.eps")
-h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
-binsize=1
-plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+1),col="green", main='PT frequency on INC2', 
-sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
-xlab='PT (ms)', ylab=expression('Frequency'))
-dev.off()
-
-x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
-setEPS()
-postscript("4_sec_pt_hist.eps")
-h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
-binsize=1
-plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+2),col="green", main='PT frequency on INC4', 
-sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
-xlab='PT (ms)', ylab=expression('Frequency'))
-dev.off()
-
[…]

### 2017-02-23 pt_characterization/new_data/emp_v5_hist.r

@@ -1,341 +1,0 @@
-x = read.csv(file="128_sec.dat",head=TRUE,sep="\t")
-#x <- subset(x, x$ITERNUM <= 10)
-uol = mean(x$METIME) + 2*sd(x$METIME)
-x.sub = subset(x, x$METIME <= uol)
-x.ol = subset(x, x$METIME > uol)
-nrow(x.sub)
-nrow(x.ol)
-uol = mean(x.sub$PRTIME) + 2*sd(x.sub$PRTIME)
-dol = mean(x.sub$PRTIME) - 2*sd(x.sub$PRTIME)
-x.ref = subset(x.sub, x.sub$PRTIME <= uol & x.sub$PRTIME >= dol)
-x.ol2 = subset(x.sub, x.sub$PRTIME > uol | x.sub$PRTIME < dol)
-nrow(x.ref)
-nrow(x.ol2)
-min(x.ref$PRTIME)
-max(x.ref$PRTIME)
-round(mean(x.ref$PRTIME),0)
-round(sd(x.ref$PRTIME),1)
-sd(x.ref$PRTIME)/mean(x.ref$PRTIME)
-
-### 1 sec.dat
-x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
-x <- subset(x, x$ITERNUM <= 10)
-#x <- subset(x, x$ITERNUM <= 800)
-uol = mean(x$METIME) + 2*sd(x$METIME)
-x.sub = subset(x, x$METIME <= uol)
-x.ol = subset(x, x$METIME > uol)
-nrow(x.sub)
-nrow(x.ol)
-uol = mean(x.sub$PRTIME) + 2*sd(x.sub$PRTIME)
-dol = mean(x.sub$PRTIME) - 2*sd(x.sub$PRTIME)
-x.ref = subset(x.sub, x.sub$PRTIME <= uol & x.sub$PRTIME >= dol)
-x.ol2 = subset(x.sub, x.sub$PRTIME > uol | x.sub$PRTIME < dol)
-nrow(x.ref)
-nrow(x.ol2)
-min(x.ref$PRTIME)
-max(x.ref$PRTIME)
-round(mean(x.ref$PRTIME),0)
-round(sd(x.ref$PRTIME),1)
-sd(x.ref$PRTIME)/mean(x.ref$PRTIME)
-
-### 2 sec.dat
-x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
[…]

### 2017-02-23 pt_characterization/recent_data/sodb9/et_hist_sodb9.r

@@ -1,4 +1,7 @@
+### sodb9
+numSamples <- 300
 x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
 postscript("1_sec_et_hist_v5.eps")
 binsize=1
@@ -19,6 +22,7 @@ axis(side=1, at=xaxtl, labels=xaxtl)
 dev.off()
 
 x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
 postscript("2_sec_et_hist_v5.eps")
 binsize=1
@@ -39,6 +43,7 @@ axis(side=1, at=xaxtl, labels=xaxtl)
 dev.off()
 
 x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
 postscript("4_sec_et_hist_v5.eps")
 binsize=1
@@ -49,20 +54,21 @@ nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
 h = hist(x$METIME, right=F, breaks=nbins,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
+ymax <- 80
 xmin <-min(x$METIME)
 xmax <-max(x$METIME)
-plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC4', 
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-2,xmax+2),col="blue", main='ET frequency on INC4', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='ET (ms)', ylab=expression('Frequency'))
-xaxtl <- seq(xmin-1,xmax+1,by=1)
 axis(side=1, at=xaxtl, labels=xaxtl)
 dev.off()
 
 x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
 postscript("8_sec_et_hist_v5.eps")
 binsize=1
-x <- subset(x, x$ITERNUM != 629 & x$ITERNUM != 658)
[…]

### 2017-02-23 pt_characterization/recent_data/sodb9/pt_hist_sodb9.r

@@ -1,0 +1,234 @@
+### sodb9
+numSamples <- 300
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+x <- subset(x, x$ITERNUM <= numSamples)
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- 250
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC2', 
[…]

### 2017-02-23 mlppi-comsis17/ComSISTemplateLaTeX/ComSIS-typeinst.tex
comsis version of MLPPI

\documentclass[runningheads]{comsis2}

\def\journalissue{Computer Science and Information Systems 0(0):1--6}
\def\paperidnum{DOI: N/A}
\setcounter{page}{1}

\usepackage[dvipdfm]{hyperref}
\usepackage{verbatim}
\usepackage{graphicx}

\title{Authors' Instructions for the Preparation\\of Camera-Ready Contributions with \LaTeX}
\titlerunning{Authors' Instructions}
\author{Vladan Deved\v{z}i\'c\inst{1} \and Marijana Despotovi\'c\inst{1} \and Ivan Lukovi\'c\inst{2} \and
        Violeta Damjanovi\'c\inst{3} \and Milo\v{s} Radovanovi\'c\inst{4}}
\authorrunning{Author(s)}
\institute{Faculty of Organizational Sciences, POB 52\\
  11000 Belgrade, Serbia\\
  \email{\{devedzic,maja\}@fon.rs}
  \and
  Faculty of Technical Sciences, Trg D.\ Obradovi\'ca 6\\
  21000 Novi Sad, Serbia\\
  \email{ivan@uns.ac.rs}
  \and
  Postal Savings Bank, 27.\ marta 71\\
  11000 Belgrade, Serbia\\
  \email{vdamjanovic@posted.co.rs}
  \and
  Dept.\ of Mathematics and Informatics, Trg D.\ Obradovi\'ca 4\\
  21000 Novi Sad, Serbia\\
  \email{radacha@dmi.uns.ac.rs}}

\begin{document}

\maketitle

\begin{abstract}
The abstract should summarize the contents of the paper and should contain at
least 70 and at most 150 words. It should be set using the \texttt{abstract}
environment.

\vspace{6pt}\textbf{Keywords:} computer science, information systems, \LaTeX,
typesetting.
\end{abstract}

\section{Introduction}

These instructions and the corresponding \LaTeX\ style are based
on the corresponding Springer instructions and \LaTeX\ document class for
preparing camera ready papers to be published in the Springer series Lecture
Notes in Computer Science.

The preparation of manuscripts which are to be reproduced by photo-offset
requires special care. Papers submitted in a technically unsuitable form will
be returned for retyping, or canceled if the volume cannot otherwise be
finished on~time.

\section{Manuscript Preparation}

You are encouraged to use \LaTeX\ for the preparation of your camera-ready
manuscript together with the corresponding document class comsis.cls (see
Sect.~\ref{sec:latex}), and send all necessary source files. Please follow
these instructions closely in order to make the volume look as uniform as
possible. Please also send a corresponding pdf file.

We would like to stress that the class/style files should not be manipulated
and that the guidelines regarding font sizes and format should be adhered to.
This is to ensure that the end product is as homogeneous as possible.

\subsection{Printing Area}

The printing area is 125mm $\times$ 197mm. The text should be justified to
occupy the full line width, so that the right margin is not ragged, with words
hyphenated as appropriate. The length of the text on a page should not be less
than~180mm.

\subsection{Layout, Typeface, Font Sizes, and Numbering}

The default font for all text formatting is Times. Use 10-point type for the
name(s) of the author(s) and 9-point type for the address(es) and the abstract.
For the main text, please use 10-point type and single-line spacing. Italic type
may be used to emphasize words in running text. Bold type and underlining should
be avoided. With these sizes, the interline distance should be such that some 45
lines occur on a full-text page.

\subsubsection{Headings.} Headings should be capitalized (i.e., nouns, verbs, and
all other words except articles, prepositions, and conjunctions should be set
with an initial capital) and should, with the exception of the title, be
aligned to the left. Words joined by a hyphen are subject to a special rule.
If the first word can stand alone, the second word should be capitalized.
Headings and appropriate \LaTeX\ commands are enumerated in
Table~\ref{tbl:headings}.

\begin{table}
\caption{Font sizes of headings. Table captions should always be positioned
\emph{above} the tables. Tables should be horizontally centered. The final
sentence of a table caption should end without a period}\label{tbl:headings}
\centering
\setlength\tabcolsep{0.5em}
\begin{tabular}{lll}
\hline
Heading level      & Example                                    & \LaTeX\ command \\ \hline \\[-2ex]
Title (centered)   & \textbf{\Large Authors' Instructions\dots} & \texttt{$\backslash$title} \\
1st-level heading  & \textbf{\large 1.\ Introduction}           & \texttt{$\backslash$section} \\
2nd-level heading  & \textbf{2.1.\ Printing Area}               & \texttt{$\backslash$subsection} \\
3rd-level heading  & \textbf{Headings.} Text follows\dots       & \texttt{$\backslash$subsubsection} \\
4th-level heading  & \textit{Remark.} Text follows\dots         & \texttt{$\backslash$paragraph} \\ \hline
\end{tabular}
\end{table}

Here are some examples of headings: ``Criteria to Disprove Context-Freeness of
Collage Languages'', ``On Correcting the Intrusion of Tracing
Non-deterministic Programs by Software'', ``A User-Friendly and Extendable
Data Distribution System'', ``Multi-flip Networks: Parallelizing GenSAT'',
``Self-determinations of Man''.

\subsubsection{Lemmas, Propositions, and Theorems.} The numbers accorded to lemmas,
propositions, and theorems etc.\ should appear in consecutive order, starting
with the number 1, and not, for example, with the number 11.

\subsection{Figures and Photographs}

Please produce your figures electronically, if possible, and refer to them
from your source file using, for example, the
\texttt{$\backslash$includegraphics} command from package \texttt{graphicx}.
Format the figures using the standard \texttt{figure} environment, and center
them horizontally.

Check that in line drawings, lines are not interrupted and have constant
width. Grids and details within the figures must be clearly readable and may
not be written one on top of the other. Line drawings should have a resolution
of at least 800dpi (preferably 1200dpi). For digital halftones 300dpi is
usually sufficient. The lettering in figures should have a height of 2mm
(10-point type). Figures should be scaled up or down accordingly. Please do
not use any absolute coordinates in figures. If possible, the files of figures
(e.g.\ eps files) should not contain binary data, but be saved in ASCII
format.

Figures should be numbered and should have a caption which should always be
positioned under the figures, in contrast to the caption belonging to a table,
which should always appear above the table. The final sentence of a caption,
be it for a table or a figure, should end without a period. The
\texttt{$\backslash$caption} command will center the captions between the
margins and set them in 9-point type (Fig.~\ref{fig:afigure} shows an
example). The distance between text and figure should be about~8mm, the
distance between figure and caption about~5mm.

\begin{figure}
\centering
\includegraphics[width=.7\textwidth]{figure}
\caption{One kernel at~$x_s$ (\emph{dotted kernel}) or two kernels at $x_i$ and~$x_j$
(\emph{left and right}) lead to the same summed estimate at~$x_s$. This shows
a figure consisting of different types of lines. Elements of the figure described
in the caption should be set in italics, in parentheses, as shown in this sample
caption. The last sentence of a figure caption should generally end without a period}\label{fig:afigure}
\end{figure}

\paragraph{Remark 1.} In the printed volumes, illustrations are generally black and
white (halftones), and only in exceptional cases, and if the author is
prepared to cover the extra cost for color reproduction, are color pictures
accepted. If color illustrations are necessary, please send us color-separated
files if possible. Color pictures are welcome in the electronic version at no
additional cost.

\paragraph{Remark 2.} To ensure that the reproduction of your illustrations is of
reasonable quality we advise against the use of shading. The contrast should
be as pronounced as possible. This particularly applies for screenshots.

\subsection{Formulas}

Displayed equations or formulas are centered and set on a separate line (with
an extra line or halfline space above and below). Displayed expressions should
be numbered for reference. The numbers should be consecutive within each
section or within the contribution, with numbers enclosed in parentheses and
set on the right margin. For example,
\begin{equation}
x + y = z\,.
\end{equation}

Please punctuate a displayed equation in the same way as ordinary text but
with a small space before the end punctuation.

\subsection{Program Code}

Program listings or program commands in the text are normally set in
typewriter font (Courier).

Example of a Computer Program from Jensen K., Wirth N. (1991) Pascal user
manual and report. Springer, New York:

\begin{verbatim}
  program Inflation (Output)
    {Assuming annual inflation rates of 7%, 8%, and
    10%,...   years};
    const  MaxYears = 10;
    var    Year: 0..MaxYears;
           Factor1, Factor2, Factor3: Real;
    begin
      Year := 0;
      Factor1 := 1.0; Factor2 := 1.0; Factor3 := 1.0;
      WriteLn('Year 7% 8% 10%'); WriteLn;
      repeat
        Year := Year + 1;
        Factor1 := Factor1 * 1.07;
        Factor2 := Factor2 * 1.08;
        Factor3 := Factor3 * 1.10;
        WriteLn(Year:5,Factor1:7:3,Factor2:7:3,
          Factor3:7:3)
      until Year = MaxYears
  end.
\end{verbatim}

\subsection{Footnotes}

The superscript numeral used to refer to a footnote appears in the text either
directly after the word to be discussed or -- in relation to a phrase or a
sentence -- following the punctuation sign (comma, semicolon, or period).
Footnotes are set at the bottom of the normal text area, with a line of
about~5cm immediately above them.\footnote{The footnote numeral is set flush
left and the text follows with the usual word spacing. Second and subsequent
lines are indented. Footnotes should end with a period.}

\subsection{Citations}

The list of references is headed ``References'' and is not assigned a number
in the decimal system of headings. The list should be set in small print and
placed at the end of your contribution, in front of the appendix, if one
exists. Please do not insert a pagebreak before the list of references if the
page is not completely filled. An example is given at the end of this
information sheet. For citations in the text please use the
\texttt{$\backslash$cite} command, and organize your references using Bib\TeX,
with \texttt{$\backslash$bibliographystyle\{splncs03\}}, producing numbered
citations in square brackets, such as: \cite{Agrawal94VLDB}, \cite{Enc97LNCS},
\cite{DatabaseBook02}, \dots

\subsection{Page Numbering and Running Heads}

Page numbers and running heads are automatically generated by \LaTeX. If the
title of your paper is too long for the header please specify a running title
using the \texttt{$\backslash$titlerunning} command. Other parameters will be
allocated by the publisher.

\subsection{Printing Quality}

You are normally not expected to send a printed copy of your manuscript if you
provide the source and pdf files matching exactly each other when printed.
Only in exceptional cases, you will be asked to send sheets which are printed
on one side only. In such a case, please use a high-resolution printer,
preferably a laser printer with at least 300dpi. We prefer the text to be
centered on the pages (i.e., equal margins left and right and top and bottom).
The format of the paper (A4, Letter, etc.) is irrelevant.

\section{Using \LaTeX}\label{sec:latex}

We encourage using \LaTeX\ for preparing final manuscripts. We provide the
document class comsis.cls to help \LaTeX\ users prepare their camera-ready
manuscript and to enable us to use their source files. The necessary style
files and documentation can be downloaded from the ComSIS Web page at
\url{http://www.comsis.org/information.php}.

\section{Supplementary Material}

If you wish to include color illustrations in the electronic version in place
of or in addition to any black and white illustrations in the printed version,
please provide the managing editor and the editorial assistant with the
appropriate files.

If you have supplementary material, e.g., executable files, video clips, or
audio recordings, on your server, simply send the managing editor and the
editorial assistant a short description of the supplementary material and
inform them of the URL at which it can be found. We will add the description
of the supplementary material to the online version of the corresponding
ComSIS volume and create a link to your server. Alternatively, if this
supplementary material is not to be updated at any stage, then it can be sent
directly to the managing editor and the editorial assistant, together with all
the other files.

\section{Copyright Form}

For details regarding the transfer of copyright, please see the ``Copyright
and Use Agreement'' section under ``Information for Contributors'' at the
ComSIS Web page. Generally, we ask contributing authors to complete and sign
the ``Transfer of Copyright'' agreement before the paper may be published. (It
is sufficient if one author from each contribution signs the form on behalf of
all the other authors.) The copyright form is located on our Web page at
\url{http://www.comsis.org/download/ComSISCopyright.doc}. The printed form should be
completed and signed and sent to the editorial assistant by normal mail.
Alternatively, a scanned copy of the form may be sent by e-mail.

\section{Checklist}

When submitting your camera-ready manuscript to the managing editor and the
editorial assistant, please make sure you include the following:
\begin{itemize}
  \item your source (input) files, including the figures;
  \item a single-sided printout (not a photocopy) of the final version of your contribution, if requested by the managing editor;
  \item any style files, templates, and special fonts you may have used;
  \item the completed and signed copyright form.
\end{itemize}

If supplementary material is available, please provide the editorial
assistant~with:
\begin{itemize}
  \item a short description of the supplementary material;
  \item the supplementary material or the URL at which it can be found;
  \item the files of color figures for the electronic version.
\end{itemize}

\nocite{Temporal97TDS,CS95,Ontology02}

\bibliographystyle{splncs03}
\bibliography{example}

\end{document}

### 2017-02-23 mlppi-comsis17/cover-letter.tex
comsis version of MLPPI

--- 2017-01-25 sedona-ieice16/submission/misc/cover-letter.tex
@@
 
-\opening{Dear Editor-in-Chief:}
+\opening{Dear Prof. Vladimir Kurbalija (or Editor-in-Chief):}
 
-\noindent This letter to be submitted to IEICE Transactions on Information and Systems 
-presents the following contributions. No version of this letter was submitted earlier.
+\noindent An earlier version of this submission was published in Proceedings of the 21st ACM \hbox{International} Conference on Information and Knowledge Management (CIKM). 
+This version to be submitted to Computer Science and Information Systems includes the following substantial extensions.
 
@@
 
-\item We provide empirical evidence that measuring program time 
-can be seriously affected by extant system daemons.
+\item Expanded Algorithms: we expand the description of the pre-processing, initial and optimized phase algorithms shortened in the earlier version.
+Furthermore, this submission contains three additional algorithms as follows:
 
-\item We propose a novel timing protocol, called {\em SEDONA} (Selective Elimination through Detection of infrequent, lOng-ruNning dAemons), 
-that identifies infrequent, long-running daemons that impact the timing results for that program.
+\begin{itemize}
 
-\item We evaluate the performance of the protocol with rigorous experiments, starting from a simple program in pure-computation mode 
-to a popular industrial benchmark suite.
+\item splitting overlapping ranges in Algorithm~2 on page~4,
+\item checking the allowed field count limit on the split ranges on in Section 2.1.4 on page~5,
+\item computing scan cost in Algorithm~3 on page~6, and
+\item computing query cost in Algorithm~5 on page~7.
 
-\item Experimental results show a strong support for the effectiveness of our SEDONA timing protocol. 
+\end{itemize}
+
+\item Complexity Analysis: we conduct the asymptotic running time analysis on each phase in Section 3 on page 8.
+
+\item Experiments: we provide additional runtime statistics on the workloads used in the experiments in Table 1 on page 10. 
+In addition, we compare the quality of a human expert's recommendation with that of our wizard in Figure 5 on page 10.
+
+\item Expanded Literature Review: the related work section on page~9 elaborates in detail on the similarities and differences between our wizard and existing tools.
+
 \end{itemize}

### 2017-02-23 mlppi-comsis17/old_paper.tex
comsis version of MLPPI

\documentclass[paper]{ieice}

%\usepackage[margin=1in]{geometry}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{balance}
\usepackage{subfigure}
\usepackage{float}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\usepackage{latexsym}
\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{comment}
\usepackage[dvips]{graphicx}          % graphicx package for including ps files

\setcounter{page}{1}
%\breakauthorline{}% breaks lines after the n-th author

\field{D}
%\SpecialIssue{}
%\SpecialSection{}
%\theme{}
\title{MLPPI Wizard: An Automated Multi-Level Partitioning Tool on Analytical Workloads}
\titlenote{An earlier version of this article was presented as a short paper at ACM CIKM'12.}
%\title[title for header]{title}
%\titlenote{}
\authorlist{% fill arguments of \authorentry, otherwise error will be caused. 
 \authorentry[yksuh@kisti.re.kr]{Young-Kyoon SUH}{m}{labelA}
 %\authorentry{Ahmad GHAZAL}{n}{labelB}
 \authorentry{Alain CROLOTTE}{n}{labelC}
 \authorentry{Pekka KOSTAMAA}{n}{labelC}
% \authorentry{name}{membership}{affiliate label}
% \authorentry{name}{membership}{affiliate label}[present affiliate label]
% \authorentry[e-mail address]{name}{membership}{affiliate label}
% \authorentry[e-mail address]{name}{membership}{affiliate label}[present affiliate label]
}
\affiliate[labelA]{Korea Inst. of Sci. and Tech. Info. (KISTI), 245 Daehangno, Yuseong-gu, Daejeon, Korea, 34141}
%\affiliate[labelB]{Oracle Endeca, Cambridge, MA 02142, USA}
\affiliate[labelC]{Teradata Corporation, El Segundo, CA 90245, USA}
%\paffiliate[present affiliate label]{Presently, the author is with the }

\received{2016}{8}{18}
%\revised{}{}{}
%\finalreceived{2011}{1}{1}

\def\form#1{$\langle{#1}\rangle$}
\def\range#1{$[{#1}]$}
\def\openrange#1{$({#1})$}
\def\lopenrange#1{$({#1}]$}
\def\ropenrange#1{$[{#1})$}
\def\product#1{$(#1)$}
\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\long\def\todo#1{\hilitey{{\bf TODO:} {\em #1}}}
\long\def\shorten#1{}
\long\def\comment#1{}
\def\MAX{\mbox{\sl MAX}}
\def\LIMIT{\mbox{\sl LIMIT}}
\def\getQueryCost{\mbox{\sl getQueryCost}}
\def\getScanCost{\mbox{\sl getScanCost}}
\def\BigO{\mbox{\sl O}}
\def\NULL{\mbox{\sl NULL}}
\def\SolA{\mbox{\sl A}}
\def\SolB{\mbox{\sl B}}
\def\SolC{\mbox{\sl C}}
\def\SolS{\mbox{\sl S}}
\def\Qone{\mbox{\sl q1}}
\def\Qtwo{\mbox{\sl q2}}
\def\Sone{\mbox{\sl s1}}
\def\Stwo{\mbox{\sl s2}}
\def\SolSone{\mbox{\sl S1}}
\def\SolStwo{\mbox{\sl S2}}
\def\SolSthree{\mbox{\sl S3}}
\def\SolSfour{\mbox{\sl S4}}
\def\SolSfive{\mbox{\sl S5}}

\begin{document}

\maketitle

\begin{summary}
This article introduces a new tool to \hbox{recommend} 
an optimized multi-level partitioning solution, called 
\hbox{Multi-Level} Partitioned Primary Index (MLPPI), for a fact table accessed by a query workload. 
This tool uses a novel greedy algorithm for search space enumeration. The space is driven by query \hbox{predicates}. 
This \hbox{technique} fits very well Teradata MLPPI, 
as it is based on a \hbox{general} \hbox{framework} \hbox{using} general
expressions, ranges and case expressions for partition 
definitions. 
The tool takes \hbox{advantage} of the optimizer’s cost model to prune the search space for yielding a final recommendation. 
The tool resides completely on the client and interacts with the optimizer via APIs 
used to 1) simplify the workload queries, 
to 2) extract the predicates, and 
to 3) estimate the workload execution cost on a candidate recommendation. 
The predicate-driven method can be applied to any clustering or partitioning schemes, 
based on simple field expressions and complex SQL predicates. 
We demonstrate that the recommendation from the tool outperforms a human expert. The experiments also show that the recommendation is scalable with the workload complexity and the size of the fact table.

\end{summary}
\begin{keywords}
Star Schema, Fact Table, Multi-Level Partitioning
\end{keywords}

\section{Introduction}
\label{sec:intro}

Tables and materialized views in Teradata are hash distributed based 
on a user-specified column or set of columns called {\em primary index}. 
Each virtual processor, a unit of parallelism called {\em AMP} in the \hbox{Teradata} \hbox{relational} 
database management system (DBMS), 
receives a subset of the data and stores it in hash order.

Teradata users typically choose primary index fields, 
so that the data is evenly distributed among the AMPs. 
The primary index fields are also chosen to \hbox{reflect} join fields 
in workloads to accomplish cheaper local joins that do not require 
data shuffling.

PPI (Partitioned Primary Index)~\cite{sinclair:ppi} is an optional horizontal 
partitioning scheme applied locally on each AMP's data. 
Note that in other database products, this type of partitioning would 
most likely to be called {\em clustering}. 
In the rest article we use the term \hbox{{\em partitioning}} only, not clustering, to avoid confusion. 
Database \hbox{administrators} (DBAs) usually choose the columns for PPI, 
based on join fields and single table \hbox{predicates} to optimize the queries included 
in the workload. 
The PPI columns are used to physically cluster data with the same values 
together in contiguous data blocks. 
This allows ``partition elimination'' in scans and joins for performance improvement.
PPI can be specified as single or multiple (or nested) levels. 
This type of partitioning scheme is known as 
Multi-Level PPI ({\em MLPPI)}~\cite{klindt09mlppi}.

By allowing non-qualified partitions to be eliminated, MLPPI can reduce 
significantly the amount of data to be scanned to answer a query. 
But a large number of partitions can create significant overhead, 
particularly in joins and table maintenance operations 
such as inserts and deletes, so that the selection of partitions can be usually a balancing act.

Figure~\ref{fig:exam} exemplifies an MLPPI table with potential partitioning schemes. 
This example is a subset of the {\tt LINEORDER} table from the Star Schema Benchmark (SSB)~\cite{oneil:ssb}. 
Note that the full table with modified data types was actually used in our experiments.

\vspace{-.15in}

\begin{figure}[htp!]
\begin{center}
{\small
\begin{tabular}{l}
{\tt CREATE TABLE LINEORDER(} \\
\hspace{0.1in}{\tt LO\_ORDERKEY INTEGER,} \\
\hspace{0.1in}{\tt LO\_QUANTITY INTEGER,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT INTEGER} \\
{\tt )} \\
{\tt PRIMARY INDEX ( LO\_ORDERKEY )} \\
{\tt PARTITION BY (} \\
\hspace{0.1in}{\tt CASE\_N(} \\
\hspace{0.2in}{\tt LO\_DISCOUNT >= 7,} \\
\hspace{0.2in}{\tt NO CASE OR UNKNOWN),} \\
\hspace{0.1in}{\tt CASE\_N(}\\
\hspace{0.2in}{\tt LO\_QUANTITY < 25,} \\
\hspace{0.2in}{\tt LO\_QUANTITY >= 25 AND LO\_QUANTITY <= 30,} \\
\hspace{0.2in}{\tt LO\_QUANTITY > 30 AND LO\_QUANTITY <= 35,} \\
\hspace{0.2in}{\tt NO CASE OR UNKNOWN)} \\
{\tt );} \\
\end{tabular}
}
\end{center}
%\vspace{-0.1in}
\caption{An Example of MLPPI Table\label{fig:exam}}
\end{figure}

\vspace{-.15in}

In the definition of {\tt LINEORDER} in the figure, the primary index 
is {\tt LO\_ORDERKEY}. 
The \hbox{primary} index \hbox{dictates} the AMP on which a row will be located, 
while the partitioning of data will be dictated by the values and 
ranges associated with {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}. 
{\tt LO\_DISCOUNT}, for example, has two ranges for \hbox{values} 
larger than or equal to 7 and one ({\tt NO CASE OR UNKNOWN}) 
for all the other values, while {\tt LO\_QUANTITY} has four ranges. 
As a result, the relational table will have a total of 2$\times$4 = 8 partitions as follows.

\begin{center}
\begin{tabular}{|c|p{6cm}|}\hline 
Partition & Condition \\ \hline
1		& {\scriptsize  {\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY < 25}} \\ \hline
2		& {\scriptsize  {\tt LO\_DISCOUNT >= 7 \&\& 25 <= LO\_QUANTITY <= 30}} \\ \hline
3		& {\scriptsize {\tt LO\_DISCOUNT >= 7 \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
4		& {\scriptsize {\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY no case}}\\ \hline
5       & {\scriptsize {\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY < 25}} \\ \hline 
6		& {\scriptsize {\tt LO\_DISCOUNT no case \&\& 25 <= LO\_QUANTITY < 30}} \\\hline								  	        
7		& {\scriptsize {\tt LO\_DISCOUNT no case \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
8		& {\scriptsize  {\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY no case}} \\ \hline
\end{tabular}
\end{center}

Whatever partition set is chosen, 
the selection must be semantically correct; namely, 
the mapping of rows to partitions must be an injection. 
In other words, the constraints must form a covering of 
the entire range, so that a row will belong to exactly one and one partition. 
The Teradata optimizer then applies \hbox{partition} elimination for queries 
that specify conditions on {\tt LO\_DISCOUNT} and/or {\tt LO\_QUANTITY}. 
For \hbox{example}, only partitions 1 and 2 are needed for the query 
``{\tt SELECT * FROM LINEORDER WHERE LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
Similarly, partitions 1 and 5 are sufficient to answer the query 
``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}.'' 
The respective sizes of the partitions are a factor of the data distribution. 
To see MLPPI in greater detail, check our references~\cite{sinclair:ppi,klindt09mlppi}.

Next, using the full version of {\tt LINEORDER} as defined in SSB~\cite{oneil:ssb} 
we provide examples of partitioning schemes that 
a DBA may define based on a small query set. 
Consider a query set $Q$ consisting of two queries $q$1 and $q$2 from the SSB set~\cite{oneil:ssb}, as shown below.

\begin{center}
{\small
\begin{tabular}{rl}
$q$1:	& {\tt SELECT SUM(l.LO$\_$EXTENDEDPRICE*l.LO\_DISCOUNT)} \\ 
		& {\tt FROM LINEORDER l, DDATE d} \\
		& {\tt WHERE l.LO\_ORDERDATE = d.D\_DATEKEY} \\
		& {\tt AND d.D\_YEAR = `1993'} \\
		& {\tt AND l.LO\_DISCOUNT IN (1, 4, 5)} \\
        & {\tt AND l.LO\_QUANTITY <= 30} \\ 
		&								   \\	        
$q$2:	& {\tt SELECT c.C\_NATION, SUM(l.LO\_REVENUE)} \\ 
		& {\tt FROM CUSTOMER c, LINEORDER l} \\
		& {\tt WHERE l.LO\_CUSTKEY = c.LO\_CUSTKEY} \\
		& {\tt AND c.C\_REGION=`EUROPE'} \\
		& {\tt AND l.LO\_DISCOUNT >= 7} \\
		& {\tt AND l.LO\_QUANTITY >= 25 AND l.LO\_QUANTITY <= 35} \\
		& {\tt GROUP BY c.C\_NATION} \\
		& {\tt ORDER BY revenue desc}. \\
\end{tabular}
}
\end{center}

Let's now focus on the {\tt LINEORDER} fact table and 
the predicates involving {\tt LINEORDER} fields only. 
There are five constraints identified by the query number 
and the sequence number of the constraint in each query, as illustrated below.
\begin{center}
\begin{tabular}{|c|l|}\hline 
$q$1.1 & {\tt LO\_DISCOUNT IN (1,4,5)} \\ \hline
$q$1.2 & {\tt LO\_QUANTITY <= 30} \\ \hline
$q$2.1 & {\tt LO\_DISCOUNT >= 7} \\ \hline
$q$2.2 & {\tt LO\_QUANTITY >= 25}\\ \hline
$q$2.3 & {\tt LO\_QUANTITY <= 35} \\ \hline 
\end{tabular}
\end{center}

At this point, the DBA needs to look at options 
based only on two fields and a total of five constraints. 
There are many possibilities from a {\em fine-grained} partition set 
to a loser definition.

Considering for the time being the column {\tt LO\_DISCOUNT}, 
one solution is to identify each value 
for the {\tt IN} predicate in $q$1.1 
and use $q$2.1 as is. This yields the following partitioning expression 
for {\tt LO\_DISCOUNT}:
\begin{center}
\begin{tabular}{l}
{\tt CASE\_N(} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 1,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 4,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT = 5,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT >= 7,} \\
\hspace{0.1in}{\tt NO CASE OR UNKNOWN} \\
{)}. \\
\end{tabular}
\end{center}
The above expression minimizes the size of the partitions 
by focusing exactly on the values required to satisfy the constraints, 
but creates a large number of small partitions.

Another possibility is to look at the maximum and minimum 
values in the {\tt IN} set and to build an {\tt AND} clause equivalent to 
a between clause yielding the following:
\begin{center}
\begin{tabular}{l}
{\tt CASE\_N(} \\
\hspace{0.1in}{\tt LO\_DISCOUNT >= 1 AND LO\_DISCOUNT <= 5,} \\
\hspace{0.1in}{\tt LO\_DISCOUNT >= 7,} \\
\hspace{0.1in}{\tt NO CASE OR UNKNOWN} \\
{)}. \\
\end{tabular}
\end{center}

\noindent The above partitioning solution decreases the number of partitions, 
compared to the previous solution but still focuses sharply on the constraints 
with still relatively small partitions.

Another possibility is to use the partitioning associated with {\tt LO\_DISCOUNT} shown 
in the beginning of this section with only two partitions.

\begin{figure*}[t!]
\centering
\includegraphics[scale=0.5]{architecture}
%\vspace{-0.2in}
\caption{The MLPPI Wizard Architecture\label{fig:arch}}
\end{figure*}

Similar considerations apply to the partitioning associated with {\tt LO\_QUANTITY} 
this time with ranges and covering problems difficult to deal with. 
The resulting partitioning scheme for the table includes both {\tt LO\_DISCOUNT}
and {\tt LO\_QUANTITY}, so that the number of possible combinations is the product of 
the potential combinations for each field. 
Also, there are two queries, and one partitioning scheme may be better for one query or the other. 
As a result, the DBA will be faced with a daunting combinatorial search problem 
and no clear basis to decide on which combination is the best. 
This state of affairs begs for a {\em tool} to assist the DBA.

In the sequel, we introduce an automated tool, called {\em MLPPI wizard}. 
Indeed, the \hbox{wizard} is the first physical database design tool 
developed at Teradata.
The tool is based on a novel technique using a greedy algorithm for 
search space enumeration. 
This tool based on a general framework allowing general expressions, 
ranges and case expressions for \hbox{partition} \hbox{definitions} 
is particularly well-suited for an MLPPI definition. \linebreak
The {\em predicate-driven} \hbox{technique} used by the tool can be applied 
to any clustering or partitioning based on simple fields and expressions or complex SQL predicates. 
The wizard also borrows the optimizer's cost model 
to prune the search space and reach a final solution.

Figure~\ref{fig:arch} illustrates how our tool recommends 
the final MLPPI customized for a given workload consisting of 
a set of queries and corresponding weights. 
The wizard passes through a series of phases: {\em Preprocessing}, 
{\em Initial}, and {\em Optimized} Phases. 
In the preprocessing phase the tool produces the most granular partitioning (MGP) based on the predicates collected from every query.
In the initial phase the wizard refines the passed MGP 
by merging a pair of partitions with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}) 
and then yields an initial (feasible) MLPPI based on the refined partitions.
In the \hbox{optimization} phase, the tool makes a further attempt to optimize the initial MLPPI. 
We provide in greater \hbox{detail} each phase, 
using the query set $Q$ consisting of $q$1 and $q2$ in the rest of the article.

\shorten{
An initial MLPPI is produced through the top two phases reflecting practical constraints by field or partition. 
It is further optimized in the last phase.
In the preprocessing phase, 
for each query its predicate(s) is extracted via an API call (on {\tt explain}) made to the server. 
Then, the wizard produces the most granular partitioning (MGP) based on the 
predicates collected from every query. 
If the MGP has too many fields over a field count limit, 
then the wizard \hbox{incrementally} throws away a field from the MGP until the limit.

In the initial phase, the passed MGP gets reduced if it includes more partitions than a partition count limit allowed by an MLPPI. 
The refinement can be \hbox{incrementally} done by merging a pair of partitions 
with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}), 
until the limit. 
After the phase, an initial (feasible) MLPPI is produced based on the refined partitions.

In the optimization phase, the tool makes a further attempt to optimize the initial MLPPI. 
Similar to the initial phase, more partitions can be reduced by merging ones with the 
least {\em query cost} (as will be described in Section~\ref{sec:opt_phase}). 
If the ongoing solution cannot be further improved, the tool makes the final MLPPI recommendation. 
We will provide more details about each phase later, 
using the query set $Q$ consisting of $q$1 and $q2$ 
in the rest of the article.}

To the best of our knowledge, there is no existing industrial tool 
to solve the {\it {\bf multi-level partitioning}} problem except our wizard. 
The wizard is contained completely on the client side, as opposed to 
previous work~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} 
requiring optimizer code extension. 
Our tool simply uses existing APIs to simplify the queries, capture fact table 
predicates and costs, and uses these items to make a recommendation. 
This makes the wizard extensible and portable to different releases 
of the Teradata DBMS server.

The article is a substantial extension of prior work~\cite{Suh12}, 
and organized in the following way. 
Section~\ref{sec:algorithms} presents the multi-level \hbox{partitioning} 
algorithms for the wizard. 
In turn, Section~\ref{sec:analysis} conducts an analysis of the complexity of 
the algorithms. 
Section~\ref{sec:experiment} reports the performance evaluation results. 
Section~\ref{sec:related_work} states how \hbox{Teradata} MLPPI wizard 
is different from some existing tools proposed by other vendors. 
Lastly, Section~\ref{sec:conclusion} summarizes our discussion.

\section{The MLPPI wizard}
\label{sec:algorithms}

In this section we describe in detail each phase of the wizard using the query set $Q$ provided in Section~\ref{sec:intro}.

\subsection{The Preprocessing Phase}
\label{sec:pre_phase}

Algorithm~\ref{algo:pre_phase} sketches 
the steps in the preprocessing phase. 
This phase consists of a total of six steps.

\vspace{-.1in}

\begin{algorithm}	
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$Q$ (input query set) }
\Output{$R$ (non-overlapping range set), $M$ (query-to-range-set map)}
\hspace{.05in} Query Simplification (on $Q$)\\
\hspace{.05in} Range Extraction (from $Q$) \\
\hspace{.05in} Non-Overlapping Range ($R$) Construction \\
\hspace{.05in} Field Count Limit Check (on $R$) \\
\hspace{.05in} Query-to-Range-Set Map ($M$) Construction \\
\hspace{.05in} Partition Count Limit Check (on $R$) \\
}
\caption{The Preprocessing Phase}
\label{algo:pre_phase}
\end{algorithm}

\vspace{-.3in}

\subsubsection{Query Simplification}

The first step of this phase is to simplify the predicate(s) of each query 
by removing redundant conditions. 
For instance, if a query predicate has the condition of 
``{\tt LO\_DISCOUNT IN (1, 4, 5) AND LO\_DISCOUNT IN (2, 3, 4, 5)}'', 
then the predicate can be simplified as ``{\tt LO\_DISCOUNT IN (4, 5)}.'' 
This can be done via an API call to the DBMS server. 
In the running example this simplification is not required.

\vspace{-.3in}

\subsubsection{Range Extraction}

The following step is to gather the simplified predicates and 
then obtain ranges from the collected predicates. 
This can also be done through an API call to the server. 
The query predicate is of the form  
\begin{center}
\begin{tabular}{l} 
\form{\tt variable} \form{\tt op} \form{\tt constant},
\end{tabular}
\end{center}
where \form{\tt variable} is a field from {\tt LINEORDER}, 
\form{\tt op} is in 
\begin{center}
\begin{tabular}{l} 
\{{\tt $ = $ }, {\tt $ < $}, {\tt $ <= $}, {\tt $ >= $}, {\tt $ > $}, 
{\tt IN}\},
\end{tabular}
\end{center}
and \form{\tt constant} represents a constant value(s). 
All \form{\tt op}s are self-explanatory. 
In particular, `{\tt IN}' is a list predicate implying `{\tt OR}' operator.
In the running example, 
we can collect a set of predicates $P$ = \{$p$1,$p$2,$p$3,$p$4\} from $Q$, where

\vspace{-.15in}

\begin{center}
\begin{tabular}{l} 
$p$1: {\tt l.LO\_DISCOUNT IN (1, 4, 5)} \\ 
$p$2: {\tt l.LO\_DISCOUNT >= 7} \\ 
$p$3: {\tt l.LO\_QUANTITY <= 30} \\ 
$p$4: {\tt l.LO\_QUANTITY >= 25} {\tt AND} {\tt l.LO\_QUANTITY <= 35}. \\ 
\end{tabular}
\end{center}

%\vspace{-.15in}

In turn, the wizard constructs a bi-directional {\em map} between a query and 
associated predicates, so that for each query the corresponding predicate(s) can be 
found directly in the map, and vice versa. 
In the example, a map, called $M$1, is built 
between $Q$ and $P$, and $M$1 is filled with the following entries:

$M$1:
\begin{center}
\begin{tabular}{ll} 
\form{q1, \{p1, p3\}} \\
\form{q2, \{p2, p4\}}.
\end{tabular}
\end{center}

Next, we gather fields referenced by predicates in $P$, extract 
ranges from the predicates, and group the ranges on each of the fields. 
Assume $I$ to be the list of fact table fields referenced by $P$. 
In the workload $Q$, we can see that there are two fields used by $P$: 
{\tt LO\_DISCOUNT}, {\tt LO\_QUANTITY}. 
These are added to $I$.

After that, we construct a set of distinct ranges on each field in $I$, using the predicates in $P$. 
Then, the whole range set becomes a kind of two-dimensional array, called $R$. 
The array $R$ captures the range representation of the predicates. 
Each entry under a field in $R$ represents a pair of (start and end) values forming the range. 
In the range representation, infinity ($\infty$) or -infinity (-$\infty$) can 
be used for unbounded ranges. 
``['' or ``]'' are used for closed ranges, and ``('' or ``)'' for open ranges. 
In particular, {\tt IN} predicate is represented by multiple ranges. 
For instance, the above $p$1 can be represented by the following three ranges 
\range{1, 1}, \range{4, 4}, \range{5, 5}; however, the last two consecutive 
ranges can be consolidated as \range{4, 5}. 
In the continuing example the range set $R$ is formed by the entries below.

$R$:
\begin{center}
\begin{tabular}{c|c} 
 {\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
\range{1, 1} & \lopenrange{\infty, 30}\\
\range{4, 5} & \range{25, 35} \\
\ropenrange{7, \infty} &  \\
\end{tabular}
\end{center}

Then, each predicate can be mapped to its associated ranges in $R$. That is, 
we can have another map, called $M$2, between $P$ and $R$. 
Let $R$[$i$, $j$] denote the interval value for the $i$-th field and $j$-th range in $R$. 
$R$[1, 3], for instance, indicates range \ropenrange{7, \infty} under {\tt LO\_DISCOUNT}. 
In the running example, $M$2 is constructed as follows:

$M$2:
\begin{center}
\begin{tabular}{l}
$\langle${$p$1}, \{$R$[1, 1],$R$[1, 2]\}$\rangle$ \\
$\langle${$p$2}, \{$R$[1 ,3]\}$\rangle$ \\ 
$\langle${$p$3}, \{$R$[2, 1]\}$\rangle$ \\
$\langle${$p$4}, \{$R$[2, 2]\}$\rangle$. \\
\end{tabular}
\end{center}

\subsubsection{Non-Overlapping Range Construction}

Since there could be multiple predicates on the same field across queries, 
extracted ranges may overlap each other. 
The overlapping ranges must be broken without any common portion, so that 
a pair of consecutive, \hbox{non-overlapping} ranges can be 
considered for a merge in further phases.

Algorithm~\ref{algo:split_overlap} describes the way of breaking 
the \hbox{overlap} of ranges.

\vspace{-.1in}

\begin{algorithm}[h]
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$R$ having overlapping ranges}
\Output{$R$ with consecutive, non-overlapping ranges}
	\ForEach{field $i$ $\in$ $R$}{	
		$L$ $\leftarrow$ Get the range set of $i$ from $R$ and sort by start value \\
		$j$ = 0 \;
		\While{$j$ $<$ $|L|$-1}{
			$r_{j}$, $r_{j+1}$ $\leftarrow$ Adjacent ranges in $L$ \\
			\If{$r_{j}$.$end$ $\geq$ $r_{j+1}$.$start$}{
				Make a split by modifying values of $r_{j}$ and $r_{j+1}$. \\
				Insert into $L$ an intermediate range if any. \\
				Re-sort ranges in $L$ and reset $j$ to 0. \\
			}\lElse{$j$++;}
		}
		Update $R$ with the final $L$. 
	} 
}
\caption{Splitting Overlapping Ranges}
\label{algo:split_overlap}
\end{algorithm}

\vspace{-.1in}

\noindent For each field, the tool gets its associated ranges and sorts them 
by start value. 
It then checks whether or not adjacent ranges overlap each other.
If that is the case, then we make a split between the ranges. 
In the example, the split is applied on the 
overlapping ranges, \lopenrange{\infty, 30} and \range{25, 35}, under {\tt LO\_QUANTITY}. 
Subsequently, a common, intermediate range (\range{25, 30}) is newly created and inserted into the range set 
belonging to the {\tt LO\_QUANTITY} field, in order to fill the gap. 
Then, $R$ in the example is shown as follows.

\
{\it $R$}:
\begin{center}
\begin{tabular}{c|c} 
 {\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
\range{1, 1}      & \openrange{\infty, 25} \\
\range{4, 5}      & \range{25, 30} \\
\ropenrange{7, \infty} & \range{31, 35} \\
\end{tabular}
\end{center}

The update of $R$ affects the existing map $M$2. 
In the running example, $p$3 and $p$4 influenced by the split get mapped to 
new range sets, \{$R$[2, 1], $R$[2, 2]\} and \{$R$[2, 2], $R$[2, 3]\}, respectively. 
Of course, the range sets mapped to $p$1 and $p$2 referencing {\tt LO\_DISCOUNT} remain unchanged. 
As a result, in the example we obtain the updated $M$2 as follows:

$M$2:
\begin{center}
\begin{tabular}{l}
$\langle${$p$1}, \{$R$[1, 1],$R$[1, 2]\}$\rangle$ \\
$\langle${$p$2}, \{$R$[1, 3]\}$\rangle$ \\ 
$\langle${$p$3}, \{$R$[2, 1], $R$[2, 2]\}$\rangle$ \\
$\langle${$p$4}, \{$R$[2, 2], $R$[2, 3]\}$\rangle$. \\
\end{tabular}
\end{center}

\vspace{.1in}

\subsubsection{Field Count Limit Check}

%The DBMS or user may impose a limit on the number of fields in the final 
%MLPPI recommendation. 
At present, the Teradata DBMS has a limit (64) of the fields 
that can be used in an MLPPI definition.
If the number of fields present in $R$ exceeds the field count limit, 
we determine which field(s) should be thrown away to satisfy the limit. 
(The star schema fact table we use has much fewer fields than the limit, 
and thus, this step will not be executed.)
To choose victim fields, the tool computes the weighted sum of 
{\em query cost} of queries regarding each field. 
The sum can be obtained by adding up every query cost on an MLPPI using 
only the ranges under the field. 
(We will cover how to measure the query cost in Section~\ref{sec:opt_phase}.) 
We incrementally discard a field with the highest query cost sum 
until the limit is reached. 
Accordingly, we can update the existing $M$2. 
One might say that the MLPPI may not be exploited if too many ranges, 
exceeding a partition count limit (or 65,536 as of now) are found in one field. 
But we assume that such an extreme case is not expected. Even if 
it happens, we can utilize the merges of consecutive ranges to make 
the MLPPI feasible.

\subsubsection{Query-to-Range-Set Map Construction}

Now, the tool can create a bi-directional {\em query-to-range-set} map, called $M$, 
between $Q$ and $R$, using the existing maps $M$1 and $M$2. 
It leverages a transitive property from $M$1 to $M$2.
Once $M$ is constructed, 
the intermediate maps ($M$1 and $M$2) are not used in the rest of phases. 
$M$ is shown as below:

{\it $M$}:
\begin{center}
\begin{tabular}{ll} 
$\langle${$q$1}, \{$R$[1, 1], $R$[1, 2], $R$[2, 1], $R$[2, 2]\}$\rangle$ \\
$\langle${$q$2}, \{$R$[1, 3], $R$[2, 2], $R$[2, 3]\}$\rangle$.
\end{tabular}
\end{center}

In the subsequent phases $M$ gets used for computing costs and updated 
along with a merge of ranges, and $R$ is refined for an MLPPI recommendation.

\subsubsection{Partition Count Limit Check}

In this regard, the range set $R$ can be used to define an MLPPI using each field as one level. 
However, if the current number of partitions by $R$ is greater than 
the partition count limit, 
then it is not possible to make a feasible MLPPI based on ranges in $R$.

The actual partition count limit is ample enough to pass the running example, 
but to continue our discussion, assume that in our discussion the limit is {\it 15}.
From $R$ in the running example, we can obtain a total of (number of ranges in {\tt LO\_DISCOUNT}) 
$\cdot$(number of ranges in {\tt LO\_QUANTITY}) = (3+1) $\cdot$ (3+1) = {\it 16}
partitions. 
Because the total partitions surpass the limit, 
the tool should pass through the initial phase in which 
fewer partitions than the limit are made. 
(Otherwise, the wizard will proceed to the optimized phase immediately.) 
Note that one more range is counted per field in the calculation. 
The added range is equivalent to the ``{\tt NO CASE OR UNKNOWN)}'' case in Figure~\ref{fig:exam}.

The tool is now ready to proceed to the next phases, 
with $R$ (input range set, or MGP) and the prepared $M$ (query-to-range-set map). 
%In the following subsections, we describe how the MLPPI solution for $Q$ 
%can be eventually made using $M$ and $R$.

\subsection{The Initial Phase}
\label{sec:init_phase}

In this phase, we incrementally merge a range pair in $R$ to reduce partitions. 
The merge continues until the number of ongoing partitions 
drop below the partition count limit. 
Once reaching the limit, we can have a feasible MLPPI with 
the remaining partitions.

By and large, overall I/O cost may be increased by the merge. 
Since every row is hash-sorted in each AMP of the Teradata DBMS server, 
typically the full scan on a partition is done to retrieve 
all the rows matching a given query. 
In the case of a merged partition, we may read the non-qualifying rows 
that would not be seen before the merge, thereby paying more I/O to answer 
the query.  
To minimize the merge overhead, we pick up the range pair incurring 
the least I/O increase in a heuristic fashion.

To choose the desirable range pair for a merge, 
the tool leverages the scan cost of a query on a range pair.
Algorithm~\ref{algo:comp_scan_cost} represents how 
the scan cost can be computed on the range pair that \hbox{influences} a query. 
The scan cost represents the I/O cost to answer the query when the range pair is merged. It is 
modeled as the {\em number of blocks} that are to be read 
for the target query. To compute the scan cost of a query $q$, we use 
a corresponding scan cost query ($s$), and 
it can be constructed as follows: 
\begin{center}
$s$: {\tt SELECT * FROM $F$ WHERE} {\it CP},
\end{center} 
where $F$ is a fact table, and 
{\it CP} indicates the predicates  
restored from the ranges mapped to $q$ from $M$.

\begin{algorithm}[t]
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$q$ (query), $rp$ (range pair), $M$ (query-to-range-set map)}
\Output{Scan cost for answering $q$ when considering a merge of $rp$}
	$r_{m}$ $\leftarrow$ Consolidate $rp$ \;
	$L$ $\leftarrow$ Copy the range set mapped to $q$ from $M$\; 
	Delete ranges in $rp$ from and insert $r_{m}$ into $L$. \\
	$s$ $\leftarrow$ ``{\tt SELECT * FROM FACT$\_$TABLE WHERE }'' \;
	\ForEach{range $r$ $\in$ $L$}{	
		Restore a predicate $p$ from $r$. \\
		Add $p$ to {\tt WHERE} clause of $s$.
	}
	Make an API call with $s$ to the server. \\
	Extract the spool size (in bytes) from the result. \\
	$blc$ $\leftarrow$ Calculate the block counts from the spool size. \\
	Update $q$'s scan cost to $blc$. \\
	{\bf return} $blc$ \;
}
\caption{Scan Cost Computation}
\label{algo:comp_scan_cost}
\end{algorithm}

If $q$ turns out to be affected by the merge of a range pair, 
then the range set associated with $q$ in $M$ is temporarily updated 
by removing the parent ranges and adding the merged one. 
The altered range set is remapped to $q$ in $M$ and then 
translated to the equivalent predicates, 
so that {\it CP} for $s$ based on the predicates can be built.
Unless the merge range pair influences $q$, then 
{\it CP} can be simply built based on the 
predicates restored from the existing ranges mapped to $q$.

To help understand the scan cost construction, let us 
consider a range pair ($rp$) of \hbox{$R$[2, 2]} and $R$[2, 3] in the running example. 
$rp$ produces the merged range, \range{25, 35}, \hbox{under} {\tt LO\_QUANTITY}. 
The merge by $rp$ affects both $q$1 and $q$2 in the workload.

Thus, the range sets of $q$1 and $q$2 on {\tt LO\_QUANTITY} are 
altered to \{\openrange{\infty, 25}, \range{25, 35}\} and \{\range{25, 35}\}, 
respectively. Of course, no change is made to the existing range sets 
of the queries on {\tt LO\_DISCOUNT}. 
Therefore, the corresponding scan cost queries for $q$1 and $q$2 
can be built as below:
\begin{center}
\begin{tabular}{rl}
$\Sone$:	& {\tt SELECT * FROM LINEORDER l} \\
		& {\tt WHERE ((l.LO\_DISCOUNT = 1)} \\
		& {\tt OR (l.LO\_DISCOUNT >= 4} \\ 
		& {\tt AND l.LO\_DISCOUNT <= 5))} \\
		& {\tt AND l.LO\_QUANTITY <= 35}  \\ 
$\Stwo$:	& {\tt SELECT * FROM LINEORDER l} \\
        & {\tt WHERE l.LO\_DISCOUNT >= 7} \\
        & {\tt AND (l.LO\_QUANTITY >= 25 AND} \\ 
        & {\tt l.LO\_QUANTITY <= 35)}.
\end{tabular}
\end{center}

It is possible to combine all the (consecutive) ranges in the (temporarily) altered set of $q$1, 
and hence, a single predicate ({\tt l.LO\_QUANTITY <= 35}) can be built.

The wizard sends such a scan cost query to the server via an API call, 
and it extracts from the server's response the {\em spool size} (in bytes), or the size of scan query result. 
Then, the tool computes as scan cost the {\em block counts} using the spool size. 
If the merge of a range pair does not impact on any query in the workload, 
the existing (previously computed) scan cost of queries 
will be reused for the range pair. 
That is, the wizard will only recompute 
the scan cost of a query {\em iff} the query is affected by 
the merge of two consecutive ranges.

In this way, the weighted sum of scan cost of queries, $T_{s}$ can be computed 
for each range pair. 
$T_{s}$ can be defined as follows:
\begin{center}
\begin{tabular}{c} 
$T_{s}$ = $\sum_{i=1}^{n}$($sc_{i}${$\cdot$}{$w_{i}$}),
\end{tabular}
\end{center}
where $n$ is the number of queries in $Q$, $sc_{i}$ is the scan cost of a query $q_{i}$ in $Q$, and 
$w_{i}$ denotes the (\hbox{non-negative}) weight associated with $q_{i}$.

Once all range pairs are examined, the tool chooses the range pair 
with the least $T_{s}$ for a merge. 
If several range pairs end up with the same least $T_{s}$, 
then the tool applies the heuristic of favoring the range pair that produces 
fewer number of partitions when merged, to make 
it faster to reach the partition count limit.

In the example, suppose that the running range pair $rp$ produces the least $T_{s}$, 
and thus, the tool merges $rp$. 
Thus we can update both $R$ and $M$ as follows.

$R$:
\begin{center}
\begin{tabular}{c|c} 
{\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
\range{1, 1} & \openrange{\infty, 25} \\
\range{4, 5} & \range{25, 35}		  \\
\ropenrange{7, \infty} &  \\
\end{tabular}
\end{center}

$M$:
\begin{center}
\begin{tabular}{ll} 
$\langle${$q$1}, \{$R$[1, 1], $R$[1, 2], $R$[2, 1], $R$[2, 2]\}$\rangle$ \\
$\langle${$q$2}, \{$R$[1, 3], $R$[2, 2]\}$\rangle$
\end{tabular}
\end{center}

Note that we see that $q$2 may have the most customized partition based on the updated $M$. 
Only the single partition formed by ($R$[1, 3]$\cap${$R$[2, 2]}) is sufficient 
to retrieve all the qualifying 
rows for $q$2; thus, the other partitions can be simply eliminated. 
In the meantime, to answer $q$1, we need to read the four partitions formed by 
the top two ranges of each field. 
Unfortunately, because the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
contain the non-qualifying rows for $q$1, $R$ cannot provide $q$1 with as much benefit as $q$2. 
However, $R$ can be a good compromise to satisfy both queries, in that 
the MLPPI derived by $R$ can potentially minimize the total execution cost 
of $Q$.

\begin{algorithm}[t]
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$M$ (query-to-range-set map), $R$ (input range set)}
\Output{$R$ with \# partitions $\leq$ partition limit}
	$W$ $\leftarrow$ Query weights \\
	\While{(\# partitions by $R$ $>$ partition limit)}{
		\ForEach{a range pair ($rp$) in $R$}{	 
			$T_{s}$ $\leftarrow$ 0 \;
			\ForEach{$q$ in $M$}{	
				$w$ $\leftarrow$ Get $q$'s weight from $W$ \\
				$L$ $\leftarrow$ Get the range set mapped to $q$ in $M$ \\
				\uIf(\tcp*[f]{Affected}) {(($rp$ $\cap$ $L$) $\neq$ $\varnothing$)}{
					$T_{s}$ += $w{\cdot}$({\it getSC}($q$,$rp$,$M$)) \tcp*[r]{See Algo.~\ref{algo:comp_scan_cost} regarding {\it getSC}()}
				}
				%\lElse{ $T_{s}$ += $w{\cdot}$($q$'s existing scan cost)}
			}
			Associate $T_{s}$ with $rp$.
		}
		Find $rp$(s) with the least $T_{s}$. \\
		If a tie happens, select the $rp$ to make fewer partitions when consolidated. \\
		Update $M$ and $R$ by the chosen $rp$.
	}
	{\bf return} $R$\;
}
\caption{The Initial Phase}
\label{algo:init_phase}
\end{algorithm}

Algorithm~\ref{algo:init_phase} represents the initial phase 
that has been proposed so far. 
The algorithm starts with a query-to-range set map ($M$) and an input range set ($R$). 
Until the number of the partitions in $R$ falls below the partition limit threshold, 
the algorithm determines the range pair yielding the least total execution cost ($T_{s}$) 
and updates $M$ and $R$. At the end, the algorithm produces a final recommendation based on $R$.
%\vspace{-.1in}

Now that the total partition counts ($4${$\times$}$3$=$12$) by $R$ falls  
below the assumed limit ($15$), the wizard 
proceeds to the optimized phase with $R$ without repeating the initial phase.

\subsection{The Optimized Phase}
\label{sec:opt_phase}

%We verified the effectiveness of the tool by running it on a workload based
In this regard, using $R$ we have an initial MLPPI recommendation for the {\tt LINEORDER} fact table.
The recommendation can be used sufficiently, but 
having fewer partitions, induced by further merges, 
can enhance the overall workload performance for the following reasons. 
First, multiple file contexts by many partitions can incur 
huge overhead that impacts on the query optimizer. 
There also exists an operational threshold 
that the optimizer can handle the maximum number of the partitions at a time.
Therefore, further reducing partitions greatly helps the optimizer to manage these partitions.

A similar algorithm, as suggested in the initial phase, 
can be applied to this optimized phase. 
Instead of the scan cost, the {\em query cost} can be used thanks to a feasible MLPPI. 
The query cost is modeled as the estimated processing time 
of a query on a ``faked'' fact table applying the MLPPI definition 
reflecting the merge of a range pair. 
This cost estimation technique is similar 
to the ``\hbox{what-if}" approach~\cite{chaudhuri1998autoadmin}.

\begin{algorithm}[t]
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$q$ (query), $rp$ (range pair), $M$ (query-to-range-set map), $R$ (input range set)}
\Output{Query cost to answer $q$ when considering a merge of $rp$}
	Create an empty shadow table $H$ having the same definition as the fact table $F$, 
	including indexes and all constraints (check and referential integrity constraints). \\
	Propagate all (field and index) statistics of $F$ to $H$. \\
	\If{$F$ has materialized views (MVs)}{
		Create the equivalent MVs on $H$. \\
		Propagate the statistics of the MVs on $F$ to those of $H$. \\
	}
	$r_{m}$ $\leftarrow$ Merge $rp$ \;
	$L$ $\leftarrow$ Copy ranges associated with $q$ from $M$\; 
	Remove ranges in $rp$ from and add $r_{m}$ to $L$. \\
	$R'$ $\leftarrow$ Update $R$ with $L$ \\
	Alter $H$ using a fictitious MLPPI with $R'$. \\
	Construct and send to the server an $H$-based shadow query $h$ equivalent to $q$. \\	
	$ept$ $\leftarrow$ Estimated processing time of $h$ extracted from the server response \;
	Update $q$'s query cost to $ept$. \\
	{\bf return $ept$} 
}
\caption{Query Cost Computation}
\label{algo:comp_query_cost}
\end{algorithm}

Algorithm~\ref{algo:comp_query_cost} illustrates the steps for computing 
the query cost of a query $q$, 
given a range pair ($rp$) affecting $q$ when merged. 
First, we create an empty shadow table $H$ with the same definition as $F$, 
including all indexes and constraints such as check and referential integrity 
ones. Next, we propagate all statistics of $F$ to $H$. 
This covers field and index statistics. 
If $F$ has materialized views (MVs), then we create the equivalent MVs on $H$ and subsequently, 
propagate the statistics of the MVs on $F$ to the new MVs on $H$. 
After that, we alter $H$ by an MLPPI with $R$' applying the merge of $rp$. 
The wizard then builds a {\em shadow} query $h$ equivalent to the input query $q$ 
but replacing $F$ in $q$ with $H$, and makes an API call to the {\tt EXPLAIN} tool with $h$. 
Eventually, the tool extracts from the result and returns the estimated elapsed time 
of $h$ as the query cost of $q$. 
As done in the initial phase, the query cost of a query is recomputed only if the merge affects the query.

For each range pair the wizard computes the weighted sum ($T_{q}$) of query cost of queries. 
$T_{q}$ can be similarly defined as $T_{s}$, described in Section~\ref{sec:init_phase}.
%and thus the definition of $T_{q}$ is omitted.
Let the current least $T_{q}$ be $T$ and the existing least $T_{q}$ be $T_{p}$. 
If $T$ $<=$ $T_{p}$, then for the next iteration $T_{p}$ is updated to $T$. 
In turn, the tool merges the range pair that produces $T$ and updates $M$ and $R$ along with the merge. 
The wizard repeats the process until 1) no ranges in $R$ remain, or 2) pre-defined iterations are reached. 
The optimized phase is finished if $T$ $\geq$ $T_{p}$.

In the running example, suppose that in the first round, ranges $R$[1, 1] and $R$[1, 2] are chosen for a merge. 
Then, the wizard produces a merged range \range{1, 5} \hbox{under} {\tt LO\_DISCOUNT} and proceeds to the next round. 
If a range pair selected for the subsequent merge fails to improve $T_{p}$, 
then, the tool exits the optimized phase and \hbox{produces} the final MLPPI recommendation for the given workload $Q$ 
as shown in Figure~\ref{fig:mlppi}.

\vspace{-.1in}

\begin{figure}[htp!]
\begin{center}
\begin{tabular}{|l|} \hline
{\tt ALTER TABLE LINEORDER} \\
{\tt MODIFY PRIMARY INDEX} \\
{\tt PARTITION BY(} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 1 AND LO\_DISCOUNT $\leq$ 5,} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 7,} \\ 
	\hspace{.2in}{\tt NO CASE OR UNKNOWN),} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_QUANTITY $<$ 25,} \\
	\hspace{.2in}{\tt LO\_QUANTITY $\geq$ 25 AND LO\_QUANTITY $\leq$ 35,} \\
	\hspace{.2in}{\tt NO CASE OR UNKNOWN)} \\ 
{\tt );} \\ \hline
\end{tabular}
\end{center}
\caption{The MLPPI Recommendation for Workload $Q$\label{fig:mlppi}}
\end{figure}

%\vspace{-.1in}

\begin{algorithm}[t]
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$M$ (query-to-range-set map), $R$ (input range set)}
\Output{MLPPI such that the total query cost is minimal}
	$W$ $\leftarrow$ Query weights \\
	$T_{p}$ $\leftarrow$ Total query cost sum on an initial MLPPI by $R$ \\
	\While{(Pre-defined iterations or $R$ $\neq$ $\varnothing$)}{
		\ForEach{range pair $rp$ in $R$}{	 
			$T_{q}$ $\leftarrow$ 0 \;
			\ForEach{$q$ in $M$}{	
				$w$ $\leftarrow$ Get $q$'s weight from $W$ \\
				$L$ $\leftarrow$ Get the range set mapped to $q$ in $M$ \\
				\uIf(\tcp*[f]{Affected}) 
				{(($rp$ $\cap$ $L$) $\neq$ $\varnothing$)}{
					$T_{q}$ += $w${$\cdot$}({\it getQC}($q$,$rp$,$M$,$R$)) 	
					\tcp*[r]{See Algo.~\ref{algo:comp_query_cost} regarding {\it getQC}()}				
				}
				\lElse{ $T_{q}$ += $w${$\cdot$}($q$'s existing query cost)}
			}
			Map $T_{q}$ to $rp$.
		}
		$T$ $\leftarrow$ The least $T_{q}$ that has been so far seen.  \\
		\lIf{$T$ $>$ $T_{p}$}{ {\bf break} \;}
		\Else{
			$T_{p}$ $\leftarrow$ $T$ \tcp*[r]{Query Cost Sum Update} 
			Find $rp$(s) with $T$. \\
			If a tie happens, select the $rp$ to make fewer partitions when consolidated. \\
			Update $M$ and $R$ by the chosen $rp$. \\
		}
	}
	{\bf return} an MLPPI by $R$ \;
}
\caption{The Optimized Phase}
\label{algo:opt_phase}
\end{algorithm}

\vspace{-.2in}

Algorithm~\ref{algo:opt_phase} exhibits the proposed optimized phase, 
which also applies the same heuristic to favor a range pair 
that produces fewer partitions to break a tie.  
The order of the different partitioning levels may have some impact on 
the execution time. A range condition on lower levels like the query 
in Section~\ref{sec:intro} 
``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}'' 
requires scanning non-consecutive partitions. 
This may \hbox{incur} some overhead at run time. 
To reduce this \hbox{effect}, we sort the partition \hbox{levels} based on number of 
partitions in descending order \hbox{before} making a final \hbox{recommendation}. 
The solution in \hbox{Figure}~\ref{fig:exam} follows this heuristic 
making the two-partition case in the first level 
followed by the four-partition case in the second level. 
The order in Figure~\ref{fig:mlppi} can go \hbox{either} way 
since both levels have the same number of \hbox{partitions}.

\section{Analysis}
\label{sec:analysis}

In this section we analyze the time complexity of 
the proposed preprocessing, initial, and optimized phases.
For the time complexity we compute the logical running time of the wizard, using the {\em number of API calls} 
made to the server during the phases.

%In this section the complexity of the proposed algorithms 
%is analyzed.

\subsection{The Preprocessing Phase}

\begin{lemma}
\label{lemma:pre_phase}  
The running time complexity for building a query-to-range map is 
$\BigO$($N${$\cdot$}$C${$\cdot$}$V$), where $N$ is the number of queries 
in a workload, $C$ is the total number of fields in the fact table, 
and $V$ is the max number of values in a field. 
\end{lemma}

{\bf Proof}. Let us consider a worst case such that given a workload of 
$N$ queries, each query references all the fields of the fact table, and 
each field is associated with at most $V$ values by the predicates of the queries. 
Each query may have a single-value range. 
A range consisting of only a single value per field can be mapped to each query. 
Thus, the running time complexity for the map construction 
is $\BigO$($N${$\cdot$}$C${$\cdot$}$V$). \qed

{\bf Comment:} In practice, our experiments showed that the bound 
$\BigO$($N${$\cdot$}$C${$\cdot$}$V$) is overly pessimistic.

\subsection{The Initial and Optimized Phases}

\begin{lemma}
\label{lemma:init_phase}  
The running time complexity for the initial or optimized phases is 
$\BigO$($M^{2}${$\cdot$}$N$), 
where $M$ is the number of range pairs, and $N$ is the number of queries in 
a given workload. 
\end{lemma}

{\bf Proof}. Suppose that the merge of every range pair influences all the queries. 
Every iteration, either the scan or query costs need to be recomputed 
for each query. 
In the first round the \hbox{re-computation} cost is paid as many times as $M${$\cdot$}$N$, 
and a chosen range pair is merged. In the subsequent round 
$M$-1 range pairs remains, and the following cost amounts to ($M$-1){$\cdot$}$N$. 
In the worst case we may end up consolidating all range pairs, thus having no partition on 
a target table. The total calls to the server can be increased up to

\begin{center}
$M${$\cdot$}$N$+($M$-1){$\cdot$}$N$+$\cdots$+$N$ = ($\sum_{i=1}^M i$){$\cdot$}$N$ = $\frac{M(M-1)}{2}${$\cdot$}$N$. 
\end{center}

\noindent Hence, the running time complexity is $\BigO$($M^{2}${$\cdot$}$N$). $\qed$

%\vspace{0.1in}

Our experiments showed that $\BigO$($M^{2}${$\cdot$}$N$), the running time complexity, 
is overly \hbox{pessimistic}. 
In \hbox{practice}, the number of calls made was much fewer than the \hbox{theoretic} bound, 
since it was very rare for a range pair to be associated with all queries in a workload. 
In \hbox{Section}~\ref{sec:stat} we will show how many API calls take place on the workloads 
used for evaluation.

\section{Experiment}
\label{sec:experiment}

In this section, we describe our environment settings, 
report the statistics measured in our experiments, and 
demonstrate the performance of the MLPPI wizard (WIZARD), compared with 
that of no partitioning and the partitioning by a human expert (EXP).

\subsection{Environment Settings}
\label{sec:env_set}

The MLPPI wizard was implemented as a prototype 
on top of the Teradata DBMS server. 
It was written in Java. 
The performance of the wizard was evaluated 
on the Teradata DBMS server machine running Unix.

When it comes to workload generation, we took advantage of 
our simple star schema query \hbox{generator}. 
The generator assumes that 1) available operators, 2) fields, 
and 3) the minimum and \hbox{maximum} values of the fields are already known. 
For each query, the generator first randomly selects the number of single-table 
predicates to create. 
Then, it arbitrarily chooses a field and a specific operator 
for each predicate. 
If {\tt IN} operator on a chosen field is picked up, the query generator 
determines the number of values to add to the {\tt IN} list and then chooses random values within the min-max value 
range of the field. 
In this way the generator builds a query involving the generated predicates.

A generated query is based on a template that joins {\tt LINEORDER} and {\tt DDATE} with constraints defined by the predicates. 
This template is common in customer cases like reports and form 
templates. 
Using the query generator we generated two workloads consisting of 10 queries (10Q) and 20 queries (20Q).

For the query workloads we populated the fact table with a scale of 1TByte (1TB) and 3TBytes (3TB).

\begin{table*}[t]
\caption{The statistics measured on the used workloads\label{tab:workload_stat}}
\begin{center}
{\scriptsize
\begin{tabular}{|l|l|l|l|l|} \hline
		   			   							   & \multicolumn{2}{|c|}{Init. Phase} & \multicolumn{2}{|c|}{Opt. Phase} \\ \cline{2-5}
		   			   							   & 10Q & 20Q & 10Q & 20Q \\ \hline
{\# Input Partitions}  							   & 1,008,000 & 110,739,200 & 51,840 & 59,520 \\ \hline
{\# Input Range Pairs} 							   & 441 & 4,180 & 24 & 48\\ \hline
{\# Total Iterations}  							   & 14 & 55 & 1 & 1 \\ \hline
{\# Total API Calls (Worst Calls by Lemma~\ref{lemma:init_phase}})   & 1,305 (1,944,810) & 31,915 (349,448,000) & 78 (5,760) & 388 (46,080) \\ \hline
{\# Average API calls per Range Pair} 		  	   & 2.96 & 7.63 & 3.25 & 8.08 \\ \hline
%{\# avg. range pairs compared per iteration} 	   & 31.5 & 76.0 & 24 & 48 \\ \hline
%{\# max. range pairs compared within an iteration} & 38 & 103 & 24 & 48 \\ \hline
%{\# min. range pairs compared within an iteration} & 25 & 49 & N/A & N/A\\ \hline
%{\# max. API Calls made per range pair} 		   & 4.92 & 16.14 & & \\ \hline
%{\# min. API Calls made per range pair} 		   & 1.92 & & & \\ \hline
%{\# avg. API calls made per iteration}		 	   & 93.21 & 580 & 78 & 388 \\ \hline
%{\# max. API calls made within an iteration} 	   & 123 & 791 & 78 & 388 \\ \hline
%{\# min. API calls made within an iteration} 	   & 73 & 381 & N/A & N/A \\ \hline
\end{tabular}
}
\end{center}
\end{table*}

\subsection{Statistics Report}
\label{sec:stat}

Table~\ref{tab:workload_stat} summarizes the WIZARD's execution statistics, 
which were obtained while the recommendations for the \hbox{workloads} were produced. 
The statistics includes input partitions, 
range pairs, number of API calls made, and total iterations 
observed in each phase.

About 1 million partitions were initially derived for 10Q whereas roughly 0.11 billion partitions for 20Q. 
Although the workload size doubled, the total number of partitions was increased exponentially. 
Much more predicates with different bindings produced two orders of magnitude more ranges in 20Q than those of 10Q. 
While the huge input partitions were made, the number of collected range pairs per field was relatively small.

The total number of iterations in the initial phases 
tended to be proportional to the input \hbox{partitions} 
\hbox{generated} from the workloads. 
Note that there \hbox{happened} only one iteration in the optimized phase. 
The reason was that since the partitions produced by the \hbox{initial} phase were customized enough, 
no more \hbox{iterations} were \hbox{necessary} for further optimization.

As mentioned in Section~\ref{sec:analysis}, the total calls made to 
the server per workload were by far fewer than the theoretical bound shown in parentheses. 
The worst call counts were calculated regarding the number of queries and range pairs observed in each phase. 
The average number of calls per range pair was limited within 10. %across workloads and phases.

\begin{figure*}[t]
\vspace{-.2in}
\centering
\subfigure[1TB]{
	\includegraphics[scale=0.6]{1T_total_time}
	\label{fig:1T_eval_res}
}
\subfigure[3TB]{
	\includegraphics[scale=0.6]{3T_total_time}
	\label{fig:3T_eval_res}
}
\vspace{-0.1in}
\caption{Total Elapsed Time\label{fig:total_elapsed_time}}
\end{figure*}

\begin{figure*}[t]
\vspace{-.2in}
\centering
\subfigure[1TB]{
	\includegraphics[scale=0.6]{1T_quality}
	\label{fig:1T_quality}
}
\subfigure[3TB]{
	\includegraphics[scale=0.6]{3T_quality}
	\label{fig:3T_quality}
}
\vspace{-0.1in}
\caption{Recommendation Quality Comparison\label{fig:recomm_qual}}
\vspace{-.1in}
\end{figure*}

\subsection{Performance Evaluation}

The main focus of our demonstration is to see the performance (quality) 
of MLPPI \hbox{recommendations} made by the MLPPI wizard (WIZARD), 
compared with the performance of no partitioning (NO PPI) and EXP (partitioning 
by a human expert).

%Our evalution results are exhibited in Figures ~\ref{fig:eval_res} 
%and ~\ref{fig:improvement}.
%Figure~\ref{fig:eval_res} shows the times (in seconds) taken 
%to run the 10Q or 20Q workloads over the 1TB or 3TB fact tables. 
%Based on the run results, Figure~\ref{fig:improvement} displays 
%how much improvement could be achieved by the EXP or the WIZARD 
%recommendations against the NO PPI one.

Figures~\ref{fig:total_elapsed_time} and \ref{fig:recomm_qual} exhibit our evaluation results.
Figure~\ref{fig:total_elapsed_time} shows the times (in seconds) taken 
to run the 10Q and 20Q workloads over the 1TB and 3TB fact tables. 
Based on the run results, Figure~\ref{fig:recomm_qual} shows 
how much improvement was gained by the EXP and WIZARD solutions, compared with the NO PPI solution.

Overall, the WIZARD recommendations were very effective at partitioning 
the fact tables by leveraging the predicates in the workloads. 
On average, the \hbox{total} execution time of the workloads on the WIZARD solutions was 
about 3.5 and 2 times faster than those of NO PPI and EXP, respectively. 
Also, the quality of the \hbox{WIZARD} solutions outperformed that of the EXP solutions; 
the WIZARD solutions yielded an average of about 77\% improvement on the NO PPI solutions.

Furthermore, the WIZARD recommendations 
scaled very well with growing workload size and table size, 
as illustrated in Figure~\ref{fig:recomm_qual}. 
Although we doubled the workload scale from 10Q to 20Q, 
the performance improvement of the WIZARD solution stayed the same in Figure~\ref{fig:1T_quality} 
and got almost negligibly decreased in Figure~\ref{fig:3T_quality}. 
This result showed that our WIZARD \hbox{solution} was scalable over increasing workload size.

We also increased the table scale from 1TB to 3TB. 
For 10Q the performance improvement (about 77\%) was almost the same over the increasing scales, 
and for 20Q it was decreased from 76\% to 74\%, which was \hbox{almost} negligible.
The WIZARD partitioning \hbox{recommendation} scaled well with increasing table size.

In sum, our experiment results attest the effectiveness of the proposed algorithms for WIZARD 
and show the superiority of the recommendations of WIZARD.

\shorten{Regarding the scale of 1TB in Figure~\ref{fig:1T_quality}, 
although we doubled the workload scale (from 10Q to 20Q) 
the performance improvement of the WIZARD solution compared with that of NO PPI was not decrease although we doubled the workload scale (from 10Q to 20Q).
In Figure~\ref{fig:3T_quality}, for the 3TB scale the improvement of WIZARD was slightly decreased, but it was negligible.
In sum, the performance of the WIZARD solution was scalable over increasing workload and table scale.

%Also, we scaled up the table from 1TB to 3TB on the fixed workload, but 
%the \hbox{recommendation} quality was not degraded at all. 
These results attest the effectiveness of our algorithms, 
and show the superiority of our recommendations.}

\section{Related Work}
\label{sec:related_work}

Physical database design~\cite{finkelstein1988physical,labio1997physical,rozen1991framework,Zilio1998} 
has been discussed in \hbox{academic} research and industrial sectors in the past years.
The major DBMS vendors (e.g., IBM, MS, and Oracle) have driven much of the work. 
Their specific interests have been in automating the physical design for table 
partitioning~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated,sheet2009oracle,saa2007oracle,rao2002automating}, indexes/materialized views~\cite{agrawal2000automated,Agrawal04:sqlserver,dash2011cophy,Kimura11,itwsqlserver,valentin2000db2,zilio2004recommending}, 
and integration~\cite{Zilio04:db2design}.

%Lightstone's work~\cite{Lightstone04:db2auto} introduces the MDC (Multidimensional 
%Clustering) Advisor that automates the selection of MDC 
%keys customized for a specific workload on DB2 Universal Databases to use MDC structure. 
%DB2 Advisor~\cite{Zilio04:db2design} is an integrated commercial 
%tool that captures a variety of aspects of physical database design 
%by automation in a comprehensive way. 
%Oracle Partitioning Advisor~\cite{sheet2009oracle} for recommending 
%table partitionings is  integrated into 
%the SQL Access Advisor~\cite{saa2007oracle}.

%Since data clustering significantly influences on the performance of 
%multidimensional queries, the MDC advisor helps DBAs that may have 
%nontrivial troubles in finding proper clustering dimensions and 
%granularity. 
%To enumerate and determine possible clustering dimensions, 
%his work relies on {\it what-if} query cost modeling, data sampling and a search algorithm.

%DB2 Advisor~\cite{Zilio04:db2design} is an integrated commercial 
%tool that captures a variety of aspects of physical database design 
%by automation in a comprehensive way. 
%It makes an attempt to cover a variety of design choices-partitioning, 
%MDC, materialized views and indexes. 
%DB2 Advisor leverages the DB2 query optimizer for yielding a recommendation, and 
%takes a hybrid searching approach for considering different features.

%Oracle has a tool, called Partitioning Advisor~\cite{sheet2009oracle}, 
%to make a partitioning recommendation for tables. 
%The tool is now integrated into 
%the SQL Access Advisor~\cite{saa2007oracle} that recommends indexes and materialized 
%views. 
%However, its technical details are not yet released in public. 
%
%Also, MS SQL Server provides some similar, automated tools~\cite{agrawal2000automated,agrawal04:integrating,
%itwsqlserver} for table partitioning, materialized views or indexes fitting to a given workload.
%The optimizer cost model is exploited to acquire estimates for 
%candidate materialized views or indexes. 
%In particular, Agrawal's work~\cite{agrawal04:integrating} presents 
%an approach of integrating vertical and horizontal partitioning schemes 
%into their physical database design, considering scalability.

Some of the tools in IBM DB2~\cite{Lightstone04:db2auto}, 
Oracle~\cite{sheet2009oracle}, and MS SQL Server~\cite{agrawal04:integrating} 
appear to be similar to our MLPPI wizard. 
However, in light of problem scope and approach, the wizard 
is fundamentally different from the existing tools excluding Oracle Partitioning \hbox{Advisor}~\cite{sheet2009oracle} 
providing no published technical details.

DB2 MDC Advisor~\cite{Lightstone04:db2auto} actually tackles 
a different problem of automatically recommending the most well-suited MDC 
keys for a given workload. 
Agrawal's work~\cite{agrawal04:integrating} 
discusses another problem of merging \hbox{single-level} range partitionings on objects such as tables and indexes. 
This article addresses the multi-level \hbox{partitioning} problem. 
Hence, the existing solutions cannot be directly applied to our MLPPI wizard.

Regarding the approach, DB2 MDC Advisor~\cite{Lightstone04:db2auto} uses 
the search space driven by {\em fields}. In contrast, our search 
space is driven by query-predicates, which is superior because 
\hbox{predicates} are more customized and specific to a workload than fields. 
Agrawal's horizontal partitioning scheme~\cite{agrawal04:integrating} 
also uses the search space driven by simple range predicates, but 
his technique has a \hbox{shortcoming} as follows. 
His work produces a solution for each individual query and attempt 
to merge the solutions. But this approach cannot reach 
an optimized solution in a global \hbox{perspective}.
Our wizard generates the whole search space upfront and in turn merges partitions, leading to a globally-optimized \hbox{solution}. 
Moreover, only a single column is considered in his work~\cite{agrawal04:integrating}, 
whereas our tool deals with multiple fields.

Implementations of previous tools~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} \hbox{required} instrumentation for optimizer code.
These instrumentations are needed to facilitate the required 
information for the physical design tools API calls. 
The \hbox{instrumentation} code need to be enhanced and tested for new database 
releases that add complexity and additional cost for software upgrades. 
The Teradata optimizer has a rich set of existing APIs originally coded 
for system and workload management tools. 
The APIs are sufficient to avoid the costly optimizer code change.

Also, Nehme's work~\cite{nehme2011automated}, deeply-integrated with 
optimizer, reveals a concern about the quality of the recommendations 
made by some tools, shallowly integrated with optimizer. 
But we observed in our experiments that the quality of 
the wizard's solutions was much superior to that of the base solutions. 
In addition, some might argue that in our loosely-coupled approach 
the cost to invoke the optimizer might be significant. 
But the measured call counts were much fewer than 
the theoretical bound, since most calls were made only when queries were affected by a merge.  
%If the remote communication cost is expensive, the tool can also be run, 
%{\em co-located} with the server.

%The existing work requires their database engine to be extended 
%to support their features. 
%This strong coupling may impose a nontrivial burden on their servers whenever 
%a new feature of the existing tools is desired. 
%As illustrated in Figure~\ref{fig:arch}, our tool is completely outside 
%the DBMS server. Thus, we are free of the constraint. 
%A literature~\cite{nehme2011automated} claims that 
%such loosely-coupled approaches might suffer from the reduced performance of a tool 
%and the poor quality of the partitioning recommendations. 
%The powerful Teradata APIs, however, help the wizard receive from the server 
%everything needed for yielding a MLPPI recommendation. 
%Moreover, as demonstrated in Figure~\ref{fig:perf_eval}, the quality of the 
%partitioning solutions by the wizard was also much superior to that of the base solutions.
%Some might argue that the cost of invoking the optimizer could be significant as well. 
%Again, the actual calls are mainly made only when queries are affected by a merge. 
%As a result, they were much fewer than 
%the theoretical bound as shown in Table~\ref{tab:workload_stat}.  
%To further save the remote communication cost, the tool can also be run, 
%{\em co-located} with the server. 
%Therefore, we believe that the shallowly-integrated strategy perfectly matches 
%with the wizard.

Lastly, there has been some previous work~\cite{nehme2011automated,rao2002automating,tatarowicz2012lookup} 
regarding table partitioning in multi-node \hbox{systems}, but 
our problem is discussed in the context of a single node system, as in the existing work~\cite{agrawal04:integrating}. 
Database cracking~\cite{idreoskm07}
%, self-reorganizing column values in accordance with query workload, 
assumes a single node environment, but 
it does not address our multi-level partitioning problem.

\section{Summary}
\label{sec:conclusion}
%We verified the effectiveness of the tool by running it on a workload based

Given workloads, it is difficult for DBAs to select appropriate fields 
in partitioning the fact table due to large search space. 
DBAs cannot easily determine how granular partitions should be made 
for the workloads.

To address this concern, we presented the MLPPI wizard to recommend a fact-table partitioning for a given star schema workload. 
Since \hbox{query-predicates} are exploited 
to capture necessary ranges for fields and define partitions, 
the MLPPI \hbox{recommendation} can be very optimized, customized to the workload.

We proposed the wizard's algorithms consisting of the three phases. 
The wizard \hbox{incrementally} reduced its search space by merging 
the range pair with the least scan or query cost, and eventually reached an MLPPI recommendation. 
In addition, we analyzed the running complexity for the initial and optimized phases. 
Despite the theoretically high bound, in practice, the wizard made much fewer calls in the phases.

We measured the performance of the recommendation by the wizard using 
our workloads.  
We demonstrated that the produced MLPPI solutions by 
the wizard could reduce the total elapsed time by more than a factor of two, 
compared with those of no partitioning approach and partitioning 
done by a human expert.

\section*{Acknowledgments} 
This work was carried out while the first author visited Teradata. 
The work was supported in part under the EDISON program (grant number: 2011-0020576).

\begin{thebibliography}{99}
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{\texttt{#1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
\expandafter\ifx\csname href\endcsname\relax
  \def\href#1#2{#2} \def\path#1{#1}\fi

{\footnotesize
%\vspace{-.1in}

\bibitem{Suh12}
{Y.-K.~Suh et. al.}, {A New Tool for Multi-level Partitioning in Teradata}, 
in: CIKM'12, pp. 2214--2218.

\bibitem{sinclair:ppi}
P.~Sinclair, {Using PPIs to Improve Performance},
  \url{http://www.teradata.com/tdmo/v08n03/pdf/AR5731.pdf}
  %(accessed August 27, 2015) (2008).

\bibitem{klindt09mlppi}
J.~Klindt, {Single-level and Multilevel Partitioned \hbox{Primary} Indexes},
  \url{http://www.teradata.com/white-papers/Single}\linebreak\url{-level-and-Multilevel-Partitioned-Primary-Indexes}\linebreak\url{-eb1889/}
  %(accessed July 27, 2015) (2009).

\bibitem{oneil:ssb}
{P.~O’Neil et. al.}, {The Star Schema Benchmark (SSB)},
  \url{http://www.cs.umb.edu/~poneil/StarSchemaB.PDF} 
  %(accessed \hbox{August} 26, 2015) (2007).

\bibitem{agrawal04:integrating}
{S.~Agrawal et. al.}, {Integrating Vertical and Horizontal \hbox{Partitioning} into Automated Physical Database Design}, 
in: \hbox{SIGMOD} '04, pp. 359--370.

\bibitem{Lightstone04:db2auto}
{S.~Lightstone et. al.}, {Automated Design of
  \hbox{Multi-dimensional} Clustering Tables for Relational Databases}, in:
  VLDB '04, pp. 1170--1181.

\bibitem{nehme2011automated}
{R.~Nehme et. al.}, {Automated Partitioning Design in Parallel
  Database Systems}, in: SIGMOD '11, pp. 1137--1148.

\bibitem{chaudhuri1998autoadmin}
{S.~Chaudhuri et. al.}, {AutoAdmin `What-If' Index Analysis Utility}, ACM SIGMOD Rec. 27~(2) (1998) 367--378.

\bibitem{finkelstein1988physical}
{S.~Finkelstein et. al.}, {Physical Database Design for \hbox{Relational} Databases}, ACM TODS, 13~(1) (1988) 91--128.

\bibitem{labio1997physical}
{W.~Labio et. al.}, {Physical Database Design for Data Warehouses}, in: ICDE '97, pp. 277--288.

\bibitem{rozen1991framework}
{S.~Rozen et. al.}, {A Framework for Automating Physical
  Database Design}, in: VLDB '91, pp. 401--411.

\bibitem{Zilio1998}
{C.~Zilio}, {Physical Database Design Decision Algorithms and Concurrent
  \hbox{Reorganization} for Parallel Database Systems}, Ph.D. thesis,
  Dept. of CS, Univ. of Toronto (1998).

\bibitem{sheet2009oracle}
{Oracle}, {Oracle White Paper---Partitioning with \hbox{Oracle} Database 11g
  \hbox{Release 2}},
  \url{http://www.oracle.com/technetw}\linebreak\url{ork/database/options/partitioning/twp-partitioning}\linebreak\url{-11gr2-2009-09-130569.pdf}
  %(accessed June 23, 2014) (2009).

\bibitem{saa2007oracle}
{Oracle}, {SQL Access Advisor},
  \url{http://docs.oracle.com/cd/B19306_01/server.102/b14211/advisor.htm}
  %(accessed March 30, 2012) (2007).

\bibitem{rao2002automating}
{J.~Rao et. al.}, {Automating
  Physical Database Design in A Parallel Database}, in: SIGMOD '02, pp. 558--569.

\bibitem{agrawal2000automated}
{S.~Agrawal et. al.}, {Automated Selection of Materialized Views and Indexes in SQL Databases}, in:
  {VLDB '00}, pp. 496--505.

\bibitem{Agrawal04:sqlserver}
{S.~Agrawal et. al.}, {Database Tuning Advisor for Microsoft
  SQL Server 2005: Demo}, in: SIGMOD '05, pp. 930--932.

\bibitem{dash2011cophy}
{D.~Dash}, {CoPhy: A
  Scalable, Portable, and Interactive Index Advisor for Large Workloads},
  in: PVLDB, 4~(6) (2011) 362--372.

\bibitem{Kimura11}
{H.~Kimura et. al.}, {Compression Aware
  \hbox{Physical} Database Design}, in: PVLDB, 4~(10) (2011) 657--668.

\bibitem{itwsqlserver}
{Microsoft SQL Server 2000}, {Index Tuning Wizard SQL Server 2000},
  \url{http://technet.microsoft.com/en-us/library/cc966541.aspx} 
  %(2000).

\bibitem{valentin2000db2}
{G.~Valentin}, {DB2 Advisor: An Optimizer Smart Enough to Recommend Its Own
  Indexes}, in: {ICDE '00}, pp. 101--110.

\bibitem{zilio2004recommending}
{D.~Zilio et. al.},
  {Recommending Materialized Views and Indexes with the IBM DB2 Design
  Advisor}, in: ICAC '04, pp. 180--188.

\bibitem{Zilio04:db2design}
{D.~Zilio et. al.}, {DB2 Design Advisor:
  Integrated Automatic Physical Database}, in: VLDB '04, pp. 1087--1097.

\bibitem{tatarowicz2012lookup}
{A.~Tatarowicz et. al.}, {Lookup Tables: Fine-grained Partitioning for Distributed Databases},
  in: ICDE'12, pp. 102--113.

\bibitem{idreoskm07}
{S.~Idreos et. al.}, {Database Cracking}, in: CIDR'07, pp. 68--78.
}

\vspace{-.8in}

\profile[yksuh.eps]{Young-Kyoon Suh}{
received MS and PhD degrees in \hbox{computer science} from KAIST, Korea, and the University of \hbox{Arizona}, USA, in 2005 and 2015, respectively. 
He is a senior researcher at KISTI. His research interests include database \hbox{design} and \hbox{systems}, 
science of databases, and big data management.}

%\profile[ahmad.eps]{Ahmad Ghazal}
%{received the PhD degree in computer science from Univ. of Illinois at Chicago.  
%He is now at \hbox{Oracle} Endeca. He is a well-known industry expert in query optimization theory and practice. 
%%He peer-reviewed submissions to the Journal of Systems and Software, IEEE TKDE and DOLAP. 
%His research interest lies in query optimization, big data systems, and physical database design.}

\vspace{-.2in}

\profile[alain.eps]{Alain Crolette}
{received a PhD degree from Univ. of California at Los \hbox{Angeles} (UCLA).  
He is a performance specialist at Teradata. 
He also worked as senior benchmark consultant at Teradata.
Prior to joining Teradata, he was also a technical \hbox{consultant} at Abacus Programming Corporation. 
His research focuses on big data systems, query optimization, and physical database design.}

\profile[pekka.eps]{Pekka Kostamaa}
{is a director of Product Management at Teradata. 
He worked as VP of Engineering and Big Data Lab at Teradata Aster.
Prior to VP, he was a director in the Department of Advanced Development at Teradata.
He is interested in query optimization and physical database design.}

%\vspace\fill

\end{thebibliography}
\end{document}

### 2017-02-23 mlppi-comsis17/cover-letter.tex
ComSis cover letter

@@
 
-\opening{Dear Prof. Vladimir Kurbalija (or Editor-in-Chief):}
+\opening{Dear Prof. Mirjana Ivanovi\'c (Editor-in-Chief):}
 
@@
 
-\item Expanded Algorithms: we expand the description of the pre-processing, initial and optimized phase algorithms shortened in the earlier version.
+\item Expanded Algorithms: we substantially expand the description of the pre-processing, initial and optimized phase algorithms shortened in the earlier version.
 Furthermore, this submission contains three additional algorithms as follows:

### 2017-02-24 mlppi-comsis17/paper.tex
MLPPI wizard

\documentclass[runningheads]{comsis2}

%\usepackage[ruled,linesnumbered]{algorithm2e}
%\usepackage{amssymb,amsthm,amsmath}
\usepackage{balance}
\usepackage{subfigure}
\usepackage{float}
%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{lemma}[theorem]{Lemma}
\usepackage{latexsym}
\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{comment}
\usepackage[dvips]{graphicx}          % graphicx package for including ps files

%% Necessary definitions for the running heads
\def\journalissue{Computer Science and Information Systems ?(?):??--??}
\def\paperidnum{DOI: N/A}
\setcounter{page}{1}

\title{An Automated Multi-level Partitioning Tool on Analytical Workloads}
%\titlenote{An earlier version of this article was presented as a short paper at ACM CIKM'12.}
%% Use this if the title is too long for the running heads
%\titlerunning{An Automated Multi-level Partitioning Tool on Analytical Workloads}

\author{Young-Kyoon Suh\inst{1} \and Alain Crolotte\inst{2} \and Pekka Kostamaa\inst{2}}
%Correspondence to Young-Kyoon Suh, now at Korea Institute of Science and Technology Information
%% Use this the list of authors is too long for the running heads
%\authorrunning{Correspondence to Young-Kyoon Suh, now at Korea Institute of Science and Technology Information}

\institute{Department of Computer Science, The University of Arizona, Tucson, AZ 85721, USA\\ 
  \email{yksuh@cs.arizona.edu}
  \and
  Teradata Corporation, El Segundo, CA 90245, USA\\
  \email{\{alain.crolotte,pekka.kostamaa\}@teradata.com}
%  \and
%  Korea Institute of Science and Technology Information, Daejeon 34141 Republic of Korea\\ 
%  \email{yksuh@kisti.re.kr}
}

\def\form#1{$\langle{#1}\rangle$}
\def\range#1{$[{#1}]$}
\def\openrange#1{$({#1})$}
\def\lopenrange#1{$({#1}]$}
\def\ropenrange#1{$[{#1})$}
\def\product#1{$(#1)$}
\definecolor{grey}{RGB}{200,200,200}
\newcommand{\hilite}[1]{\colorbox{grey}{#1}}
\newcommand{\hilitey}[1]{\colorbox{yellow}{#1}}
\newcommand{\hiliting}[1]{\colorbox{grey}{#1}}
\long\def\todo#1{\hilitey{{\bf TODO:} {\em #1}}}
\long\def\shorten#1{}
\long\def\comment#1{}
\def\MAX{\mbox{\sl MAX}}
\def\LIMIT{\mbox{\sl LIMIT}}
\def\getQueryCost{\mbox{\sl getQueryCost}}
\def\getScanCost{\mbox{\sl getScanCost}}
\def\BigO{\mbox{\sl O}}
\def\NULL{\mbox{\sl NULL}}
\def\SolA{\mbox{\sl A}}
\def\SolB{\mbox{\sl B}}
\def\SolC{\mbox{\sl C}}
\def\SolS{\mbox{\sl S}}
\def\Qone{\mbox{\sl q1}}
\def\Qtwo{\mbox{\sl q2}}
\def\Sone{\mbox{\sl s1}}
\def\Stwo{\mbox{\sl s2}}
\def\SolSone{\mbox{\sl S1}}
\def\SolStwo{\mbox{\sl S2}}
\def\SolSthree{\mbox{\sl S3}}
\def\SolSfour{\mbox{\sl S4}}
\def\SolSfive{\mbox{\sl S5}}

\begin{document}

\maketitle

\begin{abstract} 
Typically, it is a daunting task for a database administrator (DBA) to devise a solution to partitioning a huge fact table accessed by query workloads, for better performance. To relieve such a burden for the DBA, we introduce an intelligent physical database designer to recommend an optimized partitioning on the fact table. This designer uses a greedy algorithm for search space enumeration. This space is driven by predicates of a given query workload. The designer takes advantage of the cost model of a query optimizer to prune the search space. The wizard resides completely on a client and interacts with the optimizer via APIs. Thus, there is no overhead to instrument the optimizer code. Furthermore, the predicate-driven method can be applied to any clustering or partitioning schemes. We show that the designer’s recommendation outperforms a human expert’s solution. We also demonstrate that the recommendation scale very well with increasing workload and growing fact table.

\vspace{6pt}\textbf{Keywords:} keyword1, keyword2.
\end{abstract}

\section{Introduction}

The text of the paper\dots

%\bibliographystyle{splncs03}
%\bibliography{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% For the final version of the paper: %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Author information
%\vspace{4ex}\noindent
%\textbf{Author One} is\dots
%
%\bigskip\noindent
%\textbf{Author Two} is\dots
%
%\bigskip\noindent
%\textbf{Author Three} is\dots

%% Reception and acceptance information
%\bigskip
%\paragraph{Received: Month DD, 20YY; Accepted: Month DD, 20YY.}

\end{document}

### 2017-02-28 mlppi-comsis17/paper.tex
ComSIS working draft

@@
 
-%\usepackage[ruled,linesnumbered]{algorithm2e}
+\usepackage[ruled,linesnumbered]{algorithm2e}
+%\usepackage{algorithm}
 %\usepackage{amssymb,amsthm,amsmath}
+\usepackage{amssymb}
 \usepackage{balance}
@@
 
-\title{An Automated Multi-level Partitioning Tool on Analytical Workloads}
+\title{MLPPI Wizard: An Automated Multi-level Partitioning Tool on Analytical Workloads}
 %\titlenote{An earlier version of this article was presented as a short paper at ACM CIKM'12.}
@@
 \begin{abstract} 
-Typically, it is a daunting task for a database administrator (DBA) to devise a solution to partitioning a huge fact table accessed by query workloads, for better performance.
-To relieve such a burden for the DBA, we introduce an intelligent physical database designer to recommend an optimized partitioning on the fact table.
-This designer uses a greedy algorithm for search space enumeration.
+Typically, it is a daunting task for a database administrator (DBA) to figure out how to partition a huge fact table accessed by query workloads for better performance.
+To relieve such a burden, we introduce an intelligent partitioning tool to recommend an optimized partitioning on the fact table.
+This tool uses a greedy algorithm for search space enumeration.
 This space is driven by predicates of a given query workload.
-The designer takes advantage of the cost model of a query optimizer to prune the search space.
-The wizard resides completely on a client and interacts with the optimizer via APIs.
+The tool takes advantage of the cost model of a query optimizer to prune the search space.
+The tool resides completely on a client and interacts with the optimizer via APIs.
 Thus, there is no overhead to instrument the optimizer code.
-Furthermore, the predicate-driven method can be applied to any clustering or partitioning schemes.
-We show that the designer’s recommendation outperforms a human expert’s solution.
+Furthermore, the predicate-driven method can be applied to any clustering or partitioning scheme.
+We show that the tool's recommendation outperforms a human expert's solution.
 We also demonstrate that the recommendation scale very well with increasing workload and growing fact table.
 
-\vspace{6pt}\textbf{Keywords:} keyword1, keyword2.
+\vspace{6pt}\textbf{Keywords:} Star Schema, Fact Table, Multi-Level Partitioning.
 \end{abstract}
 
+%\section{Introduction}
+%
+%The text of the paper\dots
+
 \section{Introduction}
+\label{sec:intro}
 
-The text of the paper\dots
+Tables and materialized views in Teradata are hash distributed based 
+on a user-specified column or set of columns called {\em primary index}. 
+Each virtual processor, a unit of parallelism called {\em AMP} in the \hbox{Teradata} \hbox{relational} 
+database management system (DBMS), 
+receives a subset of the data and stores it in hash order.
 
-%\bibliographystyle{splncs03}
-%\bibliography{example}
+Teradata users typically choose primary index fields, 
+so that the data is evenly distributed among the AMPs. 
+The primary index fields are also chosen to \hbox{reflect} join fields 
+in workloads to accomplish cheaper local joins that do not require 
+data shuffling.
+
+PPI (Partitioned Primary Index)~\cite{sinclair:ppi} is an optional horizontal 
+partitioning scheme applied locally on each AMP's data. 
+Note that in other database products, this type of partitioning would 
+most likely to be called {\em clustering}. 
+In the rest article we use the term \hbox{{\em partitioning}} only, not clustering, to avoid confusion. 
+Database \hbox{administrators} (DBAs) usually choose the columns for PPI, 
+based on join fields and single table \hbox{predicates} to optimize the queries included 
+in the workload. 
+The PPI columns are used to physically cluster data with the same values 
+together in contiguous data blocks. 
+This allows ``partition elimination'' in scans and joins for performance improvement.
+PPI can be specified as single or multiple (or nested) levels. 
+This type of partitioning scheme is known as 
+Multi-Level PPI ({\em MLPPI)}~\cite{klindt09mlppi}.
+
+By allowing non-qualified partitions to be eliminated, MLPPI can reduce 
+significantly the amount of data to be scanned to answer a query. 
+But a large number of partitions can create significant overhead, 
+particularly in joins and table maintenance operations 
+such as inserts and deletes, so that the selection of partitions can be usually a balancing act.
+
+Figure~\ref{fig:exam} exemplifies an MLPPI table with potential partitioning schemes. 
+This example is a subset of the {\tt LINEORDER} table from the Star Schema Benchmark (SSB)~\cite{oneil:ssb}. 
+Note that the full table with modified data types was actually used in our experiments.
+
+\begin{figure}[t]
+\begin{center}
+{\small
+\begin{tabular}{l}
+{\tt CREATE TABLE LINEORDER(} \\
+\hspace{0.1in}{\tt LO\_ORDERKEY INTEGER,} \\
+\hspace{0.1in}{\tt LO\_QUANTITY INTEGER,} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT INTEGER} \\
+{\tt )} \\
+{\tt PRIMARY INDEX ( LO\_ORDERKEY )} \\
+{\tt PARTITION BY (} \\
+\hspace{0.1in}{\tt CASE\_N(} \\
+\hspace{0.2in}{\tt LO\_DISCOUNT >= 7,} \\
+\hspace{0.2in}{\tt NO CASE OR UNKNOWN),} \\
+\hspace{0.1in}{\tt CASE\_N(}\\
+\hspace{0.2in}{\tt LO\_QUANTITY < 25,} \\
+\hspace{0.2in}{\tt LO\_QUANTITY >= 25 AND LO\_QUANTITY <= 30,} \\
+\hspace{0.2in}{\tt LO\_QUANTITY > 30 AND LO\_QUANTITY <= 35,} \\
+\hspace{0.2in}{\tt NO CASE OR UNKNOWN)} \\
+{\tt );} \\
+\end{tabular}
+}
+\end{center}
+\vspace{-0.2in}
+\caption{An Example of MLPPI Table\label{fig:exam}}
+\end{figure}
+
+In the definition of {\tt LINEORDER} in the figure, the primary index 
+is {\tt LO\_ORDERKEY}. 
+The \hbox{primary} index \hbox{dictates} the AMP on which a row will be located, 
+while the partitioning of data will be dictated by the values and 
+ranges associated with {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}. 
+{\tt LO\_DISCOUNT}, for example, has two ranges for \hbox{values} 
+greater than or equal to 7 and one ({\tt NO CASE OR UNKNOWN}) 
+for all the other values, while \linebreak {\tt LO\_QUANTITY} has four ranges. 
+As a result, the relational table will have a total of 2$\times$4 = 8 partitions as follows.
+
+\begin{center}
+\begin{tabular}{|c|p{6cm}|}\hline 
+Partition & Condition \\ \hline
+1		& {\scriptsize  {\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY < 25}} \\ \hline
+2		& {\scriptsize  {\tt LO\_DISCOUNT >= 7 \&\& 25 <= LO\_QUANTITY <= 30}} \\ \hline
+3		& {\scriptsize {\tt LO\_DISCOUNT >= 7 \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
+4		& {\scriptsize {\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY no case}}\\ \hline
+5       & {\scriptsize {\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY < 25}} \\ \hline 
+6		& {\scriptsize {\tt LO\_DISCOUNT no case \&\& 25 <= LO\_QUANTITY < 30}} \\\hline								  	        
+7		& {\scriptsize {\tt LO\_DISCOUNT no case \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
+8		& {\scriptsize  {\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY no case}} \\ \hline
+\end{tabular}
+\end{center}
+
+Whatever partition set is chosen, 
+the selection must be semantically correct; namely, 
+the mapping of rows to partitions must be an injection. 
+In other words, the constraints must form a covering of 
+the entire range, so that a row will belong to exactly one and one partition. 
+The Teradata optimizer then applies \hbox{partition} elimination for queries 
+that specify conditions on {\tt LO\_DISCOUNT} and/or {\tt LO\_QUANTITY}. 
+For \hbox{example}, only partitions 1 and 2 are needed for the query 
+``{\tt SELECT * FROM LINEORDER WHERE LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
+Similarly, partitions 1 and 5 are sufficient to answer the query 
+``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}.'' 
+\linebreak The respective sizes of the partitions are a factor of the data distribution. 
+To see MLPPI in greater detail, check our references~\cite{sinclair:ppi,klindt09mlppi}.
+
+Next, using the full version of {\tt LINEORDER} as defined in SSB~\cite{oneil:ssb} 
+we provide examples of partitioning schemes that 
+a DBA may define based on a small query set. 
+Consider a query set $Q$ consisting of two queries $q$1 and $q$2 from the SSB set~\cite{oneil:ssb}, as shown below.
+
+\begin{center}
+{\small
+\begin{tabular}{rl}
+$q$1:	& {\tt SELECT SUM(l.LO$\_$EXTENDEDPRICE*l.LO\_DISCOUNT)} \\ 
+		& {\tt FROM LINEORDER l, DDATE d} \\
+		& {\tt WHERE l.LO\_ORDERDATE = d.D\_DATEKEY} \\
+		& {\tt AND d.D\_YEAR = `1993'} \\
+		& {\tt AND l.LO\_DISCOUNT IN (1, 4, 5)} \\
+        & {\tt AND l.LO\_QUANTITY <= 30} \\ 
+		&						   \\	        
+$q$2:	& {\tt SELECT c.C\_NATION, SUM(l.LO\_REVENUE)} \\ 
+		& {\tt FROM CUSTOMER c, LINEORDER l} \\
+		& {\tt WHERE l.LO\_CUSTKEY = c.LO\_CUSTKEY} \\
+		& {\tt AND c.C\_REGION=`EUROPE'} \\
+		& {\tt AND l.LO\_DISCOUNT >= 7} \\
+		& {\tt AND l.LO\_QUANTITY >= 25 AND l.LO\_QUANTITY <= 35} \\
+		& {\tt GROUP BY c.C\_NATION} \\
+		& {\tt ORDER BY revenue desc}. \\
+\end{tabular}
+}
+\end{center}
+
+Let's now focus on the {\tt LINEORDER} fact table and 
+the predicates involving {\tt LINEORDER} fields only. 
+There are five constraints identified by the query number 
+and the sequence number of the constraint in each query, as illustrated below.
+%\vspace{-0.05in}	
+\begin{center}
+\begin{tabular}{|c|l|}\hline 
+$q$1.1 & {\tt LO\_DISCOUNT IN (1,4,5)} \\ \hline
+$q$1.2 & {\tt LO\_QUANTITY <= 30} \\ \hline
+$q$2.1 & {\tt LO\_DISCOUNT >= 7} \\ \hline
+$q$2.2 & {\tt LO\_QUANTITY >= 25}\\ \hline
+$q$2.3 & {\tt LO\_QUANTITY <= 35} \\ \hline 
+\end{tabular}
+\end{center}
+%\vspace{-0.05in}	
+At this point the DBA needs to consider options 
+based only on two fields and the five constraints. 
+There are many possibilities from a {\em fine-grained} partition set 
+to a loser definition.
+
+Considering for the time being the column {\tt LO\_DISCOUNT}, 
+one solution is to identify each value 
+for the {\tt IN} predicate in $q$1.1 
+and use $q$2.1 as is. This yields the following partitioning expression 
+for {\tt LO\_DISCOUNT}:
+\begin{center}
+\begin{tabular}{l}
+{\tt CASE\_N(} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT = 1,} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT = 4,} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT = 5,} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT >= 7,} \\
+\hspace{0.1in}{\tt NO CASE OR UNKNOWN} \\
+{)}. \\
+\end{tabular}
+\end{center}
+The above expression minimizes the size of the partitions 
+by focusing exactly on the values required to satisfy the constraints, 
+but creates a large number of small partitions.
+
+Another possibility is to look at the maximum and minimum 
+values in the {\tt IN} set and to build an {\tt AND} clause equivalent to 
+a between clause yielding the following:
+\begin{center}
+\begin{tabular}{l}
+{\tt CASE\_N(} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT >= 1 AND LO\_DISCOUNT <= 5,} \\
+\hspace{0.1in}{\tt LO\_DISCOUNT >= 7,} \\
+\hspace{0.1in}{\tt NO CASE OR UNKNOWN} \\
+{)}. \\
+\end{tabular}
+\end{center}
+
+\noindent The above partitioning solution decreases the number of partitions, 
+compared to the previous solution but still focuses sharply on the constraints 
+with still relatively small partitions.
+
+Another possibility is to use the partitioning associated with {\tt LO\_DISCOUNT} shown 
+in the beginning of this section with only two partitions.
+
+Similar considerations apply to the partitioning associated with {\tt LO\_QUANTITY} 
+this time with ranges and covering problems difficult to deal with. 
+The resulting partitioning scheme for the table includes both {\tt LO\_DISCOUNT}
+and {\tt LO\_QUANTITY}, so that the number of possible combinations is the product of 
+the potential combinations for each field. 
+Also, there are two queries, and one partitioning scheme may be better for one query or the other. 
+As a result, the DBA will be faced with a daunting combinatorial search problem 
+and no clear basis to decide on which combination is the best. 
+This state of affairs begs for a {\em tool} to assist the DBA.
+
+In the sequel, we introduce an automated tool, called {\em MLPPI wizard}. 
+Indeed, the \hbox{wizard} is the first physical database design tool 
+developed at Teradata.
+The tool is based on a novel technique using a greedy algorithm for 
+search space enumeration. 
+This tool based on a general framework allowing general expressions, 
+ranges and case expressions for \hbox{partition} \hbox{definitions} 
+is particularly well-suited for an MLPPI definition. \linebreak
+The {\em predicate-driven} \hbox{technique} used by the tool can be applied 
+to any clustering or partitioning based on simple fields and expressions or complex SQL predicates. 
+The wizard also borrows the optimizer's cost model 
+to prune the search space and reach a final solution.
+
+\begin{figure*}[htp!]
+\centering
+\includegraphics[scale=0.5]{architecture}
+%\vspace{-0.2in}
+\caption{The MLPPI Wizard Architecture\label{fig:arch}}
+\end{figure*}
+
+Figure~\ref{fig:arch} illustrates how our tool recommends 
+the final MLPPI customized for a given workload consisting of 
+a set of queries and corresponding weights. 
+The wizard passes through a series of phases: {\em Preprocessing}, 
+{\em Initial}, and {\em Optimized} Phases. 
+In the preprocessing phase the tool produces the most granular partitioning (MGP) based on the predicates collected from every query.
+In the initial phase the wizard refines the passed MGP 
+by merging a pair of partitions with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}) 
+and then yields an initial (feasible) MLPPI based on the refined partitions.
+In the \hbox{optimization} phase, the tool makes a further attempt to optimize the initial MLPPI. 
+We provide in greater \hbox{detail} each phase, 
+using the query set $Q$ consisting of $q$1 and $q2$ in the rest of the article.
+
+\shorten{
+An initial MLPPI is produced through the top two phases reflecting practical constraints by field or partition. 
+It is further optimized in the last phase.
+In the preprocessing phase, 
+for each query its predicate(s) is extracted via an API call (on {\tt explain}) made to the server. 
+Then, the wizard produces the most granular partitioning (MGP) based on the 
+predicates collected from every query. 
+If the MGP has too many fields over a field count limit, 
+then the wizard \hbox{incrementally} throws away a field from the MGP until the limit.
+
+In the initial phase, the passed MGP gets reduced if it includes more partitions than a partition count limit allowed by an MLPPI. 
+The refinement can be \hbox{incrementally} done by merging a pair of partitions 
+with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}), 
+until the limit. 
+After the phase, an initial (feasible) MLPPI is produced based on the refined partitions.
+
+In the optimization phase, the tool makes a further attempt to optimize the initial MLPPI. 
+Similar to the initial phase, more partitions can be reduced by merging ones with the 
+least {\em query cost} (as will be described in Section~\ref{sec:opt_phase}). 
+If the ongoing solution cannot be further improved, the tool makes the final MLPPI recommendation. 
+We will provide more details about each phase later, 
+using the query set $Q$ consisting of $q$1 and $q2$ 
+in the rest of the article.}
+
+To the best of our knowledge, there is no existing industrial tool 
+to solve the {\it {\bf multi-level partitioning}} problem except our wizard. 
+The wizard is contained completely on the client side, as opposed to 
+previous work~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} 
+requiring optimizer code extension. 
+Our tool simply uses existing APIs to simplify the queries, capture fact table 
+predicates and costs, and uses these items to make a recommendation. 
+This makes the wizard extensible and portable to different releases 
+of the Teradata DBMS server.
+
+The article is a substantial extension of prior work~\cite{Suh12}, 
+and organized in the following way. 
+Section~\ref{sec:algorithms} presents the multi-level \hbox{partitioning} 
+algorithms for the wizard. 
+In turn, Section~\ref{sec:analysis} conducts an analysis of the complexity of 
+the algorithms. 
+Section~\ref{sec:experiment} reports the performance evaluation results. 
+Section~\ref{sec:related_work} states how \hbox{Teradata} MLPPI wizard 
+is different from some existing tools proposed by other vendors. 
+Lastly, Section~\ref{sec:conclusion} summarizes our discussion.
+
+\section{The MLPPI wizard}
+\label{sec:algorithms}
+
+In this section we describe in detail each phase of the wizard using the query set $Q$ provided in Section~\ref{sec:intro}.
+
+\subsection{The Preprocessing Phase}
+\label{sec:pre_phase}
+
+Algorithm~\ref{algo:pre_phase} sketches 
+the steps in the preprocessing phase. 
+This phase consists of a total of six steps.
+
+\vspace{-.1in}
+
+\begin{algorithm}	
+\caption{The Preprocessing Phase}
+\label{algo:pre_phase}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$Q$ (input query set) }
+\Output{$R$ (non-overlapping range set), $M$ (query-to-range-set map)}
+\hspace{.05in} Query Simplification (on $Q$)\\
+\hspace{.05in} Range Extraction (from $Q$) \\
+\hspace{.05in} Non-Overlapping Range ($R$) Construction \\
+\hspace{.05in} Field Count Limit Check (on $R$) \\
+\hspace{.05in} Query-to-Range-Set Map ($M$) Construction \\
+\hspace{.05in} Partition Count Limit Check (on $R$) \\
+}
+\end{algorithm}
+
+%\vspace{-.3in}
+
+\subsubsection{Query Simplification}
+
+The first step of this phase is to simplify the predicate(s) of each query 
+by removing redundant conditions. 
+For instance, if a query predicate has the condition of 
+``{\tt LO\_DISCOUNT IN (1, 4, 5) AND LO\_DISCOUNT IN (2, 3, 4, 5)}'', 
+then the predicate can be simplified as ``{\tt LO\_DISCOUNT IN (4, 5)}.'' 
+This can be done via an API call to the DBMS server. 
+In the running example this simplification is not required.
+
+%\vspace{-.3in}
+
+\subsubsection{Range Extraction}
+
+The following step is to gather the simplified predicates and 
+then obtain ranges from the collected predicates. 
+This can also be done through an API call to the server. 
+The query predicate is of the form  
+\begin{center}
+\begin{tabular}{l} 
+\form{\tt variable} \form{\tt op} \form{\tt constant},
+\end{tabular}
+\end{center}
+where \form{\tt variable} is a field from {\tt LINEORDER}, 
+\form{\tt op} is in 
+\begin{center}
+\begin{tabular}{l} 
+\{{\tt $ = $ }, {\tt $ < $}, {\tt $ <= $}, {\tt $ >= $}, {\tt $ > $}, 
+{\tt IN}\},
+\end{tabular}
+\end{center}
+and \form{\tt constant} represents a constant value(s). 
+All \form{\tt op}s are self-explanatory. 
+In particular, `{\tt IN}' is a list predicate implying `{\tt OR}' operator.
+In the running example, 
+we can collect a set of predicates $P$ = \{$p$1,$p$2,$p$3,$p$4\} from $Q$, where
+
+%\vspace{-.15in}
+
+\begin{center}
+\begin{tabular}{l} 
+$p$1: {\tt l.LO\_DISCOUNT IN (1, 4, 5)} \\ 
+$p$2: {\tt l.LO\_DISCOUNT >= 7} \\ 
+$p$3: {\tt l.LO\_QUANTITY <= 30} \\ 
+$p$4: {\tt l.LO\_QUANTITY >= 25} {\tt AND} {\tt l.LO\_QUANTITY <= 35}. \\ 
+\end{tabular}
+\end{center}
+
+%\vspace{-.15in}
+
+In turn, the wizard constructs a bi-directional {\em map} between a query and 
+associated predicates, so that for each query the corresponding predicate(s) can be 
+found directly in the map, and vice versa. 
+In the example, a map, called $M$1, is built 
+between $Q$ and $P$, and $M$1 is filled with the following entries:
+
+$M$1:
+\begin{center}
+\begin{tabular}{ll} 
+\form{q1, \{p1, p3\}} \\
+\form{q2, \{p2, p4\}}.
+\end{tabular}
+\end{center}
+
+Next, we gather fields referenced by predicates in $P$, extract 
+ranges from the predicates, and group the ranges on each of the fields. 
+Assume $I$ to be the list of fact table fields referenced by $P$. 
+In the workload $Q$, we can see that there are two fields used by $P$: 
+{\tt LO\_DISCOUNT}, {\tt LO\_QUANTITY}. 
+These are added to $I$.
+
+After that, we construct a set of distinct ranges on each field in $I$, using the predicates in $P$. 
+Then, the whole range set becomes a kind of two-dimensional array, called $R$. 
+The array $R$ captures the range representation of the predicates. 
+Each entry under a field in $R$ represents a pair of (start and end) values forming the range. 
+In the range representation, infinity ($\infty$) or -infinity (-$\infty$) can 
+be used for unbounded ranges. 
+``['' or ``]'' are used for closed ranges, and ``('' or ``)'' for open ranges. 
+In particular, {\tt IN} predicate is represented by multiple ranges. 
+For instance, the above $p$1 can be represented by the following three ranges 
+\range{1, 1}, \range{4, 4}, \range{5, 5}; however, the last two consecutive 
+ranges can be consolidated as \range{4, 5}. 
+In the continuing example the range set $R$ is formed by the entries below.
+
+$R$:
+\begin{center}
+\begin{tabular}{c|c} 
+ {\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
+\range{1, 1} & \lopenrange{\infty, 30}\\
+\range{4, 5} & \range{25, 35} \\
+\ropenrange{7, \infty} &  \\
+\end{tabular}
+\end{center}
+
+Then, each predicate can be mapped to its associated ranges in $R$. That is, 
+we can have another map, called $M$2, between $P$ and $R$. 
+Let $R$[$i$, $j$] denote the interval value for the $i$-th field and $j$-th range in $R$. 
+$R$[1, 3], for instance, indicates range \ropenrange{7, \infty} under {\tt LO\_DISCOUNT}. 
+In the running example, $M$2 is constructed as follows:
+
+$M$2:
+\begin{center}
+\begin{tabular}{l}
+$\langle${$p$1}, \{$R$[1, 1],$R$[1, 2]\}$\rangle$ \\
+$\langle${$p$2}, \{$R$[1 ,3]\}$\rangle$ \\ 
+$\langle${$p$3}, \{$R$[2, 1]\}$\rangle$ \\
+$\langle${$p$4}, \{$R$[2, 2]\}$\rangle$. \\
+\end{tabular}
+\end{center}
+
+\subsubsection{Non-Overlapping Range Construction}
+
+Since there could be multiple predicates on the same field across queries, 
+extracted ranges may overlap each other. 
+The overlapping ranges must be broken without any common portion, so that 
+a pair of consecutive, \hbox{non-overlapping} ranges can be 
+considered for a merge in further phases.
+
+Algorithm~\ref{algo:split_overlap} describes the way of breaking 
+the \hbox{overlap} of ranges.
+
+%\vspace{-.1in}
+
+\begin{algorithm}[h]
+\caption{Splitting Overlapping Ranges}
+\label{algo:split_overlap}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$R$ having overlapping ranges}
+\Output{$R$ with consecutive, non-overlapping ranges}
+	\ForEach{field $i$ $\in$ $R$}{	
+		$L$ $\leftarrow$ Get the range set of $i$ from $R$ and sort by start value \\
+		$j$ = 0 \;
+		\While{$j$ $<$ $|L|$-1}{
+			$r_{j}$, $r_{j+1}$ $\leftarrow$ Adjacent ranges in $L$ \\
+			\If{$r_{j}$.$end$ $\geq$ $r_{j+1}$.$start$}{
+				Make a split by modifying values of $r_{j}$ and $r_{j+1}$. \\
+				Insert into $L$ an intermediate range if any. \\
+				Re-sort ranges in $L$ and reset $j$ to 0. \\
+			}\lElse{$j$++;}
+		}
+		Update $R$ with the final $L$. 
+	} 
+}
+\end{algorithm}
+
+\vspace{-.1in}
+
+\noindent For each field, the tool gets its associated ranges and sorts them 
+by start value. 
+It then checks whether or not adjacent ranges overlap each other.
+If that is the case, then we make a split between the ranges. 
+In the example, the split is applied on the 
+overlapping ranges, \lopenrange{\infty, 30} and \range{25, 35}, under {\tt LO\_QUANTITY}. 
+Subsequently, a common, intermediate range (\range{25, 30}) is newly created and inserted into the range set 
+belonging to the {\tt LO\_QUANTITY} field, in order to fill the gap. 
+Then, $R$ in the example is shown as follows.
+
+\
+{\it $R$}:
+\begin{center}
+\begin{tabular}{c|c} 
+ {\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
+\range{1, 1}      & \openrange{\infty, 25} \\
+\range{4, 5}      & \range{25, 30} \\
+\ropenrange{7, \infty} & \range{31, 35} \\
+\end{tabular}
+\end{center}
+
+The update of $R$ affects the existing map $M$2. 
+In the running example, $p$3 and $p$4 influenced by the split get mapped to 
+new range sets, \{$R$[2, 1], $R$[2, 2]\} and \{$R$[2, 2], $R$[2, 3]\}, respectively. 
+Of course, the range sets mapped to $p$1 and $p$2 referencing {\tt LO\_DISCOUNT} remain unchanged. 
+As a result, in the example we obtain the updated $M$2 as follows:
+
+$M$2:
+\begin{center}
+\begin{tabular}{l}
+$\langle${$p$1}, \{$R$[1, 1],$R$[1, 2]\}$\rangle$ \\
+$\langle${$p$2}, \{$R$[1, 3]\}$\rangle$ \\ 
+$\langle${$p$3}, \{$R$[2, 1], $R$[2, 2]\}$\rangle$ \\
+$\langle${$p$4}, \{$R$[2, 2], $R$[2, 3]\}$\rangle$. \\
+\end{tabular}
+\end{center}
+
+%\vspace{.1in}
+
+\subsubsection{Field Count Limit Check}
+
+%The DBMS or user may impose a limit on the number of fields in the final 
+%MLPPI recommendation. 
+At present, the Teradata DBMS has a limit (64) of the fields 
+that can be used in an MLPPI definition.
+If the number of fields present in $R$ exceeds the field count limit, 
+we determine which field(s) should be thrown away to satisfy the limit. 
+(The star schema fact table we use has much fewer fields than the limit, 
+and thus, this step will not be executed.)
+To choose victim fields, the tool computes the weighted sum of 
+{\em query cost} of queries regarding each field. 
+The sum can be obtained by adding up every query cost on an MLPPI using 
+only the ranges under the field. 
+(We will cover how to measure the query cost in Section~\ref{sec:opt_phase}.) 
+We incrementally discard a field with the highest query cost sum 
+until the limit is reached. 
+Accordingly, we can update the existing $M$2. 
+One might say that the MLPPI may not be exploited if too many ranges, 
+exceeding a partition count limit (or 65,536 as of now) are found in one field. 
+But we assume that such an extreme case is not expected. Even if 
+it happens, we can utilize the merges of consecutive ranges to make 
+the MLPPI feasible.
+
+\subsubsection{Query-to-Range-Set Map Construction}
+
+Now, the tool can create a bi-directional {\em query-to-range-set} map, called $M$, 
+between $Q$ and $R$, using the existing maps $M$1 and $M$2. 
+It leverages a transitive property from $M$1 to $M$2.
+Once $M$ is constructed, 
+the intermediate maps ($M$1 and $M$2) are not used in the rest of phases. 
+$M$ is shown as below:
+
+{\it $M$}:
+\begin{center}
+\begin{tabular}{ll} 
+$\langle${$q$1}, \{$R$[1, 1], $R$[1, 2], $R$[2, 1], $R$[2, 2]\}$\rangle$ \\
+$\langle${$q$2}, \{$R$[1, 3], $R$[2, 2], $R$[2, 3]\}$\rangle$.
+\end{tabular}
+\end{center}
+
+In the subsequent phases $M$ gets used for computing costs and updated 
+along with a merge of ranges, and $R$ is refined for an MLPPI recommendation.
+
+\subsubsection{Partition Count Limit Check}
+
+In this regard, the range set $R$ can be used to define an MLPPI using each field as one level. 
+However, if the current number of partitions by $R$ is greater than 
+the partition count limit, 
+then it is not possible to make a feasible MLPPI based on ranges in $R$.
+
+The actual partition count limit is ample enough to pass the running example, 
+but to continue our discussion, assume that in our discussion the limit is {\it 15}.
+From $R$ in the running example, we can obtain a total of (number of ranges in {\tt LO\_DISCOUNT}) 
+$\cdot$(number of ranges in {\tt LO\_QUANTITY}) = (3+1) $\cdot$ (3+1) = {\it 16}
+partitions. 
+Because the total partitions surpass the limit, 
+the tool should pass through the initial phase in which 
+fewer partitions than the limit are made. 
+(Otherwise, the wizard will proceed to the optimized phase immediately.) 
+Note that one more range is counted per field in the calculation. 
+The added range is equivalent to the ``{\tt NO CASE OR UNKNOWN)}'' case in Figure~\ref{fig:exam}.
+
+The tool is now ready to proceed to the next phases, 
+with $R$ (input range set, or MGP) and the prepared $M$ (query-to-range-set map). 
+%In the following subsections, we describe how the MLPPI solution for $Q$ 
+%can be eventually made using $M$ and $R$.
+
+\subsection{The Initial Phase}
+\label{sec:init_phase}
+
+In this phase, we incrementally merge a range pair in $R$ to reduce partitions. 
+The merge continues until the number of ongoing partitions 
+drop below the partition count limit. 
+Once reaching the limit, we can have a feasible MLPPI with 
+the remaining partitions.
+
+By and large, overall I/O cost may be increased by the merge. 
+Since every row is hash-sorted in each AMP of the Teradata DBMS server, 
+typically the full scan on a partition is done to retrieve 
+all the rows matching a given query. 
+In the case of a merged partition, we may read the non-qualifying rows 
+that would not be seen before the merge, thereby paying more I/O to answer 
+the query.  
+To minimize the merge overhead, we pick up the range pair incurring 
+the least I/O increase in a heuristic fashion.
+
+To choose the desirable range pair for a merge, 
+the tool leverages the scan cost of a query on a range pair.
+Algorithm~\ref{algo:comp_scan_cost} represents how 
+the scan cost can be computed on the range pair that \hbox{influences} a query. 
+The scan cost represents the I/O cost to answer the query when the range pair is merged. It is 
+modeled as the {\em number of blocks} that are to be read 
+for the target query. To compute the scan cost of a query $q$, we use 
+a corresponding scan cost query ($s$), and 
+it can be constructed as follows: 
+\begin{center}
+$s$: {\tt SELECT * FROM $F$ WHERE} {\it CP},
+\end{center} 
+where $F$ is a fact table, and 
+{\it CP} indicates the predicates  
+restored from the ranges mapped to $q$ from $M$.
+
+\begin{algorithm}[t]
+\caption{Scan Cost Computation}
+\label{algo:comp_scan_cost}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$q$ (query), $rp$ (range pair), $M$ (query-to-range-set map)}
+\Output{Scan cost for answering $q$ when considering a merge of $rp$}
+	$r_{m}$ $\leftarrow$ Consolidate $rp$ \;
+	$L$ $\leftarrow$ Copy the range set mapped to $q$ from $M$\; 
+	Delete ranges in $rp$ from and insert $r_{m}$ into $L$. \\
+	$s$ $\leftarrow$ ``{\tt SELECT * FROM FACT$\_$TABLE WHERE }'' \;
+	\ForEach{range $r$ $\in$ $L$}{	
+		Restore a predicate $p$ from $r$. \\
+		Add $p$ to {\tt WHERE} clause of $s$.
+	}
+	Make an API call with $s$ to the server. \\
+	Extract the spool size (in bytes) from the result. \\
+	$blc$ $\leftarrow$ Calculate the block counts from the spool size. \\
+	Update $q$'s scan cost to $blc$. \\
+	{\bf return} $blc$ \;
+}
+\end{algorithm}
+
+If $q$ turns out to be affected by the merge of a range pair, 
+then the range set associated with $q$ in $M$ is temporarily updated 
+by removing the parent ranges and adding the merged one. 
+The altered range set is remapped to $q$ in $M$ and then 
+translated to the equivalent predicates, 
+so that {\it CP} for $s$ based on the predicates can be built.
+Unless the merge range pair influences $q$, then 
+{\it CP} can be simply built based on the 
+predicates restored from the existing ranges mapped to $q$.
+
+To help understand the scan cost construction, let us 
+consider a range pair ($rp$) of \hbox{$R$[2, 2]} and $R$[2, 3] in the running example. 
+$rp$ produces the merged range, \range{25, 35}, \hbox{under} {\tt LO\_QUANTITY}. 
+The merge by $rp$ affects both $q$1 and $q$2 in the workload.
+
+Thus, the range sets of $q$1 and $q$2 on {\tt LO\_QUANTITY} are 
+altered to \{\openrange{\infty, 25}, \range{25, 35}\} and \{\range{25, 35}\}, 
+respectively. Of course, no change is made to the existing range sets 
+of the queries on {\tt LO\_DISCOUNT}. 
+Therefore, the corresponding scan cost queries for $q$1 and $q$2 
+can be built as below:
+\begin{center}
+\begin{tabular}{rl}
+$\Sone$:	& {\tt SELECT * FROM LINEORDER l} \\
+		& {\tt WHERE ((l.LO\_DISCOUNT = 1)} \\
+		& {\tt OR (l.LO\_DISCOUNT >= 4} \\ 
+		& {\tt AND l.LO\_DISCOUNT <= 5))} \\
+		& {\tt AND l.LO\_QUANTITY <= 35}  \\ 
+$\Stwo$:	& {\tt SELECT * FROM LINEORDER l} \\
+        & {\tt WHERE l.LO\_DISCOUNT >= 7} \\
+        & {\tt AND (l.LO\_QUANTITY >= 25 AND} \\ 
+        & {\tt l.LO\_QUANTITY <= 35)}.
+\end{tabular}
+\end{center}
+
+It is possible to combine all the (consecutive) ranges in the (temporarily) altered set of $q$1, 
+and hence, a single predicate ({\tt l.LO\_QUANTITY <= 35}) can be built.
+
+The wizard sends such a scan cost query to the server via an API call, 
+and it extracts from the server's response the {\em spool size} (in bytes), or the size of scan query result. 
+Then, the tool computes as scan cost the {\em block counts} using the spool size. 
+If the merge of a range pair does not impact on any query in the workload, 
+the existing (previously computed) scan cost of queries 
+will be reused for the range pair. 
+That is, the wizard will only recompute 
+the scan cost of a query {\em iff} the query is affected by 
+the merge of two consecutive ranges.
+
+In this way, the weighted sum of scan cost of queries, $T_{s}$ can be computed 
+for each range pair. 
+$T_{s}$ can be defined as follows:
+\begin{center}
+\begin{tabular}{c} 
+$T_{s}$ = $\sum_{i=1}^{n}$($sc_{i}${$\cdot$}{$w_{i}$}),
+\end{tabular}
+\end{center}
+where $n$ is the number of queries in $Q$, $sc_{i}$ is the scan cost of a query $q_{i}$ in $Q$, and 
+$w_{i}$ denotes the (\hbox{non-negative}) weight associated with $q_{i}$.
+
+Once all range pairs are examined, the tool chooses the range pair 
+with the least $T_{s}$ for a merge. 
+If several range pairs end up with the same least $T_{s}$, 
+then the tool applies the heuristic of favoring the range pair that produces 
+fewer number of partitions when merged, to make 
+it faster to reach the partition count limit.
+
+In the example, suppose that the running range pair $rp$ produces the least $T_{s}$, 
+and thus, the tool merges $rp$. 
+Thus we can update both $R$ and $M$ as follows.
+
+$R$:
+\begin{center}
+\begin{tabular}{c|c} 
+{\tt LO\_DISCOUNT} & {\tt LO\_QUANTITY} \\ \hline
+\range{1, 1} & \openrange{\infty, 25} \\
+\range{4, 5} & \range{25, 35}		  \\
+\ropenrange{7, \infty} &  \\
+\end{tabular}
+\end{center}
+
+$M$:
+\begin{center}
+\begin{tabular}{ll} 
+$\langle${$q$1}, \{$R$[1, 1], $R$[1, 2], $R$[2, 1], $R$[2, 2]\}$\rangle$ \\
+$\langle${$q$2}, \{$R$[1, 3], $R$[2, 2]\}$\rangle$
+\end{tabular}
+\end{center}
+
+Note that we see that $q$2 may have the most customized partition based on the updated $M$. 
+Only the single partition formed by ($R$[1, 3]$\cap${$R$[2, 2]}) is sufficient 
+to retrieve all the qualifying 
+rows for $q$2; thus, the other partitions can be simply eliminated. 
+In the meantime, to answer $q$1, we need to read the four partitions formed by 
+the top two ranges of each field. 
+Unfortunately, because the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
+contain the non-qualifying rows for $q$1, $R$ cannot provide $q$1 with as much benefit as $q$2. 
+However, $R$ can be a good compromise to satisfy both queries, in that 
+the MLPPI derived by $R$ can potentially minimize the total execution cost 
+of $Q$.
+
+\begin{algorithm}[t]
+\caption{The Initial Phase}
+\label{algo:init_phase}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$M$ (query-to-range-set map), $R$ (input range set)}
+\Output{$R$ with \# partitions $\leq$ partition limit}
+	$W$ $\leftarrow$ Query weights \\
+	\While{(\# partitions by $R$ $>$ partition limit)}
+	{
+		\ForEach{a range pair ($rp$) in $R$}
+		{	 
+			$T_{s}$ $\leftarrow$ 0 \;
+			\ForEach{$q$ in $M$}
+			{	
+				$w$ $\leftarrow$ Get $q$'s weight from $W$ \\
+				$L$ $\leftarrow$ Get the range set mapped to $q$ in $M$ \\
+				\uIf(\tcp*[f]{Affected}) {(($rp$ $\cap$ $L$) $\neq$ $\varnothing$)}
+				{
+					$T_{s}$ += $w{\cdot}$({\it getSC}($q$,$rp$,$M$)) \tcp*[r]{See Algo.~\ref{algo:comp_scan_cost} regarding {\it getSC}()}
+				}
+				%\lElse{ $T_{s}$ += $w{\cdot}$($q$'s existing scan cost)}
+			}
+			Associate $T_{s}$ with $rp$.
+		}
+		Find $rp$(s) with the least $T_{s}$. \\
+		If a tie happens, select the $rp$ to make fewer partitions when consolidated. \\
+		Update $M$ and $R$ by the chosen $rp$.
+	}
+	{\bf return} $R$\;
+}
+\end{algorithm}
+
+Algorithm~\ref{algo:init_phase} represents the initial phase 
+that has been proposed so far. 
+The algorithm starts with a query-to-range set map ($M$) and an input range set ($R$). 
+Until the number of the partitions in $R$ falls below the partition limit threshold, 
+the algorithm determines the range pair yielding the least total execution cost ($T_{s}$) 
+and updates $M$ and $R$. At the end, the algorithm produces a final recommendation based on $R$.
+%\vspace{-.1in}
+
+Now that the total partition counts ($4${$\times$}$3$=$12$) by $R$ falls  
+below the assumed limit ($15$), the wizard 
+proceeds to the optimized phase with $R$ without repeating the initial phase.
+
+\subsection{The Optimized Phase}
+\label{sec:opt_phase}
+
+%We verified the effectiveness of the tool by running it on a workload based
+In this regard, using $R$ we have an initial MLPPI recommendation for the {\tt LINEORDER} fact table.
+The recommendation can be used sufficiently, but 
+having fewer partitions, induced by further merges, 
+can enhance the overall workload performance for the following reasons. 
+First, multiple file contexts by many partitions can incur 
+huge overhead that impacts on the query optimizer. 
+There also exists an operational threshold 
+that the optimizer can handle the maximum number of the partitions at a time.
+Therefore, further reducing partitions greatly helps the optimizer to manage these partitions.
+
+A similar algorithm, as suggested in the initial phase, 
+can be applied to this optimized phase. 
+Instead of the scan cost, the {\em query cost} can be used thanks to a feasible MLPPI. 
+The query cost is modeled as the estimated processing time 
+of a query on a ``faked'' fact table applying the MLPPI definition 
+reflecting the merge of a range pair. 
+This cost estimation technique is similar 
+to the ``\hbox{what-if}" approach~\cite{chaudhuri1998autoadmin}.
+
+\begin{algorithm}[t]
+\caption{Query Cost Computation}
+\label{algo:comp_query_cost}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$q$ (query), $rp$ (range pair), $M$ (query-to-range-set map), $R$ (input range set)}
+\Output{Query cost to answer $q$ when considering a merge of $rp$}
+	Create an empty shadow table $H$ having the same definition as the fact table $F$, 
+	including indexes and all constraints (check and referential integrity constraints). \\
+	Propagate all (field and index) statistics of $F$ to $H$. \\
+	\If{$F$ has materialized views (MVs)}{
+		Create the equivalent MVs on $H$. \\
+		Propagate the statistics of the MVs on $F$ to those of $H$. \\
+	}
+	$r_{m}$ $\leftarrow$ Merge $rp$ \;
+	$L$ $\leftarrow$ Copy ranges associated with $q$ from $M$\; 
+	Remove ranges in $rp$ from and add $r_{m}$ to $L$. \\
+	$R'$ $\leftarrow$ Update $R$ with $L$ \\
+	Alter $H$ using a fictitious MLPPI with $R'$. \\
+	Construct and send to the server an $H$-based shadow query $h$ equivalent to $q$. \\	
+	$ept$ $\leftarrow$ Estimated processing time of $h$ extracted from the server response \;
+	Update $q$'s query cost to $ept$. \\
+	{\bf return $ept$} 
+}
+\end{algorithm}
+
+Algorithm~\ref{algo:comp_query_cost} illustrates the steps for computing 
+the query cost of a query $q$, 
+given a range pair ($rp$) affecting $q$ when merged. 
+First, we create an empty shadow table $H$ with the same definition as $F$, 
+including all indexes and constraints such as check and referential integrity 
+ones. Next, we propagate all statistics of $F$ to $H$. 
+This covers field and index statistics. 
+If $F$ has materialized views (MVs), then we create the equivalent MVs on $H$ and subsequently, 
+propagate the statistics of the MVs on $F$ to the new MVs on $H$. 
+After that, we alter $H$ by an MLPPI with $R$' applying the merge of $rp$. 
+The wizard then builds a {\em shadow} query $h$ equivalent to the input query $q$ 
+but replacing $F$ in $q$ with $H$, and makes an API call to the {\tt EXPLAIN} tool with $h$. 
+Eventually, the tool extracts from the result and returns the estimated elapsed time 
+of $h$ as the query cost of $q$. 
+As done in the initial phase, the query cost of a query is recomputed only if the merge affects the query.
+
+For each range pair the wizard computes the weighted sum ($T_{q}$) of query cost of queries. 
+$T_{q}$ can be similarly defined as $T_{s}$, described in Section~\ref{sec:init_phase}.
+%and thus the definition of $T_{q}$ is omitted.
+Let the current least $T_{q}$ be $T$ and the existing least $T_{q}$ be $T_{p}$. 
+If $T$ $<=$ $T_{p}$, then for the next iteration $T_{p}$ is updated to $T$. 
+In turn, the tool merges the range pair that produces $T$ and updates $M$ and $R$ along with the merge. 
+The wizard repeats the process until 1) no ranges in $R$ remain, or 2) pre-defined iterations are reached. 
+The optimized phase is finished if $T$ $\geq$ $T_{p}$.
+
+\begin{figure}[t]
+\begin{center}
+\begin{tabular}{|l|} \hline
+{\tt ALTER TABLE LINEORDER} \\
+{\tt MODIFY PRIMARY INDEX} \\
+{\tt PARTITION BY(} \\
+\hspace{.1in}{\tt CASE\_N(} \\
+	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 1 AND LO\_DISCOUNT $\leq$ 5,} \\
+	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 7,} \\ 
+	\hspace{.2in}{\tt NO CASE OR UNKNOWN),} \\
+\hspace{.1in}{\tt CASE\_N(} \\
+	\hspace{.2in}{\tt LO\_QUANTITY $<$ 25,} \\
+	\hspace{.2in}{\tt LO\_QUANTITY $\geq$ 25 AND LO\_QUANTITY $\leq$ 35,} \\
+	\hspace{.2in}{\tt NO CASE OR UNKNOWN)} \\ 
+{\tt );} \\ \hline
+\end{tabular}
+\end{center}
+\caption{An MLPPI Recommendation for Workload $Q$\label{fig:mlppi}}
+\end{figure}
+
+In the running example, suppose that in the first round, ranges $R$[1, 1] and $R$[1, 2] are chosen for a merge. 
+Then, the wizard produces a merged range \range{1, 5} \hbox{under} {\tt LO\_DISCOUNT} and proceeds to the next round. 
+If a range pair selected for the subsequent merge fails to improve $T_{p}$, 
+then, the tool exits the optimized phase and \hbox{produces} the final MLPPI recommendation for the given workload $Q$ 
+as shown in Figure~\ref{fig:mlppi}.
+
+Algorithm~\ref{algo:opt_phase} exhibits the proposed optimized phase, 
+which also applies the same heuristic to favor a range pair 
+that produces fewer partitions to break a tie.  
+The order of the different partitioning levels may have some impact on 
+the execution time. A range condition on lower levels like the query 
+in Section~\ref{sec:intro} 
+``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}'' 
+requires scanning non-consecutive partitions. 
+This may \hbox{incur} some overhead at run time. 
+To reduce this \hbox{effect}, we sort the partition \hbox{levels} based on number of 
+partitions in descending order \hbox{before} making a final \hbox{recommendation}. 
+The solution in \hbox{Figure}~\ref{fig:exam} follows this heuristic 
+making the two-partition case in the first level 
+followed by the four-partition case in the second level. 
+The order in Figure~\ref{fig:mlppi} can go \hbox{either} way 
+since both levels have the same number of \hbox{partitions}.
+
+\begin{algorithm}[t]
+\caption{The Optimized Phase}
+\label{algo:opt_phase}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$M$ (query-to-range-set map), $R$ (input range set)}
+\Output{MLPPI such that the total query cost is minimal}
+	$W$ $\leftarrow$ Query weights \\
+	$T_{p}$ $\leftarrow$ Total query cost sum on an initial MLPPI by $R$ \\
+	\While{(Pre-defined iterations or $R$ $\neq$ $\varnothing$)}{
+		\ForEach{range pair $rp$ in $R$}{	 
+			$T_{q}$ $\leftarrow$ 0 \;
+			\ForEach{$q$ in $M$}{	
+				$w$ $\leftarrow$ Get $q$'s weight from $W$ \\
+				$L$ $\leftarrow$ Get the range set mapped to $q$ in $M$ \\
+				\uIf(\tcp*[f]{Affected}) 
+				{(($rp$ $\cap$ $L$) $\neq$ $\varnothing$)}{
+					$T_{q}$ += $w${$\cdot$}({\it getQC}($q$,$rp$,$M$,$R$)) 	
+					\tcp*[r]{See Algo.~\ref{algo:comp_query_cost} regarding {\it getQC}()}				
+				}
+				\lElse{ $T_{q}$ += $w${$\cdot$}($q$'s existing query cost)}
+			}
+			Map $T_{q}$ to $rp$.
+		}
+		$T$ $\leftarrow$ The least $T_{q}$ that has been so far seen.  \\
+		\lIf{$T$ $>$ $T_{p}$}{ {\bf break} \;}
+		\Else{
+			$T_{p}$ $\leftarrow$ $T$ \tcp*[r]{Query Cost Sum Update} 
+			Find $rp$(s) with $T$. \\
+			If a tie happens, select the $rp$ to make fewer partitions when consolidated. \\
+			Update $M$ and $R$ by the chosen $rp$. \\
+		}
+	}
+	{\bf return} an MLPPI by $R$ \;
+}
+\end{algorithm}
+
+\vspace{-.1in}
+%\vspace\fill
+
+\section{Analysis}
+\label{sec:analysis}
+
+In this section we analyze the time complexity of 
+the proposed preprocessing, initial, and optimized phases.
+For the time complexity we compute the logical running time of the wizard, using the {\em number of API calls} 
+made to the server during the phases.
+
+%In this section the complexity of the proposed algorithms 
+%is analyzed.
+
+\subsection{The Preprocessing Phase}
+
+\begin{lemma}
+\label{lemma:pre_phase}  
+The running time complexity for building a query-to-range map is 
+$\BigO$($N${$\cdot$}$C${$\cdot$}$V$), where $N$ is the number of queries 
+in a workload, $C$ is the total number of fields in the fact table, 
+and $V$ is the max number of values in a field. 
+\end{lemma}
+
+{\bf Proof}. Let us consider a worst case such that given a workload of 
+$N$ queries, each query references all the fields of the fact table, and 
+each field is associated with at most $V$ values by the predicates of the queries. 
+Each query may have a single-value range. 
+A range consisting of only a single value per field can be mapped to each query. 
+Thus, the running time complexity for the map construction 
+is $\BigO$($N${$\cdot$}$C${$\cdot$}$V$). \qed
+
+{\bf Comment:} In practice, our experiments showed that the bound 
+$\BigO$($N${$\cdot$}$C${$\cdot$}$V$) is overly pessimistic.
+
+\subsection{The Initial and Optimized Phases}
+
+\begin{lemma}
+\label{lemma:init_phase}  
+The running time complexity for the initial or optimized phases is 
+$\BigO$($M^{2}${$\cdot$}$N$), 
+where $M$ is the number of range pairs, and $N$ is the number of queries in 
+a given workload. 
+\end{lemma}
+
+{\bf Proof}. Suppose that the merge of every range pair influences all the queries. 
+Every iteration, either the scan or query costs need to be recomputed 
+for each query. 
+In the first round the \hbox{re-computation} cost is paid as many times as $M${$\cdot$}$N$, 
+and a chosen range pair is merged. In the subsequent round 
+$M$-1 range pairs remains, and the following cost amounts to ($M$-1){$\cdot$}$N$. 
+In the worst case we may end up consolidating all range pairs, thus having no partition on 
+a target table. The total calls to the server can be increased up to
+
+\begin{center}
+$M${$\cdot$}$N$+($M$-1){$\cdot$}$N$+$\cdots$+$N$ = ($\sum_{i=1}^M i$){$\cdot$}$N$ = $\frac{M(M-1)}{2}${$\cdot$}$N$. 
+\end{center}
+
+\noindent Hence, the running time complexity is $\BigO$($M^{2}${$\cdot$}$N$). $\qed$
+
+%\vspace{0.1in}
+
+Our experiments showed that $\BigO$($M^{2}${$\cdot$}$N$), the running time complexity, 
+is overly pessimistic. 
+In \hbox{practice}, the number of calls made was much fewer than the \hbox{theoretic} bound, 
+since it was very rare for a range pair to be associated with all queries in a workload. 
+In \hbox{Section}~\ref{sec:stat} we will show how many API calls take place on the workloads 
+used for evaluation.
+
+\section{Experiment}
+\label{sec:experiment}
+
+In this section, we describe our environment settings, 
+report the statistics measured in our experiments, and 
+demonstrate the performance of the MLPPI wizard (WIZARD), compared with 
+that of no partitioning and the partitioning by a human expert (EXP).
+
+\subsection{Environment Settings}
+\label{sec:env_set}
+
+The MLPPI wizard was implemented as a prototype 
+on top of the Teradata DBMS server. 
+It was written in Java. 
+The performance of the wizard was evaluated 
+on the Teradata DBMS server machine running Unix.
+
+When it comes to workload generation, we took advantage of 
+our simple star schema query \hbox{generator}. 
+The generator assumes that 1) available operators, 2) fields, 
+and 3) the minimum and \hbox{maximum} values of the fields are already known. 
+For each query, the generator first randomly selects the number of single-table 
+predicates to create. 
+Then, it arbitrarily chooses a field and a specific operator 
+for each predicate. 
+If {\tt IN} operator on a chosen field is picked up, the query generator 
+determines the number of values to add to the {\tt IN} list and then chooses random values within the min-max value 
+range of the field. 
+In this way the generator builds a query involving the generated predicates.
+
+A generated query is based on a template that joins {\tt LINEORDER} and {\tt DDATE} with constraints defined by the predicates. 
+This template is common in customer cases like reports and form 
+templates. 
+Using the query generator we generated two workloads consisting of 10 queries (10Q) and 20 queries (20Q).
+
+For the query workloads we populated the fact table with a scale of 1TByte (1TB) and 3TBytes (3TB).
+
+\begin{table*}[t]
+\caption{The statistics obtained on the used workloads\label{tab:workload_stat}}
+\begin{center}
+{\scriptsize
+\begin{tabular}{|l|l|l|l|l|} \hline
+		   			   							   & \multicolumn{2}{|c|}{Init. Phase} & \multicolumn{2}{|c|}{Opt. Phase} \\ \cline{2-5}
+		   			   							   & 10Q & 20Q & 10Q & 20Q \\ \hline
+{\# Input Partitions}  							   & 1,008,000 & 110,739,200 & 51,840 & 59,520 \\ \hline
+{\# Input Range Pairs} 							   & 441 & 4,180 & 24 & 48\\ \hline
+{\# Total Iterations}  							   & 14 & 55 & 1 & 1 \\ \hline
+{\# Total API Calls (Worst Calls by Lemma~\ref{lemma:init_phase}})   & 1,305 (1,944,810) & 31,915 (349,448,000) & 78 (5,760) & 388 (46,080) \\ \hline
+{\# Average API calls per Range Pair} 		  	   & 2.96 & 7.63 & 3.25 & 8.08 \\ \hline
+{\# Average number of range pairs compared per iteration} 	   & 31.5 & 76.0 & 24 & 48 \\ \hline
+{\# Maximum number of range pairs compared per an iteration} & 38 & 103 & 24 & 48 \\ \hline
+{\# Minimum number of range pairs compared per an iteration} & 25 & 49 & N/A & N/A\\ \hline
+%{\# max. API Calls made per range pair} 		   & 4.92 & 16.14 & & \\ \hline
+%{\# min. API Calls made per range pair} 		   & 1.92 & & & \\ \hline
+{\# Average API calls made per iteration}		   & 93.21 & 580 & 78 & 388 \\ \hline
+{\# Maximum number of API calls made per an iteration} 	   & 123 & 791 & 78 & 388 \\ \hline
+{\# Minimum number of API calls made per an iteration} 	   & 73 & 381 & N/A & N/A \\ \hline
+\end{tabular}
+}
+\end{center}
+\vspace{-0.3in}
+\end{table*}
+
+\subsection{Statistics Report}
+\label{sec:stat}
+
+Table~\ref{tab:workload_stat} summarizes the WIZARD's execution statistics, 
+which were obtained while the recommendations for the \hbox{workloads} were produced. 
+The statistics includes input partitions, 
+range pairs, number of API calls made, and total iterations 
+observed in each phase.
+
+About 1 million partitions were initially derived for 10Q whereas roughly 0.11 billion partitions for 20Q. 
+Although the workload size doubled, the total number of partitions was increased exponentially. 
+Much more predicates with different bindings produced two orders of magnitude more ranges in 20Q than those of 10Q. 
+While the huge input partitions were made, the number of collected range pairs per field was relatively small.
+
+The total number of iterations in the initial phases 
+tended to be proportional to the input \hbox{partitions} 
+\hbox{generated} from the workloads. 
+Note that there \hbox{happened} only one iteration in the optimized phase. 
+The reason was that since the partitions produced by the \hbox{initial} phase were customized enough, 
+no more \hbox{iterations} were \hbox{necessary} for further optimization.
+
+As mentioned in Section~\ref{sec:analysis}, the total calls made to 
+the server per workload were by far fewer than the theoretical bound shown in parentheses. 
+The worst call counts were calculated regarding the number of queries and range pairs observed in each phase. 
+The average number of calls per range pair was limited within 10. %across workloads and phases.
+
+\begin{figure*}[t]
+\vspace{-.2in}
+\centering
+\subfigure[1TB]{
+	\includegraphics[scale=0.6]{1T_total_time}
+	\label{fig:1T_eval_res}
+}
+\subfigure[3TB]{
+	\includegraphics[scale=0.6]{3T_total_time}
+	\label{fig:3T_eval_res}
+}
+\vspace{-0.1in}
+\caption{Total Elapsed Time\label{fig:total_elapsed_time}}
+\end{figure*}
+
+\begin{figure*}[t]
+\vspace{-.2in}
+\centering
+\subfigure[1TB]{
+	\includegraphics[scale=0.6]{1T_quality}
+	\label{fig:1T_quality}
+}
+\subfigure[3TB]{
+	\includegraphics[scale=0.6]{3T_quality}
+	\label{fig:3T_quality}
+}
+\vspace{-0.1in}
+\caption{Recommendation Quality Comparison\label{fig:recomm_qual}}
+\vspace{-.1in}
+\end{figure*}
+
+\subsection{Performance Evaluation}
+
+The main focus of our demonstration is to see the performance (quality) 
+of MLPPI recommendations made by the MLPPI wizard (WIZARD), 
+compared with the performance of no partitioning (NO PPI) and EXP (partitioning 
+by a human expert).
+
+%Our evalution results are exhibited in Figures ~\ref{fig:eval_res} 
+%and ~\ref{fig:improvement}.
+%Figure~\ref{fig:eval_res} shows the times (in seconds) taken 
+%to run the 10Q or 20Q workloads over the 1TB or 3TB fact tables. 
+%Based on the run results, Figure~\ref{fig:improvement} displays 
+%how much improvement could be achieved by the EXP or the WIZARD 
+%recommendations against the NO PPI one.
+
+Figures~\ref{fig:total_elapsed_time} and \ref{fig:recomm_qual} exhibit our evaluation results.
+Figure~\ref{fig:total_elapsed_time} shows the times (in seconds) taken 
+to run the 10Q and 20Q workloads over the 1TB and 3TB fact tables. 
+Based on the run results, Figure~\ref{fig:recomm_qual} shows 
+how much improvement was gained by the EXP and WIZARD solutions, compared with the NO PPI solution.
+
+Overall, the WIZARD recommendations were very effective at partitioning 
+the fact tables by leveraging the predicates in the workloads. 
+On average, the \hbox{total} execution time of the workloads on the WIZARD solutions was 
+about 3.5 and 2 times faster than those of NO PPI and EXP, respectively. 
+Also, the quality of the \hbox{WIZARD} solutions outperformed that of the EXP solutions; 
+the WIZARD solutions yielded an average of about 77\% improvement on the NO PPI solutions.
+
+Furthermore, the WIZARD recommendations 
+scaled very well with growing workload size and table size, 
+as illustrated in Figure~\ref{fig:recomm_qual}. 
+Although we doubled the workload scale from 10Q to 20Q, 
+the performance improvement of the WIZARD solution stayed the same in Figure~\ref{fig:1T_quality} 
+and got almost negligibly decreased in Figure~\ref{fig:3T_quality}. 
+This result showed that our WIZARD \hbox{solution} was scalable over increasing workload size.
+
+We also increased the table scale from 1TB to 3TB. 
+For 10Q the performance improvement (about 77\%) was almost the same over the increasing scales, 
+and for 20Q it was decreased from 76\% to 74\%, which was \hbox{almost} negligible.
+The WIZARD partitioning \hbox{recommendation} scaled well with increasing table size.
+
+In sum, our experiment results attest the effectiveness of the proposed algorithms for WIZARD 
+and show the superiority of the recommendations of WIZARD.
+
+\shorten{Regarding the scale of 1TB in Figure~\ref{fig:1T_quality}, 
+although we doubled the workload scale (from 10Q to 20Q) 
+the performance improvement of the WIZARD solution compared with that of NO PPI was not decrease although we doubled the workload scale (from 10Q to 20Q).
+In Figure~\ref{fig:3T_quality}, for the 3TB scale the improvement of WIZARD was slightly decreased, but it was negligible.
+In sum, the performance of the WIZARD solution was scalable over increasing workload and table scale.
+
+%Also, we scaled up the table from 1TB to 3TB on the fixed workload, but 
+%the \hbox{recommendation} quality was not degraded at all. 
+These results attest the effectiveness of our algorithms, 
+and show the superiority of our recommendations.}
+
+\section{Related Work}
+\label{sec:related_work}
+
+Physical database design~\cite{finkelstein1988physical,labio1997physical,rozen1991framework,Zilio1998} 
+has been discussed in \hbox{academic} research and industrial sectors in the past years.
+The major DBMS vendors (e.g., IBM, MS, and Oracle) have driven much of the work. 
+Their specific interests have been in automating the physical design for table 
+partitioning~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated,sheet2009oracle,saa2007oracle,rao2002automating}, indexes/materialized views~\cite{agrawal2000automated,Agrawal04:sqlserver,dash2011cophy,Kimura11,itwsqlserver,valentin2000db2,zilio2004recommending},
+and integration~\cite{Zilio04:db2design}.
+
+%Lightstone's work~\cite{Lightstone04:db2auto} introduces the MDC (Multidimensional 
+%Clustering) Advisor that automates the selection of MDC 
+%keys customized for a specific workload on DB2 Universal Databases to use MDC structure. 
+%DB2 Advisor~\cite{Zilio04:db2design} is an integrated commercial 
+%tool that captures a variety of aspects of physical database design 
+%by automation in a comprehensive way. 
+%Oracle Partitioning Advisor~\cite{sheet2009oracle} for recommending 
+%table partitionings is  integrated into 
+%the SQL Access Advisor~\cite{saa2007oracle}.
+
+%Since data clustering significantly influences on the performance of 
+%multidimensional queries, the MDC advisor helps DBAs that may have 
+%nontrivial troubles in finding proper clustering dimensions and 
+%granularity. 
+%To enumerate and determine possible clustering dimensions, 
+%his work relies on {\it what-if} query cost modeling, data sampling and a search algorithm.
+
+%DB2 Advisor~\cite{Zilio04:db2design} is an integrated commercial 
+%tool that captures a variety of aspects of physical database design 
+%by automation in a comprehensive way. 
+%It makes an attempt to cover a variety of design choices-partitioning, 
+%MDC, materialized views and indexes. 
+%DB2 Advisor leverages the DB2 query optimizer for yielding a recommendation, and 
+%takes a hybrid searching approach for considering different features.
+
+%Oracle has a tool, called Partitioning Advisor~\cite{sheet2009oracle}, 
+%to make a partitioning recommendation for tables. 
+%The tool is now integrated into 
+%the SQL Access Advisor~\cite{saa2007oracle} that recommends indexes and materialized 
+%views. 
+%However, its technical details are not yet released in public. 
+%
+%Also, MS SQL Server provides some similar, automated tools~\cite{agrawal2000automated,agrawal04:integrating,
+%itwsqlserver} for table partitioning, materialized views or indexes fitting to a given workload.
+%The optimizer cost model is exploited to acquire estimates for 
+%candidate materialized views or indexes. 
+%In particular, Agrawal's work~\cite{agrawal04:integrating} presents 
+%an approach of integrating vertical and horizontal partitioning schemes 
+%into their physical database design, considering scalability.
+
+Some of the tools in IBM DB2~\cite{Lightstone04:db2auto}, 
+Oracle~\cite{sheet2009oracle}, and MS SQL Server~\cite{agrawal04:integrating} 
+appear to be similar to our MLPPI wizard. 
+However, in light of problem scope and approach, the wizard 
+is fundamentally different from the existing tools excluding Oracle Partitioning \hbox{Advisor}~\cite{sheet2009oracle} 
+providing no published technical details.
+
+DB2 MDC Advisor~\cite{Lightstone04:db2auto} actually tackles 
+a different problem of automatically recommending the most well-suited MDC 
+keys for a given workload. 
+Agrawal's work~\cite{agrawal04:integrating} 
+discusses another problem of merging \hbox{single-level} range partitionings on objects such as tables and indexes. 
+This article addresses the multi-level \hbox{partitioning} problem. 
+Hence, the existing solutions cannot be directly applied to our MLPPI wizard.
+
+Regarding the approach, DB2 MDC Advisor~\cite{Lightstone04:db2auto} uses 
+the search space driven by {\em fields}. In contrast, our search 
+space is driven by query-predicates, which is superior because 
+\hbox{predicates} are more customized and specific to a workload than fields. 
+Agrawal's horizontal partitioning scheme~\cite{agrawal04:integrating} 
+also uses the search space driven by simple range predicates, but 
+his technique has a \hbox{shortcoming} as follows. 
+His work produces a solution for each individual query and attempt 
+to merge the solutions. But this approach cannot reach 
+an optimized solution in a global \hbox{perspective}.
+Our wizard generates the whole search space upfront and in turn merges partitions, leading to a globally-optimized \hbox{solution}. 
+Moreover, only a single column is considered in his work~\cite{agrawal04:integrating}, 
+whereas our tool deals with multiple fields.
+
+Implementations of previous tools~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} \hbox{required} instrumentation for optimizer code.
+These instrumentations are needed to facilitate the required 
+information for the physical design tools API calls. 
+The \hbox{instrumentation} code need to be enhanced and tested for new database 
+releases that add complexity and additional cost for software upgrades. 
+The Teradata optimizer has a rich set of existing APIs originally coded 
+for system and workload management tools. 
+The APIs are sufficient to avoid the costly optimizer code change.
+
+Also, Nehme's work~\cite{nehme2011automated}, deeply-integrated with 
+optimizer, reveals a concern about the quality of the recommendations 
+made by some tools, shallowly integrated with optimizer. 
+But we observed in our experiments that the quality of 
+the wizard's solutions was much superior to that of the base solutions. 
+In addition, some might argue that in our loosely-coupled approach 
+the cost to invoke the optimizer might be significant. 
+But the measured call counts were much fewer than 
+the theoretical bound, since most calls were made only when queries were affected by a merge.  
+%If the remote communication cost is expensive, the tool can also be run, 
+%{\em co-located} with the server.
+
+%The existing work requires their database engine to be extended 
+%to support their features. 
+%This strong coupling may impose a nontrivial burden on their servers whenever 
+%a new feature of the existing tools is desired. 
+%As illustrated in Figure~\ref{fig:arch}, our tool is completely outside 
+%the DBMS server. Thus, we are free of the constraint. 
+%A literature~\cite{nehme2011automated} claims that 
+%such loosely-coupled approaches might suffer from the reduced performance of a tool 
+%and the poor quality of the partitioning recommendations. 
+%The powerful Teradata APIs, however, help the wizard receive from the server 
+%everything needed for yielding a MLPPI recommendation. 
+%Moreover, as demonstrated in Figure~\ref{fig:perf_eval}, the quality of the 
+%partitioning solutions by the wizard was also much superior to that of the base solutions.
+%Some might argue that the cost of invoking the optimizer could be significant as well. 
+%Again, the actual calls are mainly made only when queries are affected by a merge. 
+%As a result, they were much fewer than 
+%the theoretical bound as shown in Table~\ref{tab:workload_stat}.  
+%To further save the remote communication cost, the tool can also be run, 
+%{\em co-located} with the server. 
+%Therefore, we believe that the shallowly-integrated strategy perfectly matches 
+%with the wizard.
+
+Lastly, there has been some previous work~\cite{nehme2011automated,rao2002automating,tatarowicz2012lookup} 
+regarding table partitioning in multi-node \hbox{systems}, but 
+our problem is discussed in the context of a single node system, as in the existing work~\cite{agrawal04:integrating}. 
+Database cracking~\cite{idreoskm07}
+%, self-reorganizing column values in accordance with query workload, 
+assumes a single node environment, but 
+it does not address our multi-level partitioning problem.
+
+\section{Summary}
+\label{sec:conclusion}
+%We verified the effectiveness of the tool by running it on a workload based
+
+Given workloads, it is difficult for DBAs to select appropriate fields 
+in partitioning the fact table due to large search space. 
+DBAs cannot easily determine how granular partitions should be made 
+for the workloads.
+
+To address this concern, we presented the MLPPI wizard to recommend a fact-table partitioning for a given star schema workload. 
+Since \hbox{query-predicates} are exploited 
+to capture necessary ranges for fields and define partitions, 
+the MLPPI \hbox{recommendation} can be very optimized, customized to the workload.
+
+We proposed the wizard's algorithms consisting of the three phases. 
+The wizard incrementally reduced its search space by merging 
+the range pair with the least scan or query cost, and eventually reached an MLPPI recommendation. 
+In addition, we analyzed the running complexity for the initial and optimized phases. 
+Despite the theoretically high bound, in practice, the wizard made much fewer calls in the phases.
+
+We measured the performance of the recommendation by the wizard using 
+our workloads.  
+We demonstrated that the produced MLPPI solutions by 
+the wizard could reduce the total elapsed time by more than a factor of two, 
+compared with those of no partitioning approach and partitioning 
+done by a human expert.
+
+\section*{Acknowledgments} 
+This work was carried out while the first author visited Teradata. 
+The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
+%\clearpage
+%\vspace\fill
+
+%1.
+Agrawal, R., Srikant, R.: Fast algorithms for mining association rules.
+In: Proceedings of the 20th International Conference on Very Large Databases.
+pp.
+487–499.
+Morgan Kaufmann, San- tiago, Chile (1994)
+%2.
+Bruce,K.B.,Cardelli,L.,Pierce,B.C.:Comparingobjectencodings.In:Abadi,M.,Ito,T.(eds.) Theoretical Aspects of Computer Software, Lecture Notes in Computer Science, vol.
+1281, pp.
+415–438.
+Springer-Verlag, Berlin Heidelberg New York (1997)
+%3. Garcia-Molina, H., Ullman, D.J., Widom, J.: Database Systems: The Complete Book. Prentice Hall, New Jersey, USA (2002)
+%4. van Leeuwen, J. (ed.): Computer Science Today. Recent Trends and Developments, Lecture Notes in Computer Science, vol. 1000. Springer-Verlag, Berlin Heidelberg New York (1995)
+%5. Ribicˇre,M.,Charlton,P.:Ontologyoverview.MotorolaLabs,Paris(2002),[Online].Available:
+%http://www.fipa.org/docs/input/f-in-00045/f-in-00045.pdf (current October 2003)
+%6. Wang,X.,Bettini,C.,Brodsky,A.,Jajoida,S.:Logicaldesignfortemporaldatabaseswithmul-
+%tiple granularities. ACM Transactions on Database Systems 22(2), 115–170 (1997)
+
+%\begin{thebibliography}{99}
+%\expandafter\ifx\csname url\endcsname\relax
+%  \def\url#1{\texttt{#1}}\fi
+%\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+%\expandafter\ifx\csname href\endcsname\relax
+%  \def\href#1#2{#2} \def\path#1{#1}\fi
+%
+%{
+%\bibitem{Suh12}
+%{Y.-K.~Suh, A.~Ghazal, A.~Crolotte, and P.~Kostamaa}, {A New Tool for Multi-level Partitioning in Teradata}, 
+%in: CIKM'12, pp. 2214--2218.
+%  
+%\bibitem{sinclair:ppi}
+%P.~Sinclair, {Using PPIs to Improve Performance},
+%  \url{http://www.teradata.com/tdmo/v08n03/pdf/AR5731.pdf}
+%  (accessed August 27, 2015).
+%
+%\bibitem{klindt09mlppi}
+%J.~Klindt, {Single-level and Multilevel Partitioned Primary \linebreak Indexes},
+%  \url{http://www.teradata.com/white-papers/Single-level-and-}\linebreak\url{Multilevel-Partitioned-Primary-Indexes-eb1889/}
+%  (accessed July 27 2015).
+%
+%\bibitem{oneil:ssb}
+%{P.~O’Neil, B.~O'Neil, and X.~Chen}, {The Star Schema Benchmark (SSB)},
+%  \url{http://www.cs.umb.edu/~poneil/StarSchemaB.PDF} 
+%  (accessed February 28, 2017).
+%
+%\bibitem{agrawal04:integrating}
+%{S.~Agrawal, V.~Narasayya, and B.~Yang}, {Integrating Vertical and Horizontal Partitioning into Automated Physical Database Design}, 
+%in: \hbox{SIGMOD} '04, pp. 359--370.
+%
+%\bibitem{Lightstone04:db2auto}
+%{S.~Lightstone and B.~Bhattacharjee}, {Automated Design of
+%  Multi-dimensional Clustering Tables for Relational Databases}, in:
+%  VLDB '04, pp. 1170--1181.
+%
+%\bibitem{nehme2011automated}
+%{R.~Nehme and N.~Bruno}, {Automated Partitioning Design in Parallel
+%  Database Systems}, in: SIGMOD '11, pp. 1137--1148.
+%
+%\bibitem{chaudhuri1998autoadmin}
+%{S.~Chaudhuri and V.~Narasayya}, {AutoAdmin `What-If' Index Analysis Utility}, ACM SIGMOD Rec. 27~(2) (1998) 367--378.
+%
+%\bibitem{finkelstein1988physical}
+%{S.~Finkelstein, M. Schkolnick, and P. Tiberio}, {Physical Database Design for Relational Databases}, ACM Trans. on Datab. Syst. (TODS), 13~(1) (1988) 91--128.
+%
+%\bibitem{labio1997physical}
+%{W.~Labio, D.~Quass, and B.~Adelberg}, {Physical Database Design for Data Warehouses}, in: ICDE '97, pp. 277--288.
+%
+%\bibitem{rozen1991framework}
+%{S.~Rozen and D.~Shasha}, {A Framework for Automating Physical Database Design}, in: VLDB '91, pp. 401--411.
+%
+%\bibitem{Zilio1998}
+%{C.~Zilio}, {Physical Database Design Decision Algorithms and Concurrent
+%  \hbox{Reorganization} for Parallel Database Systems}, Ph.D. thesis,
+%  Dept. of CS, Univ. of Toronto (1998).
+%
+%\bibitem{sheet2009oracle}
+%{Oracle}, {Oracle White Paper---Partitioning with \hbox{Oracle} Database 11g
+%  \hbox{Release 2}},
+%  \url{http://www.oracle.com/technetwork/database/options/partitioning/twp-partitioning-11gr2-2009-09-130569.pdf}
+%  (accessed June 23, 2014).
+%
+%\bibitem{saa2007oracle}
+%{Oracle}, {SQL Access Advisor},
+%  \url{http://docs.oracle.com/cd/B19306_01/server.102/b14211/advisor.htm}
+%  (accessed March 30, 2012).
+%
+%\bibitem{rao2002automating}
+%{Jun Rao, Chun Zhang, Nimrod Megiddo, and Guy Lohman}, {Automating
+%  Physical Database Design in A Parallel Database}, in: SIGMOD '02, pp. 558--569.
+%
+%\bibitem{agrawal2000automated}
+%{S.~Agrawal et. al.}, {Automated Selection of Materialized Views and Indexes in SQL Databases}, in:
+%  {VLDB '00}, pp. 496--505.
+%
+%\bibitem{Agrawal04:sqlserver}
+%{S.~Agrawal et. al.}, {Database Tuning Advisor for Microsoft
+%  SQL Server 2005: Demo}, in: \hbox{SIGMOD} '05, pp. 930--932.
+%
+%\bibitem{dash2011cophy}
+%{D.~Dash}, {CoPhy: A
+%  Scalable, Portable, and Interactive Index Advisor for Large Workloads},
+%  in: PVLDB, 4~(6) (2011) 362--372.
+%
+%\bibitem{Kimura11}
+%{H.~Kimura et. al.}, {Compression Aware
+%  \hbox{Physical} Database Design}, in: PVLDB, 4~(10) (2011) 657--668.
+%
+%\bibitem{itwsqlserver}
+%{Microsoft SQL Server 2000}, {Index Tuning Wizard SQL Server 2000},
+%  \url{http://technet.microsoft.com/en-us/library/cc966541.aspx} 
+%  (accessed on March 21, 2014).
+%
+%\bibitem{valentin2000db2}
+%{G.~Valentin}, {DB2 Advisor: An Optimizer Smart Enough to Recommend Its Own
+%  Indexes}, in: {ICDE '00}, pp. 101--110.
+%
+%\bibitem{zilio2004recommending}
+%{D.~Zilio et. al.},
+%  {Recommending Materialized Views and Indexes with the IBM DB2 Design
+%  Advisor}, in: ICAC '04, pp. 180--188.
+%
+%\bibitem{Zilio04:db2design}
+%{D.~Zilio et. al.}, {DB2 Design Advisor:
+%  Integrated Automatic Physical Database}, in: VLDB '04, pp. 1087--1097.
+%
+%\bibitem{tatarowicz2012lookup}
+%{A.~Tatarowicz et. al.}, {Lookup Tables: Fine-grained Partitioning for Distributed Databases},
+%  in: ICDE'12, pp. 102--113.
+%
+%\bibitem{idreoskm07}
+%{S.~Idreos et. al.}, {Database Cracking}, in: CIDR'07, pp. 68--78.
+%}
+%
+%\end{thebibliography}
+
+\bibliographystyle{splncs03}
+\bibliography{paper}
 

### 2017-03-01 pt_characterization/data/emp_et_hist.r
New data arrangement

@@ -1,0 +1,283 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_et_hist.eps")
+xmax <- max(x$METIME)
+xmin <- min(x$METIME)
+h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+binsize=1
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+#xaxtl <- seq(xmin-1,xmax+1,by=1)
+#axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_et_hist.eps")
+xmax <- max(x$METIME)
+xmin <- min(x$METIME)
+h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+binsize=1
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-2,xmax+2),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+#xaxtl <- seq(xmin-1,xmax+1,by=1)
+#axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_et_hist.eps")
+xmax <- max(x$METIME)
+xmin <- min(x$METIME)
+h = hist(x$METIME, right=F,breaks=xmax-xmin+1,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+binsize=1
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-5,xmax+5),col="blue", main='ET frequency on INC4', 
[…]

### 2017-03-01 pt_characterization/data/emp_pt_hist.r

@@ -1,0 +1,199 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-1,max(x$PRTIME)+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(min(x$PRTIME)-1,max(x$PRTIME)+1,by=1)
+axis(side=1, at=xaxtl, labels=xaxtl)
+#axis(side=1, at=seq(min(h$mids)-1,max(h$mids)+1,1), labels=seq(min(x$PRTIME),max(x$PRTIME)+1,by=1))
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+1),col="green", main='PT frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist.eps")
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+2),col="green", main='PT frequency on INC4', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
[…]

### 2017-03-01 pt_characterization/data/empv5_et_hist.r

@@ -1,0 +1,311 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+#xaxtl <- seq(xmin-1,xmax+1,by=1)
+#axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_et_hist_v5.eps")
+binsize=1
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="blue", main='ET frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+#xaxtl <- seq(xmin-1,xmax+1,by=1)
+#axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
[…]

### 2017-03-01 pt_characterization/data/empv5_pt_hist.r

@@ -1,0 +1,266 @@
+x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("1_sec_pt_hist_v5.eps")
+binsize=1
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+xmin <-min(x$PRTIME)
+xmax <-max(x$PRTIME)
+plot(h, ylim=c(0,1000), xaxt='n',freq=TRUE,xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC1', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin-1,xmax+1,by=binsize)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="2_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("2_sec_pt_hist_v5.eps")
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
+binsize=1
+plot(h, axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+3),col="green", main='PT frequency on INC2', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()
+
+x = read.csv(file="4_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("4_sec_pt_hist_v5.eps")
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
[…]

### 2017-03-02 pt_characterization/recent_data/paper.tex
Draft for March 2

@@
 \begin{tabular}{|p{2cm}|p{3cm}|p{6cm}|p{4cm}|} \hline
-Machine & Task Length & Description & Time Length\\ \hline
-{\tt sodb9} &  INC1$\sim$INC1024 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10 / 2017-02-12 $\sim$ 2017-02-19\\ \hline
-{\tt sodb10} & INC2048 & Runs of 300 samples & 2017-02-12 $\sim$ 2017-02-20\\ \hline
+Machine & Task Length (sec) & Description & Experiment Period\\ \hline
+{\tt sodb9} &  INC1$\sim$INC1024 & Each run with 300 samples & 2017-02-09 $\sim$ 2017-02-10 / 2017-02-12 $\sim$ 2017-02-19\\ \hline
+{\tt sodb10} & INC2048 & A run of 300 samples & 2017-02-12 $\sim$ 2017-02-20\\ \hline
+{\tt sodb12} & INC4096 & A run of 300 samples & 2017-02-12 $\sim$ 2017-02-27\\ \hline
+{\tt sodb12} & INC8192 & Two runs, each with 40 samples & 2017-01-16 $\sim$ 2017-01-20 / 2017-01-25 $\sim$ 2017-01-29\\ \hline
+{\tt sodb12} & INC16384 & Two runs, each with 40 samples & 2017-01-05 $\sim$ 2017-01-13 / 2017-01-29 $\sim$ 2017-02-06\\ \hline 
 \end{tabular}
@@
 \vspace{-.2in}
-\caption{Detailed description of INC data used for histograms\label{tab:exp_notes}}
+\caption{Notes on experiment runs used for histograms\label{tab:exp_notes}}
 \end{table}
 
-Now we show histograms of elapsed time (ET) and process time (PT) 
-over increasing INC's task lengths.
+Now we show histograms of elapsed time (ET) and process time (PT) of INC.
 

### 2017-03-02 pt_characterization/recent_data/sodb9/res.tex
Draft for March 2

@@
 the task length of INC increases from 1 second to 2048 seconds. 
-The detailed description of the base data are from Table~\ref{tab:exp_notes}.
+The detailed description of the base data is from Table~\ref{tab:exp_notes}.
 
@@
 	}
+	\caption{ET Histograms of INC16 ... INC64~\label{fig:s9_et_hist2}}
+\end{figure}
+
+\begin{figure}[hp!]
+	\centering
 	\subfigure[ET frequency on INC128]{
@@
 	}
-	\caption{ET Histograms of INC16 ... INC128~\label{fig:s9_et_hist2}}
+	\subfigure[ET frequency on INC256]{
+		\includegraphics[scale=0.43]{sodb9/256_sec_et_hist_v5.eps}
+		\label{fig:inc256_et_hist_v5}
+	}
+	\subfigure[ET frequency on INC512]{
+		\includegraphics[scale=0.43]{sodb9/512_sec_et_hist_v5.eps}
+		\label{fig:inc512_et_hist_v5}
+	}
+	\subfigure[ET frequency on INC1024]{
+		\includegraphics[scale=0.43]{sodb9/1024_sec_et_hist_v5.eps}
+		\label{fig:inc1024_et_hist_v5}
+	}
+	\caption{ET Histograms of INC128 ... INC1024~\label{fig:s9_et_hist3}}
 \end{figure}
+
+\pagebreak
 
@@
 	\centering
-	\subfigure[ET frequency on INC256]{
-		\includegraphics[scale=0.43]{sodb9/256_sec_et_hist_v5.eps}
-		\label{fig:inc256_hist_v5}
-	}
-	\subfigure[ET frequency on INC512]{
-		\includegraphics[scale=0.43]{sodb9/512_sec_et_hist_v5.eps}
-		\label{fig:inc512_hist_v5}
-	}
-	\subfigure[ET frequency on INC1024]{
-		\includegraphics[scale=0.43]{sodb9/1024_sec_et_hist_v5.eps}
-		\label{fig:inc1024_hist_v5}
-	}
 	\subfigure[ET frequency on INC2048]{
 		\includegraphics[scale=0.43]{sodb10/2048_sec_et_hist_v5.eps}
-		\label{fig:inc2048_hist_v5}
+		\label{fig:inc2048_et_hist_v5}
 	}
-	\caption{ET Histograms of INC256 ... INC2048~\label{fig:s9_et_hist3}}
+	\subfigure[ET frequency on INC4096]{
+		\includegraphics[scale=0.43]{sodb12/4096_sec_et_hist_v5.eps}
+		\label{fig:inc4096_et_hist_v5}
+	}
+	\subfigure[ET frequency on INC8192]{
+		\includegraphics[scale=0.43]{sodb12/8192_sec_et_hist2_v5.eps}
+		\label{fig:inc8192_et_hist_v5}
+	}
+	\subfigure[ET frequency on INC16384]{
+		\includegraphics[scale=0.43]{sodb12/16384_sec_et_hist2_v5.eps}
+		\label{fig:inc16384_et_hist_v5}
+	}
+	\caption{ET Histograms of INC2048 ... INC16384~\label{fig:s9_et_hist4}}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}[hp!]
+	\centering
+	\subfigure[ET frequency on INC8192]{
+		\includegraphics[scale=0.43]{sodb12/8192_sec_et_hist_v5.eps}
+		\label{fig:inc8192_et_hist_all_v5}
+	}
+	\subfigure[ET frequency on INC16384]{
+		\includegraphics[scale=0.43]{sodb12/16384_sec_et_hist_v5.eps}
+		\label{fig:inc16384_et_hist_all_v5}
+	}
+	\caption{ET Histograms of INC8192... INC16384 [combined with 2015's run]~\label{fig:s9_et_hist5}}
 \end{figure}
@@
 	}
+	\caption{PT Histograms of INC16 ... INC64\label{fig:s9_pt_hist2}}
+\end{figure}
+
+\begin{figure}[hp!]
+	\centering
 	\subfigure[PT frequency on INC128]{
@@
 	}
-	\caption{PT Histograms of INC16 ... INC128~\label{fig:s9_pt_hist2}}
-\end{figure}
-
-\begin{figure}[hp!]
-	\centering
 	\subfigure[PT frequency on INC256]{
@@
 	}
+	\caption{PT Histograms of INC256 ... INC1024~\label{fig:s9_pt_hist3}}
+\end{figure}
+
+\begin{figure}[hp!]
+	\centering
 	\subfigure[PT frequency on INC2048]{
@@
 	}
-	\caption{PT Histograms of INC256 ... INC2048~\label{fig:s9_pt_hist3}}
+	\subfigure[PT frequency on INC4096]{
+		\includegraphics[scale=0.43]{sodb12/4096_sec_pt_hist_v5.eps}
+		\label{fig:inc4096_hist_v5}
+	}
+	\subfigure[PT frequency on INC8192]{
+		\includegraphics[scale=0.43]{sodb12/8192_sec_pt_hist2_v5.eps}
+		\label{fig:inc8192_hist_v5}
+	}
+	\subfigure[PT frequency on INC16384]{
+		\includegraphics[scale=0.43]{sodb12/16384_sec_pt_hist2_v5.eps}
+		\label{fig:inc16384_hist_v5}
+	}
+	\caption{PT Histograms of INC2048 and INC16384~\label{fig:s9_pt_hist4}}
 \end{figure}
+
+\pagebreak
+
+\begin{figure}[hp!]
+	\centering
+	\subfigure[PT frequency on INC8192]{
+		\includegraphics[scale=0.43]{sodb12/8192_sec_pt_hist_v5.eps}
+		\label{fig:inc8192_hist_all_v5}
+	}
+	\subfigure[PT frequency on INC16384]{
+		\includegraphics[scale=0.43]{sodb12/16384_sec_pt_hist_v5.eps}
+		\label{fig:inc16384_hist_all_v5}
+	}
+	\caption{PT Histograms of INC8192 and INC16384 [combined with 2015's run]~\label{fig:s9_pt_hist5}}
+\end{figure}

### 2017-03-02 pt_characterization/recent_data/sodb12/et_hist_sodb12.r

@@ -1,0 +1,90 @@
+## sodb12
+## 8192_sec
+
+x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8192_sec_et_hist2_v5.eps")
+x <- subset(x, x$ITERNUM > 40) 
+x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+binsize=10
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- max(h$counts)
+ymax <- 10
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-50,xmax+50),col="blue", main='ET frequency on INC8192', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+dev.off()
+
+x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8192_sec_et_hist_v5.eps")
+x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+binsize=50
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+xmin <-min(x$METIME)
[…]

### 2017-03-02 pt_characterization/recent_data/sodb12/pt_hist_sodb12.r

@@ -1,0 +1,81 @@
+x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8192_sec_pt_hist2_v5.eps")
+x <- subset(x, x$ITERNUM > 40) 
+x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+xmax <- max(x$PRTIME)
+xmin <- min(x$PRTIME)
+binsize=10
+nbins <- ceiling((xmax-xmin) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+xmin <-xmin-17
+xmax <-xmax+15
+plot(h, ylim=c(0,15), xaxt='n',freq=TRUE,xlim=c(xmin,xmax),col="green", main='PT frequency on INC8192', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), xlab='PT (ms)', ylab=expression('Frequency'))
+xaxtl <- seq(xmin,xmax,by=binsize)
+axis(side=1, at=xaxtl, labels=xaxtl)
+dev.off()
+
+x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("8192_sec_pt_hist_v5.eps")
+x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+xmax <- max(x$PRTIME)
+xmin <- min(x$PRTIME)
+binsize=10
+nbins <- ceiling((xmax-xmin) / binsize)
[…]

### 2017-03-02 pt_characterization/recent_data/sodb9/et_hist_sodb9.r

@@ -256,3 +256,25 @@ sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""),
 xlab='ET (ms)', ylab=expression('Frequency'))
 dev.off()
 
+### run on sodb12
+x = read.csv(file="../sodb12/4096_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("../sodb12/4096_sec_et_hist_v5.eps")
+x <- subset(x, x$ITERNUM != 5 & x$ITERNUM != 26 & x$ITERNUM != 47 & x$ITERNUM != 68 & x$ITERNUM != 89 & x$ITERNUM != 110 & x$ITERNUM != 131 & x$ITERNUM != 152 & x$ITERNUM != 173 & x$ITERNUM != 195 & x$ITERNUM != 215 & x$ITERNUM != 236 & x$ITERNUM != 258 & x$ITERNUM != 278 & x$ITERNUM != 299) 
+x_up = mean(x$METIME) + 2*sd(x$METIME)
+x_dn = mean(x$METIME) - 2*sd(x$METIME)
+binsize=5
+x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
+nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
+h = hist(x$METIME, right=F, breaks=nbins,plot=F)
+xmin <-min(x$METIME)
+xmax <-max(x$METIME)
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- max(h$counts)
+ymax <- 50
+plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-50,xmax+50),col="blue", main='ET frequency on INC4096', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='ET (ms)', ylab=expression('Frequency'))
+dev.off()
+

### 2017-03-02 pt_characterization/recent_data/sodb9/pt_hist_sodb9.r

@@ -232,3 +232,24 @@ plot(h, ylim=c(0,ymax), axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-4,max(x$PRTIM
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='PT (ms)', ylab=expression('Frequency'))
 dev.off()
+
+### run on sodb12
+x = read.csv(file="../sodb12/4096_sec.dat",head=TRUE,sep="\t")
+setEPS()
+postscript("../sodb12/4096_sec_pt_hist_v5.eps")
+x <- subset(x, x$ITERNUM != 5 & x$ITERNUM != 26 & x$ITERNUM != 47 & x$ITERNUM != 68 & x$ITERNUM != 89 & x$ITERNUM != 110 & x$ITERNUM != 131 & x$ITERNUM != 152 & x$ITERNUM != 173 & x$ITERNUM != 195 & x$ITERNUM != 215 & x$ITERNUM != 236 & x$ITERNUM != 258 & x$ITERNUM != 278 & x$ITERNUM != 299) 
+x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
+x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
+x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
+xmax <- max(x$PRTIME)
+xmin <- min(x$PRTIME)
+nbins <- ceiling((xmax-xmin) / binsize)
+h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
+binsize=1
+ymax <- max(h$counts)
+ymax <- ceiling(ymax/100)*100
+ymax <- 15
+plot(h, ylim=c(0,ymax), axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-4,max(x$PRTIME)+10),col="green", main='PT frequency on INC4096', 
+sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
+xlab='PT (ms)', ylab=expression('Frequency'))
+dev.off()

### 2017-03-02 pt_characterization/new_data/empv5_et_hist.r
2015's run summary

@@ -287,11 +287,12 @@ sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""),
 xlab='ET (ms)', ylab=expression('Frequency'))
 dev.off()
 
+#8192 sec
+
 x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
 setEPS()
 postscript("8192_sec_et_hist1_v5.eps")
-#x <- subset(x, x$ITERNUM <= 40) 
-x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
+x <- subset(x, x$ITERNUM <= 40) 
 x_up = mean(x$METIME) + 2*sd(x$METIME)
 x_dn = mean(x$METIME) - 2*sd(x$METIME)
 binsize=50
@@ -303,16 +304,18 @@ xmax <-max(x$METIME)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
 ymax <- max(h$counts)
-ymax <- 50
+ymax <- 8
 plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-200,xmax+200),col="blue", main='ET frequency on INC8192', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='ET (ms)', ylab=expression('Frequency'))
 dev.off()
 
+
+### 16384 sec
 x = read.csv(file="16384_sec.dat",head=TRUE,sep="\t")
 x <- subset(x, x$ITERNUM <= 40)
 setEPS()
-postscript("16384_sec_et_hist_v5.eps")
+postscript("16384_sec_et_hist1_v5.eps")
 x <- subset(x, x$ITERNUM != 10 & x$ITERNUM != 16)
 x_up = mean(x$METIME) + 2*sd(x$METIME)
 x_dn = mean(x$METIME) - 2*sd(x$METIME)
@@ -331,23 +334,4 @@ sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""),
 xlab='ET (ms)', ylab=expression('Frequency'))
 dev.off()
 
-x = read.csv(file="16384_sec.dat",head=TRUE,sep="\t")
-setEPS()
[…]

### 2017-03-02 pt_characterization/new_data/empv5_pt_hist.r

@@ -223,8 +223,8 @@ dev.off()
 x = read.csv(file="8192_sec.dat",head=TRUE,sep="\t")
 setEPS()
 postscript("8192_sec_pt_hist1_v5.eps")
-#x <- subset(x, x$ITERNUM <= 40) 
-x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
+x <- subset(x, x$ITERNUM <= 40) 
+#x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
 x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
 x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
 x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
@@ -235,7 +235,7 @@ nbins <- ceiling((xmax-xmin) / binsize)
 h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
 xmin <-xmin-17
 xmax <-xmax+15
-plot(h, ylim=c(0,15), xaxt='n',freq=TRUE,xlim=c(xmin,xmax),col="green", main='PT frequency on INC8192', 
+plot(h, ylim=c(0,10), xaxt='n',freq=TRUE,xlim=c(xmin,xmax),col="green", main='PT frequency on INC8192', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), xlab='PT (ms)', ylab=expression('Frequency'))
 xaxtl <- seq(xmin,xmax,by=binsize)
 axis(side=1, at=xaxtl, labels=xaxtl)
@@ -244,7 +244,8 @@ dev.off()
 x = read.csv(file="16384_sec.dat",head=TRUE,sep="\t")
 setEPS()
 postscript("16384_sec_pt_hist1_v5.eps")
[…]

### 2017-03-03 mlppi-comsis17/paper.tex
MLPPI paper bibliography
Intro updated

@@
 
-%\section{Introduction}
-%
-%The text of the paper\dots
-
 \section{Introduction}
@@
 
-Tables and materialized views in Teradata are hash distributed based 
+Over the past decades much attention has been paid 
+to improve the performance on analytical workloads in a data warehousing environment. 
+Typically, a relational database management system (\hbox{DBMS}) 
+has been exploited to process such analytical queries faster 
+and assist its customers to make a timely business-decision in 
+rapidly changing enterprise data warehousing.
+Teradata DBMS has been a leading database product in this data warehousing marketplace 
+that has been increasingly more competitive than ever.
+
+To optimize the performance of analytical processing, 
+tables and materialized views in the Teradata DBMS are hash-distributed based 
 on a user-specified column or set of columns called {\em primary index}. 
-Each virtual processor, a unit of parallelism called {\em AMP} in the \hbox{Teradata} \hbox{relational} 
-database management system (DBMS), 
+Each virtual processor, a unit of parallelism called {\em AMP} in 
+the \hbox{Teradata} \hbox{DBMS}, 
 receives a subset of the data and stores it in hash order.
 
-Teradata users typically choose primary index fields, 
+Users typically choose primary index fields in the DBMS, 
 so that the data is evenly distributed among the AMPs. 
-The primary index fields are also chosen to \hbox{reflect} join fields 
+The primary index fields are also chosen to reflect join fields 
 in workloads to accomplish cheaper local joins that do not require 
@@
 \vspace{-0.2in}
-\caption{An Example of MLPPI Table\label{fig:exam}}
+\caption{An Example of an MLPPI Table in the Teradata DBMS\label{fig:exam}}
 \end{figure}
@@
 \begin{center}
-\begin{tabular}{|c|p{6cm}|}\hline 
+\begin{tabular}{|c|l|}\hline 
 Partition & Condition \\ \hline
-1		& {\scriptsize  {\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY < 25}} \\ \hline
-2		& {\scriptsize  {\tt LO\_DISCOUNT >= 7 \&\& 25 <= LO\_QUANTITY <= 30}} \\ \hline
-3		& {\scriptsize {\tt LO\_DISCOUNT >= 7 \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
-4		& {\scriptsize {\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY no case}}\\ \hline
-5       & {\scriptsize {\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY < 25}} \\ \hline 
-6		& {\scriptsize {\tt LO\_DISCOUNT no case \&\& 25 <= LO\_QUANTITY < 30}} \\\hline								  	        
-7		& {\scriptsize {\tt LO\_DISCOUNT no case \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
-8		& {\scriptsize  {\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY no case}} \\ \hline
+1		& {{\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY < 25}} \\ \hline
+2		& {{\tt LO\_DISCOUNT >= 7 \&\& 25 <= LO\_QUANTITY <= 30}} \\ \hline
+3		& {{\tt LO\_DISCOUNT >= 7 \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
+4		& {{\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY no case}}\\ \hline
+5       & {{\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY < 25}} \\ \hline 
+6		& {{\tt LO\_DISCOUNT no case \&\& 25 <= LO\_QUANTITY < 30}} \\\hline								  	        
+7		& {{\tt LO\_DISCOUNT no case \&\& 30 < LO\_QUANTITY <= 35}} \\ \hline
+8		& {{\tt LO\_DISCOUNT no case \&\& LO\_QUANTITY no case}} \\ \hline
 \end{tabular}
@@
         & {\tt AND l.LO\_QUANTITY <= 30} \\ 
-		&						   \\	        
 $q$2:	& {\tt SELECT c.C\_NATION, SUM(l.LO\_REVENUE)} \\ 
@@
 		& {\tt GROUP BY c.C\_NATION} \\
-		& {\tt ORDER BY revenue desc}. \\
+		& {\tt ORDER BY revenue desc} \\
 \end{tabular}
@@
 and the sequence number of the constraint in each query, as illustrated below.
-%\vspace{-0.05in}	
+\vspace{-0.03in}	
 \begin{center}
@@
 for {\tt LO\_DISCOUNT}:
-\begin{center}
+
+\begin{center}
+{\footnotesize
 \begin{tabular}{l}
@@
 \end{tabular}
+}
 \end{center}
@@
 \includegraphics[scale=0.5]{architecture}
-%\vspace{-0.2in}
-\caption{The MLPPI Wizard Architecture\label{fig:arch}}
+\vspace{-0.1in}
+\caption{The MLPPI wizard architecture\label{fig:arch}}
 \end{figure*}
@@
 
-The article is a substantial extension of prior work~\cite{Suh12}, 
-and organized in the following way. 
-Section~\ref{sec:algorithms} presents the multi-level \hbox{partitioning} 
-algorithms for the wizard. 
-In turn, Section~\ref{sec:analysis} conducts an analysis of the complexity of 
-the algorithms. 
-Section~\ref{sec:experiment} reports the performance evaluation results. 
-Section~\ref{sec:related_work} states how \hbox{Teradata} MLPPI wizard 
-is different from some existing tools proposed by other vendors. 
-Lastly, Section~\ref{sec:conclusion} summarizes our discussion.
+The article is a substantial extension of prior work~\cite{Suh12}. 
+The rest of this article is \hbox{organized} in the following way. 
+In the following section we propose a \hbox{multi-level} partitioning algorithm---consisting of a series of phases---used by our MLPPI wizard, 
+which is applicable to any clustering or partitioning scheme in an arbitrary DBMS. 
+In turn, we conduct a detailed analysis of the complexity of the algorithms. 
+Next, we report the performance evaluation results. 
+We then elaborate on how our wizard is distinguished from several existing tools proposed by other DBMS vendors. 
+Finally, we conclude this article by summarizing our discussion.
 
@@
 
-In this section we describe in detail each phase of the wizard using the query set $Q$ provided in Section~\ref{sec:intro}.
+In this section we describe in detail each phase of the wizard using the query set $Q$ \hbox{provided} in Section~\ref{sec:intro}.
 
@@
 
+{\footnotesize
 \bibliographystyle{splncs03}
 \bibliography{paper}
+}
 

### 2017-03-06 mlppi-comsis17/paper.tex
Intro updated on MLPPI paper

@@
 
+%\subsection{Background}
+%\label{sec:background}
+
+\paragraph{Background:} 
 To optimize the performance of analytical processing, 
@@
 receives a subset of the data and stores it in hash order. 
-
 Users typically choose primary index fields in the DBMS, 
@@
 The primary index fields are also chosen to reflect join fields 
-in workloads to accomplish cheaper local joins that do not require 
-data shuffling.
-
-PPI (Partitioned Primary Index)~\cite{sinclair:ppi} is an optional horizontal 
-partitioning scheme applied locally on each AMP's data. 
-Note that in other database products, this type of partitioning would 
-most likely to be called {\em clustering}. 
-In the rest article we use the term \hbox{{\em partitioning}} only, not clustering, to avoid confusion. 
+in workloads to accomplish cheaper local joins that do not require data shuffling.
+
+Partitioned Primary Index (PPI)~\cite{sinclair:ppi} is an optional horizontal 
+partitioning scheme applied locally on the data belonging to each AMP. 
+In some database products, this type of partitioning is termed {\em clustering}, 
+but hereafter we call the term \hbox{{\em partitioning}} only to avoid confusion.
 Database \hbox{administrators} (DBAs) usually choose the columns for PPI, 
@@
 in the workload. 
-The PPI columns are used to physically cluster data with the same values 
+The PPI columns are used to physically group data with the same values 
 together in contiguous data blocks. 
-This allows ``partition elimination'' in scans and joins for performance improvement.
-PPI can be specified as single or multiple (or nested) levels. 
+This grouping enables ``partition elimination'' in scans and joins for performance enhancement.
+PPI can be specified as single or multiple levels. 
 This type of partitioning scheme is known as 
-Multi-Level PPI ({\em MLPPI)}~\cite{klindt09mlppi}.
-
-By allowing non-qualified partitions to be eliminated, MLPPI can reduce 
-significantly the amount of data to be scanned to answer a query. 
+\hbox{Multi-Level} PPI ({\em MLPPI)}~\cite{klindt09mlppi}.
+
+By allowing non-qualified partitions to be eliminated, MLPPI can considerably 
+reduce the amount of data to be scanned to answer a query. 
 But a large number of partitions can create significant overhead, 
@@
 
-Figure~\ref{fig:exam} exemplifies an MLPPI table with potential partitioning schemes. 
+Figure~\ref{fig:exam} exemplifies an MLPPI table with a defined partitioning option. 
 This example is a subset of the {\tt LINEORDER} table from the Star Schema Benchmark (SSB)~\cite{oneil:ssb}. 
-Note that the full table with modified data types was actually used in our experiments.
+(Note that the full table with modified data types was actually used in our experiments.)
 
@@
 In the definition of {\tt LINEORDER} in the figure, the primary index 
-is {\tt LO\_ORDERKEY}. 
-The \hbox{primary} index \hbox{dictates} the AMP on which a row will be located, 
+is on {\tt LO\_ORDERKEY}. 
+The \hbox{primary} index \hbox{dictates} the AMP on which a row will be located 
 while the partitioning of data will be dictated by the values and 
 ranges associated with {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}. 
-{\tt LO\_DISCOUNT}, for example, has two ranges for \hbox{values} 
-greater than or equal to 7 and one ({\tt NO CASE OR UNKNOWN}) 
-for all the other values, while \linebreak {\tt LO\_QUANTITY} has four ranges. 
-As a result, the relational table will have a total of 2$\times$4 = 8 partitions as follows.
-
-\begin{center}
+For example, {\tt LO\_DISCOUNT} has two ranges: (i)~one for \hbox{values} $\geq$ 7 
+and (ii)~another for all the other values (or {\tt NO CASE OR UNKNOWN}). 
+{\tt LO\_QUANTITY} has four ranges. 
+Hence, the relation will have a total of 2$\times$4 = 8 partitions as follows.
+
+\begin{center}
+{\footnotesize
 \begin{tabular}{|c|l|}\hline 
-Partition & Condition \\ \hline
+Partition Number & Condition \\ \hline
 1		& {{\tt LO\_DISCOUNT >= 7 \&\& LO\_QUANTITY < 25}} \\ \hline
@@
 \end{tabular}
-\end{center}
-
-Whatever partition set is chosen, 
+}
+\end{center}
+
+Whatever partition is chosen, 
 the selection must be semantically correct; namely, 
@@
 that specify conditions on {\tt LO\_DISCOUNT} and/or {\tt LO\_QUANTITY}. 
-For \hbox{example}, only partitions 1 and 2 are needed for the query 
+For \hbox{example}, only partitions~1 and~2 are needed for the query 
 ``{\tt SELECT * FROM LINEORDER WHERE LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
-Similarly, partitions 1 and 5 are sufficient to answer the query 
+Similarly, partitions~1 and~5 are sufficient to answer the query 
 ``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}.'' 
 \linebreak The respective sizes of the partitions are a factor of the data distribution. 
-To see MLPPI in greater detail, check our references~\cite{sinclair:ppi,klindt09mlppi}.
-
-Next, using the full version of {\tt LINEORDER} as defined in SSB~\cite{oneil:ssb} 
-we provide examples of partitioning schemes that 
-a DBA may define based on a small query set. 
-Consider a query set $Q$ consisting of two queries $q$1 and $q$2 from the SSB set~\cite{oneil:ssb}, as shown below.
-
+To see MLPPI in greater detail, check our references~\cite{klindt09mlppi,sinclair:ppi}.
+
+%\vspace{-.1in}
+
+%Next, using {\tt LINEORDER} defined in the SSB we provide examples of partitioning options that 
+%a DBA may define based on a small query set. 
+\paragraph{Challenge:}
+Consider a query set $Q$ having two queries $q$1 and $q$2 on the SSB, as shown below.
+\vspace{-.05in}
 \begin{center}
@@
 }
-\end{center}
-
-Let's now focus on the {\tt LINEORDER} fact table and 
+\vspace{-.01in}
+\end{center}
+
+A DBA may focus on the {\tt LINEORDER} fact table and 
 the predicates involving {\tt LINEORDER} fields only. 
-There are five constraints identified by the query number 
-and the sequence number of the constraint in each query, as illustrated below.
-\vspace{-0.03in}	
-\begin{center}
-\begin{tabular}{|c|l|}\hline 
-$q$1.1 & {\tt LO\_DISCOUNT IN (1,4,5)} \\ \hline
-$q$1.2 & {\tt LO\_QUANTITY <= 30} \\ \hline
-$q$2.1 & {\tt LO\_DISCOUNT >= 7} \\ \hline
-$q$2.2 & {\tt LO\_QUANTITY >= 25}\\ \hline
-$q$2.3 & {\tt LO\_QUANTITY <= 35} \\ \hline 
-\end{tabular}
-\end{center}
-%\vspace{-0.05in}	
-At this point the DBA needs to consider options 
+He can figure out the following five constraints, identified by the query number 
+and the sequence number of predicates in each query: 
+\vspace{-.05in}
+\begin{center}
+\begin{tabular}{cl}
+$q$1.1: & {\tt LO\_DISCOUNT IN (1,4,5)} \\ 
+$q$1.2: & {\tt LO\_QUANTITY <= 30} \\ 
+$q$2.1: & {\tt LO\_DISCOUNT >= 7} \\ 
+$q$2.2: & {\tt LO\_QUANTITY >= 25}\\ 
+$q$2.3: & {\tt LO\_QUANTITY <= 35}. \\ 
+\end{tabular}
+\end{center}
+\vspace{-.05in}
+At this point the DBA needs to take into account options 
 based only on two fields and the five constraints. 
-There are many possibilities from a {\em fine-grained} partition set 
-to a loser definition.
+There are many other possibilities from a {\em fine-grained} partition set 
+to a {\em coarser-grained} definition.
 
@@
 for {\tt LO\_DISCOUNT}:
-
+\vspace{-.05in}
 \begin{center}
@@
 The above expression minimizes the size of the partitions 
-by focusing exactly on the values required to satisfy the constraints, 
+by focusing exactly on the values required to satisfy the constraints 
 but creates a large number of small partitions.
@@
 a between clause yielding the following:
+\vspace{-.1in}
 \begin{center}
@@
 \end{tabular}
+\vspace{-.1in}
 \end{center}
@@
 
-Another possibility is to use the partitioning associated with {\tt LO\_DISCOUNT} shown 
-in the beginning of this section with only two partitions.
-
-Similar considerations apply to the partitioning associated with {\tt LO\_QUANTITY} 
-this time with ranges and covering problems difficult to deal with. 
-The resulting partitioning scheme for the table includes both {\tt LO\_DISCOUNT}
-and {\tt LO\_QUANTITY}, so that the number of possible combinations is the product of 
+(Another possibility would be to use the partitioning associated with {\tt LO\_DISCOUNT} shown 
+in the beginning of this section with only two partitions.)
+
+Similar considerations apply to the partitioning associated with {\tt LO\_QUANTITY}. 
+The resulting partitioning definition for the table includes both {\tt LO\_DISCOUNT}
+and \linebreak {\tt LO\_QUANTITY}, and thus the number of possible combinations is the product of 
 the potential combinations for each field.
-Also, there are two queries, and one partitioning scheme may be better for one query or the other. 
-As a result, the DBA will be faced with a daunting combinatorial search problem 
+
+Furthermore, given two queries one partitioning scheme may be favorable for one query but not for the other, or vice versa. 
+As a result, the DBA becomes faced with a daunting combinatorial search problem 
 and no clear basis to decide on which combination is the best. 
-This state of affairs begs for a {\em tool} to assist the DBA.
-
-In the sequel, we introduce an automated tool, called {\em MLPPI wizard}. 
-Indeed, the \hbox{wizard} is the first physical database design tool 
-developed at Teradata.
-The tool is based on a novel technique using a greedy algorithm for 
-search space enumeration. 
-This tool based on a general framework allowing general expressions, 
-ranges and case expressions for \hbox{partition} \hbox{definitions} 
-is particularly well-suited for an MLPPI definition. \linebreak
-The {\em predicate-driven} \hbox{technique} used by the tool can be applied 
-to any clustering or partitioning based on simple fields and expressions or complex SQL predicates. 
+This state of affairs begs for a {\em tool} for the DBA.
+
+%\subsection{Solution: An Automated Partitioning Tool}
+%\label{sec:solution}
+\paragraph{Solution:} 
+To ease a burden of partitioning decision, 
+we introduce a novel physical database design tool, called {\em MLPPI wizard}, 
+designed to recommend an effective partitioning solution for a given workload.
+%Indeed, the \hbox{wizard} is the first physical database design tool 
+%developed at Teradata.
+This tool uses a greedy algorithm for search space enumeration 
+and is based on a general framework allowing general expressions, 
+ranges and case expressions for partition definitions. 
+%is particularly well-suited for an MLPPI definition. 
+We emphasize that the \hbox{predicate-driven} algorithm used by the tool can be applied 
+to {\em any} clustering or partitioning scheme based on simple fields and expressions or complex SQL predicates 
+in an {\em arbitrary} DBMS. 
 The wizard also borrows the optimizer's cost model 
@@
 \centering
-\includegraphics[scale=0.5]{architecture}
+\includegraphics[scale=0.48]{architecture}
 \vspace{-0.1in}
@@
 {\em Initial}, and {\em Optimized} Phases. 
-In the preprocessing phase the tool produces the most granular partitioning (MGP) based on the predicates collected from every query.
+In the pre-processing phase the tool produces the most granular partitioning (MGP) based on the predicates collected from every query.
 In the initial phase the wizard refines the passed MGP 
@@
 In the \hbox{optimization} phase, the tool makes a further attempt to optimize the initial MLPPI. 
-We provide in greater \hbox{detail} each phase, 
-using the query set $Q$ consisting of $q$1 and $q2$ in the rest of the article.
-
-\shorten{
-An initial MLPPI is produced through the top two phases reflecting practical constraints by field or partition. 
-It is further optimized in the last phase.
-In the preprocessing phase, 
-for each query its predicate(s) is extracted via an API call (on {\tt explain}) made to the server. 
-Then, the wizard produces the most granular partitioning (MGP) based on the 
-predicates collected from every query. 
-If the MGP has too many fields over a field count limit, 
-then the wizard \hbox{incrementally} throws away a field from the MGP until the limit.
-
-In the initial phase, the passed MGP gets reduced if it includes more partitions than a partition count limit allowed by an MLPPI. 
-The refinement can be \hbox{incrementally} done by merging a pair of partitions 
-with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}), 
-until the limit. 
-After the phase, an initial (feasible) MLPPI is produced based on the refined partitions.
-
-In the optimization phase, the tool makes a further attempt to optimize the initial MLPPI. 
-Similar to the initial phase, more partitions can be reduced by merging ones with the 
-least {\em query cost} (as will be described in Section~\ref{sec:opt_phase}). 
-If the ongoing solution cannot be further improved, the tool makes the final MLPPI recommendation. 
-We will provide more details about each phase later, 
-using the query set $Q$ consisting of $q$1 and $q2$ 
-in the rest of the article.}
-
-To the best of our knowledge, there is no existing industrial tool 
-to solve the {\it {\bf multi-level partitioning}} problem except our wizard. 
-The wizard is contained completely on the client side, as opposed to 
-previous work~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} 
-requiring optimizer code extension. 
+We elaborate in greater detail on each phase, 
+using the running query set $Q$ consisting of $q$1 and $q2$ in the rest of the article.
+
+%\shorten{
+%An initial MLPPI is produced through the top two phases reflecting practical constraints by field or partition. 
+%It is further optimized in the last phase.
+%In the preprocessing phase, 
+%for each query its predicate(s) is extracted via an API call (on {\tt explain}) made to the server. 
+%Then, the wizard produces the most granular partitioning (MGP) based on the 
+%predicates collected from every query. 
+%If the MGP has too many fields over a field count limit, 
+%then the wizard \hbox{incrementally} throws away a field from the MGP until the limit. 
+%
+%In the initial phase, the passed MGP gets reduced if it includes more partitions than a partition count limit allowed by an MLPPI. 
+%The refinement can be \hbox{incrementally} done by merging a pair of partitions 
+%with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}), 
+%until the limit. 
+%After the phase, an initial (feasible) MLPPI is produced based on the refined partitions.
+%
+%In the optimization phase, the tool makes a further attempt to optimize the initial MLPPI. 
+%Similar to the initial phase, more partitions can be reduced by merging ones with the 
+%least {\em query cost} (as will be described in Section~\ref{sec:opt_phase}). 
+%If the ongoing solution cannot be further improved, the tool makes the final MLPPI recommendation. 
+%We will provide more details about each phase later, 
+%using the query set $Q$ consisting of $q$1 and $q2$ 
+%in the rest of the article.}
+
+To the best of our knowledge, there is few extant industrial tool 
+to addressing the {\it {\bf \hbox{multi-level} partitioning}} problem, except our wizard. 
+The wizard resides completely on the client side as opposed to 
+several previous tools~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} 
+that require optimizer code instrumentation. 
 Our tool simply uses existing APIs to simplify the queries, capture fact table 
 predicates and costs, and uses these items to make a recommendation. 
-This makes the wizard extensible and portable to different releases 
-of the Teradata DBMS server.
+%This makes the wizard extensible and portable to different releases 
+%of the Teradata DBMS server. 
+This \hbox{loosely-coupled} approach makes the wizard extensible and portable to different releases 
+of the database server.
 

### 2017-03-07 mlppi-comsis17/paper.tex
MLPPI updates

@@
 
-(Another possibility would be to use the partitioning associated with {\tt LO\_DISCOUNT} shown 
-in the beginning of this section with only two partitions.)
+%(Another possibility would be to use the partitioning associated with {\tt LO\_DISCOUNT} shown 
+%in the beginning of this section with only two partitions.)
 
@@
 
-%\subsection{Solution: An Automated Partitioning Tool}
-%\label{sec:solution}
 \paragraph{Solution:} 
@@
 \centering
-\includegraphics[scale=0.48]{architecture}
+\includegraphics[scale=0.45]{architecture}
 \vspace{-0.1in}
@@
 a set of queries and corresponding weights. 
-The wizard passes through a series of phases: {\em Preprocessing}, 
+Until yielding a final partitioning recommendation, 
+the wizard passes through a series of phases: {\em Preprocessing}, 
 {\em Initial}, and {\em Optimized} Phases. 
-In the pre-processing phase the tool produces the most granular partitioning (MGP) based on the predicates collected from every query.
-In the initial phase the wizard refines the passed MGP 
-by merging a pair of partitions with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}) 
-and then yields an initial (feasible) MLPPI based on the refined partitions.
-In the \hbox{optimization} phase, the tool makes a further attempt to optimize the initial MLPPI. 
-We elaborate in greater detail on each phase, 
+%In the pre-processing phase the tool produces the most granular partitioning (MGP) based on the predicates collected from every query.
+%In the initial phase the wizard refines the passed MGP 
+%by merging a pair of partitions with the least (scan) cost 
+%and then yields an initial (feasible) MLPPI based on the refined partitions.
+%In the \hbox{optimization} phase, the tool makes a further attempt to optimize the initial MLPPI. 
+We elaborate in greater detail on each of the three phases, 
 using the running query set $Q$ consisting of $q$1 and $q2$ in the rest of the article. 
-
-%\shorten{
-%An initial MLPPI is produced through the top two phases reflecting practical constraints by field or partition. 
-%It is further optimized in the last phase.
-%In the preprocessing phase, 
-%for each query its predicate(s) is extracted via an API call (on {\tt explain}) made to the server. 
-%Then, the wizard produces the most granular partitioning (MGP) based on the 
-%predicates collected from every query. 
-%If the MGP has too many fields over a field count limit, 
-%then the wizard \hbox{incrementally} throws away a field from the MGP until the limit. 
-%
-%In the initial phase, the passed MGP gets reduced if it includes more partitions than a partition count limit allowed by an MLPPI. 
-%The refinement can be \hbox{incrementally} done by merging a pair of partitions 
-%with the least {\em scan cost} (as will be explained in \hbox{Section~\ref{sec:init_phase}}), 
-%until the limit. 
-%After the phase, an initial (feasible) MLPPI is produced based on the refined partitions.
-%
-%In the optimization phase, the tool makes a further attempt to optimize the initial MLPPI. 
-%Similar to the initial phase, more partitions can be reduced by merging ones with the 
-%least {\em query cost} (as will be described in Section~\ref{sec:opt_phase}). 
-%If the ongoing solution cannot be further improved, the tool makes the final MLPPI recommendation. 
-%We will provide more details about each phase later, 
-%using the query set $Q$ consisting of $q$1 and $q2$ 
-%in the rest of the article.}
-
-To the best of our knowledge, there is few extant industrial tool 
-to addressing the {\it {\bf \hbox{multi-level} partitioning}} problem, except our wizard. 
-The wizard resides completely on the client side as opposed to 
+As illustrated in Figure~\ref{fig:arch}, the wizard resides completely on the client side as opposed to 
 several previous tools~\cite{agrawal04:integrating,Lightstone04:db2auto,nehme2011automated} 
 that require optimizer code instrumentation. 
-Our tool simply uses existing APIs to simplify the queries, capture fact table 
-predicates and costs, and uses these items to make a recommendation. 
+Our tool simply invokes the server's existing APIs used to simplify the queries, capture fact table 
+predicates, and estimate processing costs. 
 %This makes the wizard extensible and portable to different releases 
 %of the Teradata DBMS server. 
-This \hbox{loosely-coupled} approach makes the wizard extensible and portable to different releases 
-of the database server.
+%This \hbox{loosely-coupled} interaction 
+%between both sides makes the wizard extensible and portable to different releases 
+%of the database server. 
+To the best of our knowledge, our wizard  
+is the very first physical database design tool 
+to address such a {\em \hbox{multi-level} partitioning} problem.
+
+\paragraph{Contribution:} This article provides 
+the following contributions.
+
+\begin{itemize}
+\item We propose a physical database design tool, termed MLPPI wizard, to recommend a multi-level partitioning definition 
+for reducing the processing cost of a given query workload on a star schema benchmark. 
+\item This tool is totally outside the database server, thereby incurring no overhead of instrumenting query optimizer's code.
+\item We empirically show that a partitioning recommendation of our tool outperforms that of human expert 
+over increasing workload size and growing fact table.
+
+\end{itemize}
 

### 2017-03-07 subopt-tods16/paper.tex
SubOpt updates

@@
 \usepackage{amssymb}
+\usepackage{subfigure}
 
@@
 
-\section{Introduction}
-DBMSes underlie all information systems and hence optimizing their performance is of
+\section{Introduction}\label{sec:intro}
+DBMSes underlie information systems and hence optimizing their performance is of
 critical importance. The DBMS's query optimizer plays an important role. But
@@
 \item We find that for a surprisingly large portion of queries, the plan
-  chosen  by the query optimizer is not the best plan, for some cardinality,
+  chosen  by the query optimizer is not the best plan, for some cardinality
   of the underlying tables.
 
-\item We propose and test a {\em predictive model} for \hbox{DBMSes} to better understand the factors causing suboptimality.
-
-\item \todo{Rick:Mention generational result}
+\item We propose a {\em predictive model} for \hbox{DBMSes} to better
+  understand the factors causing suboptimality.
+
+\item We test the six quite specific hypotheses that arise from that
+  model across a wide range of DBMSes, queries, and data.
+
+\item We show through correlational and regression analysis that four
+  hypothesized interactions are strongly supported, one is partially supported, and
+  an interaction hypothesized to go one direction went quite weakly the
+  other direction, lending credibility to our particular model.
+
+\item Through an innovative
+  analysis, we track the net cumulative benefit of a succession of operators
+  added to the DBMS, and show that current relational DBMSes may have
+  already reached the stage where it is difficult to provide performance
+  improvement via new operators, and in fact performance may degrade.
 
@@
   engineering directions.
+
+\item The model and analyses presented here taken as a whole imply that a
+  new approach to query optimization, fundamentally
+  different from that utilized over the last forty years, may be required
+  to get past a fundamental limitation uncovered in this research.
 \end{itemize}
@@
 understand how existing query optimizers can be further improved, and
-indeed, fundamental limitations inherent in these optimizers.
-
-\begin{figure*}[t]
-\centering
-%originally 30pc
-%\epsfig{figure=figures/plan769.pdf,width=36pc}
-\includegraphics[width=30pc]{figures/subopt_flutter.pdf}
-%\includegraphics[scale=0.42]{figures/subopt_flutter.pdf}
-\caption{An Example of Suboptimality and Fluttering\label{fig:query769}}
-\end{figure*}
+indeed, whether  fundamental limitation inherent in these optimizers exist
+and whether such limitations are indeed already being encountered.
 
@@
 The following section introduces the methodology we will follow, that of
-empirical generalization. We then present a predictive, causal model of
-suboptimality and empirically test eight specific hypotheses derived from that
-model\c2j{. These are the first scientific results that we are aware of that
-  test a general causal model of \hbox{DBMSes}.}{ on three \hbox{DBMSes}.
-  These are the first scientific results that we are aware of that apply
-  {\em across} \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
-  specific algorithm.} This model has implications for research in
-engineering more efficient \hbox{DBMSes}.
+{\em empirical generalization}~\cite{cohenbook}. We present in
+Sections~\ref{sec:model} and~\ref{sec:operationalization} a predictive,
+causal model of suboptimality and state seven specific hypotheses derived
+from that model. We then test these hypotheses across a number of queries,
+data, and cardinalities, whose results provide strong support for the
+validity of the proposed model.  These are the first scientific results that
+we are aware of that apply {\em across} \hbox{DBMSes}, rather than on a
+single, specific \hbox{DBMS} or on a specific algorithm. In
+Section~\ref{sec:diminishing} presents a follow-on analysis that
+demonstrates a fundamental limit to the number of operators that a
+cost-based optimizer can support, and indicates that perhaps extant DBMSes
+have hit that limit. This model has implications listed in
+Section~\ref{sec:engineering} for research in engineering more efficient
+\hbox{DBMSes}.
 
@@
 cardinalities. So Plan~P0 was chosen for 2M tuples, switching to Plan~P1 at
-1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through the
+1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through
+the plan
 sequence P0, P1, P0, P1, P2, P0, P2, and finally P3 at 40,000 tuples.
+
+\begin{figure*}[t]
+\centering
+%originally 30pc
+%\epsfig{figure=figures/plan769.pdf,width=36pc}
+\includegraphics[width=30pc]{figures/subopt_flutter.pdf}
+%\includegraphics[scale=0.42]{figures/subopt_flutter.pdf}
+\caption{An Example of Suboptimality and Fluttering\label{fig:query769}}
+\end{figure*}
 
@@
 \begin{figure}[h!]\centering
+%\todo{Young: were is the figure!}
 \includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
@@
 plan.
-
-\vspace{0.6em}\noindent
-{\bf Hypothesis 4}: Plan space complexity (CEPS) and Number of operators
-with discontinuity will be positively correlated.
 
@@
 \vspace{0.6em}\noindent
-{\bf Hypothesis 5}: The Number of repeats will positively correlate with CEPS.
+{\bf Hypothesis 4}: Plan space complexity (CEPS) and (a)~Number of operators
+with discontinuity and (b) Number of repeats will be positively correlated.
 
@@
 is to increase the strength of the interaction between query
-complexity to plan space complexity.
+complexity to plan space complexity. (Note that as the presence of primary
+key is required for secondary indexes, we don't include that former
+independent variable here.)
 
 \vspace{0.6em}\noindent
-{\bf Hypothesis 6}: Presence of secondary indexes will strengthen
+{\bf Hypothesis 5}: Presence of secondary indexes will strengthen
 the correlations between the Query Complexity construct, specifically, the
 variables Number of
-correlation names, Presence of aggregate, Presence of primary key
-attribute, and Presence of subquery, and
+correlation names, Presence of aggregate, and Presence of subquery, and
 the Plan Space Complexity construct, specifically, variables Number of repeats, CEPS, and Number
-of operators with discontinuity (correlations a--l listed above in
+of operators with discontinuity (correlations a--f and j--l listed above in
 Hypothesis 2).
@@
 \vspace{0.6em}\noindent
-{\bf Hypothesis 7}: Suboptimality will be positively correlated with Plan
+{\bf Hypothesis 6}: Suboptimality will be positively correlated with Plan
 space complexity, that is, (a)~Number of repeats, (b)~CEPS, and (c)~Number
@@
 suboptimality might arise, through a theory and its elaborated causal model,
-which implies seven specific hypotheses.
+which implies six specific hypotheses (some with multiple components).
 We now need to move to {\em prediction}.
@@
 
-It is important to note that all manipulation must be done {\em
-  outside} the \hbox{DBMS}. For proprietary systems we do not have access to the
-internal code. We do not know ({\em cannot} know) all the plans that were
+It is important to note that manipulation must be done {\em
+  outside} the \hbox{DBMS}, as we do not have access to the
+internal code of proprietary DBMSes. Hence, we do not know ({\em cannot} know) all the plans that were
 considered, nor the details of how the plans are selected. But such access
@@
 The {\tt EXPLAIN PLAN} facility specifies the operators(s) \hbox{employed} in that
-plan. For each \hbox{DBMS}, we collect all the unique operators from all the plans
+plan. For each \hbox{DBMS}, we collect the unique operators from the plans
 and count the number of these distinct operators, each used by for least one
 query at at least one cardinality by that \hbox{DBMS}. The number of operators used
-by a \hbox{DBMS} ranged from 8 to 53 across all the queries and data sets
+by a \hbox{DBMS} ranged from 8 to 53 across the queries and data sets
 used in the Exhaustive, Exhaustive with Keys, and Exploratory Experiments,
@@
 eliminated\c2j{.}{\shorten{({\tt cartesianPossible="false"})}}  We do this by
-connecting all the correlation names that appear in the FROM statement via
+connecting the correlation names that appear in the FROM statement via
 equi-joins on random attributes. The comparisons are all
@@
 This skew factor translates to 10 distinct values. 
-So we generate 200K spans, each with 10 values. 
+So we generate 200K spans, each with ten values. 
 To shorten, we delete 1,000 spans, which is equivalent to removing 10K values. 
@@
 being evaluated on one of the 200 cardinalities for the variable table. It
-is ``effective'' because it was chosen, as opposed to all of the
+is ``effective'' because it was chosen, as opposed to of the
 plans that were considered but not chosen. (Recall that for proprietary
@@
 study. We
-collected all query plans at all possible cardinalities (200 in all) and
+collected the query plan at each possible cardinalities (200 in all) and
 looked for ``jumps,'' that is, when the cost model for an operator in the
@@
 each plan at a particular cardinality. Each distinct point type depicts an
-individual hash-join operator. Note that when two identical query plans
+individual hash-join operator. \todo{Young: I don't understand what this
+  sentence adds'' Rick: Yes, please delete the sentence.}Note that when two identical query plans
 appear at two different cardinalities, we consider all the plan operators
@@
 This computes a series of slope values.
-We then compute the standard deviation of all the slope
+We then compute the standard deviation of the slope
 values.  For example, examining Figure~\ref{fig:discontinuity}, the slopes
@@
 \subsection{Suboptimality}\label{sec:suboptimality}
-We now consider the one dependent variable.
+We now consider the one dependent variable at the core of this investigation.
 How might suboptimality be observed? We have developed a system, \azdb, that
@@
 
-We modify the cardinality to produce multiple execution
-plans for a given query. For one of the \hbox{DBMSes}, we can modify the stored
-table statistics directly. For the other \hbox{DBMSes}, we had to do so indirectly,
-by varying the size of the table and running the optimizer on tables of
-different size. As we vary the cardinality, we
-collects all the plans that the optimizer felt were appropriate for that
-query.
+We modify the cardinality to produce multiple execution plans for a given
+query. For one of the \hbox{DBMSes}, we can modify the stored table
+statistics directly. For the other \hbox{DBMSes}, we had to do so
+indirectly, by varying the size of the table and running the optimizer on
+tables of different size. As we vary the cardinality, we collect the plans
+that the optimizer felt were appropriate for that query at the various cardinalities.
 
@@
 As we will see in Section~\ref{sec:experiments}, we observed only 3,347
-violations (0.73\%) of non-strict monotonicity, for the largest experiment
+violations (0.74\%) of non-strict monotonicity, for the largest experiment
 (Experiment 7: Confirmatory) across all
@@
 At each Q@C pair, Suboptimality is coded as four levels (0--3), based on the distance in 
-standard deviations, upto 3 standard deviations. 
+standard deviations, up to three standard deviations (we chose this cutoff
+because 99.7\% of Q@C pairs are less than or equal to three standard
+deviations). 
 \shorten{\begin{itemize}
@@
 \noindent}
-We then sum this value over all pairs of cardinalities with different plans
-(the change points) for the query.
-
-\todo{Sabah: this paragraph talks about a suboptimality of up to 133, but the
-  above papra says there are just 4 levels. That is inconsistent. The reason is that the raw value ranges from 0 to 133. 
-After the log transformation, the values range from 0 and 2.13. Also, do
-  we still use a log transformation?}
+We then sum this value over the Q@C pairs with different plans
+(the change points) to arrive at a single integer for the query instance.
+
 Note that while \azdb\ examines the plan at every cardinality, it only has
@@
 
-No runs violated the experiment-wide sanity checks. For the largest
+No run violated the experiment-wide sanity checks. For the largest
 experiment, the Confirmatory experiment 7 discussed below, approximately
-10.5\% of the query exectuions and 5.1\%
-of the queryies-at-cardinality (Q@Cs) were dropped due to query execution
+10.5\% of the query executions and 5.1\%
+of the queries-at-cardinality (Q@Cs) were dropped due to query execution
 and Q@C sanity checks. As a result, excessive variation in calculated query
-time was observed in only 0.003\% of the Q@Cs. A total of 0.35\% of Q@C adjacent
-pairs violated relaxed monotonicity and 0.68\% of the Q@Cs violated strict
+time was observed in only 0.003\% of the Q@Cs. A total of 0.43\% of Q@C adjacent
+pairs violated relaxed monotonicity and 0.74\% of the Q@Cs violated strict
 monotonicity, which is acceptable.
@@
 \doubleblind{In fact, over the course of the research described here, we
-  found that currently available techniques did not offer sufficient query time
+  found that currently available techniques for measuring query time did not offer sufficient query time
   precision, and so we developed TTPv1~\cite{TTPv1} to provide that needed
-  precision. As our research in suboptimality prooceeded, we realized that
-  to get a actually recognized suboptimality for two adjacent cardinalities
-  separated by only 10,000 rows (just a few pages), or even in the case of
+  precision. As our research in suboptimality proceeded, we realized that
+  to detect suboptimality for two adjacent cardinalities
+  separated by only 10,000 rows (just a few database pages), or even in the case of
   MySQL, 300 rows, we needed even finer resolution. This provided the
@@
 In the following, we describe in detail the data used by our experiments and
-the experimental scenarios we defined. More details on the data sets used
+the experimental scenarios we defined. More details on both
 can be found in Appendix~\ref{sec:app}.
@@
   Figure~\ref{fig:empirical}, from one system to several systems and then to
-  a general theory. We don't reveal the identity of the \hbox{DBMS} we studied, for two reasons. First, commercial \hbox{DBMSes}
+  a general theory. We don't reveal the identity of the \hbox{DBMSes} we studied, for two reasons. First, commercial \hbox{DBMSes}
 include in their user agreements requirements not to release 
@@
 doesn't matter, as we are studying phenomena about cost-based optimizers
-{\em in general}, and so are interested in making statements that apply to
-all the experimental subjects in our study.
+{\em in general}, and so are interested in making statements that apply across
+the experimental subjects in our study.
 
@@
 1 & Monotonicity 		& --- & 38 & 60 & 12,000 & 12,000 & 12,000\\
-2 & Initial Exhaustive 		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
+2 & Refined Monotonicity 		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
 3 & Exhaustive 			& TTPv2 & 1,544 & 160 & 32,000 & 320,000 & 319,980\\
@@
     became available,} we performed our
-second experiment, termed ``Initial Exhaustive,'' which more accurately tested the
-monotonicity assumption, as also described in Section~\ref{sec:suboptimality}.
+second experiment, termed ``Refined Monotonicity,'' which more accurately tested the
+(strict) monotonicity assumption, as also described in
+Section~\ref{sec:suboptimality}.\todo{Young: I don't see any such discussion
+  in that section..., Rick: Please take a look at the definition of the strict monotonicity assumption. We discuss this there.}
 This experiment involved 160 query instances, 32,000 \QatC s, and thus
-320,000 QEs, requiring 1,672 cumulative hours.  The (very small) percentage
+320,000 QEs, requiring 1,672 cumulative hours.  The (very small\todo{Young:
+  how much?, Rick: 1.55\%}) percentage % (16312+8215)/(995321+587546) = 1.55%
 of strict monotonicity violations observed was consistent with the remaining
@@
 
-We used the plans generated from Initial Exhaustive to classify operators as
+We later reran this experiment using TTPv2 (we term this simply
+``Exhaustive''). Note that the number of retained QEs went up a lot (that is
+the main benefit of TTPv2), while the number of hours required actually
+went down (another benefit).
+
+We then used the plans generated from the Exhaustive Experiment to classify operators as
 continuous or discontinuous, as discussed in detail in
@@
 
-We later reran this experiment using TTPv2 (we term this simply
-``Exhaustive''). Note that the number of retained QEs went up a lot (that is
-the main benefit of TTPv2), while the number of hourse required actually
-went down (another benefit).
-
-We also ran an experiment (number 3 in the table) on the Exhaustive query set but using the data
-set with primary keys, termed ``Exhaustive with Keys.'' However, in this
+We also ran an experiment (termed ``Exhaustive with Keys'') on the Exhaustive query set but using the data
+set with primary keys,  However, in this
 case we did not actually execute 
@@
 \hbox{DBMS} to identify additional discontinuous operators. \comment{Thus this third
-experiment was very fast.}
-
-The fourth experiment was for initial exploratory analysis of a prior
-version of the causal model. (This fourth experiment used
+experiment was very fast.}\todo{Sabah:were any additional discontinous
+  operators found? Rick: I don't know about this. I redirect this question to Sabah.}
+
+\todo{Rick: Shouldn't this experiment  be before Experiment 3 (Exhaustive),
+  because it used TTPv1?}The fifth experiment was for initial exploratory analysis of a prior
+version of the causal model. (Note that this experiment used
 the first version of the Tucson Timing Protocol (TTPv1)~\cite{TTPv1}.)
@@
 for a total of 1,910 queries and a total of 7,640 query instances (over the
-four DBMSes).
-Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed, roughly 13 per query.
-\todo{Young, where did this number come from?:}The protocol accurately timed 53,547
-{\QatC} pairs as change points.\
-
-In this Confirmatory Experiment, we observed 3,347 strict
-monotonicity violations (0.74\%), and 1,966 relaxed monotonic violations
-(0.43\%) (out of a total of 452,684 pairs), which
-provides further confidence that monotonicity also applies to
-operators found in queries over data in various contexts (as described above) 
-and that our operationalization of suboptimality is a valid one.
-
-In the remainder of this section, we focus using the measured independent,
-latent, and dependent variables in the Confirmatory experiment
-to test predictions that arise out of our causal model in Figure~\ref{fig:model}.
+four \hbox{DBMSes}).
+Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed,
+roughly 13 per query. The protocol accurately timed
+both sides of adjacent {\QatC}s (53,547 change points in all).
+
+\todo{Young: these percentages  are different from those in Section 6.1.
+Why?
+Rick: A good catch.
+Those numbers are out-of-date: ones from the TODS metrology paper.
+I replaced them with the current numbers mentioned in this para.
+Now consistent.}In this Confirmatory Experiment, we observed 3,347 strict monotonicity
+violations (0.74\%), and 1,966 relaxed monotonicity violations (0.43\%) (out
+of a total of 452,684 pairs
+of two Q@Cs having identical plans and query instance), \todo{Rick:merge
+  somehow with Section 6.1:}which provides
+further confidence that monotonicity also applies to operators found in
+queries over data in various contexts (as described above) and that our
+operationalization of suboptimality is a valid one.
+
+In the remainder of this section, we focus using the measured independent
+and dependent variables in the Confirmatory experiment
+to test the predictions that arise out of our causal model in Figure~\ref{fig:model}.
 
@@
 result of several years of programming effort and about 30 months of
-experimental runs. First, {\em every} \hbox{DBMS} exhibits suboptimality. Thus, this
-phenomenon is likely to be a fundamental aspect of either the algorithm
-(cost-based optimization) or the creator of the algorithm (human information
-processing). Our model includes both effects. Perhaps surprisingly, 
-%a little over half (1579 queries out of the 3056 query instances
-more than half (3,933 queries out of the 6,983\todo{Rick: I have 6967 in my set since 16 were dropped 
-because of missing one side of change point}  
-query instances that emerged from the measurement protocol) exhibited suboptimality
+experimental runs, summarized in Table~\ref{tab:run_stat}.
+
+Let illustrate the descriptive statistics with the Confirmatory
+experiment. In that experiment, we started with 7,640 query instances. These
+resulted in 99,558 Q@Cs, each with ten query executions (QEs). Some of these
+QEs violated one or more of the (many) sanity checks specified within the
+TPPv1 and TTPv2 protocols, leaving a remaining 890,631
+retained QEs. Some of the query instances then voliated one or more of the
+Q@C sanity checks in the protocols (657 in all), leaving a remaining 6,983 retained query
+instances. Additionally, 16 query instances were dropped {\em after} the
+protocol because they were missing data for the only change point in the
+query.
+
+Therefore, we were left with 6,967 queries, of which we make three general observations.
+
+First, perhaps surprisingly, 
+more than half (3,933 queries out of those that emerged) exhibited suboptimality
 somewhere in the range of cardinality of the varying table. 
-Most instances of those queries (3,370) had the meximum value of suboptimality at a Q@C pair (i.e., level 3). 
+{\em Every} \hbox{DBMS} exhibits suboptimality.
+
+Secondly, most (3,370) of those suboptimal queries (86\%) had the maximum value of
+suboptimality  (i.e., level 3) at some Q@C pair.
 \shorten{a big number for a query could be many pairs, each at a small
@@
 querynum from Cnfm_Analysis_SuboptDefn where subopt_sd>=3);}
-CEPS ranged from 1 to 24,
-plans across the cardinality range. The number of discontinuous operators observed in plans for a query
-averaged 9.4, with 85 being the maximum. Number of repeats ranged from 0 to 108, with the mean being 4.99.
-
-\subsection{Correlational Analysis}
-We tested Hypotheses 1--7 using the strength and significance of correlations
-of variables involved. These hypotheses 
-predicts eleven positive relationships. Table~\ref{tab:hyptest} lists the hypotheses followed by the
-correlation observed when testing each hypothesis.  (``NS'' denotes not
-significant at the 0.05 level, the accepted standard for
-significance. ``---'' denotes no prediction arising from the model.) As can
-be seen, most of the predictions (nine) arising from the causal model are
-supported and significant.  The two
-exceptions are Hypotheses 1 and 2. Hypothesis 1 was not supported because
-the correlation between number of operators available and suboptimality was negative,
-while we predicted positive correlation.  \hbox{Hypothesis}~2 was not significant.
-
-\subsection{Regression Analysis}
-\todo{Sabah: revisit this entire section, as we changed our model}
-We ran a regression over the independent variables of the
-model that predict suboptimality over the data from the Confirmatory experiment. 
-Our model explained \% of the variance of the suboptimality dependent
-variable.
-
-We also did regressions on the causal variables for CEPS, number of repeats and for number of
-operators with discontinuity. Our model explained \% of the
-variance for CEPS, \% of the variance for number of repeats, and \% of the variance for
-discontinuity. \shorten{need path analysis}
-
-Hypothesis 5 predicts that the presence of secondary indexes will be a moderator,
-affecting the interactions between query complexity with plan space
-complexity. It thus provides
-predictions that the strength of such interactions will decrease in the
-presence of secondary keys. When
-secondary indexes were not specified, our model explains 
-\shorten{\% of the variance
-of suboptimality,} \% of the variance of CEPS, and \% of the variance
-of discontinuity, with all of the independent variables. When
-secondary indexes were defined on the underlying tables, our model explains
-\% of the variance of suboptimality, \% of the variance of CEPS, and
-\% of the variance of discontinuity. These findings are all consistent with our
-hypothesis predicting a positive moderation effect of primary keys.
-
-\todo{Sabah: revisit. Not sure about the purpose of this paragraph}
-We also examined the individual contributions to each of these
-three variables (e.g., suboptimality) from their causal variables (in the
-case of suboptimality, from the four variables of correlation names, presence
-of an aggregate, CEPS, and discontinuity). 
-Table~\ref{tab:hyptest8} shows that the strength of all interactions goes
-down as predicted, except for ones involving aggregates. Recall
-that the predictive model indicates weaker effects generally for aggregates,
-as they are evaluated late in the query, and the primary key attributes are
-not necessarily included in the grouping attributes.
-
-\todo{Sabah: redo hypotheses in these tables, for example, H1 doesn't go to
-  subopt, then add the confirmatory numbers after we discuss them}
+
+The phenomenon of query suboptimality that we observed is likely to be a
+fundamental aspect of either the algorithm (cost-based optimization) or the
+creator of the algorithm (human information processing). Our model includes
+both effects.
+
+Third, concerning the causal factors of suboptimality in our model,
+\begin{itemize}
+\item the cardinality of the effective plan space (CEPS) ranged from 1 to 24
+plans across the cardinality range,
+\item the number of discontinuous operators observed in plans for a query
+averaged 9.4, with 85 being the maximum, and
+\item the number of repeats ranged from 0 to 108, with the mean being 4.99.
+\end{itemize}
+
+\subsection{Correlational Analysis}\label{sec:corr}
+We tested Hypotheses 1--6 using the strength and significance of
+correlations of variables involved. These hypotheses predict 21 main effects
+and 12 moderating effects. Table~\ref{tab:hyptest} lists the hypotheses
+followed by the correlation observed when testing each hypothesis.  (``NS''
+denotes not significant at the $p < 0.05$ level, the accepted standard for
+significance. For each dependent variable, we used the Bonferroni correction
+to control the familywise error rate. ``---'' denotes no prediction arising
+from the model.)  As can be seen, most of the main effects (18) arising from
+the causal model are supported and significant.  The four exceptions are
+Hypotheses 1c, 2g, 2j, and 3. Hypothesis 1c was not supported because the
+correlation between Number of operators available and Number of operators
+with discontinuity was negative, while we predicted positive
+correlation. Similarly Hypothesis 3 was not supported because the
+correlation between Percentage of skewed data and Suboptimality was
+negative, while we predicted positive correlation. (In both cases, the
+strength of the negative correlation was small: -0.09 and -0.05, respectively.) \hbox{Hypotheses}~2g and
+2j, both involving Number of repeats, were not significant.
+
 \begin{table}[t]
-\tbl{Testing Hypotheses 1--7: Correlations on the Confiratory Study\label{tab:hyptest}}{%
+\tbl{Testing Hypotheses 1--7: Correlations on the Confirmatory Study\label{tab:hyptest}}{%
 %\begin{center}
@@
 \hline
-Operators in \hbox{DBMS} & ---&  {\bf H1a:}$-$0.16 & {\bf H1b:} & {\bf H1c:} \\
-Correlation names & {\bf }  & {\bf H2a:} 0.54 & {\bf H2b:}  &{\bf H2c:} 0.49\\
-Skew & {\bf H3:}  & --- & ---  & ---\\
-Presence of aggregate & & {\bf H2d:}  & {\bf H2e:}  & {\bf H2f:} \\
-Presence of primary key & & {\bf H2g:}  & {\bf H2h:}  & {\bf H2i:} \\
-Presence of subquery & & {\bf H2j:}  & {\bf H2k:}  & {\bf H2l:} \\
-Repeats& {\bf H7a:}  & --- & {\bf H5:}  & ---\\
-CEPS &  {\bf H7b:} 0. & --- &  --- &{\bf H4:}  \\
-Discontinuity & {\bf H7c:} 0.31 & --- & --- & ---\\
+Operators in \hbox{DBMS} & ---&  {\bf H1a:} 0.35 & {\bf H1b:} 0.27& {\bf H1c:} -0.09\\
+Correlation names & --- & {\bf H2a:} 0.17 & {\bf H2b:} 0.46 &{\bf H2c:} 0.34\\
+Presence of aggregate & -- & {\bf H2d:} 0.04 & {\bf H2e:} 0.10 & {\bf H2f:} 0.29\\
+Presence of primary key & -- & {\bf H2g:} NS & {\bf H2h:} 0.15 & {\bf H2i:} 0.14\\
+Presence of subquery & -- & {\bf H2j:} NS & {\bf H2k:} 0.26 & {\bf H2l:} 0.20\\
+Skew & {\bf H3:} -0.05 & --- & ---  & ---\\
+Repeats& {\bf H6a:} 0.62 & --- & --  & ---\\
+CEPS &  {\bf H6b:} 0.61 & {\bf H4b:} 0.45 &  --- &{\bf H4a:} 0.72 \\
+Discontinuity & {\bf H6c:} 0.47 & --- & --- & ---\\
 \end{tabular}
@@
 \begin{table}[t]
-\tbl{Testing Hypothesis 6: Interaction Strength on the Confirmatory Study\label{tab:hyptest8}}{%
+\tbl{Testing Hypothesis 5: Interaction Strength on the Confirmatory Study\label{tab:hyptest5}}{%
 %\begin{center}
@@
 \hline
-Correlation names &  & 0. &0. &0.  & 0. & 0. \\
-Presence of aggregate  & 0. & 0. & 0. & 0. & 0. & 0. \\
-Presence of primary key  &  & 0. & --- & --- & --- & --- \\
-Presence of subquery  &  & 0. & --- & ---  & --- & --- 
+Correlation names & 0.17 & 0.15 &0.49 &0.39  & 0.36 & 0.27 \\
+Presence of aggregate  & 0.04 & NS & 0.12 & NS & 0.31 & 0.22 \\
+Presence of primary key  & -0.03 & NA & 0.06 & NA & 0.09 & NA \\
+Presence of subquery  & 0.03 & NS & 0.18 & 0.17  & 0.14 & 0.13 
 \end{tabular}
@@
 \end{table}
+
+\subsection{Regression Analysis}\label{sec:regression}
+Regression analysis is required to test Hypothesis 5, which predicts a positive
+moderation in the presence of query complexity, specifically the Number of
+correlation names, Presence of aggregate, and Presence of subquery
+independent variables. Thus our causal model predicts that the interaction
+strength should increase when secondary indexes appear in the realtional schema.
+In Table~\ref{tab:hyptest5}, we examine the interaction strength both when
+the secondary indexes are absent (``Not SI'') and when they are present
+(``SI''). As before, ``NS'' denotes no significant at the $p < 0.05$ level (three
+of these interactions beome not significant when secondary indexes are
+added). ``NA'' denotes inapplicable: as the Presence of primary key is
+required for secondary indexes, it is not possible to compute the
+interaction strength in the presence of secondary indexes.
+
+This table shows that the strength of all six applicable and significant
+interactions goes {\em down} just a little (on average by 0.05), which is
+opposite to our prediction. Hence, this negative moderation seems small but
+robust. One possibility is that rather than adding
+complexity, secondary indexes may have {\em reduced} the complexity of query
+optimization, since the role of secondary indexes within query optimization
+is well-understood.
+
+A second use of regression is to compute the amount of variance explained,
+as another way of testing the moderation effect.
+Hypothesis 5 thus predicts that the strength of the interactions between
+query complexity with plan space complexity will increase in the
+presence of secondary keys. When
+secondary indexes were not specified, our model explains 
+7.23\% of the variance of number of repeats, 40.9\% of the variance of CEPS, and 35.7\% of the variance
+of discontinuity\todo{Sabah: don't understand this ending phrase:}, with all of the independent variables. When
+secondary indexes were defined on the underlying tables, our model explains
+7.23\% of the variance of number of repeats, 41.1\% of the variance of CEPS, and
+38.9\% of the variance of discontinuity. These findings are all consistent with our
+hypothesis predicting a positive moderation effect of secondary
+indexes. Hence, the jury is still out on that hypothesis.
+
+A third use of regression is to compute the amount of variance explained
+for the intervening and dependent variables, to indicate how much of the
+underlying phenomena are explained holistically by the model.
+
+We did regressions on the causal variables for CEPS, number of repeats and for number of
+operators with discontinuity. Our model explained 41.1\% of the
+variance for CEPS, 7.23\% of the variance for number of repeats, and 38.9\% of the variance for
+discontinuity. \shorten{need path analysis}
+
+We also ran a regression over the independent variables of the model that
+predict suboptimality over the data from the Confirmatory experiment. Our
+model explained 53.6\% of the variance of the suboptimality dependent
+variable. That means that over half of the variance of suboptimality is
+explained by the independent and intevening variables in our model, the ten
+variables within our five constructs of Optimizer Complexity, Schema
+Complexity, Query Complexity, Data Complexity, and Plan Space Complexity.
+
+To state this a different way, all other possible causes for suboptimality
+will in concert have less predictive power than the constructs and specific
+variables included in the causal model introduced here.
 
@@
 the presence of an aggregate and CEPS were significant at the 0.05 level, yet
-our model explains fully 53\% of the suboptimality variance (and 56\% of the
-CEPS variance and \todo{Sabah:??} of the discontinuity variance). This might imply
+our model explains fully \% of the suboptimality variance (and \% of the
+CEPS variance and \% of the discontinuity variance). This might imply
 that heuristic optimization itself introduces perhaps 10--20% of the variance.
@@
 The size of the regression coefficients also provides useful
-information. From the regressions run on all of the data, for the
+information. From the regressions run on the data, for the
 statistically significant causal factors, the highest coefficients, all in
-predicting the suboptimality dependent variable, were CEPS at 0.42, number
-of tables involved at 0.23, number of discontinuous variables at 0.16,
-number of operators available in \hbox{DBMS} at 0.06. Note though that number of correlation
+predicting the suboptimality dependent variable, were CEPS at , number
+of tables involved at , number of discontinuous variables at ,
+number of operators available in \hbox{DBMS} at . Note though that number of correlation
 variables is highly correlated with CEPS; the regression will apportion the variance of the dependent
@@
 encapsulates our theory for how cost-based query optimizers might select a
-suboptimal plan for a query at a cardinality. This model implies the eight
+suboptimal plan for a query at a cardinality. This model implies the six
 specific hypotheses listed in Section~\ref{sec:model}. We then performed
-four experiments to refine our operationalizations (such as discontinuity,
-via the Exhaustive Experiment and number of operators available with the additional
-Exhaustive with Keys Experiment), to test fundamental assumptions (such as
-monotonicity, via the Monotonicity Experiment), and to test and make minor
-refinements to our model (via the Exploratory Experiment). While exploring
-the data, one must be cognizant of the possibility of Type 1 errors: false
-positives that lead one to believe a relationship exists when it doesn't.
-
-To control for Type 1 errors, we then performed the Confirmatory 
+six experiments,
+\begin{itemize}
+\item to refine our operationalizations: specifically discontinuity,
+via the Initial \hbox{Exhaustive} and Exhaustive \hbox{Experiments}, and number of operators available, via the
+Exhaustive with Keys \hbox{Experiment},
+\item to test fundamental assumptions, specifically
+monotonicity, via the Monotonicity \hbox{Experiment}, and
+\item to test and make minor
+refinements to our model: the Initial Exploratory and \hbox{Exploratory}
+Experiments.
+\end{itemize}
+During this exploration, \doubleblind{once we completed the TTPv2 protocol,}{when the TTPv2 protocol became available,} we reran
+\hbox{experiments} to avail ourselves of the increased precision due to a much
+higher percentage of retained QEs and thus Q@Cs and query instances.
+
+Throughout these six experiments, we were cognizant of the
+possibility of Type 1 errors: false positives that lead one to believe a
+relationship exists when it doesn't.
+To control for such errors, we then performed the final Confirmatory 
 Experiment on a completely different data set consisting of many more 
@@
 
-Correlational analysis provides significant support for the model, except
-for Hypotheses 1 and 2, with the implication that the influence of the
-number of operators available remains largely unexplored. Via regression
-analysis, the model explains 34\% of the variance of suboptimality (overall:
-with primary keys a little lower and without primary keys a little higher
-due to that moderating effect), 31\% of CEPS, and 35\% of discontinuity. In
-all but one case, the causal influence of independent and latent variables is significant
-($p<0.05$). The direction of the regression coefficients, again in all but one
-case as predicted, also provides strong support for the model.
-
 Now that the causal model has been found to be supported by the confirmatory
-analysis, we turn to possible implications of this model.
+\hbox{analysis}, we turn to possible implications of this model.
 
@@
 
-Our model includes three causal factors of suboptimality: \hbox{DBMS} optimizer
-complexity, query complexity, and plan space complexity. Of these factors,
-the regression \hbox{coefficient} that was highest was for CEPS
-(cf.~Table~\ref{tab:hyptest8}: 0.51--0.54, normalized) The next \hbox{highest}
-regression factor was number of tables involved, which is highly correlated
-with CEPS (cf.~Table~\ref{tab:hyptest}: 0.54). These two observations imply that the
+\todo{Sabah: update with final numbers:}
+Our model includes two causal factors of suboptimality: \hbox{DBMS} data complexity, and plan space complexity. Of these factors,
+the regression \hbox{coefficient} that was highest was for number of repeats (0.46, normalized). The next \hbox{highest}
+regression factor was CEPS (0.27, normalized). These two observations imply that the
 number of plans being considered is a major determinate of
 suboptimality, indicating that choosing among many plans is hard, despite
-many decades of research and development.
+many decades of research and development. \todo{Rick: Please confirm this paragraph is still valid}
 
 The next most influential factor is the number of discontinuous operators
-(cf.~Table~\ref{tab:hyptest8}: 0.40--0.41, normalized),
+(0.176, normalized),
 \hbox{implicating} the cost model.
-
-\shorten{Sabah, how does the following sound?}The one following that is the number of operators provided by the \hbox{DBMS}.
-This factor is totally correlated with the \hbox{DBMS}, and so
-is confounded with other specificities of the coding of each \hbox{DBMS}. That said,
-it is still the case that as the number of operators goes up, across
-all the \hbox{DBMSes} studied, the amount of suboptimality increases. This
-\hbox{factor} also strongly impacts CEPS, and so indirectly impacts suboptimality
-through that path.
 
@@
 
-\section{Diminishing Returns?}
-In order to increase query performance, \hbox{DBMSes} are extended over time with
-new \hbox{relational} query operators. Also over time, \hbox{DBMSes} are extended with new
-storage and indexing structures,
-which themselves elicit new query operators. Each subsequent generation of
-the \hbox{DBMS} thus supports an ever-expanding collection of operators,
-thus the current version is simply the most recent within a series of \hbox{DBMS}
-{\em generations}.
+\section{Diminishing Returns?}\label{sec:diminishing}
+In order to increase query performance, \hbox{DBMSes} are extended over time
+with new \hbox{relational} query operators. Also over time, \hbox{DBMSes}
+are extended with new storage and indexing structures, which themselves
+elicit new query operators. For example, nested loop join was probably the
+first implementation of this needed algebraic operator. B-tree index join
+was only possible when such indexes were added (quite early in the history
+of modern DBMSes). Adding hashing to a DBMS enables a hash join operator;
+adding hash indexes enables (perhaps several variants of) hash-index join. Each
+subsequent generation of the \hbox{DBMS} thus supports an ever-expanding
+collection of operators, with the current incarnation the most recent
+within a series of \hbox{DBMS} {\em generations}.
 
 \subsection{A Gedanken Experiment}\label{sec:gedanken}
-Consider a Gedanken experiment that analyzes the plan for each query at
-each cardinality, that is, for each Q@C, for each \hbox{DBMS} generation. In early
-generations, there will be fewer plan changes as the cardinality increases
-for a given query, just \hbox{because} there are fewer operators available. For
-later generations, some of the Q@Cs will be associated with different plans,
-enabled by the new operators added by subsequent generations. As our causal
-model in Figure~\ref{fig:model} predicts, the effective plan space for a
-query may grow in subsequent generations, due to the increased number of
-operators available.
-
-In many (hopefully most) cases, the new plan will be more efficient than
-staying with the previous plan selected in a previous generation. After all,
-that is the very reason the new operator(s) were added to that generation.
-However, in the presence of suboptimality,
-sometimes the new plan at that Q@C is {\em not} preferable,
-as that \hbox{DBMS} generation's query optimizer selected the wrong
-plan.  Indeed, the query optimizer also evolves and improves with each new
+Consider a Gedanken experiment that examines the plan for each query at
+each cardinality, that is, each Q@C, {\em for each \hbox{DBMS} generation}. In early
+generations, there will be few plan changes for a given query as the cardinality varies
+simply \hbox{because} there are few operators available. For
+subsequent generations, some of the Q@Cs will be associated with different plans,
+enabled by the new operators that were added.
+
+In many (hopefully most) cases, a new plan selected by a generation will be
+more efficient than the plan selected by the immediate previous
+generation. After all, that is the very reason the new operator(s) were
+added to that subsequent generation.  However, in the presence of
+suboptimality, sometimes the new plan at that Q@C is {\em not} preferable,
+as that \hbox{DBMS} generation's query optimizer selected the wrong plan.
+Indeed, the query optimizer also evolves and improves with each new
 generation, in part to minimize the chance of selecting the wrong plan.
 
-Our predictive model suggests that suboptimality is causally
-impacted by the independent variable of CEPS (cardinality of effective plan
-space\todo{Rick:Connect with number of operators available and with Section~\ref{sec:root}}. A
-possible implication is that as the generations of the \hbox{DBMS} add more and more
-operators, suboptimality will also increase. \todo{Rick:No: causality is with
-  CEPS, not number of operators!}
-
-\todo{Rick:Make point that our causal analysis is across \hbox{DBMSes} with different
-  numbers of operators. Here we're talking about {\em within} a single \hbox{DBMS},
-  across multiple generations, each with an additional operator, and thus a
-  higher CEPS. In fact, it would be useful to see if generation number was correlated with CEPS.}
+Our predictive model in Figure~\ref{fig:model} suggests that suboptimality is causally impacted by the
+independent variable of CEPS (cardinality of effective plan space), and CEPS
+is naturally connected to the number of operators available\todo{Rick:
+  connect with Section~\ref{sec:root} after Sabah updates that section; also
+  discuss the exponential increase in CEPS possible}. A possible implication
+is that as the evolution of generations of the \hbox{DBMS} add more and more
+operators, suboptimality will also increase.
+
+It is important to emphasize that our causal analysis to this point has been
+{\em across multiple \hbox{DBMSes}}, each with a different set of
+operators and even number of operators. Let's reprise from Section~\ref{sec:intro} on
+page~\pageref{sec:intro}: Our goal, up to this point, has been ``to \hbox{understand} cost-based query
+optimizers
+as a {\em general} class of computational artifacts and to come up with
+insights and ultimately with predictive theories about how such optimizers,
+again, as a general class, behave.''
+
+The present discussion has a quite different focus. Here we're talking about
+the query optimizer {\em within} a single \hbox{DBMS}, across multiple
+generations, each generation adding one or more operators.
+
+Let's then refine our Gedanken experiment. Let's assume each subsequent
+generation adds a single operator. Starting with a fixed set of queries,
+for each DBMS, we run each generation on each Q@C and then sum up the query
+times to compute a {\em per-generation (total) time} for that DBMS. We can
+also sum over the four DBMSes, to see how this  {\em overall per-generation
+  time} (a single number for each generation number) varies with generation.
 
 The underlying question then becomes, does an additional operator made
-available in a subsequent generation of the \hbox{DBMS} actually help or hurt?
+available in a subsequent generation of the \hbox{DBMS} actually help or
+hurt? More fully, is the predicted increase in suboptimality originating
+from that added operator, a causal effect discovered and validated by the process
+of {\em science}, specifically the application of \hbox{empirical} generalization illustrated
+in Figure~\ref{fig:empirical}, compensated for by the increased performance
+afforded by that operator, a benefit realized by the application of {\em
+  engineering}, the articulation, elaboration, and implementation of new
+storage structures and query evaluation algorithms?
+
+Fortunately, the engineering has already been done: current DBMSes have at
+their disposal multiple query operators, which have been perfected over the
+decades. Adopting the engineering perspective would predict that the overall per-generation
+time would {\em monotonically decrease as the generations add query evaluation
+operators}.
+
+Interestingly, adopting the scientific perspective reaches a different conclusion. Our
+causal model in Figure~\ref{fig:model} predicts that as the number of
+operators increases, the latent measures of plan space complexity increase,
+which therefore also {\em increase} suboptimality. As the generations of the
+\hbox{DBMS} contain successively greater numbers of operators, suboptimality
+will also increase with generation number.  Given that each new operator
+will improve a {\em shrinking}
+subset of Q@Cs for any given query, while the suboptimality occurs in a portion over an
+{\em expanding} subset of Q@Cs, our causal model implies that the curve of
+per-generation time will fall with each successive generation, but then
+start to level off as suboptimality becomes more prevalent (specifically,
+the prediction is that the second derivative is either. Eventually, the
+per-generation time will either asymptotically approach a horizontal line
+(with the first derivative approaching 0 from below), or worse: the first
+derivative will change to positive, with the per-generation time over the
+queries we are studying actually {\em increasing} with \hbox{DBMS}
+generation. In either case, the causal model predicts a point where {\em the
+\hbox{increase} in execution time due to suboptimality obviates the decrease
+enabled by the new operator}.
+
+Put in simpler terms, engineering considerations alone imply that DBMSes can
+continue to improve, whereas also adopting the scientific perspective predicts
+that eventually each DBMS will hit a wall, beyond which improvement is not possible.
+
+Does such a limit exist, and if so, how close are modern \hbox{DBMSes} to that limit?
 
@@
 
-We then examine the plans containing exactly two operators (in the case of
-MySQL, one of which is
-full table scan, as a two-operator plan not having that operator would
-necessarily be constructed by generation 3 or higher). Using MySQL again, there are two sets of
-such plans, those with the full table scan and full table scan with join
-operators (888 Q@Cs) and
-those with the full table scan and ref operators (112 Q@Cs). We pick full
-table scan with join as the operator added by Generation 2, given its
-prevalence of Q@Cs (a total of 1705 in the full set). Each
+We then examine the plans containing exactly two operators.
+Using MySQL again, there are two such : one with the Full Table Scan and
+Full Table Scan with Join operators (888 Q@Cs) and
+one with the Full Table Scan and Ref operators (112 Q@Cs). Full
+Table Scan with Join is thus the operator added by Generation 2, given its
+prevalence of Q@Cs. Each
 subsequent generation adds that operator that maximizes the number of plans
-that operator will eventual enable. So Generation 3 adds the eq\_ref operator,
-as that operator enables 633 plans eventually. Generation 4 adds the ref
-operator and Generation 5 adds the index operator. There are six plans that
-involve all five operators.
+that operator will eventual enable. So Generation 3 adds the Eq\_Ref operator,
+as that operator enables 633 plans eventually. Generation 4 adds the Ref
+operator and Generation 5 adds the Index operator.
 
@@
 there is a query that has an astonishingly high number of change points,
-172, all of that generation.
-
-There's another limitation we're hitting, one that is much more fundamental.
+172, all waffling between plans within that generation.
+
+This analysis indicates that there is something concerning that seems to get
+critical around generation 8.  And indeed, for this data set, we'll see in
+Section~\ref{sec:hitthewall} that the efficacy of the query optimizer
+bottoms out around that generation.
+
+\subsection{Using Change Points}\label{sec:usingCP}
+We now consider how to use data already collected, that is the {\em change
+  points} discussed in Section~\ref{sec:experiments}: adjacent
+cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
+plans. Specifically, how can change points be used to evaluate the
+effectiveness of different generations of a \hbox{DBMS}?
+
+Define $\hbox{\em ops}(p)$ for a given plan $p$ to be the set of operators
+present in that plan, with some operators perhaps repeated in that plan. 
+A generation $g$ (also a set of operators) is {\em applicable} to a plan $p$ if
+$\hbox{\em ops}(p) \subseteq g$. By
+definition, if a generation is applicable to a given plan, it is
+applicable to all subsequent generations. The smallest such generation is
+termed the {\em minimally applicable generation}, or {\em mingen}.
+\shorten{
+\shorten{We couldn't make this one work...
+\todo{Young:replace in the above para with an actual SQL example query that was associated
+  with two MySQL generations in a Q@C pair. use the generation numbers from
+  the next para}
+
+\todo{Rick: I found one example but the cardinality information 
+was different than what's stated in the current prose...}
+
+%%runid = 1097, querynum = 46, max_card = 60000, upper_card = 29100, lower_ard = 28800, upper_gen = 4, lower_gen = 5
+%%max_cqt: 230, upper_cqt: 110, lower_cqt: 120, lower_extrapolated_cqt: 110
+%% experimentname: op-pk-30K-100q-idx, pk: true, sec_idx: true
+%% plan 0: ALL:Full Table Scan,ref
+%% plan 1: eq_ref,index,ref
+%% plan 0 at upper_card, plan 1 at lower_card
+\noindent
+\hspace{3ex}{\small\begin{verbatim}
+        SELECT t3.id2, t3.id4, t0.id3, t2.id1 
+        FROM ft_HT3 t2, ft_HT1 t1, ft_HT2 t3, ft_HT1 t0  
+        WHERE  (t2.id2=t1.id2 AND t1.id2=t3.id1 AND t3.id1=t0.id4)
+\end{verbatim}}
+}
+}
+As change point consists of a pair of adjacent Q@Cs, we have two generations
+to consider. Lets start with pairs for which $\hbox{\em
+mingen}(\hbox{\em lower}) > \hbox{\em mingen}(\hbox{\em
+  upper})$. An example is shown in Figure~\ref{fig:posback}. In this case,
+the set of operators in Plan A for the Q@C at cardinality 900K requires at
+least generation 3, whereas the set of operators for Plan B for the Q@C at
+cardinality 910K requires but generation 2. (Note that the generations in
+this example
+are consecutive, but that is not required. Sometimes the generations within a
+pair are wildly different.)
 
@@
 %originally 30pc
-\includegraphics[width=30pc]{figures/gen_benefit.pdf}
-\caption{Cumulative Average Relative Delta over Generations\label{fig:benefit}}
+\includegraphics[width=30pc]{figures/posback.pdf}
+\caption{An Example of a Positive Backward Delta\label{fig:posback}}
 \end{figure*}
 
-The center graph shows the ``net benefit,'' that is, the cumulative average
-relative delta
-of both the suboptimal decision and non-suboptimal decision change
-points. (Note that each point is not simply the sum of the non-suboptimal
-and suboptimal cumulative
-deltas, as the number of change points differ for each generation, but rather the cumulative
-relative delta over all of the change points for that and prior generations.) There is a dramatic
-rise at generation 5, with the rest of the generations presenting a
-relatively smooth growth, moving from negative relative delta (not
-suboptimal) to positive relative delta (suboptimal), crossing around
-generation 10. This highly aggregated result, extracting the 3000-odd
-change-point pairs from almost 100-thousand Q@Cs having the specified
-properties of interest, implies that these four particular \hbox{DBMSes}, taken
-together, might be close to hitting the wall, where adding an operator
-actually slows down the average query.
-
-More specifically, we can't say whether any of these DBMSes have actually
+To examine the wisdom of picking Plan A for 900K (whose time was measured as
+the {\em higher} point shown (the one above the dashed line), we find the closest Q@C
+also having Plan A. There are two illustrated here, at cardinalities 30K and
+2000K, respectively, with the closer one at 30K. (If there is no such Q@C,
+we can't do the extrapolation, and simply remove that change point from
+further consideration.)
+
+We use the slope of the dashed line between the measured query time of Plan
+B at cardinalities 30K and 910K to get an estimated query time at 900k.
+This realizes an estimate of how fast the query using Plan B would have run
+on the database having the variable table with a cardinality of 900K. In
+this particular case, Plan B looks like it would have run {\em faster} (in
+fact, was was faster even at 910K in this particular example).
+
+As illustrated in Figure~\ref{fig:discontinuity}, plans can be
+discontinuous, an example being external merge sort. For that operator, as
+the cardinality increases, the number of passes  will at some point by
+incremented, but the time is linear for a set number of passes.
+Such behavior will, for plan B present at
+the two granularities of 30K and 910K, slightly overestimate the runtime of
+that plan at 900K, and thus {\em underestimate} the penalty of going with
+Plan A rather than Plan B. That said,
+given that we are extrapolating from the actual run time at a very
+close cardinality, our extrapolated estimate should be close.
+
+We term this situation a {\em backward extrapolation}, because we are extrapolating
+backward from a cardinality of 910K to one of 900K.
+
+With this extrapolation, we can compute the {\em relative delta}, defined as the
+measured time of Plan A (the chosen one) at the lower cardinality (900K) minus the
+extrapolated time of Plan B at that same cardinality, divided by the
+measured time of Plan A. The relative delta is thus scaled by original
+measured time at that granularity, and thus has a maximum value of 1.
+
+In this case, the relative delta is positive (the measured time for the plan
+associated with the higher generation is greater than the extrapolated
+time), which indicates that the optimizer chose the {\em wrong plan}: here,
+Plan B should have been chosen.
+
+On the other hand, a {\em negative} relative delta (where the measured
+time for the plan associated with the higher generation is lower than the
+extrapolated time, such as that illustrated in Figure~\ref{fig:negback})
+implies that the additional operator(s) available to the minimally
+applicable generation of the upper plan were indeed beneficial, in that that
+plan was faster. (In such cases, we again divide by the larger
+value, so the minimum is~-1.)
+
+Summarizing, a
+positive relative delta (extrapolated in either the forward or backward direction) implies the presence of suboptimality: the wrong
+plan was chosen by the \hbox{DBMS} generation, perhaps due to the greater number of
+operators available. A negative relative delta (forward or backward) implies the {\em absence} of
+suboptimality: the chosen plan was faster at that cardinality than the one
+chosen at the adjacent cardinality. (Note that this is a slightly more expansive definition
+of suboptimality than that used earlier and illustrated in
+Figure~\ref{fig:query769}.)
+
+There are six orthogonal possibilities for each change point (that
+is, a pair of adjacent Q@Cs), a total of 66,769 pairs/change
+points from the Exploratory and Confirmatory Experiments. (There is also the
+case of a query instance containing a lone Q@C, which we don't consider further.)
+
+\begin{enumerate}
+\item The pair of Q@Cs share the same generation: $\hbox{\em mingen}(\hbox{\em lower}) =
+\hbox{\em mingen}(\hbox{\em upper})$  (62,903 Q@C pairs, 94.2\%), which we don't consider further.
+
+\item The extrapolation yielded a computed query
+time that was negative, which we also drop (10 pairs, 0.01\%).
+
+\item The extrapolation from above and indicating
+no suboptimality, termed a {\em negative backward relative
+  delta}, as
+  exemplified in Figure~\ref{fig:negback}, examined earlier (492 pairs,
+  0.7\%).
+
+\item The extrapolation, termed a {\em positive backward
+    relative delta} and exemplified in Figure~\ref{fig:posback}, indicates
+  a suboptimal plan (583 pairs, 0.9\%).
+
+\item The extrapolation was a {\em negative forward relative delta},
+  indicating no suboptimality (1218 pairs, 1.8\%).
+
+\item A extrapolation was from below (consider Figure~\ref{fig:negback} but with
+  Plan A from generation 5; we would then need to extrapolate from the
+  closest Plan B, which is {\em at a smaller cardinality}), in a {\em forward
+    direction}, to compute a {\em positive forward relative delta},
+  indicating a suboptimal plan (795 pairs, 1.2\%).
+\end{enumerate}
+%1 of rel del: 2, -1 of rel del: 21 (neg backward:3, pos backward:18)
+From these six possibilities, we thus retain (3) and (5), which indicate a
+\hbox{suboptimal} plan at the higher generation (1710 pairs), and (4) and (6), which indicate a
+non-suboptimal plan at the higher generation (1378 pairs).
+
+\subsection{Trends Across \hbox{DBMS} Generations}\label{sec:trends}
+The analysis in the previous section is for {\em a single pair of adjacent Q@Cs for a single query
+running on a specific \hbox{DBMS}},
+providing a {\em relative delta} for the minimally \hbox{applicable} generation. A
+positive relative delta indicates that the later generation chose a
+suboptimal plan; a negative relative delta indicates the later generation
+did not. The relative delta is a percentage difference, and so is not \hbox{affected}
+by the absolute magnitude of the run time nor by the cardinality in
+question. Indeed, because it is a percentage difference, the relative delta
+is not affected by the query nor even which \hbox{DBMS} is involved. We associate
+this relative delta with the later generation, for it is that generation
+which had the choice between the two plans for that query.
+
+Consider how the {\em average} relative delta might behave across successive
+generations.
+The average relative delta is a characterization of the aggregate
+impact of that generation, providing a quantitative estimate of the benefit
+of adding that operator. (We use average so that each point is not impacted
+by the number of pairs over which that point is computed.)  To emphasize,
+the average aggregate delta aggregates over change points, queries, and even
+DBMSes, providing a single average relative delta for each generation.
+
+What does our causal model in Figure~\ref{fig:model}, supported by the
+correlational and regression analyses of the Confirmatory Experiment in
+Sections~\ref{sec:corr} and \ref{sec:regression}, say about this? That
+causal model asserts that as the number of operators increases in
+subsequent, the latent measures in plan space complexity increase, which
+therefore impacts suboptimality, also in a positive direction.
+
+Our Gedanken experiment in
+Section~\ref{sec:gedanken} takes this behavior and predicts that
+as the generations contain successively greater numbers of
+operators, suboptimality will also increase.
+
+If we plot the performance of the DBMS on the {\em y}-axis, say as the
+total time for a workload consisting of a set of queries over a
+prescribed data set, for a sequence of {\em DBMS generations} listed on the
+{\em x}-axis, the average relative delta is in some way a characterization
+of the {\em slope} of this relationship. A negative average relative delta
+implies that the indicated generation is doing a good job, with less
+suboptimality, and so the total workload execution time will go down; a
+positive average relative delta implies that the suboptimal decisions are
+dominating, indicated by the total workload execution time going up for that
+generation.
+
+We expect that the average relative delta for the first few generations will
+be negative, reflecting new operators that improves some plans, a natural
+result of the \hbox{efforts} of \hbox{DBMS} developers to increase performance over
+successive generations of their \hbox{DBMS}.  With a negative slope, the
+performance plot will decrease over successive DBMS generations.
+
+That said, all is not rosy in this picture. Each new operator is applicable
+to a successively smaller portion of the queries, and perhaps over a
+successively smaller portion of the cardinality space.  As already noted,
+our causal model predicts that the prevalence of suboptimality would
+increase as the number of operators available increased.  It seems that even
+with DBMS implementers doing smart things, these two considerations predict
+that the average relative delta (itself a slope of the performance curve)
+will {\em increase}, as suboptimality (a positive relative delta) becomes
+more prevalent.
+
+This analysis suggests then that the performance curve will thus start to
+level off. That raises the
+possibility of the performance curve either asymptotically approaching a horizontal line
+(the first derivative approaching 0), or worse: the first derivative (the
+average relative delta)
+changing to positive, with the average time over the queries we are studying
+actually {\em increasing} with \hbox{DBMS} generation.
+
+So the question comes down to this:  {\em by
+how much} is the increase in efficiency enabled by a new operator (for those
+Q@Cs utilizing a plan containing that operator) greater than the decrease in
+efficiency resulting from suboptimality (for a subset of those Q@Cs)
+resulting from adding that operator? Is there a
+point where the decrease due to suboptimality obviates the increase enabled
+by the new operator: a DBMS generation where the average relative delta
+becomes positive? How close might modern \hbox{DBMSes} be to that limit?
+
+\subsection{Have Modern DBMSes Hit The Wall?}\label{sec:hitthewall}
+We partition the relevant four sets of change points discussed in the
+Section~\ref{sec:usingCP} into two groups: (i)~those
+with a {\em positive} relative delta (either forward or backward), denoting a
+{\em suboptimal decision} by the query optimizer at the later generation
+(corresponding to a higher generation number)
+and (ii)~those with a {\em negative} relative delta (either forward or
+backward) relative delta, indicating a {\em non-suboptimal decision} by the
+query optimizer at the later generation. (We can't state unequivocally that
+the plan is optimal, because there may be yet another plan involving the
+operators within that generation that is even faster.)
+
+Let's first examine those change points for which the query optimizer made a
+good decision, the non-suboptimal change points; see
+Table~\ref{tab:non-subopt}.
+
+The first thing to note is that this data aggregates the results over the
+four \hbox{DBMSes}, which had generations ranging from five to 13. Thus this is the
+first study we are aware of that compares the generational trends of \hbox{DBMSes}
+over quite disparate code bases, thus getting at fundamental trends.
+
+Overall, only a small number of change points, 1378, or about 2\% of the
+total, satisfied the requirements listed above, including having a different
+generation number for the lower and upper plans. In fact, there were no such
+change points for the first three generations nor for the last
+generation. The second column states the number of change points added by
+that generation and the average relative delta across just those change
+points. The last column states the {\em cumulative relative delta} for that
+generation, defined as
+the sum of the average relative delta for those change points associated
+with plans having only operators in its generation, divided by the number of
+change points. This is the relevant number for a generation, as it includes
+all plans that would have been emitted by that generation, using the
+operators at that generation's disposal.
+
+While the numbers jump around quite a bit, especially for the first few
+generations. The thing to focus on is the last column, where the cumulative
+relative delta starting off at a low of
+-0.28 (recall that low negative value is good, as it indicates the
+additional operator was effective at lowering the execution time) and slowly decreasing to -0.225:
+larger negative numbers trending to lower negative
+numbers, indicating decreasing benefits of optimization.
+
+This general behavior matches our prediction arising from the structural causal model that it
+gets harder to squeeze out performance gains as operators are added to the
+\hbox{DBMS} over successive generations.
+
+We now examine those change points for which the query optimizer made a poor
+decision, in that we can surmise that there was a better plan (the one right
+next to it in the adjacent Q@C pair). Table~\ref{tab:subopt} provides the
+same information across the \hbox{DBMS} generations for the suboptimal change
+points: those for which the
+  relative deltas are positive, indicating a poor decision, as the query
+  time increased. While there are
+  still only a small number of change points, there are a greater number of
+  {\em suboptimal} change points, with a relatively greater number showing up at more
+  recent generations. But more strikingly, the cumulative relative delta has
+  the opposite behavior to the non-suboptimal change points: it {\em
+    increases} over the generations (indicating greater suboptimality), starting at 0.08 and ending at 0.25, a
+  value {\em higher} than that of the non-suboptimal change points.
+
+The next-to-last column brings the non-suboptimal and suboptimal together, stating
+the {\em net cumulative relative delta}, gathering the change points
+with plans that would have been generated by that DBMS generation
+together. We see that the net starts out \hbox{at -0.24}, indicating that the new
+operators are {\em decreasing} the query time for the
+workload. Unfortunately, this happy situation starts to deteriorate: with
+each successive generation, the improvement is less. This is as predicted:
+the optimizer is struggling with both more options (plans over the available
+operators) to select from and a diminished opportunity to make a significant
+improvement.
+
+We also see that at Generation 10, the net actually becomes very slightly positive: meaning that the performance
+curve has hit a minimum and is now on its way up, towards slower
+performance. This trend increases in later generations.
+
+\begin{table}[t]
+\tbl{Non-Suboptimal Change Points\label{tab:non-subopt}}{%
+%\begin{center}
+\begin{tabular}{c|c|c|c}
+{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}\\
+& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}\\
+\hline
+1 & --- & --- & 0.00\\
+2 & --- & --- & 0.00\\
+3 & --- & --- & 0.00\\
+4 & 24 & -0.28 & -0.28\\
+5 & 282 &-0.118 & -0.130\\
+6 & 309 &-0.274 & -0.203\\
+7 & 336 & -0.317 & -0.243\\
+8 & 43 & -0.20 & -0.241\\
+9 & 4 & -0.1 & -0.241\\
+10 & 17 &-0.29 & -0.241\\
+11 & 357 & -0.178 & -0.225\\
+12 & 6 & -0.2 & -0.225\\
+13 & 0 & --- & -0.225\\
+{\em Cumulative}  &1378 & ---& -0.225\\
+\end{tabular}
+%\end{center}
+}
+\end{table}
+
+\begin{table}[t]
+\tbl{Suboptimal Change Points\label{tab:subopt}}{%
+%\begin{center}
+\begin{tabular}{c|c|c|c|c|c}
+{\em Generation}&{\em Number of} &{\em Average}&{\em Cumulative}&{\em Net
+  Cumulative}&{\em Relative}\\
+& {\em Change Points} & {\em Relative Delta} & {\em Relative Delta}& {\em
+  Relative Delta}&{\em ``Performance''}\\
+\hline
+1 & --- & ---   & 0.00  &---   &---\\
+2 & --- & ---   & 0.00  &---   &--\\
+3 & --- & ---   & 0.00  &---   &0\\
+4 & 2   & 0.2   & 0.2   &-0.24 &-0.24\\
+5 & 211 & 0.076 & 0.077 &-0.045&-0.285\\
+6 & 86  & 0.16  & 0.102 &-0.103&-0.308\\
+7 & 263 & 0.303 & 0.196 &-0.0800&-0.388\\
+8 & 273 & 0.303 & 0.231 &-0.0256&-0.401\\
+9  & 3  & 0.04  & 0.230 &-0.0257&-0.426\\
+10 & 168& 0.319 & 0.245 &0.008&-0.426\\
+11 & 639& 0.237 & 0.242 &0.0296&-0.396\\
+12 & 53 & 0.35  & 0.245 &0.0346&-0.361\\
+13 & 12 & 0.40  & 0.246 &0.0360&-0.325\\
+{\em Cumulative}& 1710& ---  & 0.246 &0.0360&---\\
+\end{tabular}
+%\end{center}
+}
+\end{table}
+
+The final column integrates the average relative delta to produce a unit-less
+{\em relative performance}, a simulation of how the four DBMSes together
+would have performed (say, total time) on the workload of the Confirmatory
+Experiment, or rather the 1378 change points selected by the criteria in
+Section~\ref{sec:usingCP}. (We emphasize that our experiment estimates just
+the first derivative of the performance graph, which is the net cumulative
+relative delta, and then only for a small
+number of Q@Cs, and then only as a relative measure, between -1 and 1. It is
+thus only very roughly indicative of the {\em shape} of the performance graph.)
+The relative performance gets better (faster, indicated by a falling
+value), but that trend again stops at generation 10, after which the
+relative performance actually gets worse (slower, indicated by a rising value).
+
+\begin{figure}
+    \subfigure[Cumulative Average Relative Delta over Generations]{
+        %\includegraphics[width=20pc]{figures/gen_benefit.pdf}
+        %\includegraphics[scale=0.56]{figures/gen_benefit.pdf}
+        %\includegraphics[width=19pc, height=5.5cm]{figures/gen_benefit.pdf}
+        \includegraphics[scale=0.735]{figures/gen_benefit.pdf}
+        \label{fig:benefit}
+    }
+    \hspace{-0.17in}
+    \subfigure[Performance over Generations]{
+        %\includegraphics[scale=0.6]{overall/machine_pt_re.eps}
+        %\includegraphics[width=12pc]{figures/perf_gen.pdf}
+        %\includegraphics[scale=0.56]{figures/perf_gen.pdf}
+        %\includegraphics[width=14.5pc, height=5.5cm]{figures/perf_gen.pdf}
+        \includegraphics[scale=0.735]{figures/perf_gen.pdf}
+        \label{fig:performance}
+    }
+    \caption{Trend of Performance Improvement over Generations}
+    \label{fig:machine_comp}
+\end{figure}
+
+Figure~\ref{fig:benefit} and \ref{fig:performance} tell the story
+graphically. Figure~\ref{fig:benefit} plots the last column of
+Table~\ref{tab:non-subopt} as the top line (non-suboptimal cumulative
+relative data), the fourth column of Table~\ref{tab:subopt} as the bottom
+line (suboptimal cumulative relative delta), and the fifth column of
+Table~\ref{tab:subopt} as the middle line (net cumulative relative delta)
+in the center. This middle line has a least squares slope of -0.024, which
+is consistent with the transition from a generation being net beneficial to
+actually being slightly suboptimal around generation 10 and more so at later
+generations. Figure~\ref{fig:performance} plots the last column of
+Table~\ref{tab:subopt}, the relative performance across generation. This
+illustrates visually that the performance time falls (gets better) until
+around generation 10, after which adding an operator might have an overall
+detrimental impact, actually slowing down the query.
+
+In summary, the engineering perspective in Section~\ref{sec:gedanken}
+predicts that the overall per-generation time would ``monotonically decrease
+as the generations add query evaluation operators.'' The scientific
+perspective reaches the opposite conclusion: ``the
+per-generation time will either asymptotically approach a horizontal line
+(with the first derivative approaching 0 from below), or worse: the first
+derivative will change to positive, with the per-generation time over the
+queries we are studying actually increasing with \hbox{DBMS}
+generation. In either case, the causal model predicts a point where the
+\hbox{increase} in execution time due to suboptimality obviates the decrease
+enabled by the new operator.'' The results of this study clearly support
+the latter (cf.~Figure~\ref{fig:performance}).
+
+This highly aggregated result, extracting 3000-odd change-point pairs having
+the specified properties of interest stated in Section~\ref{sec:trends} and
+drawn from almost \hbox{100-thousand} Q@Cs implies that these four
+particular \hbox{DBMSes}, taken together, might be close to or have already
+hit the wall, where adding an operator actually slows down the average
+query. And recall the analysis in Section~\ref{sec:CPQ} in which a
+related problem cropped up around generation 8.
+
+It is important to emphasize that we can't say whether any of these DBMSes have actually
 transitioned to where, for this class of queries, the errors of the
@@
 
-We should emphasize all the provisos mentioned earlier. We are looking at
-quite simple queries, over a quite limited range of data, with only a small
-percentage (3\%) of change points identified in our experimental
-protocol. On the other hand, simpler queries are thought to be easier to
-generate good plans than more complex queries (as Hypothesis 2 in
-Section~\ref{sec:interactions} states), relational data is generally much less uniform
-in its structure (we used only integers), values (the values in our
-tables are quite evenly distributed), schema (the schemas of our tables
-are identical), and range of sizes (the smallest table is 1/200-th of the
-largest table), which should minimize the suboptimality.
+We reiterate the provisos mentioned earlier. In all of these
+experiments, we are looking at quite simple queries, over a quite limited
+range of data, with only a small percentage (3\%) of change points
+identified in our experimental protocol. On the other hand, simpler queries
+are thought to be easier to generate good plans than more complex queries
+(as Hypothesis 2 in Section~\ref{sec:interactions} states). Also, relational
+data is generally much less uniform in its (i)~data types (we used only
+integer data), (ii)~values (the values in our tables are quite evenly
+distributed), (ii)~schemas (the schemas of our tables are identical and
+extremely simple), and (iii)~range of table cardinalities (the smallest
+table is 1/200-th of the largest table). The simpliciy of queries, schemas,
+and tables should independently, and certainly in concert, minimize the
+suboptimality.
 
@@
 
-\todo{Rick:discuss implications of generational study}
+The generational study in Section~\ref{sec:diminishing} though implies that
+the challenge in daunting. That study validates the implications of the
+causal model in Figure~\ref{fig:model}, which correctly predict  the almost
+inexorable rise in net cumulative relative delta, which implies that the
+optimizer will eventually hit the wall where it is no longer improving.
 
@@
 
+While there has been a wealth of research over the last forty years on
+engineering approaches and refinements to improve the performance of query
+optimization, this is the first paper to the authors' knowledge that focuses
+on {\em suboptimality} and the first to take a scientific approach (in the
+sense of {\em empirical generalization} towards this important aspect.
+
 This paper makes the following contributions in
-an \hbox{attempt} to gain new understanding of this component.
+an \hbox{attempt} to gain new understanding of this
+component.\todo{Rick:make consistent with list at the beginning of the paper}
 \begin{itemize}
@@
 
-\item Tests {\em seven hypotheses} deductively derived from the predictive causal model. A
+\item Tests {\em six hypotheses} deductively derived from the predictive causal model. A
 correlational analysis and a regression analysis
@@
 qualitatively confirming what was informally known.
-
-\item Provides a {\em space of hypotheses} that in concert through further
-  investigation can refine and support (or not) this model or suggest a
-  better model.
 
@@
 
-\item Thereby identifies {\em specific directions for engineering interventions}.
-
 \end{itemize}
 This paper thus suggests a framework of casual model elaboration and
-directed engineering efforts that could reduce the prevalence of query suboptimality to an acceptably
-low level.
+directed engineering efforts.
 
 \section{Future Work}\label{sec:future}
-There are several directions we wish to take this work. With the model refinements
+There are several directions this work could be take. \todo{RICK IS HERE}With the model refinements
 we propose here, additional directed pointers to engineering efforts should
-emerge.
+emerge.\todo{Rick:Don't say ``we'', but just say the research needs to be done}
 
@@
 many subqueries are themselves suboptimal, does that causally impact whether
-the overall query is suboptimal? Finally, we may be able to determine
+the overall query is suboptimal? It would also be useful to look at the
+moderation effect for Presence of secondary indexes: why precisely do these seem to
+actually decrease plan space complexity, and does this moderation continue
+as queries get more complex? Finally, we may be able to determine
 whether certain query rewrite techniques are employed within each \hbox{DBMS},
@@
 \section{Details on the Experiments}\label{sec:app}
-Table~\ref{tab:run_stat} lists the run statistics of the seven experiments
-  used in this paper. In this appendix we provide more detailed information
-  on the experiments.
-
-Table~\ref{tab:run_stat2} gives some of those details. We'll walk through
-the columns in succession.\todo{Young: please fill in last column}
+Table~\ref{tab:run_stat} in Section~\ref{sec:experiments} lists the run statistics of the seven experiments
+  used in this \hbox{paper}. In this appendix we provide more detailed information
+  on the experiments, walking through the columns in succession of
+Table~\ref{tab:run_stat2}, given below.
+
+The third column identifies the data set(s) used in each experiment, that is, the
+specific tables being queried. There are four data sets, named A, B, C, and
+D.
+
+We first discuss the features shared between the four data sets. As
+introduced in Section~\ref{sec:motivation}, the queries referenced tables
+{\tt ft\_HT1}, {\tt ft\_HT2}, {\tt ft\_HT3}, and {\tt ft\_HT4}. All four
+tables contain four columns, each of type integer. The specific values of
+the rows for all but the first column depend
+on the percentage of skewed data. Section~\ref{sec:datacomplexity} provides
+the algorithm for generating the values for different values of skew; this
+algorithm is used in columns two--four, which for any row will have
+identical values. The first column holds a unique integer starting from 1
+and going to 60K or 2M, for use in an optional primary key.
+
+There was one version of the
+last three tables, for use with MySQL, with cardinality 60K, and one version
+for the rest of the DBMSes, with cardinality 2M.
+
+We generate 200 versions of {\tt ft\_HT1}, termed the {\em variable table}. For MySQL, these version contain
+300, 600, 900, 1200, $\ldots$, 59,700, and 60,000 rows; for the rest of the
+DBMSes, these versions contain 10,000, 20,000, 30,000, $\ldots$, 1,970,000,
+and 2M rows, as introduced in Section~\ref{sec:motivation}.
+
+We now differentiate the data sets, elaborating on the discussion in Sections~\ref{sec:datasets}--\ref{sec:experiments}. Data Set C is perhaps the simplest to
+describe: it specifies no primary key, has no duplicate rows, and has no skew
+(of course, for any of the four tables). Data Set A differs from Data Set C
+only in that there is skew.  As summarized at the end of
+Section~\ref{sec:datacomplexity}, we use five skewness values: 0
+(approximately), 0.001, 0.1, 0.5, and 1.0.
+
+Data Set B is similar to Data Set A, adding the specification of the first
+column as the primary key. And Data Set D is similar to Data Set B, adding
+the specification that the other three columns should each be associated
+with a secondary index, only only that one column.
+We see the confirmatory examined a much larger variation of data sets than
+the exploratory studies.
+
+The next column of Table~\ref{tab:run_stat2} concerns the {\em Lab
+  Shelf}. \hbox{\azdb} utilizes the metaphor of a bookshelf of lab
+notebooks. Here, each shelf is associated with a version of \hbox{\azdb} itself. For the experiments in this paper, we used at various times over
+the last three years lab shelves (that is, program versions) 5.19, 5.2,
+6.0, and 7.1. Versions 5.19 and 5.2 were very similar; both implemented
+TTPv1.  Version 6.0 also implemented TTPv1, but collected more query
+measures that were not relevant for this paper. Version 7.1 implemented
+TTPv2.
+
+The \hbox{\azdb\ system also includes support} for {\em experiment
+  scenarios}, each of which
+is a small amount~ (a few hundred lines) of Java code that actually performs the experiment, such as varying the
+cardinality and running different queries on the data. The only difference
+in the scenario code across these experiments was in accommodating the
+details of the data set (that is, creating secondary indexes and data skew)
+\todo{Young:was this in the scenario code? It seems that this should be
+  outside of that: Rick: Yes, this is part of the scenario code.
+Specifically, if the presence of secondary indexes is true, or the presence of data skew is false, then in the scenario code an experiment subject creates the indexes or makes no skew when populating tables.
+After table population is done, the code studies each query, or detects a change point and makes query executions at that change point.}and adding robustness features, such as using exponential backoff when the
+network connection to the DBMS was temporarily lost.
+
+The bottom line is that while the lab shelf and experiment scenario varied
+somewhat, the only important aspect was the {\em Protocol} column of
+Table~\ref{tab:run_stat}.
+
+We now turn to the fifth column, ``what was examined?'' Here there are just two
+possibilities, all 200 cardinalities or just the cardinalities at which the
+query plan changed. The sixth column, ``what was timed?'', indicates that experiments three and four, Exhaustive and
+Exhaustive with Keys, described in Sections~\ref{sec:discontinuity} and \ref{sec:experiments}, just
+collected query plans, and thus took much less time to run.
+
+The last column states how many Q@Cs the experiments measured (each with 10
+QEs), termed {\em raw}, and how many Q@Cs were retained after the protocol
+implied by the fourth column and listed explicitly in the third column of
+Table~\ref{tab:run_stat} dropped query executions and Q@Cs via its many
+  sanity checks.
 
@@
 \begin{tabular}{c|c|c|c|c|c|c}
-& {\em Experiment}& {\em Data Sets} & {\em Lab Shelves} & {\em What was} &{\em Was query}&{\em Number of}\\
-&                 & {\em Used}      &                   & {\em Examined?}&{\em Timed?}   &{\em Retained (Raw) Q@Cs}\\
+& {\em Experiment}& {\em Data Sets} & {\em Lab Shelves} & {\em What was} &{\em Was Query}&{\em Number of}\\
+&                 & {\em Used}      &                   & {\em Examined?}&{\em Timed?}   &{\em Retained (Raw)}\\
+&&&&&& {\em Q@Cs}\\
 \hline
-1 & Monotonicity 	& A & 6.0     &all cardinalities&yes& 12000 (12000)\\
-2 & Initial Exhaustive 	& A & 5.19 + 6.0&all cardinalities&yes& 27948 (32000)\\
-3 & Exhaustive 		& A & 7.1     &all cardinalities&no & 29515 (32000)\\
-4 & Exhaustive with Keys & B& 6.0     &change points&no& --- \\
-5 & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8171 (8842)\\
-6 & Exploratory 	& A + B&7.1   &change points&yes& 12100 (12560)\\
-7 & Confirmatory 	&A + B + C +D&7.1&change points&yes& 94502 (99558)\\
+1 & Monotonicity 	& A & 6.0     &all cardinalities&yes& 12,000 (12,000)\\
+2 & Refined Monotonicity& A & 5.19 + 6.0&all cardinalities&yes& 27,948 (32,000)\\
+3 & Exhaustive 		& A & 7.1     &all cardinalities&no & 29,515 (32,000)\\
+4 & Exhaustive with Keys & B& 6.0     &change points   &no& --- \\
+5 & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8,171 (8,842)\\
+6 & Exploratory 	& A + B&7.1   &change points   &yes& 12,100 (12,560)\\
+7 & Confirmatory 	&A + B + C +D&7.1&change points&yes& 94,502 (99,558)\\
+\multicolumn{2}{c|}{\em Total}         &            & &               &   & 184,236 (196,960)\\
 \end{tabular}
@@
 
-The third column states the data set used in each experiment, that is, the
-specific tables being queried. There are four data sets, named A, B, C, and
-D.
-
-We first discuss the features shared between the four data sets. As introduced in
-Section~\ref{sec:motivation}, the queries referenced tables {\tt ft\_HT1},
-{\tt ft\_HT2}, {\tt ft\_HT3}, and {\tt ft\_HT4}. All four tables contain four
-columns, each of type integer. The specific values of the
-rows for all four columns depend on the percentage of skewed
-data. Section~\ref{sec:datacomplexity} provides the algorithm for generated the
-values for different values of skew; this algorithm is used in each column,
-so all four columns in any row will have identical values.\todo{Young: is
-  this true?}
-
-There was one version of the
-last three tables, for use with MySQL, with cardinality 60K, and one version
-for the rest of the DBMSes, with cardinality 2M.
-
-We generate 200 versions of {\tt ft\_HT1}, termed the {\em variable table}. For MySQL, these version contain
-300, 600, 900, 1200, $\ldots$, 59,700, and 60,000 rows; for the rest of the
-DBMSes, these versions contain 10,000, 20,000, 30,000, $\ldots$, 1,970,000,
-and 2M rows, as introduced in Section~\ref{sec:motivation}.
-
-We now differentiate the data sets, elaboring on the discussion in Section~\ref{sec:datasets}--\ref{sec:experiments}. Data Set C is perhaps the simplest to
-describe: it specifies no primary key, has no duplicate rows, and has no skew
-(of course, for any of the four tables). Data Set A differs from Data Set C
-only in that there is skew.  As summarized at the end of
-Section~\ref{sec:datacomplexity}, we use five skewness values: 0
-(approximately), 0.001, 0.1, 0.5, and 1.0.
-
-Data Set B is similar to Data Set A, adding the specification of the first
-column as the primary key. And Data Set D is similar to Data Set B, adding
-the specification that the other three columns should each be associated
-with a secondary index, only only that one column.
-We see the confirmatory examined a much larger variation of data sets than
-the exploratory studies.
-
-The next column of Table~\ref{tab:run_stat2} concerns the {\em Lab
-  Shelf}. \hbox{\azdb} utilizes the metaphor of a bookshelf of lab
-notebooks. Here, each shelf is associated with a version of \hbox{\azdb} itself. For the experiments in this paper, we used at vearious times over
-the last three years labh shelves (that is, program versions) 5.19, 5.2,
-6.0, and 7.1. Versions 5.19 and 5.2 were very similar; both implemented
-TTPv1.  Version 6.0 also implemented TTPv1, but collected more query
-measures that were not relevant for this paper. Version 7.1 implemented
-TTPv2.
-
-The \hbox{\azdb\ system also includes support} for {\em experiment scenarios}, which
-are Java code that actually performs the experiment, such as varying the
-cardinality and running different queries on the data. The only difference
-in the scenario code across these experiments was in accommodating the
-details of the data set (that is, creating secondary indexes and data skew)
-and adding robustness features, such as using exponential backoff when the
-network connection to the DBMS was temporarily lost.
-
-The bottom line is that while the lab shelf and experiment scenario varied
-somewhat, the only important aspect was the {\em Protocol} column of
-Table~\ref{tab:run_stat}.
-
-We now turn to the fifth column, ``what was examined?'' Here there are just two
-possibilities, all 200 cardinalities or just the cardinalities at which the
-query plan changed. The sixth column, ``wheat was timed?'', indicates that experiments three and four, Exhaustive and
-Exhaustive with Keys, described in Sections~\ref{sec:discontinuity} and \ref{sec:experiments}, just
-collected query plans, and thus took much less time to run.
-
 The following sets of queries were used in the seven experiments.
 
-\todo{Young: how does the following look? Any changes needed?}
 \begin{description}
-\item[QSa] 100 queries over the four tables, generated as described in Section~\ref{sec:querycomplexity}
-\item[QSb] 100 queries, generated the same way
-\item[QSc] 100 queries
-\item[QSd] 100 queries
-\item[QSe] 100 queries
-\item[QSf] 100 queries
-\item[QSg] 100 queries
-\item[QSh] 100 queries
-\item[QSi] 100 queries
-\item[QSj] 100 queries
-\item[QSl] A primary key query set consisting of 230 queries from {\em
-  QSa}--{\em QSf}
-\item[QSm] A primary key query set consisting of 160 queries: 40 queries
-  drawn from each of {\em QSg}--{\em QSj}
-%\item[QSlm] A primary key query set consisting of 390 queries from QSl and QSm
-\item[QSn] A primary key query set 4 with 110 new queries
-\item[QSo] A no-skew query set consisting of 100 queries without aggregates for no skew
-\item[QSp] A subquery query set consisting of 100 queries, each with a subquery
-\item[QSq] A subquery query set consisting of 100 queries, each with a subquery
-\item[QSr] A primary key query set consisting of 100 queries drawn from {\em
-  QSl} and {\em QSm}
-\item[QSab'] 100 queries drawn from {\em QSa} and {\em QSb} for which the where
-     selection involved a primary key column
-\todo{Young what does this mean: ``This query set was revised to QSI and
-  QSm.''? Als, it doesn't seem like this set is used anywhere...}
+\item[QSa] 100 queries over the four tables, generated as described in Section~\ref{sec:querycomplexity} (on Data Set A)
+\item[QSb] 100 queries (on Data Set A)
+\item[QSc] 100 queries (on Data Set A)
+\item[QSd] 100 queries (on Data Set A)
+\item[QSe] 100 queries (on Data Set A)
+\item[QSf] 100 queries (on Data Set A)
+\item[QSg] 100 queries (on Data Set A)
+\item[QSh] 100 queries (on Data Set A)
+\item[QSi] 100 queries (on Data Set A)
+\item[QSj] 100 queries (on Data Set A)
+\item[QSl] A query set consisting of 230 queries drawn from \hbox{\em
+  QSa}--{\em QSf} (on Data Set B)
+\item[QSm] A query set consisting of 160 queries: 40 queries
+  drawn from each of {\em QSg}--{\em QSj} (on Data Set B)
+\item[QSn] A query set consisting of 110 new queries (on Data Set B)
+\item[QSo] A query set consisting of 100 queries without aggregates (on Data
+  Set C)\todo{Young: are these new queries, or drawn from prior query sets? Rick: New queries.}
+\item[QSp] A subquery query set consisting of 100 queries, each with a subquery (on Data Sets A, B, and D)
+\item[QSq] A subquery query set consisting of 100 queries, each with a subquery (on Data Set D)
+\item[QSr] A query set consisting of the 390 queries in {\em QSl} + {\em QSm} (on Data Set B)
+%\item[QSs] A primary key query set consisting of 100 queries drawn from {\em QSr}
+\item[QSs] A query set consisting of 100 queries drawn from {\em QSr} (on Data Set D)
 \end{description}
-
-\todo{Rick: discuss last column}
 
@@
 
-Experiment 2 (Initial Exhaustive) used the first 50 queries from {\em QSa}
+Experiment 2 (Refinded Monotonicity) used the first 50 queries from {\em QSa}
 for the three other DBMSes plus the ten queries for MySQL from Experiment 1, for
@@
 Experiment 4 (Exhaustive with Keys) used the first 50 queries from {\em QSa}
-for all four DBMSes, for a total of 200  query instances.
+for the four DBMSes, for a total of 200  query instances.
 
@@
 one DBMS), the first 20 queries from {\em QSa}--{\em QSf} (for another DBMS), the first
-10 (primay key) queries from {\em QSa} (for all four DBMSes), and the first 10
+10 (primary key) queries from {\em QSa} (for the four DBMSes), and the first 10
 (non-primary key) queries from {\em QSa} (for the other two DBMSes), for a total of 780 query instances.
@@
 Experiment 6 (Exploratory) used {\em QSa} and {\em QSb}, plus the first 100 (primary
-key) queries from {\em QSl} and {\em QSm}, for all four DBMSes, for a total of 1200 query
+key) queries from {\em QSr}, for the four DBMSes, for a total of 1200 query
 instances.
 
-Experiment 7 (Confirmatory) used {\em QSc}--{\em QSr}, along with {\em QSp} for primary key and
-subquery, {\em QSp} for subquery, {\em QSp} and {\em QSq} for primary key and secondary index and
-subquery), all across all four DBMSes, for a total of 7,640 query instances.
+Experiment 7 (Confirmatory) used {\em QSc}--{\em QSj},
+along with {\em QSr} except the (first 100) queries included in Experiment 6,
+{\em QSn} for primary key (for two runs, or 220 queries),
+{\em QSo} for no data skew,
+{\em QSp} for primary key and subquery,
+{\em QSp} for subquery,
+{\em QSp} and {\em QSq} for primary key and secondary index and subquery,
+{\em QSs} for primary key and secondary index,
+all across the four DBMSes, for a total of 7,640 query instances.
 

### 2017-03-09 mlppi-comsis17/paper.tex
ComSIS paper

@@
 
-In this section we describe in detail each phase of the wizard using the query set $Q$ \hbox{provided} in Section~\ref{sec:intro}.
+As illustrated in Figure~\ref{fig:arch}, 
+the MLPPI wizard has the three phases for a final partitioning recommendation. 
+In this section we delve into each phase. For better exposition we use the running 
+query set $Q$ \hbox{provided} in Section~\ref{sec:intro}.
 
@@
 
-Algorithm~\ref{algo:pre_phase} sketches 
-the steps in the preprocessing phase. 
-This phase consists of a total of six steps.
+The first phase is called {\em Preprocessing} 
+to extract query predicates and check any constraint in partitioning. 
+This phase consists of a total of six steps, as shown in Algorithm~\ref{algo:pre_phase}. 
+%We elaborate on each step in the subsequent section.
 
@@
 
-The first step of this phase is to simplify the predicate(s) of each query 
-by removing redundant conditions. 
+The first step is to simplify the predicates of queries in a given workload.
+In other words, we remove redundant conditions among the queries. 
 For instance, if a query predicate has the condition of 
 ``{\tt LO\_DISCOUNT IN (1, 4, 5) AND LO\_DISCOUNT IN (2, 3, 4, 5)}'', 
-then the predicate can be simplified as ``{\tt LO\_DISCOUNT IN (4, 5)}.'' 
-This can be done via an API call to the DBMS server. 
-In the running example this simplification is not required.
+then the predicate can be simplified as \linebreak ``\hbox{{\tt LO\_DISCOUNT}} {\tt IN (4, 5)}.'' 
+This can be done via an API call to the database server. 
+This simplification is not required for the running example.
 
@@
 
-The following step is to gather the simplified predicates and 
-then obtain ranges from the collected predicates. 
-This can also be done through an API call to the server. 
+In the subsequent step we gather the simplified predicates and 
+then extract ranges from the collected predicates. 
+This task can also be conducted through an API call to the server. 
 The query predicate is of the form 

### 2017-03-10 pt_characterization/repeatability_test/paper.tex
Chracterizing draft

--- 2017-02-23 pt_characterization/recent_data/paper.tex
@@
 \begin{document}
-\title{Histogram on Recent Runs by EMPv5}
+\title{A Study of Experiment Repeatability}
 
@@
 \section{Description}
-This document presents histograms based on the most recent runs 
-of a program under test, called {\em INC}, over its increasing task lengths.
+This document presents histograms rendered based on empirical data 
+of a program under test, called {\em INC}. 
+Our experiment increases a task length of {\em INC} 
+from one second to 4096 seconds by a factor of two.
+We already ran this experiment last month (February 2017), 
+but we reran the experiment this month (March 2017) 
+to check if experiment repeatability holds true.
 
@@
 \begin{tabular}{|p{2cm}|p{3cm}|p{6cm}|p{4cm}|} \hline
-Machine & Task Length & Description & Time Length\\ \hline
-{\tt sodb9} &  INC1$\sim$INC1024 & Runs of 300 samples & 2017-02-09 $\sim$ 2017-02-10 / 2017-02-12 $\sim$ 2017-02-19\\ \hline
-{\tt sodb10} & INC2048 & Runs of 300 samples & 2017-02-12 $\sim$ 2017-02-20\\ \hline
+Machine & Task Length (sec) & Description & Experiment Period\\ \hline
+{\tt sodb9} &  INC1$\sim$INC512 & Each run with 1000 samples & 2017-03-02 $\sim$ 2017-03-07\\ \hline
+%{\tt sodb10} & INC2048 & A run of 300 samples & 2017-02-12 $\sim$ 2017-02-20\\ \hline
+%{\tt sodb12} & INC4096 & A run of 300 samples & 2017-02-12 $\sim$ 2017-02-27\\ \hline
+%{\tt sodb12} & INC8192 & Two runs, each with 40 samples & 2017-01-16 $\sim$ 2017-01-20 / 2017-01-25 $\sim$ 2017-01-29\\ \hline
+%{\tt sodb12} & INC16384 & Two runs, each with 40 samples & 2017-01-05 $\sim$ 2017-01-13 / 2017-01-29 $\sim$ 2017-02-06\\ \hline 
 \end{tabular}

### 2017-03-10 pt_characterization/repeatability_test/sodb9/res.tex
Chracterizing draft

--- 2017-02-23 pt_characterization/recent_data/sodb9/res.tex
@@
 	}
-	\caption{ET Histograms of INC16 ... INC128~\label{fig:s9_et_hist2}}
-\end{figure}
-
-\begin{figure}[hp!]
-	\centering
 	\subfigure[ET frequency on INC256]{
 		\includegraphics[scale=0.43]{sodb9/256_sec_et_hist_v5.eps}
-		\label{fig:inc256_hist_v5}
+		\label{fig:inc256_et_hist_v5}
 	}
@@
 		\includegraphics[scale=0.43]{sodb9/512_sec_et_hist_v5.eps}
-		\label{fig:inc512_hist_v5}
+		\label{fig:inc512_et_hist_v5}
 	}
-	\subfigure[ET frequency on INC1024]{
-		\includegraphics[scale=0.43]{sodb9/1024_sec_et_hist_v5.eps}
-		\label{fig:inc1024_hist_v5}
-	}
-	\subfigure[ET frequency on INC2048]{
-		\includegraphics[scale=0.43]{sodb10/2048_sec_et_hist_v5.eps}
-		\label{fig:inc2048_hist_v5}
-	}
-	\caption{ET Histograms of INC256 ... INC2048~\label{fig:s9_et_hist3}}
+%	\subfigure[ET frequency on INC1024]{
+%		\includegraphics[scale=0.43]{sodb9/1024_sec_et_hist_v5.eps}
+%		\label{fig:inc1024_et_hist_v5}
+%	}
+	\caption{ET Histograms of INC128 ... INC512~\label{fig:s9_et_hist3}}
 \end{figure}
+
+%\pagebreak
+%
+%\begin{figure}[hp!]
+%	\centering
+%	\subfigure[ET frequency on INC2048]{
+%		\includegraphics[scale=0.43]{sodb10/2048_sec_et_hist_v5.eps}
+%		\label{fig:inc2048_et_hist_v5}
+%	}
+%	\subfigure[ET frequency on INC4096]{
+%		\includegraphics[scale=0.43]{sodb12/4096_sec_et_hist_v5.eps}
+%		\label{fig:inc4096_et_hist_v5}
+%	}
+%	\subfigure[ET frequency on INC8192]{
+%		\includegraphics[scale=0.43]{sodb12/8192_sec_et_hist2_v5.eps}
+%		\label{fig:inc8192_et_hist_v5}
+%	}
+%	\subfigure[ET frequency on INC16384]{
+%		\includegraphics[scale=0.43]{sodb12/16384_sec_et_hist2_v5.eps}
+%		\label{fig:inc16384_et_hist_v5}
+%	}
+%	\caption{ET Histograms of INC2048 ... INC16384~\label{fig:s9_et_hist4}}
+%\end{figure}
 
@@
 	}
-	\subfigure[PT frequency on INC1024]{
-		\includegraphics[scale=0.43]{sodb9/1024_sec_pt_hist_v5.eps}
-		\label{fig:inc1024_hist_v5}
-	}
-	\subfigure[PT frequency on INC2048]{
-		\includegraphics[scale=0.43]{sodb10/2048_sec_pt_hist_v5.eps}
-		\label{fig:inc2048_hist_v5}
-	}
-	\caption{PT Histograms of INC256 ... INC2048~\label{fig:s9_pt_hist3}}
+%	\subfigure[PT frequency on INC1024]{
+%		\includegraphics[scale=0.43]{sodb9/1024_sec_pt_hist_v5.eps}
+%		\label{fig:inc1024_hist_v5}
+%	}
+	\caption{PT Histograms of INC256 ... INC512~\label{fig:s9_pt_hist3}}
 \end{figure}
+
+%\begin{figure}[hp!]
+%	\centering
+%	\subfigure[PT frequency on INC2048]{
+%		\includegraphics[scale=0.43]{sodb10/2048_sec_pt_hist_v5.eps}
+%		\label{fig:inc2048_hist_v5}
+%	}
+%	\subfigure[PT frequency on INC4096]{
+%		\includegraphics[scale=0.43]{sodb12/4096_sec_pt_hist_v5.eps}
+%		\label{fig:inc4096_hist_v5}
+%	}
+%	\subfigure[PT frequency on INC8192]{
+%		\includegraphics[scale=0.43]{sodb12/8192_sec_pt_hist2_v5.eps}
+%		\label{fig:inc8192_hist_v5}
+%	}
+%	\subfigure[PT frequency on INC16384]{
+%		\includegraphics[scale=0.43]{sodb12/16384_sec_pt_hist2_v5.eps}
+%		\label{fig:inc16384_hist_v5}
+%	}
+%	\caption{PT Histograms of INC2048 and INC16384~\label{fig:s9_pt_hist4}}
+%\end{figure}

### 2017-03-10 pt_characterization/recent_data/sodb12/et_hist_sodb12.r

@@ -8,7 +8,7 @@ x <- subset(x, x$ITERNUM > 40)
 x <- subset(x, x$ITERNUM != 84 & x$ITERNUM != 95 & x$ITERNUM != 105 & x$ITERNUM != 116 & x$ITERNUM != 45 & x$ITERNUM != 56 & x$ITERNUM != 66 & x$ITERNUM != 77)
 x_up = mean(x$METIME) + 2*sd(x$METIME)
 x_dn = mean(x$METIME) - 2*sd(x$METIME)
-binsize=10
+binsize=20
 x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
 nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
 h = hist(x$METIME, right=F, breaks=nbins,plot=F)
@@ -17,7 +17,7 @@ xmax <-max(x$METIME)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
 ymax <- max(h$counts)
-ymax <- 10
+ymax <- 20
 plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-50,xmax+50),col="blue", main='ET frequency on INC8192', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='ET (ms)', ylab=expression('Frequency'))

### 2017-03-10 pt_characterization/repeatability_test/sodb9/et_hist_sodb9.r

@@ -1,5 +1,5 @@
 ### sodb9
-numSamples <- 300
+numSamples <- 1000
 x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
 x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
@@ -54,7 +54,7 @@ nbins <- ceiling((max(x$METIME)-min(x$METIME)) / binsize)
 h = hist(x$METIME, right=F, breaks=nbins,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-ymax <- 80
+ymax <- 300
 xmin <-min(x$METIME)
 xmax <-max(x$METIME)
 plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-2,xmax+2),col="blue", main='ET frequency on INC4', 
@@ -68,7 +68,6 @@ x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
 postscript("8_sec_et_hist_v5.eps")
 binsize=1
-x <- subset(x, x$ITERNUM != 27)
 x_up = mean(x$METIME) + 2*sd(x$METIME)
 x_dn = mean(x$METIME) - 2*sd(x$METIME)
 x = subset(x, x$METIME >= x_dn & x$METIME <= x_up)
@@ -78,7 +77,7 @@ xmin <-min(x$METIME)
 xmax <-max(x$METIME)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-ymax <- 150
+ymax <- 400
 plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-binsize-5,xmax+binsize+1),col="blue", main='ET frequency on INC8', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
 xlab='ET (ms)', ylab=expression('Frequency'))
@@ -98,7 +97,7 @@ xmin <-min(x$METIME)
 xmax <-max(x$METIME)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-ymax <- 100
+ymax <- 400
 plot(h, freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-3,xmax+2),col="blue", main='ET frequency on INC16', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
[…]

### 2017-03-10 pt_characterization/repeatability_test/sodb9/pt_hist_sodb9.r

@@ -1,5 +1,5 @@
 ### sodb9
-numSamples <- 300
+numSamples <- 1000
 x = read.csv(file="1_sec.dat",head=TRUE,sep="\t")
 x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
@@ -33,7 +33,7 @@ nbins <- ceiling((max(x$PRTIME)-min(x$PRTIME)) / binsize)
 h = hist(x$PRTIME, right=F, breaks=nbins,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-ymax <- 250
+ymax <- 700
 xmin <-min(x$PRTIME)
 xmax <-max(x$PRTIME)
 plot(h, axes = TRUE,freq=TRUE,ylim=c(0,ymax), xlim=c(xmin-1,xmax+1),col="green", main='PT frequency on INC2', 
@@ -53,7 +53,7 @@ x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
 h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-ymax <- 200
+ymax <- 400
 binsize=1
 plot(h, ylim=c(0,ymax), axes = TRUE,freq=TRUE,xlim=c(min(x$PRTIME)-2,max(x$PRTIME)+4),col="green", main='PT frequency on INC4', 
 sub=paste("(n=",nrow(x),", bin_size=",binsize,"ms)",sep=""), 
@@ -64,14 +64,13 @@ x = read.csv(file="8_sec.dat",head=TRUE,sep="\t")
 x <- subset(x, x$ITERNUM <= numSamples)
 setEPS()
 postscript("8_sec_pt_hist_v5.eps")
-x <- subset(x, x$ITERNUM != 27)
 x_up = mean(x$PRTIME) + 2*sd(x$PRTIME)
 x_dn = mean(x$PRTIME) - 2*sd(x$PRTIME)
 x = subset(x, x$PRTIME >= x_dn & x$PRTIME <= x_up)
 h = hist(x$PRTIME, right=F,breaks=max(x$PRTIME)-min(x$PRTIME)+1,plot=F)
 ymax <- max(h$counts)
 ymax <- ceiling(ymax/100)*100
-ymax <- 100
+ymax <- 400
 binsize=1
[…]

### 2017-03-10 mlppi-comsis17/paper.tex
New updates on ComSIS paper

@@
 then extract ranges from the collected predicates. 
-This task can also be conducted through an API call to the server. 
+This task can also be performed through an API call to the server. 
 The query predicate is of the form 
@@
 \end{center}
-where \form{\tt variable} is a field from {\tt LINEORDER}, 
+where \form{\tt variable} is a field (column) from {\tt LINEORDER}, 
 \form{\tt op} is in 
@@
 
-In turn, the wizard constructs a bi-directional {\em map} between a query and 
-associated predicates, so that for each query the corresponding predicate(s) can be 
-found directly in the map, and vice versa. 
-In the example, a map, called $M$1, is built 
-between $Q$ and $P$, and $M$1 is filled with the following entries:
+Next, we construct a bi-directional {\em map} between a query and 
+associated predicates, so that for each query the corresponding predicate(s) 
+can be found directly in the map, and vice versa. 
+In the example, we get a map, called $M$1, built between $Q$ and $P$. 
+$M$1 gets filled with the following entries:
 
@@
 
-Next, we gather fields referenced by predicates in $P$, extract 
-ranges from the predicates, and group the ranges on each of the fields. 
+For each field referenced by the predicates, 
+we then build a set of distinct ranges on that field. 
+Specifically, we gather fields referenced by predicates in $P$, extract 
+ranges from the predicates, and group the ranges of each of the referenced fields. 
 Assume $I$ to be the list of fact table fields referenced by $P$. 
-In the workload $Q$, we can see that there are two fields used by $P$: 
-{\tt LO\_DISCOUNT}, {\tt LO\_QUANTITY}. 
-These are added to $I$.
-
-After that, we construct a set of distinct ranges on each field in $I$, using the predicates in $P$. 
-Then, the whole range set becomes a kind of two-dimensional array, called $R$. 
-The array $R$ captures the range representation of the predicates. 
+In the workload $Q$ we add to $I$ two fields, {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}, used by $P$. 
+The whole range set then becomes a kind of two-dimensional array, called $R$, capturing 
+the range representation of the gathered predicates. 
 Each entry under a field in $R$ represents a pair of (start and end) values forming the range. 
@@
 ranges can be consolidated as \range{4, 5}. 
-In the continuing example the range set $R$ is formed by the entries below.
+In the continuing example the field-wise distinct range set ($R$) is formed by the entries below.
 
@@
 
-Then, each predicate can be mapped to its associated ranges in $R$. That is, 
-we can have another map, called $M$2, between $P$ and $R$. 
+We then build another map, called $M$2, between $P$ and $R$. 
+That is, we associate each predicate in $P$ with a corresponding range in $R$. 
 Let $R$[$i$, $j$] denote the interval value for the $i$-th field and $j$-th range in $R$. 

### 2017-03-13 mlppi-comsis17/paper.tex
Section 2.2 updates

@@
 For \hbox{example}, only partitions~1 and~2 are needed for the query 
-``{\tt SELECT * FROM LINEORDER WHERE LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
+``{\tt SELECT * FROM LINEORDER WHERE} \linebreak {\tt LO\_DISCOUNT >= 7 AND} {\tt LO\_QUANTITY <= 30}.'' 
 Similarly, partitions~1 and~5 are sufficient to answer the query 
-``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}.'' 
-\linebreak The respective sizes of the partitions are a factor of the data distribution. 
+``{\tt SELECT * FROM LINEORDER WHERE} \linebreak {\tt LO\_QUANTITY < 25}.'' 
+The respective sizes of the partitions are a factor of the data distribution. 
 To see MLPPI in greater detail, check our references~\cite{klindt09mlppi,sinclair:ppi}.
@@
 \paragraph{Challenge:}
-Consider a query set $Q$ having two queries $q$1 and $q$2 on the SSB, as shown below.
+Consider the below query set $Q$ having two queries $q$1 and $q$2 on the SSB.
 \vspace{-.05in}
@@
 A DBA may focus on the {\tt LINEORDER} fact table and 
-the predicates involving {\tt LINEORDER} fields only. 
+the predicates involving \linebreak {\tt LINEORDER} fields only. 
 He can figure out the following five constraints, identified by the query number 
@@
 \end{center}
-The above expression minimizes the size of the partitions 
+This expression minimizes the size of the partitions 
 by focusing exactly on the values required to satisfy the constraints 
@@
 
-\vspace{-.1in}
+%\vspace{-.1in}
 
@@
 
-%\vspace{-.3in}
-
 \subsubsection{Query Simplification}
@@
 
-For each field referenced by the predicates, 
-we then build a set of distinct ranges on that field. 
+Now we build a set of distinct ranges for each field referenced by the predicates. 
 Specifically, we gather fields referenced by predicates in $P$, extract 
 ranges from the predicates, and group the ranges of each of the referenced fields. 
-Assume $I$ to be the list of fact table fields referenced by $P$. 
+Suppose $I$ to be a list of fact table fields referenced by $P$. 
 In the workload $Q$ we add to $I$ two fields, {\tt LO\_DISCOUNT} and {\tt LO\_QUANTITY}, used by $P$. 
-The whole range set then becomes a kind of two-dimensional array, called $R$, capturing 
-the range representation of the gathered predicates. 
+The whole range of the gathered predicates can be represented by a kind of two-dimensional array, called $R$.
+
 Each entry under a field in $R$ represents a pair of (start and end) values forming the range. 
-In the range representation, infinity ($\infty$) or -infinity (-$\infty$) can 
-be used for unbounded ranges. 
-``['' or ``]'' are used for closed ranges, and ``('' or ``)'' for open ranges. 
+When it comes to range representation for an entry, 
+``['' and ``]'' are used for closed ranges while ``('' and ``)'' for open ranges. 
+Also, infinity ($\infty$) and -infinity (-$\infty$) can be used for unbounded ranges. 
 In particular, {\tt IN} predicate is represented by multiple ranges. 
@@
 
-We then build another map, called $M$2, between $P$ and $R$. 
+In turn, we build another map, called $M$2, between $P$ and $R$. 
 That is, we associate each predicate in $P$ with a corresponding range in $R$. 
@@
 
-Since there could be multiple predicates on the same field across queries, 
-extracted ranges may overlap each other. 
-The overlapping ranges must be broken without any common portion, so that 
-a pair of consecutive, \hbox{non-overlapping} ranges can be 
-considered for a merge in further phases.
-
-Algorithm~\ref{algo:split_overlap} describes the way of breaking 
-the \hbox{overlap} of ranges.
-
-%\vspace{-.1in}
-
-\begin{algorithm}[h]
+Since there may be multiple predicates on the same field across queries, 
+extracted ranges for the field may overlap each other. 
+The overlapping ranges must be broken without any common portion, in order 
+that we can consider a pair of consecutive, \hbox{non-overlapping} ranges for 
+a merge in subsequent phases.
+
+\begin{algorithm}[t]
 \caption{Splitting Overlapping Ranges}
@@
 
-\vspace{-.1in}
-
-\noindent For each field, the tool gets its associated ranges and sorts them 
-by start value. 
-It then checks whether or not adjacent ranges overlap each other.
-If that is the case, then we make a split between the ranges. 
+Algorithm~\ref{algo:split_overlap} describes the way of breaking 
+the \hbox{overlap} of ranges. 
+For each field, the algorithm 
+gets associated ranges of each field and sorts them by start value. 
+We then check whether or not adjacent ranges overlap each other.
+If that is the case, we make a split between the ranges. 
 In the example, the split is applied on the 
-overlapping ranges, \lopenrange{\infty, 30} and \range{25, 35}, under {\tt LO\_QUANTITY}. 
-Subsequently, a common, intermediate range (\range{25, 30}) is newly created and inserted into the range set 
+overlapping ranges, \lopenrange{\infty, 30} and \range{25, 35} in {\tt LO\_QUANTITY}. 
+Subsequently, a common (or intermediate) range (\range{25, 30}) is created and inserted into the range set 
 belonging to the {\tt LO\_QUANTITY} field, in order to fill the gap. 
-Then, $R$ in the example is shown as follows.
-
-\
+$R$ is shown in the following.
+
 {\it $R$}:
@@
 new range sets, \{$R$[2, 1], $R$[2, 2]\} and \{$R$[2, 2], $R$[2, 3]\}, respectively. 
-Of course, the range sets mapped to $p$1 and $p$2 referencing {\tt LO\_DISCOUNT} remain unchanged. 
-As a result, in the example we obtain the updated $M$2 as follows:
-
+Of course, the range sets corresponding to $p$1 and $p$2 referencing {\tt LO\_DISCOUNT} remain unchanged. 
+As a result, in the running example we obtain the updated $M$2 in the following.
+
+\pagebreak
 $M$2:
@@
 
-%\vspace{.1in}
-
 \subsubsection{Field Count Limit Check}
 
-%The DBMS or user may impose a limit on the number of fields in the final 
-%MLPPI recommendation. 
-At present, the Teradata DBMS has a limit (64) of the fields 
-that can be used in an MLPPI definition.
+One question that occurs is what if a DBMS has 
+a limit of fields that can be used for a partitioning definition. 
+(In the case of Teradata DBMS, up to 64 fields can be allowed for 
+an MLPPI definition.)
 If the number of fields present in $R$ exceeds the field count limit, 
@@
 and thus, this step will not be executed.)
-To choose victim fields, the tool computes the weighted sum of 
-{\em query cost} of queries regarding each field. 
+To choose victim fields, our tool computes the weighted sum of 
+{\em query cost}, to be discussed in Section~\ref{sec:opt_phase}, 
+of queries regarding each field. 
 The sum can be obtained by adding up every query cost on an MLPPI using 
 only the ranges under the field. 
-(We will cover how to measure the query cost in Section~\ref{sec:opt_phase}.) 
-We incrementally discard a field with the highest query cost sum 
+We incrementally discard a field with the largest sum of query cost 
 until the limit is reached. 
-Accordingly, we can update the existing $M$2. 
-One might say that the MLPPI may not be exploited if too many ranges, 
-exceeding a partition count limit (or 65,536 as of now) are found in one field. 
+Correspondingly, we can update the existing $M$2. 
+One might say that such an MLPPI may not be exploited if too many ranges 
+over a partition count limit (65,536 in Teradata DBMS) are found in one field.
 But we assume that such an extreme case is not expected. Even if 
-it happens, we can utilize the merges of consecutive ranges to make 
-the MLPPI feasible.
+such a case happens, the merges of consecutive ranges can make the MLPPI feasible.
 
@@
 
-Now, the tool can create a bi-directional {\em query-to-range-set} map, called $M$, 
-between $Q$ and $R$, using the existing maps $M$1 and $M$2. 
-It leverages a transitive property from $M$1 to $M$2.
-Once $M$ is constructed, 
-the intermediate maps ($M$1 and $M$2) are not used in the rest of phases. 
-$M$ is shown as below:
+Using the existing maps $M$1 and $M$2, 
+the tool can create the final bi-directional {\em query-to-range-set} map, say $M$, 
+between $Q$ and $R$. 
+To construct $M$, we leverage a transitive property from $M$1 to $M$2.
+Once we compute $M$, there is no need to keep 
+the intermediate maps ($M$1 and $M$2) for the subsequent phases. 
+In the running example, we can build $M$ in the following.
+
+\pagebreak
 
@@
 
-In the subsequent phases $M$ gets used for computing costs and updated 
+\noindent In the subsequent phases $M$ gets used for computing costs and updated 
 along with a merge of ranges, and $R$ is refined for an MLPPI recommendation.
@@
 
-In this regard, the range set $R$ can be used to define an MLPPI using each field as one level. 
+The range set $R$ can be used to define an MLPPI using each field as one level. 
 However, if the current number of partitions by $R$ is greater than 
-the partition count limit, 
-then it is not possible to make a feasible MLPPI based on ranges in $R$.
+the aforementioned partition count limit, 
+then it is not possible to make a feasible MLPPI based on the ranges in $R$.
 
@@
 but to continue our discussion, assume that in our discussion the limit is {\it 15}.
-From $R$ in the running example, we can obtain a total of (number of ranges in {\tt LO\_DISCOUNT}) 
+From $R$ in the running example, we obtain a total of (number of ranges in {\tt LO\_DISCOUNT}) 
 $\cdot$(number of ranges in {\tt LO\_QUANTITY}) = (3+1) $\cdot$ (3+1) = {\it 16}
-partitions. 
-Because the total partitions surpass the limit, 
-the tool should pass through the initial phase in which 
-fewer partitions than the limit are made. 
-(Otherwise, the wizard will proceed to the optimized phase immediately.) 
-Note that one more range is counted per field in the calculation. 
-The added range is equivalent to the ``{\tt NO CASE OR UNKNOWN)}'' case in Figure~\ref{fig:exam}.
+partitions. Note here that one reserved range covering the rest but the identified 
+is added by default to each field in the calculation, 
+and the default range is equivalent to the ``{\tt NO CASE OR UNKNOWN)}'' case in Figure~\ref{fig:exam}. 
+Since the total partitions surpass the limit, 
+the tool should go through the initial phase in which 
+we make fewer partitions than the limit. 
+(Otherwise, the wizard can bypass the initial phase and then immediately proceed to the optimized phase.)
 
@@
 
-In this phase, we incrementally merge a range pair in $R$ to reduce partitions. 
+This phase incrementally merges a range pair in $R$ to reduce partitions. 
 The merge continues until the number of ongoing partitions 
-drop below the partition count limit. 
-Once reaching the limit, we can have a feasible MLPPI with 
-the remaining partitions.
-
-By and large, overall I/O cost may be increased by the merge. 
-Since every row is hash-sorted in each AMP of the Teradata DBMS server, 
-typically the full scan on a partition is done to retrieve 
-all the rows matching a given query. 
+drops below the partition count limit. 
+After finishing the merge, the tool can derive a feasible MLPPI with the remaining partitions.
+
+By and large, overall I/O cost may be increased by the merge, 
+in that the full scan on a partition is performed to retrieve all the rows potentially matching a given query. 
 In the case of a merged partition, we may read the non-qualifying rows 
 that would not be seen before the merge, thereby paying more I/O to answer 
-the query.  
-To minimize the merge overhead, we pick up the range pair incurring 
+the query. To minimize the merge overhead, we pick up the range pair incurring 
 the least I/O increase in a heuristic fashion. 
-
 To choose the desirable range pair for a merge, 
-the tool leverages the scan cost of a query on a range pair.
+the tool leverages the {\em scan cost} of a query on a range pair.
+
 Algorithm~\ref{algo:comp_scan_cost} represents how 
-the scan cost can be computed on the range pair that \hbox{influences} a query. 
-The scan cost represents the I/O cost to answer the query when the range pair is merged. It is 
-modeled as the {\em number of blocks} that are to be read 
-for the target query. To compute the scan cost of a query $q$, we use 
-a corresponding scan cost query ($s$), and 
-it can be constructed as follows: 
+the scan cost can be computed on the range pair that \hbox{influences} a given query. 
+The scan cost represents the I/O cost to answer the query when the range pair is merged. 
+It is defined as the {\em number of blocks} that are to be read 
+for the given query. To compute the scan cost of a query $q$, we write 
+a corresponding scan cost query ($s$), which can be constructed as follows: 
 \begin{center}
@@
 where $F$ is a fact table, and 
-{\it CP} indicates the predicates  
-restored from the ranges mapped to $q$ from $M$.
+{\it CP} indicates a predicate(s) reconstructed from 
+a range(s) mapped to $q$ in $M$.
 
@@
 then the range set associated with $q$ in $M$ is temporarily updated 
-by removing the parent ranges and adding the merged one. 
+by removing the parent ranges and adding the merged range. 
 The altered range set is remapped to $q$ in $M$ and then 
-translated to the equivalent predicates, 
-so that {\it CP} for $s$ based on the predicates can be built.
-Unless the merge range pair influences $q$, then 
-{\it CP} can be simply built based on the 
-predicates restored from the existing ranges mapped to $q$.
+translated to the equivalent predicates. Then {\it CP} for $s$ can be built based on the predicates.
+Unless the merge range pair influences $q$, 
+we can simply derive {\it CP} from the existing ranges mapped to $q$.
 
@@
 
-\begin{figure}[t]
+In the running example, suppose that in the first round, ranges $R$[1, 1] and $R$[1, 2] are chosen for a merge. 
+Then, the wizard produces a merged range \range{1, 5} \hbox{under} {\tt LO\_DISCOUNT} and proceeds to the next round. 
+If a range pair selected for the subsequent merge fails to improve $T_{p}$, 
+then, the tool exits the optimized phase and \hbox{produces} the final MLPPI recommendation for the given workload $Q$ 
+as shown in Figure~\ref{fig:mlppi}.
+
+\begin{figure}[h]
 \begin{center}
@@
 \end{figure}
-
-In the running example, suppose that in the first round, ranges $R$[1, 1] and $R$[1, 2] are chosen for a merge. 
-Then, the wizard produces a merged range \range{1, 5} \hbox{under} {\tt LO\_DISCOUNT} and proceeds to the next round. 
-If a range pair selected for the subsequent merge fails to improve $T_{p}$, 
-then, the tool exits the optimized phase and \hbox{produces} the final MLPPI recommendation for the given workload $Q$ 
-as shown in Figure~\ref{fig:mlppi}.
 
@@
 The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
-%\clearpage
-%\vspace\fill
-
-%1.
-Agrawal, R., Srikant, R.: Fast algorithms for mining association rules.
-In: Proceedings of the 20th International Conference on Very Large Databases.
-pp.
-487–499.
-Morgan Kaufmann, San- tiago, Chile (1994)
-%2.
-Bruce,K.B.,Cardelli,L.,Pierce,B.C.:Comparingobjectencodings.In:Abadi,M.,Ito,T.(eds.) Theoretical Aspects of Computer Software, Lecture Notes in Computer Science, vol.
-1281, pp.
-415–438.
-Springer-Verlag, Berlin Heidelberg New York (1997)
-%3. Garcia-Molina, H., Ullman, D.J., Widom, J.: Database Systems: The Complete Book. Prentice Hall, New Jersey, USA (2002)
-%4. van Leeuwen, J. (ed.): Computer Science Today. Recent Trends and Developments, Lecture Notes in Computer Science, vol. 1000. Springer-Verlag, Berlin Heidelberg New York (1995)
-%5. Ribicˇre,M.,Charlton,P.:Ontologyoverview.MotorolaLabs,Paris(2002),[Online].Available:
-%http://www.fipa.org/docs/input/f-in-00045/f-in-00045.pdf (current October 2003)
-%6. Wang,X.,Bettini,C.,Brodsky,A.,Jajoida,S.:Logicaldesignfortemporaldatabaseswithmul-
-%tiple granularities. ACM Transactions on Database Systems 22(2), 115–170 (1997)
-
-%\begin{thebibliography}{99}
-%\expandafter\ifx\csname url\endcsname\relax
-%  \def\url#1{\texttt{#1}}\fi
-%\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
-%\expandafter\ifx\csname href\endcsname\relax
-%  \def\href#1#2{#2} \def\path#1{#1}\fi
-%
-%{
-%\bibitem{Suh12}
-%{Y.-K.~Suh, A.~Ghazal, A.~Crolotte, and P.~Kostamaa}, {A New Tool for Multi-level Partitioning in Teradata}, 
-%in: CIKM'12, pp. 2214--2218.
-%  
-%\bibitem{sinclair:ppi}
-%P.~Sinclair, {Using PPIs to Improve Performance},
-%  \url{http://www.teradata.com/tdmo/v08n03/pdf/AR5731.pdf}
-%  (accessed August 27, 2015).
-%
-%\bibitem{klindt09mlppi}
-%J.~Klindt, {Single-level and Multilevel Partitioned Primary \linebreak Indexes},
-%  \url{http://www.teradata.com/white-papers/Single-level-and-}\linebreak\url{Multilevel-Partitioned-Primary-Indexes-eb1889/}
-%  (accessed July 27 2015).
-%
-%\bibitem{oneil:ssb}
-%{P.~O’Neil, B.~O'Neil, and X.~Chen}, {The Star Schema Benchmark (SSB)},
-%  \url{http://www.cs.umb.edu/~poneil/StarSchemaB.PDF} 
-%  (accessed February 28, 2017).
-%
-%\bibitem{agrawal04:integrating}
-%{S.~Agrawal, V.~Narasayya, and B.~Yang}, {Integrating Vertical and Horizontal Partitioning into Automated Physical Database Design}, 
-%in: \hbox{SIGMOD} '04, pp. 359--370.
-%
-%\bibitem{Lightstone04:db2auto}
-%{S.~Lightstone and B.~Bhattacharjee}, {Automated Design of
-%  Multi-dimensional Clustering Tables for Relational Databases}, in:
-%  VLDB '04, pp. 1170--1181.
-%
-%\bibitem{nehme2011automated}
-%{R.~Nehme and N.~Bruno}, {Automated Partitioning Design in Parallel
-%  Database Systems}, in: SIGMOD '11, pp. 1137--1148.
-%
-%\bibitem{chaudhuri1998autoadmin}
-%{S.~Chaudhuri and V.~Narasayya}, {AutoAdmin `What-If' Index Analysis Utility}, ACM SIGMOD Rec. 27~(2) (1998) 367--378.
-%
-%\bibitem{finkelstein1988physical}
-%{S.~Finkelstein, M. Schkolnick, and P. Tiberio}, {Physical Database Design for Relational Databases}, ACM Trans. on Datab. Syst. (TODS), 13~(1) (1988) 91--128.
-%
-%\bibitem{labio1997physical}
-%{W.~Labio, D.~Quass, and B.~Adelberg}, {Physical Database Design for Data Warehouses}, in: ICDE '97, pp. 277--288.
-%
-%\bibitem{rozen1991framework}
-%{S.~Rozen and D.~Shasha}, {A Framework for Automating Physical Database Design}, in: VLDB '91, pp. 401--411.
-%
-%\bibitem{Zilio1998}
-%{C.~Zilio}, {Physical Database Design Decision Algorithms and Concurrent
-%  \hbox{Reorganization} for Parallel Database Systems}, Ph.D. thesis,
-%  Dept. of CS, Univ. of Toronto (1998).
-%
-%\bibitem{sheet2009oracle}
-%{Oracle}, {Oracle White Paper---Partitioning with \hbox{Oracle} Database 11g
-%  \hbox{Release 2}},
-%  \url{http://www.oracle.com/technetwork/database/options/partitioning/twp-partitioning-11gr2-2009-09-130569.pdf}
-%  (accessed June 23, 2014).
-%
-%\bibitem{saa2007oracle}
-%{Oracle}, {SQL Access Advisor},
-%  \url{http://docs.oracle.com/cd/B19306_01/server.102/b14211/advisor.htm}
-%  (accessed March 30, 2012).
-%
-%\bibitem{rao2002automating}
-%{Jun Rao, Chun Zhang, Nimrod Megiddo, and Guy Lohman}, {Automating
-%  Physical Database Design in A Parallel Database}, in: SIGMOD '02, pp. 558--569.
-%
-%\bibitem{agrawal2000automated}
-%{S.~Agrawal et. al.}, {Automated Selection of Materialized Views and Indexes in SQL Databases}, in:
-%  {VLDB '00}, pp. 496--505.
-%
-%\bibitem{Agrawal04:sqlserver}
-%{S.~Agrawal et. al.}, {Database Tuning Advisor for Microsoft
-%  SQL Server 2005: Demo}, in: \hbox{SIGMOD} '05, pp. 930--932.
-%
-%\bibitem{dash2011cophy}
-%{D.~Dash}, {CoPhy: A
-%  Scalable, Portable, and Interactive Index Advisor for Large Workloads},
-%  in: PVLDB, 4~(6) (2011) 362--372.
-%
-%\bibitem{Kimura11}
-%{H.~Kimura et. al.}, {Compression Aware
-%  \hbox{Physical} Database Design}, in: PVLDB, 4~(10) (2011) 657--668.
-%
-%\bibitem{itwsqlserver}
-%{Microsoft SQL Server 2000}, {Index Tuning Wizard SQL Server 2000},
-%  \url{http://technet.microsoft.com/en-us/library/cc966541.aspx} 
-%  (accessed on March 21, 2014).
-%
-%\bibitem{valentin2000db2}
-%{G.~Valentin}, {DB2 Advisor: An Optimizer Smart Enough to Recommend Its Own
-%  Indexes}, in: {ICDE '00}, pp. 101--110.
-%
-%\bibitem{zilio2004recommending}
-%{D.~Zilio et. al.},
-%  {Recommending Materialized Views and Indexes with the IBM DB2 Design
-%  Advisor}, in: ICAC '04, pp. 180--188.
-%
-%\bibitem{Zilio04:db2design}
-%{D.~Zilio et. al.}, {DB2 Design Advisor:
-%  Integrated Automatic Physical Database}, in: VLDB '04, pp. 1087--1097.
-%
-%\bibitem{tatarowicz2012lookup}
-%{A.~Tatarowicz et. al.}, {Lookup Tables: Fine-grained Partitioning for Distributed Databases},
-%  in: ICDE'12, pp. 102--113.
-%
-%\bibitem{idreoskm07}
-%{S.~Idreos et. al.}, {Database Cracking}, in: CIDR'07, pp. 68--78.
-%}
-%
-%\end{thebibliography}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% For the final version of the paper: %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Author information
-%\vspace{4ex}\noindent
-%\textbf{Author One} is\dots
-%
-%\bigskip\noindent
-%\textbf{Author Two} is\dots
-%
-%\bigskip\noindent
-%\textbf{Author Three} is\dots
-
-%% Reception and acceptance information
-%\bigskip
-%\paragraph{Received: Month DD, 20YY; Accepted: Month DD, 20YY.}
-
 \end{document}

### 2017-03-14 subopt-tods16/paper.tex
SubOpt paper updates

@@
 \acmArticle{1}
-\acmYear{2016}
-\acmMonth{2}
+\acmYear{2017}
+\acmMonth{3}
 
@@
 \usepackage{subfigure}
+\usepackage{nicefrac}
 
@@
 hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We
-observe that \todo{Rick: check these}suboptimality prevalence
+observe that suboptimality prevalence
 correlates positively with the number of operators available in the \hbox{DBMS}
@@
 by the optimizer (another proxy for optimizer complexity) and with
-the number of correlation names (a measure of query complexity), providing
-empirical support for this model as well as implications for fundamental
-improvements of these optimizers. The results of this study also uncovered evidence of a potential fundamental
-limitation of cost-based optimization.\todo{Rick:be more specific! Couple
-  with contributions and conclusions}
+the number of correlation names, number of aggregates, and presence of a subquery (all measures of query complexity), providing
+empirical support for this model (in fact, these causal factors explain over
+half of the variance of suboptimality) as well as implications for fundamental
+improvements of these optimizers. We articulate for the first time a
+possible upper bound on the number of operators a DBMS may be able to support.
 This paper thus provides
@@
 generalization, proposes a novel causal model for query suboptimality, and
-tests several hypotheses deriving from this causal model.
+demonstrates that an upper bound on the number of operators is probable and
+may have already been reached by one or more extant DBMSes.
 \end{abstract}
@@
 \item We test the six quite specific hypotheses that arise from that
-  model across a wide range of DBMSes, queries, and data.
-
-\item We show through correlational and regression analysis that four
-  hypothesized interactions are strongly supported, one is partially supported, and
-  an interaction hypothesized to go one direction went quite weakly the
-  other direction, lending credibility to our particular model.
+  model across a wide range of DBMSes, queries, and data, showing
+  through correlational and regression analysis that four
+  hypothesized interactions are strongly supported, one is weakly in the
+  other direction, and
+  the sixth is partially supported, lending credibility to our particular model.
 
@@
 \begin{figure}[h!]\centering
-%\todo{Young: were is the figure!}
 \includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
@@
 
-This is why we needed a new methodology. We want to understand
-cost-based optimization deeply. This means that we need to go beyond the
-examination of a single query optimizer, as is done in almost every paper
-over the forty-year history of query optimization research, to study multiple instances of
-that optimization architecture, in an effort to achieve generalizable
-results.
+This is why we needed a new methodology. We want to understand cost-based
+optimization deeply. This means that we need to go beyond the examination of
+a single query optimizer, as is done in almost every paper (with a few
+important exceptions~\cite{harish07,Haritsa10,Leis15}) over the forty-year history of
+query optimization research, to study multiple instances of that
+optimization architecture, in an effort to achieve generalizable results.
 
@@
 Early investigation of plan suboptimality resulted in approaches such as
-dynamic query-reoptimization~\cite{Avnur,kabra98}, which exploit more
+dynamic query-reoptimization~\cite{Avnur,Bellamkonda13,kabra98,Li07}, which exploit more
 accurate runtime statistics that appear while a query is being executed, to
@@
 \vspace{0.6em}
-We don't feel that skewed data (in the ``data complexity''
-construct) will impact plan space
-complexity, but rather that it will negatively impact accuracy of plan cost
-estimation, and thus increase suboptimality.
+We feel that skewed data (in the ``data complexity''
+construct) will not impact plan space complexity. Rather, our expectation is
+that it will negatively impact accuracy of plan cost estimation, and thus
+increase suboptimality directly.
 
@@
 possibly do a better job, while also adding complexity to the optimization
-process. We hypothesize that this factor has a more complex role in the
+process.  The quality of query optimization ``very much depends on how much
+the query engine relies on [cardinality] estimates and on how complex the
+physical database design is, i.e., the number of indexes
+available.''~\cite{Leis15}.
+
+We hypothesize that this factor has a more complex role in the
 model, serving as a {\em moderator} of two interactions previously
@@
 
-We define the skew quite simply: ``the reciprocal of the number of distinct
+We define the skew as ``the reciprocal of the number of distinct
 values,'' so $0 < \hbox{\em skew} \leq 1$.
-For 2M distinct values, the skew would be $\frac{1}{2M}= 0.000005$, which is practically 0. 
+For 2M distinct values, the skew would be $\nicefrac{1}{2M}= 0.000005$, which is practically 0. 
 For exactly one distinct value, the skew would be 1.0. For two distinct values, 
@@
 
-Let's first illustrate with a skew factor of 1.0. 
+We use two values of skew: $\nicefrac{1}{2M}$ (termed {\em tiny}) and
+$\nicefrac{1}{10K}$ (termed {\em small}).
+\shorten{Let's first illustrate with a skew factor of 1.0. 
 This translates to exactly 1 value for the entire table, or 2M spans.
@@
 
-Let's see an example at the other end of the spectrum, of a skew factor of
-1/2M.  This skew factor, which is almost close to 0, gets translated to 2M
-distinct values.  So we generate a single span of 2M values.  To shorten the
+At the other end of the spectrum, consider a skew factor of
+1/2M.  This skew factor, which is almost 0, gets translated to 2M
+distinct values.  So}For the former, we generate a single span of 2M values.  To shorten the
 table, it makes no sense to remove this single span in its entirety. But we
@@
 (the skew has changed from .0000005 to .00001).
-
-We use five skewness values: 0.000005 (which we term only slightly
+For the latter, we generate 2,000 spans each of 10K tuples, and drop a span
+to reduce to 1.99M tuples, repeating.
+\shorten{We use five skewness values: 0.000005 (which we term only slightly
 misleadingly as ``no skew''), 0.001, 0.1, 0.5, and 1.0,
-for this independent variable.
+for this independent variable.}
 
@@
 To classify each operator as continuous or discontinuous, we use the data
-from the Exhaustive Experiment. We
+from the ``Exhaustive'' experiments to be described in detail in Section~\ref{sec:experiments}. We
 used a small subset of queries to be used later in testing the model, to
@@
 
-The Exhaustive experiment gathers the cost model for each operator in each
+The Exhaustive experiments gather the cost model for each operator in each
 plan for each cardinality (in our case, for cardinalities ranging from 10K
@@
 each plan at a particular cardinality. Each distinct point type depicts an
-individual hash-join operator. \todo{Young: I don't understand what this
-  sentence adds'' Rick: Yes, please delete the sentence.}Note that when two identical query plans
-appear at two different cardinalities, we consider all the plan operators
-found at the same position within these two plans to be identical. This
+individual hash-join operator.  This
 figure shows only a small portion of the 200 cardinalities, and is typical
@@
 \vspace{1em}
-As we will see in Section~\ref{sec:experiments}, we observed only 3,347
-violations (0.74\%) of non-strict monotonicity, for the largest experiment
-(Experiment 7: Confirmatory) across all
-the \hbox{DBMSes} we studied, justifying our conclusion that the \hbox{DBMSes} under
+That said, the suspicious monotonicity violations raised the fundamental question of whether
+we were using a sufficiently reliable timing method. The answer was {\em no}, as
+the experiment used a (popular but) naive measurement technique 
+based on end-to-end timing of which the results are 
+typically perturbed by system noise potentially incurred during the measurement. 
+We realized that the accuracy of this timing data was
+insufficient for both this monotonicity test but also for our suboptimality study. 
+We then switched to a new protocol, TTPv1~\cite{TTPv1}, and later TTPv2~\cite{TTPv2},
+which fortunately is able to obtain quite accurate and precise timing.
+
+Specifically, as we will see in Section~\ref{sec:experiments}, 
+using the protocol we observed only 3,347 violations (0.74\%) of non-strict monotonicity, 
+for the largest experiment (Experiment 7: Confirmatory) across all
+the \hbox{DBMSes} we studied. 
+Hence, this result justified our conclusion that the \hbox{DBMSes} under
 study are indeed monotonic.
@@
 No run violated the experiment-wide sanity checks. For the largest
-experiment, the Confirmatory experiment 7 discussed below, approximately
+experiment, the Confirmatory Experiment (number 7) discussed below, approximately
 10.5\% of the query executions and 5.1\%
@@
   this table}{, physically, of such tables} as the experiments are being performed. 
-\c2j{We used a}{We organized three configurations for the data sets. The first configuration
+\c2j{We used a}{We generated four configurations for the data sets. The first configuration
 is a small} data set with four tables, each with four integer-typed
 attributes. Each table is populated with one million rows, and thus the size
-of each table is roughly 16Mbyte. We also produced a version of the data set with primary
-keys (of the first attribute) for all tables.
+of each table is roughly 16Mbyte, with tiny skew (cf.~Section~\ref{sec:datacomplexity}). We also produced versions of the data set
+with (i)~small skew, (ii)~primary keys (of the first attribute), and (iii)~primary
+keys and secondary indexes, for all tables, as detailed in Appendix~\ref{sec:appdatasets}.
 
@@
 1 & Monotonicity 		& --- & 38 & 60 & 12,000 & 12,000 & 12,000\\
-2 & Refined Monotonicity 		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
-3 & Exhaustive 			& TTPv2 & 1,544 & 160 & 32,000 & 320,000 & 319,980\\
-4 & Exhaustive with Keys 	& --- & 28 & 200 & 40,000 & --- & ---\\
-5 & Initial Exploratory 	& TTPv1 & 560 & 780 & 8,842 & 88,420 & 68,891\\%40512+26367+2012
+2 & Exhaustive		& TTPv1 & 1,672 & 160 & 32,000 & 320,000 & 244,787\\
+3\shorten{4} & Exhaustive with Keys 	& --- & 28 & 200 & 40,000 & --- & ---\\
+4\shorten{5} & Initial Exploratory 	& TTPv1 & 560 & 780 & 8,842 & 88,420 & 68,891\\%40512+26367+2012
+5\shorten{3} & Refined Exhaustive 	& TTPv2 & 1,544 & 160 & 32,000 & 320,000 & 319,980\\
 6 & Exploratory 		& TTPv2 & 1,663 & 1,200 & 12,560 & 125,600 & 114,377\\
@@
 
-Table~\ref{tab:run_stat} exhibits the statistics of running a number of 
+We performed seven separate experiments, each looking at a different
+aspect. \hbox{Table~\ref{tab:run_stat}} exhibits the statistics of running a number of 
 queries in our experiments. All but the last column list the number of query
@@
 column gives the number of query executions retained by the protocol. One of
-the primary goals of TTPv2 was to reduce the number of QEs discarded due to
-phantom processes that were indirectly detected. Experiments 3, 6, and 7
-benefited from that protocol.
-
-We performed six separate experiments, each looking at a different
-aspect. It is important to emphasize that while the {\em queries} all came
+the primary goals of TTPv2 was (i)~to reduce the number of QEs discarded due to
+phantom processes that were indirectly detected and (ii)~to collect relevant measures 
+directly associated with I/O, such as BlockIO Delay time~\cite{TTPv2}. 
+Experiments 5, 6, and~7 benefited from that protocol. Other details about
+the experiments can be found in Appendix~\ref{sec:otherdetails}.
+
+The queries came from 19 query sets, summarized below with more details in
+\hbox{Appendix~\ref{sec:querysets}}. It is important to emphasize that while the {\em queries} all came
 from the same query pool and while the data sets were also shared by the
-experiments, the {\em query executions} for the five experiments are
-disjoint. As a side comment, we mention that for all four \hbox{DBMSes}, the query
+experiments (see the Appendix for details), the {\em query executions} for
+the six experiments, except Experiment~3 with no timing and hence no QEs, 
+are disjoint. As a side comment, we mention that for all four \hbox{DBMSes}, the query
 plans generated by a \hbox{DBMS} for a particular \QatC\ of a particular query
-varied between the experiments, but not between the query executions (QEs) of that query
-instance at that cardinality, by virtue of the way we designed the
-measurement protocol.
+varied between the experiments, but not between the query executions (QEs)
+of that Q@C, by virtue of the way \doubleblind{we designed the
+measurement protocol}{the measurement protocol was designed}.
 
@@
     became available,} we performed our
-second experiment, termed ``Refined Monotonicity,'' which more accurately tested the
-(strict) monotonicity assumption, as also described in
-Section~\ref{sec:suboptimality}.\todo{Young: I don't see any such discussion
-  in that section..., Rick: Please take a look at the definition of the strict monotonicity assumption. We discuss this there.}
-This experiment involved 160 query instances, 32,000 \QatC s, and thus
-320,000 QEs, requiring 1,672 cumulative hours.  The (very small\todo{Young:
-  how much?, Rick: 1.55\%}) percentage % (16312+8215)/(995321+587546) = 1.55%
+second experiment, termed ``Exhaustive,'' which more accurately tested the 
+monotonicity assumption. 
+This experiment involved 160 query instances, 32,000 \QatC s (thus the name:
+we timed each query at {\em all} cardinalities, generating 2,000 \QatC s for
+each query instance), and thus
+320,000 QEs (ten QEs for each \QatC ), requiring 1,672 cumulative hours.  The (very small: 1.6\%)
+percentage % (16312+8215)/(995321+587546) = 1.6%
 of strict monotonicity violations observed was consistent with the remaining
@@
 our definition of suboptimality, which requires monotonicity.
-
-We later reran this experiment using TTPv2 (we term this simply
-``Exhaustive''). Note that the number of retained QEs went up a lot (that is
-the main benefit of TTPv2), while the number of hours required actually
-went down (another benefit).
 
@@
 
-We also ran an experiment (termed ``Exhaustive with Keys'') on the Exhaustive query set but using the data
-set with primary keys,  However, in this
+We also ran an experiment (termed ``Exhaustive with Keys'') on the 
+Exhaustive query set but using the data set with primary keys,  However, in this
 case we did not actually execute
 the queries, but rather just examined the plans that were returned from the
-\hbox{DBMS} to identify additional discontinuous operators. \comment{Thus this third
-experiment was very fast.}\todo{Sabah:were any additional discontinous
-  operators found? Rick: I don't know about this. I redirect this question to Sabah.}
-
-\todo{Rick: Shouldn't this experiment  be before Experiment 3 (Exhaustive),
-  because it used TTPv1?}The fifth experiment was for initial exploratory analysis of a prior
-version of the causal model. (Note that this experiment used
-the first version of the Tucson Timing Protocol (TTPv1)~\cite{TTPv1}.)
-The earlier model had fewer independent
+\hbox{DBMS} to identify additional discontinuous operators. (Thus this third
+experiment was very fast.)\todo{Sabah:were any additional discontinuous
+  operators found? Rick: Yes, I think they were. like {\tt eq\_ref} on MySQL. But for more details 
+I redirect this question to Sabah.}
+
+The fourth experiment was an ``Initial Exploratory'' 
+analysis of a prior version of the causal model.
+That earlier model had fewer independent
 variables yet also had more complex interactions. Specifically, the model
@@
 %% The following shorten corresponds to TTPv1.
-The fifth experiment was an ``Initial Exploratory'' analysis of the causal model, run on
-a representative sample of queries and data sets: (a)~600 queries of one
+In the experiment we ran a representative sample of queries and data sets: (a)~600 queries of one
 \hbox{DBMS}, (b)~120 queries of another \hbox{DBMS}, (c)~10 queries
@@
 from each of the four \hbox{DBMSes} on the primary key data set, for exploration
-across all combinations, a total of 780 query instances. These query sets
-are enumerated in more detail in Appendix~\ref{sec:app}.
-We ran the \hbox{DBMSes} on the {\em
+across all combinations, a total of 780 query instances.  (Again, consult
+Appendix~\ref{sec:querysets} for details on these query sets.)
+We ran the \hbox{DBMSes} only on the {\em
   change points}: adjacent cardinalities (10K apart; 300 for one \hbox{DBMS}) with different
-plans. The protocol retained about 78\% of the QEs.
+plans (unlike the Exhaustive experiments). The protocol retained about 78\% of the QEs.
+
+As a large number of QEs (roughly a quarter) were discarded by TTPv1 in
+these experiments, and as these discarded QEs were correlated with query
+time (and thus perhaps with suboptimality),
+\doubleblind{we sought to develop a better measurement protocol, which was
+  the impetus for TTPv2~\cite{TTPv2}}{we switched to a better measurement
+  protocol that had recently become available: TTPv2~\cite{TTPv2}}.
+%Our subsequent experiments indeed benefited from the improved protocol, producing 
+%highly lossless ($\sim$5\% Q@C drop) but very clean ($\sim$1\% violation) data.
+The fifth experiment was to rerun the Exhaustive experiment  (we term this simply
+``Refined Exhaustive'') using that protocol. The main purpose of this experiment 
+was to reexamine monotonicity within in the enhanced
+protocol.\todo{Sabah:didn't we also check discontinuity once more? If so,
+  were any new discontinuous operators found in this third Exhaustive experiment?}
+
+We observed that the monotonicity violation rate decreased
+from 1.54\% to 1.02\% for strict monotonicity and from 0.61\% to 0.60\% for
+non-strict monotonicity, thereby reaffirming our monotonicity assumption
+(the first benefit). Moreover, the number of retained QEs went up
+significantly (compare244K to 320K, a second and the primary benefit of TTPv2), 
+while the number of hours required actually went down somewhat (a third benefit).
+These benefits reassured us of the quality of data 
+for subsequent experiments via the improved protocol.
 
@@
 (b)~100 queries from each of the four \hbox{DBMSes} on the primary key data set,
-for exploration across all combinations, a total of 1,200 query instances (= 300 $\times$ 4 (DBMSes)).
-We termed this sixth experiment ``Exploratory.''  We ran the DBMSes on the {\em
-  change points}: adjacent cardinalities (10K apart; 300 for one DBMS) with different
-plans. This experiment required 1,663 hours. The experimental
-methodology retaining 12,100 {\QatC}s (3.7\% were dropped), covering 1,123 query
+for exploration across all combinations, a total of 1,200 query instances. % (= 300 $\times$ 4 (DBMSes))
+We termed this sixth experiment ``Exploratory.'' This experiment 
+retained 12,100 {\QatC}s (3.7\% were dropped), concerning 1,123 query
 instances (6.4\% were dropped). Only 443 strict monotonicity violations (0.78\%) and 301 relaxed
-ones (0.54\%) were observed. This exploratory analysis allowed us to refine
+violation (0.54\%) were observed. This exploratory analysis allowed us to refine
 the operationalizations.
@@
 with primary keys, 
-(c)~100 queries on the data set without skew and primary keys,  
+(c)~100 queries on the data set with tiny skew and primary keys,  
 (d)~100 queries with a subquery on the data set without primary keys,  
@@
 for a total of 1,910 queries and a total of 7,640 query instances (over the
-four \hbox{DBMSes}).
-Again, we ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed,
+four \hbox{DBMSes}). (Again, see Appendix~\ref{sec:querysets} for details.)
+We ran the \hbox{DBMSes} at the 99,558 {\QatC}s that were observed,
 roughly 13 per query. The protocol accurately timed
@@
 
-\todo{Young: these percentages  are different from those in Section 6.1.
-Why?
-Rick: A good catch.
-Those numbers are out-of-date: ones from the TODS metrology paper.
-I replaced them with the current numbers mentioned in this para.
-Now consistent.}In this Confirmatory Experiment, we observed 3,347 strict monotonicity
+In this Confirmatory Experiment, we observed 3,347 strict monotonicity
 violations (0.74\%), and 1,966 relaxed monotonicity violations (0.43\%) (out
 of a total of 452,684 pairs
-of two Q@Cs having identical plans and query instance), \todo{Rick:merge
-  somehow with Section 6.1:}which provides
+of two Q@Cs having identical plans and query instance), which provides
 further confidence that monotonicity also applies to operators found in
@@
 QEs violated one or more of the (many) sanity checks specified within the
-TPPv1 and TTPv2 protocols, leaving a remaining 890,631
-retained QEs. Some of the query instances then voliated one or more of the
+TTPv2 protocol, leaving a remaining 890,631
+retained QEs. A few of the query instances then violated one or more of the
 Q@C sanity checks in the protocols (657 in all), leaving a remaining 6,983 retained query
@@
 
-Therefore, we were left with 6,967 queries, of which we make three general observations.
+We were then left with 6,967 queries, of which we make three general observations.
 
@@
 correlations of variables involved. These hypotheses predict 21 main effects
-and 12 moderating effects. Table~\ref{tab:hyptest} lists the hypotheses
+and 9 moderating effects. Table~\ref{tab:hyptest} lists the hypotheses
 followed by the correlation observed when testing each hypothesis.  (``NS''
@@
 to control the familywise error rate. ``---'' denotes no prediction arising
-from the model.)  As can be seen, most of the main effects (18) arising from
+from the model.)  As can be seen, most of the main effects (19) arising from
 the causal model are supported and significant.  The four exceptions are
-Hypotheses 1c, 2g, 2j, and 3. Hypothesis 1c was not supported because the
+Hypotheses 1c, 2g, 2j, and 3.
+
+Hypothesis 1c was not supported because the
 correlation between Number of operators available and Number of operators
 with discontinuity was negative, while we predicted positive
-correlation. Similarly Hypothesis 3 was not supported because the
-correlation between Percentage of skewed data and Suboptimality was
-negative, while we predicted positive correlation. (In both cases, the
-strength of the negative correlation was small: -0.09 and -0.05, respectively.) \hbox{Hypotheses}~2g and
-2j, both involving Number of repeats, were not significant.
-
-\begin{table}[t]
-\tbl{Testing Hypotheses 1--7: Correlations on the Confirmatory Study\label{tab:hyptest}}{%
-%\begin{center}
-\resizebox{85mm}{!}{
-\begin{tabular}{c|c|c|c|c}
-{\em Variable}& {\em Suboptimality}&{\em Repeats}&{\em CEPS} & {\em Discontinuity}\\
-\hline
-Operators in \hbox{DBMS} & ---&  {\bf H1a:} 0.35 & {\bf H1b:} 0.27& {\bf H1c:} -0.09\\
-Correlation names & --- & {\bf H2a:} 0.17 & {\bf H2b:} 0.46 &{\bf H2c:} 0.34\\
-Presence of aggregate & -- & {\bf H2d:} 0.04 & {\bf H2e:} 0.10 & {\bf H2f:} 0.29\\
-Presence of primary key & -- & {\bf H2g:} NS & {\bf H2h:} 0.15 & {\bf H2i:} 0.14\\
-Presence of subquery & -- & {\bf H2j:} NS & {\bf H2k:} 0.26 & {\bf H2l:} 0.20\\
-Skew & {\bf H3:} -0.05 & --- & ---  & ---\\
-Repeats& {\bf H6a:} 0.62 & --- & --  & ---\\
-CEPS &  {\bf H6b:} 0.61 & {\bf H4b:} 0.45 &  --- &{\bf H4a:} 0.72 \\
-Discontinuity & {\bf H6c:} 0.47 & --- & --- & ---\\
-\end{tabular}
-}
-%\end{center}
-}
-\end{table}
-
-\begin{table}[t]
-\tbl{Testing Hypothesis 5: Interaction Strength on the Confirmatory Study\label{tab:hyptest5}}{%
-%\begin{center}
-\resizebox{85mm}{!}{
-\begin{tabular}{c|c|c|c|c|c|c}
-{\em Variable}& \multicolumn{2}{|c|}{{\em Repeats}}&\multicolumn{2}{|c|}{{\em CEPS}} & \multicolumn{2}{|c}{{\em Discontinuity}}\\
-%\hline
-{\em }&{Not SI\em }&{SI\em }&{Not SI\em }&{SI\em }&{Not SI\em }&{SI\em }\\
-%&{\em }&{\em }&{\em }&{\em }\\
-\hline
-Correlation names & 0.17 & 0.15 &0.49 &0.39  & 0.36 & 0.27 \\
-Presence of aggregate  & 0.04 & NS & 0.12 & NS & 0.31 & 0.22 \\
-Presence of primary key  & -0.03 & NA & 0.06 & NA & 0.09 & NA \\
-Presence of subquery  & 0.03 & NS & 0.18 & 0.17  & 0.14 & 0.13 
-\end{tabular}
-}
-%\end{center}
-}
-\end{table}
-
-\subsection{Regression Analysis}\label{sec:regression}
-Regression analysis is required to test Hypothesis 5, which predicts a positive
-moderation in the presence of query complexity, specifically the Number of
-correlation names, Presence of aggregate, and Presence of subquery
-independent variables. Thus our causal model predicts that the interaction
-strength should increase when secondary indexes appear in the realtional schema.
-In Table~\ref{tab:hyptest5}, we examine the interaction strength both when
-the secondary indexes are absent (``Not SI'') and when they are present
-(``SI''). As before, ``NS'' denotes no significant at the $p < 0.05$ level (three
-of these interactions beome not significant when secondary indexes are
-added). ``NA'' denotes inapplicable: as the Presence of primary key is
-required for secondary indexes, it is not possible to compute the
-interaction strength in the presence of secondary indexes.
+correlation. The strength of the negative correlation was small: \hbox{-0.09}.
+
+Hypotheses ~2g and 2j, both involving Number of repeats, were not significant.
+
+Hypothesis 3 was
+not supported because the correlation between Presence of skew (recall from
+Section~\ref{sec:datacomplexity} that absence is skew is operationalized as
+{\em tiny} skew and presence of skew is operationalized as {\em small} skew)
+and Suboptimality was in the opposite direction. Specifically, we observed a negative correlation,
+indicating that as the skew increases from tiny to small, Suboptimality
+decreases, though only slightly: \hbox{-0.05}.
+
+In Table~\ref{tab:hyptest5}, we examine the interaction strength of
+Hypothesis 5, between the Query Complexity and Plan Complexity constructs, when
+the moderator (Presence of secondary indexes) has a value of absent (``Not
+SI'') and when it has a value of present
+(``SI''). As before, ``{\em NS}'' denotes no significant at the $p < 0.05$
+level; three of these interactions become not significant when secondary
+indexes are added. (Recall that it is not possible to compute the
+interaction strength for Presence of Primary Key Attribute in the presence
+of secondary indexes.)
 
@@
 
-A second use of regression is to compute the amount of variance explained,
-as another way of testing the moderation effect.
-Hypothesis 5 thus predicts that the strength of the interactions between
-query complexity with plan space complexity will increase in the
-presence of secondary keys. When
+\begin{table}[t]
+\tbl{Testing Hypotheses 1--7: Correlations on the Confirmatory Study\label{tab:hyptest}}{%
+%\begin{center}
+%\resizebox{150mm}{!}{
+\begin{tabular}{c|c|c|c|c}
+{\em Variable}& {\em Suboptimality}&{\em Repeats}&{\em CEPS} & {\em Discontinuity}\\
+\hline
+\hline
+Operators in \hbox{DBMS} & ---&  {\bf H1a:} 0.35 & {\bf H1b:} 0.27& {\bf H1c:} -0.09\\
+\hline
+Correlation names & --- & {\bf H2a:} 0.17 & {\bf H2b:} 0.46 &{\bf H2c:} 0.34\\
+\hline
+Presence of aggregate & -- & {\bf H2d:} 0.04 & {\bf H2e:} 0.10 & {\bf H2f:} 0.29\\
+\hline
+Presence of primary key & -- & {\bf H2g:} {\em NS} & {\bf H2h:} 0.15 & {\bf H2i:} 0.14\\
+\hline
+Presence of subquery & -- & {\bf H2j:} {\em NS} & {\bf H2k:} 0.26 & {\bf H2l:} 0.20\\
+\hline
+Skew & {\bf H3:} -0.05 & --- & ---  & ---\\
+\hline
+Repeats& {\bf H6a:} 0.62 & --- & --  & ---\\
+\hline
+CEPS &  {\bf H6b:} 0.61 & {\bf H4b:} 0.45 &  --- &{\bf H4a:} 0.72 \\
+\hline
+Discontinuity & {\bf H6c:} 0.47 & --- & --- & ---\\
+\end{tabular}
+}
+%\end{center}
+%}
+\end{table}
+
+\begin{table}[t]
+\tbl{Testing Hypothesis 5: Interaction Strength on the Confirmatory Study\label{tab:hyptest5}}{%
+%\begin{center}
+%\resizebox{85mm}{!}{
+\begin{tabular}{c|c|c|c|c|c|c}
+{\em Variable}& \multicolumn{2}{|c|}{{\em Repeats}}&\multicolumn{2}{|c|}{{\em CEPS}} & \multicolumn{2}{|c}{{\em Discontinuity}}\\
+%\hline
+{\em }&{Not SI\em }&{SI\em }&{Not SI\em }&{SI\em }&{Not SI\em }&{SI\em }\\
+%&{\em }&{\em }&{\em }&{\em }\\
+\hline
+\hline
+Correlation names & 0.17 & 0.15 &0.49 &0.39  & 0.36 & 0.27 \\
+\hline
+Presence of aggregate  & 0.04 & {\em NS} & 0.12 & {\em NS} & 0.31 & 0.22 \\
+\hline
+Presence of subquery  & 0.03 & {\em NS} & 0.18 & 0.17  & 0.14 & 0.13\\
+\end{tabular}
+}
+%\end{center}
+%}
+\end{table}
+
+\subsection{Regression Analysis}\label{sec:regression}
+One use of regression is as a further test of Hypothesis 5, which predicts a (positive)
+moderation in the presence of query complexity, specifically the Number of
+correlation names, Presence of aggregate, and Presence of subquery
+independent variables. Thus our causal model predicts that the interaction
+strength should increase when secondary indexes appear in the relational
+schema. (The {\em direction} of moderation, whether positive or negative, is not involved in this test; rather,
+an increase in interaction strength only supports that there is a moderating
+effect, in either direction.) When
 secondary indexes were not specified, our model explains 
 7.23\% of the variance of number of repeats, 40.9\% of the variance of CEPS, and 35.7\% of the variance
-of discontinuity\todo{Sabah: don't understand this ending phrase:}, with all of the independent variables. When
+of discontinuity. When
 secondary indexes were defined on the underlying tables, our model explains
@@
 38.9\% of the variance of discontinuity. These findings are all consistent with our
-hypothesis predicting a positive moderation effect of secondary
-indexes. Hence, the jury is still out on that hypothesis.
-
-A third use of regression is to compute the amount of variance explained
+hypothesis predicting a moderation effect of secondary indexes.
+
+A second use of regression is to compute the amount of variance explained
 for the intervening and dependent variables, to indicate how much of the
 underlying phenomena are explained holistically by the model.
-
-We did regressions on the causal variables for CEPS, number of repeats and for number of
+\shorten{We did regressions on the causal variables for CEPS, number of repeats and for number of
 operators with discontinuity. Our model explained 41.1\% of the
@@
 
-We also ran a regression over the independent variables of the model that
+} So we also ran a regression over the independent variables of the model that
 predict suboptimality over the data from the Confirmatory experiment. Our
@@
 variable. That means that over half of the variance of suboptimality is
-explained by the independent and intevening variables in our model, the ten
-variables within our five constructs of Optimizer Complexity, Schema
+explained by our operationalizations of Optimizer Complexity, Schema
 Complexity, Query Complexity, Data Complexity, and Plan Space Complexity.
@@
 specific hypotheses listed in Section~\ref{sec:model}. We then performed
-six experiments,
+a series of experiments,
 \begin{itemize}
 \item to refine our operationalizations: specifically discontinuity,
-via the Initial \hbox{Exhaustive} and Exhaustive \hbox{Experiments}, and number of operators available, via the
+via the Exhaustive \hbox{Experiment}, and number of operators available, via the
 Exhaustive with Keys \hbox{Experiment},
 \item to test fundamental assumptions, specifically
-monotonicity, via the Monotonicity \hbox{Experiment}, and
+monotonicity, via the Monotonicity and Refined Exhaustive \hbox{Experiments}, and
 \item to test and make minor
@@
 Our goal in this paper has been to understand cost-based query optimizers
-as a {\em general} class of computational artifacts and to articulate and test
+as a \hbox{\em general} class of computational artifacts and to articulate and test
 a predictive model \hbox{characterizing} how such optimizers,
@@
 
-\todo{Sabah: update with final numbers:}
 Our model includes two causal factors of suboptimality: \hbox{DBMS} data complexity, and plan space complexity. Of these factors,
@@
 suboptimality, indicating that choosing among many plans is hard, despite
-many decades of research and development. \todo{Rick: Please confirm this paragraph is still valid}
+many decades of research and development.
 
@@
 (0.176, normalized),
-\hbox{implicating} the cost model.
+\hbox{implicating} the cost model. The final factor is Presence skew (-0.057, normalized), a small effect that has the salutary effect of
+decreasing suboptimality.
 
@@
 
-Our predictive model in Figure~\ref{fig:model} suggests that suboptimality is causally impacted by the
-independent variable of CEPS (cardinality of effective plan space), and CEPS
-is naturally connected to the number of operators available\todo{Rick:
-  connect with Section~\ref{sec:root} after Sabah updates that section; also
-  discuss the exponential increase in CEPS possible}. A possible implication
+Our predictive model in Figure~\ref{fig:model} suggests that suboptimality
+is causally impacted by the Plan Space Complexity construct, specifically, the
+constituent variables of CEPS (cardinality of effective plan space), Number
+of repeats, and 
+Number of operators with discontinuity. This Plan Space Complexity construct
+is naturally connected to the Number of operators available. Specifically,
+our Confirmatory Experiment showed that the one variable of Number of
+operators available contributed 9.21\% of the variance of Suboptimality.
+
+A possible implication of our validated causal model
 is that as the evolution of generations of the \hbox{DBMS} add more and more
-operators, suboptimality will also increase.
+operators, suboptimality will increase.
 
@@
 The present discussion has a quite different focus. Here we're talking about
-the query optimizer {\em within} a single \hbox{DBMS}, across multiple
-generations, each generation adding one or more operators.
+the query optimizer {\em across multiple
+generations}, each generation adding one or more operators. (That said, as
+we'll show shortly, we can still study this phenomenon of increasing
+suboptimality across DBMSes; this remains a scientific question.)
 
@@
 extremely simple), and (iii)~range of table cardinalities (the smallest
-table is 1/200-th of the largest table). The simpliciy of queries, schemas,
+table is $1/{200}^{th}$ of the largest table). The simplicity of queries, schemas,
 and tables should independently, and certainly in concert, minimize the
@@
 \item We also show that it may well be useful to explicitly take {\em \hbox{cardinality} estimate
-  uncertainty} into account.
-
-\item This research indicates that aggregates are {\em not} a problem, so
-  that aspect of query optimization is in good shape.\todo{Rick: check with
-    Sabah's results and modify around here if needed.}
+  uncertainty} into account. (Leis et al.~\doubleblind{independently }{}came
+  to the same conclusion: ``The results ... demonstrate that the
+  state-of-the-art in cardinality estimation is far from
+  perfect.''~\cite{Leis15}).
+
+\item This research indicates that aggregates are less of a problem, so
+  that aspect of query optimization is in reasonable shape.
 
@@
 engineering approaches and refinements to improve the performance of query
-optimization, this is the first paper to the authors' knowledge that focuses
-on {\em suboptimality} and the first to take a scientific approach (in the
-sense of {\em empirical generalization} towards this important aspect.
+optimization, this is the first paper to the authors' knowledge t0 take a
+scientific approach (in the sense of {\em empirical generalization}),
+towards the important phenomenon of {\em suboptimality}.
 
@@
 an \hbox{attempt} to gain new understanding of this
-component.\todo{Rick:make consistent with list at the beginning of the paper}
+component.
 \begin{itemize}
@@
 provided {\em strong support} for our model, across \hbox{DBMSes}, thus
-qualitatively confirming what was informally known.
+qualitatively confirming what was informally articulated but never
+coherently tested.
 
 \item Uncovers compelling {\em evidence} (a)~that
-suboptimality correlates with two operationalizations of query complexity,
-(b)~that suboptimality correlates with two operationalizations of plan space
-complexity, (c)~that query complexity is a contributor to plan space
-\hbox{complexity,} and (d)~that schema complexity, as operationalized by the
-presence of primary key attributes, moderates these three
-interactions.\todo{Rick:Check again}
+suboptimality correlates with four operationalizations of query complexity,
+(b)~that suboptimality correlates with three operationalizations of plan space
+complexity, (c)~that the presence of secondary indexes is a helpful contributor to plan space
+\hbox{complexity,} (d)~that schema complexity, as operationalized by the
+presence of secondary indexes, helpfully though weakly moderates these three
+interactions, and (e)~that the presence of skewed data may diminish
+suboptimality slightly.
 
@@
 identified in our model: optimizer complexity, query complexity, and plan
-space complexity, in concert predicted a significant portion of the variance
-of suboptimality\todo{Rick: state. Do we want to keep the following?}.  It is doubtful that any other factor, as yet unknown,
-will itself predict as much variance as the factors we studied in this
+space complexity, in concert predicted a significant portion (53.6\%) of the variance
+of suboptimality.  No other factor, as yet unknown, or combination of factors,
+will themselves predict as much variance as the factors we studied in this
 paper. That said, it is certain that there remain several unknown causal factors;
@@
 \item Applies a novel experiment over pairs of adjacent Q@Cs to show that
-  such an upper bound is probably and may have already been reached with one
+  such an upper bound is probable and may have already been reached by one
   or more our our subject DBMSes.
@@
 \section{Future Work}\label{sec:future}
-There are several directions this work could be take. \todo{RICK IS HERE}With the model refinements
-we propose here, additional directed pointers to engineering efforts should
-emerge.\todo{Rick:Don't say ``we'', but just say the research needs to be done}
-
-\todo{Rick: keep following shorten?}\shorten{restate: Again, the purpose of our work is to present
-  a methodological perspective that can be further improved to study more
-  complex interactions present in the input to query optimizers.}
-
-\shorten{Operationally, doing experiments over four \hbox{DBMSes} was challenging. We
-had to install and configure each of these \hbox{DBMSes} and get our
-experimental apparatus to work uniformly across these systems. But in terms
-of testing the predictions, four is a small number, which meant that we had
-insufficient data to resolve Hypothesis 2 and had a low correlation for
-Hypothesis 1. We plan to augment \azdb\ to enable other \hbox{DBMSes} to be
-studied; our theory makes specific predictions on what we may observe with
-these other systems. Specifically, the interaction between number of \hbox{DBMS}
-operators and both plan complexity and suboptimality is difficult to tease
-apart, as that independent variable is entirely correlated to \hbox{DBMS}. Adding
-\hbox{DBMSes} to our empirical study should help understand this interaction
-better. In addition, we plan to apply path analysis, to separate direct and
-indirect effects.}
-
-We want to look into query flutter and thrashing in greater detail, as those phenomena
-provide concrete indicators of problematic optimizer behavior. One
-possible methodological approach is to utilize SQL optimizer hints, such as
-``{\tt \*+ SEMIJOIN}'' in MySQL and ``{\tt enable\_hashjoin(false)}'' in PostgreSQL,
-to encourage the optimizer to produce more plans at a given cardinality,
-that can then be timed to make more explicit the entire plan space (recall that
-CEPS, the cardinality of the effective plan space, is no greater and
-probably much smaller than the cardinality of the plan space).
-
-The investigations in this paper were based on very simple SPJ (select-project-join)
-queries.  We wish to also consider {\em schema complexity}: 
-foreign keys and indexes and {\em query complexity}: complex predicates, subqueries,
-and user-defined data types, methods, and operators. \c2j{}{We also want
-  to manipulate \hbox{DBMSes} internally (at least for those that are open-source),
-  turning on and off the rules and observing suboptimality.} Our causal
+There are at least three fundamental directions that could be taken in
+future work: (i)~model testing and refinement, (ii)~following up on engineering
+implications drawn from the predictive causal model, and (iii)~applying the empirical
+generalization perspective to other aspects of DBMSes.
+
+The model proposed and tested here, shown in Figure~\ref{fig:model}, is
+relatively simple, with six constructs and their interactions. Some of the
+predicted interactions were not borne out in the confirmatory testing. Might
+further testing support or reject the hypothesized correlation between Number
+of repeats (in the Plan complexity construct) and Presence of primary key
+attribute and Presence of subquery (in the Query complexity construct? Why did the hypothesized moderation
+  by the Presence of secondary indexes (in the Schema complexity construct)
+  {\em decrease} the influence of Query complexity on Plan space complexity?
+Finally, why did the Presence of skewed data (actually, the increase in skew
+from tiny to small, in the Data complexity construct)
+  {\em decrease} Suboptimality? (As others have noted, ``cardinality
+  estimates are usually computed based on simplifying assumptions like
+  uniformity and independence. In real-world data sets, these assumptions
+  are {\em frequently} wrong, which may lead to sub-optimal and sometimes
+  disastrous plans ... the cardinality estimators of the major relational
+  database systems produce bad estimates for many realistic queries, in
+  particular for multi-join queries.''~\cite{Leis15})
+
+Our causal
 model is extensible, in that we can add other factors, as long as their proper
-operationalization can be established, and additional causal links.
-So for
-sub-queries, we can add nesting level, type of sub-query (scalar,
-correlated, etc.), and number of ``effective joins''  (as some queries can
-be rewritten into joins with the outer level).
-It would also be interesting to study how a suboptimal subquery can affect
-the suboptimality of the containing query.  For instance, is it true that if
-many subqueries are themselves suboptimal, does that causally impact whether
-the overall query is suboptimal? It would also be useful to look at the
-moderation effect for Presence of secondary indexes: why precisely do these seem to
-actually decrease plan space complexity, and does this moderation continue
-as queries get more complex? Finally, we may be able to determine
-whether certain query rewrite techniques are employed within each \hbox{DBMS},
-introducing another ``\hbox{DBMS} complexity'' factor into our model.
+operationalization can be established, and additional causal links. It would be
+useful to consider
+\begin{itemize}
+\item {\em schema complexity:} foreign keys and
+uniqueness and other kinds of constraints, 
+\item {\em query complexity:} complex predicates, multiple nesting of
+  subqueries and different types of subqueries, query operators such as {\tt
+    EXCEPT} and {\tt UNION}, other
+SQL clauses such as {\tt ORDER BY} and {\tt GROUP BY}, and user-defined data
+types, methods, and operators,
+\item {\em data complexity:} data that doesn't exhibit uniformity or independence
+  data, and
+\item {\em plan space complexity:} the actual space of plans considered by
+  the optimizer (discussed below).
+\end{itemize}
+
+Our Confirmatory experiment used only two values of skew, tiny
+($\nicefrac{1}{2M}$) and small ($\nicefrac{1}{10K}$), for the data complexity
+construct. It would be useful to extend the study to (much) larger
+values. (This could increase the number of duplicates in joins, which may
+dramatically increase the query time.) But more to the point, our definition
+of skew utilizes a uniform distribution of values, albeit across a narrowing
+range as skew increases. It would be useful to look at other distributions,
+such as an increasing distribution and a normal distribution, as well as a
+very spiky distribution. It would also be useful to run experiments over
+entirely different benchmarks, including those based on real-world data and
+queries such as the Join Order Benchmark~\cite{Leis15}, appropriately
+modified to provide a range of values for the various constructs such as
+data skew.
 
@@
 
-\shorten{During our long-running experiments, the operating systems on which
-the experimental subject \hbox{DBMSes} are installed were upgraded. This event caused
-a significant increase in the number of phantom processes observed
-during query executions, forcing us to eliminate many query executions for
-statistical analysis, in that our model does not yet factor in a proper
-measurement for the effects of phantom processes. But our proposed model
-can be enhanced so that various additional parameters, such as operating system
-configurations, can be examined under regression analyses, once these
-parameters are properly operationalized.}
-
-By employing an extensible causal model, many complex factors can be studied
+It may be useful to look into query flutter and thrashing in greater detail, as those phenomena
+provide concrete indicators of problematic optimizer behavior. One
+possible methodological approach is to utilize SQL optimizer hints, such as
+``{\tt \*+ SEMIJOIN}'' in MySQL and ``{\tt enable\_hashjoin(false)}'' in PostgreSQL,
+to encourage the optimizer to produce more plans at a given cardinality,
+that can then be timed to make more explicit the entire plan space (recall that
+CEPS, the cardinality of the effective plan space, is no greater and
+probably much smaller than the cardinality of the plan space).
+
+Along with adding such factors to the model, one can watch whether these
+other factors impact (at all, and if so, how much) the original constructs
+and interactions. So for example it would be interesting to study how a suboptimal subquery can affect
+the suboptimality of the containing query.  For instance, is it true that if
+many subqueries are themselves suboptimal, does that causally impact whether
+the overall query is suboptimal?
+
+It would also be useful to study other relational DBMSes, as the model
+should apply to any such DBMS using a cost-based optimizer.
+
+All of the above activities are in the tradition of science, specifically
+empirical generalization (cf.~Figure~\ref{fig:empirical}).
+
+The second fundamental direction takes the validated model and draws
+engineering implications. Our model has provided specific directions for implementation interventions:
+refining the cost model of discontinuous operators, improving buffer
+allocation algorithms, and accommodating cardinality estimate uncertainty.
+One can further ask, for each causal interaction in the model,
+interactions that have been validated in empirical studies, what is it about
+the cost-based query optimizer that that interaction is evidenced? And then
+one can ask, what changes to that optimizer might ameliorate that
+interaction?
+
+It may be that that interaction is {\em baked into the optimizer}. For
+example, the study in Section~\ref{sec:diminishing} was predicated on some
+basic aspects of adding operators: each successive operator {\em benefits} a smaller region
+of query/cardinality (specifically, Q@Cs) and is susceptible to
+suboptimality on a larger region of Q@Cs
+(cf.~Section~\ref{sec:gedanken}). That seems like a fundamental limitation.
+
+But other interactions may be avoidable, through selection of alternative
+algorithms. After all, this study only generalizes over four DBMSes.
+
+Engineering also provides alternative methodologies to studying the model
+variables, which we have examined only from the ``outside'' of the DBMS. It
+would be useful to manipulate \hbox{DBMSes} internally (at least for those
+that are open-source or are available within the DBMS vendor), turning on
+and off the rules and observing suboptimality. (An example of this is Leis
+et al.'s study of cardinality estimation, cost model, and plan enumeration~\cite{Leis15}.) And it may be possible to
+determine whether certain query rewrite techniques are employed within each
+\hbox{DBMS}, introducing another ``\hbox{DBMS} complexity'' factor into our
+model.
+
+If extant cost-based query optimizers are hitting a wall,
+it might be necessary to fundamentally revisit query optimization,
+to come up with an entirely new approach that is less impacted by number
+of operators and by CEPS and thus avoids the flutter phenomenon.
+
+A third fundamental direction is to return to science, but examine dependent
+measures other than suboptimality. This paper demonstrates that by employing an extensible causal model, many complex factors can be studied
 via a systematic, statistically sound, scientific manner to better
 understand the causal factors and their interactions.
-
-In addition to the refinements of the causal model just discussed, our
-research has provided specific directions for implementation interventions:
-refining the cost model of discontinuous operators, improving buffer
-allocation algorithms, and accommodating cardinality estimate uncertainty.
-
-\todo{Rick: revise: talk about generational study, how we're hitting a wall,
-  and how it might be necessary to fundamentally revisit query optimization,
-  to come up with an entirely new approach that is less impacted by number
-  of operators and by CEPS and thus avoids the flutter phenomenon.}
-
-The methodology introduced in this paper and the causal model that results
-has suggested both the scope of the problem of query suboptimality and a
-number of specific engineering efforts that can now be carried out. Our
-ultimate goal is a refined causal model that can fully explain how query
-suboptimality arises in cost-based optimizers, thereby enabling engineering
-solutions that reduce and eventually
-effectively eliminate query suboptimality.
-
-\doubleblind{\todo{Rick:Consider also mentioning somewhere Young's ergalics study
-  on another kind of thrashing.}}{}
+\doubleblind{Another recent example is Suh's validated causal
+predictive model for DBMS thrashing~\cite{Suh17}. }{}What other areas of the
+rich field of databases might be amenable to this approach?
+
+The methodology utilized in this paper to introduce a predictive causal model has
+suggested the scope of the problem of query suboptimality, a number of
+contributing factors, and a collection of specific engineering efforts that
+can now be considered. An ultimate goal is a refined causal model that
+fully explains how query suboptimality arises in cost-based optimizers,
+thereby enabling engineering solutions that address this important issue.
 
@@
 
+%\todo{Young: get citations to show up as Name, year}
 %\vfill\pagebreak
@@
 
-\bibliographystyle{abbrv}
-\newcommand{\etalchar}[1]{$^{#1}$}
+\bibliographystyle{ACM-Reference-Format-Journals}
+\bibliography{paper}
+\shorten{\newcommand{\etalchar}[1]{$^{#1}$}
 \begin{thebibliography}{99}
@@
   Conference}, pp.~119--130, Baltimore, Maryland, 2005.
+
+\bibitem{Bellamkonda13} S.~Bellamkonda, H.-G.~Li, U.~Jagtop, Y.~Zhu,
+V.~Liang, and T.~Cruanes, ``Adaptive and big data scale parallel execution
+in Oracle,'' {\em PVLDB} 6(11):1102--1113, 2013.
 
@@
 
+\bibitem{Leis15}
+V.~Leis, A.~Gubichev, A.~Mirchev, P.~Boncz, A.~Kemper, and T.~Neumann, ``How
+Good Are Query Optimizers, Really?''' {\em Proceedings of the VLDB
+  Endowment} 9(3): 204--215, 2015.
+
+\bibitem{Li07}
+Q.~Li, M.~Shao, V.~Markl, K.~S.~Beyer, L.~S.~Colby, and G.~M.~Lohman,
+``Adaptively reordering joins during query execution,'' in {\em Proceedings
+  of the IEEE International Conference on Data Engineering}, pp.~26--35, 2007.\
 \bibitem
@@
 
+\bibitem{Suh17}
+Y.-K.~Suh, R.~T.~Snodgrass, and S.~Currim, ``An Empirical Study
+of Transaction Throughput Thrashing Across Relational DBMSes,''
+{\em Information Systems}, 54 pages, 2017.
 \bibitem
@@
 
-%\bibitem
-%%{Kabra:1998:EMR:276304.276315}
-%{Kabra}
-%N.~Kabra, D.~J.~DeWitt, ``Efficient Mid-Query Re-Optimization of Sub-Optimal Query Execution Plans'', in {\em Proceedings of the ACM SIGMOD Conference},
-%pp.~106--117, 1998.
-
-%\bibitem
-%%{Avnur:2000:ECA:342009.335420}
-%{Avnur}
-%R.~Avnur, J.~M.~Hellerstein, ``Eddies: Continuously Adaptive Query Processing'',
-%in {\em Proceedings of the ACM SIGMOD Conference}, pp.~261--272, 2000.
-
 \end{thebibliography}
+}
 
@@
 
+\subsection{Data Sets}\label{sec:appdatasets}
 The third column identifies the data set(s) used in each experiment, that is, the
@@
 the algorithm for generating the values for different values of skew; this
-algorithm is used in columns two--four, which for any row will have
+algorithm is used in the second to fourth columns, which for any row will have
 identical values. The first column holds a unique integer starting from 1
@@
 
+\subsection{Other Details}\label{sec:otherdetails}
 The next column of Table~\ref{tab:run_stat2} concerns the {\em Lab
@@
 in the scenario code across these experiments was in accommodating the
-details of the data set (that is, creating secondary indexes and data skew)
-\todo{Young:was this in the scenario code? It seems that this should be
+details of the data set (that is, creating secondary indexes and data skew).
+\shorten{\todo{Young:was this in the scenario code? It seems that this should be
   outside of that: Rick: Yes, this is part of the scenario code.
 Specifically, if the presence of secondary indexes is true, or the presence of data skew is false, then in the scenario code an experiment subject creates the indexes or makes no skew when populating tables.
-After table population is done, the code studies each query, or detects a change point and makes query executions at that change point.}and adding robustness features, such as using exponential backoff when the
-network connection to the DBMS was temporarily lost.
+After table population is done, the code studies each query, or detects a change point and makes query executions at that change point.}}
 
@@
 possibilities, all 200 cardinalities or just the cardinalities at which the
-query plan changed. The sixth column, ``what was timed?'', indicates that experiments three and four, Exhaustive and
-Exhaustive with Keys, described in Sections~\ref{sec:discontinuity} and \ref{sec:experiments}, just
-collected query plans, and thus took much less time to run.
+query plan changed. The sixth column, ``what was timed?'', indicates that Exhaustive with Keys,
+described in Section~\ref{sec:experiments}, just
+collected query plans, not timing any of them.
 
@@
 QEs), termed {\em raw}, and how many Q@Cs were retained after the protocol
-implied by the fourth column and listed explicitly in the third column of
-Table~\ref{tab:run_stat} dropped query executions and Q@Cs via its many
-  sanity checks.
+(listed in the third column of Table~\ref{tab:run_stat}) dropped query
+executions and Q@Cs via its many sanity checks. (Note that we don't list QEs
+and retained QEs in Table~\ref{tab:run_stat} for Exhaustive with Keys simply
+because we did not use timing data in that experiment.)
 
@@
 1 & Monotonicity 	& A & 6.0     &all cardinalities&yes& 12,000 (12,000)\\
-2 & Refined Monotonicity& A & 5.19 + 6.0&all cardinalities&yes& 27,948 (32,000)\\
-3 & Exhaustive 		& A & 7.1     &all cardinalities&no & 29,515 (32,000)\\
-4 & Exhaustive with Keys & B& 6.0     &change points   &no& --- \\
-5 & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8,171 (8,842)\\
+2 & Exhaustive & A & 5.19 + 6.0&all cardinalities&yes& 27,948 (32,000)\\
+3\shorten{4} & Exhaustive with Keys & B& 6.0     &change points   &no& 40,000 (40,000) \\
+4\shorten{5} & Initial Exploratory & A + B&5.19 + 5.2 + 6.0&change points&yes& 8,171 (8,842)\\
+5\shorten{3} & Refined Exhaustive & A & 7.1     &all cardinalities&yes & 29,515 (32,000)\\
 6 & Exploratory 	& A + B&7.1   &change points   &yes& 12,100 (12,560)\\
 7 & Confirmatory 	&A + B + C +D&7.1&change points&yes& 94,502 (99,558)\\
-\multicolumn{2}{c|}{\em Total}         &            & &               &   & 184,236 (196,960)\\
+\multicolumn{2}{c|}{\em Total}&      &   &             &   & 184,236 (196,960)\\
 \end{tabular}
@@
 
+\subsection{Query Sets}\label{sec:querysets}
 The following sets of queries were used in the seven experiments.
@@
 \item[QSj] 100 queries (on Data Set A)
-\item[QSl] A query set consisting of 230 queries drawn from \hbox{\em
-  QSa}--{\em QSf} (on Data Set B)
-\item[QSm] A query set consisting of 160 queries: 40 queries
-  drawn from each of {\em QSg}--{\em QSj} (on Data Set B)
-\item[QSn] A query set consisting of 110 new queries (on Data Set B)
-\item[QSo] A query set consisting of 100 queries without aggregates (on Data
-  Set C)\todo{Young: are these new queries, or drawn from prior query sets? Rick: New queries.}
-\item[QSp] A subquery query set consisting of 100 queries, each with a subquery (on Data Sets A, B, and D)
-\item[QSq] A subquery query set consisting of 100 queries, each with a subquery (on Data Set D)
-\item[QSr] A query set consisting of the 390 queries in {\em QSl} + {\em QSm} (on Data Set B)
-%\item[QSs] A primary key query set consisting of 100 queries drawn from {\em QSr}
-\item[QSs] A query set consisting of 100 queries drawn from {\em QSr} (on Data Set D)
+\item[QSk] A query set consisting of the 390 queries drawn from {\em QSa}--{\em QSj} (on Data Set B)
+\item[QSl] A query set consisting of 110 new queries (on Data Set B)
+\item[QSm] A query set consisting of 100 queries without aggregates (on Data Set C)
+\item[QSn] A subquery query set consisting of 100 queries, each with a subquery (on Data Sets A, B, and D)
+\item[QSo] A subquery query set consisting of 100 queries, each with a subquery (on Data Set D)
+\item[QSp] A query set consisting of 100 queries drawn from {\em QSk} (on Data Set D)
 \end{description}
@@
 
-Experiment 2 (Refinded Monotonicity) used the first 50 queries from {\em QSa}
+Experiment 2 (Exhaustive) used the first 50 queries from {\em QSa}
 for the three other DBMSes plus the ten queries for MySQL from Experiment 1, for
-a total of 160 query instances. Experiment 3 (Exhaustive) used the same 160 queries.
-
-Experiment 4 (Exhaustive with Keys) used the first 50 queries from {\em QSa}
+a total of 160 query instances. Experiment 5 (Refined Exhaustive) used the same 160 queries.
+
+Experiment 3 (Exhaustive with Keys) used the first 50 queries from {\em QSa}
 for the four DBMSes, for a total of 200 query instances.
@@
 Experiment 6 (Exploratory) used {\em QSa} and {\em QSb}, plus the first 100 (primary
-key) queries from {\em QSr}, for the four DBMSes, for a total of 1200 query
+key) queries from {\em QSk}, for the four DBMSes, for a total of 1200 query
 instances.
@@
 Experiment 7 (Confirmatory) used {\em QSc}--{\em QSj},
-along with {\em QSr} except the (first 100) queries included in Experiment 6,
-{\em QSn} for primary key (for two runs, or 220 queries),
-{\em QSo} for no data skew,
-{\em QSp} for primary key and subquery,
-{\em QSp} for subquery,
-{\em QSp} and {\em QSq} for primary key and secondary index and subquery,
-{\em QSs} for primary key and secondary index,
+along with {\em QSk} except the (first 100) queries included in Experiment 6,
+{\em QSl} for primary key (for two runs, or 220 queries),
+{\em QSm} for no data skew,
+{\em QSn} for primary key and subquery,
+{\em QSn} for subquery,
+{\em QSn} and {\em QSo} for primary key and secondary index and subquery,
+{\em QSp} for primary key and secondary index,
 all across the four DBMSes, for a total of 7,640 query instances.

### 2017-03-14 mlppi-comsis17/paper.tex
Done before the Optimized Phase

@@
 
-\institute{Department of Computer Science, The University of Arizona, Tucson, AZ 85721, USA\\ 
+\institute{Department of Computer Science, The University of Arizona, AZ 85721, USA\footnote{Currently, he is at Korea Institute of Science and Technology Information (KISTI).
+This work was conducted while he interned at Teradata during his PhD in the University of Arizona.}\\
+  %\email{yksuh@cs.arizona.edu}
   \email{yksuh@cs.arizona.edu}
@@
   \email{\{alain.crolotte,pekka.kostamaa\}@teradata.com}
-%  \and
-%  Korea Institute of Science and Technology Information, Daejeon 34141 Republic of Korea\\ 
-%  \email{yksuh@kisti.re.kr}
 }
@@
 
-By and large, overall I/O cost may be increased by the merge, 
+Overall, I/O cost may be increased by the merge, 
 in that the full scan on a partition is performed to retrieve all the rows potentially matching a given query. 
@@
 the least I/O increase in a heuristic fashion. 
-To choose the desirable range pair for a merge, 
-the tool leverages the {\em scan cost} of a query on a range pair.
+To choose a range pair for a merge, 
+we calculate the {\em scan cost} of a query on the merged range pair.
 
@@
 	\ForEach{range $r$ $\in$ $L$}{	
-		Restore a predicate $p$ from $r$. \\
+		Reconstruct a predicate $p$ from $r$. \\
 		Add $p$ to {\tt WHERE} clause of $s$.
@@
 
-If $q$ turns out to be affected by the merge of a range pair, 
+If $q$ turns out being affected by the merge of a range pair, 
 then the range set associated with $q$ in $M$ is temporarily updated 
@@
 
-To help understand the scan cost construction, let us 
+To see in the Algorithm~\ref{algo:comp_scan_cost} how a scan cost query is built, 
 consider a range pair ($rp$) of \hbox{$R$[2, 2]} and $R$[2, 3] in the running example. 
@@
 The merge by $rp$ affects both $q$1 and $q$2 in the workload. 
-
 Thus, the range sets of $q$1 and $q$2 on {\tt LO\_QUANTITY} are 
-altered to \{\openrange{\infty, 25}, \range{25, 35}\} and \{\range{25, 35}\}, 
+correspondingly altered to \{\openrange{\infty, 25}, \range{25, 35}\} and \{\range{25, 35}\}, 
 respectively. Of course, no change is made to the existing range sets 
@@
 Therefore, the corresponding scan cost queries for $q$1 and $q$2 
-can be built as below:
+can be built as shown below:
+\vspace{-.08in}
 \begin{center}
@@
 \end{center} 
-
-It is possible to combine all the (consecutive) ranges in the (temporarily) altered set of $q$1, 
-and hence, a single predicate ({\tt l.LO\_QUANTITY <= 35}) can be built.
-
-The wizard sends such a scan cost query to the server via an API call, 
-and it extracts from the server's response the {\em spool size} (in bytes), or the size of scan query result. 
-Then, the tool computes as scan cost the {\em block counts} using the spool size. 
+\vspace{-.08in}
+\noindent Since the altered ranges (\openrange{\infty, 25} and \range{25, 35}) of $q$1 
+are consecutive, we can simply build the merged, single predicate ({\tt l.LO\_QUANTITY <= 35}) for $\Sone$.
+
+In turn, the wizard sends each scan cost query to the server through an API call. 
+The tool extracts from the server's response the {\em spool size} (in bytes), which is 
+the result size of the sent scan cost query, and then 
+calculates as scan cost the {\em block counts} ($blc$) using the spool size. 
 If the merge of a range pair does not impact on any query in the workload, 
-the existing (previously computed) scan cost of queries 
-will be reused for the range pair. 
-That is, the wizard will only recompute 
-the scan cost of a query {\em iff} the query is affected by 
-the merge of two consecutive ranges.
-
-In this way, the weighted sum of scan cost of queries, $T_{s}$ can be computed 
+the existing (previously computed) scan cost of queries is reused for the range pair. 
+In other words, the wizard only recomputes the scan cost of a query {\em only if} 
+the query is affected by the merge of two consecutive ranges. 
+The weighted sum of scan cost of queries, $T_{s}$ can be computed 
 for each range pair. 
@@
 the top two ranges of each field. 
-Unfortunately, because the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
-contain the non-qualifying rows for $q$1, $R$ cannot provide $q$1 with as much benefit as $q$2. 
-However, $R$ can be a good compromise to satisfy both queries, in that 
-the MLPPI derived by $R$ can potentially minimize the total execution cost 
-of $Q$.
+As the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
+contain the non-qualifying rows for $q$1, unfortunately, 
+$R$ cannot provide $q$1 with as much benefit as $q$2. 
+$R$, however, can be a good compromise to satisfy both queries, in that 
+the MLPPI derived by $R$ can potentially minimize the total execution cost of $Q$.
 
@@
 Algorithm~\ref{algo:init_phase} represents the initial phase 
-that has been proposed so far. 
-The algorithm starts with a query-to-range set map ($M$) and an input range set ($R$). 
-Until the number of the partitions in $R$ falls below the partition limit threshold, 
-the algorithm determines the range pair yielding the least total execution cost ($T_{s}$) 
-and updates $M$ and $R$. At the end, the algorithm produces a final recommendation based on $R$.
-%\vspace{-.1in}
-
-Now that the total partition counts ($4${$\times$}$3$=$12$) by $R$ falls  
-below the assumed limit ($15$), the wizard 
-proceeds to the optimized phase with $R$ without repeating the initial phase.
+that has been described so far. 
+Given a \linebreak \hbox{query-to-range} set map ($M$) and an input range set ($R$), 
+the algorithm determines which range pair to yield the least total execution cost ($T_{s}$) 
+and updates $M$ and $R$ using the chosen range pair, 
+until the number of the partitions by $R$ falls below a pre-defined partition limit count.
+Finally, the algorithm produces the final recommendation based on $R$.
+
+In the running example the total partition counts ($4${$\times$}$3$=$12$) by $R$ is under 
+the assumed limit ($15$), which meets the partitioning limit. 
+Hence, the wizard can enter the optimized phase with $R$, 
+without repeating the initial phase.
 
@@
 %We verified the effectiveness of the tool by running it on a workload based
-In this regard, using $R$ we have an initial MLPPI recommendation for the {\tt LINEORDER} fact table.
-The recommendation can be used sufficiently, but 
-having fewer partitions, induced by further merges, 
-can enhance the overall workload performance for the following reasons. 
+After passing the initial phase, 
+we now have an initial MLPPI recommendation based on $R$, for the {\tt LINEORDER} fact table.
+The initial recommendation can be used sufficiently.
+
+But via additional merges we may further make fewer partitions, 
+which actually can enhance the overall workload performance for the following reasons. 
 First, multiple file contexts by many partitions can incur 
@@
 as shown in Figure~\ref{fig:mlppi}. 
-
+\vspace{-.25in}
 \begin{figure}[h]
@@
 \end{center}
-\caption{An MLPPI Recommendation for Workload $Q$\label{fig:mlppi}}
+\vspace{-.2in}
+\caption{The Final MLPPI Recommendation for the Given Workload $Q$\label{fig:mlppi}}
+\vspace{-.2in}
 \end{figure}
@@
 
-{\bf Comment:} In practice, our experiments showed that the bound 
-$\BigO$($N${$\cdot$}$C${$\cdot$}$V$) is overly pessimistic.
+But we empirically observed that the bound of $\BigO$($N${$\cdot$}$C${$\cdot$}$V$) 
+was overly pessimistic.
 
@@
 $\BigO$($M^{2}${$\cdot$}$N$), 
-where $M$ is the number of range pairs, and $N$ is the number of queries in 
-a given workload. 
+where $M$ is the number of range pairs, and $N$ is the number of queries in a workload. 
 \end{lemma}
@@
 
-Our experiments showed that $\BigO$($M^{2}${$\cdot$}$N$), the running time complexity, 
-is overly pessimistic. 
-In \hbox{practice}, the number of calls made was much fewer than the \hbox{theoretic} bound, 
-since it was very rare for a range pair to be associated with all queries in a workload. 
-In \hbox{Section}~\ref{sec:stat} we will show how many API calls take place on the workloads 
-used for evaluation.
+Again, our experiments showed that $\BigO$($M^{2}${$\cdot$}$N$), the running time complexity, 
+was overly pessimistic. Specifically, 
+the number of calls made in our experiments was much fewer than that bound, as 
+it was almost rare for a range pair to involve all queries in a workload 
+(as seen in Table~\ref{tab:workload_stat}).
+
+In the following, we evaluate the proposed algorithm.
 
@@
 Note that there \hbox{happened} only one iteration in the optimized phase. 
-The reason was that since the partitions produced by the \hbox{initial} phase were customized enough, 
+The reason was that as the partitions produced by the \hbox{initial} phase were customized enough, 
 no more \hbox{iterations} were \hbox{necessary} for further optimization.
@@
 
-\section*{Acknowledgments} 
-This work was carried out while the first author visited Teradata. 
-The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
+%\section*{Acknowledgments} 
+%This work was carried out while the first author visited Teradata. 
+%The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
 

### 2017-03-15 mlppi-comsis17/paper.tex
Algorithm rewriting done

@@
 \end{center} 
-
+\vspace{-.1in}
 \noindent In the subsequent phases $M$ gets used for computing costs and updated 
@@
 a corresponding scan cost query ($s$), which can be constructed as follows: 
+\vspace{-.1in}
 \begin{center}
@@
 \end{center} 
+\vspace{-.1in}
 where $F$ is a fact table, and 
@@
 can be built as shown below:
-\vspace{-.08in}
+\vspace{-.05in}
 \begin{center}
@@
 $\Sone$:	& {\tt SELECT * FROM LINEORDER l} \\
-		& {\tt WHERE ((l.LO\_DISCOUNT = 1)} \\
-		& {\tt OR (l.LO\_DISCOUNT >= 4} \\ 
-		& {\tt AND l.LO\_DISCOUNT <= 5))} \\
+		& {\tt WHERE ((l.LO\_DISCOUNT = 1) OR}\\
+		& {\tt (l.LO\_DISCOUNT >= 4 AND l.LO\_DISCOUNT <= 5))} \\
 		& {\tt AND l.LO\_QUANTITY <= 35}  \\ 
@@
         & {\tt WHERE l.LO\_DISCOUNT >= 7} \\
-        & {\tt AND (l.LO\_QUANTITY >= 25 AND} \\ 
-        & {\tt l.LO\_QUANTITY <= 35)}.
-\end{tabular}
-\end{center} 
-\vspace{-.08in}
+        & {\tt AND (l.LO\_QUANTITY >= 25 AND l.LO\_QUANTITY <= 35)}.
+\end{tabular}
+\end{center} 
+\vspace{-.05in}
 \noindent Since the altered ranges (\openrange{\infty, 25} and \range{25, 35}) of $q$1 
@@
 
+\vspace{-0.1in}
 $M$:
@@
 As the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
-contain the non-qualifying rows for $q$1, unfortunately, 
+contain the \hbox{non-qualifying} rows for $q$1, unfortunately, 
 $R$ cannot provide $q$1 with as much benefit as $q$2. 
@@
 the MLPPI derived by $R$ can potentially minimize the total execution cost of $Q$.
+
+Algorithm~\ref{algo:init_phase} represents the initial phase 
+that has been described so far. 
+Given a \linebreak \hbox{query-to-range} set map ($M$) and an input range set ($R$), 
+the algorithm determines which range pair to yield the least total execution cost ($T_{s}$) 
+and updates $M$ and $R$ using the chosen range pair, 
+until the number of the partitions by $R$ falls below a pre-defined partition limit count.
+Finally, the algorithm produces the final recommendation based on $R$.
 
@@
 
-Algorithm~\ref{algo:init_phase} represents the initial phase 
-that has been described so far. 
-Given a \linebreak \hbox{query-to-range} set map ($M$) and an input range set ($R$), 
-the algorithm determines which range pair to yield the least total execution cost ($T_{s}$) 
-and updates $M$ and $R$ using the chosen range pair, 
-until the number of the partitions by $R$ falls below a pre-defined partition limit count.
-Finally, the algorithm produces the final recommendation based on $R$.
-
 In the running example the total partition counts ($4${$\times$}$3$=$12$) by $R$ is under 
@@
 
-But via additional merges we may further make fewer partitions, 
+But we emphasize that additional merges may make fewer partitions, 
 which actually can enhance the overall workload performance for the following reasons. 
@@
 A similar algorithm, as suggested in the initial phase, 
-can be applied to this optimized phase. 
-Instead of the scan cost, the {\em query cost} can be used thanks to a feasible MLPPI. 
-The query cost is modeled as the estimated processing time 
-of a query on a ``faked'' fact table applying the MLPPI definition 
-reflecting the merge of a range pair. 
+can be applied. 
+In this phase, we use {\em query cost}, analogous to scan cost, 
+which can be obtained via a feasible MLPPI. 
+Specifically, the query cost is modeled as the estimated processing time 
+of a query on a {\it faked} fact table applying the MLPPI definition that reflects the merge of a range pair. 
 This cost estimation technique is similar 
@@
 
-Algorithm~\ref{algo:comp_query_cost} illustrates the steps for computing 
-the query cost of a query $q$, 
-given a range pair ($rp$) affecting $q$ when merged. 
-First, we create an empty shadow table $H$ with the same definition as $F$, 
+\begin{algorithm}[t]
+\caption{The Optimized Phase}
+\label{algo:opt_phase}
+\SetKwInOut{Input}{input}
+\SetKwInOut{Output}{output}
+\SetKw{func}{Function}
+\SetKwBlock{begin}{}{end}
+{
+\Input{$M$ (query-to-range-set map), $R$ (input range set)}
+\Output{The MLPPI on which the total query cost is minimal}
+	$W$ $\leftarrow$ Query weights \\
+	$T_{p}$ $\leftarrow$ Total query cost sum on an initial MLPPI by $R$ \\
+	\While{(Pre-defined iterations or $R$ $\neq$ $\varnothing$)}{
+		\ForEach{range pair $rp$ in $R$}{	 
+			$T_{q}$ $\leftarrow$ 0 \;
+			\ForEach{$q$ in $M$}{	
+				$w$ $\leftarrow$ Get $q$'s weight from $W$ \\
+				$L$ $\leftarrow$ Get the range set mapped to $q$ in $M$ \\
+				\uIf(\tcp*[f]{Affected}) 
+				{(($rp$ $\cap$ $L$) $\neq$ $\varnothing$)}{
+					$T_{q}$ += $w${$\cdot$}({\it getQC}($q$,$rp$,$M$,$R$)) 	
+					\tcp*[r]{See Algo.~\ref{algo:comp_query_cost} regarding {\it getQC}()}				
+				}
+				\lElse{ $T_{q}$ += $w${$\cdot$}($q$'s existing query cost)}
+			}
+			Map $T_{q}$ to $rp$.
+		}
+		$T$ $\leftarrow$ The least $T_{q}$ that has been so far seen.  \\
+		\lIf{$T$ $>$ $T_{p}$}{ {\bf break} \;}
+		\Else{
+			$T_{p}$ $\leftarrow$ $T$ \tcp*[r]{Query Cost Sum Update} 
+			Find $rp$(s) with $T$. \\
+			If a tie happens, select the $rp$ to make fewer partitions when consolidated. \\
+			Update $M$ and $R$ by the chosen $rp$. \\
+		}
+	}
+	{\bf return} an MLPPI by $R$ \;
+}
+\end{algorithm}
+
+Algorithm~\ref{algo:comp_query_cost} illustrates the steps for computing the query cost of a query $q$. 
+Given a range pair ($rp$) affecting $q$ when merged, 
+we first create an empty shadow table $H$ with the same definition as $F$, 
 including all indexes and constraints such as check and referential integrity 
@@
 propagate the statistics of the MVs on $F$ to the new MVs on $H$. 
-After that, we alter $H$ by an MLPPI with $R$' applying the merge of $rp$. 
-The wizard then builds a {\em shadow} query $h$ equivalent to the input query $q$ 
-but replacing $F$ in $q$ with $H$, and makes an API call to the {\tt EXPLAIN} tool with $h$. 
-Eventually, the tool extracts from the result and returns the estimated elapsed time 
+After that, we alter $H$ by an MLPPI with the updated $R$ ($R$'), applying the merge of $rp$. 
+The algorithm then builds a {\em shadow} query $h$, equivalent to the input query $q$,
+which is built by replacing $F$ in $q$ with $H$. 
+The algorithm in turn makes an API call to the {\tt EXPLAIN} tool with $h$. 
+Eventually, the algorithm extracts from the result and returns the estimated elapsed time 
 of $h$ as the query cost of $q$. 
@@
 
-For each range pair the wizard computes the weighted sum ($T_{q}$) of query cost of queries. 
-$T_{q}$ can be similarly defined as $T_{s}$, described in Section~\ref{sec:init_phase}.
-%and thus the definition of $T_{q}$ is omitted.
-Let the current least $T_{q}$ be $T$ and the existing least $T_{q}$ be $T_{p}$. 
-If $T$ $<=$ $T_{p}$, then for the next iteration $T_{p}$ is updated to $T$. 
-In turn, the tool merges the range pair that produces $T$ and updates $M$ and $R$ along with the merge. 
-The wizard repeats the process until 1) no ranges in $R$ remain, or 2) pre-defined iterations are reached. 
-The optimized phase is finished if $T$ $\geq$ $T_{p}$.
+Algorithm~\ref{algo:opt_phase} proposes the algorithm of the optimized phase. 
+$W$ indicates a set of weights assigned to individual queries in a given workload. 
+$T_{p}$ keeps track of the existing least query cost sum, and initially is computed on an initial MLPPI. 
+For each range pair, the algorithm 
+computes the respective weighted sum ($T_{q}$) of query costs of the queries in a given workload. 
+($T_{q}$ can be similarly defined as $T_{s}$, described in Section~\ref{sec:init_phase}, 
+and thus the definition of $T_{q}$ is omitted here.)
+Let the current least $T_{q}$ be $T$. 
+If $T$ $<=$ $T_{p}$ (existing least), then for the next iteration $T_{p}$ is updated to $T$. 
+We perform the merge on the range pair that produces that $T$ and update $M$ and $R$ along with the merge. 
+The algorithm repeats the process until (1)~no more ranges in $R$ are left, or (2)~predefined iterations are reached. 
+If $T$ $\geq$ $T_{p}$, then the optimized phase is terminated.
+
+Note that to break a tie between range pairs, 
+we apply the same heuristic to favor one of them that produces fewer partitions. 
+The order of the different partitioning levels may have some impact on 
+the execution time. A range condition on lower levels like the query 
+in Section~\ref{sec:intro} 
+``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}'' 
+requires scanning non-consecutive partitions. 
+This may incur some overhead at run time. 
+To reduce this effect, we sort the partition levels based on the number of 
+partitions in descending order before making the final recommendation. 
+The solution in Figure~\ref{fig:exam} follows this heuristic 
+making the two-partition case in the first level 
+followed by the four-partition case in the second level. 
+The order in Figure~\ref{fig:mlppi} can go \hbox{either} way 
+since both levels have the same number of \hbox{partitions}.
 
@@
 as shown in Figure~\ref{fig:mlppi}.
+
 \vspace{-.25in}
-\begin{figure}[h]
+\begin{figure}[h!]
 \begin{center}
@@
 
-Algorithm~\ref{algo:opt_phase} exhibits the proposed optimized phase, 
-which also applies the same heuristic to favor a range pair 
-that produces fewer partitions to break a tie.  
-The order of the different partitioning levels may have some impact on 
-the execution time. A range condition on lower levels like the query 
-in Section~\ref{sec:intro} 
-``{\tt SELECT * FROM LINEORDER WHERE LO\_QUANTITY < 25}'' 
-requires scanning non-consecutive partitions. 
-This may \hbox{incur} some overhead at run time. 
-To reduce this \hbox{effect}, we sort the partition \hbox{levels} based on number of 
-partitions in descending order \hbox{before} making a final \hbox{recommendation}. 
-The solution in \hbox{Figure}~\ref{fig:exam} follows this heuristic 
-making the two-partition case in the first level 
-followed by the four-partition case in the second level. 
-The order in Figure~\ref{fig:mlppi} can go \hbox{either} way 
-since both levels have the same number of \hbox{partitions}.
-
-\begin{algorithm}[t]
-\caption{The Optimized Phase}
-\label{algo:opt_phase}
-\SetKwInOut{Input}{input}
-\SetKwInOut{Output}{output}
-\SetKw{func}{Function}
-\SetKwBlock{begin}{}{end}
-{
-\Input{$M$ (query-to-range-set map), $R$ (input range set)}
-\Output{MLPPI such that the total query cost is minimal}
-	$W$ $\leftarrow$ Query weights \\
-	$T_{p}$ $\leftarrow$ Total query cost sum on an initial MLPPI by $R$ \\
-	\While{(Pre-defined iterations or $R$ $\neq$ $\varnothing$)}{
-		\ForEach{range pair $rp$ in $R$}{	 
-			$T_{q}$ $\leftarrow$ 0 \;
-			\ForEach{$q$ in $M$}{	
-				$w$ $\leftarrow$ Get $q$'s weight from $W$ \\
-				$L$ $\leftarrow$ Get the range set mapped to $q$ in $M$ \\
-				\uIf(\tcp*[f]{Affected}) 
-				{(($rp$ $\cap$ $L$) $\neq$ $\varnothing$)}{
-					$T_{q}$ += $w${$\cdot$}({\it getQC}($q$,$rp$,$M$,$R$)) 	
-					\tcp*[r]{See Algo.~\ref{algo:comp_query_cost} regarding {\it getQC}()}				
-				}
-				\lElse{ $T_{q}$ += $w${$\cdot$}($q$'s existing query cost)}
-			}
-			Map $T_{q}$ to $rp$.
-		}
-		$T$ $\leftarrow$ The least $T_{q}$ that has been so far seen.  \\
-		\lIf{$T$ $>$ $T_{p}$}{ {\bf break} \;}
-		\Else{
-			$T_{p}$ $\leftarrow$ $T$ \tcp*[r]{Query Cost Sum Update} 
-			Find $rp$(s) with $T$. \\
-			If a tie happens, select the $rp$ to make fewer partitions when consolidated. \\
-			Update $M$ and $R$ by the chosen $rp$. \\
-		}
-	}
-	{\bf return} an MLPPI by $R$ \;
-}
-\end{algorithm}
-
-\vspace{-.1in}
-%\vspace\fill
-
 \section{Analysis}
@@
 made to the server during the phases.
-
-%In this section the complexity of the proposed algorithms 
-%is analyzed.
 
@@
 space is driven by query-predicates, which is superior because 
-\hbox{predicates} are more customized and specific to a workload than fields. 
+predicates are more customized and specific to a workload than fields.
+
 Agrawal's horizontal partitioning scheme~\cite{agrawal04:integrating} 
 also uses the search space driven by simple range predicates, but 
-his technique has a \hbox{shortcoming} as follows. 
+his technique has a shortcoming as follows. 
 His work produces a solution for each individual query and attempt 
 to merge the solutions. But this approach cannot reach 
-an optimized solution in a global \hbox{perspective}.
-Our wizard generates the whole search space upfront and in turn merges partitions, leading to a globally-optimized \hbox{solution}. 
-Moreover, only a single column is considered in his work~\cite{agrawal04:integrating}, 
-whereas our tool deals with multiple fields.
+an optimized solution in a global perspective.
+Our wizard generates the whole search space upfront and in turn merges partitions, leading to a \hbox{globally-optimized} solution. 
+While his work considers a single column only, we deal with multiple fields.
 
@@
 releases that add complexity and additional cost for software upgrades. 
-The Teradata optimizer has a rich set of existing APIs originally coded 
-for system and workload management tools. 
-The APIs are sufficient to avoid the costly optimizer code change.
+But our algorithm is based existing APIs supported by an optimizer (like Teradata), 
+which requires no code change in the optimizer.
+%The Teradata optimizer with a rich set of existing APIs originally coded 
+%for system and workload management tools. 
+%The APIs are sufficient to avoid the costly optimizer code change.
 

### 2017-03-16 mlppi-comsis17/paper.tex
Experiment Setup rewritten

@@
 In this section we analyze the time complexity of 
-the proposed preprocessing, initial, and optimized phases.
-For the time complexity we compute the logical running time of the wizard, using the {\em number of API calls} 
-made to the server during the phases.
+the proposed algorithms: preprocessing, initial, and optimized phases. 
+The metric of this complexity analysis on each phase is the logical running time of the wizard, 
+equivalent to the number of API calls made to the database server during that phase.
 
@@
 
-{\bf Proof}. Let us consider a worst case such that given a workload of 
+{\bf Proof}. Consider the worst case such that given a workload of 
 $N$ queries, each query references all the fields of the fact table, and 
@@
 Each query may have a single-value range. 
-A range consisting of only a single value per field can be mapped to each query. 
+A range consisting of only one value per field can be mapped to each query. 
 Thus, the running time complexity for the map construction 
@@
 
-But we empirically observed that the bound of $\BigO$($N${$\cdot$}$C${$\cdot$}$V$) 
+But note that in our experiments the bound of $\BigO$($N${$\cdot$}$C${$\cdot$}$V$) 
 was overly pessimistic.
@@
 \label{lemma:init_phase}  
-The running time complexity for the initial or optimized phases is 
+The running time complexity for the initial (or optimized) phase is 
 $\BigO$($M^{2}${$\cdot$}$N$), 
@@
 {\bf Proof}. Suppose that the merge of every range pair influences all the queries. 
-Every iteration, either the scan or query costs need to be recomputed 
-for each query. 
-In the first round the \hbox{re-computation} cost is paid as many times as $M${$\cdot$}$N$, 
-and a chosen range pair is merged. In the subsequent round 
-$M$-1 range pairs remains, and the following cost amounts to ($M$-1){$\cdot$}$N$. 
-In the worst case we may end up consolidating all range pairs, thus having no partition on 
-a target table. The total calls to the server can be increased up to
+Every iteration, either the scan or query costs need to be recomputed for each query. 
+In the first round we pay the \hbox{re-computation} cost of $M${$\cdot$}$N$ 
+and merge a victim range pair. In the subsequent round, $M$-1 range pairs remain, 
+so the cost amounts to ($M$-1){$\cdot$}$N$. 
+In the worst case we end up consolidating all range pairs, thus having no partition on 
+the target table. The total calls made to the server can be added up to
 
@@
 was overly pessimistic. Specifically, 
-the number of calls made in our experiments was much fewer than that bound, as 
-it was almost rare for a range pair to involve all queries in a workload 
+the number of calls made in our experiments was by far fewer than that bound. 
+That was because it was very rare for a range pair to involve all queries in a given workload 
 (as seen in Table~\ref{tab:workload_stat}).
 
-In the following, we evaluate the proposed algorithm.
+%In the following we evaluate the performance of our wizard using the proposed algorithms.
 
@@
 
-In this section, we describe our environment settings, 
-report the statistics measured in our experiments, and 
-demonstrate the performance of the MLPPI wizard (WIZARD), compared with 
-that of no partitioning and the partitioning by a human expert (EXP).
+In this section we evaluate the performance of our wizard using the proposed algorithms.
+We first describe our environment settings. 
+We then report the statistics measured during our experiments. 
+Finally, we compare the performance of the recommendation 
+by the MLPPI wizard with (1)~that of no partitioning 
+and (2)~the partitioning of a human expert (EXP) under different setups.
 
@@
 
-The MLPPI wizard was implemented as a prototype 
-on top of the Teradata DBMS server. 
-It was written in Java. 
-The performance of the wizard was evaluated 
+\paragraph{Development:} 
+We implemented our MLPPI wizard as a prototype on top of the \hbox{Teradata} DBMS server. 
+The wizard was written in Java. 
+We evaluated the performance of the \hbox{wizard} 
 on the Teradata DBMS server machine running Unix.
 
-When it comes to workload generation, we took advantage of 
+\paragraph{Workload:} When it comes to workload generation, we took advantage of 
 our simple star schema query \hbox{generator}. 
-The generator assumes that 1) available operators, 2) fields, 
-and 3) the minimum and \hbox{maximum} values of the fields are already known. 
+The generator assumes that (i)~available operators, (ii)~fields, 
+and (iii)~the minimum and \hbox{maximum} values of the fields are already known. 
 For each query, the generator first randomly selects the number of single-table 
 predicates to create. 
-Then, it arbitrarily chooses a field and a specific operator 
+The generator then arbitrarily chooses a field and a specific operator 
 for each predicate. 
@@
 range of the field. 
-In this way the generator builds a query involving the generated predicates.
-
-A generated query is based on a template that joins {\tt LINEORDER} and {\tt DDATE} with constraints defined by the predicates. 
+In this way the generator makes a query involving the generated predicates.
+
+A generated query is based on a template that joins {\tt LINEORDER} and 
+{\tt DDATE} with constraints defined by the predicates. 
 This template is common in customer cases like reports and form 
 templates. 
-Using the query generator we generated two workloads consisting of 10 queries (10Q) and 20 queries (20Q).
-
-For the query workloads we populated the fact table with a scale of 1TByte (1TB) and 3TBytes (3TB).
+Using the query generator we generated two workloads consisting of 10 queries~(10Q) and 20 queries~(20Q).
+
+For the query workloads we populated the fact table with a scale factor of 1TByte~(1TB) and 3TBytes~(3TB).
 

### 2017-03-17 mlppi-comsis17/paper.tex
Update on stats

@@
 Finally, we compare the performance of the recommendation 
-by the MLPPI wizard with (1)~that of no partitioning 
+by the MLPPI wizard (WIZARD) with (1)~that of no partitioning (NO PPI)
 and (2)~the partitioning of a human expert (EXP) under different setups.
@@
 \begin{center}
-{\scriptsize
-\begin{tabular}{|l|l|l|l|l|} \hline
-		   			   							   & \multicolumn{2}{|c|}{Init. Phase} & \multicolumn{2}{|c|}{Opt. Phase} \\ \cline{2-5}
-		   			   							   & 10Q & 20Q & 10Q & 20Q \\ \hline
-{\# Input Partitions}  							   & 1,008,000 & 110,739,200 & 51,840 & 59,520 \\ \hline
-{\# Input Range Pairs} 							   & 441 & 4,180 & 24 & 48\\ \hline
-{\# Total Iterations}  							   & 14 & 55 & 1 & 1 \\ \hline
-{\# Total API Calls (Worst Calls by Lemma~\ref{lemma:init_phase}})   & 1,305 (1,944,810) & 31,915 (349,448,000) & 78 (5,760) & 388 (46,080) \\ \hline
-{\# Average API calls per Range Pair} 		  	   & 2.96 & 7.63 & 3.25 & 8.08 \\ \hline
-{\# Average number of range pairs compared per iteration} 	   & 31.5 & 76.0 & 24 & 48 \\ \hline
-{\# Maximum number of range pairs compared per an iteration} & 38 & 103 & 24 & 48 \\ \hline
-{\# Minimum number of range pairs compared per an iteration} & 25 & 49 & N/A & N/A\\ \hline
+{\scriptsize %
+\begin{tabular}{|c|l|l|l|l|l|} \hline
+\multicolumn{2}{|c|}{\multirow{2}{*}{Item}} & \multicolumn{2}{|c|}{The Initial Phase} & \multicolumn{2}{|c|}{The Optimized Phase} \\ \cline{3-6}
+ \multicolumn{2}{|c|}{} & 10Q & 20Q & 10Q & 20Q \\ \hline
+1 & {\# of input partitions}  	 & 1,008,000 & 110,739,200 & 51,840 & 59,520 \\ \hline
+2 & {\# of input range pairs} 	 & 441 & 4,180 & 24 & 48\\ \hline
+ % {\# of total iterations}   & 14 & 55 & 1 & 1 \\ \hline
+3 & {\# of total API calls (\# of worst calls by Lemma~\ref{lemma:init_phase}})   & 1,305 (1,944,810) & 31,915 (349,448,000) & 78 (5,760) & 388 (46,080) \\ \hline
+4 & {\# of average API calls per range pair} 	& 3 & 8 & 3 & 8 \\ \hline
+5 & {\# of average range pairs compared per iteration} 	   & 32 & 76 & 24 & 48 \\ \hline
+6 & {The maximum \# of range pairs compared per an iteration} & 38 & 103 & 24 & 48 \\ \hline
+7 &{The minimum \# of range pairs compared per an iteration} & 25 & 49 & 24 & 48\\ \hline
 %{\# max. API Calls made per range pair} 		   & 4.92 & 16.14 & & \\ \hline
 %{\# min. API Calls made per range pair} 		   & 1.92 & & & \\ \hline
-{\# Average API calls made per iteration}		   & 93.21 & 580 & 78 & 388 \\ \hline
-{\# Maximum number of API calls made per an iteration} 	   & 123 & 791 & 78 & 388 \\ \hline
-{\# Minimum number of API calls made per an iteration} 	   & 73 & 381 & N/A & N/A \\ \hline
+8 & {\# of average API calls made per iteration}		   & 93 & 580 & 78 & 388 \\ \hline
+9 & {The maximum \# of API calls made per an iteration} 	   & 123 & 791 & 78 & 388 \\ \hline
+10 & {The minimum \# of API calls made per an iteration} 	   & 73 & 381 & 78 & 388 \\ \hline
 \end{tabular}
@@
 
-Table~\ref{tab:workload_stat} summarizes the WIZARD's execution statistics, 
+Table~\ref{tab:workload_stat} summarizes the execution statistics of 
 which were obtained while the recommendations for the \hbox{workloads} were produced. 
@@
 
-About 1 million partitions were initially derived for 10Q whereas roughly 0.11 billion partitions for 20Q. 
-Although the workload size doubled, the total number of partitions was increased exponentially. 
-Much more predicates with different bindings produced two orders of magnitude more ranges in 20Q than those of 10Q. 
-While the huge input partitions were made, the number of collected range pairs per field was relatively small.
-
-The total number of iterations in the initial phases 
-tended to be proportional to the input \hbox{partitions} 
-\hbox{generated} from the workloads. 
-Note that there \hbox{happened} only one iteration in the optimized phase. 
-The reason was that as the partitions produced by the \hbox{initial} phase were customized enough, 
-no more \hbox{iterations} were \hbox{necessary} for further optimization.
-
-As mentioned in Section~\ref{sec:analysis}, the total calls made to 
-the server per workload were by far fewer than the theoretical bound shown in parentheses. 
+About 1 million partitions were initially derived for 10Q, and roughly 110 million partitions for 20Q. 
+Although the workload size doubled from 10Q to 20Q, the total number of partitions increased exponentially. 
+Much more predicates with different bindings produced about 110x more ranges in 20Q than those of 10Q. 
+While a huge number of input partitions were made, the number of range pairs collected per field was relatively small.
+
+The total number of iterations in the initial phase 
+tended to be proportional to the input partitions 
+generated from the 10Q and 20Q workloads. 
+Note that the optimized phase ran only once for those two workloads.
+Since the partitions produced by the initial phase were customized enough,
+no more iterations were necessary for further optimization.
+
+As mentioned in Section~\ref{sec:analysis}, for each workload the number of the total calls made to 
+the database server were extremely small, compared to that of the theoretical worst calls shown in the parentheses. 
 The worst call counts were calculated regarding the number of queries and range pairs observed in each phase. 
-The average number of calls per range pair was limited within 10. %across workloads and phases.
+Furthermore, the average number of calls per range pair (was limited within 10 across the workloads and phases.
 
